### 2025-03-01 paper/paper.tex
move paper here

%%
%% This is file `sample-sigplan.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `all,proceedings,bibtex,sigplan')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigplan.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[sigplan,screen]{acmart}%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

\newif\ifdraft\drafttrue{}
\newif\iflater\latertrue{}

\PassOptionsToPackage{names,dvipsnames}{xcolor}

\settopmatter{printfolios=false,printccs=false,printacmref=false}
\setcopyright{none}

\usepackage{listings}
\usepackage{subcaption}

\definecolor{dkred}{rgb}{0.7,0,0}
\definecolor{dkblue}{rgb}{0,0,0.7}
\definecolor{notered}{rgb}{0.85,0,0}
\definecolor{dkpurple}{HTML}{4e02eb}
\definecolor{dkgreen}{HTML}{006329}
\definecolor{teal}{HTML}{007982}
\definecolor{fuchsia}{HTML}{8C368C}

\newcommand{\comm}[3]{\ifdraft\textcolor{#1}{[#2: #3]}\fi}
\newcommand{\cn}{\ifdraft{\color{blue} [CITE]}\fi}
\newcommand{\hg}[1]{\comm{dkred}{HG}{#1}}
\newcommand{\tr}[1]{\comm{dkblue}{TR}{#1}}
\newcommand{\bcp}[1]{\comm{dkpurple}{BCP}{#1}}
\newcommand{\jwc}[1]{\comm{dkgreen}{JWC}{#1}}
\newcommand{\outline}[1]{\ifdraft{\color{teal} #1}\fi}

\newcommand{\bq}{\texttt{base\_quickcheck}}
\newcommand{\name}{\texttt{waffle\_house}}
\newcommand{\tool}{Waffle House}

\newcommand{\secref}[1]{\S\ref{#1}}

\newcommand{\proposecut}[1]{\ifdraft{\color{gray} [#1]}\fi}
\newcommand{\proposechange}[2]{\ifdraft{\color{gray} {#1}
    $\rightarrow$ #2}\else #2\fi}

\theoremstyle{definition}
\newtheorem{definition}{Definition}

\newtheorem{claim}{Claim}

\AtBeginDocument{%
  \providecommand\BibTeX{{%
    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmlicensed}
\copyrightyear{2025}
\acmYear{2025}
\acmDOI{XXXXXXX.XXXXXXX}
%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[Conference acronym 'XX]{Make sure to enter the correct
  conference title from your rights confirmation email}{June 03--05,
  2018}{Woodstock, NY}
%%
%%  Uncomment \acmBooktitle if the title of the proceedings is different
%%  from ``Proceedings of ...''!
%%
%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
%%  June 03--05, 2018, Woodstock, NY}
\acmISBN{978-1-4503-XXXX-X/2025/02}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% For managing citations, it is recommended to use bibliography
%% files in BibTeX format.
%%
%% You can then either use BibTeX with the ACM-Reference-Format style,
%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
%% support for advanced citation of software artefact from the
%% biblatex-software package, also separately available on CTAN.
%%
%% Look at the sample-*-biblatex.tex files for templates showcasing
%% the biblatex styles.
%%

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Fail Faster}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Cynthia Richey}
\affiliation{%
  \institution{University of Pennsylvania}
  \city{Philadelphia}
  \state{Pennsylvania}
  \country{USA}
}

\author{Joseph W. Cutler}
\affiliation{%
  \institution{University of Pennsylvania}
  \city{Philadelphia}
  \state{Pennsylvania}
  \country{USA}}

\author{Harrison Goldstein}
\affiliation{%
  \institution{University of Maryland}
  \city{College Park}
  \state{Maryland}
  \country{USA}
}

\author{Benjamin C. Pierce}
\affiliation{%
 \institution{University of Pennsylvania}
 \city{Philadelphia}
 \state{Pennsylvania}
 \country{USA}}
%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
\renewcommand{\shortauthors}{Richey et al.}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract} \tr{Abstract needs a lot of work.}
Property-based testing (PBT) is a software testing approach that verifies a system against a large suite of automatically-generated inputs. Since testing pipelines typically run under strict time constraints, PBT generators must produce inputs as quickly as possible to maximize the likelihood of finding bugs in the time available. However, existing PBT libraries often prioritize generality at the cost of performance. We introduce \texttt{waffle\_house}, a high-performance generator library that uses staging, a lightweight compilation technique, to eliminate many common generator abstractions. To evaluate \texttt{waffle\_house}, we design a novel benchmarking methodology that compares generators based on program equivalence, isolating performance improvements from differences in input distribution. Using this methodology, we compare \texttt{waffle\_house} to a leading generator library, and, through extensive evaluation over a diverse range of generators, demonstrate that \texttt{waffle\_house} significantly improves generation speed and resource efficiency while matching \texttt{base\_quickcheck}'s expressiveness.
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%

\begin{CCSXML}
<ccs2012>
   <concept>
       <concept_id>10011007</concept_id>
       <concept_desc>Software and its engineering</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
   <concept>
       <concept_id>10011007.10011006</concept_id>
       <concept_desc>Software and its engineering~Software notations and tools</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
   <concept>
       <concept_id>10011007.10011006.10011072</concept_id>
       <concept_desc>Software and its engineering~Software libraries and repositories</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
   <concept>
       <concept_id>10011007.10011006.10011041.10011046</concept_id>
       <concept_desc>Software and its engineering~Translator writing systems and compiler generators</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
 </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering}
\ccsdesc[500]{Software and its engineering~Software notations and tools}
\ccsdesc[500]{Software and its engineering~Software libraries and repositories}
\ccsdesc[500]{Software and its engineering~Translator writing systems and compiler generators}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{Property-based testing, Generators, Staging, Meta-programming}
%% A "teaser" image appears between the author and affiliation
%% information and the body of the document, and typically spans the
%% page.

\received{--}
\received[revised]{--}
\received[accepted]{--}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Introduction}
Property-based testing (PBT) is a widely-used testing framework consisting of
two key components: a set of \textit{properties} that a system must satisfy, and
a large number of \textit{inputs} to that system. In contrast to traditional
unit testing, where inputs are hand-written, PBT users create a random-input
\textit{generator} that automatically produces inputs. Sometimes, this process
can itself be automated through the use of PBT libraries that synthesize
generators from type definitions; however, when inputs are required to maintain
invariants not communicated by the type, users must write a custom generator (or
filter out a broad class of invalid inputs).

Writing ``good" generators—those that uncover more bugs in the system under
test—is challenging. Much effort has been dedicated to this problem, often by
developing sophisticated domain-specific languages for specifying constrained
generators. However, these approaches do not always reflect the needs of real
PBT users. Under ideal conditions, a PBT suite would continue to run
until it stopped finding bugs, real-world usage patterns are very different; a
study of expert PBT users reported time budgets of between 50 milliseconds and
30 seconds for their test suites~\cite{inpractice}. Consequently, generation
time is a significant factor in PBT efficacy---that is, if a generator can
produce inputs twice as fast, it has twice as many chances within a given time
bound to find a bug. Therefore, it is imperative that PBT libraries are built
with efficiency in mind.

To better understand and address performance challenges in the landscape of PBT
generators, we explore popular libraries across several widely-used programming
languages. We find that the flexibility of these libraries comes at a cost to 
performance,
introducing abstraction overhead, frequent boxing and unboxing, and avoidable
allocations (which lead to costly garbage collection pauses). We focus primarily 
on OCaml's \bq, selected for its efficiency relative to other OCaml PBT libraries 
and its integration with industrial build systems, where it is expected to run 
within tight time constraints. These
inefficiencies raise a fundamental question: how can PBT libraries produce code
that is both efficient \textit{and} general? By analyzing
\bq\ and similar libraries in other languages \tr{Rust?
More?}, we provide a broader perspective on common performance challenges in PBT
and explore potential strategies for more efficient generator implementations.

Our solution is \name, a generator library that preserves
\bq’s functionality while improving performance through
\textit{staged metaprogramming} (staging). Staging is a lightweight,
domain-specific compilation technique that allows us to completely eliminate the
runtime overhead of many common generator abstractions. For instance, in
\name, monadic operations are zero-cost.
\name’s staged eliminates unnecessary allocations,
generates static control structures, and leverages unboxed integer libraries to
achieve performance improvements over \texttt{base\_quickcheck}, making it
well-suited to real-world PBT applications.

We present a new methodology for evaluating the relative effectiveness of PBT
generators. Prior work has compared generators based on bug-finding ability: 
generators are considered equally effective if they find bugs in a
system at the same rate over a large number of trials ~\cite{etna}. However,
because \name\ preserves the semantics of
\bq, a \name\ generator produces exactly
the same inputs as its \bq\ counterpart. By comparing
generators based on \textit{program equality}, we ensure that any speedups in
bug-finding ability stem from performance enhancements rather than variations in
input distribution. This approach establishes a foundation for benchmarking
optimizations that enhance performance without compromising expressiveness.

Finally, we conduct an extensive evaluation, implementing both type-derived and
custom generators in \name\ and \bq. Our
generators produce a diverse range of inputs, including recursive data
structures, lambda calculus terms, and regular expressions. We benchmark
generation time, resource usage, and time-to-failure for each pair of staged and
unstaged generators, demonstrating that \name\ achieves
significant gains in speed and resource.

In summary, we dramatically improve the bug-finding effectiveness of PBT generators
by optimizing them for speed. In speicific, we make the following contributions:
\begin{enumerate}
    \item An empirical analysis of the sources of inefficiency in PBT
    generators. 
    \item A library, \name, which offers efficient generator functions through the
    use of staged metaprogramming; this is the first known application of metaprogramming to PBT.
    \item The insight that generators should be compared by program equality 
    \item An evaluation demonstrating the improved performance of generators constructed
    using our library in a controlled comparison.

\end{enumerate}

\section{Background}
\subsection{Staged Metaprogramming}
\subsection{Property-Based Testing}
\subsection{Existing PBT Generators}

\tr{Here's where we talk about what's going on in the world, and ultimately make the argument that \bq\ is the best tool to reproduce.}

\subsection{\bq}

\tr{Discuss BQ, including type-derived generators.}

\section{Why is \bq\ slow, and how can we fix it?}

\jwc{
  \begin{itemize}
    \item How does BQ work?
    \begin{itemize}
      \item The basic generator type: \texttt{'a generator = size:int -> random:SR.t -> 'a}. Size and random seed to deterministic value.
      \item This gets a monad intance in the obvious way
    \end{itemize}
  \end{itemize}
}

\tr{Each of these consist of an explanation of the problem and pseudocode outlining the solution.}
\subsubsection{Monadic bind}

\jwc{
  \begin{itemize}
    \item Naked binds like \texttt{Bind}
  \end{itemize}
}

\subsubsection{Weighted unions}
% \subsubsection{Function call overhead}
\subsubsection{Boxed integers}

\section{\name: A Library for Staging Generators}

\tr{Now, we formally introduce the library, describe all its functionality, how things work, etc.}

\jwc{Note: tradeoff between maintaining the effect ordering and even more performance. }

\jwc{Note with bind that we don't split because it's imperatively updated. Note that this omits the need to use SR, but we use it to maintain backwards compat.}

\subsection{Comparing Generators}
\subsubsection{Existing Approaches}
\subsubsection{Equality up to Program Equality}
\subsubsection{Effect Ordering}
\subsubsection{PBT for PBT}
\subsubsection{Does maintaining program equality in \name\ affect performance?}
\section{Evaluation}
\subsubsection{Implementation}
\subsection{Benchmarking speed \& resource usage}
\subsection{Impact on bug-finding ability}
\section{Conclusion \& Future Work}
%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{bib}

%%
%% If your work has an appendix, this is the place to put it.
\end{document}
\endinput
%%
%% End of file `sample-sigplan.tex'.

### 2025-03-01 paper/paper.tex
changes

@@
 %% IMPORTANT NOTICE:
+
 %% 
@@
 domain-specific compilation technique that allows us to completely eliminate the
-runtime overhead of many common generator abstractions. For instance, in
+runtime overhead of many common generator abstractions. \jwc{This relies on the fact that the code of generators is known statically at compile time} For instance, in
 \name, monadic operations are zero-cost.
@@
 \subsection{Staged Metaprogramming}
+
+\jwc{A good quick presentation of staged metaprogramming is here, in section 6.
+https://www.cl.cam.ac.uk/~jdy22/papers/a-typed-algebraic-approach-to-parsing.pdf
+}
+
 \subsection{Property-Based Testing}
-\subsection{Existing PBT Generators}
+\subsection{Other Languages and Libraries}
 
-\tr{Here's where we talk about what's going on in the world, and ultimately make the argument that \bq\ is the best tool to reproduce.}
+\tr{Here's where we talk about what's going on in the world, and ultimately make the argument that \bq{} is the best tool to reproduce.}
+
+\jwc{Note that this explanation require some prior note of exactly *why* BQ is slow... i.e. the abstraction overhead of the library is high.}
+
+\jwc{
+  \begin{itemize}
+    \item Scala.
+Functional abstractions like QC generators are known to be costly in Scala, that's why they have LMS.
+Example: parser combinators (``On Staged Parser Combinators for Efficient Data Processing''), functional data structures (\href{https://ppl.stanford.edu/papers/popl13_rompf.pdf}{Link}), web programming (``Efficient High-Level Abstractions for Web Programming'').
+    % \item Haskell: GHC does a lot of these optimizations already, since the code is pure. Since QC generators are relatively small programs,
+    % GHC has little trouble specializing them. Of course, this is not guaranteed. A version of this idea can easily be ported to the original QC with template haskell, to guarantee
+    % the highest-performance generators.
+    % \item Rust: Not GC'd, so no alloc overhead but bind'd generators still dispatch through runtime data.
+  \end{itemize}
+}
 
@@
       \item The basic generator type: \texttt{'a generator = size:int -> random:SR.t -> 'a}. Size and random seed to deterministic value.
-      \item This gets a monad intance in the obvious way
+      \item Also two inverse functions, \texttt{create}, and \texttt{generate}.
+      \item This gets a monad intance in the obvious way (show code).
     \end{itemize}
+    \item Note that \emph{extensionally} \texttt{generate (create (fun ~size ~random -> e)) ~size ~random = e}, but the OCaml compiler does not always perform this optimization, or do the inlining required to expose it.
+    (When sufficiently obfuscated behind returns and binds ...) This program compiles to code that (1) allocates the closure for `e', (2) passes it to create (which returns the closure), and then calls (3) generate, which immediately jumps into the closure.
   \end{itemize}
 }
+
+In OCaml, garbage collection is the killer.
+
+This generator generates a list of bools. To generate a list of 100 bools, it allocates \texttt{4808.00w (minor)}, and runs for an average of
+3.9us. Merely representing a list of bools of length 100 requires 300 words in OCaml, so this generator allocates 16 times more than the theoretical minimum.
+
+\begin{lstlisting}
+let gen_list_bool =
+  fixed_point (fun gl ->
+    let%bind n = size in
+    if n <= 0 then return [] else
+      let%bind x = bool in
+      let%bind xs = with_size ~size:(n-1) gl in
+      return (x :: xs)
+  )
+\end{lstlisting}
+
+On the other hand, this generator allocates only 303w, and runs for 0.7us, to generate the same list of 100 bools.
+\begin{lstlisting}
+  let rec gen_list_bool_faster ~size ~random  = 
+    if size <= 0 then []
+    else (Unboxed_splitmix.bool random)::(gen_list_bool_faster ~size:(size - 1) ~random)
+\end{lstlisting}
 
@@
 \section{\name: A Library for Staging Generators}
+
+\jwc{
+  \begin{itemize}
+    \item To eliminate this runtime overhead and generate efficient code, we do a usual ``binding-time analysis''.
+    \item The crux of this binding-time analysis is that while generators are `just' values of type \texttt{'a t}, in practice they are always \emph{completely statically defined}: one never
+    computes a generator dynamically at test time.
+  \end{itemize}
+}
+
+\jwc{Here!}
+
+\begin{lstlisting}
+type 'a gen = int code -> Random.t code -> 'a
+\end{lstlisting}
 

### 2025-03-03 handwritten-rust/src/main.rs
main rs

@@ -1,7 +1,36 @@
 
 use std::collections::LinkedList;
-use bolero_generator::{driver::*, gen, one_of, combinator};
+use bolero::TypeGenerator;
+use bolero_generator::{driver::*, gen, one_of, ValueGenerator};
 
+#[derive(Clone)]
+enum LL {
+    Nil,
+    Cons(u32,Box<LL>)
+}
+
+
+impl ValueGenerator for LL {
+    type Output = LL;
+
+    fn generate<D : Driver>(&self, driver : &mut D) -> Option<LL> {
+        Some(self.clone())
+    }
+
+}
+
+fn g<D : Driver>(sz : u32) -> dyn TypeGenerator {
+    let g = gen::<bool>().and_then_gen(|b|
+        if(sz == 0 || b) {
+            LL::Nil
+        } else {
+            g(sz / 2).and_then_gen(|l|
+                LL::Nil
+            )
+        }
+    );
+    return g.generate(driver);
+}
 
 fn main() {
     let mut driver = ByteSliceDriver::new(&[0xF,0xA,3,4,6,7,8,9], &Options::default());

### 2025-03-03 paper/paper.tex
switch to acmsmall

@@
-%%
-%% This is file `sample-sigplan.tex',
-%% generated with the docstrip utility.
-%%
-%% The original source files were:
-%%
-%% samples.dtx  (with options: `all,proceedings,bibtex,sigplan')
-%% 
-%% IMPORTANT NOTICE:
-
-%% 
-%% For the copyright see the source file.
-%% 
-%% Any modified versions of this file must be renamed
-%% with new filenames distinct from sample-sigplan.tex.
-%% 
-%% For distribution of the original source see the terms
-%% for copying and modification in the file samples.dtx.
-%% 
-%% This generated file may be distributed as long as the
-%% original source files, as listed above, are part of the
-%% same distribution. (The sources need not necessarily be
-%% in the same archive or directory.)
-%%
-%%
-%% Commands for TeXCount
-%TC:macro \cite [option:text,text]
-%TC:macro \citep [option:text,text]
-%TC:macro \citet [option:text,text]
-%TC:envir table 0 1
-%TC:envir table* 0 1
-%TC:envir tabular [ignore] word
-%TC:envir displaymath 0 word
-%TC:envir math 0 word
-%TC:envir comment 0 0
-%%
-%% The first command in your LaTeX source must be the \documentclass
-%% command.
-%%
-%% For submission and review of your manuscript please change the
-%% command to \documentclass[manuscript, screen, review]{acmart}.
-%%
-%% When submitting camera ready or to TAPS, please change the command
-%% to \documentclass[sigconf]{acmart} or whichever template is required
-%% for your publication.
-%%
-%%
-\documentclass[sigplan,screen]{acmart}%%
+\documentclass[sigplan,screen,acmsmall]{acmart}%%
 %% \BibTeX command to typeset BibTeX logo in the docs
@@
     \item Scala.
-Functional abstractions like QC generators are known to be costly in Scala, that's why they have LMS.
+Functional abstractions like QC generators are known to be costly in Scala, that's why they have LMS (in Scala 2, and Macros in Scala 3).
 Example: parser combinators (``On Staged Parser Combinators for Efficient Data Processing''), functional data structures (\href{https://ppl.stanford.edu/papers/popl13_rompf.pdf}{Link}), web programming (``Efficient High-Level Abstractions for Web Programming'').
-    % \item Haskell: GHC does a lot of these optimizations already, since the code is pure. Since QC generators are relatively small programs,
-    % GHC has little trouble specializing them. Of course, this is not guaranteed. A version of this idea can easily be ported to the original QC with template haskell, to guarantee
-    % the highest-performance generators.
-    % \item Rust: Not GC'd, so no alloc overhead but bind'd generators still dispatch through runtime data.
+    Al of this should still work in scala.
+Could easily be incorporated into ScalaCheck, with minor modification: ScalaCheck uses a state monad to thread around the seed, instead of a stateful one (like BQ), or a splittable one (like Haskell).
+So you have to adapt to that.
+But same diff.
+    \item Haskell: GHC does a lot of these optimizations already, since the code is pure. Since QC generators are relatively small programs,
+    GHC has little trouble specializing them. Of course, this is not guaranteed. A version of this idea can easily be ported to the original QC with template haskell, to guarantee
+    the highest-performance generators.
+    \item Rust: Not GC'd, so no alloc overhead but bind'd generators still dispatch through runtime data.
   \end{itemize}

### 2025-03-03 staged-scala/src/main/scala/Main.scala
staged scala impl

@@ -1,5 +1,29 @@
+import scala.quoted.*
+
+import org.scalacheck.Gen
+import org.scalacheck.rng.Seed
+
+import stagedgen.StGen
+
+def fooImpl(using Quotes) = {
+  val e = StGen.splat(StGen.pure('{55}))
+  println(s"Generated code: ${e.show}")
+  e
+}
+
+inline def foo() = {
+  ${fooImpl}
+}
+
+def sample[T](f : Gen.Parameters => Seed => Option[T]) = {
+  f(Gen.Parameters.default)(Seed.random())
+}
+
+
+def msg = "I was compiled by Scala 3. :)"
+
 @main def hello(): Unit =
   println("Hello world!")
   println(msg)
 
-def msg = "I was compiled by Scala 3. :)"
+

### 2025-03-03 staged-scala/src/main/scala/benchmark/GenBm.scala

@@ -1,0 +1,33 @@
+package benchmark
+
+import java.util.concurrent.TimeUnit
+
+import scala.quoted.*
+import org.openjdk.jmh.annotations._
+import org.scalacheck.Gen
+import org.scalacheck.rng.*
+import stagedgen.StGen
+
+@BenchmarkMode(Array(Mode.AverageTime))
+@OutputTimeUnit(TimeUnit.NANOSECONDS)
+@Warmup(iterations = 2, time = 1, timeUnit = TimeUnit.SECONDS)
+@Measurement(iterations = 3, time = 1, timeUnit = TimeUnit.SECONDS)
+@Fork(1)
+@State(Scope.Thread)
+
+  def complexGen: Gen[(Long,Long)] = for {
+    x <- Gen.choose(1,1000)
+    y <- Gen.choose(0,x)
+  } yield (x,y)
+
+class GenBm {
+@Benchmark
+  def generateComplexBase(): Option[(Long,Long)] = {
+    complexGen.sample
+  }
+
+@Benchmark
+  def generateComplexSt(): Option[(Long,Long)] = {
+    StGen.complexStGen(Gen.Parameters.default)(Seed.random())
+  }
+}

### 2025-03-03 staged-scala/src/main/scala/stagedgen/Cps.scala

@@ -1,0 +1,51 @@
+package stagedgen
+
+import scala.quoted.*
+import org.scalacheck.rng.Seed
+
+abstract class Cps[T] { self =>
+  def apply[Z : Type](cont: T => Expr[Z]): Expr[Z]
+
+  def map[S](f : T => S) : Cps[S] = {
+    new Cps {
+      def apply[Z : Type](cont: S => Expr[Z]): Expr[Z] = {
+        self.apply((x : T) => cont(f(x)))
+      }
+    }
+  }
+
+  def flatMap[S](f : T => Cps[S]) : Cps[S] = {
+    new Cps {
+      // I think this is correct?
+      def apply[Z : Type](cont: S => Expr[Z]): Expr[Z] = {
+        self.apply((x : T) => f(x).apply(cont))
+      }
+    }
+  }
+}
+
+object Cps {
+  def pure[T](t: T): Cps[T] = new Cps[T] {
+    def apply[Z : Type](cont: T => Expr[Z]): Expr[Z] = cont(t)
+  }
+
+  def run[T : Type](c : Cps[Expr[T]]) : Expr[T] = {
+    c.apply(x => x)
+  }
+
+  def letInsert[T : Type](ex : Expr[T])(using Quotes) : Cps[Expr[T]] = new Cps[Expr[T]] {
+    def apply[Z : Type](cont: Expr[T] => Expr[Z]): Expr[Z] =
+      '{
+        val x = ${ex}
+        ${cont('{x})}
+      }
+  }
+
+  def splitPair[S: Type, T : Type](ex : Expr[(S,T)])(using Quotes) : Cps[(Expr[S],Expr[T])] = new Cps[(Expr[S],Expr[T])] {
+    def apply[Z : Type](cont: ((Expr[S],Expr[T])) => Expr[Z]): Expr[Z] =
+      '{
+        val (x,y) = ${ex}
+        ${cont('{x},'{y})}
+      }
+  }
+}

### 2025-03-03 staged-scala/src/main/scala/stagedgen/StGen.scala

@@ -1,0 +1,124 @@
+package stagedgen
+
+import org.scalacheck.Gen
+import org.scalacheck.rng.Seed
+import scala.quoted.*
+import stagedgen.Cps
+
+class IllegalBoundsError[A](low: A, high: A)
+    extends IllegalArgumentException(s"invalid bounds: low=$low, high=$high")
+
+private def chLng(l: Long, h: Long)(p: Gen.Parameters, seed: Seed): (Long,Seed) = {
+      if (h < l) {
+        throw new IllegalBoundsError(l, h)
+      } else if (h == l) {
+        ((l),seed)
+      } else if (l == Long.MinValue && h == Long.MaxValue) {
+        val (n,s) = seed.long
+        ((n),s)
+      } else if (l == Int.MinValue && h == Int.MaxValue) {
+        val (n, s) = seed.long
+        ((n.toInt.toLong), s)
+      } else if (l == Short.MinValue && h == Short.MaxValue) {
+        val (n, s) = seed.long
+        ((n.toShort.toLong), s)
+      } else if (l == 0L && h == Char.MaxValue) {
+        val (n, s) = seed.long
+        ((n.toChar.toLong), s)
+      } else if (l == Byte.MinValue && h == Byte.MaxValue) {
+        val (n, s) = seed.long
+        ((n.toByte.toLong), s)
+      } else {
+        val d = h - l + 1
+        if (d <= 0) {
+          var tpl = seed.long
+          var n = tpl._1
+          var s = tpl._2
+          while (n < l || n > h) {
+            tpl = s.long
+            n = tpl._1
+            s = tpl._2
+          }
+          ((n), s)
+        } else {
+          val (n, s) = seed.long
+          ((l + (n & 0x7fffffffffffffffL) % d), s)
+        }
+      }
+    }
+
+abstract class StGen[T] { self =>
+  def doApply(p : Expr[Gen.Parameters],seed:Expr[Seed]) : Cps[(Option[T],Expr[Seed])]
+
+  
+
+  def map[S](f : T => S) : StGen[S] = {
+    StGen.gen((p,seed) =>
+        self.doApply(p,seed).map((ot,seed2) => (ot.map(f),seed2))
+    )
+  }
+
+  def flatMap[S](f : T => StGen[S]) : StGen[S] = {
+    StGen.gen((p,seed) =>
[…]

### 2025-03-04 staged-scala/src/main/scala/Main.scala
splittable

@@ -5,25 +5,8 @@ import org.scalacheck.rng.Seed
 
 import stagedgen.StGen
 
-def fooImpl(using Quotes) = {
-  val e = StGen.splat(StGen.pure('{55}))
-  println(s"Generated code: ${e.show}")
-  e
-}
-
-inline def foo() = {
-  ${fooImpl}
-}
-
-def sample[T](f : Gen.Parameters => Seed => Option[T]) = {
-  f(Gen.Parameters.default)(Seed.random())
-}
-
-
-def msg = "I was compiled by Scala 3. :)"
 
 @main def hello(): Unit =
-  println("Hello world!")
-  println(msg)
-
-
+  val u = StGen.complexStGen
+  val v = u(Gen.Parameters.default)(Seed.random())
+  println(v)

### 2025-03-04 staged-scala/src/main/scala/stagedgen/Cps.scala

@@ -29,6 +29,12 @@ object Cps {
     def apply[Z : Type](cont: T => Expr[Z]): Expr[Z] = cont(t)
   }
 
+  def cps[T](f : [Z : Type] => (T => Expr[Z]) => Expr[Z]) : Cps[T] = {
+    new Cps[T]{
+      def apply[Z : Type](cont: T => Expr[Z]): Expr[Z] = f(cont)
+    }
+  }
+
   def run[T : Type](c : Cps[Expr[T]]) : Expr[T] = {
     c.apply(x => x)
   }

### 2025-03-04 staged-scala/src/main/scala/stagedgen/Splittable.scala

@@ -1,0 +1,39 @@
+package stagedgen
+
+import scala.quoted.*
+import stagedgen.Cps
+
+trait Splittable[T : Type,S] {
+    extension (a : Expr[T]) def split(using Quotes) : Cps[S]
+}
+
+given (using Quotes) : Splittable[Boolean,Boolean] with
+    extension (a : Expr[Boolean])
+        def split(using Quotes) : Cps[Boolean] = Cps.cps([Z : Type] => (k : Boolean => Expr[Z]) => '{
+            if ${a} then ${k(true)} else ${k(false)}
+        })
+
+given [A : Type, B : Type](using Quotes) : Splittable[(A,B),(Expr[A],Expr[B])] with
+    extension (a : Expr[(A,B)])
+        def split(using Quotes) : Cps[(Expr[A],Expr[B])] = Cps.cps([Z : Type] => k => '{
+            val (x,y) = ${a}
+            ${k('{x},'{y})}
+        })
+
+given [A : Type](using Quotes) : Splittable[Option[A],Option[Expr[A]]] with
+    extension (a : Expr[Option[A]])
+        def split(using Quotes) : Cps[Option[Expr[A]]] = Cps.cps([Z : Type] => k => '{
+            ${a} match {
+                case None => ${k(None)}
+                case Some(v) => ${k(Some('{v}))}
+            }
+        })
+
+given [A : Type](using Quotes) : Splittable[List[A],Either[Unit,(Expr[A],Expr[List[A]])]] with
+    extension (a : Expr[List[A]])
+        def split(using Quotes) : Cps[Either[Unit,(Expr[A],Expr[List[A]])]] = Cps.cps([Z : Type] => k => '{
+            ${a} match {
+                case Nil => ${k(Left(()))}
+                case hd :: tl => ${k(Right(('{hd},'{tl})))}
+            }
+        })

### 2025-03-04 staged-scala/src/main/scala/stagedgen/StGen.scala

@@ -49,7 +49,6 @@ private def chLng(l: Long, h: Long)(p: Gen.Parameters, seed: Seed): (Long,Seed) 
 
 abstract class StGen[T] { self =>
   def doApply(p : Expr[Gen.Parameters],seed:Expr[Seed]) : Cps[(Option[T],Expr[Seed])]
-
   
 
   def map[S](f : T => S) : StGen[S] = {
@@ -68,8 +67,6 @@ abstract class StGen[T] { self =>
         )
     )
   }
-
-  
 }
 
 object StGen {
@@ -90,7 +87,7 @@ object StGen {
   def chooseLong(lo : Expr[Long],hi:Expr[Long])(using Quotes) : StGen[Expr[Long]] = {
     StGen.gen((p,seed) =>
         for {
-            (x,y) <- Cps.splitPair('{chLng(${lo},${hi})(${p},${seed})})
+            (x,y) <- ('{chLng(${lo},${hi})(${p},${seed})}).split
         } yield (Some(x),y)
     )
   }
@@ -110,12 +107,15 @@ object StGen {
       println(s"Generator: $s")
   }
 
-  def complexStGenImpl (using Quotes): Expr[Gen.Parameters => Seed => Option[(Long,Long)]] =
-    StGen.splat(for {
+  def complexStGenImpl (using q : Quotes): Expr[Gen.Parameters => Seed => Option[(Long,Long)]] = {
+    val e = StGen.splat(for {
         x <- StGen.chooseLong('{1},'{1000})
         y <- StGen.chooseLong('{0},x)
     } yield '{(${x},${y})}
     )
+    q.reflect.report.info(s"Generated code: ${e.show}")
+    e
+  }
 
   inline def complexStGen = {
     ${complexStGenImpl}

### 2025-03-04 paper/paper.tex
blah

@@
 
-\subsubsection{Weighted unions}
+\subsubsection{Expensive Combinators}
+\jwc{Many combinators like \texttt{union} and \texttt{weighted_union} needlessly allocate large data structures in the generation hot path.
+Using a combination of staging and more careful algorithm design, we can avoid this overhead.
+}
 % \subsubsection{Function call overhead}
@@
 \subsection{Comparing Generators}
+
+\jwc{NOTE: we should test generator speed across both languages, but speed -> bugfinding ability in only OCaml.}
+\jwc{
+  Baseline generators to test speed in both languages:
+  \begin{itemize}
+    \item Single int
+    \item Pair of ints, constrained
+    \item List of ints
+  \end{itemize}
+}
+
 \subsubsection{Existing Approaches}

### 2025-03-05 paper/paper.tex
things
skeleton further

@@
 \section{Background}
+
+\subsection{Property-Based Testing}
+\jwc{Mention the shape of PBT dsls, give running example here (or maybe earlier!). Monadic dsls.}
+
+\begin{lstlisting}
+let int_pair : (int * int) Bq.t =
+  let%bind x = int 0 100 in
+  let%bind y = int 0 x in
+  return (x,y)
+\end{lstlisting}
+
+\jwc{\texttt{int} here is really \texttt{int\_uniform\_inclusive}. I also think we should completely omit named arguments in the paper so we don't confuse with splices.}
+
 \subsection{Staged Metaprogramming}
@@
 
-\subsection{Property-Based Testing}
 \subsection{Other Languages and Libraries}
@@
     \begin{itemize}
-      \item The basic generator type: \texttt{'a generator = size:int -> random:SR.t -> 'a}. Size and random seed to deterministic value.
+      \item The basic generator type: \texttt{'a generator = int -> SR.t -> 'a}. Size and random seed to deterministic value. (note: \texttt{SR.t} is a mutable seed)
       \item Also two inverse functions, \texttt{create}, and \texttt{generate}.
       \item This gets a monad intance in the obvious way (show code).
+      \item Also show the code for \texttt{int}, how it calls the underlying SR function.
     \end{itemize}
-    \item Note that \emph{extensionally} \texttt{generate (create (fun ~size ~random -> e)) ~size ~random = e}, but the OCaml compiler does not always perform this optimization, or do the inlining required to expose it.
+    \item Note that \emph{extensionally} \texttt{generate (create (fun size random -> e)) size random = e}, but the OCaml compiler does not always perform this optimization, or do the inlining required to expose it.
     (When sufficiently obfuscated behind returns and binds ...) This program compiles to code that (1) allocates the closure for `e', (2) passes it to create (which returns the closure), and then calls (3) generate, which immediately jumps into the closure.
+    \item Show allocation benchmarks of the running example, versus the version where you inline everything.
   \end{itemize}
@@
 
-In OCaml, garbage collection is the killer.
+% In OCaml, garbage collection is the killer.
 
-This generator generates a list of bools. To generate a list of 100 bools, it allocates \texttt{4808.00w (minor)}, and runs for an average of
-3.9us. Merely representing a list of bools of length 100 requires 300 words in OCaml, so this generator allocates 16 times more than the theoretical minimum.
+% This generator generates a list of bools. To generate a list of 100 bools, it allocates \texttt{4808.00w (minor)}, and runs for an average of
+% 3.9us. Merely representing a list of bools of length 100 requires 300 words in OCaml, so this generator allocates 16 times more than the theoretical minimum.
 
-\begin{lstlisting}
-let gen_list_bool =
-  fixed_point (fun gl ->
-    let%bind n = size in
-    if n <= 0 then return [] else
-      let%bind x = bool in
-      let%bind xs = with_size ~size:(n-1) gl in
-      return (x :: xs)
-  )
-\end{lstlisting}
+% \jwc{Mention ppx_jane and let syntax}
 
-On the other hand, this generator allocates only 303w, and runs for 0.7us, to generate the same list of 100 bools.
-\begin{lstlisting}
-  let rec gen_list_bool_faster ~size ~random  = 
-    if size <= 0 then []
-    else (Unboxed_splitmix.bool random)::(gen_list_bool_faster ~size:(size - 1) ~random)
-\end{lstlisting}
+% \begin{lstlisting}
+% let gen_list_bool =
+%   fixed_point (fun gl ->
+%     let%bind n = size in
+%     if n <= 0 then return [] else
+%       let%bind x = bool in
+%       let%bind xs = with_size ~size:(n-1) gl in
+%       return (x :: xs)
+%   )
+% \end{lstlisting}
+
+% On the other hand, this generator allocates only 303w, and runs for 0.7us, to generate the same list of 100 bools.
+% \begin{lstlisting}
+%   let rec gen_list_bool_faster ~size ~random  = 
+%     if size <= 0 then []
+%     else (Unboxed_splitmix.bool random)::(gen_list_bool_faster ~size:(size - 1) ~random)
+% \end{lstlisting}
 
@@
 \subsubsection{Expensive Combinators}
-\jwc{Many combinators like \texttt{union} and \texttt{weighted_union} needlessly allocate large data structures in the generation hot path.
+\jwc{Many combinators like \texttt{union} and \texttt{weighted\_union} needlessly allocate large data structures in the generation hot path.
 Using a combination of staging and more careful algorithm design, we can avoid this overhead.
 }
+
+\begin{lstlisting}
+  let of_weighted_list alist =
+  let weights, values = List.unzip alist in
+  let value_array = Array.of_list values in
+  let total_weight, cumulative_weight_array =
+    let array = Array.init (Array.length value_array) ~f:(fun _ -> 0.) in
+    let sum =
+      List.foldi weights ~init:0. ~f:(fun index acc weight ->
+        let acc' = acc +. weight in
+        array.(index) <- acc';
+        acc'
+      )
+    in
+    sum, array
+  in
+  create (fun ~size:_ ~random ->
+    let choice = Splittable_random.float random ~lo:0. ~hi:total_weight in
+    match
+      Array.binary_search
+        cumulative_weight_array
+        ~compare:Float.compare
+        `First_greater_than_or_equal_to
+        choice
+    with
+    | Some index -> value_array.(index)
+    | None -> assert false
+   )
+\end{lstlisting}
+
+\jwc{This builds the cdf of the distribution, samples from 0 to the total, and binary searches through the array to find the bucket.}
+\jwc{Allocating this array is unnecessary: we can simply compute the total weight, sample \texttt{x} between 0 and the total, and then walk the list accumulating the sum, until the accumulator exceeds \texttt{x}.
+Since in practice the list of possible options is quite small (usually at most 10 in practice), the linear time scan is going to be much faster.}
+\jwc{(The point of this para will just be to emphasize that it's almost always better to be allocation-aware, instead of algorithmically clever.)}
+
+\jwc{Moreover, the possible options is almost always statically known in practice: you use weightedunion to generate (say) a datatype with possible
+variants, whose options are known at compile time.
+Put more simply, you basically always call weighted union with an \emph{explicit list}: For this reason, you should not have to incur the cost of allocating this list at runtime.}
+
 % \subsubsection{Function call overhead}
@@
 
-\jwc{Here!}
+\subsection{Basic Design}
 
 \begin{lstlisting}
-type 'a gen = int code -> Random.t code -> 'a
+type 'a gen = int code -> Random.t code -> 'a code
+
+let return (cx : 'a code) : 'a gen = fun size_c random_c -> cx
+
+let bind (g : 'a gen) (k : 'a -> 'b gen) : 'b gen =
+  fun size_c random_c ->
+    .<
+      let x = .~(g size_c random_c) in
+      .~(
+        (k .<x>.) size_c random_c
+      )
+    >.
+
+let int (lo : int code) (hi : int code) : int code gen =
+  fun size_c random_c ->
+    .< SR.int .~random_c .~lo .~hi >.
 \end{lstlisting}
 
-\tr{Now, we formally introduce the library, describe all its functionality, how things work, etc.}
+\jwc{Conversion to a bq generator}
+
+\begin{lstlisting}
+let to_bq (g : 'a code gen) : ('a Bq.t) code =
+  .<
+    Bq.create (fun size random ->
+      .~(g .<size>. .<random>.)
+    )
+  >.
+\end{lstlisting}
+
+\jwc{Then show the running example, with }
+
+\begin{lstlisting}
+let int_pair_staged : ((int * int) code) gen =
+  let%bind x = int 0 100 in
+  let%bind y = int 0 x in
+  return (x,y)
+
+let int_pair : (int * int) Bq.t code = to_bq int_pair_staged
+====
+.<
+  Bq.create (fun size random ->
+    let x = SR.int random 0 100 in
+    let y = SR.int random 0 x in
+    (x,y)
+  )
+.>
+\end{lstlisting}
+
+\jwc{So if you write with the \texttt{gen} combinators and then only call \texttt{to\_bq} at the end, you get something that includes no creates/generates!}
+
+\jwc{Show benchmark difference between these two generators.}
+
+\subsection{Staging More Combinators}
+
+\jwc{Weighted unions: recall that we should not need to materialize}
+
+\subsection{Basic Design}
 

### 2025-03-05 paper/paper.tex
further skeleton tech section

@@
 
-\subsection{Basic Design}
+% let rec genpick n ws =
+% match ws with
+% | [] -> { rand_gen = fun ~size_c:_ ~random_c:_ -> Codecps.return .< failwith "Fell of the end of pick list" >. }
+% | (k,g) :: ws' ->
+%       { rand_gen = 
+%         fun ~size_c ~random_c ->
+%           Codecps.bind (Codecps.split_bool .< Float.compare .~(v2c n) .~(v2c k) <= 0 >.) (fun leq ->
+%             if leq then
+%               g.rand_gen ~size_c ~random_c
+%             else
+%               Codecps.bind (Codecps.let_insertv .< .~(v2c n) -. .~(v2c k) >.) @@ fun n' ->
+%               (* let%bind n' =  in *)
+%               (genpick n' ws').rand_gen ~size_c ~random_c
+%         )
+%       }
+
+\begin{lstlisting}
+  let weighted_union (weighted_elts : (float code * 'a code) list) : 'a t =
+    let sum_code = List.foldr _ in
+    let go (rc : int code) (weighted_elts : (float code * 'a code) list) : 'a code =
+      match weighted_elts with
+      | [] -> .< failwith "Error" >.
+      | (wc,xc) :: elts' ->
+        .<
+          if .~rc <= .~wc then .~xc
+          else
+            let r' = .~rc - .~wc in
+            .~(go .<r'>. elts')
+        >.
+
+    fun size_c random_c ->
+      .<
+        let sum = .~sum_code in
+        let r = SR.int .~random_c 0 sum in
+        .~(go .<r>. gwcs)
+      >.
+\end{lstlisting}
+
+Then, if we write
+
+\begin{lstlisting}
+let grades : charBq.t gen = to_bq (
+  weighted_union [
+    (.<3.0>., 'a');
+    (.<2.0>., 'b');
+    (.<1.0>., 'c');
+  ]
+)
+===
+Bq.create (
+  fun size random ->
+    let sum = 3.0 + 2.0 + 1.0 in
+    let r = SR.int random 0 sum in
+    if r <= 3.0 then 'a'
+    else
+      let r' = r - 3.0 in
+      if r' <= 2.0 then 'b'
+      else
+        let r'' = r' - 2.0 in
+        if r'' <= 1.0 then 'c'
+        else
+          failwith "Error"
+)
+\end{lstlisting}
+
+\subsection{``The Trick'' and Effect Ordering}
 

### 2025-03-05 paper/paper.tex
move structure around a bit more
progress

@@
 
+\jwc{ofc, sometimes you only know the list at runtime (see the STLC generator for an example), so you the library also includes this.}
+
 % \subsubsection{Function call overhead}
@@
 
+\jwc{Note that some of this is a lie, but it's a good first cut: point to the section on \emph{The Trick} to see the real version.}
+
 \begin{lstlisting}
@@
 
-let bind (g : 'a gen) (k : 'a -> 'b gen) : 'b gen =
+let bind (g : 'a gen) (k : 'a code -> 'b gen) : 'b gen =
   fun size_c random_c ->
@@
 let int_pair_staged : ((int * int) code) gen =
-  let%bind x = int 0 100 in
-  let%bind y = int 0 x in
-  return (x,y)
-
+  bind (int .<0>. .<100>.) (fun cx ->
+    bind (int .<0> cx) (fun cy ->
+      return .<(.~cx,.~cy)>.
+    )
+  )
 let int_pair : (int * int) Bq.t code = to_bq int_pair_staged
@@
   weighted_union [
-    (.<3.0>., 'a');
-    (.<2.0>., 'b');
-    (.<1.0>., 'c');
+    (.<3.0>., .<'a'>.);
+    (.<2.0>., .<'b'>.);
+    (.<1.0>., .<'c'>.);
   ]
@@
 
-\subsection{``The Trick'' and Effect Ordering}
+\subsection{Effect Ordering, and Equivalence is Program Equality}
+\jwc{``Wait, why was the defintion of bind like that''?
+Careful readers might note that the definition of bind was a little strange. In particular, why not just do this?}
+\begin{lstlisting}
+  let bind (g : 'a gen) (k : 'a code -> 'b gen) : 'b gen =
+    fun size_c random_c ->
+      k (g size_c random_c) size_c random_c
+\end{lstlisting}
+\jwc{The answer is that this doesn't generate the right code! In particular, this definition of bind means that
+\texttt{bind (int 0 1) (fun x -> return .<(.~x,.~x)>.)} would generate \texttt{fun size random -> (int random 0 1, int random 0 1)},
+which is \emph{not} the desired behavior. (Cite cbn/cbv randomness here).
+A splice is a (hygenic) syntactic replacment, so
+if a function \texttt{f : 'a code -> 'b code} splices its argument multiple times into its output, the argument code will be directly copied in.
+This is why we insert a let-binding in the definition of bind, to ensure that
+the explicit sequencing of bind is preserved into explicit sequencing in the
+generated code.
+}
 
-\jwc{Note: tradeoff between maintaining the effect ordering and even more performance. }
+\jwc{
+The library is carefully design to preserve exactly the effect order of BQ. If
+you write the same generator in both, they run identically: for the same size
+and random seed, they produce the same value.
+}
 
-\jwc{Note with bind that we don't split because it's imperatively updated. Note that this omits the need to use SR, but we use it to maintain backwards compat.}
+\jwc{Note: tradeoff between maintaining the effect ordering and even more performance.}
 
-\subsection{Comparing Generators}
+\subsection{A Real Monad Instance, CPS, and ``The Trick''}
+
+\jwc{
+  \begin{itemize}
+    \item Careful readers might note that the types of return and bind are not actually the right types.
+    \item This means that they aren't compatible with ppx-jane syntax.
+    \item We could try a different type, \texttt{type 'a gen = int code -> SR.t code -> 'a}.
+This has an actual monad instance (show), and ou can write teh combinators of before, with \texttt{'a code gen} everywhere you had \texttt{'a gen} before.
+    \item But because the result type isn't always in \texttt{code}, \texttt{bind} can't perform the let-insertion needed to preserve effect ordering.
+    \item Instead, we turn to a classic technique from the multistage programming literature: writing our staged program in continuation-passing style.
+    \item We define \texttt{type 'a codecps = 'z. ('a -> 'z code) -> 'z code}. This is the polymorphic CPS monad, with the continuation type always in `code'.
+    (Cite: ``improving binding times without explicit cps conversion'', ``Multi-stage programming with functors and monads: Eliminating abstraction overhead from generic code'', and ``Closure-Free Functional Programming in a Two-Level Type Theory'')
+    \item This lets us define the function \texttt{let\_insert : 'a code -> 'a codecps} as \texttt{let\_insert cx = fun k -> k .< let x = .~cx in .~(k .<x>.) >.}.
+    \item We then define \texttt{type 'a gen = int code -> SR.t code -> 'a codecps}, which lets us have it both ways. the effectful programs like \texttt{int} do their \emph{own} let-insertion
+  \end{itemize}
+}
+
+\jwc{This gives us the final definitions:}
+
+\begin{lstlisting}
+module CodeCps = struct
+  type 'a t = 'z. ('a -> 'z code) -> 'z code
+
+  let run (t : 'a code t) : 'a code = t (fun x -> x)
+
+  let let_insert (cx : 'a code) : 'a code t =
+    fun k -> k .< let x = .~cx in .~(k .<x>.) >.
+end
+
+module Gen = struct
+  type 'a t = int code -> SR.t code -> 'a CodeCps.t
+
+  let return (x : 'a) : 'a t = fun _ _ -> x
+
+  let bind (g : 'a t) (f : 'a -> 'b t) =
+    fun size random ->
+      f (g size random) size random
+
+  let int (lo : int code) (hi : int code) : int code t =
+    fun size random ->
+      let_insert .< SR.int .~random .~lo .~hi >.
+end
+\end{lstlisting}
+
+\jwc {
+And this has the desired effect that
+\texttt{bind (int .<0>. .<1.>) (fun x -> return .<(.~x,.~x)>.)} generates
+\texttt{fun size random -> let x = SR.int random 0 1 in (x,x)}
+}
+
+\subsubsection{The Trick}
+
+\subsection{Staged Type-Derived Generators}
+
+\jwc{Todo: Thia}
+
+\section{Evaluation}
+
+\subsubsection{Implementation}
+\subsection{Benchmarking speed \& resource usage}
 
@@
 }
-
-\subsubsection{Existing Approaches}
-\subsubsection{Equality up to Program Equality}
-\subsubsection{Effect Ordering}
-\subsubsection{PBT for PBT}
-\subsubsection{Does maintaining program equality in \name\ affect performance?}
-\section{Evaluation}
-\subsubsection{Implementation}
-\subsection{Benchmarking speed \& resource usage}
 \subsection{Impact on bug-finding ability}

### 2025-03-05 paper/paper.tex
some things, plus buiild issues

@@
 \tr{Each of these consist of an explanation of the problem and pseudocode outlining the solution.}
-\subsubsection{Monadic bind}
+\subsubsection{Monadic Bind}
 
@@
 % \subsubsection{Function call overhead}
-\subsubsection{Boxed integers}
+\subsubsection{Boxed RNG}
+
+\jwc{
+  \begin{itemize}
+    \item The pRNG at the heart of BQ is the hottest part of the hot path: generating a single int with \texttt{int st lo hi} samples from the underlying rng many many times.
+    \item BQ's prng is the SplitMix algorithm, implemented in the splittablerandom package.
+    \item The precise details of how SplitMix works are unimportant for this paper, but the core of it does an enormous amount of 64-bit int arithmetic.
+    \item Because of implementation details related to the garbage collector, OCaml's \texttt{int} type is 63 bits, so an \texttt{Int64.t} is actually \emph{boxed}: a pointer to a custom block, not scanned by the garbage collector.
+    \item Because of this, all \texttt{Int64.t} operations (both arithmetic and bitwise operations) must allocate!
+A single call to \texttt{int} in Bq calls the samples the RNG many times, and each sample (call to \texttt{next\_int64}) allocates 9 words (2 multiplications, one addition, and 6 logical operations).
+  \end{itemize}
+}
+
+\jwc{We discuss ways to fix this in section ...}
+
+\jwc{We note that this inefficiency is not due to the ``abstraction overhead'' of the DSL itself, but simply due to language implementation and algorithm choices.
+We include this to demonstrate the amount of slack in the system, and so we can demonstrate the additional testing-time savings in the evaluation section.
+}
 
@@
     \item Instead, we turn to a classic technique from the multistage programming literature: writing our staged program in continuation-passing style.
-    \item We define \texttt{type 'a codecps = 'z. ('a -> 'z code) -> 'z code}. This is the polymorphic CPS monad, with the continuation type always in `code'.
+    \item We define \texttt{type 'a codecps = 'z.
+('a -> 'z code) -> 'z code}.
+This is the polymorphic CPS monad, with the continuation type always in `code'.
+(Footnote: this is an instance of the \emph{codensity} monad, cf ``Asymptotic Improvement of Computations over Free Monads'')
     (Cite: ``improving binding times without explicit cps conversion'', ``Multi-stage programming with functors and monads: Eliminating abstraction overhead from generic code'', and ``Closure-Free Functional Programming in a Two-Level Type Theory'')
     \item This lets us define the function \texttt{let\_insert : 'a code -> 'a codecps} as \texttt{let\_insert cx = fun k -> k .< let x = .~cx in .~(k .<x>.) >.}.
-    \item We then define \texttt{type 'a gen = int code -> SR.t code -> 'a codecps}, which lets us have it both ways. the effectful programs like \texttt{int} do their \emph{own} let-insertion
+    \item of course, \texttt{codecps} has the usual monad instance for the CPS monad with a polymorphic result type.
+    \item We then define \texttt{type 'a gen = int code -> SR.t code -> 'a codecps}, which lets us have it both ways.
+the effectful programs like \texttt{int} do their \emph{own} let-insertion, and we get a real monad instance.
   \end{itemize}
@@
 module CodeCps = struct
-  type 'a t = 'z. ('a -> 'z code) -> 'z code
+  type 'a t = { cps : 'z. ('a -> 'z code) -> 'z code }
 
-  let run (t : 'a code t) : 'a code = t (fun x -> x)
+  let return x = {cps = fun k -> k x}
+
+  let bind (x : 'a t) (f : 'a -> 'b t) : 'b t =
+    {cps = fun k -> x.cps (fun a -> (f a).cps k)}
+
+  let run (t : 'a code t) : 'a code = t.cps (fun x -> x)
 
   let let_insert (cx : 'a code) : 'a code t =
-    fun k -> k .< let x = .~cx in .~(k .<x>.) >.
+    {cps = fun k -> k .< let x = .~cx in .~(k .<x>.) >.}
 end
@@
     fun size random ->
-      f (g size random) size random
+      CodeCps.bind (g size random) (fun x ->
+        (f x) size random
+      )
 
@@
 \jwc{Todo: Thia}
+
+\subsection{Faster Random Number Generators}
+
+\jwc{
+  \begin{itemize}
+    \item 
+  \end{itemize}
+}
 

### 2025-03-06 paper/paper.tex
push
more edits to structure

@@
 \usepackage{subcaption}
+
+\lstset{
+  basicstyle=\small\fontfamily{cmtt}\selectfont, % Computer Modern Typewriter
+}
 
@@
 
+\jwc{In this paper, we'll use OCaml to illustrate the ideas, but the concepts are portable to other languages: see Section~\ref{subsection:other-langs}}
+
 \subsection{Property-Based Testing}
-\jwc{Mention the shape of PBT dsls, give running example here (or maybe earlier!). Monadic dsls.}
+\jwc{Usual spiel about pbt and generators: here's the BQ generator library, it has return and bind, this is what they're for.}
+
+\begin{lstlisting}
+module Bq = struct
+  type 'a t = int -> SR.t -> 'a
+
+  let return (x : 'a) : 'a t = fun size seed -> x
+
+  let bind (g : 'a t) (f : 'a -> 'b t) : 'b t =
+    fun size seed ->
+      let x = g size seed in
+      (f x) size seed
+
+  let gen_int (lo : int) (hi : int) : int t =
+    fun size seed -> SR.int seed lo hi
+end
+\end{lstlisting}
+
+\jwc{\texttt{gen\_int} here is really \texttt{int\_uniform\_inclusive}. I also think we should completely omit named arguments in the paper so we don't confuse with splices.}
+
+\jwc{Here's an example of a generator: you can sample from it and it gives you pairs of ints, one less than the other.}
 
@@
 let int_pair : (int * int) Bq.t =
-  let%bind x = int 0 100 in
-  let%bind y = int 0 x in
+  let%bind x = Bq.gen_int 0 100 in
+  let%bind y = Bq.gen_int 0 x in
   return (x,y)
@@
 
-\jwc{\texttt{int} here is really \texttt{int\_uniform\_inclusive}. I also think we should completely omit named arguments in the paper so we don't confuse with splices.}
+\jwc{This desugars to...}
+
+\begin{lstlisting}
+let int_pair : (int * int) Bq.t =
+  Bq.bind (Bq.gen_int 0 100) (fun x ->
+    Bq.bind (Bq.gen_int 0 x) (fun y ->
+      Bq.return (x,y)
+  ))
+\end{lstlisting}
 
@@
 
+\jwc{Staged metaprogramming has a long history as a technique for eliminating the runtime overhead of abstractions,
+specializing programs at an earlier stage to run more efficiently at a later stage.}
+
+\jwc{
+  \begin{itemize}
+    \item The \texttt{'a code} type, quote, splice, stage distinction.
+    \item The difference between \texttt{('a -> 'b) code} and \texttt{'a code -> 'b code}
+  \end{itemize}
+}
+
 \subsection{Other Languages and Libraries}
+\label{subsection:other-langs}
 
@@
     \item Show allocation benchmarks of the running example, versus the version where you inline everything.
+    \item Let's look at the running example: inline it all the way.
   \end{itemize}
 }
+
+\begin{lstlisting}
+let int_pair : (int * int) Bq.t =
+  Bq.bind (Bq.gen_int 0 100) (fun x ->
+    Bq.bind (Bq.gen_int 0 x) (fun y ->
+      Bq.return (x,y)
+  ))
+\end{lstlisting}
 
@@
 
+\jwc{Usual introduction to this section, corresponding to how we talked about it in the intro}
+
 \jwc{
@@
     computes a generator dynamically at test time.
+    \item The size and the random seed are only known at run-time (the later stage), but the code of the generator itself is known at compile time.
+    \item This means the generator type should be \texttt{int code -> SR.t code -> 'a code}: a statically known-function from dynamically-known int and seed to dynamically-determined result.
   \end{itemize}
@@
 
+A generator of \texttt{'a}s is then a (compile-time) function \texttt{int code -> Random.t code -> 'a code}.
+
 \jwc{Note that some of this is a lie, but it's a good first cut: point to the section on \emph{The Trick} to see the real version.}
 
+\jwc{
+  \begin{itemize}
+    \item The simplest combinators are presented below.
+    \item \texttt{return} is the constant generator, given code for \texttt{x : 'a}, it gives the generator which always generates \texttt{x}
+    \item \texttt{bind} is like BQ's bind --- sequencing generators by passing the result of one to another --- except that the continuation gets access to the
+    \texttt{code} for the result of running the first generator.
+At compile time, we know only that \texttt{g} will return \emph{some} \texttt{a}, but we do not get access to the particular \texttt{a}.
+  \end{itemize}
+}
+
 \begin{lstlisting}
-type 'a gen = int code -> Random.t code -> 'a code
 
-let return (cx : 'a code) : 'a gen = fun size_c random_c -> cx
+module Gen = struct
+  type 'a t = int code -> Random.t code -> 'a code
 
-let bind (g : 'a gen) (k : 'a code -> 'b gen) : 'b gen =
+  let return (cx : 'a code) : 'a t = fun size_c random_c -> cx
+
+  let bind (g : 'a t) (k : 'a code -> 'b t) : 'b t =
     fun size_c random_c ->
@@
 
-let int (lo : int code) (hi : int code) : int code gen =
+  let int (lo : int code) (hi : int code) : int t =
     fun size_c random_c ->
       .< SR.int .~random_c .~lo .~hi >.
+end
 \end{lstlisting}
@@
 \begin{lstlisting}
-let to_bq (g : 'a code gen) : ('a Bq.t) code =
+let to_bq (g : 'a code Gen.t) : ('a Bq.t) code =
   .<
@@
 \begin{lstlisting}
-let int_pair_staged : ((int * int) code) gen =
-  bind (int .<0>. .<100>.) (fun cx ->
-    bind (int .<0> cx) (fun cy ->
-      return .<(.~cx,.~cy)>.
+let int_pair_staged : ((int * int) code) Gen.t =
+  Gen.bind (int .<0>. .<100>.) (fun cx ->
+    Gen.bind (int .<0> cx) (fun cy ->
+      Gen.return .<(.~cx,.~cy)>.
     )
   )
-let int_pair : (int * int) Bq.t code = to_bq int_pair_staged
+let int_pair : (int * int) Bq.t code = Gen.to_bq int_pair_staged
 ====
@@
 
-\jwc{So if you write with the \texttt{gen} combinators and then only call \texttt{to\_bq} at the end, you get something that includes no creates/generates!}
+\jwc{So if you write with the \texttt{Gen} combinators and then only call \texttt{to\_bq} at the end, you get something that includes no creates/generates!}
 
@@
 
+\jwc{This is 3-stage metaprogramming! There's PPX-time, compile time, and run-time.}
+
 \subsection{Faster Random Number Generators}

### 2025-03-06 paper/paper.tex
fix lstlisting
work

@@
 \lstset{
-  basicstyle=\small\fontfamily{cmtt}\selectfont, % Computer Modern Typewriter
+  mathescape=true,
+  frame=none,
+  xleftmargin=10pt,
+  stepnumber=1,
+  belowcaptionskip=\bigskipamount,
+  captionpos=b,
+  % language=haskell,
+  keepspaces=true,
+  tabsize=2,
+  emphstyle={\bf},
+  % commentstyle=\it\color{dkgreen},
+  stringstyle=\mdseries\ttfamily,
+  showspaces=false,
+  keywordstyle=\bfseries\ttfamily,
+  columns=flexible,
+  basicstyle=\footnotesize\ttfamily,
+  showstringspaces=false,
+  % morecomment=[l]\%,
+  % moredelim=**[is][\color{dkgreen}]{@}{@}
 }
@@
     \item The difference between \texttt{('a -> 'b) code} and \texttt{'a code -> 'b code}
+    \item We can convert one way, but not the other.
+    \item Functions \texttt{'a code -> 'b code} ``fuse''.
+    \begin{itemize}
+      \item Consider writing \texttt{even . succ}. This
+      \item Consider \texttt{even_c : int code -> bool code = fun cx -> .<.~cx mod 2 == 0>.} and \texttt{succ_c : int code -> int code = fun cx -> .< .~x + 1 >.}
+      \item If you do \texttt{to_dyn (even_c . succ_c)}, you get \texttt{fun x -> (x + 1) mod 2 == 0}. Composing functions from code to code, and then \emph{only at the end} stamping out
+      a dynamic function value ensures that the resulting code is free of abstractions.
+    \end{itemize}
+    and 
   \end{itemize}
@@
 
+\jwc{We can simplify this further by getting rid of the size parameter. Make the story even cleaner, I think!}
+
 \jwc{
   \begin{itemize}
-    \item How does BQ work?
+    \item How does (simplified) BQ work?
     \begin{itemize}
       \item The basic generator type: \texttt{'a generator = int -> SR.t -> 'a}. Size and random seed to deterministic value. (note: \texttt{SR.t} is a mutable seed)
-      \item Also two inverse functions, \texttt{create}, and \texttt{generate}.
       \item This gets a monad intance in the obvious way (show code).
@@
     \end{itemize}
-    \item Note that \emph{extensionally} \texttt{generate (create (fun size random -> e)) size random = e}, but the OCaml compiler does not always perform this optimization, or do the inlining required to expose it.
-    (When sufficiently obfuscated behind returns and binds ...) This program compiles to code that (1) allocates the closure for `e', (2) passes it to create (which returns the closure), and then calls (3) generate, which immediately jumps into the closure.
-    \item Show allocation benchmarks of the running example, versus the version where you inline everything.
+    % \item Note that \emph{extensionally} \texttt{generate (create (fun size random -> e)) size random = e}, but the OCaml compiler does not always perform this optimization, or do the inlining required to expose it.
+    % (When sufficiently obfuscated behind returns and binds ...) This program compiles to code that (1) allocates the closure for `e', (2) passes it to create (which returns the closure), and then calls (3) generate, which immediately jumps into the closure.
+    \item Show benchmarks of the running example, versus the version where you inline everything. ()
     \item Let's look at the running example: inline it all the way.
@@
   ))
+
+let int_pair_inlined : (int * int) Bq.t
+  fun sr ->
+    let x = Splittable_random.int sr ~lo:0 ~hi:100 in
+    let y = Splittable_random.int sr ~lo:0 ~hi:x in
+    (x,y)
 \end{lstlisting}
 
-% In OCaml, garbage collection is the killer.
+\jwc{Show benchmark difference between these two generators: the benchmark is in \texttt{waffle-house/handwritten-ocaml/bin/basic-compare.ml}. It's about 2x. (see comment in latex here.)}
 
-% This generator generates a list of bools. To generate a list of 100 bools, it allocates \texttt{4808.00w (minor)}, and runs for an average of
-% 3.9us. Merely representing a list of bools of length 100 requires 300 words in OCaml, so this generator allocates 16 times more than the theoretical minimum.
+\jwc{The native code OCaml compiler, even with -O3, fails to specialize this code and eliminate the overhead of this abstraction --- inlining all of the funciton definitions and then performing beta-reductions speeds up sampling by a factor of 2.}
 
-% \jwc{Mention ppx_jane and let syntax}
+% ┌───────┬──────────┬─────────┬────────────┐
+% │ Name  │ Time/Run │ mWd/Run │ Percentage │
+% ├───────┼──────────┼─────────┼────────────┤
+% │ bq    │  70.47ns │  78.00w │    100.00% │
+% │ fused │  35.86ns │  78.00w │     50.89% │
+% └───────┴──────────┴─────────┴────────────┘
 
-% \begin{lstlisting}
-% let gen_list_bool =
-%   fixed_point (fun gl ->
-%     let%bind n = size in
-%     if n <= 0 then return [] else
-%       let%bind x = bool in
-%       let%bind xs = with_size ~size:(n-1) gl in
-%       return (x :: xs)
-%   )
-% \end{lstlisting}
-
-% On the other hand, this generator allocates only 303w, and runs for 0.7us, to generate the same list of 100 bools.
-% \begin{lstlisting}
-%   let rec gen_list_bool_faster ~size ~random  = 
-%     if size <= 0 then []
-%     else (Unboxed_splitmix.bool random)::(gen_list_bool_faster ~size:(size - 1) ~random)
-% \end{lstlisting}
+\jwc{The overhead of the abstraction is 2x.
+The reality is actually worse: actual base-quickcheck includes even more indirection in its type.}
 
@@
   \begin{itemize}
-    \item Naked binds like \texttt{Bind}
+    \item In cases where the compiler cannot statically eliminate them, binds allocate short-lived closures: we allocate a closure for the continuation, and then immediately jump into it.
+    \item Each closure allocation is relatively cheap (they're going to be minor allocations since they're short-lived), but doing lots of allocation in a generation hot loop adds up fast, and each minor allocation brings us closer to the next costly GC pause.
   \end{itemize}
@@
 
-\section{\name: A Library for Staging Generators}
+\section{\name: A Library for Staged Generators}
 
@@
 At compile time, we know only that \texttt{g} will return \emph{some} \texttt{a}, but we do not get access to the particular \texttt{a}.
+    \item Bind takes code for the size and seed, and returns code that first let-binds the result of the first generator.
+    The function application \texttt{g size_c random_c} runs during the first stage, producing the code for the \texttt{a}.
+This is then spliced into the quote: the specialized code of the generator \texttt{g}
+    put on the right-hand side of the let. Then, we splice in the code for the continuation, passing first the variable x, and the same size and random seed.
+    \item \texttt{}
   \end{itemize}
@@
         let x = .~(g size_c random_c) in
-        .~(
-          (k .<x>.) size_c random_c
-        )
+        .~(k .<x>. size_c random_c)
       >.
@@
       .< SR.int .~random_c .~lo .~hi >.
+
+  let to_bq (g : 'a code Gen.t) : ('a Bq.t) code =
+  .<
+    fun size random -> .~(g .<size>. .<random>.)
+  >.
 end
@@
 
-\begin{lstlisting}
-let to_bq (g : 'a code Gen.t) : ('a Bq.t) code =
-  .<
-    Bq.create (fun size random ->
-      .~(g .<size>. .<random>.)
-    )
-  >.
-\end{lstlisting}
-
 \jwc{Then show the running example, with }
@@
 let int_pair_staged : ((int * int) code) Gen.t =
-  Gen.bind (int .<0>. .<100>.) (fun cx ->
-    Gen.bind (int .<0> cx) (fun cy ->
+  Gen.bind (Gen.int .<0>. .<100>.) (fun cx ->
+    Gen.bind (Gen.int .<0> cx) (fun cy ->
       Gen.return .<(.~cx,.~cy)>.
@@
 .<
-  Bq.create (fun size random ->
+  fun size random ->
     let x = SR.int random 0 100 in
@@
     (x,y)
-  )
 .>
@@
 
-\jwc{So if you write with the \texttt{Gen} combinators and then only call \texttt{to\_bq} at the end, you get something that includes no creates/generates!}
-
-\jwc{Show benchmark difference between these two generators.}
+\jwc{So if you write with the \texttt{Gen} combinators and then only call \texttt{to\_bq} at the end, you get something that inlines everything all the way down!}
 
@@
 
-\subsection{A Real Monad Instance, CPS, and ``The Trick''}
+\subsection{Size Parameters, A Real Monad Instance, CPS, and ``The Trick''}
+
+\jwc{Introduce the size parameter here. (if you deleted it beforehand.)}
 
@@
 
-  let return (x : 'a) : 'a t = fun _ _ -> x
+  let return (x : 'a) : 'a t = fun _ _ -> Codecps.return x
 

### 2025-03-07 staged-ocaml/lib/randomness/random_stub.c
oof, really rough bug with C ffi

@@ -138,8 +138,8 @@ CAMLprim value int_c_unchecked(value state_val, value lo_val, value hi_val) {
   CAMLparam3(state_val,lo_val,hi_val);
   state_t st = State_val(state_val);
 
-  int64_t lo = (int64_t) Int_val(lo_val);
-  int64_t hi = (int64_t) Int_val(hi_val);
+  int64_t lo = (int64_t) Long_val(lo_val);
+  int64_t hi = (int64_t) Long_val(hi_val);
 
   int64_t result = next_int(&st,lo,hi);
   CAMLreturn(Val_int(result));
@@ -191,8 +191,8 @@ CAMLprim value int_c_log_uniform(value state_val, value lo_val, value hi_val) {
   CAMLparam3(state_val,lo_val,hi_val);
   state_t st = State_val(state_val);
 
-  int64_t lo = (int64_t) Int_val(lo_val);
-  int64_t hi = (int64_t) Int_val(hi_val);
+  int64_t lo = (int64_t) Long_val(lo_val);
+  int64_t hi = (int64_t) Long_val(hi_val);
 
   int64_t result = next_int_log_uniform(&st,lo,hi);
   CAMLreturn(Val_int(result));

### 2025-03-07 staged-ocaml/lib/randomness/sr_random_stub.c

@@ -117,8 +117,8 @@ CAMLprim value int_c_sr_unchecked(value sr_state_val, value lo_val, value hi_val
   state_t* st = (state_t*) alloca(sizeof(state_t));
   fill_from_value(sr_state_val,st);
 
-  int64_t lo = (int64_t) Int_val(lo_val);
-  int64_t hi = (int64_t) Int_val(hi_val);
+  int64_t lo = (int64_t) Long_val(lo_val);
+  int64_t hi = (int64_t) Long_val(hi_val);
 
   int64_t result = next_int_sr(st,lo,hi);
   CAMLreturn(Val_int(result));

### 2025-03-09 staged-ocaml/lib/randomness/common.c
HUGE, fixed the boxing of floats

@@ -1,28 +1,0 @@
-#include <caml/mlvalues.h>
-#include <caml/memory.h>
-#include <caml/alloc.h>
-#include <caml/custom.h>
-#include <stdint.h>
-#include <stdbool.h>
-#include <string.h>
-#include <stdio.h>
-#include <math.h>
-// This bug took me forever to figure out. A "logical shift right" requires
-// casting to uint64_t first.
-int64_t mix_bits(int64_t z, int64_t n) {
-  return z ^ ((uint64_t) z >> n);
-}
-
-int64_t mix64(int64_t z) {
-  z = mix_bits(z,33) * 0xff51afd7ed558ccdL;
-  z = mix_bits(z,33) * 0xc4ceb9fe1a85ec53L;
-  return mix_bits(z, 33);
-}
-
-bool remainder_is_unbiased(int64_t draw, int64_t remainder, int64_t draw_max, int64_t remainder_max){
-  return (draw - remainder <= draw_max - remainder_max);
-}
-
-double unit_float_from_int64(int64_t n){
-  return ((uint64_t) n >> 11) * pow(2,-53);
-}

### 2025-03-09 staged-ocaml/lib/randomness/common.h

@@ -1,22 +1,0 @@
-#ifndef COMMON_H_   /* Include guard */
-#define COMMON_H_
-
-#include <caml/mlvalues.h>
-#include <caml/memory.h>
-#include <caml/alloc.h>
-#include <caml/custom.h>
-#include <stdint.h>
-#include <stdbool.h>
-#include <string.h>
-#include <stdio.h>
-#include <math.h>
-
-int64_t mix_bits(int64_t z, int64_t n);
-
-int64_t mix64(int64_t z);
-
-bool remainder_is_unbiased(int64_t draw, int64_t remainder, int64_t draw_max, int64_t remainder_max);
-
-double unit_float_from_int64(int64_t n);
-
-#endif

### 2025-03-09 staged-ocaml/lib/randomness/random_stub.c

@@ -8,7 +8,23 @@
 #include <stdio.h>
 #include <math.h>
 
-#include "common.h"
+static inline int64_t mix_bits(int64_t z, int64_t n) {
+  return z ^ ((uint64_t) z >> n);
+}
+
+static inline int64_t mix64(int64_t z) {
+  z = mix_bits(z,33) * 0xff51afd7ed558ccdL;
+  z = mix_bits(z,33) * 0xc4ceb9fe1a85ec53L;
+  return mix_bits(z, 33);
+}
+
+static inline bool remainder_is_unbiased(int64_t draw, int64_t remainder, int64_t draw_max, int64_t remainder_max){
+  return (draw - remainder <= draw_max - remainder_max);
+}
+
+static inline double unit_float_from_int64(int64_t n){
+  return ((uint64_t) n >> 11) * pow(2,-53);
+}
 
 
 typedef struct state {

### 2025-03-09 staged-ocaml/lib/randomness/sr_random_stub.c

@@ -8,7 +8,25 @@
 #include <stdio.h>
 #include <math.h>
 
-#include "common.h"
+static inline int64_t mix_bits(int64_t z, int64_t n) {
+  return z ^ ((uint64_t) z >> n);
+}
+
+static inline int64_t mix64(int64_t z) {
+  z = mix_bits(z,33) * 0xff51afd7ed558ccdL;
+  z = mix_bits(z,33) * 0xc4ceb9fe1a85ec53L;
+  return mix_bits(z, 33);
+}
+
+static inline bool remainder_is_unbiased(int64_t draw, int64_t remainder, int64_t draw_max, int64_t remainder_max){
+  return (draw - remainder <= draw_max - remainder_max);
+}
+
+static inline double unit_float_from_int64(int64_t n){
+  return ((uint64_t) n >> 11) * pow(2,-53);
+}
+
+
 
 
 /*
@@ -102,14 +120,21 @@ CAMLprim value bool_c_sr(value sr_state_val) {
   CAMLreturn(Val_bool(next_bool_sr(st)));
 }
 
-CAMLprim value float_c_sr_unchecked(value sr_state_val, value lo_val, value hi_val){
+
+
+CAMLprim double float_c_sr_unchecked_unboxed(value sr_state_val, value lo_val, value hi_val){
   CAMLparam3(sr_state_val,lo_val,hi_val);
   state_t* st = (state_t*) alloca(sizeof(state_t));
   fill_from_value(sr_state_val,st);
   double lo = Double_val(lo_val);
   double hi = Double_val(hi_val);
   double result = next_float_sr(st,lo,hi);
-  CAMLreturn(caml_copy_double(result));
+  CAMLreturn(result);
+}
+
+CAMLprim value float_c_sr_unchecked(value sr_state_val, value lo_val, value hi_val){
+  CAMLparam3(sr_state_val,lo_val,hi_val);
+  CAMLreturn(caml_copy_double(float_c_sr_unchecked_unboxed(sr_state_val,lo_val,hi_val)));
 }
 
 CAMLprim value int_c_sr_unchecked(value sr_state_val, value lo_val, value hi_val) {

### 2025-03-09 staged-ocaml/lib/randomness/random_stub.c
whew.

@@ -141,13 +141,18 @@ CAMLprim value bool_c(value state_val) {
   CAMLreturn(Val_bool(next_bool(&st)));
 }
 
-CAMLprim value float_c_unchecked(value state_val, value lo_val, value hi_val){
+CAMLprim double float_c_unchecked_unboxed(value state_val, value lo_val, value hi_val){
   CAMLparam3(state_val,lo_val,hi_val);
   state_t st = State_val(state_val);
   double lo = Double_val(lo_val);
   double hi = Double_val(hi_val);
   double result = next_float(&st,lo,hi);
-  CAMLreturn(caml_copy_double(result));
+  CAMLreturn(result);
+}
+
+CAMLprim value float_c_unchecked(value state_val, value lo_val, value hi_val){
+  CAMLparam3(state_val,lo_val,hi_val);
+  CAMLreturn(caml_copy_double(float_c_unchecked_unboxed(state_val,lo_val,hi_val)));
 }
 
 CAMLprim value int_c_unchecked(value state_val, value lo_val, value hi_val) {

### 2025-03-09 staged-ocaml/lib/randomness/random_stub.c
unbox more

@@ -141,18 +141,16 @@ CAMLprim value bool_c(value state_val) {
   CAMLreturn(Val_bool(next_bool(&st)));
 }
 
-CAMLprim double float_c_unchecked_unboxed(value state_val, value lo_val, value hi_val){
-  CAMLparam3(state_val,lo_val,hi_val);
+CAMLprim double float_c_unchecked_unboxed(value state_val, double lo, double hi){
+  CAMLparam1(state_val);
   state_t st = State_val(state_val);
-  double lo = Double_val(lo_val);
-  double hi = Double_val(hi_val);
   double result = next_float(&st,lo,hi);
   CAMLreturn(result);
 }
 
 CAMLprim value float_c_unchecked(value state_val, value lo_val, value hi_val){
   CAMLparam3(state_val,lo_val,hi_val);
-  CAMLreturn(caml_copy_double(float_c_unchecked_unboxed(state_val,lo_val,hi_val)));
+  CAMLreturn(caml_copy_double(float_c_unchecked_unboxed(state_val,Double_val(lo_val),Double_val(hi_val))));
 }
 
 CAMLprim value int_c_unchecked(value state_val, value lo_val, value hi_val) {
@@ -167,7 +165,7 @@ CAMLprim value int_c_unchecked(value state_val, value lo_val, value hi_val) {
 }
 
 
-int64_t min_represented_by_n_bits(int64_t n) {
+static inline int64_t min_represented_by_n_bits(int64_t n) {
   if (n == 0) {
       return 0;
   } else {
@@ -175,20 +173,20 @@ int64_t min_represented_by_n_bits(int64_t n) {
   }
 }
 
-int64_t max_represented_by_n_bits(int64_t bits) {
+static inline int64_t max_represented_by_n_bits(int64_t bits) {
   return (1LL << bits) - 1;
 }
 
 
-int64_t max(int64_t a, int64_t b) {
+static inline int64_t max(int64_t a, int64_t b) {
   return (a > b) ? a : b;
 }
 
-int64_t min(int64_t a, int64_t b) {
+static inline int64_t min(int64_t a, int64_t b) {
   return (a < b) ? a : b;
 }
 
-uint64_t bits_to_represent(int64_t t) {
+static uint64_t bits_to_represent(int64_t t) {
   uint64_t n = 0;
   
   while (t > 0) {

### 2025-03-09 staged-ocaml/lib/randomness/sr_random_stub.c

@@ -122,19 +122,19 @@ CAMLprim value bool_c_sr(value sr_state_val) {
 
 
 
-CAMLprim double float_c_sr_unchecked_unboxed(value sr_state_val, value lo_val, value hi_val){
-  CAMLparam3(sr_state_val,lo_val,hi_val);
+CAMLprim double float_c_sr_unchecked_unboxed(value sr_state_val, double lo, double hi){
+  CAMLparam1(sr_state_val);
   state_t* st = (state_t*) alloca(sizeof(state_t));
   fill_from_value(sr_state_val,st);
-  double lo = Double_val(lo_val);
-  double hi = Double_val(hi_val);
+  // double lo = Double_val(lo_val);
+  // double hi = Double_val(hi_val);
   double result = next_float_sr(st,lo,hi);
   CAMLreturn(result);
 }
 
 CAMLprim value float_c_sr_unchecked(value sr_state_val, value lo_val, value hi_val){
   CAMLparam3(sr_state_val,lo_val,hi_val);
-  CAMLreturn(caml_copy_double(float_c_sr_unchecked_unboxed(sr_state_val,lo_val,hi_val)));
+  CAMLreturn(caml_copy_double(float_c_sr_unchecked_unboxed(sr_state_val,Double_val(lo_val),Double_val(hi_val))));
 }
 
 CAMLprim value int_c_sr_unchecked(value sr_state_val, value lo_val, value hi_val) {
@@ -146,6 +146,60 @@ CAMLprim value int_c_sr_unchecked(value sr_state_val, value lo_val, value hi_val
   int64_t hi = (int64_t) Long_val(hi_val);
 
   int64_t result = next_int_sr(st,lo,hi);
+  CAMLreturn(Val_int(result));
+}
+
+
+static inline int64_t min_represented_by_n_bits(int64_t n) {
+  if (n == 0) {
+      return 0;
+  } else {
+      return 1LL << (n - 1);
+  }
+}
+
+static inline int64_t max_represented_by_n_bits(int64_t bits) {
+  return (1LL << bits) - 1;
+}
+
+
+static inline int64_t max(int64_t a, int64_t b) {
+  return (a > b) ? a : b;
+}
+
+static inline int64_t min(int64_t a, int64_t b) {
+  return (a < b) ? a : b;
+}
+
+static uint64_t bits_to_represent(int64_t t) {
+  uint64_t n = 0;
+  
[…]

### 2025-03-09 staged-ocaml/lib/randomness/random_stub.c
start to pivot floats over

@@ -8,21 +8,21 @@
 #include <stdio.h>
 #include <math.h>
 
-static inline int64_t mix_bits(int64_t z, int64_t n) {
+int64_t mix_bits(int64_t z, int64_t n) {
   return z ^ ((uint64_t) z >> n);
 }
 
-static inline int64_t mix64(int64_t z) {
+int64_t mix64(int64_t z) {
   z = mix_bits(z,33) * 0xff51afd7ed558ccdL;
   z = mix_bits(z,33) * 0xc4ceb9fe1a85ec53L;
   return mix_bits(z, 33);
 }
 
-static inline bool remainder_is_unbiased(int64_t draw, int64_t remainder, int64_t draw_max, int64_t remainder_max){
+bool remainder_is_unbiased(int64_t draw, int64_t remainder, int64_t draw_max, int64_t remainder_max){
   return (draw - remainder <= draw_max - remainder_max);
 }
 
-static inline double unit_float_from_int64(int64_t n){
+double unit_float_from_int64(int64_t n){
   return ((uint64_t) n >> 11) * pow(2,-53);
 }
 
@@ -222,4 +222,54 @@ CAMLprim value print(value state_val) {
   state_t st = State_val(state_val);
   printf("C-Seed: %ld\n", st.seed);
   CAMLreturn(Val_unit);
+}
+
+static inline int64_t to_int64_preserve_order(double t){
+  if(t == 0.0) {
+    return 0L;
+  } else if(t > 0.0){
+    return ((int64_t) t);
+  } else {
+    return (- ((int64_t) (-t)));
+  }
+}
+
+static inline double of_int64_preserve_order(double x){
+  if(x >= 0L){
+    return ((double) x);
+  } else {
+    return -((double) (-x));
+  }
+}
+
+CAMLprim double one_ulp_up_c_unboxed(double x){
+  CAMLparam0();
+  double res;
+  if(isnan(x)){
+    res = NAN;
+  } else {
+    res = of_int64_preserve_order(to_int64_preserve_order(x) + 1L);
+  }
+  return res;
+}
+
+CAMLprim value one_ulp_up_c(value x_val){
+  CAMLparam1(x_val);
+  CAMLreturn(caml_copy_double(one_ulp_up_c_unboxed(Double_val(x_val))));
+}
+
+CAMLprim double one_ulp_down_c_unboxed(double x){
+  CAMLparam0();
+  double res;
[…]

### 2025-03-09 staged-ocaml/lib/randomness/sr_random_stub.c

@@ -209,3 +209,53 @@ CAMLprim value int_c_sr_log_uniform(value sr_state_val, value lo_val, value hi_v
 //   printf("C-Seed: %ld\n", st.seed);
 //   CAMLreturn(Val_unit);
 // }
+
+static inline int64_t to_int64_preserve_order(double t){
+  if(t == 0.0) {
+    return 0L;
+  } else if(t > 0.0){
+    return ((int64_t) t);
+  } else {
+    return (- ((int64_t) (-t)));
+  }
+}
+
+static inline double of_int64_preserve_order(double x){
+  if(x >= 0L){
+    return ((double) x);
+  } else {
+    return -((double) (-x));
+  }
+}
+
+CAMLprim double one_ulp_up_c_sr_unboxed(double x){
+  CAMLparam0();
+  double res;
+  if(isnan(x)){
+    res = NAN;
+  } else {
+    res = of_int64_preserve_order(to_int64_preserve_order(x) + 1L);
+  }
+  return res;
+}
+
+CAMLprim value one_ulp_up_c_sr(value x_val){
+  CAMLparam1(x_val);
+  CAMLreturn(caml_copy_double(one_ulp_up_c_sr_unboxed(Double_val(x_val))));
+}
+
+CAMLprim double one_ulp_down_c_sr_unboxed(double x){
+  CAMLparam0();
+  double res;
+  if(isnan(x)){
+    res = NAN;
+  } else {
+    res = of_int64_preserve_order(to_int64_preserve_order(x) - 1L);
+  }
+  return res;
+}
+
+CAMLprim value one_ulp_down_c_sr(value x_val){
+  CAMLparam1(x_val);
+  CAMLreturn(caml_copy_double(one_ulp_down_c_sr_unboxed(Double_val(x_val))));
+}

### 2025-03-09 staged-ocaml/lib/randomness/random_stub.c
shit

@@ -145,6 +145,7 @@ CAMLprim double float_c_unchecked_unboxed(value state_val, double lo, double hi)
   CAMLparam1(state_val);
   state_t st = State_val(state_val);
   double result = next_float(&st,lo,hi);
+  printf("RESULT: %f\n",result);
   CAMLreturn(result);
 }
 
@@ -224,21 +225,33 @@ CAMLprim value print(value state_val) {
   CAMLreturn(Val_unit);
 }
 
+static inline double bitcast_int64_to_double(int64_t value) {
+  double result;
+  memcpy(&result, &value, sizeof(double));
+  return result;
+}
+
+static inline int64_t bitcast_double_to_int64(double value) {
+  int64_t result;
+  memcpy(&result, &value, sizeof(int64_t));
+  return result;
+}
+
 static inline int64_t to_int64_preserve_order(double t){
   if(t == 0.0) {
     return 0L;
   } else if(t > 0.0){
-    return ((int64_t) t);
+    return (bitcast_double_to_int64(t));
   } else {
-    return (- ((int64_t) (-t)));
+    return (- (bitcast_double_to_int64(-t)));
   }
 }
 
-static inline double of_int64_preserve_order(double x){
+static inline double of_int64_preserve_order(int64_t x){
   if(x >= 0L){
-    return ((double) x);
+    return (bitcast_int64_to_double(x));
   } else {
-    return -((double) (-x));
+    return -(bitcast_int64_to_double(-x));
   }
 }
 

### 2025-03-09 staged-ocaml/lib/randomness/sr_random_stub.c

@@ -210,21 +210,33 @@ CAMLprim value int_c_sr_log_uniform(value sr_state_val, value lo_val, value hi_v
 //   CAMLreturn(Val_unit);
 // }
 
+static inline double bitcast_int64_to_double(int64_t value) {
+  double result;
+  memcpy(&result, &value, sizeof(double));
+  return result;
+}
+
+static inline int64_t bitcast_double_to_int64(double value) {
+  int64_t result;
+  memcpy(&result, &value, sizeof(int64_t));
+  return result;
+}
+
 static inline int64_t to_int64_preserve_order(double t){
   if(t == 0.0) {
     return 0L;
   } else if(t > 0.0){
-    return ((int64_t) t);
+    return (bitcast_double_to_int64(t));
   } else {
-    return (- ((int64_t) (-t)));
+    return (- (bitcast_double_to_int64(-t)));
   }
 }
 
-static inline double of_int64_preserve_order(double x){
+static inline double of_int64_preserve_order(int64_t x){
   if(x >= 0L){
-    return ((double) x);
+    return (bitcast_int64_to_double(x));
   } else {
-    return -((double) (-x));
+    return -(bitcast_int64_to_double(-x));
   }
 }
 

### 2025-03-10 staged-ocaml/lib/randomness/random_stub.c
fix floats, faster bitcast with UB

@@ -141,17 +141,17 @@ CAMLprim value bool_c(value state_val) {
   CAMLreturn(Val_bool(next_bool(&st)));
 }
 
-CAMLprim double float_c_unchecked_unboxed(value state_val, double lo, double hi){
-  CAMLparam1(state_val);
+double float_c_unchecked_unboxed(value state_val, double lo, double hi){
+  // CAMLparam1(state_val);
   state_t st = State_val(state_val);
   double result = next_float(&st,lo,hi);
-  printf("RESULT: %f\n",result);
-  CAMLreturn(result);
+  return result;
 }
 
 CAMLprim value float_c_unchecked(value state_val, value lo_val, value hi_val){
   CAMLparam3(state_val,lo_val,hi_val);
-  CAMLreturn(caml_copy_double(float_c_unchecked_unboxed(state_val,Double_val(lo_val),Double_val(hi_val))));
+  value v = caml_copy_double(float_c_unchecked_unboxed(state_val,Double_val(lo_val),Double_val(hi_val)));
+  CAMLreturn(v);
 }
 
 CAMLprim value int_c_unchecked(value state_val, value lo_val, value hi_val) {
@@ -226,15 +226,17 @@ CAMLprim value print(value state_val) {
 }
 
 static inline double bitcast_int64_to_double(int64_t value) {
-  double result;
-  memcpy(&result, &value, sizeof(double));
-  return result;
+  return *(double*)&value;
+  // double result;
+  // memcpy(&result, &value, sizeof(double));
+  // return result;
 }
 
 static inline int64_t bitcast_double_to_int64(double value) {
-  int64_t result;
-  memcpy(&result, &value, sizeof(int64_t));
-  return result;
+  return *(int64_t*)&value;
+  // int64_t result;
+  // memcpy(&result, &value, sizeof(int64_t));
+  // return result;
 }
 
 static inline int64_t to_int64_preserve_order(double t){
@@ -255,8 +257,7 @@ static inline double of_int64_preserve_order(int64_t x){
   }
 }
 
-CAMLprim double one_ulp_up_c_unboxed(double x){
-  CAMLparam0();
+double one_ulp_up_c_unboxed(double x){
   double res;
   if(isnan(x)){
     res = NAN;
@@ -271,8 +272,7 @@ CAMLprim value one_ulp_up_c(value x_val){
[…]

### 2025-03-10 staged-ocaml/lib/randomness/sr_random_stub.c

@@ -122,14 +122,13 @@ CAMLprim value bool_c_sr(value sr_state_val) {
 
 
 
-CAMLprim double float_c_sr_unchecked_unboxed(value sr_state_val, double lo, double hi){
-  CAMLparam1(sr_state_val);
+double float_c_sr_unchecked_unboxed(value sr_state_val, double lo, double hi){
   state_t* st = (state_t*) alloca(sizeof(state_t));
   fill_from_value(sr_state_val,st);
   // double lo = Double_val(lo_val);
   // double hi = Double_val(hi_val);
   double result = next_float_sr(st,lo,hi);
-  CAMLreturn(result);
+  return result;
 }
 
 CAMLprim value float_c_sr_unchecked(value sr_state_val, value lo_val, value hi_val){
@@ -211,15 +210,11 @@ CAMLprim value int_c_sr_log_uniform(value sr_state_val, value lo_val, value hi_v
 // }
 
 static inline double bitcast_int64_to_double(int64_t value) {
-  double result;
-  memcpy(&result, &value, sizeof(double));
-  return result;
+  return *(double*)&value;
 }
 
 static inline int64_t bitcast_double_to_int64(double value) {
-  int64_t result;
-  memcpy(&result, &value, sizeof(int64_t));
-  return result;
+  return *(int64_t*)&value;
 }
 
 static inline int64_t to_int64_preserve_order(double t){
@@ -240,8 +235,7 @@ static inline double of_int64_preserve_order(int64_t x){
   }
 }
 
-CAMLprim double one_ulp_up_c_sr_unboxed(double x){
-  CAMLparam0();
+double one_ulp_up_c_sr_unboxed(double x){
   double res;
   if(isnan(x)){
     res = NAN;
@@ -256,8 +250,7 @@ CAMLprim value one_ulp_up_c_sr(value x_val){
   CAMLreturn(caml_copy_double(one_ulp_up_c_sr_unboxed(Double_val(x_val))));
 }
 
-CAMLprim double one_ulp_down_c_sr_unboxed(double x){
-  CAMLparam0();
+double one_ulp_down_c_sr_unboxed(double x){
   double res;
   if(isnan(x)){
     res = NAN;

### 2025-03-10 staged-ocaml/lib/randomness/random_stub.c
fix extremely dumb bug

@@ -44,6 +44,7 @@ static struct custom_operations state_ops = {
 };
 
 #define State_val(v) (*((state_t *) Data_custom_val(v)))
+#define State_ptr(v) ((state_t *) Data_custom_val(v))
 
 /*
 NOTE: We could get around this single allocation (which will happen at the beginning of every generator call)
@@ -137,14 +138,14 @@ double next_float(state_t *st, double lo, double hi){
 
 CAMLprim value bool_c(value state_val) {
   CAMLparam1(state_val);
-  state_t st = State_val(state_val);
-  CAMLreturn(Val_bool(next_bool(&st)));
+  state_t *st = State_ptr(state_val);
+  CAMLreturn(Val_bool(next_bool(st)));
 }
 
 double float_c_unchecked_unboxed(value state_val, double lo, double hi){
   // CAMLparam1(state_val);
-  state_t st = State_val(state_val);
-  double result = next_float(&st,lo,hi);
+  state_t *st = State_ptr(state_val);
+  double result = next_float(st,lo,hi);
   return result;
 }
 
@@ -156,12 +157,12 @@ CAMLprim value float_c_unchecked(value state_val, value lo_val, value hi_val){
 
 CAMLprim value int_c_unchecked(value state_val, value lo_val, value hi_val) {
   CAMLparam3(state_val,lo_val,hi_val);
-  state_t st = State_val(state_val);
+  state_t *st = State_ptr(state_val);
 
   int64_t lo = (int64_t) Long_val(lo_val);
   int64_t hi = (int64_t) Long_val(hi_val);
 
-  int64_t result = next_int(&st,lo,hi);
+  int64_t result = next_int(st,lo,hi);
   CAMLreturn(Val_int(result));
 }
 
@@ -209,12 +210,12 @@ int64_t next_int_log_uniform(state_t *st, int64_t lo, int64_t hi) {
 
 CAMLprim value int_c_log_uniform(value state_val, value lo_val, value hi_val) {
   CAMLparam3(state_val,lo_val,hi_val);
-  state_t st = State_val(state_val);
+  state_t *st = State_ptr(state_val);
 
   int64_t lo = (int64_t) Long_val(lo_val);
   int64_t hi = (int64_t) Long_val(hi_val);
 
-  int64_t result = next_int_log_uniform(&st,lo,hi);
+  int64_t result = next_int_log_uniform(st,lo,hi);
   CAMLreturn(Val_int(result));
 }
 

### 2025-03-10 staged-ocaml/lib/randomness/sr_random_stub.c

@@ -145,6 +145,7 @@ CAMLprim value int_c_sr_unchecked(value sr_state_val, value lo_val, value hi_val
   int64_t hi = (int64_t) Long_val(hi_val);
 
   int64_t result = next_int_sr(st,lo,hi);
+
   CAMLreturn(Val_int(result));
 }
 

### 2025-03-10 staged-ocaml/lib/randomness/sr_random_stub.c
perf work on randomness

@@ -26,9 +26,6 @@ static inline double unit_float_from_int64(int64_t n){
   return ((uint64_t) n >> 11) * pow(2,-53);
 }
 
-
-
-
 /*
 Here, a state_t is just a "view" into an existing Splittable_random.State.t.
 The seed is a pointer to the actual SR's Int64.t (techncially immutable) custom block.
@@ -39,95 +36,98 @@ typedef struct state {
   int64_t odd_gamma;
 } state_t;
 
-void fill_from_value(value v, state_t* st){
+static state_t global_state = {
+  .seed = NULL,
+  .odd_gamma = 0
+};
+
+void fill_from_value(value v){
     int64_t *seed = (int64_t*) Data_custom_val(Field(v,0));
     int64_t odd_gamma = Int64_val(Field(v,1));
-    st->seed = seed;
-    st->odd_gamma = odd_gamma;
+    global_state.seed = seed;
+    global_state.odd_gamma = odd_gamma;
 }
 
-int64_t next_seed_sr(state_t *s){
-  int64_t next = *(s->seed) + s->odd_gamma;
-  *(s->seed) = next;
+int64_t next_seed_sr(){
+  int64_t next = *global_state.seed + global_state.odd_gamma;
+  *global_state.seed = next;
   return next;
 }
 
-
-
-int64_t next_int64_sr(state_t* st){
-  return mix64(next_seed_sr(st));
+int64_t next_int64_sr(){
+  return mix64(next_seed_sr());
 }
 
-bool next_bool_sr(state_t *st){
-  int64_t draw = next_int64_sr(st);
+bool next_bool_sr(){
+  int64_t draw = next_int64_sr();
   return ((draw | 1L) == draw);
 }
 
-int64_t between_sr(state_t* st,int64_t lo,int64_t hi){
-  int64_t draw = next_int64_sr(st);
+int64_t between_sr(int64_t lo,int64_t hi){
+  int64_t draw = next_int64_sr();
   while (lo > draw || hi < draw){
-    draw = next_int64_sr(st);
+    draw = next_int64_sr();
   }
   return draw;
 }
 
-int64_t non_negative_up_to_sr(state_t *st, int64_t max){
+int64_t non_negative_up_to_sr(int64_t max){
   int64_t draw;
   int64_t remainder;
   do {
-    draw = next_int64_sr(st) & INT64_MAX;
+    draw = next_int64_sr() & INT64_MAX;
[…]

### 2025-03-11 paper/paper.tex
some updates

@@
 
-\definecolor{dkred}{rgb}{0.7,0,0}
-\definecolor{dkblue}{rgb}{0,0,0.7}
-\definecolor{notered}{rgb}{0.85,0,0}
-\definecolor{dkpurple}{HTML}{4e02eb}
-\definecolor{dkgreen}{HTML}{006329}
-\definecolor{teal}{HTML}{007982}
-\definecolor{fuchsia}{HTML}{8C368C}
-
-\newcommand{\comm}[3]{\ifdraft\textcolor{#1}{[#2: #3]}\fi}
-\newcommand{\cn}{\ifdraft{\color{blue} [CITE]}\fi}
-\newcommand{\hg}[1]{\comm{dkred}{HG}{#1}}
-\newcommand{\tr}[1]{\comm{dkblue}{TR}{#1}}
-\newcommand{\bcp}[1]{\comm{dkpurple}{BCP}{#1}}
-\newcommand{\jwc}[1]{\comm{dkgreen}{JWC}{#1}}
-\newcommand{\outline}[1]{\ifdraft{\color{teal} #1}\fi}
-
-\newcommand{\bq}{\texttt{base\_quickcheck}}
-\newcommand{\name}{\texttt{waffle\_house}}
-\newcommand{\tool}{Waffle House}
-
-\newcommand{\secref}[1]{\S\ref{#1}}
-
-\newcommand{\proposecut}[1]{\ifdraft{\color{gray} [#1]}\fi}
-\newcommand{\proposechange}[2]{\ifdraft{\color{gray} {#1}
-    $\rightarrow$ #2}\else #2\fi}
+\include{macros}
 
@@
 \author{Joseph W. Cutler}
-\affiliation{%
+\affiliation{
   \institution{University of Pennsylvania}
@@
   \state{Pennsylvania}
-  \country{USA}}
+  \country{USA}
+}
 
@@
 %% of authors' names for this purpose.
-\renewcommand{\shortauthors}{Richey et al.}
+\renewcommand{\shortauthors}{Richey and Cutler et al.}
 
@@
 \section{Introduction}
+
+\jwc{This paper is about ``Strict functional languages like OCaml and Scala'', but we'll focus on OCaml for presentation.}
+
 Property-based testing (PBT) is a widely-used testing framework consisting of
@@
 bound to find a bug. Therefore, it is imperative that PBT libraries are built
-with efficiency in mind.
+with performance in mind.
 
@@
     \item An empirical analysis of the sources of inefficiency in PBT
-    generators. 
-    \item A library, \name, which offers efficient generator functions through the
+    generators. \jwc{And an argument that PBT generator libraries should be engineered for performance}
+    \item A library, \name, which offers efficient \jwc{performant?} generator \proposechange{functions}{combinators} through the
     use of staged metaprogramming; this is the first known application of metaprogramming to PBT.
@@
 
+\jwc{For the purposes of this paper, we will only use two stages: compile time, and run time.}
+
 \jwc{
@@
       \item Consider writing \texttt{even . succ}. This
-      \item Consider \texttt{even_c : int code -> bool code = fun cx -> .<.~cx mod 2 == 0>.} and \texttt{succ_c : int code -> int code = fun cx -> .< .~x + 1 >.}
-      \item If you do \texttt{to_dyn (even_c . succ_c)}, you get \texttt{fun x -> (x + 1) mod 2 == 0}. Composing functions from code to code, and then \emph{only at the end} stamping out
-      a dynamic function value ensures that the resulting code is free of abstractions.
+      \item Consider \texttt{even\_c : int code -> bool code = fun cx -> .<.~cx mod 2 == 0>.} and \texttt{succ\_c : int code -> int code = fun cx -> .< .~x + 1 >.}
+      \item If you do \texttt{to\_dyn (even\_c . succ\_c)}, you get \texttt{fun x -> (x + 1) mod 2 == 0}. Composing functions from code to code, and then \emph{only at the end} stamping out
+      a dynamic function value eliminates the function abstraction.
     \end{itemize}
-    and 
+    \item This is the basis of (WORD). By defining a library with functions with types like \texttt{'a code -> 'b code}, we can ensure that the
+    abstractions the library introduces are fully eliminated at compile time.
   \end{itemize}
@@
 \label{subsection:other-langs}
+
+\jwc{TODO: Move this section later (cf conversation in WH meeting 3/7/25)}
 
@@
 
-\tr{Discuss BQ, including type-derived generators.}
-
 \section{Why is \bq\ slow, and how can we fix it?}
 
-\jwc{We can simplify this further by getting rid of the size parameter. Make the story even cleaner, I think!}
+\jwc{NOTE: We can simplify this further by getting rid of the size parameter. Make the story even cleaner, I think!}
 
@@
   \begin{itemize}
+    \item It's been long-known that clean functional abstractions have a runtime overhead (this should be familiar by this time in the paper).
     \item How does (simplified) BQ work?
@@
 
+\jwc{
+The problem is plain: while the abstractions that PBT libraries like base quickcheck provide are indespensible for writing idiomatic generators, the performance overhead of using them is dramatic.
+The compiler (and flambda) uses heuristics (\href{https://ocaml.org/manual/5.3/flambda.html}{here}) to decide when to specialize and inline code, and these heuristics cannot guarantee that generator abstractions are zero-cost.
+These heuristics are also extremely conservative about inlining and specializing recursive functions, which all generators of recursive data types must be.
+}
+
 \tr{Each of these consist of an explanation of the problem and pseudocode outlining the solution.}
@@
   \begin{itemize}
-    \item In cases where the compiler cannot statically eliminate them, binds allocate short-lived closures: we allocate a closure for the continuation, and then immediately jump into it.
+    \item While it's the ``correct'' abstraction for generators, using a monadic interface prevents the compiler from specializing generator code.
+Using monadic bind and return obscures the control and data flow of a generator from the compiler.
+This is
+    compounded when handling recursive functions.
+    \item In cases where the compiler cannot statically eliminate them, running a monadic bind allocates a short-lived closure: we allocate a closure for the continuation, and then immediately jump into it.
     \item Each closure allocation is relatively cheap (they're going to be minor allocations since they're short-lived), but doing lots of allocation in a generation hot loop adds up fast, and each minor allocation brings us closer to the next costly GC pause.
@@
 
-\jwc{
-  \begin{itemize}
-    \item The pRNG at the heart of BQ is the hottest part of the hot path: generating a single int with \texttt{int st lo hi} samples from the underlying rng many many times.
-    \item BQ's prng is the SplitMix algorithm, implemented in the splittablerandom package.
-    \item The precise details of how SplitMix works are unimportant for this paper, but the core of it does an enormous amount of 64-bit int arithmetic.
-    \item Because of implementation details related to the garbage collector, OCaml's \texttt{int} type is 63 bits, so an \texttt{Int64.t} is actually \emph{boxed}: a pointer to a custom block, not scanned by the garbage collector.
-    \item Because of this, all \texttt{Int64.t} operations (both arithmetic and bitwise operations) must allocate!
-A single call to \texttt{int} in Bq calls the samples the RNG many times, and each sample (call to \texttt{next\_int64}) allocates 9 words (2 multiplications, one addition, and 6 logical operations).
-  \end{itemize}
-}
+The core of any PBT generator library is a random-number generator. Following
+the original Haskell QuickCheck, \bq uses the SplitMix algorithm \cn,
+implemented as an OCaml library called \texttt{Splittable\_random}. The precise
+details of how SplitMix works are unimportant for the present paper, but the
+main operations are defined in terms of bitwise operations on 64-bit integers.
+However, due to implementation details related to the garbage collector, all
+runtime OCaml values are either (a) pointers to a block of memory, or (b) 63-bit ``immediate'' values.
+As such, the OCaml type of 64-bit ingeters, \texttt{int64}, is \emph{boxed}: it
+is a pointer to a block of memory that contains a 64-bit integer.  This means
+that \emph{all} \texttt{int64} operations (both arithmetic and bitwise) must
+allocate memory cells to contain their output.
 
-\jwc{We discuss ways to fix this in section ...}
+This has a significant impact on the performance of generators. A single call to
+one of the \bq library functions may sample from \texttt{Splittable\_random}
+multiple times, and each sample from \texttt{Splittable\_random} allocates 9
+words \jwc{this is a call to \texttt{next\_int64}}.  While seemingly small, we
+will see in Section~\ref{section:eval} that this amount of allocation has a
+significant impact on significant.
 
-\jwc{We note that this inefficiency is not due to the ``abstraction overhead'' of the DSL itself, but simply due to language implementation and algorithm choices.
-We include this to demonstrate the amount of slack in the system, and so we can demonstrate the additional testing-time savings in the evaluation section.
-}
+In Section~\ref{subsection:faster-rng}, we show how our library mitigates this
+performance hit by using OCaml's FFI to reimplement the RNG in non-allocating C code.
+We also discuss how the ``unboxed types'' extension to the OCaml compiler \cn could instead be used
+to the same effect without leaving OCaml.
+
+It is worth noting that the inefficiency of boxed \texttt{int64}s is not due to
+the abstraction overhead of the generator DSL itself, but simply due to language
+implementation and algorithm choices. We include this discussion --- and
+the corresponding evaluation
+--- to demonstrate the testing-time savings of taking seriously the notion that PBT libraries
+should be engineered for performance.
 
@@
     \item Bind takes code for the size and seed, and returns code that first let-binds the result of the first generator.
-    The function application \texttt{g size_c random_c} runs during the first stage, producing the code for the \texttt{a}.
+    The function application \texttt{g size\_c random\_c} runs during the first stage, producing the code for the \texttt{a}.
 This is then spliced into the quote: the specialized code of the generator \texttt{g}
     put on the right-hand side of the let. Then, we splice in the code for the continuation, passing first the variable x, and the same size and random seed.
-    \item \texttt{}
   \end{itemize}
@@
 \subsection{Faster Random Number Generators}
+\label{subsection:faster-rng}
 
@@
 }
+
+\jwc{All of this can be done with Jane Street bleeding-edge compiler extensions, though these are not yet compatible with MetaOCaml.}
 

### 2025-03-11 staged-ocaml/repin.sh
pinner boi

@@ -1,0 +1,23 @@
+#!/bin/bash
+
+# Run dune clean
+echo "Running 'dune clean'..."
+dune clean
+
+# Run dune build
+echo "Running 'dune build'..."
+dune build
+
+# Unpin the package
+echo "Running 'opam unpin .'..."
+opam unpin .
+
+# Delete line 21 of fast_gen.opam
+echo "Deleting line 21 of fast_gen.opam..."
+sed -i '21d' fast_gen.opam
+
+# Pin the package
+echo "Running 'opam pin .'..."
+opam pin .
+
+echo "All operations completed successfully."

### 2025-03-11 staged-ocaml/repin.sh
evil

@@ -20,4 +20,6 @@ sed -i '21d' fast_gen.opam
 echo "Running 'opam pin .'..."
 opam pin .
 
+opam pin ../ppx_staged
+
 echo "All operations completed successfully."

### 2025-03-11 paper/paper.tex
incorporate bcp comments

@@
 \jwc{This paper is about ``Strict functional languages like OCaml and Scala'', but we'll focus on OCaml for presentation.}
+
+\bcp{The current introduction says, roughly, ``We took a look at
+  base.quickcheck, noticed some inefficiencies, applied known
+  techniques from staged metaprogramming to eliminate them, and
+  observed that speed went up.''  I think we can make much stronger
+  claims, but I'm not sure about exactly how strong or what belongs in
+  the foreground...
+  \begin{itemize}
+  \item One of the first serious uses of staged metaprogramming in anger?
+  \item A use of staged metaprogramming that is novel / challenging in
+  itself, in some way?
+  \item An analysis of sources of inefficiency across a range of PBT
+  frameworks (and a solution that applies to many of them)?
+  \item A usable tool that addresses and overcomes some significant
+  implementation challenges?
+  \item Careful measurements showing where the sources of inefficiency
+  are in existing PBT tools, and which ones matter most?
+  \item (Your claim here...)?
+  \end{itemize}
+}
 
@@
 \begin{lstlisting}
-
 module Gen = struct
@@
 
+\bcp{How much faster could we go if we changed the effect ordering?
+  If it might be a lot, then should we measure this too?}
+
 \subsection{Size Parameters, A Real Monad Instance, CPS, and ``The Trick''}

### 2025-03-11 paper/paper.tex
intro comments

@@
   \begin{itemize}
-  \item One of the first serious uses of staged metaprogramming in anger?
-  \item A use of staged metaprogramming that is novel / challenging in
+  \item One of the first serious uses of staging in anger?
+  \item A use of staging that is novel / challenging in
   itself, in some way?
@@
   \end{itemize}
+}
+
+\jwc{Agree: I think something like the 3rd is the best reframing. Something
+like: ``We demonstrate that, across languages, monadic PBT generator DSLs can
+have a significant performance overhead, and present a cross-language technique
+for eliminating the overhead while preserving the idiomatic abstraction.
+''}
+
+\jwc{Another angle that I'd argue needs to come out more is that we demonstrate that
+engineering PBT libraries with performance in mind has significant impact on testing power.
 }
@@
 \bq’s functionality while improving performance through
-\textit{staged metaprogramming} (staging). Staging is a lightweight,
+\textit{multi-stage programming} (staging). Staging is a lightweight,
 domain-specific compilation technique that allows us to completely eliminate the
@@
     \item A library, \name, which offers efficient \jwc{performant?} generator \proposechange{functions}{combinators} through the
-    use of staged metaprogramming; this is the first known application of metaprogramming to PBT.
+    use of multi-stage programming; this is the first known application of staging to PBT.
     \item The insight that generators should be compared by program equality 
@@
 
-\subsection{Staged Metaprogramming}
+\subsection{Multi-Stage Programming}
+
+Muli-stage programming (also known more simply as staging) is \jwc{finish}
 
@@
 
-\jwc{Staged metaprogramming has a long history as a technique for eliminating the runtime overhead of abstractions,
+\jwc{Staging has a long history as a technique for eliminating the runtime overhead of abstractions,
 specializing programs at an earlier stage to run more efficiently at a later stage.}

### 2025-03-11 paper/paper.tex
staging is hard

@@
 
-\jwc{In this paper, we'll use OCaml to illustrate the ideas, but the concepts are portable to other languages: see Section~\ref{subsection:other-langs}}
+\jwc{In this paper, we'll use OCaml to illustrate the ideas, but the concepts
+are portable to other languages: see Section~\ref{subsection:other-langs}}
 
@@
 
-Muli-stage programming (also known more simply as staging) is \jwc{finish}
-
-\jwc{A good quick presentation of staged metaprogramming is here, in section 6.
-https://www.cl.cam.ac.uk/~jdy22/papers/a-typed-algebraic-approach-to-parsing.pdf
+\jwc{
+  \begin{itemize}
+    \item In Muli-stage programming (also known more simply as staging), programs execute over the course of multiple stages,
+with each stage producing the code to be run in the next.
+\item This is an old idea, with roots going back to quasiquotation in LISP, and picking up steam again in the 1990s with MetaML \cn{}.
+\item For the purposes of this paper, we will only consider two stages: compile time and run time. Staged programs thus execute twice: once at compile time,
+which produces more code, which is then compiled and run at run time.
+\item Staging has many applications, but chief among them is its use for \emph{optimization}. We can write programs such that during the compile time stage, they are partially evaluated to eliminate
+abstraction overhead. 
+\item Many languages have some degree of staging functionality, either provided as a library \cn or build directly into the language's implementation \cn.
+\item For this paper, we use MetaOCaml \cn for OCaml staging, but all of the staging concepts are portable to any other language with multi-stage programming functionality.
+  \end{itemize}
 }
 
-\jwc{Staging has a long history as a technique for eliminating the runtime overhead of abstractions,
-specializing programs at an earlier stage to run more efficiently at a later stage.}
+\jwc{This section is going to need work.}
 
-\jwc{For the purposes of this paper, we will only use two stages: compile time, and run time.}
+MetaOCaml's staging functionality is exposed through a type \texttt{'a code}. A
+value of type \texttt{t code} at compile time is a (potentially open) OCaml term
+of type \texttt{t}.
+
+Values of \texttt{code} type are introduced by \emph{quotes}, written \texttt{.<$\ldots$>.}. Brackets
+delay execution of a program until run time. For example, the program \texttt{.< 5 + 1 >.} has type \texttt{int code}.
+Note that this is not the same as \texttt{.< 6 >.}. Because brackets delay computation, the code is not \emph{executed}
+until the next stage (run time).
+
+Values of type \texttt{code} can be combined together using \emph{escape}, written \texttt{.\~{}($\ldots$)}.
+Escape lets you take a value of type \texttt{code}, and ``splice'' it directly into a quote.
+For example, this program \texttt{let x = .<1 * 5>. in .< .\~{}(x) + .\~{}(x) >.} evaluates to \texttt{.<(1 * 5) + (1 * 5)>.}.
 
@@
   \begin{itemize}
-    \item The \texttt{'a code} type, quote, splice, stage distinction.
+    \item The \texttt{'a code} type, quote, escape or ``splice'', stage distinction.
+    \item MetaOCaml ensures correct scoping and hygene by preventing alpha-collision when you splice open terms.
     \item The difference between \texttt{('a -> 'b) code} and \texttt{'a code -> 'b code}
@@
 But same diff.
+    \item Python. Maybe could do it?? Hypothesis + MacroPy
     \item Haskell: GHC does a lot of these optimizations already, since the code is pure. Since QC generators are relatively small programs,

### 2025-03-11 paper/paper.tex
continuing to explain

@@
 
-Values of \texttt{code} type are introduced by \emph{quotes}, written \texttt{.<$\ldots$>.}. Brackets
-delay execution of a program until run time. For example, the program \texttt{.< 5 + 1 >.} has type \texttt{int code}.
-Note that this is not the same as \texttt{.< 6 >.}. Because brackets delay computation, the code is not \emph{executed}
-until the next stage (run time).
+Values of \texttt{code} type are introduced by \emph{quotes}, written
+\texttt{.<$\ldots$>.}. Brackets delay execution of a program until run time. For
+example, the program \texttt{.< 5 + 1 >.} has type \texttt{int code}.  Note that
+this is not the same as \texttt{.< 6 >.}. Because brackets delay computation,
+the code is not \emph{executed} until the next stage (run time).
+Values of type \texttt{code} can be combined together using \emph{escape},
+written \texttt{.\~{}($\ldots$)}.  Escape lets you take a value of type
+\texttt{code}, and ``splice'' it directly into a quote.  For example, this
+program \texttt{let x = .<1 * 5>. in .< .\~{}(x) + .\~{}(x) >.} evaluates to
+\texttt{.<(1 * 5) + (1 * 5)>.}.  MetaOCaml ensures correct scoping and macro
+hygiene, ensuring that variables are not shadowed when open terms are spliced.
 
-Values of type \texttt{code} can be combined together using \emph{escape}, written \texttt{.\~{}($\ldots$)}.
-Escape lets you take a value of type \texttt{code}, and ``splice'' it directly into a quote.
-For example, this program \texttt{let x = .<1 * 5>. in .< .\~{}(x) + .\~{}(x) >.} evaluates to \texttt{.<(1 * 5) + (1 * 5)>.}.
+The power of staging for optimization away abstraction overheads comes from
+defining functions that accept and return \texttt{code} values.  A function
+\texttt{f : 'a code -> 'b code} is a function that takes a program computing a
+run-time \texttt{'a} and transforms it into a program computing a run-time
+\texttt{'b}. In particular, because \texttt{f} itself runs at compile time, the
+abstraction of using \texttt{f} is necessarily completely eliminated by
+run-time.
+A code-transforming function \texttt{'a code -> 'b code} can also be converted \emph{code for a function} --- a value of type
+\texttt{('a -> 'b) code} --- with the following program:
+\begin{lstlisting}
+let eta (f : 'a code -> 'b code) : ('a -> 'b) code = .<fun x -> .\~(f .<x>.)>.
+\end{lstlisting}
+
+This program is known as ``the trick'' in the partial evaluation and multi-stage programming literature.
+\cn{} 
+% USE THIS: https://arxiv.org/pdf/2309.08207
+
+\jwc{words}
+
+As an example, consider
+\begin{lstlisting}
+let is_even x = .< .~x mod 2 == 0 >. in
+let succ x = .< 1 + .~x >. in
+eta (succ . is_even)
+\end{lstlisting}
+This is equal to \texttt{.< fun x -> (1 + x) mod 2 == 0 >.}: the functions have been fused together, and the overhead of calling them is not present at runtime.
+
+This ``trick'' is how we use metaprogramming
 

### 2025-03-11 paper/paper.tex
msp explanation

@@
 % USE THIS: https://arxiv.org/pdf/2309.08207
+It returns a code for a function that takes an argument \texttt{x}, and then it splats in the result of calling \texttt{f} on just
+the quoted \texttt{x}.
+\jwc{Does this make any sense?}
 
-\jwc{words}
-
-As an example, consider
+The trick is best illustrated by an example. The following program reduces to \texttt{.< fun x -> (1 + x) mod 2 == 0 >}.
 \begin{lstlisting}
@@
 \end{lstlisting}
-This is equal to \texttt{.< fun x -> (1 + x) mod 2 == 0 >.}: the functions have been fused together, and the overhead of calling them is not present at runtime.
+By composing the two code-transforming functions together at compile time, and only then turning them into a run-time function,
+the functions are fused together... \jwc{words}.
+This is the basis of how staging is used to eliminate the abstraction of DSLs
+(like a generator DSL). By writing DSL combinators as compile-time functions ---
+and only calling \texttt{eta} at the end on the completed DSL program ---  we
+can ensure that any overhead of using them is eliminated by run time.
 
-This ``trick'' is how we use metaprogramming
-
-\jwc{
-  \begin{itemize}
-    \item The \texttt{'a code} type, quote, escape or ``splice'', stage distinction.
-    \item MetaOCaml ensures correct scoping and hygene by preventing alpha-collision when you splice open terms.
-    \item The difference between \texttt{('a -> 'b) code} and \texttt{'a code -> 'b code}
-    \item We can convert one way, but not the other.
-    \item Functions \texttt{'a code -> 'b code} ``fuse''.
-    \begin{itemize}
-      \item Consider writing \texttt{even . succ}. This
-      \item Consider \texttt{even\_c : int code -> bool code = fun cx -> .<.~cx mod 2 == 0>.} and \texttt{succ\_c : int code -> int code = fun cx -> .< .~x + 1 >.}
-      \item If you do \texttt{to\_dyn (even\_c . succ\_c)}, you get \texttt{fun x -> (x + 1) mod 2 == 0}. Composing functions from code to code, and then \emph{only at the end} stamping out
-      a dynamic function value eliminates the function abstraction.
-    \end{itemize}
-    \item This is the basis of (WORD). By defining a library with functions with types like \texttt{'a code -> 'b code}, we can ensure that the
-    abstractions the library introduces are fully eliminated at compile time.
-  \end{itemize}
-}
+% \jwc{
+%   \begin{itemize}
+%     \item The \texttt{'a code} type, quote, escape or ``splice'', stage distinction.
+%     \item MetaOCaml ensures correct scoping and hygene by preventing alpha-collision when you splice open terms.
+%     \item The difference between \texttt{('a -> 'b) code} and \texttt{'a code -> 'b code}
+%     \item We can convert one way, but not the other.
+%     \item Functions \texttt{'a code -> 'b code} ``fuse''.
+%     \begin{itemize}
+%       \item Consider writing \texttt{even . succ}. This
+%       \item Consider \texttt{even\_c : int code -> bool code = fun cx -> .<.~cx mod 2 == 0>.} and \texttt{succ\_c : int code -> int code = fun cx -> .< .~x + 1 >.}
+%       \item If you do \texttt{to\_dyn (even\_c . succ\_c)}, you get \texttt{fun x -> (x + 1) mod 2 == 0}. Composing functions from code to code, and then \emph{only at the end} stamping out
+%       a dynamic function value eliminates the function abstraction.
+%     \end{itemize}
+%     \item This is the basis of (WORD). By defining a library with functions with types like \texttt{'a code -> 'b code}, we can ensure that the
+%     abstractions the library introduces are fully eliminated at compile time.
+%   \end{itemize}
+% }
 

### 2025-03-11 paper/paper.tex
a whole bunch more

@@
 
-\subsection{\bq}
-
 \section{Why is \bq\ slow, and how can we fix it?}
@@
 
-\jwc{Usual introduction to this section, corresponding to how we talked about it in the intro}
-
-\jwc{
-  \begin{itemize}
-    \item To eliminate this runtime overhead and generate efficient code, we do a usual ``binding-time analysis''.
-    \item The crux of this binding-time analysis is that while generators are `just' values of type \texttt{'a t}, in practice they are always \emph{completely statically defined}: one never
-    computes a generator dynamically at test time.
-    \item The size and the random seed are only known at run-time (the later stage), but the code of the generator itself is known at compile time.
-    \item This means the generator type should be \texttt{int code -> SR.t code -> 'a code}: a statically known-function from dynamically-known int and seed to dynamically-determined result.
-  \end{itemize}
-}
+\jwc{Usual introduction to this section, corresponding to how we talked about it in the intro. ``We present a library that XYZ''.}
 
@@
 
-A generator of \texttt{'a}s is then a (compile-time) function \texttt{int code -> Random.t code -> 'a code}.
+Recall that staging a DSL requires carefully changing some of the types and
+combinators to transform \texttt{code}s.  Deciding which types \texttt{t} can
+instead be \texttt{t code} is an art known as ``binding-time analysis'' \cn:
+figuring out which parts of the DSL can be determined statically (and can be
+part of the compile-time stage), and which parts are only known dynamically (and
+hence must be \texttt{code}).
+
+The crux of our binding time analysis is that particular size random seed are
+only known at run-time (the later stage), but the code of the generator itself
+is known at compile time.  This means our library's generator type \texttt{'a
+Gen.t} should be \texttt{int code -> SR.t code -> 'a code}: a compile-time
+function from dynamically-known size and seed to dynamically-determined result.
 
@@
     let sum_code = List.foldr _ in
-    let go (rc : int code) (weighted_elts : (float code * 'a code) list) : 'a code =
+    let pick (rc : int code) (weighted_elts : (float code * 'a code) list) : 'a code =
       match weighted_elts with
@@
         let r = SR.int .~random_c 0 sum in
-        .~(go .<r>. gwcs)
+        .~(pick .<r>. weighted_elts)
       >.
@@
 
-\subsection{Size Parameters, A Real Monad Instance, CPS, and ``The Trick''}
+\subsection{Size Parameters, A Real Monad Instance, CPS, and Recursion}
 
+\subsubsection{Size Parameters}
 \jwc{Introduce the size parameter here. (if you deleted it beforehand.)}
 
+\subsubsection{CodeCPS and a Monad Instance}
 \jwc{
@@
 
+\subsubsection{Recursive Generators}
+\jwc{Describe recursion and recurisve handles}
+
 \subsubsection{The Trick}
+\jwc{Describe split --- this section is approximately experts-only, but you can just point at the Andras paper and the things it cites.}
 

### 2025-03-12 paper/paper.tex
Add anon and review
Intro

@@
-\documentclass[sigplan,screen,acmsmall]{acmart}%%
+\documentclass[sigplan,screen,acmsmall,anonymous,review]{acmart}%%
 %% \BibTeX command to typeset BibTeX logo in the docs
@@
 
-Property-based testing (PBT) is a widely-used testing framework consisting of
+Property-based testing (PBT) is a widely-used testing framework\hg{PBT isn't a
+framework, it's a ``technique'' or ``approach'' --- Base\_QuickCheck is a framework} consisting of
 two key components: a set of \textit{properties} that a system must satisfy, and
+a way of producing
 a large number of \textit{inputs} to that system. In contrast to traditional
-unit testing, where inputs are hand-written, PBT users create a random-input
+unit testing, where test inputs are hand-written, PBT users create a random input
 \textit{generator} that automatically produces inputs. Sometimes, this process
@@
 
-Writing ``good" generators—those that uncover more bugs in the system under
+\hg{I think this paragraph buries the lede. I'd start the paragraph talking
+about performance and citing the 50ms number, and then jump right into what's
+slow.}
+Writing ``good'' generators—ones that successfully uncover bugs in the system under
 test—is challenging. Much effort has been dedicated to this problem, often by
@@
 generators, we explore popular libraries across several widely-used programming
-languages. We find that the flexibility of these libraries comes at a cost to 
+languages.\hg{Again, this sentence also starts the paragraph too slowly; rather than
+focus on our exploration (which isn't really the point of the paper) lead with
+the result --- that abstraction overhead, boxing/unboxing, and allocations slow
+down QuickCheck ports} We find that the flexibility of these libraries comes at a cost to 
 performance,
@@
 on OCaml's \bq, selected for its efficiency relative to other OCaml PBT libraries 
-and its integration with industrial build systems, where it is expected to run 
+and its integration with industrial build systems\hg{Let's not try to justify
+our choice of OCaml \bq here --- we should just talk about strict functional
+languages and justify this choice later}, where it is expected to run 
 within tight time constraints. These
@@
 
-Our solution is \name, a generator library that preserves
+Our solution is \name, a generator library\hg{Can we say ``an approach to PBT
+generator libraries'' or something more general like that?} that preserves
 \bq’s functionality while improving performance through
@@
 
-We present a new methodology for evaluating the relative effectiveness of PBT
-generators. Prior work has compared generators based on bug-finding ability: 
+We \hg{also} present a new methodology for evaluating the relative effectiveness of PBT
+generators.\hg{Might be an over-claim, we'll need to write this carefully}
+Prior work has compared generators based on bug-finding ability: 
 generators are considered equally effective if they find bugs in a
@@
 optimizations that enhance performance without compromising expressiveness.
+\hg{I'm not totally convinced by this paragraph. Here's how I'd explain this:\\
+In order to justify the switching cost associated with using \name, we designed
+\name\ to be incredibly easy to benchmark: most PBT generator optimizations make
+changes up to {\em distributional equivalence}, allowing the precise sampling
+order to differ as long as the probability distribution is maintained, but
+\name's optimizations are correct up to {\em program equivalence}. This means
+that a \name\ generator produces precisely the same values in the same order as
+the original generator, only faster. With this setup, benchmarking becomes
+significantly easier than for prior work---our measurements have incredibly low
+variance, and performance gains from \name\ translate directly to improved
+bug-finding speed.}
 
@@
 In summary, we dramatically improve the bug-finding effectiveness of PBT generators
-by optimizing them for speed. In speicific, we make the following contributions:
+by optimizing them for speed. Specifically, we make the following contributions:
 \begin{enumerate}
     \item An empirical analysis of the sources of inefficiency in PBT
-    generators. \jwc{And an argument that PBT generator libraries should be engineered for performance}
+    generators. \jwc{And an argument that PBT generator libraries should be
+    engineered for performance}
+    \hg{Let's talk about this, I'm not sure I buy it}
     \item A library, \name, which offers efficient \jwc{performant?} generator \proposechange{functions}{combinators} through the
-    use of multi-stage programming; this is the first known application of staging to PBT.
+    use of multi-stage programming; this is the first known application of
+    staging to PBT.\hg{Let's also talk about this}
     \item The insight that generators should be compared by program equality
@@
     using our library in a controlled comparison.
-
 \end{enumerate}

### 2025-03-12 paper/paper.tex
blah

--- 2025-03-11 paper/paper.tex
@@
 abstraction of using \texttt{f} is necessarily completely eliminated by
-run-time.
+run time.
 A code-transforming function \texttt{'a code -> 'b code} can also be converted \emph{code for a function} --- a value of type
@@
 The crux of our binding time analysis is that particular size random seed are
-only known at run-time (the later stage), but the code of the generator itself
-is known at compile time.  This means our library's generator type \texttt{'a
+only known at run time (the later stage), but the code of the generator itself
+is known at compile time.
+
+Generators --- values of type \texttt{'a Bq.t} --- are never constructed dynamically at run time in practice,
+so the
+
+This means our library's generator type \texttt{'a
 Gen.t} should be \texttt{int code -> SR.t code -> 'a code}: a compile-time
@@
 
-\jwc{This is 3-stage metaprogramming! There's PPX-time, compile time, and run-time.}
+\jwc{This is 3-stage metaprogramming! There's PPX time, compile time, and run time.}
 

### 2025-03-12 staged-ocaml/lib/randomness/random_stub.c
repopulate seeds

@@ -283,3 +283,21 @@ CAMLprim value one_ulp_down_c(value x_val){
   CAMLparam1(x_val);
   CAMLreturn(caml_copy_double(one_ulp_down_c_unboxed(Double_val(x_val))));
 }
+
+// void fill_from_value(value v){
+//     int64_t *seed = (int64_t*) Data_custom_val(Field(v,0));
+//     int64_t odd_gamma = Int64_val(Field(v,1));
+//     global_state.seed = seed;
+//     global_state.odd_gamma = odd_gamma;
+// }
+
+CAMLprim value repopulate(value state_val, value sr_state_val) {
+  CAMLparam2(state_val, sr_state_val);
+
+  state_t s = State_val(state_val);
+
+  Int64_val(Field(sr_state_val,0)) = s.seed;
+  Int64_val(Field(sr_state_val,1)) = s.odd_gamma;
+  
+  CAMLreturn(Val_unit);
+}

### 2025-03-12 paper/paper.tex
some notes from meeting

@@
 
+\jwc{People have been using staged metaprogrammign to eliminate abstraction
+overheads in parsing for a long time, we turn that lens on generation (Which looks like parsing)}
+
 \jwc{Another angle that I'd argue needs to come out more is that we demonstrate that
 engineering PBT libraries with performance in mind has significant impact on testing power.
+}
+
+\jwc{
+Let's lead on with the quote from PBT in practice: performance engineering
+matters.  De-emphasize the abstraction overhead framing. Eliminating abstraction
+overhead is important, so is using the fastest RNG you can possibly use.
 }
@@
 \subsection{Multi-Stage Programming}
+\label{subsection:msp}
 
@@
 \section{Why is \bq\ slow, and how can we fix it?}
+\jwc{Why is a monadic generator library slow?}
 
@@
 
+\jwc{If you're not using the fasteest RNG possible, you're leaving RNG on the table.}
+
 \section{\name: A Library for Staged Generators}
@@
 
-The crux of our binding time analysis is that particular size random seed are
-only known at run time (the later stage), but the code of the generator itself
-is known at compile time.
-
-Generators --- values of type \texttt{'a Bq.t} --- are never constructed dynamically at run time in practice,
-so the
+The crux of our binding time analysis is that the particular random seed and
+size parameter \jwc{if explained} are only known at run time (the later stage),
+but the code of the generator itself is known at compile time. Generators ---
+values of type \texttt{'a Bq.t} --- are never constructed dynamically at run
+time in practice, so all of the combinators we use to build them can run at compile time.
 
@@
 
-\jwc{Note that some of this is a lie, but it's a good first cut: point to the section on \emph{The Trick} to see the real version.}
+This type, along with basic monadic generator DSL functionality can be found in Figure~\ref{fig:gen-staged-basic}.
+The monadic interface is given by a return and bind, as usual.
+\texttt{return} is the constant generator, but this time it operates on codes. Given \texttt{cx : 'a code}, the code for a \texttt{'a}, it returns the generator which always generates that value.
+\texttt{bind g k} sequences generators by passing the result of running the generator \texttt{g} to a continuation \texttt{k}. However, instead of getting access
+to the particular value generated by \texttt{g}, the continuation \texttt{k} gets access to \texttt{code} for the value sampled from \texttt{g}:
+at compile time, we only know \texttt{g} will generate \emph{some} \texttt{'a}, but not which one.
+Operationally, bind takes code for the size and seed, and returns code that (1) let-binds a variable \texttt{a} to spliced-in code that runs \texttt{g}, and then
+(2) runs the spliced-in continuation \texttt{k}.
+Both function applications \texttt{g size\_c random\_c}
+and \texttt{k .<a>. size\_c random\_c} run at compile time.
+\texttt{Gen.int} is the generator that samples an int from the RNG.
+Given any size and random seed, it returns a code block that calls \texttt{SR.int} with that random seed. Because the lower and upper bounds
+might not be known at compile time --- they may themselves be the results of calling \texttt{Gen.int} ---
+the arguments \texttt{lo} and \texttt{hi} are of type \texttt{int code}, and get spliced into the code block as arguments to \texttt{SR.int}.
+Lastly, \texttt{to_bq} turns a staged generator into code for a normal \bq
+generator. This function is just a 2-argument version of ``The Trick'' (\texttt{eta}
+from Section~\ref{subsection:msp}).
 
-\jwc{
-  \begin{itemize}
-    \item The simplest combinators are presented below.
-    \item \texttt{return} is the constant generator, given code for \texttt{x : 'a}, it gives the generator which always generates \texttt{x}
-    \item \texttt{bind} is like BQ's bind --- sequencing generators by passing the result of one to another --- except that the continuation gets access to the
-    \texttt{code} for the result of running the first generator.
-At compile time, we know only that \texttt{g} will return \emph{some} \texttt{a}, but we do not get access to the particular \texttt{a}.
-    \item Bind takes code for the size and seed, and returns code that first let-binds the result of the first generator.
-    The function application \texttt{g size\_c random\_c} runs during the first stage, producing the code for the \texttt{a}.
-This is then spliced into the quote: the specialized code of the generator \texttt{g}
-    put on the right-hand side of the let. Then, we splice in the code for the continuation, passing first the variable x, and the same size and random seed.
-  \end{itemize}
-}
-
+\begin{figure}
 \begin{lstlisting}
@@
       .<
-        let x = .~(g size_c random_c) in
-        .~(k .<x>. size_c random_c)
+        let a = .~(g size_c random_c) in
+        .~(k .<a>. size_c random_c)
       >.
@@
 \end{lstlisting}
+\caption{Basic Staged Generator Library}
+\ref{fig:gen-staged-basic}
+\end{figure}
 
-\jwc{Conversion to a bq generator}
-
-\jwc{Then show the running example, with }
+Returning to our running example, we can define a staged generator for pairs of ints
 
@@
 \end{lstlisting}
+
+\jwc{Changing some identifier names }
 

### 2025-03-12 paper/paper.tex
blah

@@
 
+\jwc{
+  Things to ensure people come away with answers to:
+  \begin{itemize}
+    \item Is there generality?
+    \item Is this important?
+    \item Did they actually solve the problem?
+  \end{itemize}
+}
+
 \section{Background}
@@
 
-\section{Why is \bq\ slow, and how can we fix it?}
+\section{Why are Monadic Generator Libraries Slow?}
 \jwc{Why is a monadic generator library slow?}
@@
 
-\section{\name: A Library for Staged Generators}
+\section{Eliminating Abstraction Overhead of Generator DSLs by Staging}
+\jwc{Emphasize that this is an \emph{COMPLETELY EQUIVALENT drop in replacement!} We din't just build a different library with different distributions.}
 
@@
 the arguments \texttt{lo} and \texttt{hi} are of type \texttt{int code}, and get spliced into the code block as arguments to \texttt{SR.int}.
-Lastly, \texttt{to_bq} turns a staged generator into code for a normal \bq
+Lastly, \texttt{to\_bq} turns a staged generator into code for a normal \bq
 generator. This function is just a 2-argument version of ``The Trick'' (\texttt{eta}

### 2025-03-12 paper/paper.tex
blah

@@
 \jwc{Why is a monadic generator library slow?}
-
 \jwc{NOTE: We can simplify this further by getting rid of the size parameter. Make the story even cleaner, I think!}
+
+\jwc{IMPORTANT: We are using OCaml because we have to pick a syntax for this section, but these abstraction overheads exist in other languages -- at least scala, rust.}
 
@@
 
+\jwc{Demonstrate this in Scala too!!!}
+
 \jwc{
@@
 \tr{Each of these consist of an explanation of the problem and pseudocode outlining the solution.}
-\subsubsection{Monadic Bind}
+\subsubsection{Monadic DSL Abstraction Overhead}
 
@@
     \item In cases where the compiler cannot statically eliminate them, running a monadic bind allocates a short-lived closure: we allocate a closure for the continuation, and then immediately jump into it.
-    \item Each closure allocation is relatively cheap (they're going to be minor allocations since they're short-lived), but doing lots of allocation in a generation hot loop adds up fast, and each minor allocation brings us closer to the next costly GC pause.
+    \item Each closure allocation is relatively cheap (they're going to be minor allocations since they're short-lived), but doing lots of allocation in a generation hot loop adds up fast.
+    \item Similar things have been noticed about monadic parsing DSLs in the past.
   \end{itemize}
@@
 
-\subsubsection{Expensive Combinators}
+\subsubsection{Combinator Abstraction Overhead}
 \jwc{Many combinators like \texttt{union} and \texttt{weighted\_union} needlessly allocate large data structures in the generation hot path.
@@
 }
+\jwc{Ensure this discussion is initially multi-language, but uses OCaml to demo.}
 
@@
 
+\jwc{
+  \begin{itemize}
+    \item Because the RNG sits at the heart of the generation hot path, RNG speed matters enormously.
+    \item To illustrate this, we demonstrate a particular inefficiency of
+    OCaml's BQ, resolve it in a later section, and then show how much you can
+    bug-finding performance just by speeding up and RNG.
+    \item We also demonstrate how using an inequivalent RNG
+    \item Scalacheck uses a faster RNG called ``xoroshiro128++'' \cn{}.
+    % https://lemire.me/blog/2019/03/19/the-fastest-conventional-random-number-generator-that-can-pass-big-crush/
+    \item One could go even faster, using \href{https://lemire.me/blog/2019/03/19/the-fastest-conventional-random-number-generator-that-can-pass-big-crush/}{even faster RNG} with less guarantees.
+  \end{itemize}
+}
+
+\jwc{The inefficiency of BQ's RNG presents a ``natural experiment'' (in the
+sense of economics): we can demonstrate how using a faster RNG can improve
+bugfinding time, but while maintainign program equivlanece. (This should go in the last paragraph here...)}
+
 The core of any PBT generator library is a random-number generator. Following
@@
 
-\subsection{Faster Random Number Generators}
+\subsection{Fast Random Number Generators}
 \label{subsection:faster-rng}
@@
   \begin{itemize}
-    \item 
+    \item Using the fastest RNG possible is important.
+    \item In OCaml, the splittalbe random RNG is not as fast as it could be, because of the Ints issue.
+    \item In this section, we describe how to resolve the ints issue by calling out ot the FFI. 
+    \item All of this can be done with Jane Street bleeding-edge compiler extensions, though these are not yet compatible with MetaOCaml.
+    \item Because different libraries 
   \end{itemize}
 }
-
-\jwc{All of this can be done with Jane Street bleeding-edge compiler extensions, though these are not yet compatible with MetaOCaml.}
 

### 2025-03-12 paper/paper.tex
reframe

@@
 }
+
+\jwc{``Property based testing is a race against time...''}
 
@@
 % \subsubsection{Function call overhead}
-\subsubsection{Boxed RNG}
+\subsubsection{Choice of RNG}
 
@@
   \begin{itemize}
-    \item Because the RNG sits at the heart of the generation hot path, RNG speed matters enormously.
-    \item To illustrate this, we demonstrate a particular inefficiency of
-    OCaml's BQ, resolve it in a later section, and then show how much you can
-    bug-finding performance just by speeding up and RNG.
-    \item We also demonstrate how using an inequivalent RNG
-    \item Scalacheck uses a faster RNG called ``xoroshiro128++'' \cn{}.
-    % https://lemire.me/blog/2019/03/19/the-fastest-conventional-random-number-generator-that-can-pass-big-crush/
-    \item One could go even faster, using \href{https://lemire.me/blog/2019/03/19/the-fastest-conventional-random-number-generator-that-can-pass-big-crush/}{even faster RNG} with less guarantees.
+    \item Because the RNG sits at the heart of the generation hot path, RNG
+    speed matters enormously. Which RNG you pick matters...
+
+    \item Different PBT libraries use different RNGs --- QuickCheck and \bq use
+    splittable_random, Scalacheck uses \href{https://burtleburtle.net/bob/rand/smallprng.html}{this one}.  These
+    are reasonably performant RNGs, but faster ones exist. Because this is the
+    core of the hot path of the generator, speeding up sampling just a little can
+    can dramatically speed up generation time.
+    \item Significantly faster RNGs that still pass pRNG benchmarks like BigCrush exist (like \href{https://github.com/lemire/testingRNG/blob/master/source/lehmer64.h}{this one}), and using them can 
+    speed up generation enormously.
+
+    \item However, showing this
+    is tricky, because different RNGs have different sampling orders, but our
+    eval is made much stronger by the fact that all of our generators are functionally equivalent.
+
+    \item In this paper demonstrate this fact by noticing a ``natural experiment'': OCaml's splittable random library is
+    slow in a way that can be improved \emph{without} changing the extensional behavior. By resolving this issue, we can demonstrate
+    just how much you can improve bug-finding performance simply by speeding up the RNG.
+
+    % To illustrate this, we demonstrate a particular inefficiency of
+    % OCaml's BQ, resolve it in a later section, and then show how much you can
+    % bug-finding performance just by speeding up and RNG.
   \end{itemize}

### 2025-03-12 paper/paper.tex
scala overhead

@@
 let int_pair : (int * int) Bq.t =
-  Bq.bind (Bq.gen_int 0 100) (fun x ->
-    Bq.bind (Bq.gen_int 0 x) (fun y ->
+  let%bind x = (Bq.gen_int 0 100) in
+  let%bind y = (Bq.gen_int 0 x) in
   Bq.return (x,y)
-  ))
 
@@
 \jwc{Show benchmark difference between these two generators: the benchmark is in \texttt{waffle-house/handwritten-ocaml/bin/basic-compare.ml}. It's about 2x. (see comment in latex here.)}
-
-\jwc{The native code OCaml compiler, even with -O3, fails to specialize this code and eliminate the overhead of this abstraction --- inlining all of the funciton definitions and then performing beta-reductions speeds up sampling by a factor of 2.}
 
@@
 
-\jwc{Demonstrate this in Scala too!!!}
+\jwc{The native code OCaml compiler, even with -O3, fails to specialize this code and eliminate the overhead of this abstraction --- inlining all of the funciton definitions and then performing beta-reductions speeds up sampling by a factor of 2.}
+
+\jwc{Same story in scala. Approximately 2x difference for the same generator... (scala uses a slightly different generator type, but the principles are the same)}
+
+\begin{lstlisting}
+def intPair : Gen[(Long,Long)] = for {
+  x <- Gen.choose(0,1000)
+  y <- Gen.choose(0,x)
+} yield (x,y)
+
+def intPairInlined : (Gen.Parameters, Seed) => (Option[(Long,Long)],Seed) = {
+ (p,seed) =>
+  val (x,seed2) = chLng(0,1000)(p,seed)
+  x match {
+    case None => (None,seed2)
+    case Some(x) =>
+      val (y,seed3) = chLng(0,x)(p,seed2)
+      y match {
+        case None => (None,seed)
+        case Some(y) => (Some(x,y),seed3)
+      }
+    }
+}
+\end{lstlisting}
+
+% [info] GenBm.generateComplexBase     avgt    3  458.325 ± 67.214  ns/op
+% [info] GenBm.generateComplexInlined  avgt    3  266.035 ±  8.559  ns/op
 
@@
     speed matters enormously. Which RNG you pick matters...
-
     \item Different PBT libraries use different RNGs --- QuickCheck and \bq use
-    splittable_random, Scalacheck uses \href{https://burtleburtle.net/bob/rand/smallprng.html}{this one}.  These
+    splittable random, Scalacheck uses \href{https://burtleburtle.net/bob/rand/smallprng.html}{this one}.  These
     are reasonably performant RNGs, but faster ones exist. Because this is the
@@
     speed up generation enormously.
-
-    \item However, showing this
-    is tricky, because different RNGs have different sampling orders, but our
+    \item However, showing this is tricky, because different RNGs have different sampling orders, but our
     eval is made much stronger by the fact that all of our generators are functionally equivalent.
-
     \item In this paper demonstrate this fact by noticing a ``natural experiment'': OCaml's splittable random library is
@@
     just how much you can improve bug-finding performance simply by speeding up the RNG.
-
-    % To illustrate this, we demonstrate a particular inefficiency of
-    % OCaml's BQ, resolve it in a later section, and then show how much you can
-    % bug-finding performance just by speeding up and RNG.
   \end{itemize}

### 2025-03-13 paper/paper.tex

@@
 
-\jwc{Same story in scala. Approximately 2x difference for the same generator... (scala uses a slightly different generator type, but the principles are the same)}
-
 \begin{lstlisting}
@@
 
+\jwc{Same story in scala. 
+Approximately 2x difference for the same generator (though an order of magnitude slower overall than bq)... (scala uses a slightly different generator type, but the principles are the same)
+This doesn't even account for more inlining we could do to eliminate the boxing/unboxing of results. (Scalacheck generators can fail, though this adds significant overhead)
+}
+
 % [info] GenBm.generateComplexBase     avgt    3  458.325 ± 67.214  ns/op
@@
 The problem is plain: while the abstractions that PBT libraries like base quickcheck provide are indespensible for writing idiomatic generators, the performance overhead of using them is dramatic.
-The compiler (and flambda) uses heuristics (\href{https://ocaml.org/manual/5.3/flambda.html}{here}) to decide when to specialize and inline code, and these heuristics cannot guarantee that generator abstractions are zero-cost.
-These heuristics are also extremely conservative about inlining and specializing recursive functions, which all generators of recursive data types must be.
+Modern compilers use heuristics to decide when to specialize and inline code \cn, and these heuristics cannot guarantee that generator abstractions are zero-cost.
+These heuristics are also necessarily conservative about inlining and specializing recursive functions, which all generators of recursive data types must be.
 }

### 2025-03-13 staged-ocaml/lib/randomness/random_stub.c
tentative?

@@ -289,8 +289,7 @@ CAMLprim value repopulate(value state_val, value sr_state_val) {
 
   state_t s = State_val(state_val);
 
-  Int64_val(Field(sr_state_val,0)) = s.seed;
-  Int64_val(Field(sr_state_val,1)) = s.odd_gamma;
-  
+  caml_modify(&Field(sr_state_val, 0), caml_copy_int64(s.seed));
+
   CAMLreturn(Val_unit);
 }

### 2025-03-13 staged-ocaml/lib/randomness/random_stub.c
more randomness nonsense

@@ -143,7 +143,6 @@ CAMLprim value bool_c(value state_val) {
 }
 
 double float_c_unchecked_unboxed(value state_val, double lo, double hi){
-  // CAMLparam1(state_val);
   state_t *st = State_ptr(state_val);
   double result = next_float(st,lo,hi);
   return result;
@@ -286,10 +285,12 @@ CAMLprim value one_ulp_down_c(value x_val){
 
 CAMLprim value repopulate(value state_val, value sr_state_val) {
   CAMLparam2(state_val, sr_state_val);
+  CAMLlocal1(seed_val);
 
   state_t s = State_val(state_val);
+  seed_val = caml_copy_int64(s.seed);
 
-  caml_modify(&Field(sr_state_val, 0), caml_copy_int64(s.seed));
+  caml_modify(&Field(sr_state_val, 0), seed_val);
 
   CAMLreturn(Val_unit);
 }

### 2025-03-13 paper/paper.tex
weighted unions and let insertion

@@
 
+\jwc{
+  \begin{itemize}
+    \item PBT generator libraries include more functions than just the monad interface and sampling, libraries include helper functions.
+    \item With \texttt{weighted\_union}, to make a weighted choices, and \texttt{fixed\_point} to define recursive generators.
+    \item Also \texttt{size} and \texttt{with\_size} to adjust the size parameter --- these are useful to ensure we terminate with recursive generators.
+  \end{itemize}
+}
+
+\jwc{This section needs to explain approximately how weighted union works --- compute a sum of the weights, sample between 0 and the sum, pick
+the first element where the partial sums to the left exceeds the sampled number.
+(This has been written about a billion times, cite one of the papers about this.)
+}
+
+\begin{lstlisting}
+let tree_of g = fixed_point (fun rg ->
+  let%bind n = size in
+  weighted_union [
+    (1, return E);
+    (n,
+      let%bind x = g in
+      let%bind l = with_size (n / 2) rg in
+      let%bind r = with_size (n / 2) rg in
+      return (Node (l,x,r))
+    )
+  ]
+)
+\end{lstlisting}
+
+\jwc{This generator generates binary trees using all the combinators, yay. Explain em.}
+
+\jwc{Explain effect ordering here!!}
+
 \subsection{Multi-Stage Programming}
 \label{subsection:msp}
+
+\jwc{I think this section should actually go later.}
 
@@
 Values of type \texttt{code} can be combined together using \emph{escape},
-written \texttt{.\~{}($\ldots$)}.  Escape lets you take a value of type
+written \texttt{.\~{}(e)} (or just \texttt{.\~{}x}, when \texttt{x} is a variable).  Escape lets you take a value of type
 \texttt{code}, and ``splice'' it directly into a quote.  For example, this
@@
 
-\section{Why are Monadic Generator Libraries Slow?}
+\section{Sources of Inefficiency in Monadic Generator Libraries}
+\label{section:motiv}
 \jwc{Why is a monadic generator library slow?}
@@
 
+\subsubsection{Monadic DSL Abstraction Overhead}
 \jwc{
@@
 \jwc{
-The problem is plain: while the abstractions that PBT libraries like base quickcheck provide are indespensible for writing idiomatic generators, the performance overhead of using them is dramatic.
+The problem is plain: while the monadic abstraction that PBT libraries like base quickcheck provide are indespensible for writing idiomatic generators, the performance overhead of using them is dramatic.
 Modern compilers use heuristics to decide when to specialize and inline code \cn, and these heuristics cannot guarantee that generator abstractions are zero-cost.
@@
 \tr{Each of these consist of an explanation of the problem and pseudocode outlining the solution.}
-\subsubsection{Monadic DSL Abstraction Overhead}
 
@@
     \item In cases where the compiler cannot statically eliminate them, running a monadic bind allocates a short-lived closure: we allocate a closure for the continuation, and then immediately jump into it.
-    \item Each closure allocation is relatively cheap (they're going to be minor allocations since they're short-lived), but doing lots of allocation in a generation hot loop adds up fast.
+    \item Each closure allocation is relatively cheap, but doing lots of allocation in a generation hot loop adds up fast.
     \item Similar things have been noticed about monadic parsing DSLs in the past.
@@
 \subsubsection{Combinator Abstraction Overhead}
-\jwc{Many combinators like \texttt{union} and \texttt{weighted\_union} needlessly allocate large data structures in the generation hot path.
-Using a combination of staging and more careful algorithm design, we can avoid this overhead.
-}
-\jwc{Ensure this discussion is initially multi-language, but uses OCaml to demo.}
-
-\begin{lstlisting}
-  let of_weighted_list alist =
-  let weights, values = List.unzip alist in
-  let value_array = Array.of_list values in
-  let total_weight, cumulative_weight_array =
-    let array = Array.init (Array.length value_array) ~f:(fun _ -> 0.) in
-    let sum =
-      List.foldi weights ~init:0. ~f:(fun index acc weight ->
-        let acc' = acc +. weight in
-        array.(index) <- acc';
-        acc'
-      )
-    in
-    sum, array
-  in
-  create (fun ~size:_ ~random ->
-    let choice = Splittable_random.float random ~lo:0. ~hi:total_weight in
-    match
-      Array.binary_search
-        cumulative_weight_array
-        ~compare:Float.compare
-        `First_greater_than_or_equal_to
-        choice
-    with
-    | Some index -> value_array.(index)
-    | None -> assert false
-   )
-\end{lstlisting}
-
-\jwc{This builds the cdf of the distribution, samples from 0 to the total, and binary searches through the array to find the bucket.}
-\jwc{Allocating this array is unnecessary: we can simply compute the total weight, sample \texttt{x} between 0 and the total, and then walk the list accumulating the sum, until the accumulator exceeds \texttt{x}.
+
+\jwc{
+Aside from the usual fact that the compiler can't ``see through'' the abstraction boundary to specialize, combinators like union and weighted union
+necessarily allocate a lot in the generation hot path.
+Even without the dumb BQ array thing, to call union and weighted union, the types mean that you have to allocate a list.
+These allocations are extremely costly, and happen every sample from the generator. If the generator is recursive (like the tree generator), they occur at each recursive call.
+}
+
+\jwc{In practice, the possible options are almost always statically known: in the binary tree generator from earlier, we can tell from the text of the code what
+the weights are, what the generators are, and how many there are.
+Put more simply, you basically always call weighted union with an \emph{explicit list}, which is then immediately traversed by the weighted union.
+For this reason, you should not have to incur the cost of allocating this list at runtime.
+However, almost no compilers (GHC is a notable exception) can figure this out and eliminate the list \cn (list fusion).
+}
+
+% \jwc{This builds the cdf of the distribution, samples from 0 to the total, and binary searches through the array to find the bucket.}
+% \jwc{Allocating this array is unnecessary: we can simply compute the total weight, sample \texttt{x} between 0 and the total, and then walk the list accumulating the sum, until the accumulator exceeds \texttt{x}.
 Since in practice the list of possible options is quite small (usually at most 10 in practice), the linear time scan is going to be much faster.}
-\jwc{(The point of this para will just be to emphasize that it's almost always better to be allocation-aware, instead of algorithmically clever.)}
-
-\jwc{Moreover, the possible options is almost always statically known in practice: you use weightedunion to generate (say) a datatype with possible
-variants, whose options are known at compile time.
-Put more simply, you basically always call weighted union with an \emph{explicit list}: For this reason, you should not have to incur the cost of allocating this list at runtime.}
-
-\jwc{ofc, sometimes you only know the list at runtime (see the STLC generator for an example), so you the library also includes this.}
+% \jwc{(The point of this para will just be to emphasize that it's almost always better to be allocation-aware, instead of algorithmically clever.)}
 
@@
 \subsubsection{Choice of RNG}
+
+\jwc{We really need to have explained the effect ordering equivalence before here.}
 
@@
     splittable random, Scalacheck uses \href{https://burtleburtle.net/bob/rand/smallprng.html}{this one}.  These
-    are reasonably performant RNGs, but faster ones exist. Because this is the
+    are reasonably performant RNGs. Because this is the
     core of the hot path of the generator, speeding up sampling just a little can
@@
     \item Significantly faster RNGs that still pass pRNG benchmarks like BigCrush exist (like \href{https://github.com/lemire/testingRNG/blob/master/source/lehmer64.h}{this one}), and using them can 
-    speed up generation enormously.
+    speed up generation enormously --- in our microbenchmarks, the lehmer generator runs 4x faster than splittable random.
     \item However, showing this is tricky, because different RNGs have different sampling orders, but our
@@
 }
-
-\jwc{The inefficiency of BQ's RNG presents a ``natural experiment'' (in the
-sense of economics): we can demonstrate how using a faster RNG can improve
-bugfinding time, but while maintainign program equivlanece. (This should go in the last paragraph here...)}
 
@@
 \subsection{Basic Design}
-
+\label{subsection:basic-design}
 Recall that staging a DSL requires carefully changing some of the types and
@@
 to the particular value generated by \texttt{g}, the continuation \texttt{k} gets access to \texttt{code} for the value sampled from \texttt{g}:
-at compile time, we only know \texttt{g} will generate \emph{some} \texttt{'a}, but not which one.
+at compile time, we only know \texttt{g} will generate \emph{some} \texttt{'a}, but not which one \footnote{Readers familiar with OCaml may notice that \texttt{return : 'a code -> 'a Gen.t} and \texttt{bind : 'a Gen.t -> ('a code -> 'b Gen.t) -> 'b Gen.t}
+do not have the correct types for a monad instance, preventing us from using \texttt{let\%bind} notation.
+We rectify this issue in Section~\ref{subsection:codecps} by importing some clever ideas from the staging literature.}.
 Operationally, bind takes code for the size and seed, and returns code that (1) let-binds a variable \texttt{a} to spliced-in code that runs \texttt{g}, and then
@@
 
-Returning to our running example, we can define a staged generator for pairs of ints
-
-\begin{lstlisting}
-let int_pair_staged : ((int * int) code) Gen.t =
+Returning to our running example, Figure~\ref{fig:running-staged} shows the
+int-pair example written with the staged \texttt{Gen.t} monad, as well as the
+inlined code that results from calling \texttt{Gen.to\_bq} (changing some identifier names for clarity).
+The code generated is identical to the manually inlined version from Section~\ref{section:motiv}, and of course runs equally fast.
+
+\begin{figure}
+\begin{lstlisting}
+let int_pair_staged : (int * int) Gen.t =
   Gen.bind (Gen.int .<0>. .<100>.) (fun cx ->
@@
   )
+
 let int_pair : (int * int) Bq.t code = Gen.to_bq int_pair_staged
-====
-.<
-  fun size random ->
+(* .< fun size random ->
         let x = SR.int random 0 100 in
@@
         (x,y)
-.>
-\end{lstlisting}
-
-\jwc{Changing some identifier names }
-
-\jwc{So if you write with the \texttt{Gen} combinators and then only call \texttt{to\_bq} at the end, you get something that inlines everything all the way down!}
-
-\subsection{Staging More Combinators}
-
-\jwc{Weighted unions: recall that we should not need to materialize}
-
-% let rec genpick n ws =
-% match ws with
-% | [] -> { rand_gen = fun ~size_c:_ ~random_c:_ -> Codecps.return .< failwith "Fell of the end of pick list" >. }
-% | (k,g) :: ws' ->
-%       { rand_gen = 
-%         fun ~size_c ~random_c ->
-%           Codecps.bind (Codecps.split_bool .< Float.compare .~(v2c n) .~(v2c k) <= 0 >.) (fun leq ->
-%             if leq then
-%               g.rand_gen ~size_c ~random_c
-%             else
-%               Codecps.bind (Codecps.let_insertv .< .~(v2c n) -. .~(v2c k) >.) @@ fun n' ->
-%               (* let%bind n' =  in *)
-%               (genpick n' ws').rand_gen ~size_c ~random_c
-%         )
-%       }
-
-\begin{lstlisting}
-  let weighted_union (weighted_elts : (float code * 'a code) list) : 'a t =
-    let sum_code = List.foldr _ in
-    let pick (rc : int code) (weighted_elts : (float code * 'a code) list) : 'a code =
-      match weighted_elts with
+    >.
+*)
+\end{lstlisting}
+\caption{Pairs of Ints, Staged}
+\ref{fig:running-staged}
+\end{figure}
+
+\jwc{Is there anything more we need to say here?}
+
+\subsection{Staging Combinators}
+
+In Section~\ref{section:motiv}, we noticed that generator combinators like \texttt{weighted\_union}
+must allocate lists in the hot path of the generator. Even though these lists are often small ---
+usually at most a few dozen elements in practice --- each allocation takes us closer to the next garbage collection.
+
+This is an ideal opportunity to exercise another use of staging: compile-time specialization.
+Because we almost always know the particular list of choices at compile time, a staged version of \texttt{weighted\_union}
+can generate \emph{different code} depending on the number of generators that are being unioned together.
+If we use weighted union on a list (known at compile time) of generators \texttt{g1}, \texttt{g2}, and \texttt{g3},
+we can emit code that never has to create a list of three elements at run time.
+
+\begin{figure}
+\begin{lstlisting}
+module Gen =
+$\dots$
+  let pick (acc : int code) (weighted_gens : (int code * 'a t) list) (size : int code) (random : Sr.t code) : 'a code =
+    match weighted_gens with
     | [] -> .< failwith "Error" >.
-      | (wc,xc) :: elts' ->
+    | (wc,g) :: gens' ->
       .<
-          if .~rc <= .~wc then .~xc
+        if .~acc <= .~wc then .~(g size random)
         else
-            let r' = .~rc - .~wc in
-            .~(go .<r'>. elts')
+          let acc' = .~acc - .~wc in
+          .~(pick .<acc'>. gens' size random)
       >.
 
-    fun size_c random_c ->
+  let weighted_union (weighted_gens : (int code * 'a t) list) : 'a t =
+    let sum_code = List.foldr (fun acc (w,_) -> .<.~acc + .~w>. ) .<0>. weighted_gens in
+    fun size random ->
       .<
@@
         let r = SR.int .~random_c 0 sum in
-        .~(pick .<r>. weighted_elts)
+        .~(pick .<r>. weighted_gens size random)
       >.
 \end{lstlisting}
-
-Then, if we write
-
-\begin{lstlisting}
-let grades : charBq.t gen = to_bq (
-  weighted_union [
-    (.<3.0>., .<'a'>.);
-    (.<2.0>., .<'b'>.);
-    (.<1.0>., .<'c'>.);
+\caption{Staged Weighted Union}
+\label{fig:staged-weighted-union}
+\end{figure}
+
+Figure~\ref{fig:staged-weighted-union} shows the code for such a staged weighted union.
+Crucially, it takes a \emph{compile-time} list \texttt{weighted\_gens} of
+generators and weights. The weights themselves might only be known at run time --- it is
+common to use the current size parameter as a weight, for instance ---
+so they are \texttt{code}s.
+\texttt{Gen.weighted\_union} begins by computing \texttt{sum\_code}, an \texttt{int code}
+that is the sum of the weights. Note that this happens at compile time: we fold over a list
+known at compile time to produce another code value. 
+We then call \texttt{SR.int} to sample a random number \texttt{r} between \texttt{0} and the sum.
+Finally, we splice in the result of calling the helper function \texttt{pick}. \texttt{pick}
+produces a tree of \texttt{if}s by again traversing the list of generators at compile time.
+This tree of \texttt{ifs} ``searches'' for the genrator corresponding to the sampled value \texttt{r}, and then runs it.
+
+Figure~\ref{fig:staged-weighted-union-example} demonstrates a use of this staged weighted union.
+Given a list (in this case constant) generators with weights ``the current size parameter'', \texttt{2}, and \texttt{1},
+the generated code first computes the sum of these numbers, samples between \texttt{0} and the sum, and then
+traverses a tree of three \texttt{if}s to find the correct value to return.
+
+\begin{figure}
+\begin{lstlisting}
+let grades : char Bq.t = Gen.to_bq (
+  Gen.bind size (fun n ->
+    Gen.weighted_union [
+      (n, Gen.return .<'a'>.);
+      (.<2>., Gen.return .<'b'>.);
+      (.<1>., Gen.return .<'c'>.);
     ]
   )
-===
-Bq.create (
-  fun size random ->
-    let sum = 3.0 + 2.0 + 1.0 in
+)
+(*
+.< fun size random ->
+    let sum = size + 2 + 1 + 0 in
     let r = SR.int random 0 sum in
-    if r <= 3.0 then 'a'
+    if r <= size then 'a'
     else
-      let r' = r - 3.0 in
-      if r' <= 2.0 then 'b'
+      let r' = r - size in
+      if r' <= 2 then 'b'
       else
-        let r'' = r' - 2.0 in
-        if r'' <= 1.0 then 'c'
+        let r'' = r' - 2 in
+        if r'' <= 1 then 'c'
         else
           failwith "Error"
-)
-\end{lstlisting}
-
-\subsection{Effect Ordering, and Equivalence is Program Equality}
-\jwc{``Wait, why was the defintion of bind like that''?
-Careful readers might note that the definition of bind was a little strange. In particular, why not just do this?}
-\begin{lstlisting}
-  let bind (g : 'a gen) (k : 'a code -> 'b gen) : 'b gen =
-    fun size_c random_c ->
-      k (g size_c random_c) size_c random_c
-\end{lstlisting}
-\jwc{The answer is that this doesn't generate the right code! In particular, this definition of bind means that
-\texttt{bind (int 0 1) (fun x -> return .<(.~x,.~x)>.)} would generate \texttt{fun size random -> (int random 0 1, int random 0 1)},
-which is \emph{not} the desired behavior. (Cite cbn/cbv randomness here).
-A splice is a (hygenic) syntactic replacment, so
-if a function \texttt{f : 'a code -> 'b code} splices its argument multiple times into its output, the argument code will be directly copied in.
-This is why we insert a let-binding in the definition of bind, to ensure that
-the explicit sequencing of bind is preserved into explicit sequencing in the
-generated code.
-}
-
-\jwc{
-The library is carefully design to preserve exactly the effect order of BQ. If
-you write the same generator in both, they run identically: for the same size
-and random seed, they produce the same value.
-}
-
-\jwc{Note: tradeoff between maintaining the effect ordering and even more performance.}
-
-\bcp{How much faster could we go if we changed the effect ordering?
-  If it might be a lot, then should we measure this too?}
-
-\subsection{Size Parameters, A Real Monad Instance, CPS, and Recursion}
-
-\subsubsection{Size Parameters}
-\jwc{Introduce the size parameter here. (if you deleted it beforehand.)}
-
-\subsubsection{CodeCPS and a Monad Instance}
+>.
+*)
+\end{lstlisting} 
+\caption{Use of Staged Weighted Union}
+\label{fig:staged-weighted-union-example}
+\end{figure}
+
+\subsection{Let-Insertion and Effect Ordering}
+Careful readers might note that the definition of \texttt{bind} in Section~\ref{subsection:basic-design}
+was more complicated than one might expect. In particular, why not define bind in a more standard way: as
+\texttt{let bind' g k = fun size random -> k (g size random) size random}, without the code block
+that let-binds the spliced code \texttt{.\~{}(g size random)}?
+Unfortunately, using \texttt{Gen.bind'} leads to incorrect code being generated.
+For example, consider \texttt{Gen.bind' (Gen.int .<0>. .<1>.) (fun x -> Gen.return .<(.~x,.~x)>.)}.
+This generates the run time code \texttt{fun size random -> (SR.int random 0 1, int SR.random 0 1)},
+which is incorrect. 
+This is not equivalent to the behavior of writing the same code with base quickcheck \jwc{Which is a property we want to hold for our eval.}.
+Instead of generating a single integer and returning it twice, it samples two different integers.
+
+This problem is intimately related to nondeterministic effects in the presence of CBN/CBV evaluation ordering \cn.
+In essence, the behavior of splice \texttt{.\~{}cx} in a staged function \texttt{f(cx : 'a code) = ...} is to \emph{copy}
+the entire block of code, effects and all. To ensure that the randomness effects of the first generator are executed only once,
+but that the value can be used in the continuation multiple times, \texttt{bind} let-binds the result of generation to a variable,
+and then passes that to the continuation.
+
+The library \jwc{ensure this is consistent with the way we talk about the project, cf cross-language} is careully designed to 
+preserve exactly the effect order of base quickcheck \jwc{and the scala version to preserve the effect ordering of ScalaCheck}.
+\jwc{This fact should go earlier. Not really sure about where this subsection should actually land, but it needs to be said somehwere.}
+
+\subsection{CodeCPS and a Monad Instance}
+\label{subsection:codecps}
 \jwc{
@@
 
-\subsubsection{Recursive Generators}
-\jwc{Describe recursion and recurisve handles}
-
 \subsubsection{The Trick}
 \jwc{Describe split --- this section is approximately experts-only, but you can just point at the Andras paper and the things it cites.}
+
+\subsection{Recursive Generators}
+\jwc{Describe recursion and recurisve handles}
 
@@
 \section{Conclusion \& Future Work}
+\section{Related Work}
+\jwc{A generator is a parser of randomness, and people have implemented lots of staged parser libraries!}
 %%

### 2025-03-13 paper/paper.tex
progress

@@
 \label{subsection:codecps}
+
+This is all great so far, but there's a subtle issue that prevents the version of the library design discussed so far
+from being used as a proper drop-in replacement for an existing generator DSL: the types of \texttt{return : 'a code -> 'a Gen.t}
+and \texttt{bind : 'a Gen.t -> ('a code -> 'b Gen.t) -> 'b Gen.t} aren't quite right.
+For the type \texttt{'a Gen.t} to actually be a monad, 
+the these types cannot mention \texttt{code}. This is not just a theoretical issue, it is a significant usability concern:
+the syntactic sugar for monadic programming (\texttt{let\%bind} in OCaml, \texttt{foreach} in Scala, \texttt{do} in Haskell, etc)
+that makes it so appealing van \emph{only} be used if the types involved are actually the proper monad function types.
+
+To support real monadic programming, we'll need to adjust the type of \texttt{'a Gen.t} slightly.
+An initial attempt is to try \texttt{type 'a t = int code -> SR.t code -> 'a}. If we strip the \texttt{code} off the result type,
+the functions \texttt{return (x : 'a) = fun \_ \_ -> x} and \texttt{bind (g : 'a t) (k : 'a -> 'b t) = fun size random -> k (g size random) size random}
+have the proper types for a monad instance.
+Then, any combinators of type \texttt{'a Gen.t} before simply become \texttt{'a code Gen.t} with this new version.
+\jwc{do we want a different name for the first cut?}
+
+However, this definition of bind doesn't have call-by value effect semantics, as
+discussed in the previous section!  And because the result type of \texttt{g} is
+just \texttt{'a} (not necessarily \texttt{'a code}), we cannot perform the
+let-insertion needed to preserve the CBV effects.  To solve this problem, we
+turn to a classic technique from the multistage programming literature: writing
+our staged program in continuation-passing style \cite{bondorf92, carrette05}.
+
+In Figure~\ref{fig:codecps-and-final-gen}, we follow \citet{kovacs24} and define the type \texttt{'a CodeCps.t = 'z.
+('a -> 'z code) -> 'z code}.
+This is the polymorphic CPS monad, with the continuation type always in \texttt{code}.
+\footnote{
+This is an instance of the \emph{codensity} monad \cite{janis08}.
+}
+
 \jwc{
   \begin{itemize}
-    \item Careful readers might note that the types of return and bind are not actually the right types.
-    \item This means that they aren't compatible with ppx-jane syntax.
-    \item We could try a different type, \texttt{type 'a gen = int code -> SR.t code -> 'a}.
-This has an actual monad instance (show), and ou can write teh combinators of before, with \texttt{'a code gen} everywhere you had \texttt{'a gen} before.
-    \item But because the result type isn't always in \texttt{code}, \texttt{bind} can't perform the let-insertion needed to preserve effect ordering.
-    \item Instead, we turn to a classic technique from the multistage programming literature: writing our staged program in continuation-passing style.
-    \item We define \texttt{type 'a codecps = 'z.
-('a -> 'z code) -> 'z code}.
-This is the polymorphic CPS monad, with the continuation type always in `code'.
-(Footnote: this is an instance of the \emph{codensity} monad, cf ``Asymptotic Improvement of Computations over Free Monads'')
-    (Cite: ``improving binding times without explicit cps conversion'', ``Multi-stage programming with functors and monads: Eliminating abstraction overhead from generic code'', and ``Closure-Free Functional Programming in a Two-Level Type Theory'')
     \item This lets us define the function \texttt{let\_insert : 'a code -> 'a codecps} as \texttt{let\_insert cx = fun k -> k .< let x = .~cx in .~(k .<x>.) >.}.
@@
 
+\begin{figure}
 \begin{lstlisting}
@@
 \end{lstlisting} 
+\caption{CodeCPS and The Final Gen Monad}
+\label{fig:codecps-and-final-gen}
+\end{figure}
 

### 2025-03-13 paper/paper.tex
skeleton for recursion

@@
 turn to a classic technique from the multistage programming literature: writing
-our staged program in continuation-passing style \cite{bondorf92, carrette05}.
-
-In Figure~\ref{fig:codecps-and-final-gen}, we follow \citet{kovacs24} and define the type \texttt{'a CodeCps.t = 'z.
-('a -> 'z code) -> 'z code}.
-This is the polymorphic CPS monad, with the continuation type always in \texttt{code}.
-\footnote{
-This is an instance of the \emph{codensity} monad \cite{janis08}.
-}
-
-\jwc{
-  \begin{itemize}
-    \item This lets us define the function \texttt{let\_insert : 'a code -> 'a codecps} as \texttt{let\_insert cx = fun k -> k .< let x = .~cx in .~(k .<x>.) >.}.
-    \item of course, \texttt{codecps} has the usual monad instance for the CPS monad with a polymorphic result type.
-    \item We then define \texttt{type 'a gen = int code -> SR.t code -> 'a codecps}, which lets us have it both ways.
-the effectful programs like \texttt{int} do their \emph{own} let-insertion, and we get a real monad instance.
-  \end{itemize}
-}
-
-\jwc{This gives us the final definitions:}
+our staged programs in continuation-passing style \cite{bondorf92}.
+
+In Figure~\ref{fig:codecps-and-final-gen}, we follow prior work \cite{kovacs24, carrette05} and define
+the type \texttt{'a CodeCps.t = 'z. ('a -> 'z code) -> 'z code}: a polymorphic continuation transformer with the result type
+always in \texttt{code}
+\footnote{This is an instance of the \emph{codensity} monad \cite{janis08}, a fact which deserves further investigation.}.
+The monad instance for this type is the standard instance for a CPS monad with polymorphic return type.
+In prior work, this type is often referred to as the ``code generation'' monad (and sometimes called \texttt{Gen}).
+This is because a value of type \texttt{('a code) CodeCps.t} is like an ``action'' that generates \texttt{code}:
+\texttt{CodeCps.run} passes the continuation transformer the identity continuation to produce a \texttt{'a code}.
+To avoid confusion with random data generators, we refer to this type as \texttt{CodeCps.t}.
+Most importantly, the \texttt{CodeCps} type supports a function \texttt{let\_insert}, which, given \texttt{cx : 'a code},
+let-binds \texttt{let x = .\~{}cx}, and then passes \texttt{.<x>.} to the continuation.
+
+We can then redefine our staged generator monad type to be
+\texttt{'a Gen.t = int code -> SR.t code -> 'a CodeCPS.t}, as shown in Figure~\ref{fig:codecps-and-final-gen}.
+\jwc{any types that were \texttt{'a Gen.t} before are now \texttt{'a code Gen.t}}.
+This gives us the best of both worlds. First, we get a monad instance for \texttt{'a Gen.t} with the correct types, which lets us
+use the monadic syntactic sugar of our chosen language. Moreover, we also get to maintain the correct
+effect ordering: effectful combinators like \texttt{Gen.int} do their \emph{own} let-insertion, ensuring
+that a program like \texttt{Gen.bind Gen.int (fun x -> ...)} generates a let-binding for the result of sampling the RNG.
+For example, \texttt{bind (int .<0>. .<1.>) (fun cx -> return .<(.~cx,.~cx)>.)} now correctly generates
+\texttt{.< fun size random -> let x = SR.int random 0 1 in (x,x) >.}. \jwc{Should we do out the whole reduction sequence here? It might be explanatory, but it also might be boring.}
+
+This design is less obviously correct, and does require some care. Rather than \texttt{bind} ensuring correct evaluation order once and for all,
+individual combinators must be carefully written to ensure that \texttt{'a code} values that contian effects are \texttt{let\_insert}'d.
+In Section~\ref{subsection:pbt-for-pbt}, we discuss how we use PBT to ensure that the OCaml is written correctly.
+
+\jwc{... and all of this works identically in Scala, too.}
 
@@
 
-  let run (t : 'a code t) : 'a code = t.cps (fun x -> x)
+  let run (t : ('a code) t) : 'a code = t.cps (fun x -> x)
 
@@
   let int (lo : int code) (hi : int code) : int code t =
-    fun size random ->
-      let_insert .< SR.int .~random .~lo .~hi >.
+    fun size random -> let_insert .< SR.int .~random .~lo .~hi >.
+
 end
@@
 
-\jwc {
-And this has the desired effect that
-\texttt{bind (int .<0>. .<1.>) (fun x -> return .<(.~x,.~x)>.)} generates
-\texttt{fun size random -> let x = SR.int random 0 1 in (x,x)}
-}
-
-\subsubsection{The Trick}
+\subsubsection{The Trick at Other Types}
+To write more interesting generators, we also need the ability to generate code that manipulates runtime values.
+For instance, consider this generator
+
+\jwc{need a better example here}
+
+\begin{figure}
+\begin{lstlisting}
+let int_or_zero : int Bq.t =
+  let%bind n = size in
+  if n <= 5 then return 0 else Bq.int
+
+let split_bool (b : bool code) : bool Gen.t =
+  fun _ _ ->
+
+let int_or_zero_staged : int 
+\end{lstlisting}
+\caption{??}
+\label{fig:trick-example}
+\end{figure}
+
 \jwc{Describe split --- this section is approximately experts-only, but you can just point at the Andras paper and the things it cites.}
@@
 \jwc{Describe recursion and recurisve handles}
+
+\begin{lstlisting}
+type ('a,'r) handle = 'r code -> 'a code t
+
+let recurse (f : ('a,'r) handle) (x : 'r code) : 'a code t =
+  fun size random ->
+    Codecps.bind ((f x) size random) @@ fun c ->
+    Codecps.let_insert c
+
+let recursive (x0 : 'r code) (step : ('a,'r) handle -> 'r code -> 'a code t) : 'a code t =
+  fun size_c random_c -> 
+    Codecps.return @@
+      .< let rec go x size random = .~(
+            Codecps.code_generate @@
+            (step (fun xc' -> fun ~size' ~random' -> Codecps.return .< go .~xc' .~size' .~random' >.) .<x>.) .<size>. .<random>.
+          )
+        in go .~x0 .~size_c .~random_c
+      >.
+\end{lstlisting}
 
@@
 \section{Evaluation}
-
-\subsubsection{Implementation}
+\subsection{Implementing and Testing Generators}
+\label{subsection:pbt-for-pbt}
+\jwc{Talk about the difftesting to ensure staging maintains semantics, and also different RNGs have identical semantics.}
+
 \subsection{Benchmarking speed \& resource usage}

### 2025-03-13 paper/paper.tex
CP 4.1

@@
 \label{subsection:basic-design}
+\hg{TODO: Signposting --- I got a few paragraphs in and realized I wasn't sure
+what I was reading}
+
 Recall that staging a DSL requires carefully changing some of the types and
-combinators to transform \texttt{code}s.  Deciding which types \texttt{t} can
-instead be \texttt{t code} is an art known as ``binding-time analysis'' \cn:
+combinators to transform \texttt{code}s.\hg{``to run at compile time rather than
+run-time''? The phrase ``transform codes'' tripped me up} Deciding which types \texttt{t} can
+instead be \texttt{t code} --- in other words,
 figuring out which parts of the DSL can be determined statically (and can be
 part of the compile-time stage), and which parts are only known dynamically (and
-hence must be \texttt{code}).
+hence must be \texttt{code}) --- is an art known as ``binding-time analysis'' \cn.
 
@@
 size parameter \jwc{if explained} are only known at run time (the later stage),
-but the code of the generator itself is known at compile time. Generators ---
-values of type \texttt{'a Bq.t} --- are never constructed dynamically at run
-time in practice, so all of the combinators we use to build them can run at compile time.
+but the code of the generator itself is known at compile time.\hg{Are we going
+to talk more about how we did this analysis?} Generators ---
+values of type \texttt{'a Bq.t} --- are always constructed statically in practice,
+so all of the combinators we use to build them can run at compile time.
 
 This means our library's generator type \texttt{'a
-Gen.t} should be \texttt{int code -> SR.t code -> 'a code}: a compile-time
+Gen.t} should have the type \texttt{int code -> SR.t code -> 'a code}: a compile-time
 function from dynamically-known size and seed to dynamically-determined result.
@@
 The monadic interface is given by a return and bind, as usual.
-\texttt{return} is the constant generator, but this time it operates on codes. Given \texttt{cx : 'a code}, the code for a \texttt{'a}, it returns the generator which always generates that value.
+\hg{Don't start a sentence with a lower-case letter, even one that's part of a
+symbol}\texttt{return} is the constant generator, but this time it operates on
+codes\hg{Is there an alternative to this phrase? Runs at compile time? Or
+operates on code? I just don't like ``codes'' used this way}. Given \texttt{cx : 'a code}, the code for a \texttt{'a}, it
+returns the generator which always generates that value.
 \texttt{bind g k} sequences generators by passing the result of running the generator \texttt{g} to a continuation \texttt{k}. However, instead of getting access
@@
 the arguments \texttt{lo} and \texttt{hi} are of type \texttt{int code}, and get spliced into the code block as arguments to \texttt{SR.int}.
-Lastly, \texttt{to\_bq} turns a staged generator into code for a normal \bq
+Lastly, \texttt{to\_bq} turns a staged generator into code for a normal \bq\
 generator. This function is just a 2-argument version of ``The Trick'' (\texttt{eta}
 from Section~\ref{subsection:msp}).
+\hg{TODO: Be really careful with formatting of the above paragraph. It's very
+easy for some inline code to get split weirdly over a line break or just be
+difficult to parse in general, and that could throw the reader off when the
+content is already pretty low-level}
 

### 2025-03-13 paper/paper.tex
CP 4.2

@@
 
-In Section~\ref{section:motiv}, we noticed that generator combinators like \texttt{weighted\_union}
+In Section~\ref{section:motiv}, we noted that generator combinators like \texttt{weighted\_union}
 must allocate lists in the hot path of the generator. Even though these lists are often small ---
@@
 This is an ideal opportunity to exercise another use of staging: compile-time specialization.
-Because we almost always know the particular list of choices at compile time, a staged version of \texttt{weighted\_union}
-can generate \emph{different code} depending on the number of generators that are being unioned together.
-If we use weighted union on a list (known at compile time) of generators \texttt{g1}, \texttt{g2}, and \texttt{g3},
-we can emit code that never has to create a list of three elements at run time.
+Since we almost always know the particular list of choices at compile time, a staged version of \texttt{weighted\_union}
+can generate \emph{different code} depending on the number of generators in the union.
+If we use weighted union on a compile-time list of generators \texttt{g1}, \texttt{g2}, and \texttt{g3},
+we can emit code that picks between the generators without realizing the list at
+run time.
 
@@
 produces a tree of \texttt{if}s by again traversing the list of generators at compile time.
-This tree of \texttt{ifs} ``searches'' for the genrator corresponding to the sampled value \texttt{r}, and then runs it.
+This tree of \texttt{ifs} ``searches'' for the generator corresponding to the
+sampled value \texttt{r}, and then runs it.
+\hg{I'm not sure this helps me. I think I'd prefer a less detailed explanation
+of the code that conveys the intuition and let the reader actually read the code
+if they want to know specifics. As it stands the explanation is just too dense
+for me to process}
 

### 2025-03-13 paper/paper.tex
CP 4.3

@@
 \texttt{let bind' g k = fun size random -> k (g size random) size random}, without the code block
-that let-binds the spliced code \texttt{.\~{}(g size random)}?
+that let-binds the spliced code \texttt{.\~{}(g size random)}?\hg{Need to spell
+this out --- the vast majority of readers won't have that precise question in
+their heads. Might even be worth comparing the two implementations here side by side}
 Unfortunately, using \texttt{Gen.bind'} leads to incorrect code being generated.
@@
 \jwc{This fact should go earlier. Not really sure about where this subsection should actually land, but it needs to be said somehwere.}
+\hg{+1, I think this feels sort of out of place here, but I also think moving it
+up would make that part harder to understand too. Is there a way to tie this in
+to our discussion about maintaining the precise set of choices? The two issues
+are different, but they're related}
 

### 2025-03-13 paper/paper.tex
CP 4.4

@@
 
-This is all great so far, but there's a subtle issue that prevents the version of the library design discussed so far
+This is all great so far,\hg{too informal} but there's a subtle issue that prevents the version of the library design discussed so far
 from being used as a proper drop-in replacement for an existing generator DSL: the types of \texttt{return : 'a code -> 'a Gen.t}
@@
 the syntactic sugar for monadic programming (\texttt{let\%bind} in OCaml, \texttt{foreach} in Scala, \texttt{do} in Haskell, etc)
-that makes it so appealing van \emph{only} be used if the types involved are actually the proper monad function types.
+that makes it so appealing van \emph{only} be used if the types involved are
+actually the proper monad function types.
+\hg{A pedantic reader may ask: Are we concerned with the monad {\em laws} as
+well, or just the type signature?}
 
@@
 our staged programs in continuation-passing style \cite{bondorf92}.
+\hg{Slow down! This is all very interesting and very technical! Try making each
+of these sentences two sentences, add some citations, and maybe spell this out
+wiht an example}
 
@@
 The monad instance for this type is the standard instance for a CPS monad with polymorphic return type.
-In prior work, this type is often referred to as the ``code generation'' monad (and sometimes called \texttt{Gen}).
+In prior work, this type is often referred to as the ``code generation'' monad
+(and sometimes\hg{, ironically,} called ``\texttt{Gen}'').
 This is because a value of type \texttt{('a code) CodeCps.t} is like an ``action'' that generates \texttt{code}:
@@
 Most importantly, the \texttt{CodeCps} type supports a function \texttt{let\_insert}, which, given \texttt{cx : 'a code},
-let-binds \texttt{let x = .\~{}cx}, and then passes \texttt{.<x>.} to the continuation.
+let-binds \texttt{let x = .\~{}cx}, and then passes \texttt{.<x>.} to the
+continuation.
+\hg{I'm getting lost in the inline code again}
 

### 2025-03-14 paper/paper.tex
explain recursive generators

@@
 
+\hg{TODO: Signposting --- I got a few paragraphs in and realized I wasn't sure
+what I was reading}
+
 \subsection{Basic Design}
 \label{subsection:basic-design}
-\hg{TODO: Signposting --- I got a few paragraphs in and realized I wasn't sure
-what I was reading}
-
-Recall that staging a DSL requires carefully changing some of the types and
-combinators to transform \texttt{code}s.\hg{``to run at compile time rather than
-run-time''? The phrase ``transform codes'' tripped me up} Deciding which types \texttt{t} can
+Recall that staging a DSL involves changing the
+combinators to run at compile time by carefully annotating their types with \texttt{code}s.
+Deciding which types \texttt{t} can
 instead be \texttt{t code} --- in other words,
@@
 but the code of the generator itself is known at compile time.\hg{Are we going
-to talk more about how we did this analysis?} Generators ---
+to talk more about how we did this analysis?} \jwc{Well it's sort of right here: you just think about it, it's the same thing as staging a parser too.}
+Generators ---
 values of type \texttt{'a Bq.t} --- are always constructed statically in practice,
@@
 The monadic interface is given by a return and bind, as usual.
-\hg{Don't start a sentence with a lower-case letter, even one that's part of a
-symbol}\texttt{return} is the constant generator, but this time it operates on
-codes\hg{Is there an alternative to this phrase? Runs at compile time? Or
-operates on code? I just don't like ``codes'' used this way}. Given \texttt{cx : 'a code}, the code for a \texttt{'a}, it
+\texttt{Return} is the constant generator, but this time it 
+runs at compile time. Given \texttt{cx : 'a code}, the code for a \texttt{'a}, it
 returns the generator which always generates that value.
@@
 \label{subsection:codecps}
-
 This is all great so far,\hg{too informal} but there's a subtle issue that prevents the version of the library design discussed so far
@@
 the syntactic sugar for monadic programming (\texttt{let\%bind} in OCaml, \texttt{foreach} in Scala, \texttt{do} in Haskell, etc)
-that makes it so appealing van \emph{only} be used if the types involved are
+that makes it so appealing can \emph{only} be used if the types involved are
 actually the proper monad function types.
@@
 However, this definition of bind doesn't have call-by value effect semantics, as
-discussed in the previous section!  And because the result type of \texttt{g} is
+discussed in the previous section!  And because the type of \texttt{g size random} is
 just \texttt{'a} (not necessarily \texttt{'a code}), we cannot perform the
@@
 wiht an example}
+\jwc{I could, but i'm just not sure this is critical.}
 
@@
 In prior work, this type is often referred to as the ``code generation'' monad
-(and sometimes\hg{, ironically,} called ``\texttt{Gen}'').
+(and sometimes, ironically, called ``\texttt{Gen}'').
 This is because a value of type \texttt{('a code) CodeCps.t} is like an ``action'' that generates \texttt{code}:
@@
 let split_bool (b : bool code) : bool Gen.t =
-  fun _ _ ->
-
-let int_or_zero_staged : int 
+  fun _ _ ->  _
+
 \end{lstlisting}
@@
 \subsection{Recursive Generators}
-\jwc{Describe recursion and recurisve handles}
-
-\begin{lstlisting}
-type ('a,'r) handle = 'r code -> 'a code t
-
-let recurse (f : ('a,'r) handle) (x : 'r code) : 'a code t =
-  fun size random ->
-    Codecps.bind ((f x) size random) @@ fun c ->
-    Codecps.let_insert c
-
-let recursive (x0 : 'r code) (step : ('a,'r) handle -> 'r code -> 'a code t) : 'a code t =
-  fun size_c random_c -> 
-    Codecps.return @@
-      .< let rec go x size random = .~(
-            Codecps.code_generate @@
-            (step (fun xc' -> fun ~size' ~random' -> Codecps.return .< go .~xc' .~size' .~random' >.) .<x>.) .<size>. .<random>.
-          )
-        in go .~x0 .~size_c .~random_c
-      >.
-\end{lstlisting}
+\jwc{Generating recursive datatypes requires recursive generators!}
+
+Different generator DSLs handle defining recursive generators differently. Some allow recursive generators
+to be defined as recursive functions (or in Haskell's case, recursive values), while others
+(like \bq) expose a fixpoint combinator to construct recursive generators.
+Generator fixpoint combinators are simliar to a usual \texttt{fix : ('a -> 'a) -> 'a} combinator,
+except that they operate at monadic type \texttt{fixed\_point : ('a Bq.t -> 'a Bq.t) -> 'a Bq.t}.
+Given a step function that takes a ``handle'' to sample from a recursive generator call, it
+ties the knot and builds a recursive generator.
+
+In our case, letting programmers define recursive generators as recursive functions is out of the question.
+With staged programming, recursion must be handled with care: it is far to easy to accidentally recursively
+define an infinite \texttt{code} value and have the program diverge at compile time, when trying to write
+a \texttt{code} representing a recursive program.
+To this end, we develop a staged recursive generator combinator\footnote{
+In reality, we actually have a more general API that allows programmers to define \emph{parameterized}
+recursive combinators, of type \texttt{'r code -> 'a code Gen.t}, for any type \texttt{'r}. See Appendix~\jwc{appendix} for details.
+}, whose API is shown in Figure~\ref{fig:staged-recursive-generator}.
+The code for this combinator can be found in Appendix~\jwc{appendix}.
+The recursion API consists of an opaque type \texttt{'a handle}, and a function \texttt{recurse} to perform recursive calls.
+Programmers can then define recursive generators by \texttt{fixed\_point}, which ties the recursive knot.
+
+% \begin{lstlisting}
+%   type ('a,'r) handle = 'r code -> 'a code t
+
+%   let recurse (f : ('a,'r) handle) (x : 'r code) : 'a code t =
+%     fun size random ->
+%       Codecps.bind ((f x) size random) @@ fun c ->
+%       Codecps.let_insert c
+
+%   let recursive (step : ('a,'r) handle -> 'r code -> 'a code t) (x0 : 'r code) : 'a code t =
+%     fun size_c random_c -> 
+%       Codecps.return @@
+%         .< let rec go x size random = .~(
+%               Codecps.code_generate @@
+%               (step (fun xc' -> fun ~size' ~random' -> Codecps.return .< go .~xc' .~size' .~random' >.) .<x>.) .<size>. .<random>.
+%             )
+%           in go .~x0 .~size_c .~random_c
+%         >.
+% end
+% \end{lstlisting}
+
+\begin{figure}
+\begin{lstlisting}
+type 'a handle
+val recurse : 'a handle -> 'a code Gen.t
+val fixed_point : ('a handle -> 'a code Gen.t) -> 'a code Gen.t
+\end{lstlisting}
+\caption{Staged Recursive Generator Combinator API}
+\label{fig:staged-recursive-generator}
+\end{figure}
 

### 2025-03-14 paper/paper.tex
some more work

@@
 \subsection{Property-Based Testing}
-\jwc{Usual spiel about pbt and generators: here's the BQ generator library, it has return and bind, this is what they're for.}
+\jwc{Usual spiel about pbt and generators: here's a monadic generic generator library (but we're using BQ to illustrage), it has return and bind, this is what they're for.}
+\jwc{You often want generators to generate for sparse precoditions so you can exercise a property reasonably well}
 
@@
 
+\jwc{
+  \begin{itemize}
+    \item In PBT in practice, we learned that in many cases, programmers don't
+    even write custom generators, instead relying on type-derived generators.
+    \item These generators just produce arbitrary values of a given type, not necessarily enforcing validity conditions.
+    \item Many generator
+  \end{itemize}
+}
 \jwc{Todo: Thia}
@@
 \subsection{Impact on bug-finding ability}
+\jwc{Staging type-derived generators (which are the most commonly used ones) can turn a timeout into a found bug!}
 \section{Conclusion \& Future Work}

### 2025-03-14 paper/paper.tex
skeleton paper

@@
     \item These generators just produce arbitrary values of a given type, not necessarily enforcing validity conditions.
-    \item Many generator
+    \item If 
+    \item Let's talk about how type-deriving works. In languages with typeclasses, it works
+    by typeclass resolution. In OCaml it works by PPX system, but the principle
+    is the same. You define rules to go from generators of a subtypes to a generator of the larger type, and then apply those rules to build up a full generator.
+    \item \begin{itemize}
+      \item For base types, you just call the associated generator
+      \item For product types, you sample from all the component gnerators, bind the values, and then tuple up the values, and return the tuple
+      \item For variant types, you use a weighted union to choose one of the component generators with a weighted union.
+      \item For recursive types, wrap the whole thing in a fixedpoint and then use the recusive handle as the ``component generator'' for all recursive instance of the type.
     \end{itemize}
+    \item This kind of generic deriving of generators works just as well for
+    staged generators. Just replace the standard combinators in question with
+    staged ones!
+    \item We built this in OCaml, but you can easily use typeclasses to do it in scala too.
+    \item Note that this is (essentially) 3-stage metaprograming.
+You're using either the PPX mechanism or typeclass resolution to generate a bunch of staged combinator calls, which then run at compile time, which then run at run time.
+  \end{itemize}
 }
 \jwc{Todo: Thia}
-
-\jwc{This is 3-stage metaprogramming! There's PPX time, compile time, and run time.}
 

### 2025-03-14 paper/paper.tex
rewrite rng motiv section

@@
 \subsubsection{Choice of RNG}
-
-\jwc{We really need to have explained the effect ordering equivalence before here.}
-
-\jwc{
-  \begin{itemize}
-    \item Because the RNG sits at the heart of the generation hot path, RNG
-    speed matters enormously. Which RNG you pick matters...
-    \item Different PBT libraries use different RNGs --- QuickCheck and \bq use
-    splittable random, Scalacheck uses \href{https://burtleburtle.net/bob/rand/smallprng.html}{this one}.  These
-    are reasonably performant RNGs. Because this is the
-    core of the hot path of the generator, speeding up sampling just a little can
-    can dramatically speed up generation time.
-    \item Significantly faster RNGs that still pass pRNG benchmarks like BigCrush exist (like \href{https://github.com/lemire/testingRNG/blob/master/source/lehmer64.h}{this one}), and using them can 
-    speed up generation enormously --- in our microbenchmarks, the lehmer generator runs 4x faster than splittable random.
-    \item However, showing this is tricky, because different RNGs have different sampling orders, but our
-    eval is made much stronger by the fact that all of our generators are functionally equivalent.
-    \item In this paper demonstrate this fact by noticing a ``natural experiment'': OCaml's splittable random library is
-    slow in a way that can be improved \emph{without} changing the extensional behavior. By resolving this issue, we can demonstrate
-    just how much you can improve bug-finding performance simply by speeding up the RNG.
-  \end{itemize}
-}
-
-The core of any PBT generator library is a random-number generator. Following
-the original Haskell QuickCheck, \bq uses the SplitMix algorithm \cn,
-implemented as an OCaml library called \texttt{Splittable\_random}. The precise
-details of how SplitMix works are unimportant for the present paper, but the
-main operations are defined in terms of bitwise operations on 64-bit integers.
-However, due to implementation details related to the garbage collector, all
-runtime OCaml values are either (a) pointers to a block of memory, or (b) 63-bit ``immediate'' values.
-As such, the OCaml type of 64-bit ingeters, \texttt{int64}, is \emph{boxed}: it
-is a pointer to a block of memory that contains a 64-bit integer.  This means
-that \emph{all} \texttt{int64} operations (both arithmetic and bitwise) must
-allocate memory cells to contain their output.
-
-This has a significant impact on the performance of generators. A single call to
-one of the \bq library functions may sample from \texttt{Splittable\_random}
-multiple times, and each sample from \texttt{Splittable\_random} allocates 9
-words \jwc{this is a call to \texttt{next\_int64}}.  While seemingly small, we
-will see in Section~\ref{section:eval} that this amount of allocation has a
-significant impact on significant.
-
-In Section~\ref{subsection:faster-rng}, we show how our library mitigates this
-performance hit by using OCaml's FFI to reimplement the RNG in non-allocating C code.
-We also discuss how the ``unboxed types'' extension to the OCaml compiler \cn could instead be used
-to the same effect without leaving OCaml.
-
-It is worth noting that the inefficiency of boxed \texttt{int64}s is not due to
-the abstraction overhead of the generator DSL itself, but simply due to language
-implementation and algorithm choices. We include this discussion --- and
-the corresponding evaluation
---- to demonstrate the testing-time savings of taking seriously the notion that PBT libraries
-should be engineered for performance.
-
-\jwc{If you're not using the fasteest RNG possible, you're leaving RNG on the table.}
+\jwc{We really need to have explained the effect ordering equivalence aspect of the evaluation before here.}
+
+The core of any PBT generator library is a (pseudo) random number generator.
+Different PBT libraries use a wide range of different RNGs.
+Following the original Haskell QuickCheck, \bq uses the SplitMix algorithm \cn,
+implemented (statefully) as an OCaml library called \texttt{Splittable\_random}. Meanwhile,
+ScalaCheck uses the JSF algorithm.
+% https://www.pcg-random.org/posts/bob-jenkins-small-prng-passes-practrand.html
+% https://burtleburtle.net/bob/rand/smallprng.html
+
+% \jwc{There are a few aspects of PBT that make it interesting when it comes to PBT choice.
+% \begin{itemize}
+%   \item Doesn't need to be cryptographic.
+%   \item doesn't even \emph{necessarily} need to pass sophisticated staistical
+%   tests.  \cn
+% \end{itemize}
+% }
+
+Because the RNG sits at the heart of the generation hot path, RNG speed matters enormously.
+As we will show in Section~\ref{eval}, speeding up sampling can dramatically speed up generation time and bugfinding speed.
+
+We argue that the RNGs that PBT libraries use are (broadly) a source of inefficiency, and their (lack of) speed is
+an impediment towards better bugfinding ability.
+
+Significantly faster RNGs (that still pass pRNG benchmarks like BigCrush) exist like \href{https://github.com/lemire/testingRNG/blob/master/source/lehmer64.h}{this one}, and using them can 
+speed up generation tremendously --- in microbenchmarks, a C implementation of the lehmer generator runs 4x faster than a C implementation of splitmix.
+
+Similarly, PBT libraries could even eschew the requirement that a source of entropy pass statistical tests like bigcrush \cn.
+This is common practice in testing already: fuzzers often simply use a buffer full of arbitrary bytes as a source of entropy!
+Bumping a pointer and reading from a buffer (which can be pipelined trivially) is likely to be significantly faster than any RNG with data-dependent arithmetic instructions.
+
+In the context of this paper, however, demonstrating the benefits of any of the above options is tricky.
+Our bug-finding evaluation's power is predicated on the fact that all of the different versions of a generator we test are extensionally identical:
+for a given seed and size parameter, they all produce the \emph{same} value.
+Using a different random number generator or different source of entropy would break this property.
+
+In Section~\ref{eval}, we demonstrate that faster RNG matters for bugfinding by exploting a ``natural experiment'':
+OCaml's \texttt{Splittable\_random} library is slow in a way that can be improved \emph{without} changing its extensional behavior,
+namely that it uses \emph{boxed} 64-bit integers. By building a version that uses much faster unoxed 64-bit integer arithmetic, we can
+demonstrate just how much bug-finding performance can be improved just by using a more performant RNG.
+
+% \jwc{
+%   \begin{itemize}
+%     \item In this paper demonstrate this fact by noticing a ``natural experiment'': OCaml's splittable random library is
+%     slow in a way that can be improved \emph{without} changing the extensional behavior. By resolving this issue, we can demonstrate
+%     just how much you can improve bug-finding performance simply by speeding up the RNG.
+%   \end{itemize}
+% }
+
+%  The precise
+% details of how SplitMix works are unimportant for the present paper, but the
+% main operations are defined in terms of bitwise operations on 64-bit integers.
+% However, due to implementation details related to the garbage collector, all
+% runtime OCaml values are either (a) pointers to a block of memory, or (b) 63-bit ``immediate'' values.
+% As such, the OCaml type of 64-bit ingeters, \texttt{int64}, is \emph{boxed}: it
+% is a pointer to a block of memory that contains a 64-bit integer.  This means
+% that \emph{all} \texttt{int64} operations (both arithmetic and bitwise) must
+% allocate memory cells to contain their output.
+
+% This has a significant impact on the performance of generators. A single call to
+% one of the \bq library functions may sample from \texttt{Splittable\_random}
+% multiple times, and each sample from \texttt{Splittable\_random} allocates 9
+% words \jwc{this is a call to \texttt{next\_int64}}.  While seemingly small, we
+% will see in Section~\ref{section:eval} that this amount of allocation has a
+% significant impact on significant.
+
+% In Section~\ref{subsection:faster-rng}, we show how our library mitigates this
+% performance hit by using OCaml's FFI to reimplement the RNG in non-allocating C code.
+% We also discuss how the ``unboxed types'' extension to the OCaml compiler \cn could instead be used
+% to the same effect without leaving OCaml.
+
+% It is worth noting that the inefficiency of boxed \texttt{int64}s is not due to
+% the abstraction overhead of the generator DSL itself, but simply due to language
+% implementation and algorithm choices. We include this discussion --- and
+% the corresponding evaluation
+% --- to demonstrate the testing-time savings of taking seriously the notion that PBT libraries
+% should be engineered for performance.
+
+% \jwc{If you're not using the fasteest RNG possible, you're leaving RNG on the table.}
 

### 2025-03-14 paper/paper.tex
more refactoring of rng section

@@
 ScalaCheck uses the JSF algorithm.
+
+Because the RNG sits at the heart of the generation hot path, RNG speed matters enormously.
+We argue that the RNGs that PBT libraries use are (broadly) a source of inefficiency, and their (lack of) speed is
+an impediment towards better bugfinding ability \cn \jwc{see comment for cites}.
 % https://www.pcg-random.org/posts/bob-jenkins-small-prng-passes-practrand.html
 % https://burtleburtle.net/bob/rand/smallprng.html
+
+Significantly faster RNGs like the Lehmer RNG \cn exist like, and using them can
+speed up generation tremendously --- in microbenchmarks, a C implementation of
+the lehmer generator runs 4x faster than a C implementation of splitmix.
+Similarly, PBT libraries could even eschew the requirement that a source of
+entropy pass statistical tests\cn.  This is common practice in testing already:
+fuzzers often simply use a buffer full of arbitrary bytes as a source of entropy
+\cn.  Bumping a pointer and reading from a buffer (which can be pipelined
+trivially) will always be significantly faster than any RNG with data-dependent
+arithmetic instructions.
+
+In the context of this paper, however, demonstrating the benefits of any of the
+above options is tricky.  Our bug-finding evaluation's power is predicated on
+the fact that all of the different versions of a generator we test are
+extensionally identical: for a given seed and size parameter, they all produce
+the \emph{same} value.  Using a different random number generator or different
+source of entropy would break this property.  In Section~\ref{eval}, we
+demonstrate that faster RNG matters for bugfinding by exploting a ``natural
+experiment'': OCaml's \texttt{Splittable\_random} library is slow in a way that
+can be improved \emph{without} changing its extensional behavior.
+In particular,
+due to implementation details related to the OCaml garbage collector,
+values of the OCaml type \texttt{int64} are not machine words, but rather
+\emph{pointers} to machine words. This means
+that \emph{all} \texttt{int64} operations (both arithmetic and bitwise) must
+allocate memory cells to contain their output, which has a significant performance benefit.
+By building a version that uses much faster ``unboxed'' 64-bit integer arithmetic, we can demonstrate just how much bug-finding
+performance can be improved just by using a more performant RNG.
 
@@
 
-Because the RNG sits at the heart of the generation hot path, RNG speed matters enormously.
-As we will show in Section~\ref{eval}, speeding up sampling can dramatically speed up generation time and bugfinding speed.
-
-We argue that the RNGs that PBT libraries use are (broadly) a source of inefficiency, and their (lack of) speed is
-an impediment towards better bugfinding ability.
-
-Significantly faster RNGs (that still pass pRNG benchmarks like BigCrush) exist like \href{https://github.com/lemire/testingRNG/blob/master/source/lehmer64.h}{this one}, and using them can 
-speed up generation tremendously --- in microbenchmarks, a C implementation of the lehmer generator runs 4x faster than a C implementation of splitmix.
-
-Similarly, PBT libraries could even eschew the requirement that a source of entropy pass statistical tests like bigcrush \cn.
-This is common practice in testing already: fuzzers often simply use a buffer full of arbitrary bytes as a source of entropy!
-Bumping a pointer and reading from a buffer (which can be pipelined trivially) is likely to be significantly faster than any RNG with data-dependent arithmetic instructions.
-
-In the context of this paper, however, demonstrating the benefits of any of the above options is tricky.
-Our bug-finding evaluation's power is predicated on the fact that all of the different versions of a generator we test are extensionally identical:
-for a given seed and size parameter, they all produce the \emph{same} value.
-Using a different random number generator or different source of entropy would break this property.
-
-In Section~\ref{eval}, we demonstrate that faster RNG matters for bugfinding by exploting a ``natural experiment'':
-OCaml's \texttt{Splittable\_random} library is slow in a way that can be improved \emph{without} changing its extensional behavior,
-namely that it uses \emph{boxed} 64-bit integers. By building a version that uses much faster unoxed 64-bit integer arithmetic, we can
-demonstrate just how much bug-finding performance can be improved just by using a more performant RNG.
-
 % \jwc{
@@
 % main operations are defined in terms of bitwise operations on 64-bit integers.
-% However, due to implementation details related to the garbage collector, all
-% runtime OCaml values are either (a) pointers to a block of memory, or (b) 63-bit ``immediate'' values.
-% As such, the OCaml type of 64-bit ingeters, \texttt{int64}, is \emph{boxed}: it
-% is a pointer to a block of memory that contains a 64-bit integer.  This means
-% that \emph{all} \texttt{int64} operations (both arithmetic and bitwise) must
-% allocate memory cells to contain their output.
+% However,
 

### 2025-03-14 paper/paper.tex
cut at rng section

@@
 implemented (statefully) as an OCaml library called \texttt{Splittable\_random}. Meanwhile,
-ScalaCheck uses the JSF algorithm.
-
-Because the RNG sits at the heart of the generation hot path, RNG speed matters enormously.
-We argue that the RNGs that PBT libraries use are (broadly) a source of inefficiency, and their (lack of) speed is
-an impediment towards better bugfinding ability \cn \jwc{see comment for cites}.
+ScalaCheck uses the JSF algorithm \cn.
 % https://www.pcg-random.org/posts/bob-jenkins-small-prng-passes-practrand.html
@@
 
-Significantly faster RNGs like the Lehmer RNG \cn exist like, and using them can
-speed up generation tremendously --- in microbenchmarks, a C implementation of
-the lehmer generator runs 4x faster than a C implementation of splitmix.
-Similarly, PBT libraries could even eschew the requirement that a source of
-entropy pass statistical tests\cn.  This is common practice in testing already:
+The RNG sits at the heart of the generation hot path.
+Even basic generators like generating a single \texttt{int} or \texttt{float} may sample from the RNG multiple times,
+and generator combinators like \texttt{list} generate many \texttt{int}s.
+Because of this, the speed of a single RNG call matters a great deal. Unfortunately,
+we argue that existing PBT libraries make relatively inefficient choices on this front,
+leading to worse bugfinding power than what is possible.
+
+For example, significantly faster RNG algorithms than SplitMix or JSF exist,
+such as the Lehmer RNG \cn exist. In microbenchmarks, the Lehmer generator almost 2x as fast as SplitMix \cn.
+% https://github.com/lemire/testingRNG/tree/master
+Moreover, PBT libraries could even consider eschewing the requirement that a source of
+entropy pass statistical tests like BigCrush\cn.  This is common practice in other areas of testing already:
 fuzzers often simply use a buffer full of arbitrary bytes as a source of entropy
-\cn.  Bumping a pointer and reading from a buffer (which can be pipelined
-trivially) will always be significantly faster than any RNG with data-dependent
-arithmetic instructions.
+\cn. Such approaches are also faster than algorithms like SplitMix: bumping a
+pointer and reading from memory (which can be pipelined trivially) will always
+be faster on modern CPUs than any RNG with data-dependent arithmetic
+instructions.
 
 In the context of this paper, however, demonstrating the benefits of any of the
-above options is tricky.  Our bug-finding evaluation's power is predicated on
+above faster options is tricky. Our bug-finding evaluation's power is predicated on
 the fact that all of the different versions of a generator we test are

### 2025-03-15 paper/paper.tex
minor updates

@@
 
-\subsubsection{Monadic DSL Abstraction Overhead}
+\subsection{Monadic DSL Abstraction Overhead}
 \jwc{
@@
 
-\subsubsection{Combinator Abstraction Overhead}
+\subsection{Combinator Abstraction Overhead}
 
@@
 % \subsubsection{Function call overhead}
-\subsubsection{Choice of RNG}
+\subsection{Choice of RNG}
 \jwc{We really need to have explained the effect ordering equivalence aspect of the evaluation before here.}
 
-The core of any PBT generator library is a (pseudo) random number generator.
+The core of any PBT generator library is a random number generator: to generate random values, we need
 Different PBT libraries use a wide range of different RNGs.
@@
 
-\section{Eliminating Abstraction Overhead of Generator DSLs by Staging}
+\section{Eliminating Abstraction Overhead of Generator DSLs by Staging and Faster RNG}
 \jwc{Emphasize that this is an \emph{COMPLETELY EQUIVALENT drop in replacement!} We din't just build a different library with different distributions.}
@@
 
-\subsection{Fast Random Number Generators}
+\subsection{Faster SplitMix with Unboxed Int}
 \label{subsection:faster-rng}
+
+\jwc{Unclear what we should call this section}
 
@@
   \begin{itemize}
-    \item Using the fastest RNG possible is important.
-    \item In OCaml, the splittalbe random RNG is not as fast as it could be, because of the Ints issue.
-    \item In this section, we describe how to resolve the ints issue by calling out ot the FFI. 
-    \item All of this can be done with Jane Street bleeding-edge compiler extensions, though these are not yet compatible with MetaOCaml.
-    \item Because different libraries 
+    \item As we discussed in Section~\ref{section:motiv}, inefficient RNG choice is a bottleneck for bugfinding power.
+    \item Faster RNGs and sources of entropy exist, though they are extensionally inequivalent.
+To demonstrate the bugfinding power of a faster RNG, we use the natural experiment provided by OCaml's inefficient implementation of SplitMix.
+    \item Unlike the previous subsections, this is not a anguage-agnostic technique. In most other PBT frameworks, the RNG operates on by machine integers, so this inefficiency does not exist.
+    \item We are simply using this as a way to demonstrate that faster RNG can translate to faster bugfinding for sampling-bounded generators.
+    \item Here's how SplitMix is slow: the core of SplitMix is nextInt64, which does allocations for each operation.
+    \item If we could use actual int64s, we wouldn't have to allocate.
+    \item So we do this:
+    \begin{itemize}
+      \item We implement SplitMix in C (which uses normal Int64s) and real int64 arithmetic, and use the OCaml FFI to call out to the C functions. T
+      \item This could be done with Jane Street bleeding-edge compiler extensions, though these are not yet implemented in a a version of the OCaml compiler that also has staging.
+    \end{itemize}
   \end{itemize}
@@
     \item Pair of ints, constrained
-    \item List of ints
+    \item List of ints without bind (use map): lots of sampling, minimal binds.
+    \item Unableled Trees of a fixed size (no weighted union) minimal sampling, lots of binds.
   \end{itemize}

### 2025-03-15 paper/paper.tex
continue to write splitmix section

--- 2025-03-14 paper/paper.tex
@@
 
-%  The precise
-% details of how SplitMix works are unimportant for the present paper, but the
-% main operations are defined in terms of bitwise operations on 64-bit integers.
-% However,
-
-% This has a significant impact on the performance of generators. A single call to
-% one of the \bq library functions may sample from \texttt{Splittable\_random}
-% multiple times, and each sample from \texttt{Splittable\_random} allocates 9
-% words \jwc{this is a call to \texttt{next\_int64}}.  While seemingly small, we
-% will see in Section~\ref{section:eval} that this amount of allocation has a
-% significant impact on significant.
-
 % In Section~\ref{subsection:faster-rng}, we show how our library mitigates this
@@
 
-\subsection{Fast Random Number Generators}
+\subsection{Faster SplitMix with Unboxed Int}
 \label{subsection:faster-rng}
+
+\jwc{Unclear what we should call this section}
+
+As we discussed in Section~\ref{section:motiv}, inefficient RNG choice is
+another bottleneck for finding bugs fast. While generator libraries by
+and large use sensible RNG algorithms, they are not explicitly chosen
+with performance in mind. Indeed, much faster RNGs \cn and fast sources
+of entropy \cn exist. \jwc{... more here?}
+To demonstrate that a faster RNG can significanty impact bugfinding power, we
+use the natural experiment provided by OCaml's inefficient implementation of
+SplitMix. By replacing this slow RNG with a faster but extensionally equivalent
+implemetnation, we are able to precisely quantify the bugfinding speedup that using a RNG gives across a range of PBT scenarios.
+
+We emphasize that while we believe that the insight that faster RNG translates to faster bug finding
+is a cross-langauge one, the specific technical contents of this section are OCaml-specific.
+In most other PBT frameworks \cn, the RNG operates on by machine
+integers, so this \emph{particular} inefficiency does not exist.
+
+The precise details of how the SplitMix algorithm works are unimportant for the
+present paper, but the critical component is that all of its operations are are
+defined in terms of arithmetic bitwise operations on 64-bit integers.  In OCaml,
+a
+% however,
+
+% this has a significant impact on the performance of generators. a single call to
+% one of the \bq library functions may sample from \texttt{splittable\_random}
+% multiple times, and each sample from \texttt{splittable\_random} allocates 9
+% words \jwc{this is a call to \texttt{next\_int64}}.  while seemingly small, we
+% will see in section~\ref{section:eval} that this amount of allocation has a
+% significant impact on significant.
 
@@
   \begin{itemize}
-    \item Using the fastest RNG possible is important.
-    \item In OCaml, the splittalbe random RNG is not as fast as it could be, because of the Ints issue.
-    \item In this section, we describe how to resolve the ints issue by calling out ot the FFI. 
-    \item All of this can be done with Jane Street bleeding-edge compiler extensions, though these are not yet compatible with MetaOCaml.
-    \item Because different libraries 
+    \item as we discussed in section~\ref{section:motiv}, 
+    \item faster rngs and sources of entropy exist, though they are extensionally inequivalent. 
+    \item 
+    \item 
+    \item here's how splitmix is slow: the core of splitmix is nextint64, which does allocations for each operation.
+    \item if we could use actual int64s, we wouldn't have to allocate.
+    \item so we do this:
+    \begin{itemize}
+      \item we implement splitmix in c (which uses normal int64s) and real int64 arithmetic, and use the ocaml ffi to call out to the c functions. t
+      \item this could be done with jane street bleeding-edge compiler extensions, though these are not yet implemented in a a version of the ocaml compiler that also has staging.
     \end{itemize}
-}
-
-\section{Evaluation}
-\subsection{Implementing and Testing Generators}
+  \end{itemize}
+}
+
+\section{evaluation}
+\subsection{implementing and testing generators}
 \label{subsection:pbt-for-pbt}
-\jwc{Talk about the difftesting to ensure staging maintains semantics, and also different RNGs have identical semantics.}
-
-\subsection{Benchmarking speed \& resource usage}
-
-\jwc{NOTE: we should test generator speed across both languages, but speed -> bugfinding ability in only OCaml.}
+\jwc{talk about the difftesting to ensure staging maintains semantics, and also different rngs have identical semantics.}
+
+\subsection{benchmarking speed \& resource usage}
+
+\jwc{note: we should test generator speed across both languages, but speed -> bugfinding ability in only ocaml.}
 \jwc{
-  Baseline generators to test speed in both languages:
+  baseline generators to test speed in both languages:
   \begin{itemize}
-    \item Single int
-    \item Pair of ints, constrained
-    \item List of ints
+    \item single int
+    \item pair of ints, constrained
+    \item list of ints without bind (use map): lots of sampling, minimal binds.
+    \item unableled trees of a fixed size (no weighted union) minimal sampling, lots of binds.
   \end{itemize}
 }
-\subsection{Impact on bug-finding ability}
-\jwc{Staging type-derived generators (which are the most commonly used ones) can turn a timeout into a found bug!}
-\section{Conclusion \& Future Work}
-\section{Related Work}
-\jwc{A generator is a parser of randomness, and people have implemented lots of staged parser libraries!}
-%%
-%% The next two lines define the bibliography style to be used, and
+\subsection{impact on bug-finding ability}
+\jwc{staging type-derived generators (which are the most commonly used ones) can turn a timeout into a found bug!}
+\section{conclusion \& future work}
+\section{related work}
+\jwc{a generator is a parser of randomness, and people have implemented lots of staged parser libraries!}
+%%
+%% the next two lines define the bibliography style to be used, and
 %% the bibliography file.
-\bibliographystyle{ACM-Reference-Format}
+\bibliographystyle{acm-reference-format}
 \bibliography{bib}
@@
 %%
-%% If your work has an appendix, this is the place to put it.
+%% if your work has an appendix, this is the place to put it.
 \end{document}
@@
 %%
-%% End of file `sample-sigplan.tex'.
+%% end of file `sample-sigplan.tex'.

### 2025-03-15 paper/paper.tex
blahg

@@
 \jwc{Unclear what we should call this section}
-
 As we discussed in Section~\ref{section:motiv}, inefficient RNG choice is
@@
 
-The precise details of how the SplitMix algorithm works are unimportant for the
+The precise details of how the SplitMix algorithm works \cn are unimportant for the
 present paper, but the critical component is that all of its operations are are
 defined in terms of arithmetic bitwise operations on 64-bit integers. In OCaml,
-a
-% however,
-
-% this has a significant impact on the performance of generators. a single call to
-% one of the \bq library functions may sample from \texttt{splittable\_random}
-% multiple times, and each sample from \texttt{splittable\_random} allocates 9
-% words \jwc{this is a call to \texttt{next\_int64}}.  while seemingly small, we
-% will see in section~\ref{section:eval} that this amount of allocation has a
-% significant impact on significant.
-
-\jwc{
-  \begin{itemize}
-    \item as we discussed in section~\ref{section:motiv}, 
-    \item faster rngs and sources of entropy exist, though they are extensionally inequivalent. 
-    \item 
-    \item 
-    \item here's how splitmix is slow: the core of splitmix is nextint64, which does allocations for each operation.
-    \item if we could use actual int64s, we wouldn't have to allocate.
-    \item so we do this:
-    \begin{itemize}
-      \item we implement splitmix in c (which uses normal int64s) and real int64 arithmetic, and use the ocaml ffi to call out to the c functions. t
-      \item this could be done with jane street bleeding-edge compiler extensions, though these are not yet implemented in a a version of the ocaml compiler that also has staging.
-    \end{itemize}
-  \end{itemize}
-}
+because of details related to the garbage collector, the 64-bit integer type \texttt{int64} is represented
+at run time as a \emph{pointer} to an unscanned block of memory containing (among other things)
+a 64-bit integer \cn.
+
+This means that all operations that return an \texttt{int64} must allocate this block of memory.
+This has a significant impact on the performance of generators. A single call to
+one of the \bq library functions --- like generating an integer uniformly in a range --- may call the \texttt{splittable\_random}
+algorithm multiple times. Each sample from \texttt{splittable\_random} allocates 9
+words \jwc{this is a call to \texttt{next\_int64}}, and each allocation brings us closer
+to the next garbage collection pause. While allocation in OCaml is fast and the garbage colletor
+runs incrementally in small slices
+\footnote{The OCaml GC is a generational collector \cn, and since these allocations are mostly very short lived, they will never be promoted.
+}, we will see in Section~\ref{section:eval} that this can have
+a large performance impact on some generators that spend most of their time sampling data.
+
+% \jwc{
+%   \begin{itemize}
+%     \item as we discussed in section~\ref{section:motiv}, 
+%     \item faster rngs and sources of entropy exist, though they are extensionally inequivalent. 
+%     \item here's how splitmix is slow: the core of splitmix is nextint64, which does allocations for each operation.
+%     \item if we could use actual int64s, we wouldn't have to allocate.
+%     \item so we do this:
+%     \begin{itemize}
+%       \item we implement splitmix in c (which uses normal int64s) and real int64 arithmetic, and use the ocaml ffi to call out to the c functions. t
+%       \item this could be done with jane street bleeding-edge compiler extensions, though these are not yet implemented in a a version of the ocaml compiler that also has staging.
+%     \end{itemize}
+%   \end{itemize}
+% }
 

### 2025-03-15 paper/paper.tex
oops fix caps

@@
 
+\label{subsection:faster-rng}
+
 \jwc{Unclear what we should call this section}
@@
 
-\section{evaluation}
-\subsection{implementing and testing generators}
+\section{Evaluation}
+\subsection{Implementing and Testing Generators}
 \label{subsection:pbt-for-pbt}
-\jwc{talk about the difftesting to ensure staging maintains semantics, and also different rngs have identical semantics.}
-
-\subsection{benchmarking speed \& resource usage}
-
-\jwc{note: we should test generator speed across both languages, but speed -> bugfinding ability in only ocaml.}
+\jwc{Talk about the difftesting to ensure staging maintains semantics, and also different RNGs have identical semantics.}
+
+\subsection{Benchmarking speed \& resource usage}
+
+\jwc{NOTE: we should test generator speed across both languages, but speed -> bugfinding ability in only OCaml.}
 \jwc{
-  baseline generators to test speed in both languages:
+  Baseline generators to test speed in both languages:
   \begin{itemize}
-    \item single int
-    \item pair of ints, constrained
-    \item list of ints without bind (use map): lots of sampling, minimal binds.
-    \item unableled trees of a fixed size (no weighted union) minimal sampling, lots of binds.
+    \item Single int
+    \item Pair of ints, constrained
+    \item List of ints without bind (use map): lots of sampling, minimal binds.
+    \item Unableled Trees of a fixed size (no weighted union) minimal sampling, lots of binds.
   \end{itemize}
 }
-\subsection{impact on bug-finding ability}
-\jwc{staging type-derived generators (which are the most commonly used ones) can turn a timeout into a found bug!}
-\section{conclusion \& future work}
-\section{related work}
-\jwc{a generator is a parser of randomness, and people have implemented lots of staged parser libraries!}
-%%
-%% the next two lines define the bibliography style to be used, and
+\subsection{Impact on bug-finding ability}
+\jwc{Staging type-derived generators (which are the most commonly used ones) can turn a timeout into a found bug!}
+\section{Conclusion \& Future Work}
+\section{Related Work}
+\jwc{A generator is a parser of randomness, and people have implemented lots of staged parser libraries!}
+%%
+%% The next two lines define the bibliography style to be used, and
 %% the bibliography file.
-\bibliographystyle{acm-reference-format}
+\bibliographystyle{ACM-Reference-Format}
 \bibliography{bib}
@@
 %%
-%% if your work has an appendix, this is the place to put it.
+%% If your work has an appendix, this is the place to put it.
 \end{document}
@@
 %%
-%% end of file `sample-sigplan.tex'.
+%% End of file `sample-sigplan.tex'.

### 2025-03-15 paper/paper.tex
wrote splitmix section

@@
 
-\subsection{Faster SplitMix with Unboxed Int}
-\label{subsection:faster-rng}
-
+\subsection{Faster SplitMix with Unboxed \texttt{int64}s}
 \label{subsection:faster-rng}
@@
 at run time as a \emph{pointer} to an unscanned block of memory containing (among other things)
-a 64-bit integer \cn.
-
-This means that all operations that return an \texttt{int64} must allocate this block of memory.
+a 64-bit integer \cn. This means that all operations that return an \texttt{int64} must allocate this block of memory.
 This has a significant impact on the performance of generators. A single call to
@@
 a large performance impact on some generators that spend most of their time sampling data.
-
-% \jwc{
-%   \begin{itemize}
-%     \item as we discussed in section~\ref{section:motiv}, 
-%     \item faster rngs and sources of entropy exist, though they are extensionally inequivalent. 
-%     \item here's how splitmix is slow: the core of splitmix is nextint64, which does allocations for each operation.
-%     \item if we could use actual int64s, we wouldn't have to allocate.
-%     \item so we do this:
-%     \begin{itemize}
-%       \item we implement splitmix in c (which uses normal int64s) and real int64 arithmetic, and use the ocaml ffi to call out to the c functions. t
-%       \item this could be done with jane street bleeding-edge compiler extensions, though these are not yet implemented in a a version of the ocaml compiler that also has staging.
-%     \end{itemize}
-%   \end{itemize}
-% }
+To circumvent this allocation and provide an equivalent version of \texttt{splittable\_random}, we
+reimplement SplitMix in C, and call out to it with the OCaml FFI. The C version of the library uses
+proper \texttt{int64\_t} arithmetic, only boxing and unboxing integers at the call boundaries between OCaml and C code.
+Ideally in the future, one would not need to call out to C for this: the Jane Street bleeding-edge OCaml compiler has support
+for unboxed types \cn, which (among other things) would let us implement a version of SplitMix that does not allocate, directly
+in OCaml. Unfortunately, the Jane Street branch of the compiler is incompatible with MetaOCaml, which we use to implement the metaprogramming
+discussed in the previous sections.
 

### 2025-03-15 paper/paper.tex
smithing words about alloc

@@
 algorithm multiple times. Each sample from \texttt{splittable\_random} allocates 9
-words \jwc{this is a call to \texttt{next\_int64}}, and each allocation brings us closer
-to the next garbage collection pause. While allocation in OCaml is fast and the garbage colletor
-runs incrementally in small slices
-\footnote{The OCaml GC is a generational collector \cn, and since these allocations are mostly very short lived, they will never be promoted.
-}, we will see in Section~\ref{section:eval} that this can have
+times \jwc{this is a call to \texttt{next\_int64}}, and each allocation brings us closer
+to the next garbage collection pause. While small allocations like these are \emph{very} fast to perform and subsequently collect in OCaml,
+\footnote{The OCaml GC is a generational collector \cn, and since these allocations are small and mostly very short lived, they will all be minor allocations, never to be promoted.}, we will see in Section~\ref{section:eval} that this can have
 a large performance impact on some generators that spend most of their time sampling data.

### 2025-03-15 paper/paper.tex
3.3

@@
 
-The core of any PBT generator library is a random number generator: to generate random values, we need
+The core of any PBT generator library is a random number generator: to generate
+random values, we need \hg{...}
 Different PBT libraries use a wide range of different RNGs.
-Following the original Haskell QuickCheck, \bq uses the SplitMix algorithm \cn,
-implemented (statefully) as an OCaml library called \texttt{Splittable\_random}. Meanwhile,
+Following the original Haskell QuickCheck implementation, \bq\ uses the SplitMix
+algorithm \cn,
+implemented as a (stateful) OCaml library called \texttt{Splittable\_random}. Meanwhile,
 ScalaCheck uses the JSF algorithm \cn.
@@
 
-The RNG sits at the heart of the generation hot path.
-Even basic generators like generating a single \texttt{int} or \texttt{float} may sample from the RNG multiple times,
-and generator combinators like \texttt{list} generate many \texttt{int}s.
+The RNG sits at the heart of the hot path.
+Even basic generators, like ones generating a single \texttt{int} or \texttt{float}, may sample from the RNG multiple times,
+and generator combinators like \texttt{list} call those basic generators. \hg{To
+generate a single list with 3 elements, \bq's list generator calls the RNG \dots times!}
 Because of this, the speed of a single RNG call matters a great deal. Unfortunately,
-we argue that existing PBT libraries make relatively inefficient choices on this front,
+\proposecut{we argue that} existing PBT libraries make relatively inefficient choices on this front,
 leading to worse bugfinding power than what is possible.
 
+\hg{This paragraph feels a bit clumsy right now. What if we started with a table
+of the RNGs used by a bunch of popular PBT frameworks and then argued that
+things like Lehmer are faster?}
 For example, significantly faster RNG algorithms than SplitMix or JSF exist,
-such as the Lehmer RNG \cn exist. In microbenchmarks, the Lehmer generator almost 2x as fast as SplitMix \cn.
+such as the Lehmer RNG \cn. In microbenchmarks, the Lehmer generator almost 2x as fast as SplitMix \cn.
 % https://github.com/lemire/testingRNG/tree/master
 Moreover, PBT libraries could even consider eschewing the requirement that a source of
-entropy pass statistical tests like BigCrush\cn.  This is common practice in other areas of testing already:
+entropy pass statistical tests like BigCrush\cn.\hg{This needs more discussion}  This is common practice in other areas of testing already:
 fuzzers often simply use a buffer full of arbitrary bytes as a source of entropy
@@
 be faster on modern CPUs than any RNG with data-dependent arithmetic
-instructions.
-
-In the context of this paper, however, demonstrating the benefits of any of the
+instructions.\hg{Should we just always be doing this? I think we want to argue
+that this is a bridge too far because you still need to generate the buffer
+ahead of time and you can run out of randomness?}
+
+\hg{Alternative framing: ``Of course, simply arguing that the RNG is on the hot
+path does not guarantee that a faster RNG leads to measurably faster
+generation; for that, we need an experiment. The most obvious experiment is to
+simply swap out the RNG for \bq\ with a totally different, faster one. But, as
+discussed previously, generators with different generation orders can be
+difficult to compare. To get around this, we exploit a ``natural experiment'': \dots''}
+In the context of this paper, however, clearly demonstrating the benefits of any of the
 above faster options is tricky. Our bug-finding evaluation's power is predicated on
 the fact that all of the different versions of a generator we test are
-extensionally identical: for a given seed and size parameter, they all produce
+extensionally\hg{Is this how we're explaining this? I'm not sure that's quite right} identical: for a given seed and size parameter, they all produce
 the \emph{same} value.  Using a different random number generator or different

### 2025-03-15 paper/paper.tex
skeleton of explanation about pw equivalence

--- 2025-03-15 paper/paper.tex
@@
 \jwc{
+Explain effect ordering here!!
+\begin{itemize}
+  \item \texttt{Bq.bind} has the CBV sampling semantics
+  \item If you \texttt{Bq.bind} \texttt{x} and then use it twice, both uses refer to the \emph{same} value.
+  \item Sampling is effectful: it mutates the internal state.
+  \item Permuting two independent binds gives you a generator that is distributionally equivalent, but not equivlaent when considered up to pointwise equality as \emph{functions} from seed to value.
+  \item Note that it's much more sensible to compare the bugfinding performance of two function-equivalent generators:
+\end{itemize}
+}
+
+\jwc{
   \begin{itemize}
@@
 \jwc{This generator generates binary trees using all the combinators, yay. Explain em.}
-
-\jwc{Explain effect ordering here!!}
 

### 2025-03-15 paper/paper.tex
intro

@@
 \jwc{``Property based testing is a race against time...''}
+\\
+\tr{Property-based testing (PBT) uses generators of random data to produce inputs to a
+system under test, which is then checked against a set of properties the system is
+ expected to satisfy.  In practice, PBT is used as a lightweight, ad hoc approach 
+ to software correctness, aimed at detecting bugs within a short testing 
+ window---typically 50 milliseconds to 30 seconds. Given this practical constraint, 
+ it is 
+ imperative that inputs be generated as fast as possible; in other words, 
+ generators must be treated as performance-sensitive code.}
+
+\tr{The locus for improvement in this area is \textit{generator libraries}: if programmers are 
+ provided with the tools to write fast generators, then they do not themselves 
+ need to be performance experts. However, existing libraries fail to achieve good performance 
+ due to two major sources of inefficiency. The first is \textit{random sampling}, or the process
+ of selecting random values. The first is random sampling, or the process of selecting random values. 
+ The second is abstraction: while generator libraries are designed for flexibility and expressiveness, 
+ this comes at the cost of performance, resulting in frequent boxing and unboxing and unnecessary 
+ allocations (which lead to costly garbage collection pauses). In strict functional languages, 
+ constructs like monadic bind introduce function call overhead, requiring each step in a generator 
+ sequence to allocate closures and thread state through function applications.
+ }
 

### 2025-03-15 paper/paper.tex
intro

--- 2025-03-15 paper/paper.tex
@@
 \usepackage{subcaption}
-
+\usepackage{xspace}
 \lstset{
@@
 \jwc{``Property based testing is a race against time...''}
-
+\\
+\tr{Property-based testing (PBT) uses generators of random data to produce inputs to a
+system under test, which is then checked against a set of properties the system is
+ expected to satisfy.  In practice, PBT is used as a lightweight, ad hoc approach 
+ to software correctness, aimed at detecting bugs within a short testing 
+ window---typically 50 milliseconds to 30 seconds. Given this practical constraint, 
+ it is 
+ imperative that inputs be generated as fast as possible; in other words, 
+ generators must be treated as performance-sensitive code.}
+
+\tr{The locus for improvement in this area is \textit{generator libraries}: if programmers are 
+ provided with the tools to write fast generators, then they do not themselves 
+ need to be performance experts. However, existing libraries fail to achieve good performance 
+ due to two major sources of inefficiency. The first is \textit{random sampling}, or the process
+ of selecting random values. (Another sentence here?)
+ The second is abstraction: generator libraries are implemented as embedded DSLS, which are flexible
+ and expressive, but also introduce significant overhead that is not necessarily optimized away by 
+ the compiler. These costs compound in the aggregate, raising a fundamental
+ question: can generator libraries produce code that is both expressive \textit{and} efficient?
+ }
+ \\
+\tr{
+  % in strict functionallanguages like OCaml and Scala
+We outline a fully general, two-stage approach to designing generator libraries, 
+which provide high-level combinator functions with dramatically reduced 
+overhead. First, \rand choice matters. We demonstrate that using a faster \rand
+significantly impacts generator performance by directly optimizing an existing RNG, which
+produces the same values as its unoptimized counterpart---but faster--thereby allowing us
+to compare the RNG's impact on an existing generator library. Second, we 
+describe a method of optimizing generator libraries using \textit{staged metaprogramming}, 
+or ``staging''. Staged metaprogramming allows for code generation, which can mitigate 
+performance challenges introduced by embedded DSLs. In particular, when a generator DSL 
+introduces overhead for dynamic control flow, staging can specialize that control flow
+and compile it statically. For example, when choosing between a list of options, a staged
+generator can avoid allocating the list of options at runtime by expanding to an 
+if-else. This approach is not tied to any particular PBT library---instead, it gives a recipe
+for improving even SOTA PBT libraries.}
+\tr{To demonstrate
+the effectiveness of our technique, we apply our technique to a leading PBT library, \bq, built with performance engineering in mind.
+Our version of this library, termed \name, is a direct drop-in replacement for the original that is completely
+semantically equivalent: generators written using it produce exactly the same values as 
+those written in the original version. However, in \name, the abstractions provided by the generator DSL are 
+zero-cost, and the \rand is highly-optimized. The direct equivalence between \name and \bq
+allows us to perform extremely fine-grained comparisons between generators---generators implemented with both versions of the
+library will produce exactly the same sequence of values, so differences in testing performance are directly attributable
+to \name's performance speedup.
+}
+\tr{
+  To test \name's effectiveness, we run a bunch of experiments to show that \name runs faster, allocates less, etc., than
+  \bq on a variety of different workloads. We demonstrate the generality of the technique by implementing \name in
+  Scala, and benchmarking the same generators in the Scala port of the library. Finally, as a sanity check, we demonstrate
+  that \name allows you to find bugs significantly faster ... .
+}
+\\
+\tr{ Introduce waffle house... }
+\tr{Harry's ``methodology'' paragraph...}
 Property-based testing (PBT) is a widely-used testing framework\hg{PBT isn't a

### 2025-03-15 paper/paper.tex
unpolished

--- 2025-03-15 paper/paper.tex
@@
 
-Property-based testing (PBT) is a widely-used testing framework\hg{PBT isn't a
-framework, it's a ``technique'' or ``approach'' --- Base\_QuickCheck is a framework} consisting of
-two key components: a set of \textit{properties} that a system must satisfy, and
-a way of producing
-a large number of \textit{inputs} to that system. In contrast to traditional
-unit testing, where test inputs are hand-written, PBT users create a random input
-\textit{generator} that automatically produces inputs. Sometimes, this process
-can itself be automated through the use of PBT libraries that synthesize
-generators from type definitions; however, when inputs are required to maintain
-invariants not communicated by the type, users must write a custom generator (or
-filter out a broad class of invalid inputs).
-
-\hg{I think this paragraph buries the lede. I'd start the paragraph talking
-about performance and citing the 50ms number, and then jump right into what's
-slow.}
-Writing ``good'' generators—ones that successfully uncover bugs in the system under
-test—is challenging. Much effort has been dedicated to this problem, often by
-developing sophisticated domain-specific languages for specifying constrained
-generators. However, these approaches do not always reflect the needs of real
-PBT users. Under ideal conditions, a PBT suite would continue to run
-until it stopped finding bugs, real-world usage patterns are very different; a
-study of expert PBT users reported time budgets of between 50 milliseconds and
-30 seconds for their test suites~\cite{inpractice}. Consequently, generation
-time is a significant factor in PBT efficacy---that is, if a generator can
-produce inputs twice as fast, it has twice as many chances within a given time
-bound to find a bug. Therefore, it is imperative that PBT libraries are built
-with performance in mind.
-
-To better understand and address performance challenges in the landscape of PBT
-generators, we explore popular libraries across several widely-used programming
-languages.\hg{Again, this sentence also starts the paragraph too slowly; rather than
-focus on our exploration (which isn't really the point of the paper) lead with
-the result --- that abstraction overhead, boxing/unboxing, and allocations slow
-down QuickCheck ports} We find that the flexibility of these libraries comes at a cost to 
-performance,
-introducing abstraction overhead, frequent boxing and unboxing, and avoidable
-allocations (which lead to costly garbage collection pauses). We focus primarily 
-on OCaml's \bq, selected for its efficiency relative to other OCaml PBT libraries 
-and its integration with industrial build systems\hg{Let's not try to justify
-our choice of OCaml \bq here --- we should just talk about strict functional
-languages and justify this choice later}, where it is expected to run 
-within tight time constraints. These
-inefficiencies raise a fundamental question: how can PBT libraries produce code
-that is both efficient \textit{and} general? By analyzing
-\bq\ and similar libraries in other languages \tr{Rust?
-More?}, we provide a broader perspective on common performance challenges in PBT
-and explore potential strategies for more efficient generator implementations.
-
-Our solution is \name, a generator library\hg{Can we say ``an approach to PBT
-generator libraries'' or something more general like that?} that preserves
-\bq’s functionality while improving performance through
-\textit{multi-stage programming} (staging). Staging is a lightweight,
-domain-specific compilation technique that allows us to completely eliminate the
-runtime overhead of many common generator abstractions. \jwc{This relies on the fact that the code of generators is known statically at compile time} For instance, in
-\name, monadic operations are zero-cost.
-\name’s staged eliminates unnecessary allocations,
-generates static control structures, and leverages unboxed integer libraries to
-achieve performance improvements over \texttt{base\_quickcheck}, making it
-well-suited to real-world PBT applications.
-
-We \hg{also} present a new methodology for evaluating the relative effectiveness of PBT
-generators.\hg{Might be an over-claim, we'll need to write this carefully}
-Prior work has compared generators based on bug-finding ability: 
-generators are considered equally effective if they find bugs in a
-system at the same rate over a large number of trials ~\cite{etna}. However,
-because \name\ preserves the semantics of
-\bq, a \name\ generator produces exactly
-the same inputs as its \bq\ counterpart. By comparing
-generators based on \textit{program equality}, we ensure that any speedups in
-bug-finding ability stem from performance enhancements rather than variations in
-input distribution. This approach establishes a foundation for benchmarking
-optimizations that enhance performance without compromising expressiveness.
-\hg{I'm not totally convinced by this paragraph. Here's how I'd explain this:\\
-In order to justify the switching cost associated with using \name, we designed
-\name\ to be incredibly easy to benchmark: most PBT generator optimizations make
-changes up to {\em distributional equivalence}, allowing the precise sampling
-order to differ as long as the probability distribution is maintained, but
-\name's optimizations are correct up to {\em program equivalence}. This means
-that a \name\ generator produces precisely the same values in the same order as
-the original generator, only faster. With this setup, benchmarking becomes
-significantly easier than for prior work---our measurements have incredibly low
-variance, and performance gains from \name\ translate directly to improved
-bug-finding speed.}
-
-Finally, we conduct an extensive evaluation, implementing both type-derived and
-custom generators in \name\ and \bq. Our
-generators produce a diverse range of inputs, including recursive data
-structures, lambda calculus terms, and regular expressions. We benchmark
-generation time, resource usage, and time-to-failure for each pair of staged and
-unstaged generators, demonstrating that \name\ achieves
-significant gains in speed and resource.
-
-In summary, we dramatically improve the bug-finding effectiveness of PBT generators
-by optimizing them for speed. Specifically, we make the following contributions:
+Property-based testing (PBT) uses generators of random data to produce inputs to a
+system under test, which is then checked against a set of properties the system is
+ expected to satisfy.  In practice, PBT is used as a lightweight, ad hoc approach 
+ to software correctness, aimed at detecting bugs within a short testing 
+ window---typically 50 milliseconds to 30 seconds. Given this practical constraint, 
+ it is imperative that inputs be generated as fast as possible; in other words, 
+ generators must be treated as performance-sensitive code.
+
+ The locus for improvement in this area is \textit{generator libraries}: if programmers are 
+ provided with the tools to write fast generators, then they do not themselves 
+ need to be performance experts. However, existing libraries fail to achieve good performance 
+ due to two major sources of inefficiency. The first is \textit{random sampling}, or the process
+ of selecting random values. (Another sentence here?)
+ The second is abstraction: generator libraries are implemented as embedded DSLS, which are flexible
+ and expressive, but also introduce significant overhead that is not necessarily optimized away by 
+ the compiler. These costs compound in the aggregate, raising a fundamental
+ question: can generator libraries produce code that is both expressive \textit{and} efficient?
+
+  % in strict functionallanguages like OCaml and Scala
+We outline a fully general, two-stage approach to designing generator libraries, 
+which provide high-level combinator functions with dramatically reduced 
+overhead. First, \rand choice matters. We demonstrate that using a faster \rand
+significantly impacts generator performance by directly optimizing an existing RNG, which
+produces the same values as its unoptimized counterpart---but faster--thereby allowing us
+to compare the RNG's impact on an existing generator library. Second, we 
+describe a method of optimizing generator libraries using \textit{staged metaprogramming}, 
+or ``staging''. Staged metaprogramming allows for code generation, which can mitigate 
+performance challenges introduced by embedded DSLs. In particular, when a generator DSL 
+introduces overhead for dynamic control flow, staging can specialize that control flow
+and compile it statically. For example, when choosing between a list of options, a staged
+generator can avoid allocating the list of options at runtime by expanding to an 
+if-else. This approach is not tied to any particular PBT library---instead, it gives a recipe
+for improving even SOTA PBT libraries.
+
+To demonstrate
+the effectiveness of our technique, we apply our technique to a leading PBT library, \bq, built with performance engineering in mind.
+Our version of this library, termed \name, is a direct drop-in replacement for the original that is completely
+semantically equivalent: generators written using it produce exactly the same values as 
+those written in the original version. However, in \name, the abstractions provided by the generator DSL are 
+zero-cost, and the \rand is highly-optimized. The direct equivalence between \name and \bq
+allows us to perform extremely fine-grained comparisons between generators---generators implemented with both versions of the
+library will produce exactly the same sequence of values, so differences in testing performance are directly attributable
+to \name's performance speedup.
+
+To test \name's effectiveness, we run a bunch of experiments to show that \name runs faster, allocates less, etc., than
+\bq on a variety of different workloads. We demonstrate the generality of the technique by implementing \name in
+Scala, and benchmarking the same generators in the Scala port of the library. Finally, as a sanity check, we demonstrate
+that \name allows you to find bugs significantly faster ...
+
+We make the following contributions:
 \begin{enumerate}
-    \item An empirical analysis of the sources of inefficiency in PBT
-    generators. \jwc{And an argument that PBT generator libraries should be
-    engineered for performance}
-    \hg{Let's talk about this, I'm not sure I buy it}
-    \item A library, \name, which offers efficient \jwc{performant?} generator \proposechange{functions}{combinators} through the
-    use of multi-stage programming; this is the first known application of
-    staging to PBT.\hg{Let's also talk about this}
-    \item The insight that generators should be compared by program equality
-    \item An evaluation demonstrating the improved performance of generators constructed
-    using our library in a controlled comparison.
+    \item We identify two key sources of inefficiencies in PBT generator libraries, which can significantly impact performance: 
+    namely, \rand choice, and abstraction overhead. We describe a general procedure to address the latter concern based on staged metaprogramming. \secref{section:motiv}
+    \item We implement a version of \bq, \name, that contains our optimizations and maintains strict program equality. \secref{section:4}
+    \item We evaluate \name by demonstrating the improved performance of generators constructed
+    using our library in a controlled comparison, as well as its impact on bug-finding ability. \secref{section:eval}
 \end{enumerate}
@@
 \section{Evaluation}
+\label{section:eval}
 \subsection{Implementing and Testing Generators}

### 2025-03-15 paper/paper.tex
lil cut

--- 2025-03-15 paper/paper.tex
@@
 \jwc{``Property based testing is a race against time...''}
-\\
-\tr{Property-based testing (PBT) uses generators of random data to produce inputs to a
-system under test, which is then checked against a set of properties the system is
- expected to satisfy.  In practice, PBT is used as a lightweight, ad hoc approach 
- to software correctness, aimed at detecting bugs within a short testing 
- window---typically 50 milliseconds to 30 seconds. Given this practical constraint, 
- it is 
- imperative that inputs be generated as fast as possible; in other words, 
- generators must be treated as performance-sensitive code.}
-
-\tr{The locus for improvement in this area is \textit{generator libraries}: if programmers are 
- provided with the tools to write fast generators, then they do not themselves 
- need to be performance experts. However, existing libraries fail to achieve good performance 
- due to two major sources of inefficiency. The first is \textit{random sampling}, or the process
- of selecting random values. The first is random sampling, or the process of selecting random values. 
- The second is abstraction: while generator libraries are designed for flexibility and expressiveness, 
- this comes at the cost of performance, resulting in frequent boxing and unboxing and unnecessary 
- allocations (which lead to costly garbage collection pauses). In strict functional languages, 
- constructs like monadic bind introduce function call overhead, requiring each step in a generator 
- sequence to allocate closures and thread state through function applications.
- }
 
@@
 Even basic generators, like ones generating a single \texttt{int} or \texttt{float}, may sample from the RNG multiple times,
-and generator combinators like \texttt{list} call those basic generators. \hg{To
+and generator combinators like \texttt{list} call those basic generators.
+\hg{To
 generate a single list with 3 elements, \bq's list generator calls the RNG \dots times!}
 Because of this, the speed of a single RNG call matters a great deal. Unfortunately,
-\proposecut{we argue that} existing PBT libraries make relatively inefficient choices on this front,
+existing PBT libraries make relatively inefficient choices on this front,
 leading to worse bugfinding power than what is possible.
@@
 things like Lehmer are faster?}
+
 For example, significantly faster RNG algorithms than SplitMix or JSF exist,
@@
 difficult to compare. To get around this, we exploit a ``natural experiment'': \dots''}
+
 In the context of this paper, however, clearly demonstrating the benefits of any of the

### 2025-03-16 paper/paper.tex
incorporate harry comment

@@
 and generator combinators like \texttt{list} call those basic generators.
-\hg{To
-generate a single list with 3 elements, \bq's list generator calls the RNG \dots times!}
+For instance, to generate a single list with 3 elements, 
+\hg{To generate a single list with 3 elements, \bq's list generator calls the RNG \dots times!}
 Because of this, the speed of a single RNG call matters a great deal. Unfortunately,
@@
 
-\hg{Alternative framing: ``Of course, simply arguing that the RNG is on the hot
+\hg{Alternative framing: ``'': \dots''}
+Of course, simply arguing that the RNG is on the hot
 path does not guarantee that a faster RNG leads to measurably faster
@@
 discussed previously, generators with different generation orders can be
-difficult to compare. To get around this, we exploit a ``natural experiment'': \dots''}
-
-In the context of this paper, however, clearly demonstrating the benefits of any of the
-above faster options is tricky. Our bug-finding evaluation's power is predicated on
-the fact that all of the different versions of a generator we test are
-extensionally\hg{Is this how we're explaining this? I'm not sure that's quite right} identical: for a given seed and size parameter, they all produce
-the \emph{same} value.  Using a different random number generator or different
-source of entropy would break this property. In Section~\ref{eval}, we
-demonstrate that faster RNG matters for bugfinding by exploting a ``natural
-experiment'': OCaml's \texttt{Splittable\_random} library is slow in a way that
+difficult to compare. To get around this, we exploit a ``natural experiment:
+OCaml's \texttt{Splittable\_random} library is slow in a way that
 can be improved \emph{without} changing its extensional behavior.
@@
 performance can be improved just by using a more performant RNG.
-
-% \jwc{There are a few aspects of PBT that make it interesting when it comes to PBT choice.
-% \begin{itemize}
-%   \item Doesn't need to be cryptographic.
-%   \item doesn't even \emph{necessarily} need to pass sophisticated staistical
-%   tests.  \cn
-% \end{itemize}
-% }
-
-% \jwc{
-%   \begin{itemize}
-%     \item In this paper demonstrate this fact by noticing a ``natural experiment'': OCaml's splittable random library is
-%     slow in a way that can be improved \emph{without} changing the extensional behavior. By resolving this issue, we can demonstrate
-%     just how much you can improve bug-finding performance simply by speeding up the RNG.
-%   \end{itemize}
-% }
-
-% In Section~\ref{subsection:faster-rng}, we show how our library mitigates this
-% performance hit by using OCaml's FFI to reimplement the RNG in non-allocating C code.
-% We also discuss how the ``unboxed types'' extension to the OCaml compiler \cn could instead be used
-% to the same effect without leaving OCaml.
-
-% It is worth noting that the inefficiency of boxed \texttt{int64}s is not due to
-% the abstraction overhead of the generator DSL itself, but simply due to language
-% implementation and algorithm choices. We include this discussion --- and
-% the corresponding evaluation
-% --- to demonstrate the testing-time savings of taking seriously the notion that PBT libraries
-% should be engineered for performance.
-
-% \jwc{If you're not using the fasteest RNG possible, you're leaving RNG on the table.}
 

### 2025-03-16 paper/paper.tex
rip out mentions of RNG

@@
 % \subsubsection{Function call overhead}
-\subsection{Choice of RNG}
+\subsection{Choice of Randomness Library}
 \jwc{We really need to have explained the effect ordering equivalence aspect of the evaluation before here.}
 
-The core of any PBT generator library is a random number generator: to generate
-random values, we need \hg{...}
-Different PBT libraries use a wide range of different RNGs.
+The core of any PBT generator library is a source of randomness: to generate
+random values of some datatype, we some access to random numbers!
+Different PBT libraries use different randomness libraries implementing different algorithms
+\footnote{The common term for such an algorithm or library is a ``Random Number
+Generator'' (RNG) we will avoid this term and instead say ``randomness library''
+to avoid confusing RNG implementations with the PBT gnenerator libraries that
+use them.}.
 Following the original Haskell QuickCheck implementation, \bq\ uses the SplitMix
-algorithm \cn,
-implemented as a (stateful) OCaml library called \texttt{Splittable\_random}. Meanwhile,
+algorithm \cn, implemented as a (stateful) OCaml library called \texttt{Splittable\_random}. Meanwhile,
 ScalaCheck uses the JSF algorithm \cn.
@@
 
-The RNG sits at the heart of the hot path.
-Even basic generators, like ones generating a single \texttt{int} or \texttt{float}, may sample from the RNG multiple times,
-and generator combinators like \texttt{list} call those basic generators.
-For instance, to generate a single list with 3 elements, 
-\hg{To generate a single list with 3 elements, \bq's list generator calls the RNG \dots times!}
-Because of this, the speed of a single RNG call matters a great deal. Unfortunately,
+The randomness library sits at the heart of the hot path.
+Even basic generators --- like ones generating a single \texttt{int} or \texttt{float} uniformly within a range --- can sample \emph{unboundedly many}
+random numbers, since they usually use versions of rejection sampling to find a value within the range \cn.
+Moreover, generator combinators like \texttt{list} usually make $O(n)$ calls to even those basic generators.
+Because of this, the speed of a single sample matters a great deal. Unfortunately,
 existing PBT libraries make relatively inefficient choices on this front,
@@
 
-For example, significantly faster RNG algorithms than SplitMix or JSF exist,
-such as the Lehmer RNG \cn. In microbenchmarks, the Lehmer generator almost 2x as fast as SplitMix \cn.
+For example, significantly faster algorithms than SplitMix or JSF exist,
+such as the Lehmer algorithm\cn. In microbenchmarks, the Lehmer algorithm runs almost 2x as fast as SplitMix \cn.
 % https://github.com/lemire/testingRNG/tree/master
 Moreover, PBT libraries could even consider eschewing the requirement that a source of
-entropy pass statistical tests like BigCrush\cn.\hg{This needs more discussion}  This is common practice in other areas of testing already:
+entropy pass statistical tests like BigCrush\cn. \hg{This needs more discussion} 
+This is common practice in other areas of testing already:
 fuzzers often simply use a buffer full of arbitrary bytes as a source of entropy
@@
 pointer and reading from memory (which can be pipelined trivially) will always
-be faster on modern CPUs than any RNG with data-dependent arithmetic
+be faster on modern CPUs than any random sampling algorithm with data-dependent arithmetic
 instructions.\hg{Should we just always be doing this? I think we want to argue
@@
 
-\hg{Alternative framing: ``'': \dots''}
-Of course, simply arguing that the RNG is on the hot
-path does not guarantee that a faster RNG leads to measurably faster
+Of course, simply arguing that the randomness library is on the hot
+path does not guarantee that a faster sampling leads to measurably faster
 generation; for that, we need an experiment. The most obvious experiment is to
-simply swap out the RNG for \bq\ with a totally different, faster one. But, as
+simply swap out the randomness library for \bq{} with a totally different, faster one. But, as
 discussed previously, generators with different generation orders can be
@@
 By building a version that uses much faster ``unboxed'' 64-bit integer arithmetic, we can demonstrate just how much bug-finding
-performance can be improved just by using a more performant RNG.
-
-\section{Eliminating Abstraction Overhead of Generator DSLs by Staging and Faster RNG}
+performance can be improved just by using a more performant randomness library.
+
+\section{Eliminating Abstraction Overhead of Generator DSLs by Staging and Faster Randomness Libraries}
+\jwc{This section needs a much better title...}
 \jwc{Emphasize that this is an \emph{COMPLETELY EQUIVALENT drop in replacement!} We din't just build a different library with different distributions.}
@@
 
-\hg{TODO: Signposting --- I got a few paragraphs in and realized I wasn't sure
-what I was reading}
-
-\subsection{Basic Design}
+\hg{TODO: Signposting --- I got a few paragraphs in and realized I wasn't sure what I was reading}
+
+\subsection{Design of a Staged Generator DSL}
 \label{subsection:basic-design}
@@
 and \texttt{k .<a>. size\_c random\_c} run at compile time.
-\texttt{Gen.int} is the generator that samples an int from the RNG.
+\texttt{Gen.int} is the generator that samples an int from the randomness library.
 Given any size and random seed, it returns a code block that calls \texttt{SR.int} with that random seed. Because the lower and upper bounds
@@
 effect ordering: effectful combinators like \texttt{Gen.int} do their \emph{own} let-insertion, ensuring
-that a program like \texttt{Gen.bind Gen.int (fun x -> ...)} generates a let-binding for the result of sampling the RNG.
+that a program like \texttt{Gen.bind Gen.int (fun x -> ...)} generates a let-binding for the result of sampling the randomness library.
 For example, \texttt{bind (int .<0>. .<1.>) (fun cx -> return .<(.~cx,.~cx)>.)} now correctly generates
@@
 \jwc{Unclear what we should call this section}
-As we discussed in Section~\ref{section:motiv}, inefficient RNG choice is
+As we discussed in Section~\ref{section:motiv}, choosing an inefficient randomness library is
 another bottleneck for finding bugs fast. While generator libraries by
-and large use sensible RNG algorithms, they are not explicitly chosen
-with performance in mind. Indeed, much faster RNGs \cn and fast sources
+and large use sensible sources of randomness, they are not explicitly chosen
+with performance in mind. Indeed, much faster randomness libraries\cn and fast sources
 of entropy \cn exist. \jwc{... more here?}
-To demonstrate that a faster RNG can significanty impact bugfinding power, we
+To demonstrate that faster random sampling can significanty impact bugfinding power, we
 use the natural experiment provided by OCaml's inefficient implementation of
-SplitMix. By replacing this slow RNG with a faster but extensionally equivalent
-implemetnation, we are able to precisely quantify the bugfinding speedup that using a RNG gives across a range of PBT scenarios.
-
-We emphasize that while we believe that the insight that faster RNG translates to faster bug finding
+SplitMix. By replacing this slow randomness library with a faster but extensionally equivalent
+implemetnation, we are able to precisely quantify the bugfinding speedup that using a randomness library gives across a range of PBT scenarios.
+
+We emphasize that while we believe that the insight that faster sampilng translates to faster bug finding
 is a cross-langauge one, the specific technical contents of this section are OCaml-specific.
-In most other PBT frameworks \cn, the RNG operates on by machine
+In the case of most other PBT frameworks \cn, the randomness library used operates on by machine
 integers, so this \emph{particular} inefficiency does not exist.

### 2025-03-16 paper/paper.tex
rough

@@
 
-Property-based testing (PBT) uses generators of random data to produce inputs to a
-system under test, which is then checked against a set of properties the system is
- expected to satisfy.  In practice, PBT is used as a lightweight, ad hoc approach 
- to software correctness, aimed at detecting bugs within a short testing 
- window---typically 50 milliseconds to 30 seconds. Given this practical constraint, 
- it is imperative that inputs be generated as fast as possible; in other words, 
- generators must be treated as performance-sensitive code.
-
- The locus for improvement in this area is \textit{generator libraries}: if programmers are 
- provided with the tools to write fast generators, then they do not themselves 
- need to be performance experts. However, existing libraries fail to achieve good performance 
- due to two major sources of inefficiency. The first is \textit{random sampling}, or the process
- of selecting random values. (Another sentence here?)
- The second is abstraction: generator libraries are implemented as embedded DSLS, which are flexible
- and expressive, but also introduce significant overhead that is not necessarily optimized away by 
- the compiler. These costs compound in the aggregate, raising a fundamental
- question: can generator libraries produce code that is both expressive \textit{and} efficient?
+Property-based testing (PBT) uses generators of random data to produce 
+inputs to a system under test, which is then checked against a set of 
+properties the system is expected to satisfy.  In practice, PBT is used 
+as a lightweight, ad hoc approach to software correctness, aimed at 
+detecting bugs within a short testing window---typically 50 milliseconds 
+to 30 seconds. Given this practical constraint, it is imperative that 
+inputs be generated as fast as possible; in other words, generators must be 
+treated as performance-sensitive code.
+
+The locus for improvement in this area is \textit{generator libraries}: if 
+programmers are provided with the tools to write fast generators, then they 
+do not themselves need to be performance experts. However, existing generator 
+libraries fail to achieve good performance due to two major sources of 
+inefficiency. The first is \textit{random sampling}, or the process of 
+selecting random values, which, depending on one's choice of \rand and
+the number of samples needed, can constitute an unexpectedly high proportion
+of a generator's runtime. The second is \textit{abstraction}: while flexible 
+and expressive, generator libraries rely on dynamic control structures that
+allocate at runtime. Allocations compound in the aggregate, leading to costly
+garbage collection pauses. This raises a fundamental 
+question: can generator libraries produce code that is both expressive 
+\textit{and} efficient?
 
   % in strict functionallanguages like OCaml and Scala
-We outline a fully general, two-stage approach to designing generator libraries, 
-which provide high-level combinator functions with dramatically reduced 
-overhead. First, \rand choice matters. We demonstrate that using a faster \rand
-significantly impacts generator performance by directly optimizing an existing RNG, which
-produces the same values as its unoptimized counterpart---but faster--thereby allowing us
-to compare the RNG's impact on an existing generator library. Second, we 
-describe a method of optimizing generator libraries using \textit{staged metaprogramming}, 
-or ``staging''. Staged metaprogramming allows for code generation, which can mitigate 
-performance challenges introduced by embedded DSLs. In particular, when a generator DSL 
-introduces overhead for dynamic control flow, staging can specialize that control flow
-and compile it statically. For example, when choosing between a list of options, a staged
-generator can avoid allocating the list of options at runtime by expanding to an 
-if-else. This approach is not tied to any particular PBT library---instead, it gives a recipe
-for improving even SOTA PBT libraries.
-
-To demonstrate
-the effectiveness of our technique, we apply our technique to a leading PBT library, \bq, built with performance engineering in mind.
-Our version of this library, termed \name, is a direct drop-in replacement for the original that is completely
-semantically equivalent: generators written using it produce exactly the same values as 
-those written in the original version. However, in \name, the abstractions provided by the generator DSL are 
-zero-cost, and the \rand is highly-optimized. The direct equivalence between \name and \bq
-allows us to perform extremely fine-grained comparisons between generators---generators implemented with both versions of the
-library will produce exactly the same sequence of values, so differences in testing performance are directly attributable
-to \name's performance speedup.
+We outline a fully general, two-stage approach to designing generator 
+libraries so that they retain their idiomatic abstractions with 
+significantly improved performance. First, we demonstrate the impact of
+RNG choice by optimizing an existing \rand. Our optimized RNG produces 
+exactly 
+the same values as its unoptimized counterpart, allowing us to directly 
+measure the impact of this choice on generator performance. Second, 
+we propose a method of streamlining generator libraries using 
+\textit{staged metaprogramming}, or ``staging''. \tr{Joe---maybe add
+a sentence here about parsing literature?} Using staging, it is possible
+to eliminate the overhead of many common generator abstractions. For example,
+where an unstaged generator DSL might represent a choice between
+a list of options as a dynamically-allocated array of closures, a staged 
+generator can expand to an if-else tree at compile 
+time---eliminating the allocation entirely. This approach is not tied to any
+particular PBT library, but rather provides a recipe for improving even
+state-of-the-art implementations.
+
+To demonstrate the effectiveness of our technique, we apply it to a leading PBT 
+library,\\ \bq, built with performance engineering in mind.
+Our version of this library, termed \name, is a drop-in replacement for \bq which 
+is completely semantically equivalent: that is,  generators written using \name 
+produce exactly the same values as those written in \bq. However, in \name, 
+the abstractions provided by the generator DSL are zero-cost, and the \rand is 
+highly-optimized. The direct equivalence between \name and \bq allows us to 
+perform extremely fine-grained comparisons between generators, as generators 
+implemented using both versions of the library will produce exactly the same 
+sequence of values; so, differences in testing performance are directly 
+attributable to \name's performance speedup.
+\tr{Not really done... }
 
@@
 
+In summary, we demonstrate that, across languages, monadic PBT generator DSLs can
+have a significant performance overhead, and present a cross-language technique
+for eliminating the overhead while preserving the idiomatic abstraction.
 We make the following contributions:

### 2025-03-16 paper/paper.tex
comments on S1

@@
 properties the system is expected to satisfy.  In practice, PBT is used 
-as a lightweight, ad hoc approach to software correctness, aimed at 
+as a lightweight, ad hoc approach to software correctness,
+\jwc{Move this bit about how it's used to the first sentence. I.e. PBT is a testing technqique where random data is fed to a system under test...}
+aimed at 
 detecting bugs within a short testing window---typically 50 milliseconds 
-to 30 seconds. Given this practical constraint, it is imperative that 
+to 30 seconds.
+\jwc{This sells us a little short: we should say something to the effect of ``Recent work on understanding how PBT is used in practice has determined that practitioners often run PBT with very short time budgets...''} Given this practical constraint, it is imperative that
 inputs be generated as fast as possible; in other words, generators must be 
@@
 
-The locus for improvement in this area is \textit{generator libraries}: if 
+The locus for improvement in this area \jwc{``This area'' sounds like it's talking about the whole area of PBT...
+Maybe ``a high-leverage way to improve the performance of generators is to...''} is \textit{generator libraries}: if
 programmers are provided with the tools to write fast generators, then they 
 do not themselves need to be performance experts. However, existing generator 
-libraries fail to achieve good performance due to two major sources of 
-inefficiency. The first is \textit{random sampling}, or the process of 
-selecting random values, which, depending on one's choice of \rand and
+libraries fail to achieve good performance \jwc{sounds too absolute --- who's to say what ``good performance'' is?} due to two major sources of 
+inefficiency. The first \jwc{swap these? we talk about them in the other order} is \textit{random sampling}, or the process of 
+selecting random \proposechange{values}{numbers} \jwc{i'd suggest we use ``values'' to mean BSTs and things}, which, depending on one's choice of \rand and
 the number of samples needed, can constitute an unexpectedly high proportion
-of a generator's runtime. The second is \textit{abstraction}: while flexible 
+of a generator's runtime \jwc{Nice, like this}. The second is \textit{abstraction}: while flexible 
 and expressive, generator libraries rely on dynamic control structures that
 allocate at runtime. Allocations compound in the aggregate, leading to costly
-garbage collection pauses. This raises a fundamental 
+garbage collection pauses.
+\jwc{Allocation cost is a slightly OCaml-specific problem compared to the more general issue that the abstractions (binds, etc) are \emph{opaque} to the compiler, so it can't optimize the code at all.} This raises a fundamental
 question: can generator libraries produce code that is both expressive 
-\textit{and} efficient?
+\textit{and} efficient? \jwc{love this end to the paragraph.}
 
   % in strict functionallanguages like OCaml and Scala
-We outline a fully general, two-stage approach to designing generator 
+We outline a \proposecut{fully general}, two-stage \jwc{two-pronged? this clashes with staging i.e. metaprogramming} approach to designing generator 
 libraries so that they retain their idiomatic abstractions with 
-significantly improved performance. First, we demonstrate the impact of
+significantly improved performance \jwc{simplify: idiomatic and performant}. First, we demonstrate the impact of
 RNG choice by optimizing an existing \rand. Our optimized RNG produces 
@@
 the same values as its unoptimized counterpart, allowing us to directly 
-measure the impact of this choice on generator performance. Second, 
-we propose a method of streamlining generator libraries using 
+measure the impact of this choice on generator performance.
+\jwc{Permute?
+I think staging \emph{then} randomness.} \jwc{Also, i don't love this as the first place we explain the drop-in equivalence of the RNG.
+It's unmotivated why wen want this.} Second,
+we propose a method of streamlining \jwc{``streamlining feels too vague here''} generator libraries using 
 \textit{staged metaprogramming}, or ``staging''. \tr{Joe---maybe add
-a sentence here about parsing literature?} Using staging, it is possible
+a sentence here about parsing literature?} \jwc{Maybe something like: ``Staging has long been used across domains as a technique of eliminating the overhead of idiomatic functional DSLs (cite a bunch of shit).''} Using staging, it is possible
 to eliminate the overhead of many common generator abstractions. For example,
+\jwc{I think we should use return/bind as the simplest example, since then you don't have to get into the weeds of allocation or closures.
+The functional abstractions of a monadic generator DSL block the compiler from generating fast code.}
+\jwc{Then present this case as ``more sophisticated use of staging'' or something?}
 where an unstaged generator DSL might represent a choice between
@@
 particular PBT library, but rather provides a recipe for improving even
-state-of-the-art implementations.
+state-of-the-art implementations \jwc{Sentence structure doesn't follow. Instead maybe ``not any particular, but a recipe for improving in general''}.
 
 To demonstrate the effectiveness of our technique, we apply it to a leading PBT 
-library,\\ \bq, built with performance engineering in mind.
+library, \bq, built with performance engineering in mind. \jwc{Unclear if this is true, I think we should just say that by our testing it's the fastest in OCaml.}
 Our version of this library, termed \name, is a drop-in replacement for \bq which 
-is completely semantically equivalent: that is,  generators written using \name 
+is completely semantically equivalent: that is, \jwc{given identical random seeds} generators written using \name 
 produce exactly the same values as those written in \bq. However, in \name, 
@@
 \tr{Not really done... }
+\jwc{The preceeding paragraph is \emph{really good}}
 
 To test \name's effectiveness, we run a bunch of experiments to show that \name runs faster, allocates less, etc., than
-\bq on a variety of different workloads. We demonstrate the generality of the technique by implementing \name in
+\bq on a variety of different workloads.
+\jwc{Move this to previous para.
+I think this para should start with ``To demonstrate the generality'', paralleling the ``to demonstrate the effectiveness''}We demonstrate the generality of the technique by implementing \name in
 Scala, and benchmarking the same generators in the Scala port of the library. Finally, as a sanity check, we demonstrate

### 2025-03-16 paper/paper.tex
less rough

--- 2025-03-16 paper/paper.tex
@@
 
-\jwc{This paper is about ``Strict functional languages like OCaml and Scala'', but we'll focus on OCaml for presentation.}
-
-\bcp{The current introduction says, roughly, ``We took a look at
-  base.quickcheck, noticed some inefficiencies, applied known
-  techniques from staged metaprogramming to eliminate them, and
-  observed that speed went up.''  I think we can make much stronger
-  claims, but I'm not sure about exactly how strong or what belongs in
-  the foreground...
-  \begin{itemize}
-  \item One of the first serious uses of staging in anger?
-  \item A use of staging that is novel / challenging in
-  itself, in some way?
-  \item An analysis of sources of inefficiency across a range of PBT
-  frameworks (and a solution that applies to many of them)?
-  \item A usable tool that addresses and overcomes some significant
-  implementation challenges?
-  \item Careful measurements showing where the sources of inefficiency
-  are in existing PBT tools, and which ones matter most?
-  \item (Your claim here...)?
-  \end{itemize}
-}
-
-\jwc{Agree: I think something like the 3rd is the best reframing. Something
-like: ``We demonstrate that, across languages, monadic PBT generator DSLs can
+%\jwc{This paper is about ``Strict functional languages like OCaml and Scala'', but we'll focus on OCaml for presentation.}
+
+%\bcp{The current introduction says, roughly, ``We took a look at
+%  base.quickcheck, noticed some inefficiencies, applied known
+%  techniques from staged metaprogramming to eliminate them, and
+%  observed that speed went up.''  I think we can make much stronger
+%  claims, but I'm not sure about exactly how strong or what belongs in
+%  the foreground...
+%  \begin{itemize}
+%  \item One of the first serious uses of staging in anger?
+%  \item A use of staging that is novel / challenging in
+%  itself, in some way?
+%  \item An analysis of sources of inefficiency across a range of PBT
+%  frameworks (and a solution that applies to many of them)?
+%  \item A usable tool that addresses and overcomes some significant
+%  implementation challenges?
+%  \item Careful measurements showing where the sources of inefficiency
+%  are in existing PBT tools, and which ones matter most?
+%  \item (Your claim here...)?
+%  \end{itemize}
+%}
+
+%\jwc{Agree: I think something like the 3rd is the best reframing. Something
+%like: ``We demonstrate that, across languages, monadic PBT generator DSLs can
+%have a significant performance overhead, and present a cross-language technique
+%for eliminating the overhead while preserving the idiomatic abstraction.
+%''}
+
+%\jwc{People have been using staged metaprogrammign to eliminate abstraction
+%overheads in parsing for a long time, we turn that lens on generation (Which looks like parsing)}
+
+%\jwc{Another angle that I'd argue needs to come out more is that we demonstrate that
+%engineering PBT libraries with performance in mind has significant impact on testing power.
+%}
+
+%\jwc{
+%Let's lead on with the quote from PBT in practice: performance engineering
+%matters.  De-emphasize the abstraction overhead framing. Eliminating abstraction
+%overhead is important, so is using the fastest RNG you can possibly use.
+%}
+
+%\jwc{``Property based testing is a race against time...''}
+
+Property-based testing (PBT) is a testing technique where random data
+produced by a \textit{generator} is fed
+to a system under test, which is then checked against a set of properties 
+it is expected to satisfy. Recent work on understanding how PBT is used in 
+practice has determined that practitioners often run PBT with very short time 
+budgets---typically 50 milliseconds to 30 seconds. Given this practical 
+constraint, it is imperative that inputs be generated as fast as possible; 
+in other words, generators must be treated as performance-sensitive code.
+A high-leverage way to improve the performance of generators is to optimize 
+\textit{generator libraries}: if
+programmers are provided with the tools to write fast generators, then they 
+do not themselves need to be performance experts.
+
+However, the performance of 
+existing generator libraries falls short of what is possible due to two major 
+sources of inefficiency. The first of these is \textit{abstraction}: while 
+flexible and expressive, generator libraries rely on complex abstractions
+and indirections that are difficult for compilers to fully simplify, resulting
+ in additional function calls and closure allocations.
+Our second insight involves the process of 
+selecting random values (``random sampling''), which, depending on one's choice of \rand and
+the number of samples needed, can constitute an unexpectedly high proportion
+of a generator's runtime. This raises a fundamental 
+question: can generator libraries produce code that is both expressive 
+\textit{and} efficient?
+%In practice, PBT is used 
+%as a lightweight, ad hoc approach to software correctness,
+%\jwc{Move this bit about how it's used to the first sentence. I.e. PBT is a testing technqique where random data is fed to a system under test...}
+%aimed at 
+%detecting bugs within a short testing window---typically 50 milliseconds 
+%to 30 seconds.
+\jwc{This sells us a little short: we should say something to the effect of ``Recent work on understanding how PBT is used in practice has determined that practitioners often run PBT with very short time budgets...''} Given this practical constraint, it is imperative that
+%inputs be generated as fast as possible; in other words, generators must be 
+%treated as performance-sensitive code.
+
+% The locus for improvement in this area \jwc{``This area'' sounds like it's talking about the whole area of PBT...
+Maybe ``a high-leverage way to improve the performance of generators is to...''} is \textit{generator libraries}: if
+
+  % in strict functionallanguages like OCaml and Scal
+To answer this, we propose a two-pronged approach to designing generator libraries 
+that preserves their expressiveness while eliminating unnecessary overhead.
+First, we propose a method of optimizing
+generator libraries using \textit{staged metaprogramming}, or ``staging'', which has 
+long been used across domains as a technique of eliminating the overhead 
+of idiomatic functional DSLs \todo{CITE!!!}.
+Using staging, it is possible to eliminate the overhead of many common 
+generator abstractions. For example,
+each call to monadic bind in \bq constructs and then immediately executes a 
+closure at runtime; meanwhile, in \name, bind resolves at compile time,
+allowing the compiler to inline away the closures. Secondly, we demonstrate the
+impact of different choices of \rand on generator performance 
+by conducting a controlled comparison. Together, these insights provide a
+recipe for improving even state-of-the-art generator libraries.
+
+%\jwc{I think we should use return/bind as the simplest example, since then you don't have to get into the weeds of allocation or closures.
+The functional abstractions of a monadic generator DSL block the compiler from generating fast code.}
+%\jwc{Then present this case as ``more sophisticated use of staging'' or something?}
+%where an unstaged generator DSL might represent a choice between
+%a list of options as a dynamically-allocated array of closures, a staged 
+%generator can expand to an if-else tree at compile 
+%time---eliminating the allocation entirely.
+% This approach is not tied to any
+%particular PBT library, but rather provides a recipe for improving even
+%state-of-the-art implementations \jwc{Sentence structure doesn't follow. Instead maybe ``not any particular, but a recipe for improving in general''}.
+
+%First, we demonstrate the impact of
+%RNG choice by optimizing an existing \rand. Our optimized RNG produces 
+%exactly 
+%the same values as its unoptimized counterpart, allowing us to directly 
+%measure the impact of this choice on generator performance.
+\jwc{Permute?
+I think staging \emph{then} randomness.} \jwc{Also, i don't love this as the first place we explain the drop-in equivalence of the RNG.
+It's unmotivated why wen want this.}
+
+To demonstrate the effectiveness of our technique, we apply it to \bq,
+which is the fastest property-based testing library in OCaml according to our tests.
+%\jwc{Unclear if this is true, I think we should just say that by our testing it's the fastest in OCaml.}
+Our version of this library, \name, is a drop-in replacement for \bq that 
+is \textit{completely semantically equivalent}: that is, given identical random seeds, generators written using \name 
+produce exactly the same values as those written in \bq. However, in \name, 
+the abstractions provided by the generator DSL are zero-cost, and the \rand is 
+highly-optimized. The direct equivalence between \name and \bq allows us to 
+perform extremely fine-grained comparisons between generators, as generators 
+implemented using both versions of the library will produce exactly the same 
+sequence of values. Therefore, differences in performance are directly 
+attributable to \name's speedup.
+
+To evaluate \name's effectiveness, 
+implement a series of workloads in each generator library, and demonstrate that 
+\name runs faster and allocates less than \bq. Further, we show that this performance benefit
+translates to bug-finding ability by running these workloads in Etna, a PBT platform that 
+simulates real-world usage by injecting bugs into a system and measuring how quickly different 
+generators can detect them.
+Finally, to demonstrate the generality of our technique, we implement \name in Scala
+and benchmark the same generators in the Scala port of the library---thereby highlighting 
+its portability
+to other strict functional languages and reinforcing its potential for widespread adoption.
+
+In summary, we demonstrate that, across languages, monadic PBT generator DSLs
 have a significant performance overhead, and present a cross-language technique
-for eliminating the overhead while preserving the idiomatic abstraction.
-''}
-
-\jwc{People have been using staged metaprogrammign to eliminate abstraction
-overheads in parsing for a long time, we turn that lens on generation (Which looks like parsing)}
-
-\jwc{Another angle that I'd argue needs to come out more is that we demonstrate that
-engineering PBT libraries with performance in mind has significant impact on testing power.
-}
-
-\jwc{
-Let's lead on with the quote from PBT in practice: performance engineering
-matters.  De-emphasize the abstraction overhead framing. Eliminating abstraction
-overhead is important, so is using the fastest RNG you can possibly use.
-}
-
-\jwc{``Property based testing is a race against time...''}
-
-Property-based testing (PBT) uses generators of random data to produce inputs to a
-system under test, which is then checked against a set of properties the system is
- expected to satisfy.  In practice, PBT is used as a lightweight, ad hoc approach 
- to software correctness, aimed at detecting bugs within a short testing 
- window---typically 50 milliseconds to 30 seconds. Given this practical constraint, 
- it is imperative that inputs be generated as fast as possible; in other words, 
- generators must be treated as performance-sensitive code.
-
- The locus for improvement in this area is \textit{generator libraries}: if programmers are 
- provided with the tools to write fast generators, then they do not themselves 
- need to be performance experts. However, existing libraries fail to achieve good performance 
- due to two major sources of inefficiency. The first is \textit{random sampling}, or the process
- of selecting random values. (Another sentence here?)
- The second is abstraction: generator libraries are implemented as embedded DSLS, which are flexible
- and expressive, but also introduce significant overhead that is not necessarily optimized away by 
- the compiler. These costs compound in the aggregate, raising a fundamental
- question: can generator libraries produce code that is both expressive \textit{and} efficient?
-
-  % in strict functionallanguages like OCaml and Scala
-We outline a fully general, two-stage approach to designing generator libraries, 
-which provide high-level combinator functions with dramatically reduced 
-overhead. First, \rand choice matters. We demonstrate that using a faster \rand
-significantly impacts generator performance by directly optimizing an existing RNG, which
-produces the same values as its unoptimized counterpart---but faster--thereby allowing us
-to compare the RNG's impact on an existing generator library. Second, we 
-describe a method of optimizing generator libraries using \textit{staged metaprogramming}, 
-or ``staging''. Staged metaprogramming allows for code generation, which can mitigate 
-performance challenges introduced by embedded DSLs. In particular, when a generator DSL 
-introduces overhead for dynamic control flow, staging can specialize that control flow
-and compile it statically. For example, when choosing between a list of options, a staged
-generator can avoid allocating the list of options at runtime by expanding to an 
-if-else. This approach is not tied to any particular PBT library---instead, it gives a recipe
-for improving even SOTA PBT libraries.
-
-To demonstrate
-the effectiveness of our technique, we apply our technique to a leading PBT library, \bq, built with performance engineering in mind.
-Our version of this library, termed \name, is a direct drop-in replacement for the original that is completely
-semantically equivalent: generators written using it produce exactly the same values as 
-those written in the original version. However, in \name, the abstractions provided by the generator DSL are 
-zero-cost, and the \rand is highly-optimized. The direct equivalence between \name and \bq
-allows us to perform extremely fine-grained comparisons between generators---generators implemented with both versions of the
-library will produce exactly the same sequence of values, so differences in testing performance are directly attributable
-to \name's performance speedup.
-
-To test \name's effectiveness, we run a bunch of experiments to show that \name runs faster, allocates less, etc., than
-\bq on a variety of different workloads. We demonstrate the generality of the technique by implementing \name in
-Scala, and benchmarking the same generators in the Scala port of the library. Finally, as a sanity check, we demonstrate
-that \name allows you to find bugs significantly faster ...
-
+for eliminating that overhead while preserving their idiomatic abstractions.
 We make the following contributions:
@@
     \item We identify two key sources of inefficiencies in PBT generator libraries, which can significantly impact performance: 
-    namely, \rand choice, and abstraction overhead. We describe a general procedure to address the latter concern based on staged metaprogramming. \secref{section:motiv}
-    \item We implement a version of \bq, \name, that contains our optimizations and maintains strict program equality. \secref{section:4}
+    namely, \rand choice, and abstraction overhead. We describe a general procedure to address the latter based on staged metaprogramming. \secref{section:motiv}
+    \item We implement a version of \bq, \name, that showcases our optimization technique while maintaining strict program equality. \secref{section:4}
     \item We evaluate \name by demonstrating the improved performance of generators constructed

### 2025-03-16 paper/paper.tex
another pass

@@
 
-Property-based testing (PBT) is a testing technique where random data
-produced by a \textit{generator} is fed
-to a system under test, which is then checked against a set of properties 
-it is expected to satisfy. Recent work on understanding how PBT is used in 
-practice has determined that practitioners often run PBT with very short time 
-budgets---typically 50 milliseconds to 30 seconds. Given this practical 
-constraint, it is imperative that inputs be generated as fast as possible; 
-in other words, generators must be treated as performance-sensitive code.
+Property-based testing (PBT) is a technique in which a \textit{generator} produces random inputs to a system under test, which is then evaluated against a set of properties the system is expected to satisfy.
+Recent studies on PBT usage have shown that practitioners often run tests with very short time budgets---typically between 50 milliseconds and 30 seconds.
+Given this constraint, generating inputs as quickly as possible is crucial for maximizing the odds of finding a bug in the limited time available.
+In other words, generators must be treated as performance-sensitive code.
 A high-leverage way to improve the performance of generators is to optimize 
@@
 programmers are provided with the tools to write fast generators, then they 
-do not themselves need to be performance experts.
+do not themselves need to be performance engineers.
 
 However, the performance of 
-existing generator libraries falls short of what is possible due to two major 
-sources of inefficiency. The first of these is \textit{abstraction}: while 
-flexible and expressive, generator libraries rely on complex abstractions
-and indirections that are difficult for compilers to fully simplify, resulting
- in additional function calls and closure allocations.
-Our second insight involves the process of 
-selecting random values (``random sampling''), which, depending on one's choice of \rand and
-the number of samples needed, can constitute an unexpectedly high proportion
-of a generator's runtime. This raises a fundamental 
+existing generator libraries falls short of what is possible. This is due to two major 
+sources of inefficiency. First, the same high-level design that gives generator libraries their flexibility and expressiveness also introduces overhead:
+layered abstractions and indirections hinder the compiler's ability to optimize computations, shifting more work to runtime.
+Second, the process of 
+selecting random values (``random sampling'') can constitute an unexpectedly high proportion
+of a generator's runtime, depending on one's choice of \rand and
+the number of samples needed. This raises a fundamental 
 question: can generator libraries produce code that is both expressive 
@@
   % in strict functionallanguages like OCaml and Scal
-To answer this, we propose a two-pronged approach to designing generator libraries 
+To answer this, we propose a two-pronged approach for designing generator libraries 
 that preserves their expressiveness while eliminating unnecessary overhead.
-First, we propose a method of optimizing
+First, we describe a method of optimizing
 generator libraries using \textit{staged metaprogramming}, or ``staging'', which has 
-long been used across domains as a technique of eliminating the overhead 
-of idiomatic functional DSLs \todo{CITE!!!}.
+long been used to eliminating overhead 
+from idiomatic functional DSLs \todo{CITE!!!}.
 Using staging, it is possible to eliminate the overhead of many common 
-generator abstractions. For example,
-each call to monadic bind in \bq constructs and then immediately executes a 
-closure at runtime; meanwhile, in \name, bind resolves at compile time,
-allowing the compiler to inline away the closures. Secondly, we demonstrate the
+generator abstractions. For example, in unstaged generator libraries, each call to monadic bind constructs and immediately executes a closure at runtime.
+Meanwhile, a staged approach can resolve the bind at compile time, enabling the compiler to inline and eliminate these closures.
+Secondly, we demonstrate the
 impact of different choices of \rand on generator performance 
@@
 
-To demonstrate the effectiveness of our technique, we apply it to \bq,
-which is the fastest property-based testing library in OCaml according to our tests.
+To demonstrate the effectiveness of our technique, we apply it to \bq, the fastest property-based testing library in OCaml according to our tests.
 %\jwc{Unclear if this is true, I think we should just say that by our testing it's the fastest in OCaml.}
-Our version of this library, \name, is a drop-in replacement for \bq that 
-is \textit{completely semantically equivalent}: that is, given identical random seeds, generators written using \name 
+Our staged version of this library,\\ \name, is a direct drop-in replacement for \bq that 
+is \textit{completely semantically equivalent}: that is, given identical random seeds, generators written using \\\name 
 produce exactly the same values as those written in \bq. However, in \name, 
@@
 highly-optimized. The direct equivalence between \name and \bq allows us to 
-perform extremely fine-grained comparisons between generators, as generators 
+perform extremely fine-grained comparisons, as generators 
 implemented using both versions of the library will produce exactly the same 
@@
 
-To evaluate \name's effectiveness, 
-implement a series of workloads in each generator library, and demonstrate that 
-\name runs faster and allocates less than \bq. Further, we show that this performance benefit
-translates to bug-finding ability by running these workloads in Etna, a PBT platform that 
-simulates real-world usage by injecting bugs into a system and measuring how quickly different 
-generators can detect them.
+To evaluate \name's effectiveness, we
+implement a series of workloads in each generator library, and show that 
+\name runs faster and allocates less than \\\bq. Further, we show that this performance benefit
+translates to bug-finding ability by running these workloads in Etna, a platform that 
+simulates real-world PBT usage by injecting bugs into a system and measuring how quickly various 
+generators detect them.
 Finally, to demonstrate the generality of our technique, we implement \name in Scala
@@
 
-In summary, we demonstrate that, across languages, monadic PBT generator DSLs
-have a significant performance overhead, and present a cross-language technique
+In summary, we establish that monadic PBT generator DSLs
+have a significant performance overhead across languages, and present a cross-language technique
 for eliminating that overhead while preserving their idiomatic abstractions.

### 2025-03-16 paper/paper.tex
minor

@@
 generator libraries using \textit{staged metaprogramming}, or ``staging'', which has 
-long been used to eliminating overhead 
+long been used for eliminating overhead 
 from idiomatic functional DSLs \todo{CITE!!!}.
-Using staging, it is possible to eliminate the overhead of many common 
-generator abstractions. For example, in unstaged generator libraries, each call to monadic bind constructs and immediately executes a closure at runtime.
+We extend this technique to generators. In unstaged generator libraries, each call to monadic bind constructs and immediately executes a closure at runtime.
 Meanwhile, a staged approach can resolve the bind at compile time, enabling the compiler to inline and eliminate these closures.
@@
     \item We identify two key sources of inefficiencies in PBT generator libraries, which can significantly impact performance: 
-    namely, \rand choice, and abstraction overhead. We describe a general procedure to address the latter based on staged metaprogramming. \secref{section:motiv}
+    namely, \rand choice, and abstraction overhead. \secref{section:motiv}
+    \item We apply staged metaprogramming to the novel domain of generators. \secref{section:}
     \item We implement a version of \bq, \name, that showcases our optimization technique while maintaining strict program equality. \secref{section:4}

### 2025-03-16 paper/paper.tex
minor

--- 2025-03-16 paper/paper.tex
@@
 generator libraries using \textit{staged metaprogramming}, or ``staging'', which has 
-long been used to eliminating overhead 
-from idiomatic functional DSLs \todo{CITE!!!}.
-Using staging, it is possible to eliminate the overhead of many common 
-generator abstractions. For example, in unstaged generator libraries, each call to monadic bind constructs and immediately executes a closure at runtime.
+long been used for opetimizing idiomatic functional DSLs \todo{CITE!!!}.
+We extend this technique to generators, completely eliminating the overhead of many common generator abstractions.
+For example, in unstaged generator libraries, each call to monadic bind constructs and immediately executes a closure at runtime.
 Meanwhile, a staged approach can resolve the bind at compile time, enabling the compiler to inline and eliminate these closures.
@@
 its portability
-to other strict functional languages and reinforcing its potential for widespread adoption.
-
-In summary, we establish that monadic PBT generator DSLs
-have a significant performance overhead across languages, and present a cross-language technique
-for eliminating that overhead while preserving their idiomatic abstractions.
+to other strict functional languages.
+
+In summary, we show that monadic PBT generator DSLs incur significant performance costs across languages, and present a general technique for improving their efficiency while preserving their idiomatic abstractions.
 We make the following contributions:

### 2025-03-16 paper/paper.tex
minor

@@
 generator libraries using \textit{staged metaprogramming}, or ``staging'', which has 
-long been used for opetimizing idiomatic functional DSLs \todo{CITE!!!}.
+long been used for optimizing idiomatic functional DSLs \todo{CITE!!!}.
 We extend this technique to generators, completely eliminating the overhead of many common generator abstractions.

### 2025-03-16 paper/paper.tex
intro pass

@@
 
-Property-based testing (PBT) is a technique in which a \textit{generator} produces random inputs to a system under test, which is then evaluated against a set of properties the system is expected to satisfy.
-Recent studies on PBT usage have shown that practitioners often run tests with very short time budgets---typically between 50 milliseconds and 30 seconds.
-Given this constraint, generating inputs as quickly as possible is crucial for maximizing the odds of finding a bug in the limited time available.
-In other words, generators must be treated as performance-sensitive code.
+Property-based testing (PBT) is a technique in which a \textit{generator}
+produces random inputs to a system under test, which is then evaluated against a
+set of properties the system is expected to satisfy.  Recent studies on PBT
+usage have shown that practitioners often run tests with very short time
+budgets --- typically between 50 milliseconds and 30 seconds \cite{inpractice} \jwc{phrasing could suggest ``race against time'' aspect.}. Given this
+constraint, generating inputs as quickly as possible is crucial for maximizing
+the odds of finding a bug in the limited time available. In other words,
+generators must be treated as performance-sensitive code.
 A high-leverage way to improve the performance of generators is to optimize 
@@
 programmers are provided with the tools to write fast generators, then they 
-do not themselves need to be performance engineers.
+do not themselves need to be performance engineers. \jwc{... to find bugs fast}
 
@@
 sources of inefficiency. First, the same high-level design that gives generator libraries their flexibility and expressiveness also introduces overhead:
-layered abstractions and indirections hinder the compiler's ability to optimize computations, shifting more work to runtime.
+layered abstractions and indirections hinder the compiler's ability to optimize computations, shifting more work to runtime. \jwc{Nice!!}
 Second, the process of 
-selecting random values (``random sampling'') can constitute an unexpectedly high proportion
+selecting random \proposechange{values}{numbers} \jwc{We should reserve ``values'' for BSTs/etc...
+otherwise this sentence is confusing.
+What else does would a generator's runtime consist of, except for generating values?}(``random sampling'') can constitute an unexpectedly high proportion
 of a generator's runtime, depending on one's choice of \rand and
-the number of samples needed. This raises a fundamental 
+the number of samples needed. This \jwc{Unclear referent here} raises a fundamental 
 question: can generator libraries produce code that is both expressive 
 \textit{and} efficient?
+\jwc{Love these first two paragraphs.}
 %In practice, PBT is used 
@@
 First, we describe a method of optimizing
-generator libraries using \textit{staged metaprogramming}, or ``staging'', which has 
-long been used for optimizing idiomatic functional DSLs \todo{CITE!!!}.
+generator libraries using \textit{multi-stage programming}, or ``staging'', which has 
+long been used for optimizing idiomatic functional DSLs \cn.
 We extend this technique to generators, completely eliminating the overhead of many common generator abstractions.
-For example, in unstaged generator libraries, each call to monadic bind constructs and immediately executes a closure at runtime.
+For example, in unstaged generator libraries, each call to monadic bind constructs and immediately executes a closure at runtime \jwc{We should probably say ``usually'' here...
+in very simple scenarios the compiler can figure it out}.
+\jwc{Todo: joe --- cite some evidence for this}
 Meanwhile, a staged approach can resolve the bind at compile time, enabling the compiler to inline and eliminate these closures.
@@
 recipe for improving even state-of-the-art generator libraries.
+\jwc{Great, like this a lot.}
 
@@
 Our staged version of this library,\\ \name, is a direct drop-in replacement for \bq that 
-is \textit{completely semantically equivalent}: that is, given identical random seeds, generators written using \\\name 
+is \textit{completely semantically equivalent}: that is, given identical random seeds, generators written using \name{}
 produce exactly the same values as those written in \bq. However, in \name, 
@@
 highly-optimized. The direct equivalence between \name and \bq allows us to 
-perform extremely fine-grained comparisons, as generators 
+perform extremely fine-grained \propose{bug-finding} comparisons, as generators 
 implemented using both versions of the library will produce exactly the same 
-sequence of values. Therefore, differences in performance are directly 
-attributable to \name's speedup.
-
-To evaluate \name's effectiveness, we
-implement a series of workloads in each generator library, and show that 
-\name runs faster and allocates less than \\\bq. Further, we show that this performance benefit
+sequence of values. Therefore, differences in \proposechange{performance}{bug-finding ability} are directly 
+attributable to \name's \proposechange{speedup}{??}.
+
+To evaluate \name's effectiveness \jwc{This could flow better from the end of the last para --- we were just talking about evaluating \name{}'s effectivness.}, we
+implement a series of workloads \jwc{this is an etna term of art} in each generator library, and show that 
+\name runs faster \proposecut{and allocates less} than \\\bq. Further, we show that this performance benefit
 translates to bug-finding ability by running these workloads in Etna, a platform that 
@@
 to other strict functional languages.
+\jwc{I think we should more explicitly point out the portability earlier, maybe up where we say we're applying it to BQ.}
 
@@
     \item We identify two key sources of inefficiencies in PBT generator libraries, which can significantly impact performance: 
-    namely, \rand choice, and abstraction overhead. \secref{section:motiv}
-    \item We apply staged metaprogramming to the novel domain of generators. \secref{section:}
-    \item We implement a version of \bq, \name, that showcases our optimization technique while maintaining strict program equality. \secref{section:4}
+    namely, \rand choice, and abstraction overhead \jwc{Swap}. \secref{section:motiv}
+    \item We apply multi-stage programming to the novel domain of generators. \jwc{Less sure this should be a top-level conrib.} \secref{section:}
+    \item We implement a version of \bq, \name, that showcases our optimization technique \jwc{Calling it an ``optimization technique'' begs the ``so they just made a thing faster'' question} while maintaining strict program equality.
+\secref{section:4}
+    \jwc{We implement \emph{two} versions of \name{}! Once for BQ and once for ScalaCheck.}
     \item We evaluate \name by demonstrating the improved performance of generators constructed
@@
 \end{enumerate}
-
-\jwc{
-  Things to ensure people come away with answers to:
-  \begin{itemize}
-    \item Is there generality?
-    \item Is this important?
-    \item Did they actually solve the problem?
-  \end{itemize}
-}
 

### 2025-03-16 paper/paper.tex
Changes on Zoom with Thia

@@
 \newif\ifdraft\drafttrue{}
-\newif\iflater\latertrue{}
+\newif\iflater\laterfalse{}
 
@@
 %% complete the rights form.
-\setcopyright{acmlicensed}
-\copyrightyear{2025}
-\acmYear{2025}
-\acmDOI{XXXXXXX.XXXXXXX}
+\setcopyright{none}
+% \copyrightyear{2025}
+% \acmYear{2025}
+% \acmDOI{XXXXXXX.XXXXXXX}
 %% These commands are for a PROCEEDINGS abstract or paper.
-\acmConference[Conference acronym 'XX]{Make sure to enter the correct
-  conference title from your rights confirmation email}{June 03--05,
-  2018}{Woodstock, NY}
+% \acmConference[Conference acronym 'XX]{Make sure to enter the correct
+%   conference title from your rights confirmation email}{June 03--05,
+%   2018}{Woodstock, NY}
 %%
@@
 %%  June 03--05, 2018, Woodstock, NY}
-\acmISBN{978-1-4503-XXXX-X/2025/02}
+% \acmISBN{978-1-4503-XXXX-X/2025/02}
 
@@
 \title{Fail Faster}
+\subtitle{\hg{TODO: Subtitle}}
 
@@
 %% article.
-\begin{abstract} \tr{Abstract needs a lot of work.}
-Property-based testing (PBT) is a software testing approach that verifies a system against a large suite of automatically-generated inputs.
-Since testing pipelines typically run under strict time constraints, PBT generators must produce inputs as quickly as possible to maximize the likelihood of finding bugs in the time available.
-However, existing PBT libraries often prioritize generality at the cost of performance.
-We introduce \texttt{waffle\_house}, a high-performance generator library that uses staging, a lightweight compilation technique, to eliminate many common generator abstractions.
-To evaluate \texttt{waffle\_house}, we design a novel benchmarking methodology that compares generators based on program equivalence, isolating performance improvements from differences in input distribution.
-Using this methodology, we compare \texttt{waffle\_house} to a leading generator library, and, through extensive evaluation over a diverse range of generators, demonstrate that \texttt{waffle\_house} significantly improves generation speed and resource efficiency while matching \texttt{base\_quickcheck}'s expressiveness.
+\begin{abstract}
+  \hg{TODO: Abstract}
 \end{abstract}
@@
 %% the work being presented. Separate the keywords with commas.
-\keywords{Property-based testing, Generators, Staging, Meta-programming}
+% \keywords{Property-based testing, Generators, Staging, Meta-programming}
 %% A "teaser" image appears between the author and affiliation
@@
 
-Property-based testing (PBT) is a technique in which a \textit{generator}
-produces random inputs to a system under test, which is then evaluated against a
-set of properties the system is expected to satisfy.  Recent studies on PBT
-usage have shown that practitioners often run tests with very short time
-budgets --- typically between 50 milliseconds and 30 seconds \cite{inpractice} \jwc{phrasing could suggest ``race against time'' aspect.}. Given this
-constraint, generating inputs as quickly as possible is crucial for maximizing
-the odds of finding a bug in the limited time available. In other words,
-generators must be treated as performance-sensitive code.
-A high-leverage way to improve the performance of generators is to optimize 
-\textit{generator libraries}: if
-programmers are provided with the tools to write fast generators, then they 
-do not themselves need to be performance engineers. \jwc{... to find bugs fast}
-
-However, the performance of 
-existing generator libraries falls short of what is possible. This is due to two major 
-sources of inefficiency. First, the same high-level design that gives generator libraries their flexibility and expressiveness also introduces overhead:
-layered abstractions and indirections hinder the compiler's ability to optimize computations, shifting more work to runtime. \jwc{Nice!!}
-Second, the process of 
-selecting random \proposechange{values}{numbers} \jwc{We should reserve ``values'' for BSTs/etc...
-otherwise this sentence is confusing.
-What else does would a generator's runtime consist of, except for generating values?}(``random sampling'') can constitute an unexpectedly high proportion
-of a generator's runtime, depending on one's choice of \rand and
-the number of samples needed. This \jwc{Unclear referent here} raises a fundamental 
-question: can generator libraries produce code that is both expressive 
-\textit{and} efficient?
-\jwc{Love these first two paragraphs.}
+Property-based testing (PBT) is a technique for software testing which uses
+random test inputs to evaluate logical properties that the system is expected to
+satisfy.  Recent studies on PBT usage~\cite{inpractice} have shown that
+practitioners interleave PBT with their software development process, sometimes
+testing their properties every time they save their code. This workflow makes
+sense because the developer who just wrote the code is often the best person to
+fix a potental bug; if a bug is found later, it will need to be fixed by someone
+with less context.  All of this means that PBT is often run with very short time
+budgets --- e.g., between 50 milliseconds and 30 seconds --- and it is important
+that bugs can be found in that time-frame.
+\jwc{phrasing could suggest ``race against time'' aspect.}\hg{Agree, can't see
+how to fit this in yet}
+
+Given these strict timelines, it is important to think of PBT as
+performance-sensitive process. One important locus for performance improvement
+is the \textit{generators} that produce random inputs to the properties under
+test. These generators are written with the help of \textit{generator
+libraries}, usually expressed as embedded domain-specific languages (DSLs) that
+provide combinators for building and composing generators. Optimizing these
+libraries is a key opportunity to speed up PBT: if programmers are
+provided with the tools to write fast generators, then they do not themselves
+need to be experts in performance tuning. \jwc{... to find bugs fast}
+
+We find that the performance of existing generator libraries falls short of what
+is possible. This is due to two major sources of inefficiency.
+First, the same high-level design that gives generator libraries their
+flexibility and expressiveness also introduces overhead: layered abstractions
+and indirections hinder the compiler's optimizations, shifting more work to
+runtime.
+Second, the \rand that underlies the generator's operation --- the source of
+random numbers that are used to make choices in the generator --- 
+can constitute an unexpectedly high proportion of a generator's runtime.
+\iflater
+\hg{TODO later, I'm not 100\% sure it's clear how this particular question is
+raised by the above. We could strengthen that connection}
+\fi
+These observations together raise a fundamental question: can generator
+libraries produce code that is both expressive \textit{and} efficient?
 %In practice, PBT is used 
@@
   % in strict functionallanguages like OCaml and Scal
-To answer this, we propose a two-pronged approach for designing generator libraries 
-that preserves their expressiveness while eliminating unnecessary overhead.
-First, we describe a method of optimizing
-generator libraries using \textit{multi-stage programming}, or ``staging'', which has 
-long been used for optimizing idiomatic functional DSLs \cn.
-We extend this technique to generators, completely eliminating the overhead of many common generator abstractions.
-For example, in unstaged generator libraries, each call to monadic bind constructs and immediately executes a closure at runtime \jwc{We should probably say ``usually'' here...
-in very simple scenarios the compiler can figure it out}.
-\jwc{Todo: joe --- cite some evidence for this}
-Meanwhile, a staged approach can resolve the bind at compile time, enabling the compiler to inline and eliminate these closures.
-Secondly, we demonstrate the
-impact of different choices of \rand on generator performance 
-by conducting a controlled comparison. Together, these insights provide a
-recipe for improving even state-of-the-art generator libraries.
-\jwc{Great, like this a lot.}
+To address this question, we propose a two-pronged approach for designing
+generator libraries that preserves their expressiveness while eliminating
+unnecessary overhead.
+First, we describe a method of optimizing generator libraries using
+\textit{multi-stage programming}, or ``staging,'' which has long been used for
+optimizing idiomatic functional DSLs~\cn.
+We extend this technique to generators, completely eliminating the overhead of
+many common generator abstractions. For example, in unstaged generator
+libraries, each call to monadic bind usually spends time constructing and
+immediately executing a closure at runtime. \jwc{Todo: joe --- cite some evidence for this}
+Meanwhile, a staged approach can resolve the bind at compile time, enabling the
+compiler to inline and eliminate these closures.
+Secondly, we demonstrate the impact of different choices of \rand on generator
+performance by conducting a controlled comparison. \hg{More words here for
+parallelism?} Together, these insights provide a recipe for improving even
+state-of-the-art generator libraries.
 
@@
 
-To demonstrate the effectiveness of our technique, we apply it to \bq, the fastest property-based testing library in OCaml according to our tests.
-%\jwc{Unclear if this is true, I think we should just say that by our testing it's the fastest in OCaml.}
-Our staged version of this library,\\ \name, is a direct drop-in replacement for \bq that 
+To demonstrate the effectiveness of our technique, we apply it to \bq, a
+high-quality PBT library in OCaml, authored by Jane Street.
+Our improved version, \name, is a direct drop-in replacement for
+\bq that 
 is \textit{completely semantically equivalent}: that is, given identical random seeds, generators written using \name{}
-produce exactly the same values as those written in \bq. However, in \name, 
+produce exactly the same sequence of values as those written in \bq. However, in \name, 
 the abstractions provided by the generator DSL are zero-cost, and the \rand is 
 highly-optimized. The direct equivalence between \name and \bq allows us to 
-perform extremely fine-grained \propose{bug-finding} comparisons, as generators 
-implemented using both versions of the library will produce exactly the same 
-sequence of values. Therefore, differences in \proposechange{performance}{bug-finding ability} are directly 
-attributable to \name's \proposechange{speedup}{??}.
-
-To evaluate \name's effectiveness \jwc{This could flow better from the end of the last para --- we were just talking about evaluating \name{}'s effectivness.}, we
-implement a series of workloads \jwc{this is an etna term of art} in each generator library, and show that 
-\name runs faster \proposecut{and allocates less} than \\\bq. Further, we show that this performance benefit
-translates to bug-finding ability by running these workloads in Etna, a platform that 
-simulates real-world PBT usage by injecting bugs into a system and measuring how quickly various 
-generators detect them.
-Finally, to demonstrate the generality of our technique, we implement \name in Scala
-and benchmark the same generators in the Scala port of the library---thereby highlighting 
-its portability
-to other strict functional languages.
-\jwc{I think we should more explicitly point out the portability earlier, maybe up where we say we're applying it to BQ.}
+perform extremely fine-grained comparisons, differences in generation speed are
+directly attributable to the staging and \rand optimizations.
+
+We evaluate the optimizations in \name with a series of case studies in each
+generator library, and show that \name runs faster than \bq.
+Further, we show that this performance benefit\hg{wc?} translates to bug-finding
+ability by running these case studies in the Etna framework~\cite{etna}, a
+platform that simulates real-world PBT usage\hg{Don't love this phrase, I'll
+think more about how we should talk about Etna} by injecting bugs into a system and
+measuring how quickly various generators detect them.
+Finally, to demonstrate the generality of our technique, we implement a version
+of \name in Scala \hg{(based on scalacheck?)} and benchmark the same generators
+in the Scala port of the library---thereby highlighting its portability to other
+strict functional languages.
+\iflater
+\jwc{I think we should more explicitly point out the portability earlier, maybe
+up where we say we're applying it to BQ.}\hg{Agree}
+\fi
 

### 2025-03-16 staged-scala/src/main/scala/BST/BespokeLimits.scala
singlebespoke bst

@@ -1,0 +1,3 @@
+package BST
+
+val bst_bespoke_limits = 10

### 2025-03-16 staged-scala/src/main/scala/BST/Impl.scala

@@ -1,0 +1,11 @@
+package BST
+
+
+enum Bst:
+  case E
+  case Node(left : Bst, key : Long, value : Nat, right : Bst)
+
+  override def toString: String = this match
+    case E => "E"
+    case Node(left, key, value, right) => 
+      s"Node($left, $key, $value, $right)"

### 2025-03-16 staged-scala/src/main/scala/BST/Nat.scala

@@ -1,0 +1,25 @@
+package BST
+
+import org.scalacheck.Gen
+import org.scalacheck.Arbitrary
+import scala.quoted.*
+
+import stagedgen.StGen
+
+case class Nat(value: Long) {}
+
+object Nat {
+  def generator (limit : Long): Gen[Nat] = 
+    Gen.Choose.chooseLong.choose(Long.MinValue, Long.MaxValue).map(i => Nat(i % limit))
+
+  def staged_generator (limit : Expr[Long]) (using Quotes) : StGen[Expr[Nat]] =
+    StGen.chooseLong('{Long.MinValue},'{Long.MaxValue}).map(cl => '{Nat(${cl} % ${limit})})
+
+  extension (a: Nat)
+    def <(b: Nat): Boolean = natOrdering.lt(a, b)
+    def <=(b: Nat): Boolean = natOrdering.lteq(a, b)
+    def >(b: Nat): Boolean = natOrdering.gt(a, b)
+    def >=(b: Nat): Boolean = natOrdering.gteq(a, b)
+  
+  implicit val natOrdering: Ordering[Nat] = Ordering.by(_.value)
+}

### 2025-03-16 staged-scala/src/main/scala/BST/SingleBespoke.scala

@@ -1,0 +1,24 @@
+package BST
+import org.scalacheck.Gen
+import org.scalacheck.Arbitrary
+
+import Bst.*
+
+object SingleBespoke {
+  def go (lo : Long, hi : Long): Gen[Bst] = {
+    Gen.size.flatMap (sz =>
+      if (lo >= hi || sz <= 1) {
+        Gen.const(Bst.E)
+      } else {
+        for {
+        k <- Gen.Choose.chooseLong.choose(lo, hi)
+        v <- Nat.generator(bst_bespoke_limits)
+        l <- Gen.resize(sz / 2, go(lo, k - 1))
+        r <- Gen.resize (sz / 2, go(k + 1, hi))
+        } yield (Bst.Node (l,k,v,r))
+      }
+    )
+  }
+
+  val gen = go(0,bst_bespoke_limits)
+}

### 2025-03-16 staged-scala/src/main/scala/BST/SingleBespokeStaged.scala

@@ -1,0 +1,40 @@
+package BST
+
+import stagedgen.StGen
+import stagedgen.Splittable
+import stagedgen.Splittable.given
+import stagedgen.SplittableCps.given
+import scala.quoted.*
+
+object SingleBespokeStaged {
+    def go(using Quotes)(lo : Expr[Long], hi : Expr[Long]) : StGen[Expr[Bst]] = {
+        StGen.recursive(
+            (rec : Expr[(Long,Long)] => StGen[Expr[Bst]]) =>
+            (lohi : Expr[(Long,Long)]) =>
+            lohi.split.flatMap((lo,hi) =>
+            StGen.size.flatMap (sz =>
+                '{${lo} >= ${hi} || ${sz} <= 1}.split.flatMap(b =>
+                if(b){
+                    StGen.pure('{Bst.E})
+                } else {
+                    for{
+                        k <- StGen.chooseLong(lo,hi)
+                        v <- Nat.staged_generator('{bst_bespoke_limits})
+                        l <- StGen.resize('{${sz}/2}, rec('{(${lo}, ${k}-1)}))
+                        r <- StGen.resize ('{${sz}/2}, rec('{(${k} + 1, ${hi})}))
+                    } yield ('{Bst.Node (${l},${k},${v},${r})})
+                }
+            )
+            )
+        )
+        )('{(${lo},${hi})})
+    }
+
+    def genStaged (using Quotes) = StGen.splat (go('{0},'{bst_bespoke_limits}))
+
+    inline def gen = {
+        ${genStaged}
+    }
+}
+
+

### 2025-03-16 staged-scala/src/main/scala/BST/TypeDerived.scala

@@ -1,0 +1,4 @@
+package BST
+
+import org.scalacheck.Gen
+

### 2025-03-16 staged-scala/src/main/scala/Main.scala

@@ -4,9 +4,13 @@ import org.scalacheck.Gen
 import org.scalacheck.rng.Seed
 
 import stagedgen.StGen
-
+import BST.Bst
+import BST.SingleBespoke
+import BST.SingleBespokeStaged
 
 @main def hello(): Unit =
-  val u = StGen.complexStGen
-  val v = u(Gen.Parameters.default)(Seed.random())
-  println(v)
+  val seed = Seed.random()
+  val Some(a) = SingleBespoke.gen.apply(Gen.Parameters.default,seed)
+  val b = SingleBespokeStaged.gen(Gen.Parameters.default.size)(seed)
+  println(a.toString)
+  println(b.toString)

### 2025-03-16 staged-scala/src/main/scala/benchmark/GenBm.scala

@@ -7,8 +7,11 @@ import org.openjdk.jmh.annotations._
 import org.scalacheck.Gen
 import org.scalacheck.rng.*
 import stagedgen.StGen
+import BST.Bst
+import BST.SingleBespoke
+import BST.SingleBespokeStaged
 
-  def complexGen: Gen[(Long,Long)] = for {
+def complexGen: Gen[(Long,Long)] = for {
     x <- Gen.choose(1,1000)
     y <- Gen.choose(0,x)
   } yield (x,y)
@@ -16,7 +19,7 @@ import stagedgen.StGen
 class IllegalBoundsError[A](low: A, high: A)
     extends IllegalArgumentException(s"invalid bounds: low=$low, high=$high")
 
-  def chLng(l: Long, h: Long)(p: Gen.Parameters, seed: Seed): (Option[Long],Seed) = {
+def chLng(l: Long, h: Long)(p: Gen.Parameters, seed: Seed): (Option[Long],Seed) = {
       if (h < l) {
         throw new IllegalBoundsError(l, h)
       } else if (h == l) {
@@ -69,11 +72,40 @@ class IllegalBoundsError[A](low: A, high: A)
       }
   }
 
-  //   def flatMap[U](f: T => Gen[U]): Gen[U] = gen { (p, seed) =>
-  //   val rt = doApply(p, seed)
-  //   rt.flatMap(t => f(t).doApply(p, rt.seed))
-  // }
 
+
+// @State(Scope.Thread)
+// class GenBm {
+// @Benchmark
+// @BenchmarkMode(Array(Mode.AverageTime))
+// @OutputTimeUnit(TimeUnit.NANOSECONDS)
+// @Warmup(iterations = 2, time = 5, timeUnit = TimeUnit.SECONDS)
+// @Measurement(iterations = 3, time = 5, timeUnit = TimeUnit.SECONDS)
+// @Fork(1)
+//   def generateComplexBase(): Option[(Long,Long)] = {
+//     complexGen.sample
+//   }
+
+// @Benchmark
+// @BenchmarkMode(Array(Mode.AverageTime))
+// @OutputTimeUnit(TimeUnit.NANOSECONDS)
+// @Warmup(iterations = 2, time = 5, timeUnit = TimeUnit.SECONDS)
+// @Measurement(iterations = 3, time = 5, timeUnit = TimeUnit.SECONDS)
+// @Fork(1)
[…]

### 2025-03-16 staged-scala/src/main/scala/stagedgen/Splittable.scala

@@ -2,38 +2,58 @@ package stagedgen
 
 import scala.quoted.*
 import stagedgen.Cps
+import stagedgen.StGen
 
-trait Splittable[T : Type,S] {
-    extension (a : Expr[T]) def split(using Quotes) : Cps[S]
+trait SplittableCps[T : Type,S] {
+    extension (a : Expr[T]) def splitCps(using Quotes) : Cps[S]
 }
 
-given (using Quotes) : Splittable[Boolean,Boolean] with
+object SplittableCps {
+given (using Quotes) : SplittableCps[Boolean,Boolean] with
     extension (a : Expr[Boolean])
-        def split(using Quotes) : Cps[Boolean] = Cps.cps([Z : Type] => (k : Boolean => Expr[Z]) => '{
+        def splitCps(using Quotes) : Cps[Boolean] = Cps.cps([Z : Type] => (k : Boolean => Expr[Z]) => '{
             if ${a} then ${k(true)} else ${k(false)}
         })
 
-given [A : Type, B : Type](using Quotes) : Splittable[(A,B),(Expr[A],Expr[B])] with
+given [A : Type, B : Type](using Quotes) : SplittableCps[(A,B),(Expr[A],Expr[B])] with
     extension (a : Expr[(A,B)])
-        def split(using Quotes) : Cps[(Expr[A],Expr[B])] = Cps.cps([Z : Type] => k => '{
+        def splitCps(using Quotes) : Cps[(Expr[A],Expr[B])] = Cps.cps([Z : Type] => k => '{
             val (x,y) = ${a}
             ${k('{x},'{y})}
         })
 
-given [A : Type](using Quotes) : Splittable[Option[A],Option[Expr[A]]] with
+given [A : Type](using Quotes) : SplittableCps[Option[A],Option[Expr[A]]] with
     extension (a : Expr[Option[A]])
-        def split(using Quotes) : Cps[Option[Expr[A]]] = Cps.cps([Z : Type] => k => '{
+        def splitCps(using Quotes) : Cps[Option[Expr[A]]] = Cps.cps([Z : Type] => k => '{
             ${a} match {
                 case None => ${k(None)}
                 case Some(v) => ${k(Some('{v}))}
             }
         })
 
-given [A : Type](using Quotes) : Splittable[List[A],Either[Unit,(Expr[A],Expr[List[A]])]] with
+given [A : Type](using Quotes) : SplittableCps[List[A],Either[Unit,(Expr[A],Expr[List[A]])]] with
[…]

### 2025-03-16 staged-scala/src/main/scala/stagedgen/StGen.scala

@@ -1,14 +1,15 @@
 package stagedgen
 
-import org.scalacheck.Gen
 import org.scalacheck.rng.Seed
 import scala.quoted.*
 import stagedgen.Cps
+import stagedgen.Splittable
+import stagedgen.SplittableCps.given
 
 class IllegalBoundsError[A](low: A, high: A)
     extends IllegalArgumentException(s"invalid bounds: low=$low, high=$high")
 
-private def chLng(l: Long, h: Long)(p: Gen.Parameters, seed: Seed): (Long,Seed) = {
+private def chLng(l: Long, h: Long)(size: Int, seed: Seed): (Long,Seed) = {
       if (h < l) {
         throw new IllegalBoundsError(l, h)
       } else if (h == l) {
@@ -48,57 +49,60 @@ private def chLng(l: Long, h: Long)(p: Gen.Parameters, seed: Seed): (Long,Seed) 
     }
 
 abstract class StGen[T] { self =>
-  def doApply(p : Expr[Gen.Parameters],seed:Expr[Seed]) : Cps[(Option[T],Expr[Seed])]
+  def doApply(p : Expr[Int],seed:Expr[Seed]) : Cps[(T,Expr[Seed])]
   
 
   def map[S](f : T => S) : StGen[S] = {
     StGen.gen((p,seed) =>
-        self.doApply(p,seed).map((ot,seed2) => (ot.map(f),seed2))
+        self.doApply(p,seed).flatMap((res,seed2) =>
+          Cps.pure((f(res),seed2))
+        )
     )
   }
 
   def flatMap[S](f : T => StGen[S]) : StGen[S] = {
-    StGen.gen((p,seed) =>
-        self.doApply(p,seed).flatMap((ot,seed2) =>
-          ot match {
-            case None => Cps.pure((None,seed))
-            case Some(t) => f(t).doApply(p,seed2)
-          }
-        )
+    StGen.gen((size,seed) =>
+        self.doApply(size,seed).flatMap((t,seed2) => f(t).doApply(size,seed2))
     )
   }
 }
 
 object StGen {
 
-  def gen[T](f : (Expr[Gen.Parameters],Expr[Seed]) => Cps[(Option[T],Expr[Seed])]) : StGen[T] = {
+  def gen[T](f : (Expr[Int],Expr[Seed]) => Cps[(T,Expr[Seed])]) : StGen[T] = {
     new StGen[T]{
-      def doApply(p : Expr[Gen.Parameters],seed:Expr[Seed]) = {
[…]

### 2025-03-16 paper/paper.tex
wibbles

@@
 fix a potental bug; if a bug is found later, it will need to be fixed by someone
-with less context.  All of this means that PBT is often run with very short time
+with less context. \jwc{Not sure how the previous sentence is relevant. It delays us getting to the point.}  All of this means that PBT is often run with very short time
 budgets --- e.g., between 50 milliseconds and 30 seconds --- and it is important
@@
 and indirections hinder the compiler's optimizations, shifting more work to
-runtime.
+run time.
 Second, the \rand that underlies the generator's operation --- the source of
 random numbers that are used to make choices in the generator --- 
-can constitute an unexpectedly high proportion of a generator's runtime.
+can constitute an unexpectedly high proportion of a generator's running time.
 \iflater
@@
 optimizing idiomatic functional DSLs~\cn.
-We extend this technique to generators, completely eliminating the overhead of
+We extend this technique to generators \jwc{``apply this technology to''? We're not extending staging, we're a client}, completely eliminating the overhead of
 many common generator abstractions. For example, in unstaged generator
@@
 highly-optimized. The direct equivalence between \name and \bq allows us to 
-perform extremely fine-grained comparisons, differences in generation speed are
-directly attributable to the staging and \rand optimizations.
+perform extremely fine-grained comparisons, \propose{as} differences in \proposechange{generation speed}{bugfinding power} are
+\proposechange{directly}{solely} attributable to the staging and \rand optimizations.
 

### 2025-03-16 paper/paper.tex
staging dsls
staging background

@@
 test. These generators are written with the help of \textit{generator
-libraries}, usually expressed as embedded domain-specific languages (DSLs) that
+libraries}, usually expressed as embedded domain-specific languages (eDSLs) that
 provide combinators for building and composing generators. Optimizing these
@@
 
+\jwc{Ensure we have a signposting table-of-contents-ish few sentences here, if it's not covered by the contribs.}
+
 \section{Background}
@@
 
-\jwc{
-  \begin{itemize}
-    \item In Muli-stage programming (also known more simply as staging), programs execute over the course of multiple stages,
+In Muli-stage programming (also known more simply as staging), programs execute over the course of multiple stages,
 with each stage producing the code to be run in the next.
-\item This is an old idea, with roots going back to quasiquotation in LISP, and picking up steam again in the 1990s with MetaML \cn{}.
-\item For the purposes of this paper, we will only consider two stages: compile time and run time. Staged programs thus execute twice: once at compile time,
+This is an old idea, with roots going back to quasiquotation in LISP \cn{}, and picking up steam again in the 1990s with MetaML \cn{}.
+For the purposes of this paper, we will primarily consider two stages: compile time and run time. Staged programs thus execute twice: once at compile time,
 which produces more code, which is then compiled and run at run time.
-\item Staging has many applications, but chief among them is its use for \emph{optimization}. We can write programs such that during the compile time stage, they are partially evaluated to eliminate
-abstraction overhead. 
-\item Many languages have some degree of staging functionality, either provided as a library \cn or build directly into the language's implementation \cn.
-\item For this paper, we use MetaOCaml \cn for OCaml staging, but all of the staging concepts are portable to any other language with multi-stage programming functionality.
-  \end{itemize}
-}
-
-\jwc{This section is going to need work.}
-
+
+\subsubsection{Staging in eDSL Construction}
+
+One of the primary uses of staging is in embedded DSL construction \cite{sheard99, lms, trattdsl}.
+While embedded DSLs are great \jwc{words}, they have well-known drawback: 
+the functional abstractions used to build eDSLs prevent compilers from generating efficient machine code \cn.
+In the literature, this effect is known as ``abstraction overhead'': the layers of abstraction that 
+make eDSLs so usable are precisely what prevents them from being fast.
+In light of this issue, staging is often used as a lightweight compiler for DSLs. The compile-time evaluation stage
+transforms the DSL code, eliminating abstractions to produce code that the host language compiler can generate fast machine code for.
+\jwc{This previous paragraph is peak ``shitty first draft''.}
+This recipe has been used to great effect across domains to stage eDSLs for stream processing \cn, parser combinators \cite{sspc, staged-parsers, krishnaswami19, flap},
+and query processing \cite{rhyme}.
+
+Many languages have some degree of staging functionality, though support varies
+from full native support (Scala 3 \cn, Haskell \cn) to experimental compiler extensions (OCaml \cite{metaocaml,macocaml}, Java \cite{mint}, Kotlin \cn).
+Moreover, many more languages have lighter-weight staging-like metaprogramming support in the form of macros.
+See Section~\ref{subsection:other-langs} for a full discussion of just how portable these ideas are. \jwc{this sentence is a placeholder lol}
+
+For this paper, we have implemented staged PBT libraries in both OCaml --- using the MetaOCaml fork of the OCaml compiler \cn --- and Scala 3 --- using
+\texttt{scala.quoted}. For presentation purposes, all staged code presented in the body of this paper is written in OCaml.
+However, we believe all of the concepts presented are portable across languages with staging support, and we have reimplemented \name in Scala 3.
+
+\subsubsection{Staging in MetaOCaml}
 MetaOCaml's staging functionality is exposed through a type \texttt{'a code}. A
@@
 \subsection{Monadic DSL Abstraction Overhead}
+
 \jwc{

### 2025-03-16 staged-scala/src/main/scala/STLC/Bespoke.scala
blah

@@ -1,0 +1,63 @@
+package STLC
+import org.scalacheck.Gen
+import org.scalacheck.Arbitrary
+
+import STLC.*
+
+object Bespoke {
+    def genTyp() : Gen[Typ] =
+        Gen.size.flatMap (sz =>
+            if (sz <= 1){
+                Gen.const(Typ.TBool)
+            } else {
+                Gen.frequency(
+                    1 -> Gen.const(Typ.TBool),
+                    sz -> (for {
+                        t1 <- Gen.resize(sz/2, genTyp())
+                        t2 <- Gen.resize(sz/2, genTyp())
+                    } yield (Typ.TFun(t1,t2)))
+                )
+            }
+        )
+
+    def genVar(g : List[Typ],t : Typ) : Gen[Option[Expr]] = {
+        val vars = g.zipWithIndex.flatMap { (t2, i) => 
+            if (t == t2) { Some(Some(Expr.Var(i))) } else { None }
+        }
+        vars match {
+            case Nil => Gen.const(None)
+            case _ => Gen.oneOf(vars)
+        }
+    }
+
+    def genConst(t : Typ) : Gen[Expr] = {
+        t match {
+            case Typ.TBool => Gen.oneOf(true,false).map(b => Expr.Bool(b))
+            case Typ.TFun(t1,t2) => genConst(t2).map(e => Expr.Abs(t1,e))
+        }
+    }
+
+    def genExactExpr(g : List[Typ], t : Typ) : Gen[Expr] = {
+        genVar(g,t).flatMap(me =>
+            me match {
+                case Some(e) => Gen.const(e)
+                case None => Gen.size.flatMap(n =>
+                    if(n <= 1){
+                        genConst(t)
+                    } else {
+                        t match {
+                            case Typ.TFun(t1, t2) => Gen.resize(n-1,genExactExpr(t1::g,t2)).map(e => Expr.Abs(t1,e))
+                            case _ => (for {
+                                t2 <- genTyp()
+                                e1 <- Gen.resize(n/2,genExactExpr(g,Typ.TFun(t2,t)))
+                                e2 <- Gen.resize(n/2,genExactExpr(g,t2))
+                            } yield (Expr.App(e1,e2)))
+                        }
+                    }
+                )
+            }
+        )
+    }
+
+    val gen = genTyp().flatMap(t => genExactExpr(Nil,t))
+}

### 2025-03-16 staged-scala/src/main/scala/STLC/Impl.scala

@@ -1,0 +1,27 @@
+package STLC
+
+enum Typ:
+    case TBool
+    case TFun(dom : Typ, cod : Typ)
+
+
+enum Expr:
+    case Var(idx : Long)
+    case Bool(v : Boolean)
+    case Abs(t : Typ, body : Expr)
+    case App(fun : Expr, arg : Expr)
+
+// let rec equal x y =
+//   match x, y with
+//   | TBool, TBool -> true
+//   | TFun(x1,x2), TFun(y1,y2) -> equal x1 y1 && equal x2 y2
+//   | _ -> false
+
+// // enum Bst:
+// //   case E
+// //   case Node(left : Bst, key : Long, value : Nat, right : Bst)
+
+// //   override def toString: String = this match
+// //     case E => "E"
+// //     case Node(left, key, value, right) => 
+// //       s"Node($left, $key, $value, $right)

### 2025-03-16 staged-scala/src/main/scala/STLC/Staged.scala

@@ -1,0 +1,68 @@
+package STLC
+
+package STLC
+
+import stagedgen.StGen
+import stagedgen.Splittable
+import stagedgen.Splittable.given
+import stagedgen.SplittableCps.given
+import scala.quoted.*
+
+import STLC.* 
+
+object Staged {
+    def genTBool(using Quotes) = StGen.pure('{Typ.TBool})
+
+    def genTyp(using Quotes) : StGen[Expr[Typ]] =
+        StGen.recursive[Typ,Unit] (
+            rh => _ =>
+                StGen.size.flatMap (sz =>
+                    '{$sz <= 1}.split.flatMap(b =>
+                        if(b) {
+                            genTBool
+                        } else {
+                            StGen.frequency (
+                                '{1} -> genTBool,
+                                sz -> (for {
+                                    t1 <- StGen.resize('{$sz/2}, rh('{()}))
+                                    t2 <- StGen.resize('{$sz/2}, rh('{()}))
+                                } yield '{Typ.TFun($t1,$t2)})
+                            )
+                        }
+                        
+                    )
+                )
+        )('{()})
+
+}
+
+// let genVar g t : expr option G.t =
+//   let vars = List.filter_mapi ~f:(fun i t' -> if Type_defn.equal t t' then Some (Some (Var i)) else None) g in
+//   match vars with
+//   | [] -> return None
+//   | _ -> of_list vars
+
+// let genConst t : expr G.t =
+//   recursive t @@ fun go t ->
+//     match t with
+//     | TBool -> map ~f:(fun b -> Bool b) bool
+//     | TFun(t1,t2) -> map ~f:(fun e -> Abs(t1,e)) (recurse go t2)
+
+// let genExactExpr n g t = recursive (n,g,t) @@ fun go (n,g,t) ->
+//   let%bind me = genVar g t in
+//   match me with
+//   | Some e -> return e
+//   | None -> if n <= 1 then genConst t else
+//             match t with
+//             | TFun (t1,t2) -> map ~f:(fun e -> Abs(t1,e)) (recurse go (n - 1,t1 :: g,t2))
+//             | _ -> let%bind t' = genTyp in
+//                    let%bind e1 = recurse go (n/2,g,TFun(t',t)) in
+//                    let%bind e2 = recurse go (n/2,g,t') in
+//                    return (App (e1,e2))
+
+// let genExpr =
+//   let%bind n = size in
+//   let%bind t = genTyp in
+//   genExactExpr n [] t
+
+// let quickcheck_generator = genExpr

### 2025-03-16 staged-scala/src/main/scala/stagedgen/StGen.scala

@@ -5,6 +5,7 @@ import scala.quoted.*
 import stagedgen.Cps
 import stagedgen.Splittable
 import stagedgen.SplittableCps.given
+import stagedgen.Splittable.given
 
 class IllegalBoundsError[A](low: A, high: A)
     extends IllegalArgumentException(s"invalid bounds: low=$low, high=$high")
@@ -89,6 +90,57 @@ object StGen {
         } yield (x,y)
     )
   }
+
+  // def letInsert[T : Type](ex : Expr[T])(using Quotes) : Cps[Expr[T]] = new Cps[Expr[T]] {
+  //   def apply[Z : Type](cont: Expr[T] => Expr[Z]): Expr[Z] =
+  //     '{
+  //       val x = ${ex}
+  //       ${cont('{x})}
+  //     }
+  // }
+
+
+  private def letInsert[T : Type](ex : Expr[T])(using Quotes) : StGen[Expr[T]] = {
+    gen((_,seed) => Cps.letInsert(ex).map(u => (u,seed)))
+  }
+
+
+  def pick[T : Type](gs : (Expr[Int],StGen[Expr[T]])*)(acc : Expr[Long])(using Quotes) : StGen[Expr[T]] = {
+    gs.toList match {
+      case Nil => StGen.gen((_,_) => Cps.error('{"Fell off the end of pick list"}))
+      case (x,g)::gs2 =>
+        '{$acc <= $x}.split.flatMap(leq =>
+          if(leq){
+            g
+          } else {
+            letInsert('{$acc - $x}).flatMap(acc2 =>
+              pick(gs2*)(acc2)
+            )
+          }
+        )
+    }
+  }
+
+  def frequency[T : Type](gs : (Expr[Int],StGen[Expr[T]])*)(using Quotes) : StGen[Expr[T]] = {
+    val sum = gs.map(_._1).foldRight('{0L})((x,y) => '{${x}.toLong + ${y}})
+    chooseLong('{0L},sum).flatMap(pick(gs*))
+  }
+
+  //  def frequency[T](gs: (Int, Gen[T])*): Gen[T] = {
+  //   val filtered = gs.iterator.filter(_._1 > 0).toVector
+  //   if (filtered.isEmpty) {
+  //     throw new IllegalArgumentException("no items with positive weights")
+  //   } else {
+  //     var total = 0L
+  //     val builder = TreeMap.newBuilder[Long, Gen[T]]
+  //     filtered.foreach { case (weight, value) =>
[…]

### 2025-03-16 staged-scala/src/main/scala/stagedgen/StGen.scala
freq

@@ -123,7 +123,7 @@ object StGen {
 
   def frequency[T : Type](gs : (Expr[Int],StGen[Expr[T]])*)(using Quotes) : StGen[Expr[T]] = {
     val sum = gs.map(_._1).foldRight('{0L})((x,y) => '{${x}.toLong + ${y}})
-    chooseLong('{0L},sum).flatMap(pick(gs*))
+    chooseLong('{1L},sum).flatMap(pick(gs*))
   }
 
   //  def frequency[T](gs: (Int, Gen[T])*): Gen[T] = {
@@ -206,17 +206,18 @@ object StGen {
   //       >.
   //   }
 
-  def complexStGenImpl (using q : Quotes): Expr[Int => Seed => (Long,Long)] = {
-    val e = StGen.splat(for {
-        x <- StGen.chooseLong('{1},'{1000})
-        y <- StGen.chooseLong('{0},x)
-    } yield '{(${x},${y})}
+  def wgImpl (using q : Quotes): Expr[Int => Seed => Int] = {
+    val e = StGen.splat(
+      StGen.frequency(
+        '{2} -> StGen.pure('{999}),
+        '{3} -> StGen.pure('{111})
+      )
     )
     e
   }
 
-  inline def complexStGen = {
-    ${complexStGenImpl}
+  inline def wg = {
+    ${wgImpl}
   }
 }
 

### 2025-03-17 staged-scala/src/main/scala/STLC/Bespoke.scala
more scalaing

@@ -20,9 +20,9 @@ object Bespoke {
             }
         )
 
-    def genVar(g : List[Typ],t : Typ) : Gen[Option[Expr]] = {
+    def genVar(g : List[Typ],t : Typ) : Gen[Option[Term]] = {
         val vars = g.zipWithIndex.flatMap { (t2, i) => 
-            if (t == t2) { Some(Some(Expr.Var(i))) } else { None }
+            if (t == t2) { Some(Some(Term.Var(i))) } else { None }
         }
         vars match {
             case Nil => Gen.const(None)
@@ -30,14 +30,14 @@ object Bespoke {
         }
     }
 
-    def genConst(t : Typ) : Gen[Expr] = {
+    def genConst(t : Typ) : Gen[Term] = {
         t match {
-            case Typ.TBool => Gen.oneOf(true,false).map(b => Expr.Bool(b))
-            case Typ.TFun(t1,t2) => genConst(t2).map(e => Expr.Abs(t1,e))
+            case Typ.TBool => Gen.oneOf(true,false).map(b => Term.Bool(b))
+            case Typ.TFun(t1,t2) => genConst(t2).map(e => Term.Abs(t1,e))
         }
     }
 
-    def genExactExpr(g : List[Typ], t : Typ) : Gen[Expr] = {
+    def genExactTerm(g : List[Typ], t : Typ) : Gen[Term] = {
         genVar(g,t).flatMap(me =>
             me match {
                 case Some(e) => Gen.const(e)
@@ -46,12 +46,12 @@ object Bespoke {
                         genConst(t)
                     } else {
                         t match {
-                            case Typ.TFun(t1, t2) => Gen.resize(n-1,genExactExpr(t1::g,t2)).map(e => Expr.Abs(t1,e))
+                            case Typ.TFun(t1, t2) => Gen.resize(n-1,genExactTerm(t1::g,t2)).map(e => Term.Abs(t1,e))
                             case _ => (for {
                                 t2 <- genTyp()
-                                e1 <- Gen.resize(n/2,genExactExpr(g,Typ.TFun(t2,t)))
-                                e2 <- Gen.resize(n/2,genExactExpr(g,t2))
-                            } yield (Expr.App(e1,e2)))
+                                e1 <- Gen.resize(n/2,genExactTerm(g,Typ.TFun(t2,t)))
+                                e2 <- Gen.resize(n/2,genExactTerm(g,t2))
[…]

### 2025-03-17 staged-scala/src/main/scala/STLC/Impl.scala

@@ -1,27 +1,37 @@
 package STLC
 
+import scala.quoted.*
+import stagedgen.Splittable
+import stagedgen.SplittableCps
+import stagedgen.Cps
+
 enum Typ:
     case TBool
     case TFun(dom : Typ, cod : Typ)
 
+given [A : Type](using Quotes) : SplittableCps[Typ,Unit | (Expr[Typ],Expr[Typ])] with
+    extension (a : Expr[Typ])
+        def splitCps(using Quotes) : Cps[Unit | (Expr[Typ],Expr[Typ])] = Cps.cps([Z : Type] => k => '{
+            ${a} match {
+                case Typ.TBool => ${k(())}
+                case Typ.TFun(dom,cod) => ${k(('{dom},'{cod}))}
+            }
+        })
 
-enum Expr:
+
+enum Term:
     case Var(idx : Long)
     case Bool(v : Boolean)
-    case Abs(t : Typ, body : Expr)
-    case App(fun : Expr, arg : Expr)
+    case Abs(t : Typ, body : Term)
+    case App(fun : Term, arg : Term)
 
-// let rec equal x y =
-//   match x, y with
-//   | TBool, TBool -> true
-//   | TFun(x1,x2), TFun(y1,y2) -> equal x1 y1 && equal x2 y2
-//   | _ -> false
-
-// // enum Bst:
-// //   case E
-// //   case Node(left : Bst, key : Long, value : Nat, right : Bst)
-
-// //   override def toString: String = this match
-// //     case E => "E"
-// //     case Node(left, key, value, right) => 
-// //       s"Node($left, $key, $value, $right)
+given [A : Type](using Quotes) : SplittableCps[Term, Expr[Long] | Expr[Boolean] | (Expr[Typ],Expr[Term]) | (Expr[Term],Expr[Term]) ] with
+    extension (a : Expr[Term])
+        def splitCps(using Quotes) : Cps[Expr[Long] | Expr[Boolean] | (Expr[Typ],Expr[Term]) | (Expr[Term],Expr[Term]) ] = Cps.cps([Z : Type] => k => '{
+            ${a} match {
+                case Term.Var(x) => ${k('{x})}
+                case Term.Bool(v) => ${k('{v})}
+                case Term.Abs(t,body) => ${k('{t},'{body})}
+                case Term.App(fun,arg) => ${k('{fun},'{arg})}
+            }
+        })

### 2025-03-17 staged-scala/src/main/scala/STLC/Staged.scala

@@ -6,9 +6,9 @@ import stagedgen.StGen
 import stagedgen.Splittable
 import stagedgen.Splittable.given
 import stagedgen.SplittableCps.given
+import STLC.* 
 import scala.quoted.*
 
-import STLC.* 
 
 object Staged {
     def genTBool(using Quotes) = StGen.pure('{Typ.TBool})
@@ -34,35 +34,64 @@ object Staged {
                 )
         )('{()})
 
+
+    def genVar(g : Expr[List[Typ]],t : Expr[Typ])(using Quotes) : StGen[Expr[Option[Term]]] = {
+        val vars : Expr[List[Option[Term]]] = '{$g.zipWithIndex.flatMap { (t2, i) => 
+            if ($t == t2) { Some(Some(Term.Var(i))) } else { None }
+        }}
+        vars.split.flatMap(vs =>
+            vs match {
+                case Left(_) => StGen.pure('{None})
+                case Right(_) => StGen.oneOfDyn(vars)
+            }
+        )
+    }
+
+    def genConst(t : Expr[Typ])(using Quotes) : StGen[Expr[Term]] = {
+        StGen.recursive[Term,Typ](rh => tc =>
+            tc.split.flatMap(t =>
+                t match {
+                    case _ => StGen.oneOf('{true},'{false}).map(b => '{Term.Bool($b)})
+                    case (t1,t2) => rh(t2).map(e => '{Term.Abs($t1,$e)})
+                }
+            )
+        )(t)
+        
+    }
+
+    def genExactTerm(g : Expr[List[Typ]], t : Expr[Typ])(using Quotes) : StGen[Expr[Term]] = {
+        StGen.recursive[Term,(List[Typ],Typ)](rh => gt => 
+            gt.split.flatMap((g,t) =>
+                genVar(g,t).flatMap(mec =>
+                    mec.split.flatMap(me =>
+                        me match {
+                            case Some(e) => StGen.pure(e)
+                            case None => StGen.size.flatMap(n =>
+                                '{$n <= 1}.split.flatMap(leq =>
+                                    if(leq){
+                                        genConst(t)
+                                    } else {
+                                       t.split.flatMap(tf =>
+                                            tf match {
+                                                case (t1,t2) => StGen.resize('{$n-1},rh('{($t1::$g,$t2)})).map(e => '{Term.Abs($t1,$e)})
+                                                case _ => (for {
+                                                    t2 <- genTyp
[…]

### 2025-03-17 staged-scala/src/main/scala/stagedgen/StGen.scala

@@ -91,14 +91,8 @@ object StGen {
     )
   }
 
-  // def letInsert[T : Type](ex : Expr[T])(using Quotes) : Cps[Expr[T]] = new Cps[Expr[T]] {
-  //   def apply[Z : Type](cont: Expr[T] => Expr[Z]): Expr[Z] =
-  //     '{
-  //       val x = ${ex}
-  //       ${cont('{x})}
-  //     }
-  // }
-
+  def chooseInt(lo : Expr[Int], hi : Expr[Int])(using Quotes) : StGen[Expr[Int]] =
+    chooseLong('{$lo.toLong},'{$hi.toLong}).map(l => '{$l.toInt})
 
   private def letInsert[T : Type](ex : Expr[T])(using Quotes) : StGen[Expr[T]] = {
     gen((_,seed) => Cps.letInsert(ex).map(u => (u,seed)))
@@ -126,21 +120,16 @@ object StGen {
     chooseLong('{1L},sum).flatMap(pick(gs*))
   }
 
-  //  def frequency[T](gs: (Int, Gen[T])*): Gen[T] = {
-  //   val filtered = gs.iterator.filter(_._1 > 0).toVector
-  //   if (filtered.isEmpty) {
-  //     throw new IllegalArgumentException("no items with positive weights")
-  //   } else {
-  //     var total = 0L
-  //     val builder = TreeMap.newBuilder[Long, Gen[T]]
-  //     filtered.foreach { case (weight, value) =>
-  //       total += weight
-  //       builder += ((total, value))
-  //     }
-  //     val tree = builder.result()
-  //     choose(1L, total).flatMap(r => tree.rangeFrom(r).head._2)
-  //   }
-  // }
+  def oneOf[T : Type](xs : T*)(using Quotes) : StGen[T] = {
+
+  }
+
+  def oneOfDyn[T : Type](xs : Expr[List[T]])(using Quotes) : StGen[Expr[T]] = {
+    for {
+      vec <- StGen.letInsert('{$xs.toVector})
+      i <- chooseInt('{0},'{$vec.size - 1})
+    } yield ('{$vec($i)})
+  }
 
   def sized[T](f: Expr[Int] => StGen[T]): StGen[T] =
     gen { (size, seed) => f(size).doApply(size, seed) }
@@ -182,29 +171,6 @@ object StGen {
       ).splitCps
     )
   }
-  //   type ('a,'r) recgen = 'r code -> 'a code t
-  // let recurse f x = {
[…]

### 2025-03-17 staged-scala/src/main/scala/STLC/Impl.scala
add oneOf, fix split for types

@@ -9,12 +9,16 @@ enum Typ:
     case TBool
     case TFun(dom : Typ, cod : Typ)
 
-given [A : Type](using Quotes) : SplittableCps[Typ,Unit | (Expr[Typ],Expr[Typ])] with
+enum TypS:
+    case TBoolS
+    case TFunS(dom : Expr[Typ], cod : Expr[Typ])
+
+given [A : Type](using Quotes) : SplittableCps[Typ,TypS] with
     extension (a : Expr[Typ])
-        def splitCps(using Quotes) : Cps[Unit | (Expr[Typ],Expr[Typ])] = Cps.cps([Z : Type] => k => '{
+        def splitCps(using Quotes) : Cps[TypS] = Cps.cps([Z : Type] => k => '{
             ${a} match {
-                case Typ.TBool => ${k(())}
-                case Typ.TFun(dom,cod) => ${k(('{dom},'{cod}))}
+                case Typ.TBool => ${k(TypS.TBoolS)}
+                case Typ.TFun(dom,cod) => ${k(TypS.TFunS('{dom},'{cod}))}
             }
         })
 

### 2025-03-17 staged-scala/src/main/scala/STLC/Staged.scala

@@ -51,8 +51,8 @@ object Staged {
         StGen.recursive[Term,Typ](rh => tc =>
             tc.split.flatMap(t =>
                 t match {
-                    case _ => StGen.oneOf('{true},'{false}).map(b => '{Term.Bool($b)})
-                    case (t1,t2) => rh(t2).map(e => '{Term.Abs($t1,$e)})
+                    case TypS.TFunS(t1,t2) => rh(t2).map(e => '{Term.Abs($t1,$e)})
+                    case TypS.TBoolS => StGen.oneOf('{true},'{false}).map(b => '{Term.Bool($b)})
                 }
             )
         )(t)
@@ -73,8 +73,8 @@ object Staged {
                                     } else {
                                        t.split.flatMap(tf =>
                                             tf match {
-                                                case (t1,t2) => StGen.resize('{$n-1},rh('{($t1::$g,$t2)})).map(e => '{Term.Abs($t1,$e)})
-                                                case _ => (for {
+                                                case TypS.TFunS(t1 : Expr[Typ],t2 : Expr[Typ]) => StGen.resize('{$n-1},rh('{($t1::$g,$t2)})).map(e => '{Term.Abs($t1,$e)})
+                                                case TypS.TBoolS => (for {
                                                     t2 <- genTyp
                                                     e1 <- StGen.resize('{$n/2},rh('{($g,Typ.TFun($t2,$t))}))
                                                     e2 <- StGen.resize('{$n/2},rh('{($g,$t2)}))

### 2025-03-17 staged-scala/src/main/scala/stagedgen/StGen.scala

@@ -121,7 +121,26 @@ object StGen {
   }
 
   def oneOf[T : Type](xs : T*)(using Quotes) : StGen[T] = {
+    val ys = xs.toList
+    val n = ys.length
+    def go(ys : List[T], i : Expr[Int]) : Cps[T] = {
+      ys match
+        case Nil => Cps.error('{"impossible"})
+        case x :: zs =>
+          '{$i == 0}.splitCps.flatMap(eqz =>
+            if(eqz) {
+              Cps.pure(x)
+            } else {
+              Cps.letInsert('{$i - 1}).flatMap(pred =>
+                go(zs,pred)
+              )
+            }
+          )
+    }
 
+    chooseInt('{0},'{${Expr(n)} - 1}).flatMap(i =>
+      gen((_,seed) => go(ys,i).map((_,seed)))
+    )
   }
 
   def oneOfDyn[T : Type](xs : Expr[List[T]])(using Quotes) : StGen[Expr[T]] = {

### 2025-03-17 staged-scala/src/main/scala/BST/BespokeLimits.scala
typederived bst

@@ -1,3 +1,0 @@
-package BST
-
-val bst_bespoke_limits = 10

### 2025-03-17 staged-scala/src/main/scala/BST/Limits.scala

@@ -1,0 +1,4 @@
+package BST
+
+val bst_bespoke_limits = 10
+val bst_type_limits = 10

### 2025-03-17 staged-scala/src/main/scala/BST/TypeDerived.scala

@@ -2,3 +2,31 @@ package BST
 
 import org.scalacheck.Gen
 
+import org.scalacheck.Gen
+import org.scalacheck.Arbitrary
+
+import Bst.*
+
+object TypeDerived {
+    def go() : Gen[Bst] = {
+        Gen.size.flatMap (sz =>
+            if(sz == 0){
+                Gen.const(Bst.E)
+            } else {
+                Gen.resize(sz - 1,
+                    Gen.frequency(
+                        1 -> Gen.const(Bst.E),
+                        1 -> (for {
+                            k <- Gen.Choose.chooseLong.choose(0, bst_type_limits)
+                            v <- Nat.generator(bst_type_limits)
+                            l <- go()
+                            r <- go()
+                        } yield Bst.Node(l,k,v,r))
+                    )
+                )
+            }
+        )
+    }
+
+    val gen = go()
+}

### 2025-03-17 staged-scala/src/main/scala/BST/TypeDerivedStaged.scala

@@ -1,0 +1,58 @@
+package BST
+
+import Bst.*
+import stagedgen.StGen
+import stagedgen.Splittable
+import stagedgen.Splittable.given
+import stagedgen.SplittableCps.given
+import scala.quoted.*
+
+object TypeDerivedStaged {
+    def go(using Quotes) : StGen[Expr[Bst]] = {
+        StGen.recursive[Bst,Unit](
+            rh => _ =>
+            StGen.size.flatMap (sz =>
+                '{$sz == 0}.split.flatMap(eqz =>
+                    if(eqz){
+                        StGen.pure('{Bst.E})
+                    } else {
+                        StGen.resize('{$sz - 1},
+                            StGen.frequency(
+                                '{1} -> StGen.pure('{Bst.E}),
+                                '{1} -> (for {
+                                    k <- StGen.chooseLong('{0}, '{bst_type_limits})
+                                    v <- Nat.staged_generator('{bst_type_limits})
+                                    l <- rh('{()})
+                                    r <- rh('{()})
+                                } yield ('{Bst.Node($l,$k,$v,$r)}))
+                            )
+                        )
+                    }
+                )
+            )
+        )('{()})
+    }
+
+    def genStaged (using Quotes) = StGen.splat(go)
+
+    inline def gen = {
+        ${genStaged}
+    }
+
+}
+
+// if(sz == 0){
+//                 Gen.const(Bst.E)
+//             } else {
+//                 Gen.resize(sz - 1,
+//                     Gen.frequency(
+//                         1 -> Bst.E,
+//                         1 -> (for {
+//                             k <- Gen.Choose.chooseLong.choose(0, bst_bespoke_limits)
+//                             v <- Nat.generator(bst_bespoke_limits)
+//                             l <- go()
+//                             r <- go()
+//                         } yield Bst.Node(l,k,v,r))
+//                     )
+//                 )
+//             }

### 2025-03-17 staged-scala/src/main/scala/Main.scala

@@ -14,8 +14,8 @@ val wg2 = Gen.frequency(
 )
 
 @main def hello(): Unit =
-  val seed = Seed.apply(10)
-  val Some(a) = wg2.apply(Gen.Parameters.default,seed)
-  val b = StGen.wg(Gen.Parameters.default.size)(seed)
+  val seed = Seed.apply(11)
+  val Some(a) = STLC.Bespoke.gen.apply(Gen.Parameters.default,seed)
+  val b = STLC.BespokeStaged.gen(Gen.Parameters.default.size)(seed)
   println(s"Unsaged: ${a.toString}")
   println(s"Staged: ${b.toString}")

### 2025-03-17 staged-scala/src/main/scala/STLC/BespokeStaged.scala

@@ -1,7 +1,5 @@
 package STLC
 
-package STLC
-
 import stagedgen.StGen
 import stagedgen.Splittable
 import stagedgen.Splittable.given
@@ -9,8 +7,7 @@ import stagedgen.SplittableCps.given
 import STLC.* 
 import scala.quoted.*
 
-
-object Staged {
+object BespokeStaged {
     def genTBool(using Quotes) = StGen.pure('{Typ.TBool})
 
     def genTyp(using Quotes) : StGen[Expr[Typ]] =
@@ -92,6 +89,16 @@ object Staged {
             )
 
         )('{($g,$t)})
+    }
+
+    def genStaged (using Quotes) = StGen.splat (
+        genTyp.flatMap(t =>
+            genExactTerm('{Nil},t)
+        )
+    )
+
+    inline def gen = {
+        ${genStaged}
     }
 
 }

### 2025-03-17 staged-scala/src/main/scala/STLC/Impl.scala

@@ -29,13 +29,13 @@ enum Term:
     case Abs(t : Typ, body : Term)
     case App(fun : Term, arg : Term)
 
-given [A : Type](using Quotes) : SplittableCps[Term, Expr[Long] | Expr[Boolean] | (Expr[Typ],Expr[Term]) | (Expr[Term],Expr[Term]) ] with
-    extension (a : Expr[Term])
-        def splitCps(using Quotes) : Cps[Expr[Long] | Expr[Boolean] | (Expr[Typ],Expr[Term]) | (Expr[Term],Expr[Term]) ] = Cps.cps([Z : Type] => k => '{
-            ${a} match {
-                case Term.Var(x) => ${k('{x})}
-                case Term.Bool(v) => ${k('{v})}
-                case Term.Abs(t,body) => ${k('{t},'{body})}
-                case Term.App(fun,arg) => ${k('{fun},'{arg})}
-            }
-        })
+// given [A : Type](using Quotes) : SplittableCps[Term, Expr[Long] | Expr[Boolean] | (Expr[Typ],Expr[Term]) | (Expr[Term],Expr[Term]) ] with
+//     extension (a : Expr[Term])
+//         def splitCps(using Quotes) : Cps[Expr[Long] | Expr[Boolean] | (Expr[Typ],Expr[Term]) | (Expr[Term],Expr[Term]) ] = Cps.cps([Z : Type] => k => '{
+//             ${a} match {
+//                 case Term.Var(x) => ${k('{x})}
+//                 case Term.Bool(v) => ${k('{v})}
+//                 case Term.Abs(t,body) => ${k('{t},'{body})}
+//                 case Term.App(fun,arg) => ${k('{fun},'{arg})}
+//             }
+//         })

### 2025-03-17 staged-scala/src/main/scala/benchmark/GenBm.scala

@@ -10,6 +10,7 @@ import stagedgen.StGen
 import BST.Bst
 import BST.SingleBespoke
 import BST.SingleBespokeStaged
+import STLC.*
 
 def complexGen: Gen[(Long,Long)] = for {
     x <- Gen.choose(1,1000)
@@ -114,8 +115,8 @@ class GenBm {
 @Benchmark
 @BenchmarkMode(Array(Mode.AverageTime))
 @OutputTimeUnit(TimeUnit.NANOSECONDS)
-@Warmup(iterations = 5, time = 5, timeUnit = TimeUnit.SECONDS)
-@Measurement(iterations = 3, time = 5, timeUnit = TimeUnit.SECONDS)
+@Warmup(iterations = 3, time = 5, timeUnit = TimeUnit.SECONDS)
+@Measurement(iterations = 5, time = 5, timeUnit = TimeUnit.SECONDS)
 @Fork(1)
   def generateBst(): Option[Bst] = {
     SingleBespoke.gen.sample
@@ -129,5 +130,25 @@ class GenBm {
 @Fork(1)
   def generateBstStaged(): Bst = {
     SingleBespokeStaged.gen(Gen.Parameters.default.size)(Seed.random())
+  }
+
+@Benchmark
+@BenchmarkMode(Array(Mode.AverageTime))
+@OutputTimeUnit(TimeUnit.NANOSECONDS)
+@Warmup(iterations = 3, time = 5, timeUnit = TimeUnit.SECONDS)
+@Measurement(iterations = 5, time = 5, timeUnit = TimeUnit.SECONDS)
+@Fork(1)
+  def generateTerm(): Option[Term] = {
+    STLC.Bespoke.gen.sample
+  }
+
+@Benchmark
+@BenchmarkMode(Array(Mode.AverageTime))
+@OutputTimeUnit(TimeUnit.NANOSECONDS)
+@Warmup(iterations = 3, time = 5, timeUnit = TimeUnit.SECONDS)
+@Measurement(iterations = 5, time = 5, timeUnit = TimeUnit.SECONDS)
+@Fork(1)
+  def generateTermStaged(): Term = {
+    STLC.BespokeStaged.gen(Gen.Parameters.default.size)(Seed.random())
   }
 }

### 2025-03-17 paper/paper.tex
wibbleg

@@
 
-In Muli-stage programming (also known more simply as staging), programs execute over the course of multiple stages,
-with each stage producing the code to be run in the next.
-This is an old idea, with roots going back to quasiquotation in LISP \cn{}, and picking up steam again in the 1990s with MetaML \cn{}.
-For the purposes of this paper, we will primarily consider two stages: compile time and run time. Staged programs thus execute twice: once at compile time,
-which produces more code, which is then compiled and run at run time.
+In Muli-stage programming (also known more simply as staging), programs execute
+over the course of multiple stages, with each stage producing the code to be run
+in the next.  This is an old idea in programming languages, with roots going
+back to quasiquotation in LISP \cn{}.  For the purposes of this paper, we will
+primarily consider two stages: compile time and run time. Staged programs thus
+execute twice: once at compile time, which produces more code, which is then
+compiled and run at run time.
 
@@
 the functional abstractions used to build eDSLs prevent compilers from generating efficient machine code \cn.
-In the literature, this effect is known as ``abstraction overhead'': the layers of abstraction that 
+In the literature, this effect is known as ``abstraction overhead'' \cite{carrette05, moller20}: the layers of abstraction that 
 make eDSLs so usable are precisely what prevents them from being fast.
@@
 \jwc{This previous paragraph is peak ``shitty first draft''.}
-This recipe has been used to great effect across domains to stage eDSLs for stream processing \cn, parser combinators \cite{sspc, staged-parsers, krishnaswami19, flap},
+This recipe has been used to great effect across domains to stage eDSLs for stream processing \cite{moller20,strymonas}, parser combinators \cite{sspc, staged-parsers, krishnaswami19, flap},
 and query processing \cite{rhyme}.
@@
 Many languages have some degree of staging functionality, though support varies
-from full native support (Scala 3 \cn, Haskell \cn) to experimental compiler extensions (OCaml \cite{metaocaml,macocaml}, Java \cite{mint}, Kotlin \cn).
+from full native support (Scala 3 \cite{scalamacros}, Haskell \cite{templatehaskell}) to compiler extensions (OCaml \cite{metaocaml,macocaml}, Java \cite{mint}).
 Moreover, many more languages have lighter-weight staging-like metaprogramming support in the form of macros.
@@
 In Section~\ref{section:motiv}, we noted that generator combinators like \texttt{weighted\_union}
-must allocate lists in the hot path of the generator. Even though these lists are often small ---
-usually at most a few dozen elements in practice --- each allocation takes us closer to the next garbage collection.
-
-This is an ideal opportunity to exercise another use of staging: compile-time specialization.
+often allocate lists in the hot path of the generator. Even though these lists are usually small ---
+at most a few dozen elements in practice --- each allocation takes us closer to the next garbage collection.
+
+This is an ideal opportunity to exercise another feature of staging: compile-time specialization.
 Since we almost always know the particular list of choices at compile time, a staged version of \texttt{weighted\_union}

### 2025-03-17 staged-scala/src/main/scala/BST/Impl.scala
ok, done with scala generators

@@ -1,5 +1,6 @@
 package BST
 
+import Nat.*
 
 enum Bst:
   case E

### 2025-03-17 staged-scala/src/main/scala/BST/SingleBespoke.scala

@@ -3,6 +3,7 @@ import org.scalacheck.Gen
 import org.scalacheck.Arbitrary
 
 import Bst.*
+import Nat.*
 
 object SingleBespoke {
   def go (lo : Long, hi : Long): Gen[Bst] = {

### 2025-03-17 staged-scala/src/main/scala/BST/SingleBespokeStaged.scala

@@ -5,6 +5,7 @@ import stagedgen.Splittable
 import stagedgen.Splittable.given
 import stagedgen.SplittableCps.given
 import scala.quoted.*
+import Nat.*
 
 object SingleBespokeStaged {
     def go(using Quotes)(lo : Expr[Long], hi : Expr[Long]) : StGen[Expr[Bst]] = {

### 2025-03-17 staged-scala/src/main/scala/Nat.scala

@@ -1,4 +1,4 @@
-package BST
+package Nat
 
 import org.scalacheck.Gen
 import org.scalacheck.Arbitrary

### 2025-03-17 staged-scala/src/main/scala/RBT/Impl.scala

@@ -1,0 +1,11 @@
+package RBT
+
+import Nat.*
+
+enum Color:
+  case Red
+  case Black
+
+enum Rbt:
+  case E
+  case Node(color : Color, left : Rbt, key : Long, value : Nat, right : Rbt)

### 2025-03-17 staged-scala/src/main/scala/RBT/Limits.scala

@@ -1,0 +1,3 @@
+package RBT
+
+val rbt_type_limits = 10

### 2025-03-17 staged-scala/src/main/scala/RBT/TypeDerived.scala

@@ -1,28 +1,29 @@
-package BST
+package RBT
 
 import org.scalacheck.Gen
 
 import org.scalacheck.Gen
 import org.scalacheck.Arbitrary
+import Rbt.*
+import Color.*
 import Nat.*
 
-import Bst.*
-
 object TypeDerived {
-    def go() : Gen[Bst] = {
+    def go() : Gen[Rbt] = {
         Gen.size.flatMap (sz =>
             if(sz == 0){
-                Gen.const(Bst.E)
+                Gen.const(Rbt.E)
             } else {
                 Gen.resize(sz - 1,
                     Gen.frequency(
-                        1 -> Gen.const(Bst.E),
+                        1 -> Gen.const(Rbt.E),
                         1 -> (for {
-                            k <- Gen.Choose.chooseLong.choose(0, bst_type_limits)
-                            v <- Nat.generator(bst_type_limits)
+                            c <- Gen.oneOf(Red,Black)
+                            k <- Gen.Choose.chooseLong.choose(0, rbt_type_limits)
+                            v <- Nat.generator(rbt_type_limits)
                             l <- go()
                             r <- go()
-                        } yield Bst.Node(l,k,v,r))
+                        } yield Rbt.Node(c,l,k,v,r))
                     )
                 )
             }

### 2025-03-17 staged-scala/src/main/scala/RBT/TypeDerivedStaged.scala

@@ -1,31 +1,33 @@
-package BST
+package RBT
 
-import Bst.*
+import Rbt.*
+import Color.*
+import Nat.*
 import stagedgen.StGen
 import stagedgen.Splittable
 import stagedgen.Splittable.given
 import stagedgen.SplittableCps.given
 import scala.quoted.*
-import Nat.*
 
 object TypeDerivedStaged {
-    def go(using Quotes) : StGen[Expr[Bst]] = {
-        StGen.recursive[Bst,Unit](
+    def go(using Quotes) : StGen[Expr[Rbt]] = {
+        StGen.recursive[Rbt,Unit](
             rh => _ =>
             StGen.size.flatMap (sz =>
                 '{$sz == 0}.split.flatMap(eqz =>
                     if(eqz){
-                        StGen.pure('{Bst.E})
+                        StGen.pure('{Rbt.E})
                     } else {
                         StGen.resize('{$sz - 1},
                             StGen.frequency(
-                                '{1} -> StGen.pure('{Bst.E}),
+                                '{1} -> StGen.pure('{Rbt.E}),
                                 '{1} -> (for {
-                                    k <- StGen.chooseLong('{0}, '{bst_type_limits})
-                                    v <- Nat.staged_generator('{bst_type_limits})
+                                    c <- StGen.oneOf('{Red},'{Black})
+                                    k <- StGen.chooseLong('{0}, '{rbt_type_limits})
+                                    v <- Nat.staged_generator('{rbt_type_limits})
                                     l <- rh('{()})
                                     r <- rh('{()})
-                                } yield ('{Bst.Node($l,$k,$v,$r)}))
+                                } yield ('{Rbt.Node($c,$l,$k,$v,$r)}))
                             )
                         )
                     }

### 2025-03-17 staged-scala/src/main/scala/benchmark/GenBm.scala

@@ -8,8 +8,7 @@ import org.scalacheck.Gen
 import org.scalacheck.rng.*
 import stagedgen.StGen
 import BST.Bst
-import BST.SingleBespoke
-import BST.SingleBespokeStaged
+import RBT.Rbt
 import STLC.*
 
 def complexGen: Gen[(Long,Long)] = for {
@@ -118,8 +117,8 @@ class GenBm {
 @Warmup(iterations = 3, time = 5, timeUnit = TimeUnit.SECONDS)
 @Measurement(iterations = 5, time = 5, timeUnit = TimeUnit.SECONDS)
 @Fork(1)
-  def generateBst(): Option[Bst] = {
-    SingleBespoke.gen.sample
+  def generateBstType(): Option[Bst] = {
+    BST.TypeDerived.gen.sample
   }
 
 @Benchmark
@@ -128,8 +127,48 @@ class GenBm {
 @Warmup(iterations = 3, time = 5, timeUnit = TimeUnit.SECONDS)
 @Measurement(iterations = 5, time = 5, timeUnit = TimeUnit.SECONDS)
 @Fork(1)
-  def generateBstStaged(): Bst = {
-    SingleBespokeStaged.gen(Gen.Parameters.default.size)(Seed.random())
+  def generateBstTypeStaged(): Bst = {
+    BST.TypeDerivedStaged.gen(Gen.Parameters.default.size)(Seed.random())
+  }
+
+@Benchmark
+@BenchmarkMode(Array(Mode.AverageTime))
+@OutputTimeUnit(TimeUnit.NANOSECONDS)
+@Warmup(iterations = 3, time = 5, timeUnit = TimeUnit.SECONDS)
+@Measurement(iterations = 5, time = 5, timeUnit = TimeUnit.SECONDS)
+@Fork(1)
+  def generateBstBespoke(): Option[Bst] = {
+    BST.SingleBespoke.gen.sample
+  }
+
+@Benchmark
+@BenchmarkMode(Array(Mode.AverageTime))
+@OutputTimeUnit(TimeUnit.NANOSECONDS)
+@Warmup(iterations = 3, time = 5, timeUnit = TimeUnit.SECONDS)
+@Measurement(iterations = 5, time = 5, timeUnit = TimeUnit.SECONDS)
+@Fork(1)
+  def generateBstBespokeStaged(): Bst = {
[…]

### 2025-03-17 paper/paper.tex
paper blah

@@
 \label{section:motiv}
-\jwc{Why is a monadic generator library slow?}
-\jwc{NOTE: We can simplify this further by getting rid of the size parameter. Make the story even cleaner, I think!}
+\jwc{In this section we answer: ``Why are monadic generator libraries slow?''}
 
@@
 \subsection{Monadic DSL Abstraction Overhead}
-
-\jwc{
-  \begin{itemize}
-    \item It's been long-known that clean functional abstractions have a runtime overhead (this should be familiar by this time in the paper).
-    \item How does (simplified) BQ work?
-    \begin{itemize}
-      \item The basic generator type: \texttt{'a generator = int -> SR.t -> 'a}. Size and random seed to deterministic value. (note: \texttt{SR.t} is a mutable seed)
-      \item This gets a monad intance in the obvious way (show code).
-      \item Also show the code for \texttt{int}, how it calls the underlying SR function.
-    \end{itemize}
-    % \item Note that \emph{extensionally} \texttt{generate (create (fun size random -> e)) size random = e}, but the OCaml compiler does not always perform this optimization, or do the inlining required to expose it.
-    % (When sufficiently obfuscated behind returns and binds ...) This program compiles to code that (1) allocates the closure for `e', (2) passes it to create (which returns the closure), and then calls (3) generate, which immediately jumps into the closure.
-    \item Show benchmarks of the running example, versus the version where you inline everything. ()
-    \item Let's look at the running example: inline it all the way.
-  \end{itemize}
-}
-
+Just how large is the abstraction overhead of monadic generator DSLs, and where does it come from?
+Figure~\ref{fig:bq-internals} shows the internals of (a simplified version of) \bq{}.
+A generator \texttt{'a Bq.t} is just a function of type~\texttt{int -> SR.t -> 'a},
+taking an \texttt{int} representing the current size parameter and a random seed \texttt{SR.t},
+and returning a generated value \texttt{'a}. It is an invariant of the library that function of this type are
+deterministic: for a fixed size and seed, it will return the same value. All of the randomness in testing comes from varying the initial seed.
+The monad functions \texttt{return} and \texttt{bind} are defined in the usual way
+for an instance of the reader monad \cn{}. \texttt{Return x} ignores the size and seed and just returns
+\texttt{x}, while \texttt{bind g k} first runs \texttt{g} and then passes the generated value to the continuation \texttt{k}.
+\bq{} uses a mutable seed, so the \texttt{random} passed to \texttt{(k a)} will be different from the one passed to \texttt{g}.
+The \texttt{gen\_int} combinator simply calls out to the randomness library \texttt{Splittable\_random}, aliased as \texttt{SR} here.
+\jwc{Should we do this?}
+
+Different PBT libraries use variations on this basic design. Haskell's QuickCheck uses a reader monad with an immutable state
+that is ``split'' at \texttt{bind}s \cn. Meanwhile, ScalaCheck (a) uses an immutable state type and state monad \cn{} to thread
+the state through, and (b) has an \texttt{Option} as its return type, to allow generation to fail.
+\jwc{However, the principles are the same, and we will use \bq{} to illustrate.}
+
+\begin{figure}
+\begin{lstlisting}
+module Bq = struct
+  type 'a t = int -> SR.t -> 'a
+
+  let return (x : 'a) : 'a t = fun _ _ -> x
+
+  let bind (g : 'a t) (k : 'a -> 'b t) : 'b t =
+    fun size random ->
+      let a = g size random in
+      (k a) size random
+
+  let gen_int (lo : int) (hi : int) : int t =
+    fun _ random -> SR.int random lo hi
+end
+\end{lstlisting}
+\caption{Internals of a Monadic Generator eDSL}
+\label{fig:bq-internals}
+\end{figure}
+
+Just how much runtime overhead does this monadic abstraction introduce? To
+illustrate, let's return to our running example of a constrained pair of
+integers, written in both \bq{} and ScalaCheck.
+Figure~\ref{fig:overhead-explanation-code} shows two versions of the generator, written in both languages.
+The first version (\texttt{int\_pair} in \bq{} and \texttt{intPair} in ScalaCheck) is written with monadic generator combinators.
+\jwc{Maybe we can avoid rewriting these if we already showed them earlier? Also maybe we duplicate the tutorial discussion of monadic generator libraries to include Scala?}
+The second version (\texttt{int\_pair\_inlined} and \texttt{intPairInlined})
+is just the first version, with (a) all generator combinator definitions inlined, and (b)
+beta-redexes reduced.
+
+\begin{figure}
 \begin{lstlisting}
@@
 
-let int_pair_inlined : (int * int) Bq.t
-  fun sr ->
+let int_pair_inlined : int -> SR.t -> int * int =
+  fun _ sr ->
     let x = Splittable_random.int sr ~lo:0 ~hi:100 in
@@
     (x,y)
-\end{lstlisting}
-
-\jwc{Show benchmark difference between these two generators: the benchmark is in \texttt{waffle-house/handwritten-ocaml/bin/basic-compare.ml}. It's about 2x. (see comment in latex here.)}
-
-% ┌───────┬──────────┬─────────┬────────────┐
-% │ Name  │ Time/Run │ mWd/Run │ Percentage │
-% ├───────┼──────────┼─────────┼────────────┤
-% │ bq    │  70.47ns │  78.00w │    100.00% │
-% │ fused │  35.86ns │  78.00w │     50.89% │
-% └───────┴──────────┴─────────┴────────────┘
-
-\jwc{The overhead of the abstraction is 2x.
-The reality is actually worse: actual base-quickcheck includes even more indirection in its type.}
-
-\jwc{The native code OCaml compiler, even with -O3, fails to specialize this code and eliminate the overhead of this abstraction --- inlining all of the funciton definitions and then performing beta-reductions speeds up sampling by a factor of 2.}
-
-\begin{lstlisting}
+
 def intPair : Gen[(Long,Long)] = for {
@@
 }
-\end{lstlisting}
-
-\jwc{Same story in scala. 
-Approximately 2x difference for the same generator (though an order of magnitude slower overall than bq)... (scala uses a slightly different generator type, but the principles are the same)
-This doesn't even account for more inlining we could do to eliminate the boxing/unboxing of results. (Scalacheck generators can fail, though this adds significant overhead)
-}
+\caption{Int Pair Generators in \bq{} and ScalaCheck}
+\label{fig:overhead-explanation-code}
+\end{figure}
+
+The performance difference the two versions of this generator (Figure~\ref{fig:overhead-explanation-perf}) is stark. In both
+languages, the inlined version takes \textbf{half} as much time to generate a
+single pair of \texttt{int}s.
+
+\begin{figure}
+\end{lstlisting}
+  \begin{tabular}{lll}
+  Library    & Generator          & Time per Generation (ns) \\
+  \bq{}         & \texttt{int\_pair}          & 70                       \\
+  \bq{}         & \texttt{int\_pair\_inlined} & 35                       \\
+  ScalaCheck & \texttt{intPair}            & 458                      \\
+  ScalaCheck & \texttt{intPairInlined}     & 266                     
+  \end{tabular}
+\label{fig:overhead-explanation-perf}
+\end{figure}
+
+% ┌───────┬──────────┬─────────┬────────────┐
+% │ Name  │ Time/Run │ mWd/Run │ Percentage │
+% ├───────┼──────────┼─────────┼────────────┤
+% │ bq    │  70.47ns │  78.00w │    100.00% │
+% │ fused │  35.86ns │  78.00w │     50.89% │
+% └───────┴──────────┴─────────┴────────────┘
+
+% \jwc{The overhead of the abstraction is 2x. The reality is actually worse: actual base-quickcheck includes even more indirection in its type.}
+
+% \jwc{The native code OCaml compiler, even with -O3, fails to specialize this code and eliminate the overhead of this abstraction --- inlining all of the funciton definitions and then performing beta-reductions speeds up sampling by a factor of 2.}
+
+% \jwc{Same story in scala. 
+% Approximately 2x difference for the same generator (though an order of magnitude slower overall than bq)... (scala uses a slightly different generator type, but the principles are the same)
+% This doesn't even account for more inlining we could do to eliminate the boxing/unboxing of results. (Scalacheck generators can fail, though this adds significant overhead)
+% }
 

### 2025-03-17 paper/paper.tex
oops
some eval

@@
 }
+\end{lstlisting}
 \caption{Int Pair Generators in \bq{} and ScalaCheck}
@@
 \begin{figure}
-\end{lstlisting}
   \begin{tabular}{lll}
@@
 \label{section:eval}
+\begin{itemize}
+  \item RQ1: Do generators written using \name run faster than those written using base quickcheck?
+  \item RQ2: Do performance speedups translate to better bug-finding ability?
+  \item RQ3: Does our technique generalize to other strict functional languages?
+\end{itemize}
+
+All experiments were run on a 64-bit Linux machine with 264 GB RAM and an Intel Xeon Platinum 8375C CPU, running Ubuntu 24.04.1 LTS.
+The \name library was implemented using the
+4.14.1+BER MetaOCaml compiler variant.  All OCaml benchmarks
+were run invoking \texttt{ocamlopt}, the native code compiler for OCaml. \todo{Joe : Scala info ...}
+
 \subsection{Implementing and Testing Generators}
@@
 \jwc{Talk about the difftesting to ensure staging maintains semantics, and also different RNGs have identical semantics.}
-
 \subsection{Benchmarking speed \& resource usage}
@@
 \subsection{Impact on bug-finding ability}
+We evaluate the bug-finding ability of three of our case studies---BST, RBT, and STLC---using Etna, a platform for assessing and comparing property-based testing (PBT) techniques. 
+Etna allows users to measure the effectiveness of different generator implementations by injecting bugs into the system under test and recording the time until a relevant 
+property fails in response. For example, in a tree with a property ensuring that the tree remains ordered, a bug might involve inserting a value that violates that ordering. 
+A \textit{task} consists of a bug-property pair, and each case study is associated with a diverse set of tasks that exercise different parts
+of the system. In particular, BST has 30 tasks, RBT has 46 tasks, and STLC has 18 tasks.
+
+For each case study, we construct both type-derived and custom generators.
+Each generator has four implementations, which are equivalent up to program equality:
+\begin {itemize}
+  \item The original \bq generator (``BQ'').
+  \item A staged version of the generator (``BQ++'').
+  \item A staged version of the generator using the C \rand (``BQC++'').
+  \item A staged version incorporating the CSR \rand (``BQCSR++'').
+\end{itemize}
+
+To ensure a fair comparison, we initialize each of these generators with the same random seed, ensuring that they produce identical trees.
+
+\subsubsection{Variance between seeds}
+The time-to-failure of a given strategy on a given seed is not necessarily representative of 
+the strategy's average time-to-fail over a large number of trials, but one might expect the speedup to be. This is true in most cases, but not all. For example, it is theoretically possible
+to choose a seed such that the \textit{first} value produced by a generator discovers the bug, which would elide any differential speedup between staged and unstaged generators.
+To account for these edge cases, we repeated the above process for 10 seeds and computed the geometric average of their speedups.
+
+% To produce \todo{FIG}, we repeated the above process for 10 seeds and computed the geometric average of their speedups. 
+%This is important because the time-to-failure of a given strategy on a given seed is not necessarily 
+%representative of the strategy's overall bug-finding ability, although it often is. For example, it is theoretically possible 
+%to choose a seed such that the \textit{first} value produced by a generator discovers the bug, but it would be misleading to report this result.
+
+\subsubsection{Variance between trials}
+With a fixed random seed, a given strategy in Etna consistently finds a bug in the same amount of time---across 1,000 trials on the same seed, 
+the observed variance in time-to-failure was less than a nanosecond.
+
 \jwc{Staging type-derived generators (which are the most commonly used ones) can turn a timeout into a found bug!}
@@
 \jwc{A generator is a parser of randomness, and people have implemented lots of staged parser libraries!}
+
 %%

### 2025-03-17 paper/paper.tex
pbt for pbt

@@
 \label{subsection:pbt-for-pbt}
-\jwc{Talk about the difftesting to ensure staging maintains semantics, and also different RNGs have identical semantics.}
+To establish the equivalence of generators up to program equality, we use a PBT harness to compare their outputs. 
+Specifically, we generate random (size, seed) pairs and evaluate whether the generators under test produce equal results. 
+If the equality property fails, we report the unequal trees; if, over 1,000 trials, no ``counterexample'' is reported, we consider the
+behavior of the generators to be equivalent.
 \subsection{Benchmarking speed \& resource usage}
@@
 property fails in response. For example, in a tree with a property ensuring that the tree remains ordered, a bug might involve inserting a value that violates that ordering. 
-A \textit{task} consists of a bug-property pair, and each case study is associated with a diverse set of tasks that exercise different parts
-of the system. In particular, BST has 30 tasks, RBT has 46 tasks, and STLC has 18 tasks.
+Each case study includes a diverse set of \textit{tasks}, where a task consists of a specific bug-property pair designed to test different aspects of the system.
 

### 2025-03-17 paper/paper.tex
wibbles

--- 2025-03-17 paper/paper.tex
@@
 Figure~\ref{fig:overhead-explanation-code} shows two versions of the generator, written in both languages.
-The first version (\texttt{int\_pair} in \bq{} and \texttt{intPair} in ScalaCheck) is written with monadic generator combinators.
+The first versions (\texttt{int\_pair} in \bq{} and \texttt{intPair} in ScalaCheck) are written with the monadic generator combinators from their respective libraries.
 \jwc{Maybe we can avoid rewriting these if we already showed them earlier? Also maybe we duplicate the tutorial discussion of monadic generator libraries to include Scala?}
-The second version (\texttt{int\_pair\_inlined} and \texttt{intPairInlined})
-is just the first version, with (a) all generator combinator definitions inlined, and (b)
-beta-redexes reduced.
+The second versions (\texttt{int\_pair\_inlined} and \texttt{intPairInlined})
+are semantically identical to the first, but have been rewritten
+by (1) inlining all generator combinator definitions, and then (2)
+repeatedly reducing simplifiable terms like \lstinline{(fun x -> e) e'} --- where an anonymous function is defined and then immediatly called~\footnote{This is usually known as a ``$\beta$-redex''.
+Note that when reducing, we do not reduce to \lstinline{e[e'/x]}, as this does not preserve the order of effects.
+}
+--- to \lstinline{let x = e' in e}.
 
@@
 
-The performance difference the two versions of this generator (Figure~\ref{fig:overhead-explanation-perf}) is stark. In both
-languages, the inlined version takes \textbf{half} as much time to generate a
-single pair of \texttt{int}s.
+The performance impact of this inling is large (Figure~\ref{fig:overhead-explanation-perf}). In both languages, the
+inlined version takes (on average) \textbf{half} as much time to generate a single pair of
+\texttt{int}s. 
+Microbenchmarks of more realistic generators (see Section~\ref{section:eval}
+\jwc{more precise ref}) show an even more dramatic performance boost for
+inlining in this manner.
+
+% Modern compilers use heuristics to decide when to specialize and inline code \cn, and these heuristics cannot guarantee that generator abstractions are zero-cost.
+% These heuristics are also necessarily conservative about inlining and specializing recursive functions, which all generators of recursive data types must be.
+% The
 
 \begin{figure}
-\end{lstlisting}
   \begin{tabular}{lll}
-  Library    & Generator          & Time per Generation (ns) \\
+  Library    & Generator          & Average Time per Generation (ns) \\
   \bq{}         & \texttt{int\_pair}          & 70                       \\
@@
   \end{tabular}
+\caption{Microbenchmarks of Generators in \bq{} (using \texttt{core\_bench} \cn) and ScalaCheck (using \texttt{jmh} \cn)}
 \label{fig:overhead-explanation-perf}
 \end{figure}
-
-% ┌───────┬──────────┬─────────┬────────────┐
-% │ Name  │ Time/Run │ mWd/Run │ Percentage │
-% ├───────┼──────────┼─────────┼────────────┤
-% │ bq    │  70.47ns │  78.00w │    100.00% │
-% │ fused │  35.86ns │  78.00w │     50.89% │
-% └───────┴──────────┴─────────┴────────────┘
 
@@
 
-\jwc{
-The problem is plain: while the monadic abstraction that PBT libraries like base quickcheck provide are indespensible for writing idiomatic generators, the performance overhead of using them is dramatic.
-Modern compilers use heuristics to decide when to specialize and inline code \cn, and these heuristics cannot guarantee that generator abstractions are zero-cost.
-These heuristics are also necessarily conservative about inlining and specializing recursive functions, which all generators of recursive data types must be.
-}
+% \jwc{
+% 
+% }
 
@@
 
-\subsection{Combinator Abstraction Overhead}
+\subsubsection{Overhead of Other Combinators}
 
@@
 
-This type, along with basic monadic generator DSL functionality can be found in Figure~\ref{fig:gen-staged-basic}.
-The monadic interface is given by a return and bind, as usual.
+This type, along with basic staged monadic generator DSL functionality can be found in Figure~\ref{fig:gen-staged-basic}.
 \texttt{Return} is the constant generator, but this time it 

### 2025-03-17 paper/paper.tex

--- 2025-03-17 paper/paper.tex
@@
 
-The performance difference the two versions of this generator (Figure~\ref{fig:overhead-explanation-perf}) is stark. In both
-languages, the inlined version takes \textbf{half} as much time to generate a
-single pair of \texttt{int}s.
+The performance impact of this inling is large (Figure~\ref{fig:overhead-explanation-perf}). In both languages, the
+inlined version takes (on average) \textbf{half} as much time to generate a single pair of
+\texttt{int}s. 
+Microbenchmarks of more realistic generators (see Section~\ref{section:eval}
+\jwc{more precise ref}) show an even more dramatic performance boost for
+inlining in this manner.
+
+% Modern compilers use heuristics to decide when to specialize and inline code \cn, and these heuristics cannot guarantee that generator abstractions are zero-cost.
+% These heuristics are also necessarily conservative about inlining and specializing recursive functions, which all generators of recursive data types must be.
+% The
 
@@
   \begin{tabular}{lll}
-  Library    & Generator          & Time per Generation (ns) \\
+  Library    & Generator          & Average Time per Generation (ns) \\
   \bq{}         & \texttt{int\_pair}          & 70                       \\

### 2025-03-17 paper/paper.tex
etna

--- 2025-03-17 paper/paper.tex
@@
 \label{section:eval}
+\begin{itemize}
+  \item \textbf{RQ1}: Do generators written using \name run faster than those written using base quickcheck?
+  \item \textbf{RQ2}: Do performance speedups translate to better bug-finding ability?
+  \item \textbf{RQ3}: Does our technique generalize to other strict functional languages?
+\end{itemize}
+
+All experiments were run on a 64-bit Linux machine with 264 GB RAM and an Intel Xeon Platinum 8375C CPU, running Ubuntu 24.04.1 LTS.
+The \name library was implemented using the
+4.14.1+BER MetaOCaml compiler variant.  All OCaml benchmarks
+were run invoking \texttt{ocamlopt}, the native code compiler for OCaml, using compiler flags \texttt{-flamda -o3}.
+
+. \todo{Joe : Scala info ...}
+
 \subsection{Implementing and Testing Generators}
 \label{subsection:pbt-for-pbt}
-\jwc{Talk about the difftesting to ensure staging maintains semantics, and also different RNGs have identical semantics.}
-
+To establish the equivalence of generators up to program equality, we use a PBT harness to compare their outputs. 
+Specifically, we generate random sizes and seeds and evaluate whether the generators under test produce equal results. 
+If the equality property fails, we report the unequal trees; if, over 1,000 trials, no ``counterexample'' is reported, we consider the
+behavior of the generators to be equivalent.
 \subsection{Benchmarking speed \& resource usage}
@@
 \subsection{Impact on bug-finding ability}
-\jwc{Staging type-derived generators (which are the most commonly used ones) can turn a timeout into a found bug!}
+To answer \textbf{RQ2}, we evaluate the bug-finding ability of three of our case studies---BST, RBT, and STLC---using Etna, a platform for assessing and comparing property-based testing (PBT) techniques.
+Etna allows users to measure the effectiveness of different generator implementations by injecting bugs into the system under test and recording the time taken for a relevant 
+property to fail in response. For example, in a tree with an ordering property, a bug injection might consist of inserting a value that violates that ordering.
+
+Each case study includes a diverse set of \textit{tasks}, where a task consists of a specific bug-property pair designed to test different aspects of the system (e.g., BST
+includes tasks that test insertion, deletion, and union operations on binary search trees).
+
+We evaluate each case study with both type-derived and custom generators.
+Each generator has four implementations, which are equivalent up to program equality:
+\begin {itemize}
+  \item The original \bq generator (``BQ'').
+  \item A staged version of the generator (``BQ++'').
+  \item A staged version of the generator using the C \rand (``BQC++'').
+  \item A staged version incorporating the CSR \rand (``BQCSR++'').
+\end{itemize}
+
+To ensure a fair comparison, we initialize each of these generators with the same random seed, ensuring that they produce identical trees.
+
+\subsubsection{Variance between seeds}
+The time-to-failure of a given strategy on a given seed is not necessarily representative of 
+the strategy's average time-to-fail over a large number of trials, so we normalize by computing the
+timing \textit{differential} between optimized and unoptimized generators (the ``speedup''). While this is true in most cases,
+it is not perfect.
+For example, it is theoretically possible
+to choose a seed such that the \textit{first} value produced by a generator discovers the bug, which would elide any differential speedup.
+To account for these edge cases, we repeated the above process for 10 seeds and computed the geometric average of their speedups.
+
+% To produce \todo{FIG}, we repeated the above process for 10 seeds and computed the geometric average of their speedups. 
+%This is important because the time-to-failure of a given strategy on a given seed is not necessarily 
+%representative of the strategy's overall bug-finding ability, although it often is. For example, it is theoretically possible 
+%to choose a seed such that the \textit{first} value produced by a generator discovers the bug, but it would be misleading to report this result.
+
+\subsubsection{Variance between trials}
+Although the variance \textit{between} seeds in Etna can be very vast, timing results are replicable for a given seed. Across 1,000 trials on the same seed, 
+the observed variance in time-to-failure was less than a nanosecond. \tr{Do we need this?}
+
 \section{Conclusion \& Future Work}

### 2025-03-17 paper/paper.tex
writing about abstraction overhead

--- 2025-03-17 paper/paper.tex
@@
 
+Because the inlined versions of the generator are identical to the un-inlined
+versions except for mechanical, semantics-preserving transformations,
+this performance difference is attributable solely to
+the different machine code generated by the compiler.
+Indeed, compilers of effectful and strict functional languages (including JIT
+compilers, in the case of Scala 3)
+use heuristics to determine if and when to perform this particular kind of
+simplification~\footnote{As we discuss in Section~\ref{asdf} \jwc{fixme}, purity
+means that Haskell is a slightly different story. GHC can and often does
+transformations of this form.}.  Even in cases as simple as
+Figure~\ref{fig:overhead-explanation-code}, the indirection of \texttt{return}
+and \texttt{bind} causes these heuristics to not fire.  Moreover, the story is
+even worse for recursive generators, as the heuristics are necessarily even more
+conservative for optimizing recursive functions.
+
 % Modern compilers use heuristics to decide when to specialize and inline code \cn, and these heuristics cannot guarantee that generator abstractions are zero-cost.
@@
   \begin{itemize}
-    \item While it's the ``correct'' abstraction for generators, using a monadic interface prevents the compiler from specializing generator code.
-Using monadic bind and return obscures the control and data flow of a generator from the compiler.
-This is
-    compounded when handling recursive functions.
     \item In cases where the compiler cannot statically eliminate them, running a monadic bind allocates a short-lived closure: we allocate a closure for the continuation, and then immediately jump into it.

### 2025-03-17 paper/paper.tex
halfway trhough 3.1

@@
 even worse for recursive generators, as the heuristics are necessarily even more
-conservative for optimizing recursive functions.
-
-% Modern compilers use heuristics to decide when to specialize and inline code \cn, and these heuristics cannot guarantee that generator abstractions are zero-cost.
-% These heuristics are also necessarily conservative about inlining and specializing recursive functions, which all generators of recursive data types must be.
-% The
+conservative for optimizing recursive functions. \jwc{unclear if we need this sentence, it's sorta obvious}
 
@@
 
-% \jwc{The overhead of the abstraction is 2x. The reality is actually worse: actual base-quickcheck includes even more indirection in its type.}
-
-% \jwc{The native code OCaml compiler, even with -O3, fails to specialize this code and eliminate the overhead of this abstraction --- inlining all of the funciton definitions and then performing beta-reductions speeds up sampling by a factor of 2.}
-
-% \jwc{Same story in scala. 
-% Approximately 2x difference for the same generator (though an order of magnitude slower overall than bq)... (scala uses a slightly different generator type, but the principles are the same)
-% This doesn't even account for more inlining we could do to eliminate the boxing/unboxing of results. (Scalacheck generators can fail, though this adds significant overhead)
-% }
-
-% [info] GenBm.generateComplexBase     avgt    3  458.325 ± 67.214  ns/op
-% [info] GenBm.generateComplexInlined  avgt    3  266.035 ±  8.559  ns/op
-
-% \jwc{
-% 
-% }
-
-\tr{Each of these consist of an explanation of the problem and pseudocode outlining the solution.}
-
-\jwc{
-  \begin{itemize}
-    \item In cases where the compiler cannot statically eliminate them, running a monadic bind allocates a short-lived closure: we allocate a closure for the continuation, and then immediately jump into it.
-    \item Each closure allocation is relatively cheap, but doing lots of allocation in a generation hot loop adds up fast.
-    \item Similar things have been noticed about monadic parsing DSLs in the past.
-  \end{itemize}
-}
+While neither of these microbenchmarks exhibit this behavior, more complex
+generators suffer further performance penalties due to \emph{closure
+allocation}. In cases where the compiler cannot statically eliminate it \jwc{``it'' = ``the allocation''},
+running a monadic bind allocates a short-lived closure: we allocate a closure
+for the continuation, and then immediately jump into it.
+In strict functional languages like OCaml and Scala,
+Each individual closure allocation is relatively cheap, but doing lots of allocation in a generation
+loop is very costly because each allocation brings us closer to the next costly GC pause \cn.
 

### 2025-03-17 paper/paper.tex
Fix

@@
 \subsubsection{Staging in eDSL Construction}
-
 One of the primary uses of staging is in embedded DSL construction \cite{sheard99, lms, trattdsl}.
@@
 Each individual closure allocation is relatively cheap, but doing lots of allocation in a generation
-loop is very costly because each allocation brings us closer to the next costly GC pause \cn.
+loop is very costly because each allocation brings us closer to the next costly GC pause \cn{}.
 

### 2025-03-17 paper/paper.tex
reordering

@@
 
-\section{Background}
-
-\jwc{In this paper, we'll use OCaml to illustrate the ideas, but the concepts
-are portable to other languages: see Section~\ref{subsection:other-langs}}
-
-\subsection{Property-Based Testing}
+\section{Sources of Inefficiency in Monadic Generator Libraries}
+\label{section:motiv}
+\jwc{In this section we answer: ``What are monadic generator libraries, and why are they slow?''}
+
+\subsection{Background: Monadic Generator DSLs}
 \jwc{Usual spiel about pbt and generators: here's a monadic generic generator library (but we're using BQ to illustrage), it has return and bind, this is what they're for.}
@@
 \begin{lstlisting}
-module Bq = struct
-  type 'a t = int -> SR.t -> 'a
-
-  let return (x : 'a) : 'a t = fun size seed -> x
-
-  let bind (g : 'a t) (f : 'a -> 'b t) : 'b t =
-    fun size seed ->
-      let x = g size seed in
-      (f x) size seed
-
-  let gen_int (lo : int) (hi : int) : int t =
-    fun size seed -> SR.int seed lo hi
+module Bq : sig
+  type 'a t
+  val return : 'a -> 'a t
+  val bind : 'a t -> ('a -> 'b t) -> 'b t
+  val gen_int : lo:int -> hi:int -> int t
 end
@@
 
-\subsection{Multi-Stage Programming}
+\jwc{IMPORTANT: We are using OCaml because we have to pick a syntax for this section, but these abstraction overheads exist in other languages -- at least scala, rust.}
+
+\subsection{Abstraction Overhead of Generator DSLs}
+Just how large is the abstraction overhead of monadic generator DSLs, and where does it come from?
+Figure~\ref{fig:bq-internals} shows the internals of (a simplified version of) \bq{}.
+A generator \texttt{'a Bq.t} is just a function of type~\texttt{int -> SR.t -> 'a},
+taking an \texttt{int} representing the current size parameter and a random seed \texttt{SR.t},
+and returning a generated value \texttt{'a}. It is an invariant of the library that function of this type are
+deterministic: for a fixed size and seed, it will return the same value. All of the randomness in testing comes from varying the initial seed.
+The monad functions \texttt{return} and \texttt{bind} are defined in the usual way
+for an instance of the reader monad \cn{}. \texttt{Return x} ignores the size and seed and just returns
+\texttt{x}, while \texttt{bind g k} first runs \texttt{g} and then passes the generated value to the continuation \texttt{k}.
+\bq{} uses a mutable seed, so the \texttt{random} passed to \texttt{(k a)} will be different from the one passed to \texttt{g}.
+The \texttt{gen\_int} combinator simply calls out to the randomness library \texttt{Splittable\_random}, aliased as \texttt{SR} here.
+\jwc{Should we do this?}
+
+Different PBT libraries use variations on this basic design. Haskell's QuickCheck uses a reader monad with an immutable state
+that is ``split'' at \texttt{bind}s \cn. Meanwhile, ScalaCheck (a) uses an immutable state type and state monad \cn{} to thread
+the state through, and (b) has an \texttt{Option} as its return type, to allow generation to fail.
+\jwc{However, the principles are the same, and we will use \bq{} to illustrate.}
+
+\begin{figure}
+\begin{lstlisting}
+module Bq = struct
+  type 'a t = int -> SR.t -> 'a
+
+  let return (x : 'a) : 'a t = fun _ _ -> x
+
+  let bind (g : 'a t) (k : 'a -> 'b t) : 'b t =
+    fun size random ->
+      let a = g size random in
+      (k a) size random
+
+  let gen_int (lo : int) (hi : int) : int t =
+    fun _ random -> SR.int random lo hi
+end
+\end{lstlisting}
+\caption{Internals of a Monadic Generator eDSL}
+\label{fig:bq-internals}
+\end{figure}
+
+Just how much runtime overhead does this monadic abstraction introduce? To
+illustrate, let's return to our running example of a constrained pair of
+integers, written in both \bq{} and ScalaCheck.
+Figure~\ref{fig:overhead-explanation-code} shows two versions of the generator, written in both languages.
+The first versions (\texttt{int\_pair} in \bq{} and \texttt{intPair} in ScalaCheck) are written with the monadic generator combinators from their respective libraries.
+\jwc{Maybe we can avoid rewriting these if we already showed them earlier? Also maybe we duplicate the tutorial discussion of monadic generator libraries to include Scala?}
+The second versions (\texttt{int\_pair\_inlined} and \texttt{intPairInlined})
+are semantically identical to the first, but have been rewritten
+by (1) inlining all generator combinator definitions, and then (2)
+repeatedly reducing simplifiable terms like \lstinline{(fun x -> e) e'} --- where an anonymous function is defined and then immediatly called~\footnote{This is usually known as a ``$\beta$-redex''.
+Note that when reducing, we do not reduce to \lstinline{e[e'/x]}, as this does not preserve the order of effects.
+}
+--- to \lstinline{let x = e' in e}.
+
+\begin{figure}
+\begin{lstlisting}
+let int_pair : (int * int) Bq.t =
+  let%bind x = (Bq.gen_int 0 100) in
+  let%bind y = (Bq.gen_int 0 x) in
+  Bq.return (x,y)
+
+let int_pair_inlined : int -> SR.t -> int * int =
+  fun _ sr ->
+    let x = Splittable_random.int sr ~lo:0 ~hi:100 in
+    let y = Splittable_random.int sr ~lo:0 ~hi:x in
+    (x,y)
+
+def intPair : Gen[(Long,Long)] = for {
+  x <- Gen.choose(0,1000)
+  y <- Gen.choose(0,x)
+} yield (x,y)
+
+def intPairInlined : (Gen.Parameters, Seed) => (Option[(Long,Long)],Seed) = {
+ (p,seed) =>
+  val (x,seed2) = chLng(0,1000)(p,seed)
+  x match {
+    case None => (None,seed2)
+    case Some(x) =>
+      val (y,seed3) = chLng(0,x)(p,seed2)
+      y match {
+        case None => (None,seed)
+        case Some(y) => (Some(x,y),seed3)
+      }
+    }
+}
+\end{lstlisting}
+\caption{Int Pair Generators in \bq{} and ScalaCheck}
+\label{fig:overhead-explanation-code}
+\end{figure}
+
+The performance impact of this inling is large (Figure~\ref{fig:overhead-explanation-perf}). In both languages, the
+inlined version takes (on average) \textbf{half} as much time to generate a single pair of
+\texttt{int}s. 
+Microbenchmarks of more realistic generators (see Section~\ref{section:eval}
+\jwc{more precise ref}) show an even more dramatic performance boost for
+inlining in this manner.
+
+Because the inlined versions of the generator are identical to the un-inlined
+versions except for mechanical, semantics-preserving transformations,
+this performance difference is attributable solely to
+the different machine code generated by the compiler.
+Indeed, compilers of effectful and strict functional languages (including JIT
+compilers, in the case of Scala 3)
+use heuristics to determine if and when to perform this particular kind of
+simplification~\footnote{As we discuss in Section~\ref{asdf} \jwc{fixme}, purity
+means that Haskell is a slightly different story. GHC can and often does
+transformations of this form.}.  Even in cases as simple as
+Figure~\ref{fig:overhead-explanation-code}, the indirection of \texttt{return}
+and \texttt{bind} causes these heuristics to not fire.  Moreover, the story is
+even worse for recursive generators, as the heuristics are necessarily even more
+conservative for optimizing recursive functions. \jwc{unclear if we need this sentence, it's sorta obvious}
+
+\begin{figure}
+  \begin{tabular}{lll}
+  Library    & Generator          & Average Time per Generation (ns) \\
+  \bq{}         & \texttt{int\_pair}          & 70                       \\
+  \bq{}         & \texttt{int\_pair\_inlined} & 35                       \\
+  ScalaCheck & \texttt{intPair}            & 458                      \\
+  ScalaCheck & \texttt{intPairInlined}     & 266                     
+  \end{tabular}
+\caption{Microbenchmarks of Generators in \bq{} (using \texttt{core\_bench} \cn) and ScalaCheck (using \texttt{jmh} \cn)}
+\label{fig:overhead-explanation-perf}
+\end{figure}
+
+While neither of these microbenchmarks exhibit this behavior, more complex
+generators suffer further performance penalties due to \emph{closure
+allocation}. In cases where the compiler cannot statically eliminate it \jwc{``it'' = ``the allocation''},
+running a monadic bind allocates a short-lived closure: we allocate a closure
+for the continuation, and then immediately jump into it.
+In strict functional languages like OCaml and Scala,
+Each individual closure allocation is relatively cheap, but doing lots of allocation in a generation
+loop is very costly because each allocation brings us closer to the next costly GC pause \cn{}.
+
+\subsubsection{Overhead of Choice Combinators}
+
+\jwc{
+Aside from the usual fact that the compiler can't ``see through'' the abstraction boundary to specialize, combinators like union and weighted union
+necessarily allocate a lot in the generation hot path.
+Even without the dumb BQ array thing, to call union and weighted union, the types mean that you have to allocate a list.
+These allocations are extremely costly, and happen every sample from the generator. If the generator is recursive (like the tree generator), they occur at each recursive call.
+}
+
+\jwc{In practice, the possible options are almost always statically known: in the binary tree generator from earlier, we can tell from the text of the code what
+the weights are, what the generators are, and how many there are.
+Put more simply, you basically always call weighted union with an \emph{explicit list}, which is then immediately traversed by the weighted union.
+For this reason, you should not have to incur the cost of allocating this list at runtime.
+However, almost no compilers (GHC is a notable exception) can figure this out and eliminate the list \cn (list fusion).
+}
+
+% \jwc{This builds the cdf of the distribution, samples from 0 to the total, and binary searches through the array to find the bucket.}
+% \jwc{Allocating this array is unnecessary: we can simply compute the total weight, sample \texttt{x} between 0 and the total, and then walk the list accumulating the sum, until the accumulator exceeds \texttt{x}.
+Since in practice the list of possible options is quite small (usually at most 10 in practice), the linear time scan is going to be much faster.}
+% \jwc{(The point of this para will just be to emphasize that it's almost always better to be allocation-aware, instead of algorithmically clever.)}
+
+% \subsubsection{Function call overhead}
+\subsection{Inefficient Randomness Libraries}
+\jwc{We really need to have explained the effect ordering equivalence aspect of the evaluation before here.}
+
+The core of any PBT generator library is a source of randomness: to generate
+random values of some datatype, we some access to random numbers!
+Different PBT libraries use different randomness libraries implementing different algorithms
+\footnote{The common term for such an algorithm or library is a ``Random Number
+Generator'' (RNG) we will avoid this term and instead say ``randomness library''
+to avoid confusing RNG implementations with the PBT gnenerator libraries that
+use them.}.
+Following the original Haskell QuickCheck implementation, \bq\ uses the SplitMix
+algorithm \cn, implemented as a (stateful) OCaml library called \texttt{Splittable\_random}. Meanwhile,
+ScalaCheck uses the JSF algorithm \cn.
+% https://www.pcg-random.org/posts/bob-jenkins-small-prng-passes-practrand.html
+% https://burtleburtle.net/bob/rand/smallprng.html
+
+The randomness library sits at the heart of the hot path.
+Even basic generators --- like ones generating a single \texttt{int} or \texttt{float} uniformly within a range --- can sample \emph{unboundedly many}
+random numbers, since they usually use versions of rejection sampling to find a value within the range \cn.
+Moreover, generator combinators like \texttt{list} usually make $O(n)$ calls to even those basic generators.
+Because of this, the speed of a single sample matters a great deal. Unfortunately,
+existing PBT libraries make relatively inefficient choices on this front,
+leading to worse bugfinding power than what is possible.
+
+\hg{This paragraph feels a bit clumsy right now. What if we started with a table
+of the RNGs used by a bunch of popular PBT frameworks and then argued that
+things like Lehmer are faster?}
+
+For example, significantly faster algorithms than SplitMix or JSF exist,
+such as the Lehmer algorithm\cn. In microbenchmarks, the Lehmer algorithm runs almost 2x as fast as SplitMix \cn.
+% https://github.com/lemire/testingRNG/tree/master
+Moreover, PBT libraries could even consider eschewing the requirement that a source of
+entropy pass statistical tests like BigCrush\cn. \hg{This needs more discussion} 
+This is common practice in other areas of testing already:
+fuzzers often simply use a buffer full of arbitrary bytes as a source of entropy
+\cn. Such approaches are also faster than algorithms like SplitMix: bumping a
+pointer and reading from memory (which can be pipelined trivially) will always
+be faster on modern CPUs than any random sampling algorithm with data-dependent arithmetic
+instructions.\hg{Should we just always be doing this? I think we want to argue
+that this is a bridge too far because you still need to generate the buffer
+ahead of time and you can run out of randomness?}
+
+Of course, simply arguing that the randomness library is on the hot
+path does not guarantee that a faster sampling leads to measurably faster
+generation; for that, we need an experiment. The most obvious experiment is to
+simply swap out the randomness library for \bq{} with a totally different, faster one. But, as
+discussed previously, generators with different generation orders can be
+difficult to compare. To get around this, we exploit a ``natural experiment:
+OCaml's \texttt{Splittable\_random} library is slow in a way that
+can be improved \emph{without} changing its extensional behavior.
+In particular,
+due to implementation details related to the OCaml garbage collector,
+values of the OCaml type \texttt{int64} are not machine words, but rather
+\emph{pointers} to machine words. This means
+that \emph{all} \texttt{int64} operations (both arithmetic and bitwise) must
+allocate memory cells to contain their output, which has a significant performance benefit.
+By building a version that uses much faster ``unboxed'' 64-bit integer arithmetic, we can demonstrate just how much bug-finding
+performance can be improved just by using a more performant randomness library.
+
+\section{Eliminating Abstraction Overhead of Generator DSLs by Staging and Faster Randomness Libraries}
+\jwc{This section needs a much better title...}
+\jwc{Emphasize that this is an \emph{COMPLETELY EQUIVALENT drop in replacement!} We din't just build a different library with different distributions.}
+
+\jwc{Usual introduction to this section, corresponding to how we talked about it in the intro. ``We present a library that XYZ''.}
+
+\hg{TODO: Signposting --- I got a few paragraphs in and realized I wasn't sure what I was reading}
+
+\subsection{Background: Multi-Stage Programming}
 \label{subsection:msp}
-
-\jwc{I think this section should actually go later.}
 
@@
 
-\subsubsection{Staging in eDSL Construction}
+\subsubsection{Uses of Staging in eDSL Construction}
 One of the primary uses of staging is in embedded DSL construction \cite{sheard99, lms, trattdsl}.
@@
 the functional abstractions used to build eDSLs prevent compilers from generating efficient machine code \cn.
-In the literature, this effect is known as ``abstraction overhead'' \cite{carrette05, moller20}: the layers of abstraction that 
+As alluded to previously \jwc{This sentence here because of the section-ordering transpose...
+may need to mention this earlier.}, this effect is known as ``abstraction overhead'' \cite{carrette05, moller20}: the layers of abstraction that
 make eDSLs so usable are precisely what prevents them from being fast.
+\jwc{Something to the effect of ``the issues we noticed in BQ and SC exist across many eDSLs''}
 In light of this issue, staging is often used as a lightweight compiler for DSLs. The compile-time evaluation stage
@@
 % }
-
-\subsection{Other Languages and Libraries}
-\label{subsection:other-langs}
-
-\jwc{TODO: Move this section later (cf conversation in WH meeting 3/7/25)}
-
-\tr{Here's where we talk about what's going on in the world, and ultimately make the argument that \bq{} is the best tool to reproduce.}
-
-\jwc{Note that this explanation require some prior note of exactly *why* BQ is slow... i.e. the abstraction overhead of the library is high.}
-
-\jwc{
-  \begin{itemize}
-    \item Scala.
-Functional abstractions like QC generators are known to be costly in Scala, that's why they have LMS (in Scala 2, and Macros in Scala 3).
-Example: parser combinators (``On Staged Parser Combinators for Efficient Data Processing''), functional data structures (\href{https://ppl.stanford.edu/papers/popl13_rompf.pdf}{Link}), web programming (``Efficient High-Level Abstractions for Web Programming'').
-    Al of this should still work in scala.
-Could easily be incorporated into ScalaCheck, with minor modification: ScalaCheck uses a state monad to thread around the seed, instead of a stateful one (like BQ), or a splittable one (like Haskell).
-So you have to adapt to that.
-But same diff.
-    \item Python. Maybe could do it?? Hypothesis + MacroPy
-    \item Haskell: GHC does a lot of these optimizations already, since the code is pure. Since QC generators are relatively small programs,
-    GHC has little trouble specializing them. Of course, this is not guaranteed. A version of this idea can easily be ported to the original QC with template haskell, to guarantee
-    the highest-performance generators.
-    \item Rust: Not GC'd, so no alloc overhead but bind'd generators still dispatch through runtime data.
-  \end{itemize}
-}
-
-\section{Sources of Inefficiency in Monadic Generator Libraries}
-\label{section:motiv}
-\jwc{In this section we answer: ``Why are monadic generator libraries slow?''}
-
-\jwc{IMPORTANT: We are using OCaml because we have to pick a syntax for this section, but these abstraction overheads exist in other languages -- at least scala, rust.}
-
-\subsection{Monadic DSL Abstraction Overhead}
-Just how large is the abstraction overhead of monadic generator DSLs, and where does it come from?
-Figure~\ref{fig:bq-internals} shows the internals of (a simplified version of) \bq{}.
-A generator \texttt{'a Bq.t} is just a function of type~\texttt{int -> SR.t -> 'a},
-taking an \texttt{int} representing the current size parameter and a random seed \texttt{SR.t},
-and returning a generated value \texttt{'a}. It is an invariant of the library that function of this type are
-deterministic: for a fixed size and seed, it will return the same value. All of the randomness in testing comes from varying the initial seed.
-The monad functions \texttt{return} and \texttt{bind} are defined in the usual way
-for an instance of the reader monad \cn{}. \texttt{Return x} ignores the size and seed and just returns
-\texttt{x}, while \texttt{bind g k} first runs \texttt{g} and then passes the generated value to the continuation \texttt{k}.
-\bq{} uses a mutable seed, so the \texttt{random} passed to \texttt{(k a)} will be different from the one passed to \texttt{g}.
-The \texttt{gen\_int} combinator simply calls out to the randomness library \texttt{Splittable\_random}, aliased as \texttt{SR} here.
-\jwc{Should we do this?}
-
-Different PBT libraries use variations on this basic design. Haskell's QuickCheck uses a reader monad with an immutable state
-that is ``split'' at \texttt{bind}s \cn. Meanwhile, ScalaCheck (a) uses an immutable state type and state monad \cn{} to thread
-the state through, and (b) has an \texttt{Option} as its return type, to allow generation to fail.
-\jwc{However, the principles are the same, and we will use \bq{} to illustrate.}
-
-\begin{figure}
-\begin{lstlisting}
-module Bq = struct
-  type 'a t = int -> SR.t -> 'a
-
-  let return (x : 'a) : 'a t = fun _ _ -> x
-
-  let bind (g : 'a t) (k : 'a -> 'b t) : 'b t =
-    fun size random ->
-      let a = g size random in
-      (k a) size random
-
-  let gen_int (lo : int) (hi : int) : int t =
-    fun _ random -> SR.int random lo hi
-end
-\end{lstlisting}
-\caption{Internals of a Monadic Generator eDSL}
-\label{fig:bq-internals}
-\end{figure}
-
-Just how much runtime overhead does this monadic abstraction introduce? To
-illustrate, let's return to our running example of a constrained pair of
-integers, written in both \bq{} and ScalaCheck.
-Figure~\ref{fig:overhead-explanation-code} shows two versions of the generator, written in both languages.
-The first versions (\texttt{int\_pair} in \bq{} and \texttt{intPair} in ScalaCheck) are written with the monadic generator combinators from their respective libraries.
-\jwc{Maybe we can avoid rewriting these if we already showed them earlier? Also maybe we duplicate the tutorial discussion of monadic generator libraries to include Scala?}
-The second versions (\texttt{int\_pair\_inlined} and \texttt{intPairInlined})
-are semantically identical to the first, but have been rewritten
-by (1) inlining all generator combinator definitions, and then (2)
-repeatedly reducing simplifiable terms like \lstinline{(fun x -> e) e'} --- where an anonymous function is defined and then immediatly called~\footnote{This is usually known as a ``$\beta$-redex''.
-Note that when reducing, we do not reduce to \lstinline{e[e'/x]}, as this does not preserve the order of effects.
-}
---- to \lstinline{let x = e' in e}.
-
-\begin{figure}
-\begin{lstlisting}
-let int_pair : (int * int) Bq.t =
-  let%bind x = (Bq.gen_int 0 100) in
-  let%bind y = (Bq.gen_int 0 x) in
-  Bq.return (x,y)
-
-let int_pair_inlined : int -> SR.t -> int * int =
-  fun _ sr ->
-    let x = Splittable_random.int sr ~lo:0 ~hi:100 in
-    let y = Splittable_random.int sr ~lo:0 ~hi:x in
-    (x,y)
-
-def intPair : Gen[(Long,Long)] = for {
-  x <- Gen.choose(0,1000)
-  y <- Gen.choose(0,x)
-} yield (x,y)
-
-def intPairInlined : (Gen.Parameters, Seed) => (Option[(Long,Long)],Seed) = {
- (p,seed) =>
-  val (x,seed2) = chLng(0,1000)(p,seed)
-  x match {
-    case None => (None,seed2)
-    case Some(x) =>
-      val (y,seed3) = chLng(0,x)(p,seed2)
-      y match {
-        case None => (None,seed)
-        case Some(y) => (Some(x,y),seed3)
-      }
-    }
-}
-\end{lstlisting}
-\caption{Int Pair Generators in \bq{} and ScalaCheck}
-\label{fig:overhead-explanation-code}
-\end{figure}
-
-The performance impact of this inling is large (Figure~\ref{fig:overhead-explanation-perf}). In both languages, the
-inlined version takes (on average) \textbf{half} as much time to generate a single pair of
-\texttt{int}s. 
-Microbenchmarks of more realistic generators (see Section~\ref{section:eval}
-\jwc{more precise ref}) show an even more dramatic performance boost for
-inlining in this manner.
-
-Because the inlined versions of the generator are identical to the un-inlined
-versions except for mechanical, semantics-preserving transformations,
-this performance difference is attributable solely to
-the different machine code generated by the compiler.
-Indeed, compilers of effectful and strict functional languages (including JIT
-compilers, in the case of Scala 3)
-use heuristics to determine if and when to perform this particular kind of
-simplification~\footnote{As we discuss in Section~\ref{asdf} \jwc{fixme}, purity
-means that Haskell is a slightly different story. GHC can and often does
-transformations of this form.}.  Even in cases as simple as
-Figure~\ref{fig:overhead-explanation-code}, the indirection of \texttt{return}
-and \texttt{bind} causes these heuristics to not fire.  Moreover, the story is
-even worse for recursive generators, as the heuristics are necessarily even more
-conservative for optimizing recursive functions. \jwc{unclear if we need this sentence, it's sorta obvious}
-
-\begin{figure}
-  \begin{tabular}{lll}
-  Library    & Generator          & Average Time per Generation (ns) \\
-  \bq{}         & \texttt{int\_pair}          & 70                       \\
-  \bq{}         & \texttt{int\_pair\_inlined} & 35                       \\
-  ScalaCheck & \texttt{intPair}            & 458                      \\
-  ScalaCheck & \texttt{intPairInlined}     & 266                     
-  \end{tabular}
-\caption{Microbenchmarks of Generators in \bq{} (using \texttt{core\_bench} \cn) and ScalaCheck (using \texttt{jmh} \cn)}
-\label{fig:overhead-explanation-perf}
-\end{figure}
-
-While neither of these microbenchmarks exhibit this behavior, more complex
-generators suffer further performance penalties due to \emph{closure
-allocation}. In cases where the compiler cannot statically eliminate it \jwc{``it'' = ``the allocation''},
-running a monadic bind allocates a short-lived closure: we allocate a closure
-for the continuation, and then immediately jump into it.
-In strict functional languages like OCaml and Scala,
-Each individual closure allocation is relatively cheap, but doing lots of allocation in a generation
-loop is very costly because each allocation brings us closer to the next costly GC pause \cn{}.
-
-\subsubsection{Overhead of Other Combinators}
-
-\jwc{
-Aside from the usual fact that the compiler can't ``see through'' the abstraction boundary to specialize, combinators like union and weighted union
-necessarily allocate a lot in the generation hot path.
-Even without the dumb BQ array thing, to call union and weighted union, the types mean that you have to allocate a list.
-These allocations are extremely costly, and happen every sample from the generator. If the generator is recursive (like the tree generator), they occur at each recursive call.
-}
-
-\jwc{In practice, the possible options are almost always statically known: in the binary tree generator from earlier, we can tell from the text of the code what
-the weights are, what the generators are, and how many there are.
-Put more simply, you basically always call weighted union with an \emph{explicit list}, which is then immediately traversed by the weighted union.
-For this reason, you should not have to incur the cost of allocating this list at runtime.
-However, almost no compilers (GHC is a notable exception) can figure this out and eliminate the list \cn (list fusion).
-}
-
-% \jwc{This builds the cdf of the distribution, samples from 0 to the total, and binary searches through the array to find the bucket.}
-% \jwc{Allocating this array is unnecessary: we can simply compute the total weight, sample \texttt{x} between 0 and the total, and then walk the list accumulating the sum, until the accumulator exceeds \texttt{x}.
-Since in practice the list of possible options is quite small (usually at most 10 in practice), the linear time scan is going to be much faster.}
-% \jwc{(The point of this para will just be to emphasize that it's almost always better to be allocation-aware, instead of algorithmically clever.)}
-
-% \subsubsection{Function call overhead}
-\subsection{Choice of Randomness Library}
-\jwc{We really need to have explained the effect ordering equivalence aspect of the evaluation before here.}
-
-The core of any PBT generator library is a source of randomness: to generate
-random values of some datatype, we some access to random numbers!
-Different PBT libraries use different randomness libraries implementing different algorithms
-\footnote{The common term for such an algorithm or library is a ``Random Number
-Generator'' (RNG) we will avoid this term and instead say ``randomness library''
-to avoid confusing RNG implementations with the PBT gnenerator libraries that
-use them.}.
-Following the original Haskell QuickCheck implementation, \bq\ uses the SplitMix
-algorithm \cn, implemented as a (stateful) OCaml library called \texttt{Splittable\_random}. Meanwhile,
-ScalaCheck uses the JSF algorithm \cn.
-% https://www.pcg-random.org/posts/bob-jenkins-small-prng-passes-practrand.html
-% https://burtleburtle.net/bob/rand/smallprng.html
-
-The randomness library sits at the heart of the hot path.
-Even basic generators --- like ones generating a single \texttt{int} or \texttt{float} uniformly within a range --- can sample \emph{unboundedly many}
-random numbers, since they usually use versions of rejection sampling to find a value within the range \cn.
-Moreover, generator combinators like \texttt{list} usually make $O(n)$ calls to even those basic generators.
-Because of this, the speed of a single sample matters a great deal. Unfortunately,
-existing PBT libraries make relatively inefficient choices on this front,
-leading to worse bugfinding power than what is possible.
-
-\hg{This paragraph feels a bit clumsy right now. What if we started with a table
-of the RNGs used by a bunch of popular PBT frameworks and then argued that
-things like Lehmer are faster?}
-
-For example, significantly faster algorithms than SplitMix or JSF exist,
-such as the Lehmer algorithm\cn. In microbenchmarks, the Lehmer algorithm runs almost 2x as fast as SplitMix \cn.
-% https://github.com/lemire/testingRNG/tree/master
-Moreover, PBT libraries could even consider eschewing the requirement that a source of
-entropy pass statistical tests like BigCrush\cn. \hg{This needs more discussion} 
-This is common practice in other areas of testing already:
-fuzzers often simply use a buffer full of arbitrary bytes as a source of entropy
-\cn. Such approaches are also faster than algorithms like SplitMix: bumping a
-pointer and reading from memory (which can be pipelined trivially) will always
-be faster on modern CPUs than any random sampling algorithm with data-dependent arithmetic
-instructions.\hg{Should we just always be doing this? I think we want to argue
-that this is a bridge too far because you still need to generate the buffer
-ahead of time and you can run out of randomness?}
-
-Of course, simply arguing that the randomness library is on the hot
-path does not guarantee that a faster sampling leads to measurably faster
-generation; for that, we need an experiment. The most obvious experiment is to
-simply swap out the randomness library for \bq{} with a totally different, faster one. But, as
-discussed previously, generators with different generation orders can be
-difficult to compare. To get around this, we exploit a ``natural experiment:
-OCaml's \texttt{Splittable\_random} library is slow in a way that
-can be improved \emph{without} changing its extensional behavior.
-In particular,
-due to implementation details related to the OCaml garbage collector,
-values of the OCaml type \texttt{int64} are not machine words, but rather
-\emph{pointers} to machine words. This means
-that \emph{all} \texttt{int64} operations (both arithmetic and bitwise) must
-allocate memory cells to contain their output, which has a significant performance benefit.
-By building a version that uses much faster ``unboxed'' 64-bit integer arithmetic, we can demonstrate just how much bug-finding
-performance can be improved just by using a more performant randomness library.
-
-\section{Eliminating Abstraction Overhead of Generator DSLs by Staging and Faster Randomness Libraries}
-\jwc{This section needs a much better title...}
-\jwc{Emphasize that this is an \emph{COMPLETELY EQUIVALENT drop in replacement!} We din't just build a different library with different distributions.}
-
-\jwc{Usual introduction to this section, corresponding to how we talked about it in the intro. ``We present a library that XYZ''.}
-
-\hg{TODO: Signposting --- I got a few paragraphs in and realized I wasn't sure what I was reading}
 
@@
 
-\section{Conclusion \& Future Work}
+\section{Other Languages and Libraries}
+\label{subsection:other-langs}
+
+\jwc{TODO: Move this section later (cf conversation in WH meeting 3/7/25)}
+
+\tr{Here's where we talk about what's going on in the world, and ultimately make the argument that \bq{} is the best tool to reproduce.}
+
+\jwc{Note that this explanation require some prior note of exactly *why* BQ is slow... i.e. the abstraction overhead of the library is high.}
+
+\jwc{
+  \begin{itemize}
+    \item Scala.
+Functional abstractions like QC generators are known to be costly in Scala, that's why they have LMS (in Scala 2, and Macros in Scala 3).
+Example: parser combinators (``On Staged Parser Combinators for Efficient Data Processing''), functional data structures (\href{https://ppl.stanford.edu/papers/popl13_rompf.pdf}{Link}), web programming (``Efficient High-Level Abstractions for Web Programming'').
+    Al of this should still work in scala.
+Could easily be incorporated into ScalaCheck, with minor modification: ScalaCheck uses a state monad to thread around the seed, instead of a stateful one (like BQ), or a splittable one (like Haskell).
+So you have to adapt to that.
+But same diff.
+    \item Python. Maybe could do it?? Hypothesis + MacroPy
+    \item Haskell: GHC does a lot of these optimizations already, since the code is pure. Since QC generators are relatively small programs,
+    GHC has little trouble specializing them. Of course, this is not guaranteed. A version of this idea can easily be ported to the original QC with template haskell, to guarantee
+    the highest-performance generators.
+    \item Rust: Not GC'd, so no alloc overhead but bind'd generators still dispatch through runtime data.
+  \end{itemize}
+}
+
 \section{Related Work}
 \jwc{A generator is a parser of randomness, and people have implemented lots of staged parser libraries!}
+
+\section{Conclusion \& Future Work}
 

### 2025-03-17 paper/paper.tex
transpose

@@
 
-\section{Eliminating Abstraction Overhead of Generator DSLs by Staging and Faster Randomness Libraries}
+\section{Speeding up Generator DSLs by Staging and Faster Randomness Libraries}
 \jwc{This section needs a much better title...}
@@
 \label{subsection:basic-design}
-Recall that staging a DSL involves changing the
+\jwc{Needs a lot more signposting at the top here.}
+
+Recall \jwc{did we talk about this} that staging a DSL involves changing the
 combinators to run at compile time by carefully annotating their types with \texttt{code}s.

### 2025-03-17 paper/paper.tex
picking nits

@@
 libraries, each call to monadic bind usually spends time constructing and
-immediately executing a closure at runtime. \jwc{Todo: joe --- cite some evidence for this}
+immediately executing a closure at run time. \jwc{Todo: joe --- cite some evidence for this}
 Meanwhile, a staged approach can resolve the bind at compile time, enabling the
@@
 
-Just how much runtime overhead does this monadic abstraction introduce? To
+Just how much run-time overhead does this monadic abstraction introduce? To
 illustrate, let's return to our running example of a constrained pair of
@@
 Put more simply, you basically always call weighted union with an \emph{explicit list}, which is then immediately traversed by the weighted union.
-For this reason, you should not have to incur the cost of allocating this list at runtime.
+For this reason, you should not have to incur the cost of allocating this list at run time.
 However, almost no compilers (GHC is a notable exception) can figure this out and eliminate the list \cn (list fusion).
@@
 ahead of time and you can run out of randomness?}
+\jwc{HG: talk about the tradeoffs here.}
+
+\jwc{The point here is that ``you should be thinking about using a more performant RNG, we'll show you that it does have real benefits (and quantify them). But there are tradeoffs here.'}
 
@@
 \subsubsection{The Trick at Other Types}
-To write more interesting generators, we also need the ability to generate code that manipulates runtime values.
+To write more interesting generators, we also need the ability to generate code that manipulates run-time values.
 For instance, consider this generator
@@
 Programmers can then define recursive generators by \texttt{fixed\_point}, which ties the recursive knot.
+\jwc{don't want to handle GADTs}
 
@@
     the highest-performance generators.
-    \item Rust: Not GC'd, so no alloc overhead but bind'd generators still dispatch through runtime data.
+    \item Rust: Not GC'd, so no alloc overhead but bind'd generators still dispatch through run-time data.
   \end{itemize}
@@
 \section{Conclusion \& Future Work}
+\jwc{Why we stopped where we stopped: can you squeeze more out of this?}
 

### 2025-03-17 paper/paper.tex
more in sig

@@
   val gen_int : lo:int -> hi:int -> int t
+
+  val weighted_union : (int * 'a t) list -> 'a t
+
+  val size : int t
+  val with_size : int -> 'a t -> 'a t
+
+  val fixed_point : ('a t -> 'a t) -> 'a t
+  $\dots$
 end

### 2025-03-17 paper/paper.tex
I'm committing.

@@
 In strict functional languages like OCaml and Scala,
-Each individual closure allocation is relatively cheap, but doing lots of allocation in a generation
-loop is very costly because each allocation brings us closer to the next costly GC pause \cn{}.
+Each individual closure allocation is relatively cheap, but doing lots of allocation in a
+generator is very costly because each allocation brings us closer to the next costly GC pause \cn{}.
+Similarly to \jwc{the previous one} this effect is magnified in recursive generators: each iteration
+through the recursive loop re-allocates closures for \texttt{bind}s, so the amount of allocation
+\emph{per generated value} scales linearly with the number of recursive generator calls.
 
 \subsubsection{Overhead of Choice Combinators}
+The monad functions \texttt{return} and \texttt{bind} aren't the only combinators that incur an
+abstraction overhead: all of them do! \jwc{I'm so bad at starting new sections...} ``Choice'' combinators like \texttt{weighted\_union}, however
+are particularly costly, because they unavoidably allocate a lot.
 
 \jwc{
-Aside from the usual fact that the compiler can't ``see through'' the abstraction boundary to specialize, combinators like union and weighted union
-necessarily allocate a lot in the generation hot path.
-Even without the dumb BQ array thing, to call union and weighted union, the types mean that you have to allocate a list.
-These allocations are extremely costly, and happen every sample from the generator. If the generator is recursive (like the tree generator), they occur at each recursive call.
+Aside from the usual fact that the compiler can't ``see through'' the
+abstraction boundary to specialize, combinators like union and weighted union
+necessarily allocate a lot in the generation hot path.  Even without the dumb BQ
+array thing, to call union and weighted union, the types mean that you have to
+allocate a list.  These allocations are extremely costly, and happen every
+sample from the generator. If the generator is recursive (like the tree
+generator), they occur at each recursive call.
 }
 
-\jwc{In practice, the possible options are almost always statically known: in the binary tree generator from earlier, we can tell from the text of the code what
-the weights are, what the generators are, and how many there are.
-Put more simply, you basically always call weighted union with an \emph{explicit list}, which is then immediately traversed by the weighted union.
-For this reason, you should not have to incur the cost of allocating this list at run time.
-However, almost no compilers (GHC is a notable exception) can figure this out and eliminate the list \cn (list fusion).
-}
+In practice, \texttt{weighted\_union}s is almost always\footnote{
+There are some generators where \texttt{weighted\_union} is passed a list which
+was itself the result of a generator: the well-typed STLC term generator used in Section~\ref{section:eval} is an example of this.
+} passed an explicitly constructed list, like
+\texttt{weighted\_union [(3,g); (1,g'); (2,g'')]}. 
+This is because the most common use case for \texttt{weighted\_union} is
+to choose between one of the different constructors of an algebraic datatype,
+the options for which are always known.
+The list passed to \texttt{weighted\_union} allocated at the call site, and then \emph{immediately}
+destructed by the function. In principle, since the elements of the list and its length $n$ are known statically,
+a compiler could unroll the loops in \texttt{weighted\_union} to depth $n$ and specialize the function to avoid allocating the list.
+In practice however almost~\jwc{Once again, GHC is a notable exception here \cn{}} no compilers perform this kind of optimization.
 

### 2025-03-18 paper/paper.tex
progress

@@
 \setcopyright{none}
-
+\usepackage{pgfplots}
+\usepackage{pgfplotstable}
+\pgfplotsset{compat=1.17}
 \usepackage{listings}
@@
 
-Property-based testing (PBT) is a technique for software testing which uses
-random test inputs to evaluate logical properties that the system is expected to
-satisfy.  Recent studies on PBT usage~\cite{inpractice} have shown that
-practitioners interleave PBT with their software development process, sometimes
-testing their properties every time they save their code. This workflow makes
-sense because the developer who just wrote the code is often the best person to
-fix a potental bug; if a bug is found later, it will need to be fixed by someone
-with less context. \jwc{Not sure how the previous sentence is relevant. It delays us getting to the point.}  All of this means that PBT is often run with very short time
-budgets --- e.g., between 50 milliseconds and 30 seconds --- and it is important
-that bugs can be found in that time-frame.
+Property-based testing (PBT) is a software testing technique which uses
+random test inputs to validate logical specifications.  Recent studies on PBT usage~\cite{inpractice} have shown that
+practitioners often interleave PBT with their software development process, sometimes
+testing their properties every time they save their code. This means that PBT is often run with very short time
+budgets---e.g., between 50 milliseconds and 30 seconds---and it is important
+that bugs can be found in that time-frame.\iflater
 \jwc{phrasing could suggest ``race against time'' aspect.}\hg{Agree, can't see
-how to fit this in yet}
-
-Given these strict timelines, it is important to think of PBT as
-performance-sensitive process. One important locus for performance improvement
+how to fit this in yet}\fi
+
+Given these strict timelines, the impact of performance in PBT cannot be overstated. One important locus for performance improvement
 is the \textit{generators} that produce random inputs to the properties under
@@
 provided with the tools to write fast generators, then they do not themselves
-need to be experts in performance tuning. \jwc{... to find bugs fast}
-
-We find that the performance of existing generator libraries falls short of what
+need to be experts in performance tuning to find bugs fast. \iflater\bcp{Incorporate fail fast somehow.}\fi
+
+However, careful measurements show that the performance of existing generator libraries falls significantly short of what
 is possible. This is due to two major sources of inefficiency.
@@
 flexibility and expressiveness also introduces overhead: layered abstractions
-and indirections hinder the compiler's optimizations, shifting more work to
+and indirections hinder compiler optimizations, shifting work to
 run time.
-Second, the \rand that underlies the generator's operation --- the source of
-random numbers that are used to make choices in the generator --- 
-can constitute an unexpectedly high proportion of a generator's running time.
+Second, the \rand that underlies the generator's operation---the source of the
+random numbers that are used to make choices in the generator---can 
+constitute an unexpectedly high proportion of a generator's running time.
 \iflater
@@
 \fi
-These observations together raise a fundamental question: can generator
-libraries produce code that is both expressive \textit{and} efficient?
+Together, these observations raise a fundamental question: can generator libraries be expressive and high-level without compromising efficiency?
 %In practice, PBT is used 
@@
 \textit{multi-stage programming}, or ``staging,'' which has long been used for
-optimizing idiomatic functional DSLs~\cn.
-We extend this technique to generators \jwc{``apply this technology to''? We're not extending staging, we're a client}, completely eliminating the overhead of
+optimizing functional DSLs~\cn.
+We apply this technique to generators, completely eliminating the overhead of
 many common generator abstractions. For example, in unstaged generator
@@
 Secondly, we demonstrate the impact of different choices of \rand on generator
-performance by conducting a controlled comparison. \hg{More words here for
-parallelism?} Together, these insights provide a recipe for improving even
-state-of-the-art generator libraries.
+performance by conducting a controlled comparison. \iflater\hg{More words here for
+parallelism?}\fi Together, these insights provide a recipe for improving existing generator libraries.
 
@@
 
-To demonstrate the effectiveness of our technique, we apply it to \bq, a
-high-quality PBT library in OCaml, authored by Jane Street.
+To demonstrate the effectiveness of our approach, we apply it to \bq, a
+high-quality PBT library in OCaml, authored by software engineers at the trading firm Jane Street.
 Our improved version, \name, is a direct drop-in replacement for
 \bq that 
-is \textit{completely semantically equivalent}: that is, given identical random seeds, generators written using \name{}
+is \textit{100\% semantically equivalent}: that is, given identical random seeds, generators written using \name{}
 produce exactly the same sequence of values as those written in \bq. However, in \name, 
 the abstractions provided by the generator DSL are zero-cost, and the \rand is 
-highly-optimized. The direct equivalence between \name and \bq allows us to 
-perform extremely fine-grained comparisons, \propose{as} differences in \proposechange{generation speed}{bugfinding power} are
-\proposechange{directly}{solely} attributable to the staging and \rand optimizations.
+highly-optimized. The semantic equivalence between \name and \bq allows us to 
+perform clean, fine-grained comparisons, as differences in generation speed are
+solely attributable to the staging and \rand optimizations.
 
@@
 Finally, to demonstrate the generality of our technique, we implement a version
-of \name in Scala \hg{(based on scalacheck?)} and benchmark the same generators
-in the Scala port of the library---thereby highlighting its portability to other
+of \name in Scala \hg{(based on scalacheck?)} and benchmark the same generators---thereby highlighting its portability to other
 strict functional languages.
@@
     \item We identify two key sources of inefficiencies in PBT generator libraries, which can significantly impact performance: 
-    namely, \rand choice, and abstraction overhead \jwc{Swap}. \secref{section:motiv}
-    \item We apply multi-stage programming to the novel domain of generators. \jwc{Less sure this should be a top-level conrib.} \secref{section:}
-    \item We implement a version of \bq, \name, that showcases our optimization technique \jwc{Calling it an ``optimization technique'' begs the ``so they just made a thing faster'' question} while maintaining strict program equality.
+    namely, and abstraction overhead, and \rand choice. \secref{section:motiv}
+    % \item We apply multi-stage programming to the novel domain of generators. \jwc{Less sure this should be a top-level conrib.} \secref{section:}
+    \item We implement our approach in OCaml and Scala, showcasing its cross-language applicability.
     \secref{section:4}
-    \jwc{We implement \emph{two} versions of \name{}! Once for BQ and once for ScalaCheck.}
-    \item We evaluate \name by demonstrating the improved performance of generators constructed
-    using our library in a controlled comparison, as well as its impact on bug-finding ability. \secref{section:eval}
+    \item We construct generators in \name and \scalaname, and show that they give substantial performance improvements. Finally, we show
+    that these performance improvements extend to bug-finding ability.
+     \secref{section:eval}
 \end{enumerate}
@@
 \label{subsection:pbt-for-pbt}
-To establish the equivalence of generators up to program equality, we use a PBT harness to compare their outputs. 
-Specifically, we generate random sizes and seeds and evaluate whether the generators under test produce equal results. 
-If the equality property fails, we report the unequal trees; if, over 1,000 trials, no ``counterexample'' is reported, we consider the
-behavior of the generators to be equivalent.
+To establish the equivalence of generators up to program equality, 
+we employ a PBT harness to compare their outputs over 1,000 random seeds. 
+If a discrepancy arises, we report the unequal trees; if no counterexample is found,
+we conclude that the generators exhibit equivalent behavior.
 \subsection{Benchmarking speed \& resource usage}
@@
 Etna allows users to measure the effectiveness of different generator implementations by injecting bugs into the system under test and recording the time taken for a relevant 
-property to fail in response. For example, in a tree with an ordering property, a bug injection might consist of inserting a value that violates that ordering.
+property to fail in response. For example, in a tree with an ordering property, a bug injection might consist of inserting a value into the tree that violates that ordering.
 
@@
 
-We evaluate each case study with both type-derived and custom generators.
+We evaluate our case studies with both type-derived and custom generators.
 Each generator has four implementations, which are equivalent up to program equality:
@@
   \item The original \bq generator (``BQ'').
-  \item A staged version of the generator (``BQ++'').
-  \item A staged version of the generator using the C \rand (``BQC++'').
-  \item A staged version incorporating the CSR \rand (``BQCSR++'').
+  \item A staged version of the generator (``Staged'').
+  \item A staged version of the generator using the C \rand (``Staged C'').
+  \item A staged version incorporating the CSR \rand (``Staged CSR'').
 \end{itemize}
@@
 the strategy's average time-to-fail over a large number of trials, so we normalize by computing the
-timing \textit{differential} between optimized and unoptimized generators (the ``speedup''). While this is true in most cases,
+timing \textit{differential} between optimized and unoptimized generators (the ``speedup''). While this works in most cases,
 it is not perfect.
 For example, it is theoretically possible
-to choose a seed such that the \textit{first} value produced by a generator discovers the bug, which would elide any differential speedup.
-To account for these edge cases, we repeated the above process for 10 seeds and computed the geometric average of their speedups.
+to choose a seed such that the \textit{first} value produced by a generator discovers the bug, which would elide any differential speedup (and likewise
+if both generators fail to find the bug).
+To account for these edge cases, we repeat the above process for 10 seeds and compute the geometric mean of their speedups.
 
@@
 \label{subsection:other-langs}
+
+\begin{figure}[h]
+  \centering
+  \vspace{0.5em} % Kept the vertical spacing
+  \begin{tabular}{lrl}
+  \toprule
+  Benchmark & \multicolumn{2}{c}{avg. time (ns/op) $\pm$ std. dev.} \\
+  \midrule
+  BST & $3,560.813$ & $\pm\>1,511.201$ \\
+  BST, Staged & $751.287$ & $\pm\>39.890$ \\
+  BST (Type-Derived) & $44,619.863$ & $\pm\>5,761.568$ \\
+  BST (Type-Derived), Staged & $6,863.209$ & $\pm\>416.188$ \\
+  RBT (Type-Derived) & $60,559.846$ & $\pm\>4,901.980$ \\
+  RBT (Type-Derived), Staged & $8,511.528$ & $\pm\>324.615$ \\
+  STLC & $44,497.623$ & $\pm\>1,970.484$ \\
+  STLC, Staged & $7,151.259$ & $\pm\>355.639$ \\
+  \bottomrule
+  \end{tabular}
+ \caption{\todo{Understand Joe's implementation of this...}}
+ \label{fig:benchmark-table}
+ \end{figure}
 

### 2025-03-18 paper/paper.tex
progress

@@
   % in strict functionallanguages like OCaml and Scal
-To address this question, we propose a two-pronged approach for designing
-generator libraries that preserves their expressiveness while eliminating
-unnecessary overhead.
+  To address this question, we provide two key insights for designing generator libraries, which eliminates unnecessary overhead while preserving expressiveness.
   First, we describe a method of optimizing generator libraries using
@@
 
-\jwc{Ensure we have a signposting table-of-contents-ish few sentences here, if it's not covered by the contribs.}
+\jwc{Ensure we have a signposting table-of-contents-ish few sentences here, if it's not covered by the contribs.}\todo{Todo Thia}
 

### 2025-03-18 paper/paper.tex
progress
chart

@@
   % in strict functionallanguages like OCaml and Scal
-  To address this question, we provide two key insights for designing generator libraries, which eliminates unnecessary overhead while preserving expressiveness.
+  To address this question, we outline two key principles for designing generator libraries that balance efficiency with flexibility.
  First, we describe a method of optimizing generator libraries using
@@
 \subsection{Implementing and Testing Generators}
+\begin{figure}[h]
+  \includegraphics[scale=0.34]{allocs.png}\includegraphics[scale=0.334]{time.png}
+\end{figure}
 \label{subsection:pbt-for-pbt}

### 2025-03-18 paper/paper.tex
Tiny comment

@@
 \jwc{In this section we answer: ``What are monadic generator libraries, and why are they slow?''}
+\bcp{That's actually not a bad section title!}
 
@@
 4.14.1+BER MetaOCaml compiler variant.  All OCaml benchmarks
-were run invoking \texttt{ocamlopt}, the native code compiler for OCaml, using compiler flags \texttt{-flamda -o3}.
+were run invoking \texttt{ocamlopt}, the native code compiler for OCaml, using compiler flag \texttt{-O3}.
 

### 2025-03-18 paper/paper.tex
etna chart

@@
  First, we describe a method of optimizing generator libraries using
-\textit{multi-stage programming}, or ``staging,'' which has long been used for
-optimizing functional DSLs~\cn.
+\textit{multi-stage programming}, or ``staging,'' which has long been used to
+optimize functional DSLs~\cn.
 We apply this technique to generators, completely eliminating the overhead of
@@
 \subsection{Implementing and Testing Generators}
-\begin{figure}[h]
-  \includegraphics[scale=0.34]{allocs.png}\includegraphics[scale=0.334]{time.png}
-\end{figure}
 \label{subsection:pbt-for-pbt}
@@
 \subsection{Benchmarking speed \& resource usage}
-
+\begin{figure}[h]
+  % \includegraphics[scale=0.33]{allocs.png}
+  \includegraphics[scale=0.45]{time.png}
+  \label{fig:benchmark-time}
+  \caption{Average time taken to generate a value with the staged generator, in nanoseconds.}
+\end{figure}
 \jwc{NOTE: we should test generator speed across both languages, but speed -> bugfinding ability in only OCaml.}
@@
 \subsection{Impact on bug-finding ability}
-To answer \textbf{RQ2}, we evaluate the bug-finding ability of three of our case studies---BST, RBT, and STLC---using Etna, a platform for assessing and comparing property-based testing (PBT) techniques.
+To answer \textbf{RQ2}, we evaluate the bug-finding ability of three of our case studies---BST, RBT, and STLC---using Etna, a platform for assessing and comparing PBT techniques. 
 Etna allows users to measure the effectiveness of different generator implementations by injecting bugs into the system under test and recording the time taken for a relevant 
-property to fail in response. For example, in a tree with an ordering property, a bug injection might consist of inserting a value into the tree that violates that ordering.
-
+property to fail in response.
+For example, in a tree with an ordering property, a bug injection might consist of inserting a value into the tree that violates that ordering.
 Each case study includes a diverse set of \textit{tasks}, where a task consists of a specific bug-property pair designed to test different aspects of the system (e.g., BST
@@
 We evaluate our case studies with both type-derived and custom generators.
-Each generator has four implementations, which are equivalent up to program equality:
-\begin {itemize}
-  \item The original \bq generator (``BQ'').
-  \item A staged version of the generator (``Staged'').
-  \item A staged version of the generator using the C \rand (``Staged C'').
-  \item A staged version incorporating the CSR \rand (``Staged CSR'').
-\end{itemize}
-
+Each generator has four implementations, all of which are equivalent up to program equality: the original \bq generator (``BQ''); a staged version of the generator (``Staged''); a staged version of the generator using the C \rand (``Staged C''); and a staged version incorporating the CSR \rand (``Staged CSR'').
 To ensure a fair comparison, we initialize each of these generators with the same random seed, ensuring that they produce identical trees.
@@
 To account for these edge cases, we repeat the above process for 10 seeds and compute the geometric mean of their speedups.
-
+\begin{figure}[h]
+  \includegraphics[scale=0.333]{speedup.png}
+  \caption{}
+\end{figure}
 % To produce \todo{FIG}, we repeated the above process for 10 seeds and computed the geometric average of their speedups. 
@@
 \subsubsection{Variance between trials}
-Although the variance \textit{between} seeds in Etna can be very vast, timing results are replicable for a given seed. Across 1,000 trials on the same seed, 
+Although the variance \textit{between} seeds in Etna can be vast, timing results are replicable for a given seed. Across 1,000 trials on the same seed, 
 the observed variance in time-to-failure was less than a nanosecond. \tr{Do we need this?}

### 2025-03-18 paper/paper.tex
little progress towards WU section

--- 2025-03-18 paper/paper.tex
@@
 \subsubsection{Overhead of Choice Combinators}
-The monad functions \texttt{return} and \texttt{bind} aren't the only combinators that incur an
-abstraction overhead: all of them do! \jwc{I'm so bad at starting new sections...} ``Choice'' combinators like \texttt{weighted\_union}, however
-are particularly costly, because they unavoidably allocate a lot.
+After the \texttt{return}, \texttt{bind}, and base-type generation combinators
+(such as \texttt{gen\_int}), the most frequently used combinators are ``choice''
+combinators like \texttt{weighted\_union} which randomly select between
+sub-generators.
+
+Like \texttt{return} and \texttt{bind}, these combinators incur a performance penalty
+at run time.
 
@@
 In practice however almost~\jwc{Once again, GHC is a notable exception here \cn{}} no compilers perform this kind of optimization.
-
-% \jwc{This builds the cdf of the distribution, samples from 0 to the total, and binary searches through the array to find the bucket.}
-% \jwc{Allocating this array is unnecessary: we can simply compute the total weight, sample \texttt{x} between 0 and the total, and then walk the list accumulating the sum, until the accumulator exceeds \texttt{x}.
-Since in practice the list of possible options is quite small (usually at most 10 in practice), the linear time scan is going to be much faster.}
-% \jwc{(The point of this para will just be to emphasize that it's almost always better to be allocation-aware, instead of algorithmically clever.)}
 

### 2025-03-18 paper/paper.tex
comments about bst

@@
   \end{tabular}
-\caption{Microbenchmarks of Generators in \bq{} (using \texttt{core\_bench} \cn) and ScalaCheck (using \texttt{jmh} \cn)}
+\caption{Microbenchmarks of Generators in \bq{} (using \texttt{core\_bench} \cn) and ScalaCheck (using \texttt{jmh} \cn). Average over 10000 generations with random seeds, and a fixed size}
 \label{fig:overhead-explanation-perf}
@@
 sub-generators.
-
 Like \texttt{return} and \texttt{bind}, these combinators incur a performance penalty
-at run time.
-
-\jwc{
-Aside from the usual fact that the compiler can't ``see through'' the
-abstraction boundary to specialize, combinators like union and weighted union
-necessarily allocate a lot in the generation hot path.  Even without the dumb BQ
-array thing, to call union and weighted union, the types mean that you have to
-allocate a list.  These allocations are extremely costly, and happen every
-sample from the generator. If the generator is recursive (like the tree
-generator), they occur at each recursive call.
-}
+at run time. Aside from the previously-discussed issues that compilers
+cannot see through the abstraction boundary to optimize these programs,
+choice combinators come with their own particular abstraction overhead.
 
@@
 was itself the result of a generator: the well-typed STLC term generator used in Section~\ref{section:eval} is an example of this.
-} passed an explicitly constructed list, like
-\texttt{weighted\_union [(3,g); (1,g'); (2,g'')]}. 
+} called with an explicitly constructed list, like
+\texttt{weighted\_union [(w1,g1); (w2,g2); (w3,g3)]}. 
 This is because the most common use case for \texttt{weighted\_union} is
@@
 the options for which are always known.
-The list passed to \texttt{weighted\_union} allocated at the call site, and then \emph{immediately}
+This list is allocated at the call site, and then \emph{immediately}
 destructed by the function. In principle, since the elements of the list and its length $n$ are known statically,
 a compiler could unroll the loops in \texttt{weighted\_union} to depth $n$ and specialize the function to avoid allocating the list.
-In practice however almost~\jwc{Once again, GHC is a notable exception here \cn{}} no compilers perform this kind of optimization.
+Unfortunately, almost~\jwc{Once again, GHC is a notable exception here \cn{}} no compilers perform this kind of optimization.
+This allocation (or rather, its tendency to cause GC pauses) --- as well as the cost of running the code 
+to traverse arbitrary lists compared to unrolled loops --- has a significant impact on 
+performance.
+
+Last, many PBT libraries --- including both \bq and ScalaCheck --- implement their \texttt{weighted\_union} combinators
+in ways that are asymptotically more efficient but slower in practice than the naive algorithm.
+Weighted union works by the Fitness Proportionate Selection algorithm \cn{}, which (1) samples a number $r$
+between $0$ and the sum of the weights, and then (2) finds the first generator for which the cumulative
+weights in the list before it exceeds $r$. This second part can be accomplished in $O(\log_2 n)$ time by
+binary searching the list. However, since the lists are short in practice, a linear scan is almost always faster \cn{}.
+Moreover, both \bq{} and ScalaCheck allocate auxiliary data structures (an array in \bq{} and a BST in ScalaCheck) to
+perform this search, which incurs further run time overhead.
 

### 2025-03-18 paper/paper.tex
some wibbles

@@
 }
+\jwc{I re-explain this later somewhere i think, so we need to remember to RM that explanation.}
 
@@
 4.14.1+BER MetaOCaml compiler variant.  All OCaml benchmarks
-were run invoking \texttt{ocamlopt}, the native code compiler for OCaml, using compiler flag \texttt{-O3}.
+were compiled with \texttt{ocamlopt}, the native code compiler for OCaml, using compiler flag \texttt{-O3}.
+The microbecnhmark
 
@@
 
-\subsection{Implementing and Testing Generators}
+\subsection{Differential Testing of Generators}
 \label{subsection:pbt-for-pbt}
@@
 we conclude that the generators exhibit equivalent behavior.
-\subsection{Benchmarking speed \& resource usage}
-\begin{figure}[h]
-  % \includegraphics[scale=0.33]{allocs.png}
-  \includegraphics[scale=0.45]{time.png}
-  \label{fig:benchmark-time}
-  \caption{Average time taken to generate a value with the staged generator, in nanoseconds.}
-\end{figure}
-\jwc{NOTE: we should test generator speed across both languages, but speed -> bugfinding ability in only OCaml.}
-\jwc{
-  Baseline generators to test speed in both languages:
-  \begin{itemize}
-    \item Single int
-    \item Pair of ints, constrained
-    \item List of ints without bind (use map): lots of sampling, minimal binds.
-    \item Unableled Trees of a fixed size (no weighted union) minimal sampling, lots of binds.
-  \end{itemize}
-}
-\subsection{Impact on bug-finding ability}
-To answer \textbf{RQ2}, we evaluate the bug-finding ability of three of our case studies---BST, RBT, and STLC---using Etna, a platform for assessing and comparing PBT techniques. 
-Etna allows users to measure the effectiveness of different generator implementations by injecting bugs into the system under test and recording the time taken for a relevant 
-property to fail in response.
-For example, in a tree with an ordering property, a bug injection might consist of inserting a value into the tree that violates that ordering.
-Each case study includes a diverse set of \textit{tasks}, where a task consists of a specific bug-property pair designed to test different aspects of the system (e.g., BST
-includes tasks that test insertion, deletion, and union operations on binary search trees).
-
-We evaluate our case studies with both type-derived and custom generators.
-Each generator has four implementations, all of which are equivalent up to program equality: the original \bq generator (``BQ''); a staged version of the generator (``Staged''); a staged version of the generator using the C \rand (``Staged C''); and a staged version incorporating the CSR \rand (``Staged CSR'').
-To ensure a fair comparison, we initialize each of these generators with the same random seed, ensuring that they produce identical trees.
-
-\subsubsection{Variance between seeds}
-The time-to-failure of a given strategy on a given seed is not necessarily representative of 
-the strategy's average time-to-fail over a large number of trials, so we normalize by computing the
-timing \textit{differential} between optimized and unoptimized generators (the ``speedup''). While this works in most cases,
-it is not perfect.
-For example, it is theoretically possible
-to choose a seed such that the \textit{first} value produced by a generator discovers the bug, which would elide any differential speedup (and likewise
-if both generators fail to find the bug).
-To account for these edge cases, we repeat the above process for 10 seeds and compute the geometric mean of their speedups.
-\begin{figure}[h]
-  \includegraphics[scale=0.333]{speedup.png}
-  \caption{}
-\end{figure}
-% To produce \todo{FIG}, we repeated the above process for 10 seeds and computed the geometric average of their speedups. 
-%This is important because the time-to-failure of a given strategy on a given seed is not necessarily 
-%representative of the strategy's overall bug-finding ability, although it often is. For example, it is theoretically possible 
-%to choose a seed such that the \textit{first} value produced by a generator discovers the bug, but it would be misleading to report this result.
-
-\subsubsection{Variance between trials}
-Although the variance \textit{between} seeds in Etna can be vast, timing results are replicable for a given seed. Across 1,000 trials on the same seed, 
-the observed variance in time-to-failure was less than a nanosecond. \tr{Do we need this?}
-
-\section{Other Languages and Libraries}
-\label{subsection:other-langs}
+\jwc{A little more here}
+
+\subsection{Benchmarking Generator Speed}
+\jwc{Intro stuff}
 
@@
 
-\jwc{TODO: Move this section later (cf conversation in WH meeting 3/7/25)}
+\begin{figure}[h]
+  % \includegraphics[scale=0.33]{allocs.png}
+  \includegraphics[scale=0.45]{time.png}
+  \label{fig:benchmark-time}
+  \caption{Average time taken to generate a value with the staged generator, in nanoseconds.}
+\end{figure}
+\jwc{NOTE: we should test generator speed across both languages, but speed -> bugfinding ability in only OCaml.}
+\jwc{
+  Baseline generators to test speed in both languages:
+  \begin{itemize}
+    \item Single int
+    \item Pair of ints, constrained
+    \item List of ints without bind (use map): lots of sampling, minimal binds.
+    \item Unableled Trees of a fixed size (no weighted union) minimal sampling, lots of binds.
+  \end{itemize}
+}
+\subsection{Benchmarking Bug-Finding Ability}
+To answer \textbf{RQ2}, we evaluate the bug-finding ability of three of our case studies---BST, RBT, and STLC---using Etna, a platform for assessing and comparing PBT techniques. 
+Etna allows users to measure the effectiveness of different generator implementations by injecting bugs into the system under test and recording the time taken for a relevant 
+property to fail in response.
+For example, in a tree with an ordering property, a bug injection might consist of inserting a value into the tree that violates that ordering.
+Each case study includes a diverse set of \textit{tasks}, where a task consists of a specific bug-property pair designed to test different aspects of the system (e.g., BST
+includes tasks that test insertion, deletion, and union operations on binary search trees).
+
+We evaluate our case studies with both type-derived and custom generators.
+Each generator has four implementations, all of which are equivalent up to program equality: the original \bq generator (``BQ''); a staged version of the generator (``Staged''); a staged version of the generator using the C \rand (``Staged C''); and a staged version incorporating the CSR \rand (``Staged CSR'').
+To ensure a fair comparison, we initialize each of these generators with the same random seed, ensuring that they produce identical trees.
+
+\subsubsection{Variance between seeds}
+The time-to-failure of a given strategy on a given seed is not necessarily representative of 
+the strategy's average time-to-fail over a large number of trials, so we normalize by computing the
+timing \textit{differential} between optimized and unoptimized generators (the ``speedup''). While this works in most cases,
+it is not perfect.
+For example, it is theoretically possible
+to choose a seed such that the \textit{first} value produced by a generator discovers the bug, which would elide any differential speedup (and likewise
+if both generators fail to find the bug).
+To account for these edge cases, we repeat the above process for 10 seeds and compute the geometric mean of their speedups.
+\begin{figure}[h]
+  \includegraphics[scale=0.333]{speedup.png}
+  \caption{}
+\end{figure}
+% To produce \todo{FIG}, we repeated the above process for 10 seeds and computed the geometric average of their speedups. 
+%This is important because the time-to-failure of a given strategy on a given seed is not necessarily 
+%representative of the strategy's overall bug-finding ability, although it often is. For example, it is theoretically possible 
+%to choose a seed such that the \textit{first} value produced by a generator discovers the bug, but it would be misleading to report this result.
+
+\subsubsection{Variance between trials}
+Although the variance \textit{between} seeds in Etna can be vast, timing results are replicable for a given seed. Across 1,000 trials on the same seed, 
+the observed variance in time-to-failure was less than a nanosecond. \tr{Do we need this?}
+
+\section{Other Languages and Libraries}
+\label{subsection:other-langs}
 
@@
   \begin{itemize}
-    \item Scala.
-Functional abstractions like QC generators are known to be costly in Scala, that's why they have LMS (in Scala 2, and Macros in Scala 3).
-Example: parser combinators (``On Staged Parser Combinators for Efficient Data Processing''), functional data structures (\href{https://ppl.stanford.edu/papers/popl13_rompf.pdf}{Link}), web programming (``Efficient High-Level Abstractions for Web Programming'').
-    Al of this should still work in scala.
-Could easily be incorporated into ScalaCheck, with minor modification: ScalaCheck uses a state monad to thread around the seed, instead of a stateful one (like BQ), or a splittable one (like Haskell).
-So you have to adapt to that.
-But same diff.
     \item Python. Maybe could do it?? Hypothesis + MacroPy

### 2025-03-18 paper/paper.tex
lol

@@
 \label{section:eval}
+
+\jwc{Initial section here about eval}
+
 \begin{itemize}
-  \item \textbf{RQ1}: Do generators written using \name run faster than those written using base quickcheck?
+  \item \textbf{RQ1}: Do generators written using \name run faster than those written with regular generator combinators?
   \item \textbf{RQ2}: Do performance speedups translate to better bug-finding ability?
-  \item \textbf{RQ3}: Does our technique generalize to other strict functional languages?
+  % \item \textbf{RQ3}: Does our technique generalize to other strict functional languages?
 \end{itemize}

### 2025-03-19 paper/paper.tex

@@
 All experiments were run on a 64-bit Linux machine with 264 GB RAM and an Intel Xeon Platinum 8375C CPU, running Ubuntu 24.04.1 LTS.
-The \name library was implemented using the
-4.14.1+BER MetaOCaml compiler variant.  All OCaml benchmarks
-were compiled with \texttt{ocamlopt}, the native code compiler for OCaml, using compiler flag \texttt{-O3}.
-The microbecnhmark
-
-. \todo{Joe : Scala info ...}
+
+The OCaml version of \name and all OCaml benchmarks
+were compiled with 4.14.1+BER MetaOCaml \texttt{ocamlopt}, the native code compiler for OCaml, using compiler flag \texttt{-O3}.
+The baseline OCaml generators were written with \bq{} 0.16.
+The Scala 3 version of \name and all Scala benchmarks
+were run on Scala 3.6.3
+and OpenJDK 21.0.6, using ScalaCheck version 1.17 as the baseline.
+We used \texttt{core\_bench} 0.16 \cn in OCaml and \texttt{jmh} 0.4.7 \cn for generator performance microbenchmarking.
 

### 2025-03-19 paper/paper.tex
SWARM
more words

@@
 This design is less obviously correct, and does require some care. Rather than \texttt{bind} ensuring correct evaluation order once and for all,
-individual combinators must be carefully written to ensure that \texttt{'a code} values that contian effects are \texttt{let\_insert}'d.
-In Section~\ref{subsection:pbt-for-pbt}, we discuss how we use PBT to ensure that the OCaml is written correctly.
+individual combinators must be carefully written to ensure that \texttt{'a code} values that contian effects are \texttt{let\_insert}'d.\footnote{To ensure that the OCaml code is written correctly, we employ a PBT harness to compare its outputs to the corresponding \bq generator over 1,000 random seeds.
+If no discrepancy arises, we conclude that the generators exhibit equivalent behavior.}
 
@@
 
-\subsection{Differential Testing of Generators}
-\label{subsection:pbt-for-pbt}
-To establish the equivalence of generators up to program equality, 
-we employ a PBT harness to compare their outputs over 1,000 random seeds. 
-If a discrepancy arises, we report the unequal trees; if no counterexample is found,
-we conclude that the generators exhibit equivalent behavior.
-\jwc{A little more here}
-
 \subsection{Benchmarking Generator Speed}
@@
 \subsection{Benchmarking Bug-Finding Ability}
-To answer \textbf{RQ2}, we evaluate the bug-finding ability of three of our case studies---BST, RBT, and STLC---using Etna, a platform for assessing and comparing PBT techniques. 
+To answer \textbf{RQ2}, we evaluate the bug-finding ability of our BST and STLC case studies using Etna, a platform for assessing and comparing PBT techniques. 
 Etna allows users to measure the effectiveness of different generator implementations by injecting bugs into the system under test and recording the time taken for a relevant 
@@
 
-We evaluate our case studies with both type-derived and custom generators.
-Each generator has four implementations, all of which are equivalent up to program equality: the original \bq generator (``BQ''); a staged version of the generator (``Staged''); a staged version of the generator using the C \rand (``Staged C''); and a staged version incorporating the CSR \rand (``Staged CSR'').
+We test both type-derived and custom generators for BST and STLC.
+Each generator has four implementations (``strategies''), all equivalent up to program equality: the original \bq generator (``BQ''); a staged version of the generator (``Staged''); and a staged version incorporating the CSR \rand (``Staged CSR'').
 To ensure a fair comparison, we initialize each of these generators with the same random seed, ensuring that they produce identical trees.
 
-\subsubsection{Variance between seeds}
-The time-to-failure of a given strategy on a given seed is not necessarily representative of 
-the strategy's average time-to-fail over a large number of trials, so we normalize by computing the
+We evaluate each strategy on every task with a 60-second timeout. 
+If a bug is not found within this limit, the task is considered unsolved. 
+Tasks where all strategies fail to find the bug are excluded from our dataset, 
+as they provide no basis for comparison. 
+Similarly, we exclude tasks where the BQ strategy completes in under 5ms,
+ as such negligible runtimes do not yield meaningful insights into relative performance. These filters ensure that the reported speedups are based on cases where optimization has a measurable impact.
+
+It is worth noting that the time-to-failure of a given strategy on a given seed is not necessarily representative of 
+the strategy's average time-to-failure over a large number of trials. We normalize by computing the
 timing \textit{differential} between optimized and unoptimized generators (the ``speedup''). While this works in most cases,
@@
 if both generators fail to find the bug).
-To account for these edge cases, we repeat the above process for 10 seeds and compute the geometric mean of their speedups.
+To account for these edge cases, we repeat the above process for 10 seeds and compute the geometric mean of the speedups.\footnote{Although the variance \textit{between} seeds in Etna can be vast, timing results are replicable for a \textit{given} seed.
+Across 1,000 trials using the same seed,
+the observed variance in time-to-failure was less than a nanosecond.}
+
 \begin{figure}[h]
   \includegraphics[scale=0.333]{speedup.png}
-  \caption{}
+  \caption{Geometric average of each task's speedup over case study and strategies, showing that staging leads to better bug-finding ability across the board.
+As in \todo{earlier figure}, benchmarks with higher numbers of binds (BST (Single-Pass), STLC) see more benefit from staging, whereas benchmarks with more sampling (BST) see a greater speedup from the CSR \rand.}
+  \label{fig:etna}
 \end{figure}
-% To produce \todo{FIG}, we repeated the above process for 10 seeds and computed the geometric average of their speedups. 
-%This is important because the time-to-failure of a given strategy on a given seed is not necessarily 
-%representative of the strategy's overall bug-finding ability, although it often is. For example, it is theoretically possible 
-%to choose a seed such that the \textit{first} value produced by a generator discovers the bug, but it would be misleading to report this result.
-
-\subsubsection{Variance between trials}
-Although the variance \textit{between} seeds in Etna can be vast, timing results are replicable for a given seed. Across 1,000 trials on the same seed, 
-the observed variance in time-to-failure was less than a nanosecond. \tr{Do we need this?}
+\begin{figure}[h]
+  \includegraphics[scale=0.437]{swarm.png}
+  \label{fig:swarm}
+  \caption{Swarm plot of individual speedups across different staging strategies and benchmarks.
+Each dot represents a speedup measurement, with larger spreads indicating greater variability (omitted for some strategies where all values cluster very close to the mean).
+Extreme outliers correspond to large speedups.}
+\end{figure}
 
@@
 \label{subsection:other-langs}
-
-\tr{Here's where we talk about what's going on in the world, and ultimately make the argument that \bq{} is the best tool to reproduce.}
 

### 2025-03-19 paper/paper.tex
i love the swarm but it's questionable

@@
 \begin{figure}[h]
-  \includegraphics[scale=0.437]{swarm.png}
+  \includegraphics[scale=0.355]{hist.png}
   \label{fig:swarm}
-  \caption{Swarm plot of individual speedups across different staging strategies and benchmarks.
-Each dot represents a speedup measurement, with larger spreads indicating greater variability (omitted for some strategies where all values cluster very close to the mean).
-Extreme outliers correspond to large speedups.}
+  \caption{Histogram of individual-task speedups across different staging strategies and benchmarks. Extreme outliers correspond to large speedups.}
 \end{figure}

### 2025-03-19 paper/paper.tex
minor

@@
 
-\jwc{Initial section here about eval}
+\jwc{Initial section here about eval}\tr{todo}
 
 \begin{itemize}
-  \item \textbf{RQ1}: Do generators written using \name run faster than those written with regular generator combinators?
+  \item \textbf{RQ1}: Do generators written using our technique run faster than those written with regular generator combinators?
   \item \textbf{RQ2}: Do performance speedups translate to better bug-finding ability?
@@
 \subsection{Benchmarking Generator Speed}
-\jwc{Intro stuff}
+
+\begin{figure}[h]
+  % \includegraphics[scale=0.33]{allocs.png}
+  \includegraphics[scale=0.45]{time.png}
+  \label{fig:time}
+  \caption{Average generation time over 10,000 trials. Across all benchmarks, we see reliable improvement from at least one of Staged SR and Staged CSR.}
+\end{figure}
 
@@
   \end{tabular}
- \caption{\todo{Understand Joe's implementation of this...}}
- \label{fig:benchmark-table}
+ \caption{Benchmarks reimplemented in ScalaCheck show similar performance improvements to \name.}
+ \label{fig:scala}
  \end{figure}
 
-\begin{figure}[h]
-  % \includegraphics[scale=0.33]{allocs.png}
-  \includegraphics[scale=0.45]{time.png}
-  \label{fig:benchmark-time}
-  \caption{Average time taken to generate a value with the staged generator, in nanoseconds.}
-\end{figure}
-\jwc{NOTE: we should test generator speed across both languages, but speed -> bugfinding ability in only OCaml.}
-\jwc{
-  Baseline generators to test speed in both languages:
-  \begin{itemize}
-    \item Single int
-    \item Pair of ints, constrained
-    \item List of ints without bind (use map): lots of sampling, minimal binds.
-    \item Unableled Trees of a fixed size (no weighted union) minimal sampling, lots of binds.
-  \end{itemize}
-}
+%\jwc{NOTE: we should test generator speed across both languages, but speed -> bugfinding ability in only OCaml.}
+%\jwc{
+%  Baseline generators to test speed in both languages:
+%  \begin{itemize}
+%    \item Single int
+%    \item Pair of ints, constrained
+%    \item List of ints without bind (use map): lots of sampling, minimal binds.
+%    \item Unableled Trees of a fixed size (no weighted union) minimal sampling, lots of binds.
+%  \end{itemize}
+%}
 \subsection{Benchmarking Bug-Finding Ability}
@@
 Each case study includes a diverse set of \textit{tasks}, where a task consists of a specific bug-property pair designed to test different aspects of the system (e.g., BST
-includes tasks that test insertion, deletion, and union operations on binary search trees).
+includes tasks that test insertion, deletion, and union operations on binary search trees). BST has 43 tasks; STLC has 18.
 
@@
 if both generators fail to find the bug).
-To account for these edge cases, we repeat the above process for 10 seeds and compute the geometric mean of the speedups.\footnote{Although the variance \textit{between} seeds in Etna can be vast, timing results are replicable for a \textit{given} seed.
+To account for these edge cases, we repeat the above process for 10 seeds and compute the geometric average of the speedups.\footnote{Although the variance \textit{between} seeds in Etna can be vast, timing results are replicable for a \textit{given} seed.
 Across 1,000 trials using the same seed,
@@
   \includegraphics[scale=0.333]{speedup.png}
-  \caption{Geometric average of each task's speedup over case study and strategies, showing that staging leads to better bug-finding ability across the board.
+  \caption{Geometric average of each task's speedup over case studies and strategies, showing that staging leads to better bug-finding ability across the board.
 As in \todo{earlier figure}, benchmarks with higher numbers of binds (BST (Single-Pass), STLC) see more benefit from staging, whereas benchmarks with more sampling (BST) see a greater speedup from the CSR \rand.}
@@
 \end{figure}
+
 \begin{figure}[h]

### 2025-03-19 paper/paper.tex
minor
pushed

--- 2025-03-19 paper/paper.tex
@@
 \subsection{Background: Monadic Generator DSLs}
-\jwc{Usual spiel about pbt and generators: here's a monadic generic generator library (but we're using BQ to illustrage), it has return and bind, this is what they're for.}
-\jwc{You often want generators to generate for sparse precoditions so you can exercise a property reasonably well}
-
+\jwc{Usual spiel about pbt and generators: here's a monadic generic generator
+library (but we're using BQ to illustrage), it has return and bind, this is what
+they're for.}
+\jwc{You often want generators to generate for sparse precoditions
+so you can exercise a property reasonably well}
+
+Property Based Testing was popularized by the QuickCheck library \cn{} in
+Haskell.  A core element of PBT as introduced in that paper is \emph{monadic
+generators}, an embedded DSL for writing the random data generators that drive
+the PBT process.  This DSL gives developers significant control over exactly
+what sorts of values are generated and what their distribution is. Both of these
+axes of control are important to adequately exercise the system under test and
+find bugs.  Indeed, the PBT literature \cn{} has repatedly demonstrated that the
+particular generator used to test a property has an enormous impact on how well
+PBT actually works \jwc{less derogatory phrasing}, and much research has gone
+into finding clever ways to generate data \cn{}.
+
+\begin{figure}
 \begin{lstlisting}
@@
   val bind : 'a t -> ('a -> 'b t) -> 'b t
-  val gen_int : lo:int -> hi:int -> int t
+  val gen_int : int -> int -> int t
 
@@
 \end{lstlisting}
+\caption{API of the \bq{} Generator DSL}
+\label{fig:bq-api}
+\end{figure}
+
+A standard generator DSL consists of a monadic interface, plus combinators for sampling random numbers, making (weighted) random choices,
+and controlling the size of generated values. Figure~\ref{fig:bq-api} shows a selection of functions from the API of th \bq{} generator
+DSL.
+
+Because we have to choose \emph{some} syntax for this paper, we use \bq{}'s as our primary running generator DSL. However,
+this is a standin for a generic generator DSL, since they basicalyl all look like this one.
+
+The ``monadic interface'' of a generator DSL consists of \texttt{return} and \texttt{bind}.
+The generator \texttt{return x} is the constant generator, always generating the value \texttt{x}.
+Running a generator \texttt{bind g k} runs the generator \texttt{g}, producing a value \texttt{a},
+and then runs the generator \texttt{k a}. Monadic \texttt{bind} allows for dependencies.
+
+The \texttt{gen\_int a b} generator samples an integer uniformly in the inclusive range from \texttt{a} to \texttt{b}.
 
@@
 
-\jwc{Here's an example of a generator: you can sample from it and it gives you pairs of ints, one less than the other.}
-
-\begin{lstlisting}
-let int_pair : (int * int) Bq.t =
-  let%bind x = Bq.gen_int 0 100 in
-  let%bind y = Bq.gen_int 0 x in
-  return (x,y)
-\end{lstlisting}
-
-\jwc{This desugars to...}
+Figure ... shows an example of a generator written with these combinators. It first samples an int between $0$ and $100$, 
+names it $x$, samples another between $0$ an $x$, and then returns the pair of them both.
+Most languages in which monadic APIs are common expose some sort of sytnactic sugar for them --- this is true of OCaml and Scala. 
+OCaml's looks like \texttt{let\%bind x = e in e'}, which desguars to \texttt{...}. Figure~\jwc{...} also shows this generator written with OCaml's syntacitc sugar\footnote{It's actually Jane's}.
+
+\begin{figure}
 
@@
 \end{lstlisting}
+
+\begin{lstlisting}
+let int_pair : (int * int) Bq.t =
+  let%bind x = Bq.gen_int 0 100 in
+  let%bind y = Bq.gen_int 0 x in
+  return (x,y)
+\end{lstlisting}
+\end{figure}
 
@@
 This design is less obviously correct, and does require some care. Rather than \texttt{bind} ensuring correct evaluation order once and for all,
-individual combinators must be carefully written to ensure that \texttt{'a code} values that contian effects are \texttt{let\_insert}'d.\footnote{To ensure that the OCaml code is written correctly, we employ a PBT harness to compare its outputs to the corresponding \bq generator over 1,000 random seeds.
-If no discrepancy arises, we conclude that the generators exhibit equivalent behavior.}
+individual combinators must be carefully written to ensure that \texttt{'a code} values that contian effects are \texttt{let\_insert}'d.
+In Section~\ref{subsection:pbt-for-pbt}, we discuss how we use PBT to ensure that the OCaml is written correctly.
 
@@
 
+\subsection{Differential Testing of Generators}
+\label{subsection:pbt-for-pbt}
+To establish the equivalence of generators up to program equality, 
+we employ a PBT harness to compare their outputs over 1,000 random seeds. 
+If a discrepancy arises, we report the unequal trees; if no counterexample is found,
+we conclude that the generators exhibit equivalent behavior.
+\jwc{A little more here}
+
 \subsection{Benchmarking Generator Speed}
@@
 \subsection{Benchmarking Bug-Finding Ability}
-To answer \textbf{RQ2}, we evaluate the bug-finding ability of our BST and STLC case studies using Etna, a platform for assessing and comparing PBT techniques. 
+To answer \textbf{RQ2}, we evaluate the bug-finding ability of three of our case studies---BST, RBT, and STLC---using Etna, a platform for assessing and comparing PBT techniques. 
 Etna allows users to measure the effectiveness of different generator implementations by injecting bugs into the system under test and recording the time taken for a relevant 
@@
 
-We test both type-derived and custom generators for BST and STLC.
-Each generator has four implementations (``strategies''), all equivalent up to program equality: the original \bq generator (``BQ''); a staged version of the generator (``Staged''); and a staged version incorporating the CSR \rand (``Staged CSR'').
+We evaluate our case studies with both type-derived and custom generators.
+Each generator has four implementations, all of which are equivalent up to program equality: the original \bq generator (``BQ''); a staged version of the generator (``Staged''); a staged version of the generator using the C \rand (``Staged C''); and a staged version incorporating the CSR \rand (``Staged CSR'').
 To ensure a fair comparison, we initialize each of these generators with the same random seed, ensuring that they produce identical trees.
 
-We evaluate each strategy on every task with a 60-second timeout. 
-If a bug is not found within this limit, the task is considered unsolved. 
-Tasks where all strategies fail to find the bug are excluded from our dataset, 
-as they provide no basis for comparison. 
-Similarly, we exclude tasks where the BQ strategy completes in under 5ms,
- as such negligible runtimes do not yield meaningful insights into relative performance. These filters ensure that the reported speedups are based on cases where optimization has a measurable impact.
-
-It is worth noting that the time-to-failure of a given strategy on a given seed is not necessarily representative of 
-the strategy's average time-to-failure over a large number of trials. We normalize by computing the
+\subsubsection{Variance between seeds}
+The time-to-failure of a given strategy on a given seed is not necessarily representative of 
+the strategy's average time-to-fail over a large number of trials, so we normalize by computing the
 timing \textit{differential} between optimized and unoptimized generators (the ``speedup''). While this works in most cases,
@@
 if both generators fail to find the bug).
-To account for these edge cases, we repeat the above process for 10 seeds and compute the geometric mean of the speedups.\footnote{Although the variance \textit{between} seeds in Etna can be vast, timing results are replicable for a \textit{given} seed.
-Across 1,000 trials using the same seed,
-the observed variance in time-to-failure was less than a nanosecond.}
-
+To account for these edge cases, we repeat the above process for 10 seeds and compute the geometric mean of their speedups.
 \begin{figure}[h]
   \includegraphics[scale=0.333]{speedup.png}
-  \caption{Geometric average of each task's speedup over case study and strategies, showing that staging leads to better bug-finding ability across the board.
-As in \todo{earlier figure}, benchmarks with higher numbers of binds (BST (Single-Pass), STLC) see more benefit from staging, whereas benchmarks with more sampling (BST) see a greater speedup from the CSR \rand.}
-  \label{fig:etna}
+  \caption{}
 \end{figure}
-\begin{figure}[h]
-  \includegraphics[scale=0.355]{hist.png}
-  \label{fig:swarm}
-  \caption{Histogram of individual-task speedups across different staging strategies and benchmarks. Extreme outliers correspond to large speedups.}
-\end{figure}
+% To produce \todo{FIG}, we repeated the above process for 10 seeds and computed the geometric average of their speedups. 
+%This is important because the time-to-failure of a given strategy on a given seed is not necessarily 
+%representative of the strategy's overall bug-finding ability, although it often is. For example, it is theoretically possible 
+%to choose a seed such that the \textit{first} value produced by a generator discovers the bug, but it would be misleading to report this result.
+
+\subsubsection{Variance between trials}
+Although the variance \textit{between} seeds in Etna can be vast, timing results are replicable for a given seed. Across 1,000 trials on the same seed, 
+the observed variance in time-to-failure was less than a nanosecond. \tr{Do we need this?}
 
@@
 \label{subsection:other-langs}
+
+\tr{Here's where we talk about what's going on in the world, and ultimately make the argument that \bq{} is the best tool to reproduce.}
 

### 2025-03-19 paper/paper.tex

--- 2025-03-19 paper/paper.tex
@@
 It is worth noting that the time-to-failure of a given strategy on a given seed is not necessarily representative of 
-the strategy's average time-to-failure over a large number of trials. We normalize by computing the
-timing \textit{differential} between optimized and unoptimized generators (the ``speedup''). While this works in most cases,
+the strategy's average time-to-failure over a large number of trials.
+We normalize by computing the relative performance of \name generators vs.
+\bq generators (the ``speedup'').
+While this works in most cases,
 it is not perfect.

### 2025-03-19 paper/paper.tex
we love the swarm plot

@@
   \includegraphics[scale=0.333]{speedup.png}
-  \caption{Geometric average of each task's speedup over case studies and strategies, showing that staging leads to better bug-finding ability across the board.
+  \caption{Geometric average of each task's speedup, showing that staging leads to better bug-finding ability across the board.
 As in \todo{earlier figure}, benchmarks with higher numbers of binds (BST (Single-Pass), STLC) see more benefit from staging, whereas benchmarks with more sampling (BST) see a greater speedup from the CSR \rand.}
@@
 \begin{figure}[h]
-  \includegraphics[scale=0.355]{hist.png}
+  \includegraphics[scale=0.355]{swarm.png}
   \label{fig:swarm}
@@
 \section{Related Work}
-\jwc{A generator is a parser of randomness, and people have implemented lots of staged parser libraries!}
+
+\subsection{Speeding up Property-Based Testing}
+\hg{\begin{itemize}
+  \item Changing the generation order~\cite{runcimanSmallcheckLazySmallcheck2008,braquehaisSimpleIncrementalDevelopment2017}
+  \item Dealing with filtering~\cite{claessenGeneratingConstrainedRandom2015}
+  \item QuickerCheck~\cite{krookQuickerCheckImplementingEvaluating2024}
+  \item Evaluation of PBT generation~\cite{shiEtnaEvaluationPlatform2023}
+\end{itemize}}
+
+\subsection{Staging Monadic Libraries}
+\jwc{A generator is a parser of
+randomness~\cite{goldsteinParsingRandomness2022}, and people have implemented
+lots of staged parser libraries!}
 
@@
 \bibliographystyle{ACM-Reference-Format}
-\bibliography{bib}
+\bibliography{bib,harry}
 

### 2025-03-19 paper/paper.tex
Working on related work
swarm is back baby

@@
   \includegraphics[scale=0.333]{speedup.png}
-  \caption{Geometric average of each task's speedup, showing that staging leads to better bug-finding ability across the board.
-As in \todo{earlier figure}, benchmarks with higher numbers of binds (BST (Single-Pass), STLC) see more benefit from staging, whereas benchmarks with more sampling (BST) see a greater speedup from the CSR \rand.}
+  \caption{Geometric average of all speedups for each strategy and benchmark, showing that staging leads to better bug-finding ability across the board.
+As in \todo{earlier figure}, benchmarks with higher numbers of binds (BST (Single-Pass); STLC) see more benefit from staging, whereas benchmarks with more sampling (BST) see a greater speedup from the CSR \rand.}
   \label{fig:etna}
@@
 \begin{figure}[h]
-  \includegraphics[scale=0.355]{swarm.png}
+  \includegraphics[scale=0.41]{swarm.png}
   \label{fig:swarm}
-  \caption{Histogram of individual-task speedups across different staging strategies and benchmarks. Extreme outliers correspond to large speedups.}
+  \caption{Distribution of individual-task speedups across strategies and benchmarks. Extreme outliers correspond to large speedups.}
 \end{figure}

### 2025-03-19 paper/paper.tex
other langs and libs

--- 2025-03-19 paper/paper.tex
@@
 
-\jwc{Note that this explanation require some prior note of exactly *why* BQ is slow... i.e. the abstraction overhead of the library is high.}
+In the years since the original QuickCheck paper \cn{}, PBT has had remarkable success
+in languages outside of
+
+Of course, users and developers of PBT libraries in \emph{other} languages might be curious about how
+the techniques brought to bear on OCaml's \bq{} and Scala's ScalaCheck in this paper might be imported to their favorite language.
+
+Indeed, while we have done our best to keep the proceeding discussion as
+language-agnostic as possible, our claims about the \name{} technique have only
+been thoroughly tested in OCaml and Scala 3, and we are only claiming (in the
+scientific sense) that we have contributed a technique applicable to PBT
+libraries in strict functional programming languages.
+
+In this section
 
@@
   \begin{itemize}
-    \item Python. Maybe could do it?? Hypothesis + MacroPy
-    \item Haskell: GHC does a lot of these optimizations already, since the code is pure. Since QC generators are relatively small programs,
+    \item Racket is the only ``real'' strict functional language we haven't tested this in. It'd definitely work.
+    \item Haskell GHC does a lot of these optimizations already, since the code is pure. Since QC generators are relatively small programs,
     GHC has little trouble specializing them. Of course, this is not guaranteed. A version of this idea can easily be ported to the original QC with template haskell, to guarantee
-    the highest-performance generators.
-    \item Rust: Not GC'd, so no alloc overhead but bind'd generators still dispatch through run-time data.
+    the highest-performance generators. But the benefit would be lower since GHC is smarter. Worth doing, though.
+    \item Rust: generator libraries are not monadic, presumably to avoid exactly these issues.
+    \item Python. Maybe could do it?? Hypothesis is the most used PBT library out there so this could be high impact.
   \end{itemize}

### 2025-03-19 paper/paper.tex
progress

@@
 One of the primary uses of staging is in embedded DSL construction \cite{sheard99, lms, trattdsl}.
+\jwc{Cite racket macros for DSLs here:}
 While embedded DSLs are great \jwc{words}, they have well-known drawback: 
 the functional abstractions used to build eDSLs prevent compilers from generating efficient machine code \cn.
+\jwc{Do racket people also use macros to eliminate run time overhead? Or is it just for extensibility?}
 As alluded to previously \jwc{This sentence here because of the section-ordering transpose...
@@
 and query processing \cite{rhyme}.
+\jwc{Cite racket things here}
 
 Many languages have some degree of staging functionality, though support varies
-from full native support (Scala 3 \cite{scalamacros}, Haskell \cite{templatehaskell}) to compiler extensions (OCaml \cite{metaocaml,macocaml}, Java \cite{mint}).
+from full native support (Scala 3 \cite{scalamacros}, Haskell \cite{templatehaskell}, Racket \cn{}) to compiler extensions (OCaml \cite{metaocaml,macocaml}, Java \cite{mint}).
 Moreover, many more languages have lighter-weight staging-like metaprogramming support in the form of macros.
@@
 \subsection{Staged Type-Derived Generators}
-
+\label{subsection:type-derived}
 \jwc{
@@
 \subsection{Benchmarking Generator Speed}
+To answer \textbf{RQ1}, we implement benchmarks in both \name and \bq. We vary the choice of \rand in our \name generators, 
+using both \bq's default \texttt{splittable\_random} (``SR'') and our improved SplitMix (``CSR''), as discussed in \secref{subsection:faster-rng}.
+
+Our benchmarks consist of lists, binary search trees (BST), and simply typed lambda calculus (STLC) terms. We implement generators for these benchmarks using a variety of 
+\textit{strategies}, which vary in structure and sophistication. These include: a derived generators for BSTs and STLC terms, using the approach outlined in \secref{subsection:type-derived}; 
+two custom BST generators, one that repeatedly inserts values into an initially empty tree and one that materializes the tree in a single pass by generating keys, values, and subtrees at each step;
+a custom, correct by construction generator for STLC, which is the most complex; and a generator of boolean lists.
 
@@
   \label{fig:time}
-  \caption{Average generation time over 10,000 trials. Across all benchmarks, we see reliable improvement from at least one of Staged SR and Staged CSR.}
+  \caption{Average generation time over 10,000 trials.}
 \end{figure}
@@
 It is worth noting that the time-to-failure of a given strategy on a given seed is not necessarily representative of 
-the strategy's average time-to-failure over a large number of trials.
-We normalize by computing the relative performance of \name generators vs.
-\bq generators (the ``speedup'').
-While this works in most cases,
+the strategy's average time-to-failure over a large number of trials. We normalize by computing the \textit{relative} performance (the ``speedup''). While this works in most cases,
 it is not perfect.
@@
 In the years since the original QuickCheck paper \cn{}, PBT has had remarkable success
-in languages outside of
-
-Of course, users and developers of PBT libraries in \emph{other} languages might be curious about how
-the techniques brought to bear on OCaml's \bq{} and Scala's ScalaCheck in this paper might be imported to their favorite language.
-
-Indeed, while we have done our best to keep the proceeding discussion as
-language-agnostic as possible, our claims about the \name{} technique have only
-been thoroughly tested in OCaml and Scala 3, and we are only claiming (in the
-scientific sense) that we have contributed a technique applicable to PBT
-libraries in strict functional programming languages.
-
-In this section
+in languages outside of the small ICFP world from which it came.
+For this reason, readers who are users and developers of PBT libraries in \emph{other} languages ---
+including non-strict or procedural ones --- might be curious about how the
+techniques brought to bear on OCaml's \bq{} and Scala's ScalaCheck in this paper
+might be imported to their favorite language.
+Below we discuss how \name{} could be applied in other languages.
+
+% While we have done our best to keep the proceeding discussion as
+% language-agnostic as possible, our claims about the \name{} technique have only
+% been thoroughly tested in OCaml and Scala 3, and we are only claiming (in the
+% scientific sense) that we have contributed a technique applicable to PBT
+% libraries in strict functional programming languages.
+
+\begin{itemize}
+  \item \textbf{Racket} is the next target that we intend to test the \name{} methodology on. The entire Racket philosophy is intertwined with
+  using macros to build small eDSLs like the ones we use to write generators, and so it is a natural fit.
+  We have not yet tested the technique in Racket for the simple reason that the authorship team of this paper does not posess
+  much Racket expertise \jwc{can we say this without unblinding?}.
+
+  \item \textbf{Haskell} birthed PBT, and so it is natural to ask why we have not yet applied \name{} to it. The basic answer is that
+  because most tasks in Haskell are accomplished with monads, GHC is designed specially to eliminate the run-time overhead of monadic
+  abstractions! Indeed, because Haskell is pure, GHC can aggressively inline and beta-reduce
+  nearly anything as part of its normal optimiation steps.
+  For complex enough programs however, the heuristics GHC uses can degrade, leading to performance overheads for monadic code.
+  For this reason, Haskell does have a multi-stage programming system called Template Haskell \cn{}, which is sometimes 
+  to build eDSLs in contexts where one does not want to rely on the optimizer's heuristics.
+  \item \textbf{Rust:} asdf
+  \item \textbf{Python:} asdf
+\end{itemize}
 
@@
 
-\subsection{Staging Monadic Libraries}
+\subsection{Staging eDSLs}
 \jwc{A generator is a parser of
@@
 lots of staged parser libraries!}
+\jwc{cite Lionel Parreaux  work, he's on PC}
 
@@
 \jwc{Why we stopped where we stopped: can you squeeze more out of this?}
+\jwc{
+  \begin{itemize}
+    \item On the staging front, we could do a little with more clever staging to codegen better...
+Avoided using floating let-insertion to ensure we had precise control over effect ordering, but we could try to relax that.
+          Recursive generators have some extra overhead for packing/unpacking accumulators, but could do a GADT cf \texttt{SOP} in 2ltt.
+    \item On the RNG front, there's not much more we can squeeze out of the RNG without breakning equiv with splitmix.
+As discussed, there are much faster things you can do as alterative sources of entropy.
+  \end{itemize}
+}
 

### 2025-03-19 paper/paper.tex
blah, some langauges and libraries

@@
 \subsection{Staged Type-Derived Generators}
-\label{subsection:type-derived}
+
 \jwc{
@@
 \subsection{Benchmarking Generator Speed}
-To answer \textbf{RQ1}, we implement benchmarks in both \name and \bq. We vary the choice of \rand in our \name generators, 
-using both \bq's default \texttt{splittable\_random} (``SR'') and our improved SplitMix (``CSR''), as discussed in \secref{subsection:faster-rng}.
-
-Our benchmarks consist of lists, binary search trees (BST), and simply typed lambda calculus (STLC) terms. We implement generators for these benchmarks using a variety of 
-\textit{strategies}, which vary in structure and sophistication. These include: a derived generators for BSTs and STLC terms, using the approach outlined in \secref{subsection:type-derived}; 
-two custom BST generators, one that repeatedly inserts values into an initially empty tree and one that materializes the tree in a single pass by generating keys, values, and subtrees at each step;
-a custom, correct by construction generator for STLC, which is the most complex; and a generator of boolean lists.
 
@@
   \label{fig:time}
-  \caption{Average generation time over 10,000 trials.}
+  \caption{Average generation time over 10,000 trials. Across all benchmarks, we see reliable improvement from at least one of Staged SR and Staged CSR.}
 \end{figure}
@@
 It is worth noting that the time-to-failure of a given strategy on a given seed is not necessarily representative of 
-the strategy's average time-to-failure over a large number of trials. We normalize by computing the \textit{relative} performance (the ``speedup''). While this works in most cases,
+the strategy's average time-to-failure over a large number of trials.
+We normalize by computing the relative performance of \name generators vs.
+\bq generators (the ``speedup'').
+While this works in most cases,
 it is not perfect.

### 2025-03-19 paper/paper.tex
more words

--- 2025-03-19 paper/paper.tex
@@
 
-Our benchmarks consist of lists, binary search trees (BST), and simply typed lambda calculus (STLC) terms. We implement generators for these benchmarks using a variety of 
+Our benchmarks consist of boolean lists, binary search trees (BST), and simply typed lambda calculus (STLC) terms. We implement generators for these benchmarks using a variety of 
 \textit{strategies}, which vary in structure and sophistication. These include: a derived generators for BSTs and STLC terms, using the approach outlined in \secref{subsection:type-derived}; 
-two custom BST generators, one that repeatedly inserts values into an initially empty tree and one that materializes the tree in a single pass by generating keys, values, and subtrees at each step;
-a custom, correct by construction generator for STLC, which is the most complex; and a generator of boolean lists.
-
+two custom BST generators, one that repeatedly inserts values into an initially empty tree and one that materializes the tree in a single pass by generating keys, values, and subtrees at each step; a generator of boolean lists that operates similarly to the single-pass BST strategy; and
+a custom, correct-by-construction generator of STLC terms, whose behavior is the most complex.
+
+The runtime of these generators over a range of sizes is shown in \secref{fig:time}.
+Staging provides only a negligible speedup for the type-derived and insertion-based BST strategies, as their overhead is largely dominated by random sampling.
+In these cases, the primary source of performance improvement comes from our enhanced SplitMix \rand.
+In contrast, STLC strategies benefit more from staging than from improvements to \rand, as they involve fewer sampling operations but rely heavily on generator combinators.
+The same holds for the boolean list benchmark, as sampling booleans is relatively inexpensive.
+Our single-pass BST strategy falls somewhere in between: it both samples frequently and makes extensive use of generator combinators.
 \begin{figure}[h]
@@
   \includegraphics[scale=0.45]{time.png}
+  \caption{Average generation time over 10,000 trials.}
   \label{fig:time}
-  \caption{Average generation time over 10,000 trials.}
 \end{figure}

### 2025-03-19 paper/paper.tex
more words

@@
 In these cases, the primary source of performance improvement comes from our enhanced SplitMix \rand.
-In contrast, STLC strategies benefit more from staging than from improvements to \rand, as they involve fewer sampling operations but rely heavily on generator combinators.
-The same holds for the boolean list benchmark, as sampling booleans is relatively inexpensive.
+In contrast, STLC strategies benefit more from staging than \rand choice, as they involve fewer sampling operations but rely heavily on generator combinators.
+The same holds for the boolean list benchmark, as sampling booleans is inexpensive relative to integers.
 Our single-pass BST strategy falls somewhere in between: it both samples frequently and makes extensive use of generator combinators.

### 2025-03-19 paper/paper.tex
words

@@
 The runtime of these generators over a range of sizes is shown in \secref{fig:time}.
-Staging provides only a negligible speedup for the type-derived and insertion-based BST strategies, as their overhead is largely dominated by random sampling.
-In these cases, the primary source of performance improvement comes from our enhanced SplitMix \rand.
+Staging provides only a negligible speedup for type-derived and insertion-based BST strategies, whose overhead is dominated by random sampling.
+In these cases, performance improvement is primarily attributable the enhanced SplitMix \rand.
 In contrast, STLC strategies benefit more from staging than \rand choice, as they involve fewer sampling operations but rely heavily on generator combinators.
-The same holds for the boolean list benchmark, as sampling booleans is inexpensive relative to integers.
+The same holds for boolean lists, as sampling booleans is inexpensive relative to integers.
 Our single-pass BST strategy falls somewhere in between: it both samples frequently and makes extensive use of generator combinators.
+
+These results show that staging and \rand choice serve distinct but complementary roles in generator performance.
+Sampling-heavy generators see greater improvements from optimized randomness sources, while combinator-heavy ones benefit more from staging.
+Since both factors influence performance in different ways, effective generator libraries incorporate both strategies to achieve strong performance across workloads.
 \begin{figure}[h]
@@
   \includegraphics[scale=0.45]{time.png}
-  \caption{Average generation time over 10,000 trials.}
+  \caption{Time to generate a value using each strategy, averaged over 10,000 trials. Performance benefits scale with generation size; lower is better.}
   \label{fig:time}

### 2025-03-19 paper/paper.tex
words

@@
 To answer \textbf{RQ1}, we implement benchmarks in both \name and \bq. We vary the choice of \rand in our \name generators, 
-using both \bq's default \texttt{splittable\_random} (``SR'') and our improved SplitMix (``CSR''), as discussed in \secref{subsection:faster-rng}.
+using both \bq's default \texttt{splittable\_random} and our improved SplitMix, as discussed in \secref{subsection:faster-rng}.
 
 Our benchmarks consist of boolean lists, binary search trees (BST), and simply typed lambda calculus (STLC) terms. We implement generators for these benchmarks using a variety of 
-\textit{strategies}, which vary in structure and sophistication. These include: a derived generators for BSTs and STLC terms, using the approach outlined in \secref{subsection:type-derived}; 
+\textit{strategies}, which vary in structure and sophistication. These include a derived generators for BSTs and STLC terms, using the approach outlined in \secref{subsection:type-derived}; 
 two custom BST generators, one that repeatedly inserts values into an initially empty tree and one that materializes the tree in a single pass by generating keys, values, and subtrees at each step; a generator of boolean lists that operates similarly to the single-pass BST strategy; and
@@
 Staging provides only a negligible speedup for type-derived and insertion-based BST strategies, whose overhead is dominated by random sampling.
-In these cases, performance improvement is primarily attributable the enhanced SplitMix \rand.
-In contrast, STLC strategies benefit more from staging than \rand choice, as they involve fewer sampling operations but rely heavily on generator combinators.
+In this case, performance improvement is primarily attributable the enhanced SplitMix \rand.
+In contrast, STLC strategies benefit more from staging, as they involve fewer sampling operations but rely heavily on generator combinators.
 The same holds for boolean lists, as sampling booleans is inexpensive relative to integers.
@@
 These results show that staging and \rand choice serve distinct but complementary roles in generator performance.
-Sampling-heavy generators see greater improvements from optimized randomness sources, while combinator-heavy ones benefit more from staging.
-Since both factors influence performance in different ways, effective generator libraries incorporate both strategies to achieve strong performance across workloads.
+Sampling-heavy generators see greater improvements from faster randomness libraries, while combinator-heavy ones benefit more from staging.
+Since both factors influence performance in different ways, generator libraries should use both to achieve strong performance over diverse workloads.
 \begin{figure}[h]
@@
   \includegraphics[scale=0.45]{time.png}
-  \caption{Time to generate a value using each strategy, averaged over 10,000 trials. Performance benefits scale with generation size; lower is better.}
+  \caption{Time to generate a value using each strategy, averaged over 10,000 trials.
+BQ is \bq; Staged SR is \name using \bq's \texttt{splittable\_random} RNG; Staged CSR is \name using \secref{subsection:faster-rng}'s optimized SplitMix.
+Performance benefits scale with generation size; lower is better.}
   \label{fig:time}

### 2025-03-19 paper/paper.tex
words

@@
 \subsection{Benchmarking Generator Speed}
+\label{subsection:benchmarking}
 To answer \textbf{RQ1}, we implement benchmarks in both \name and \bq. We vary the choice of \rand in our \name generators, 
@@
 Sampling-heavy generators see greater improvements from faster randomness libraries, while combinator-heavy ones benefit more from staging.
-Since both factors influence performance in different ways, generator libraries should use both to achieve strong performance over diverse workloads.
+Since both factors influence performance in different ways, generator libraries should use both to be performant over diverse workloads.
+\subsubsection{Scala} \tr{TODO!!!!}
 \begin{figure}[h]
@@
 Each case study includes a diverse set of \textit{tasks}, where a task consists of a specific bug-property pair designed to test different aspects of the system (e.g., BST
-includes tasks that test insertion, deletion, and union operations on binary search trees). BST has 43 tasks; STLC has 18.
-
-We test both type-derived and custom generators for BST and STLC.
-Each generator has four implementations (``strategies''), all equivalent up to program equality: the original \bq generator (``BQ''); a staged version of the generator (``Staged''); and a staged version incorporating the CSR \rand (``Staged CSR'').
-To ensure a fair comparison, we initialize each of these generators with the same random seed, ensuring that they produce identical trees.
-
-We evaluate each strategy on every task with a 60-second timeout. 
+includes tasks that test insertion, deletion, and union operations on binary search trees). In our case, BST has 43 tasks, and STLC has 18.
+
+Strategies for BST and STLC are unchanged from \secref{subsection:benchmarking}.
+To ensure a fair comparison, we initialize each of these generators with the same random seed, ensuring that they produce identical sequences of values.
+We evaluate each strategy on all tasks using a 60-second timeout.
 If a bug is not found within this limit, the task is considered unsolved. 
@@
 Similarly, we exclude tasks where the BQ strategy completes in under 5ms,
- as such negligible runtimes do not yield meaningful insights into relative performance. These filters ensure that the reported speedups are based on cases where optimization has a measurable impact.
+ as such negligible runtimes do not yield meaningful insights into relative performance. These filters ensure that reported speedups are based on cases where optimization has a measurable impact.
 
@@
 if both generators fail to find the bug).
-To account for these edge cases, we repeat the above process for 10 seeds and compute the geometric average of the speedups.\footnote{Although the variance \textit{between} seeds in Etna can be vast, timing results are replicable for a \textit{given} seed.
+To account for these edge cases, we repeat the above process over 10 seeds and compute the geometric average of the speedups.\footnote{Although the variance \textit{between} seeds in Etna can be vast, timing results are replicable for a \textit{given} seed.
 Across 1,000 trials using the same seed,

### 2025-03-19 paper/paper.tex
nit
Making some progress on section 2.1

@@
 \usepackage{xspace}
+\usepackage{wrapfig}
 \lstset{
@@
 \subsection{Background: Monadic Generator DSLs}
-\jwc{Usual spiel about pbt and generators: here's a monadic generic generator
-library (but we're using BQ to illustrage), it has return and bind, this is what
-they're for.}
-\jwc{You often want generators to generate for sparse precoditions
-so you can exercise a property reasonably well}
-
-Property Based Testing was popularized by the QuickCheck library \cn{} in
-Haskell.  A core element of PBT as introduced in that paper is \emph{monadic
-generators}, an embedded DSL for writing the random data generators that drive
-the PBT process.  This DSL gives developers significant control over exactly
-what sorts of values are generated and what their distribution is. Both of these
-axes of control are important to adequately exercise the system under test and
-find bugs.  Indeed, the PBT literature \cn{} has repatedly demonstrated that the
-particular generator used to test a property has an enormous impact on how well
-PBT actually works \jwc{less derogatory phrasing}, and much research has gone
-into finding clever ways to generate data \cn{}.
-
-\begin{figure}
+
+Since property-based testing was first introduced in Haskell's QuickCheck
+library~\cite{claessenQuickCheckLightweightTool2000}, {\em generators} have been
+a key element of the PBT process. While properties are critical for correctly
+expressing a developer's intent for a function, the success of the testing
+ultimately rests on how well the random inputs to that property exercise the
+system under test. QuickCheck's solution to this problem, which has been copied
+dozens of frameworks across as many languages, is to provide a domain-specific
+language (DSL) for writing random data generators that gives developers complete
+control over their distribution of test inputs.
+
+\begin{wrapfigure}{r}{.5\textwidth}
 \begin{lstlisting}
@@
   type 'a t
+  val gen_int : int -> int -> int t
   val return : 'a -> 'a t
   val bind : 'a t -> ('a -> 'b t) -> 'b t
-  val gen_int : int -> int -> int t
 
@@
 \end{lstlisting}
-\caption{API of the \bq{} Generator DSL}
+\caption{Some functions from the API of the \bq{} generator DSL.}
 \label{fig:bq-api}
-\end{figure}
-
-A standard generator DSL consists of a monadic interface, plus combinators for sampling random numbers, making (weighted) random choices,
-and controlling the size of generated values. Figure~\ref{fig:bq-api} shows a selection of functions from the API of th \bq{} generator
-DSL.
-
-Because we have to choose \emph{some} syntax for this paper, we use \bq{}'s as our primary running generator DSL. However,
-this is a standin for a generic generator DSL, since they basicalyl all look like this one.
-
-The ``monadic interface'' of a generator DSL consists of \texttt{return} and \texttt{bind}.
-The generator \texttt{return x} is the constant generator, always generating the value \texttt{x}.
-Running a generator \texttt{bind g k} runs the generator \texttt{g}, producing a value \texttt{a},
-and then runs the generator \texttt{k a}. Monadic \texttt{bind} allows for dependencies.
-
-The \texttt{gen\_int a b} generator samples an integer uniformly in the inclusive range from \texttt{a} to \texttt{b}.
-
-\jwc{\texttt{gen\_int} here is really \texttt{int\_uniform\_inclusive}. I also think we should completely omit named arguments in the paper so we don't confuse with splices.}
-
-Figure ... shows an example of a generator written with these combinators. It first samples an int between $0$ and $100$, 
-names it $x$, samples another between $0$ an $x$, and then returns the pair of them both.
-Most languages in which monadic APIs are common expose some sort of sytnactic sugar for them --- this is true of OCaml and Scala. 
-OCaml's looks like \texttt{let\%bind x = e in e'}, which desguars to \texttt{...}. Figure~\jwc{...} also shows this generator written with OCaml's syntacitc sugar\footnote{It's actually Jane's}.
-
-\begin{figure}
-
+\end{wrapfigure}
+
+While many DSL designs are possible, the standard design for QuickCheck-style
+generators is via an embedded {\em monadic} language. For the purposes of this
+paper, we use the sytnax from \bq, which is presented in
+Figure~\ref{fig:bq-api}. The library provides a series of basic generators, for
+example ones like \texttt{gen\_int} for random integers in a range, along with
+the ``monadic interface'' consisting of \texttt{return} and \texttt{bind}.  The
+generator \texttt{return x} is the constant generator, always generating the
+value \texttt{x}. Running a generator \texttt{bind g k} runs the generator
+\texttt{g}, producing a value \texttt{a}, and then runs the generator \texttt{k
+a}.
+\iflater
+\jwc{\texttt{gen\_int} here is really \texttt{int\_uniform\_inclusive}. I also
+think we should completely omit named arguments in the paper so we don't confuse
+with splices.}
+\fi
+
+Together, these three functions are the bare minimum for constructing arbitrary
+random data generators: \texttt{gen\_int} provides a base source of randomness,
+and \texttt{return} and \texttt{bind} allow generators to be composed to create
+larger, more complex generators. Figure~\ref{fig:simple-bind} shows a generator
+built with these operations; it first samples an int between $0$ and $100$,
+names it $x$, samples another between $0$ an $x$, and then returns the pair of
+them both. In practice, PBT generator libraries also include other functions
+that make generator construction easier, including functions like
+\texttt{weighted\_union} for making a weighted choice between different
+generators as well as functions for managing the sizes of generated values and
+combinators for recursion.
+
+\begin{figure}[h]
+  \begin{subfigure}{.49\textwidth}
 \begin{lstlisting}
@@
     Bq.bind (Bq.gen_int 0 x) (fun y ->
-      Bq.return (x,y)
-  ))
+      Bq.return (x,y)))
 \end{lstlisting}
-
+\caption{A simple generator using \texttt{bind} explicitly.}\label{fig:simple-bind}
+  \end{subfigure}
+  \begin{subfigure}{.49\textwidth}
 \begin{lstlisting}
@@
 \end{lstlisting}
+\caption{An equivalent generator using macros for \texttt{bind}.}\label{fig:simple-macro}
+  \end{subfigure}
+  \caption{Simple monadic generators for a pair of ordered integers.}\label{fig:simple-pair}
 \end{figure}
+
+Most languages in which monadic APIs are common expose some sort of syntactic
+sugar for them. In OCaml\footnote{OCaml actually has a few versions of syntax
+for monadic APIs; this is the one provided by Jane Street's libraries.}, this
+looks like \texttt{let\%bind x = e in e'}, which is a macro that desugars to
+\texttt{bind e (fun x => e')}. Figure~\ref{fig:simple-macro} shows the same
+generator as before, now written with macros.
 
@@
 Strategies for BST and STLC are unchanged from \secref{subsection:benchmarking}.
-To ensure a fair comparison, we initialize each of these generators with the same random seed, ensuring that they produce identical sequences of values.
+To ensure a fair comparison, we initialize each strategy with the same random seed, ensuring that they produce identical sequences of values.
 We evaluate each strategy on all tasks using a 60-second timeout.

### 2025-03-19 paper/paper.tex
more work on other langs

--- 2025-03-19 paper/paper.tex
@@
 \begin{itemize}
-  \item \textbf{Racket} is the next target that we intend to test the \name{} methodology on. The entire Racket philosophy is intertwined with
-  using macros to build small eDSLs like the ones we use to write generators, and so it is a natural fit.
+  \item \textbf{Racket} is the next target that we intend to test the \name{} methodology on.
+  The entire Racket philosophy is intertwined with
+  using macros to build small eDSLs like the ones we use to write generators, and so it is a natural fit. Indeed,
+  there are at least two \cn{} PBT libraries in Racket.
+  \jwc{The Racket macros literature, while extensive, is not about staging!}
   We have not yet tested the technique in Racket for the simple reason that the authorship team of this paper does not posess
@@
   \item \textbf{Haskell} birthed PBT, and so it is natural to ask why we have not yet applied \name{} to it. The basic answer is that
-  because most tasks in Haskell are accomplished with monads, GHC is designed specially to eliminate the run-time overhead of monadic
+  the Haskell compiler (GHC) is designed specially to eliminate the run-time overhead of monadic
   abstractions! Indeed, because Haskell is pure, GHC can aggressively inline and beta-reduce
   nearly anything as part of its normal optimiation steps.
+  This means that in many cases, the impact of a staged PBT library (and indeed, all staged monadic DSLs) negligible.
   For complex enough programs however, the heuristics GHC uses can degrade, leading to performance overheads for monadic code.
@@
   to build eDSLs in contexts where one does not want to rely on the optimizer's heuristics.
-  \item \textbf{Rust:} asdf
-  \item \textbf{Python:} asdf
+  In short: it is possible to port \name{} to Haskell QuickCheck, but the benefits would be much less pronounced.
+
+  \item \textbf{Rust} shares some similarities with functional languages, and
+  indeed, it is host to a number of property-based testing libraries \cn{}. However Rust does not
+  structurally encourage monadic eDSLs --- it does not have a special monadic syntax --- and so
+  PBT libraries in Rust ask programmers to define generators directly as \texttt{seed -> value} functions.
+  For this reason, \name{} does not seem directly applicable to any of the PBT libraries in Rust.
+
+  \item \textbf{Python} is home to Hypothesis, the most widely used PBT library in any language.
+
 \end{itemize}
-
-\jwc{
-  \begin{itemize}
-    \item Racket is the only ``real'' strict functional language we haven't tested this in. It'd definitely work.
-    \item Haskell GHC does a lot of these optimizations already, since the code is pure. Since QC generators are relatively small programs,
-    GHC has little trouble specializing them. Of course, this is not guaranteed. A version of this idea can easily be ported to the original QC with template haskell, to guarantee
-    the highest-performance generators. But the benefit would be lower since GHC is smarter. Worth doing, though.
-    \item Rust: generator libraries are not monadic, presumably to avoid exactly these issues.
-    \item Python. Maybe could do it?? Hypothesis is the most used PBT library out there so this could be high impact.
-  \end{itemize}
-}
 
@@
 \jwc{cite Lionel Parreaux  work, he's on PC}
+\jwc{Macros in Racket}
 

### 2025-03-19 paper/paper.tex
scala barchart

@@
 Since both factors influence performance in different ways, generator libraries should use both to be performant over diverse workloads.
-\subsubsection{Scala} \tr{TODO!!!!}
+\subsubsection{Scala}
+To confirm that the benefit of staging extends to other strict functional languages, we apply it to ScalaCheck, a widely-used PBT library written in Scala.
+We implement a subset of our strategies from \secref{subsection:benchmarking} in both staged and unstaged ScalaCheck.
+
 \begin{figure}[h]
@@
   \centering
-  \vspace{0.5em} % Kept the vertical spacing
-  \begin{tabular}{lrl}
-  \toprule
-  Benchmark & \multicolumn{2}{c}{avg. time (ns/op) $\pm$ std. dev.} \\
-  \midrule
-  BST & $3,560.813$ & $\pm\>1,511.201$ \\
-  BST, Staged & $751.287$ & $\pm\>39.890$ \\
-  BST (Type-Derived) & $44,619.863$ & $\pm\>5,761.568$ \\
-  BST (Type-Derived), Staged & $6,863.209$ & $\pm\>416.188$ \\
-  RBT (Type-Derived) & $60,559.846$ & $\pm\>4,901.980$ \\
-  RBT (Type-Derived), Staged & $8,511.528$ & $\pm\>324.615$ \\
-  STLC & $44,497.623$ & $\pm\>1,970.484$ \\
-  STLC, Staged & $7,151.259$ & $\pm\>355.639$ \\
-  \bottomrule
-  \end{tabular}
- \caption{Benchmarks reimplemented in ScalaCheck show similar performance improvements to \name.}
+  \includegraphics[scale=0.32]{scala.png}
+ \caption{Staging in ScalaCheck shows greater speedups than in \bq. SC is ScalaCheck; BQ is \bq.}
  \label{fig:scala}

### 2025-03-19 paper/paper.tex
flagging for better name

--- 2025-03-19 paper/paper.tex
@@
 \author{Cynthia Richey}
+\authornote{Both authors contributed equally to this work.}
 \affiliation{%
@@
 \author{Joseph W. Cutler}
+\authornotemark[1]
 \affiliation{
@@
 \jwc{phrasing could suggest ``race against time'' aspect.}\hg{Agree, can't see
-how to fit this in yet}\fi
+how to fit this in yet} \jwc{Also think the phrasing should say something like ``the goal is to fail as fast as possible''}\fi
 
@@
 Together, these observations raise a fundamental question: can generator libraries be expressive and high-level without compromising efficiency?
-%In practice, PBT is used 
-%as a lightweight, ad hoc approach to software correctness,
-%\jwc{Move this bit about how it's used to the first sentence. I.e. PBT is a testing technqique where random data is fed to a system under test...}
-%aimed at 
-%detecting bugs within a short testing window---typically 50 milliseconds 
-%to 30 seconds.
-\jwc{This sells us a little short: we should say something to the effect of ``Recent work on understanding how PBT is used in practice has determined that practitioners often run PBT with very short time budgets...''} Given this practical constraint, it is imperative that
-%inputs be generated as fast as possible; in other words, generators must be 
-%treated as performance-sensitive code.
-
-% The locus for improvement in this area \jwc{``This area'' sounds like it's talking about the whole area of PBT...
-Maybe ``a high-leverage way to improve the performance of generators is to...''} is \textit{generator libraries}: if
-
-  % in strict functionallanguages like OCaml and Scal
+
   To address this question, we outline two key principles for designing generator libraries that balance efficiency with flexibility.
@@
 parallelism?}\fi Together, these insights provide a recipe for improving existing generator libraries.
-
-%\jwc{I think we should use return/bind as the simplest example, since then you don't have to get into the weeds of allocation or closures.
-The functional abstractions of a monadic generator DSL block the compiler from generating fast code.}
-%\jwc{Then present this case as ``more sophisticated use of staging'' or something?}
-%where an unstaged generator DSL might represent a choice between
-%a list of options as a dynamically-allocated array of closures, a staged 
-%generator can expand to an if-else tree at compile 
-%time---eliminating the allocation entirely.
-% This approach is not tied to any
-%particular PBT library, but rather provides a recipe for improving even
-%state-of-the-art implementations \jwc{Sentence structure doesn't follow. Instead maybe ``not any particular, but a recipe for improving in general''}.
-
-%First, we demonstrate the impact of
-%RNG choice by optimizing an existing \rand. Our optimized RNG produces 
-%exactly 
-%the same values as its unoptimized counterpart, allowing us to directly 
-%measure the impact of this choice on generator performance.
-\jwc{Permute?
-I think staging \emph{then} randomness.} \jwc{Also, i don't love this as the first place we explain the drop-in equivalence of the RNG.
-It's unmotivated why wen want this.}
 
@@
 \bcp{That's actually not a bad section title!}
+\todo{Intro of this section}
 
@@
 In the years since the original QuickCheck paper \cn{}, PBT has had remarkable success
-in languages outside of the small ICFP world from which it came.
+in languages outside of the Haskell world from which it came.
 For this reason, readers who are users and developers of PBT libraries in \emph{other} languages ---
@@
 
-  \item \textbf{Haskell} birthed PBT, and so it is natural to ask why we have not yet applied \name{} to it. The basic answer is that
-  the Haskell compiler (GHC) is designed specially to eliminate the run-time overhead of monadic
-  abstractions! Indeed, because Haskell is pure, GHC can aggressively inline and beta-reduce
-  nearly anything as part of its normal optimiation steps.
-  This means that in many cases, the impact of a staged PBT library (and indeed, all staged monadic DSLs) negligible.
-  For complex enough programs however, the heuristics GHC uses can degrade, leading to performance overheads for monadic code.
-  For this reason, Haskell does have a multi-stage programming system called Template Haskell \cn{}, which is sometimes 
-  to build eDSLs in contexts where one does not want to rely on the optimizer's heuristics.
-  In short: it is possible to port \name{} to Haskell QuickCheck, but the benefits would be much less pronounced.
+  \item \textbf{Haskell} is home to the original PBT implementation \cn{}, and
+  so it is natural to ask why we have not yet applied \name{} to it. The basic
+  answer is that the Haskell compiler (GHC) is designed specially to eliminate
+  the run-time overhead of monadic abstractions! Indeed, because Haskell is
+  pure, GHC can aggressively inline and beta-reduce nearly anything as part of
+  its normal optimiation steps.  This means that in many cases, the impact of a
+  staged PBT library (and indeed, all staged monadic DSLs) negligible.  For
+  complex enough programs however, the heuristics GHC uses can degrade, leading
+  to performance overheads for monadic code.  For this reason, Haskell does have
+  a multi-stage programming system called Template Haskell \cn{}, which is
+  sometimes to build eDSLs in contexts where one does not want to rely on the
+  optimizer's heuristics.  In short: it is possible to port \name{} to Haskell
+  QuickCheck, but the benefits would be much less pronounced.
 

### 2025-03-19 paper/paper.tex
comments on intro

@@
 test. These generators are written with the help of \textit{generator
-libraries}, usually expressed as embedded domain-specific languages (eDSLs) that
+libraries}, usually expressed as embedded domain-specific languages (eDSLs) \cntxt{edsls} that
 provide combinators for building and composing generators. Optimizing these
@@
 \textit{multi-stage programming}, or ``staging,'' which has long been used to
-optimize functional DSLs~\cn.
+build DSLs ``without regret''~\cn.
 We apply this technique to generators, completely eliminating the overhead of
@@
 libraries, each call to monadic bind usually spends time constructing and
-immediately executing a closure at run time. \jwc{Todo: joe --- cite some evidence for this}
+immediately executing a closure at run time. \todo{cite some evidence for this} \jwc{Maybe this sentence isn't needed at all.}
 Meanwhile, a staged approach can resolve the bind at compile time, enabling the
@@
 high-quality PBT library in OCaml, authored by software engineers at the trading firm Jane Street.
+\jwc{Also to ScalaCheck! Let's put this up front.}
 Our improved version, \name, is a direct drop-in replacement for
@@
   \item \textbf{Python} is home to Hypothesis, the most widely used PBT library in any language.
+  \jwc{More here}
 

### 2025-03-19 paper/paper.tex
scala sec

@@
 
-The runtime of these generators over a range of sizes is shown in \secref{fig:time}.
+The runtime of these generators over a range of sizes is shown in \figref{fig:time}.
 Staging provides only a negligible speedup for type-derived and insertion-based BST strategies, whose overhead is dominated by random sampling.
@@
 \subsubsection{Scala}
-To confirm that the benefit of staging extends to other strict functional languages, we apply it to ScalaCheck, a widely-used PBT library written in Scala.
+To show that the benefit of staging extends to other strict functional languages, we apply it to ScalaCheck, a widely-used PBT library written in Scala.
 We implement a subset of our strategies from \secref{subsection:benchmarking} in both staged and unstaged ScalaCheck.
 
+As shown in \figref{fig:scala}, staged ScalaCheck achieves an even greater performance gain over ScalaCheck than \name does over \bq.
+This discrepancy arises from ScalaCheck's representation of generators as functions of type \texttt{size -> seed -> Option[A]}.
+Each generator combinator must construct and then pattern-match on these \texttt{Option} values, introducing significant boxing and unboxing overhead at each step.
+Staging allows us to eliminate this overhead, resulting in substantial speedups.
+These results illustrate that staging is a powerful \textit{and} general optimization technique, delivering substantial performance gains in strict functional languages beyond OCaml.
 \begin{figure}[h]
@@
 It is worth noting that the time-to-failure of a given strategy on a given seed is not necessarily representative of 
-the strategy's average time-to-failure over a large number of trials. We normalize by computing the \textit{relative} performance (the ``speedup''). While this works in most cases,
+the strategy's average time-to-failure over a large number of trials. We normalize by computing relative performance (the ``speedup''). While this works in most cases,
 it is not perfect.
@@
 if both generators fail to find the bug).
-To account for these edge cases, we repeat the above process over 10 seeds and compute the geometric average of the speedups.\footnote{Although the variance \textit{between} seeds in Etna can be vast, timing results are replicable for a \textit{given} seed.
+To account for these cases, we repeat the above process over 10 seeds and compute the geometric average of their speedups.\footnote{Although the variance \textit{between} seeds in Etna can be vast, timing results are replicable for a \textit{given} seed.
 Across 1,000 trials using the same seed,
@@
   \caption{Geometric average of all speedups for each strategy and benchmark, showing that staging leads to better bug-finding ability across the board.
-As in \todo{earlier figure}, benchmarks with higher numbers of binds (BST (Single-Pass); STLC) see more benefit from staging, whereas benchmarks with more sampling (BST) see a greater speedup from the CSR \rand.}
+As in \todo{earlier figure}, benchmarks with higher numbers of binds (BST (Single-Pass); STLC) see more benefit from staging, whereas benchmarks with more sampling (BST) see a greater speedup from faster SplitMix.}
   \label{fig:etna}

### 2025-03-19 paper/paper.tex
undo broken macro

@@
 %  \end{itemize}
-%}
+%
 
@@
 test. These generators are written with the help of \textit{generator
-libraries}, usually expressed as embedded domain-specific languages (eDSLs) \cntxt{edsls} that
+libraries}, usually expressed as embedded domain-specific languages (eDSLs) %\cntxt{edsls} 
+that
 provide combinators for building and composing generators. Optimizing these

### 2025-03-19 paper/paper.tex
csplitmix

@@
 To answer \textbf{RQ1}, we implement benchmarks in both \name and \bq. We vary the choice of \rand in our \name generators, 
-using both \bq's default \texttt{splittable\_random} and our improved SplitMix, as discussed in \secref{subsection:faster-rng}.
+using both \bq's default \texttt{splittable\_random} and our improved SplitMix (\csplitmix), as discussed in \secref{subsection:faster-rng}.
 
@@
 Staging provides only a negligible speedup for type-derived and insertion-based BST strategies, whose overhead is dominated by random sampling.
-In this case, performance improvement is primarily attributable the enhanced SplitMix \rand.
+In this case, performance improvement is primarily attributable the \csplitmix \rand.
 In contrast, STLC strategies benefit more from staging, as they involve fewer sampling operations but rely heavily on generator combinators.
@@
 Our single-pass BST strategy falls somewhere in between: it both samples frequently and makes extensive use of generator combinators.
+\tr{Goal is to replace this paragraph with the ref-counting chart---I think it will be WAY less boring.}
 
@@
 To show that the benefit of staging extends to other strict functional languages, we apply it to ScalaCheck, a widely-used PBT library written in Scala.
-We implement a subset of our strategies from \secref{subsection:benchmarking} in both staged and unstaged ScalaCheck.
-
-As shown in \figref{fig:scala}, staged ScalaCheck achieves an even greater performance gain over ScalaCheck than \name does over \bq.
+We implement a subset of our strategies from \secref{subsection:benchmarking} in both standard and staged ScalaCheck (``\scalaname'').
+
+As shown in \figref{fig:scala}, \scalaname achieves an even greater performance gain over ScalaCheck than \name does over \bq.
 This discrepancy arises from ScalaCheck's representation of generators as functions of type \texttt{size -> seed -> Option[A]}.
@@
 Staging allows us to eliminate this overhead, resulting in substantial speedups.
-These results illustrate that staging is a powerful \textit{and} general optimization technique, delivering substantial performance gains in strict functional languages beyond OCaml.
+
+These results illustrate that staging is a powerful \textit{and} general optimization technique, capable of delivering performance gains in strict functional languages beyond OCaml.
 \begin{figure}[h]
@@
   \caption{Time to generate a value using each strategy, averaged over 10,000 trials.
-BQ is \bq; Staged SR is \name using \bq's \texttt{splittable\_random} RNG; Staged CSR is \name using \secref{subsection:faster-rng}'s optimized SplitMix.
+BQ is \bq; Staged SR is \name using \bq's \texttt{splittable\_random} RNG; Staged CSR is \name using \secref{subsection:faster-rng}'s \csplitmix.
 Performance benefits scale with generation size; lower is better.}
@@
   \caption{Geometric average of all speedups for each strategy and benchmark, showing that staging leads to better bug-finding ability across the board.
-As in \todo{earlier figure}, benchmarks with higher numbers of binds (BST (Single-Pass); STLC) see more benefit from staging, whereas benchmarks with more sampling (BST) see a greater speedup from faster SplitMix.}
+As in \todo{earlier figure}, benchmarks with higher numbers of binds (BST (Single-Pass); STLC) see more benefit from staging, whereas benchmarks with more sampling (BST) see a greater speedup from \csplitmix.}
   \label{fig:etna}
@@
 \section{Other Languages and Libraries}
+\tr{Subtitle: ``Beyond strict functional languages?''}
 \label{subsection:other-langs}

### 2025-03-19 paper/paper.tex
nits
committed

--- 2025-03-19 paper/paper.tex
@@
 build DSLs ``without regret''~\cn.
-We apply this technique to generators, completely eliminating the overhead of
-many common generator abstractions. For example, in unstaged generator
-libraries, each call to monadic bind usually spends time constructing and
-immediately executing a closure at run time. \todo{cite some evidence for this} \jwc{Maybe this sentence isn't needed at all.}
-Meanwhile, a staged approach can resolve the bind at compile time, enabling the
-compiler to inline and eliminate these closures.
+We apply this technique to generators, completely eliminating the run-time overhead of
+many common generator abstractions.
 Secondly, we demonstrate the impact of different choices of \rand on generator
@@
 
-To demonstrate the effectiveness of our approach, we apply it to \bq, a
-high-quality PBT library in OCaml, authored by software engineers at the trading firm Jane Street.
-\jwc{Also to ScalaCheck! Let's put this up front.}
-Our improved version, \name, is a direct drop-in replacement for
-\bq that 
-is \textit{100\% semantically equivalent}: that is, given identical random seeds, generators written using \name{}
-produce exactly the same sequence of values as those written in \bq. However, in \name, 
-the abstractions provided by the generator DSL are zero-cost, and the \rand is 
-highly-optimized. The semantic equivalence between \name and \bq allows us to 
+To demonstrate the effectiveness of our approach, we apply it to two PBT generator libraries, namely \bq{} --- a
+high-quality PBT library in OCaml, authored by software engineers at the trading firm Jane Street ---
+and ScalaCheck --- a very mature PBT library and the de facto choice in Scala.
+In both languages, our staged versions are direct drop-in replacements that are
+\textit{100\% semantically equivalent}: that is, given identical random seeds, the same generator written once with \name{}
+and once with the original generator library produces exactly the same sequence of values.
+The semantic equivalence between the original generators and \name generators allows us to 
 perform clean, fine-grained comparisons, as differences in generation speed are
@@
 We evaluate the optimizations in \name with a series of case studies in each
-generator library, and show that \name runs faster than \bq.
+generator library, and show that the \name generators run faster than their unstaged equivalence.
 Further, we show that this performance benefit\hg{wc?} translates to bug-finding
@@
 measuring how quickly various generators detect them.
-Finally, to demonstrate the generality of our technique, we implement a version
-of \name in Scala \hg{(based on scalacheck?)} and benchmark the same generators---thereby highlighting its portability to other
-strict functional languages.
-\iflater
-\jwc{I think we should more explicitly point out the portability earlier, maybe
-up where we say we're applying it to BQ.}\hg{Agree}
-\fi
 
@@
 
-The runtime of these generators over a range of sizes is shown in \figref{fig:time}.
+The runtime of these generators over a range of sizes is shown in \secref{fig:time}.
 Staging provides only a negligible speedup for type-derived and insertion-based BST strategies, whose overhead is dominated by random sampling.
@@
 \subsubsection{Scala}
-To show that the benefit of staging extends to other strict functional languages, we apply it to ScalaCheck, a widely-used PBT library written in Scala.
+To confirm that the benefit of staging extends to other strict functional languages, we apply it to ScalaCheck, a widely-used PBT library written in Scala.
 We implement a subset of our strategies from \secref{subsection:benchmarking} in both staged and unstaged ScalaCheck.
 
-As shown in \figref{fig:scala}, staged ScalaCheck achieves an even greater performance gain over ScalaCheck than \name does over \bq.
-This discrepancy arises from ScalaCheck's representation of generators as functions of type \texttt{size -> seed -> Option[A]}.
-Each generator combinator must construct and then pattern-match on these \texttt{Option} values, introducing significant boxing and unboxing overhead at each step.
-Staging allows us to eliminate this overhead, resulting in substantial speedups.
-These results illustrate that staging is a powerful \textit{and} general optimization technique, delivering substantial performance gains in strict functional languages beyond OCaml.
 \begin{figure}[h]
@@
 It is worth noting that the time-to-failure of a given strategy on a given seed is not necessarily representative of 
-the strategy's average time-to-failure over a large number of trials. We normalize by computing relative performance (the ``speedup''). While this works in most cases,
+the strategy's average time-to-failure over a large number of trials. We normalize by computing the \textit{relative} performance (the ``speedup''). While this works in most cases,
 it is not perfect.
@@
 if both generators fail to find the bug).
-To account for these cases, we repeat the above process over 10 seeds and compute the geometric average of their speedups.\footnote{Although the variance \textit{between} seeds in Etna can be vast, timing results are replicable for a \textit{given} seed.
+To account for these edge cases, we repeat the above process over 10 seeds and compute the geometric average of the speedups.\footnote{Although the variance \textit{between} seeds in Etna can be vast, timing results are replicable for a \textit{given} seed.
 Across 1,000 trials using the same seed,
@@
   \caption{Geometric average of all speedups for each strategy and benchmark, showing that staging leads to better bug-finding ability across the board.
-As in \todo{earlier figure}, benchmarks with higher numbers of binds (BST (Single-Pass); STLC) see more benefit from staging, whereas benchmarks with more sampling (BST) see a greater speedup from faster SplitMix.}
+As in \todo{earlier figure}, benchmarks with higher numbers of binds (BST (Single-Pass); STLC) see more benefit from staging, whereas benchmarks with more sampling (BST) see a greater speedup from the CSR \rand.}
   \label{fig:etna}

### 2025-03-19 paper/paper.tex

--- 2025-03-19 paper/paper.tex
@@
 To ensure a fair comparison, we initialize each strategy with the same random seed, ensuring that they produce identical sequences of values.
-We evaluate each strategy on all tasks using a 60-second timeout.
+We run each strategy on all tasks using a 60-second timeout.
 If a bug is not found within this limit, the task is considered unsolved. 

### 2025-03-19 paper/paper.tex
2.1 finished ish

@@
 generators is via an embedded {\em monadic} language. For the purposes of this
-paper, we use the sytnax from \bq, which is presented in
-Figure~\ref{fig:bq-api}. The library provides a series of basic generators, for
-example ones like \texttt{gen\_int} for random integers in a range, along with
+paper, we use the syntax from \bq, which is presented in
+Figure~\ref{fig:bq-api}, but these kinds of DSLs can also be found in languages
+like Haskell, Scala, Rust, and many more.
+The library provides some basic generators, for
+example \texttt{gen\_int} for generating random integers in a range, along with
 the ``monadic interface'' consisting of \texttt{return} and \texttt{bind}.  The
@@
 names it $x$, samples another between $0$ an $x$, and then returns the pair of
-them both. In practice, PBT generator libraries also include other functions
-that make generator construction easier, including functions like
-\texttt{weighted\_union} for making a weighted choice between different
-generators as well as functions for managing the sizes of generated values and
-combinators for recursion.
+them both.
 
@@
 Most languages in which monadic APIs are common expose some sort of syntactic
-sugar for them. In OCaml\footnote{OCaml actually has a few versions of syntax
-for monadic APIs; this is the one provided by Jane Street's libraries.}, this
+sugar for them. In OCaml\footnote{OCaml actually has a few ways to implement
+monadic syntax; this is the one provided by Jane Street's libraries.}, this
 looks like \texttt{let\%bind x = e in e'}, which is a macro that desugars to
@@
 
-\jwc{
-Explain effect ordering here!!
-\begin{itemize}
-  \item \texttt{Bq.bind} has the CBV sampling semantics
-  \item If you \texttt{Bq.bind} \texttt{x} and then use it twice, both uses refer to the \emph{same} value.
-  \item Sampling is effectful: it mutates the internal state.
-  \item Permuting two independent binds gives you a generator that is distributionally equivalent, but not equivlaent when considered up to pointwise equality as \emph{functions} from seed to value.
-  \item Note that it's much more sensible to compare the bugfinding performance of two function-equivalent generators:
-\end{itemize}
-}
-
-\jwc{
-  \begin{itemize}
-    \item PBT generator libraries include more functions than just the monad interface and sampling, libraries include helper functions.
-    \item With \texttt{weighted\_union}, to make a weighted choices, and \texttt{fixed\_point} to define recursive generators.
-    \item Also \texttt{size} and \texttt{with\_size} to adjust the size parameter --- these are useful to ensure we terminate with recursive generators.
-  \end{itemize}
-}
-
-\jwc{This section needs to explain approximately how weighted union works --- compute a sum of the weights, sample between 0 and the sum, pick
-the first element where the partial sums to the left exceeds the sampled number.
-(This has been written about a billion times, cite one of the papers about this.)
-}
-\jwc{I re-explain this later somewhere i think, so we need to remember to RM that explanation.}
-
+\subsubsection{Subtleties of Bind}
+The \texttt{bind} function in monadic generator libraries sequences generators
+with a call-by-value semantics, regardless of the calling convention of the
+language in which it is implemented. Concretely, this means that sampling
+\begin{center}
+  \texttt{let\%bind x = gen\_int 0 10 in return (x + x)}
+\end{center}
+always produces an even number; \texttt{x} is only sampled once (when the
+\texttt{bind} is evaluated) and from that point on it is a pure value. This is
+in contrast to the semantics of some probabilistic programming languages that
+may have call-by-name binders. This is important because \hg{Joe why are we
+talking about this here? What are we going to use this fact for?}.
+
+Another subtle aspect of \texttt{bind} is that many implementations mutate the
+internal state of the random number generator\footnote{The original QuickCheck
+implementation in Haskell uses a splittable random number generator~\cn, and
+\texttt{bind} splits the seed.} This means that generators do not always obey
+equivalences that one might expect from a monadic program. For example,
+\begin{center}
+  \texttt{let\%bind \_x = return () in gen\_int 0 10}
+\end{center}
+does not necessarily produce the same sequence of values as \texttt{gen\_int 0
+10}; the \texttt{bind} updates the random seed in the background before
+geenrating the integer, even though it looks like a no-op. Critically, while
+these generators do not produce the same sequence of values, they do produce the
+same {\em distributions} of values. We say that generators obey standard monadic
+equalities up to ``distributional equivalence.''
+In Section~\hg{todo}, we discuss why distributional equivalence is often weaker
+than one might like when evaluating the differences between
+generators.\hg{Update this}
+
+\subsubsection{Weighted Union and Other Conveniences}
+Generator libraries also include other functions that make generator
+construction easier; some examples of these are also shown in
+Figure~\ref{fig:bq-api}.  The \texttt{weighted\_union} function is a
+particularly well-used one: it makes making a weighted choice between different
+generators, allowing the developer to combine different sub-generators into a
+single program and tune the data distribution. Also important are functions like
+\texttt{size} and \texttt{with\_size} that are used to control the sizes of
+generated values and \texttt{fixed\_point} that is used tto define recursive
+generators.
+
+\begin{wrapfigure}{r}{.5\textwidth}
 \begin{lstlisting}
@@
       let%bind r = with_size (n / 2) rg in
-      return (Node (l,x,r))
-    )
-  ]
-)
+      return (Node (l,x,r)))
+  ])
 \end{lstlisting}
-
-\jwc{This generator generates binary trees using all the combinators, yay. Explain em.}
-
-\jwc{IMPORTANT: We are using OCaml because we have to pick a syntax for this section, but these abstraction overheads exist in other languages -- at least scala, rust.}
+\caption{A generator using a variety of convenience functions.}\label{fig:tree_of}
+\end{wrapfigure}
+
+The generator in Figure~\ref{fig:tree_of} uses all of the aforementioned
+features. It uses \texttt{fixed\_point} to define a recursive generator that
+reads the current value of \texttt{size} to determine how to generate a tree.
+The generator uses \texttt{weighted\_union} to make a random choice between an
+empty tree and a node, choosing a node with weight proportional to the current
+size and choosing a leaf otherwise. When generating a node, the generator uses
+\texttt{with\_size} to reduce the value of the size parameter for future
+iterations.
 

### 2025-03-19 paper/paper.tex
last bit of intro work

--- 2025-03-19 paper/paper.tex
@@
 \textit{multi-stage programming}, or ``staging,'' which has long been used to
-build DSLs ``without regret''~\cn.
+build DSLs ``without regret''~\cn{}.
 We apply this technique to generators, completely eliminating the run-time overhead of
@@
 Secondly, we demonstrate the impact of different choices of \rand on generator
-performance by conducting a controlled comparison. \iflater\hg{More words here for
-parallelism?}\fi Together, these insights provide a recipe for improving existing generator libraries.
-
-To demonstrate the effectiveness of our approach, we apply it to two PBT generator libraries, namely \bq{} --- a
+performance by conducting a controlled comparison. Together, these insights provide a recipe for improving existing generator libraries.
+
+To demonstrate the effectiveness of our approach, we apply it to two PBT generator libraries, namely \bq{} \cn{} --- a
 high-quality PBT library in OCaml, authored by software engineers at the trading firm Jane Street ---
-and ScalaCheck --- a very mature PBT library and the de facto choice in Scala.
+and ScalaCheck \cn{} --- a very mature PBT library and the de facto choice in Scala.
 In both languages, our staged versions are direct drop-in replacements that are
@@
 Further, we show that this performance benefit\hg{wc?} translates to bug-finding
-ability by running these case studies in the Etna framework~\cite{etna}, a
+ability \jwc{The paper is called ``fail faster'' --- we shouldn't say ``bug finding ability'', we should say something like ``how fast you can squash bugs''.
+This is too passive }by running these case studies in the Etna framework~\cite{etna}, a
 platform that simulates real-world PBT usage\hg{Don't love this phrase, I'll
@@
 \begin{enumerate}
-    \item We identify two key sources of inefficiencies in PBT generator libraries, which can significantly impact performance: 
-    namely, and abstraction overhead, and \rand choice. \secref{section:motiv}
+    \item We identify two key sources of inefficiency in PBT generator libraries, which can significantly impact performance: 
+    namely and abstraction overhead and \rand choice. 
     % \item We apply multi-stage programming to the novel domain of generators. \jwc{Less sure this should be a top-level conrib.} \secref{section:}
-    \item We implement our approach in OCaml and Scala, showcasing its cross-language applicability.
-    \secref{section:4}
-    \item We construct generators in \name and \scalaname, and show that they give substantial performance improvements. Finally, we show
-    that these performance improvements extend to bug-finding ability.
-     \secref{section:eval}
+    \item We implement \name{} in both OCaml and Scala 3, showcasing its cross-language applicability.
+    \item We demonstrate that both \name{} and choosing a faster \rand yield substantial performance improvements. Finally, we show
+    that these performance improvements extend to significantly improved bug-finding ability. \jwc{This should be sell-ier: 2x speedups, etc. ``bug-finding ability'' isn't a race against time.}
 \end{enumerate}
@@
 \jwc{Ensure we have a signposting table-of-contents-ish few sentences here, if it's not covered by the contribs.}\todo{Todo Thia}
+In Section~\ref{section:motiv}, we introduce monadic generator DSLs, discuss how they work, and take an example-guided tour through some of the ways that
+they are slower than is possible.
+Then, in Section~\ref{section:faster-generators}, we explain multi-stage programming,
+iteratively develop the \name{} technique, and discuss options for faster \rand{} \jwc{fix}.
+Section~\ref{section:eval} evaluates all of the previous work, demonstrating the generator performance improvements
+and how they lead to much faster bug finding.
 
@@
 \section{Speeding up Generator DSLs by Staging and Faster Randomness Libraries}
+\label{section:faster-generators}
 \jwc{This section needs a much better title...}

### 2025-03-19 paper/paper.tex
in this case, later should be now

--- 2025-03-19 paper/paper.tex
@@
 budgets---e.g., between 50 milliseconds and 30 seconds---and it is important
-that bugs can be found in that time-frame.\iflater
+that bugs can be found in that time-frame.
 \jwc{phrasing could suggest ``race against time'' aspect.}\hg{Agree, can't see
-how to fit this in yet} \jwc{Also think the phrasing should say something like ``the goal is to fail as fast as possible''}\fi
+how to fit this in yet} \jwc{Also think the phrasing should say something like ``the goal is to fail as fast as possible''}
 

### 2025-03-19 paper/paper.tex
delete subsection about the trick, nobody cares

--- 2025-03-19 paper/paper.tex
@@
 
-\subsubsection{The Trick at Other Types}
-To write more interesting generators, we also need the ability to generate code that manipulates run-time values.
-For instance, consider this generator
-
-\jwc{need a better example here}
-
-\begin{figure}
-\begin{lstlisting}
-let int_or_zero : int Bq.t =
-  let%bind n = size in
-  if n <= 5 then return 0 else Bq.int
-
-let split_bool (b : bool code) : bool Gen.t =
-  fun _ _ ->  _
-
-\end{lstlisting}
-\caption{??}
-\label{fig:trick-example}
-\end{figure}
-
-\jwc{Describe split --- this section is approximately experts-only, but you can just point at the Andras paper and the things it cites.}
+% \subsubsection{The Trick at Other Types}
+% To write more interesting generators, we also need the ability to generate code that manipulates run-time values.
+% For instance, consider this generator
+
+% \jwc{need a better example here}
+
+% \begin{figure}
+% \begin{lstlisting}
+% let int_or_zero : int Bq.t =
+%   let%bind n = size in
+%   if n <= 5 then return 0 else Bq.int
+
+% let split_bool (b : bool code) : bool Gen.t =
+%   fun _ _ ->  _
+
+% \end{lstlisting}
+% \caption{??}
+% \label{fig:trick-example}
+% \end{figure}
+
+% \jwc{Describe split --- this section is approximately experts-only, but you can just point at the Andras paper and the things it cites.}
 

### 2025-03-19 paper/paper.tex
wip etna

@@
 It is worth noting that the time-to-failure of a given strategy on a given seed is not necessarily representative of 
-the strategy's average time-to-failure over a large number of trials. We normalize by computing relative performance (the ``speedup''). While this works in most cases,
+the strategy's average time-to-failure over a large number of trials. We normalize by computing relative performance, or the ``speedup''. While this works in most cases,
 it is not perfect.
@@
 if both generators fail to find the bug).
-To account for these cases, we repeat the above process over 10 seeds and compute the geometric average of their speedups.\footnote{Although the variance \textit{between} seeds in Etna can be vast, timing results are replicable for a \textit{given} seed.
+To account for these cases, we repeat the above process over 10 seeds.\footnote{Although the variance \textit{between} seeds in Etna can be vast, timing results are replicable for a \textit{given} seed.
 Across 1,000 trials using the same seed,
@@
 
+The average of individual-task speedups for each strategy and benchmark, across all seeds, is shown in \figref{fig:etna}.
+Trends in bug-finding ability closely reproduce trends in performance from \secref{subsection:benchmarking}.
+The distribution of individual-task speedups, shown in \figref{fig:swarm}, tell a more interesting story: benchmarks
+
 \begin{figure}[h]
   \includegraphics[scale=0.333]{speedup.png}
-  \caption{Geometric average of all speedups for each strategy and benchmark, showing that staging leads to better bug-finding ability across the board.
-As in \todo{earlier figure}, benchmarks with higher numbers of binds (BST (Single-Pass); STLC) see more benefit from staging, whereas benchmarks with more sampling (BST) see a greater speedup from \csplitmix.}
+  \caption{Geometric average of all speedups for each strategy and benchmark, showing that staging leads to better bug-finding ability across the board.}
   \label{fig:etna}
@@
   \includegraphics[scale=0.41]{swarm.png}
+  \caption{Distribution of individual-task speedups across strategies and benchmarks. Extreme outliers correspond to large speedups.}
   \label{fig:swarm}
-  \caption{Distribution of individual-task speedups across strategies and benchmarks. Extreme outliers correspond to large speedups.}
 \end{figure}

### 2025-03-19 paper/paper.tex
wibble

--- 2025-03-19 paper/paper.tex
@@
 
-In practice, \texttt{weighted\_union}s is almost always\footnote{
+In practice, \texttt{weighted\_union} is almost always\footnote{
 There are some generators where \texttt{weighted\_union} is passed a list which
@@
     fun size random -> let_insert .< SR.int .~random .~lo .~hi >.
-
 end

### 2025-03-19 paper/paper.tex
Polish: intro

@@
 practitioners often interleave PBT with their software development process, sometimes
-testing their properties every time they save their code. This means that PBT is often run with very short time
-budgets---e.g., between 50 milliseconds and 30 seconds---and it is important
-that bugs can be found in that time-frame.\iflater
-\jwc{phrasing could suggest ``race against time'' aspect.}\hg{Agree, can't see
-how to fit this in yet} \jwc{Also think the phrasing should say something like ``the goal is to fail as fast as possible''}\fi
+testing their properties every time they save their code. This means that PBT is often given a very short time
+window to find bugs --- e.g., between 50 milliseconds and 30 seconds~\cite{inpractice}.
+In other words, the goal is to fail as fast as possible. \jwc{Cut at ``fail fast''}
 
@@
 libraries}, usually expressed as embedded domain-specific languages (eDSLs) %\cntxt{edsls} 
-that
-provide combinators for building and composing generators. Optimizing these
+that provide combinators for building and composing generators. Optimizing these
 libraries is a key opportunity to speed up PBT: if programmers are
 provided with the tools to write fast generators, then they do not themselves
-need to be experts in performance tuning to find bugs fast. \iflater\bcp{Incorporate fail fast somehow.}\fi
+need to be experts in performance tuning to find bugs fast.
 
@@
 \textit{multi-stage programming}, or ``staging,'' which has long been used to
-build DSLs ``without regret''~\cn{}.
+build DSLs without abstraction overhead (``without regret''~\cn{}).
 We apply this technique to generators, completely eliminating the run-time overhead of
@@
 
-To demonstrate the effectiveness of our approach, we apply it to two PBT generator libraries, namely \bq{} \cn{} --- a
-high-quality PBT library in OCaml, authored by software engineers at the trading firm Jane Street ---
-and ScalaCheck \cn{} --- a very mature PBT library and the de facto choice in Scala.
+To demonstrate the effectiveness of our approach, we apply it to two PBT generator libraries, namely \bq{}  --- a
+high-quality PBT library in OCaml, authored by software engineers at the trading firm Jane Street \cn{} ---
+and ScalaCheck --- a very mature PBT library and the de facto choice in Scala \cn{}.
 In both languages, our staged versions are direct drop-in replacements that are
-\textit{100\% semantically equivalent}: that is, given identical random seeds, the same generator written once with \name{}
-and once with the original generator library produces exactly the same sequence of values.
+\textit{100\% semantically equivalent}: that is, given identical random seeds, a generator written with \name{}
+produces exactly the same sequence of values as the same generator written with the original generator library.
 The semantic equivalence between the original generators and \name generators allows us to 
@@
 We evaluate the optimizations in \name with a series of case studies in each
-generator library, and show that the \name generators run faster than their unstaged equivalence.
-Further, we show that this performance benefit\hg{wc?} translates to bug-finding
-ability \jwc{The paper is called ``fail faster'' --- we shouldn't say ``bug finding ability'', we should say something like ``how fast you can squash bugs''.
-This is too passive }by running these case studies in the Etna framework~\cite{etna}, a
+generator library, and show that the \name generators run faster than their unstaged equivalents.
+Further, we show generators that run faster also find bugs faster
+% ability \jwc{The paper is called ``fail faster'' --- we shouldn't say ``bug finding ability'', we should say something like ``how fast you can squash bugs''. This is too passive }
+by running these case studies in the Etna framework~\cite{etna}., a
 platform that simulates real-world PBT usage\hg{Don't love this phrase, I'll
@@
 In summary, we show that monadic PBT generator DSLs incur significant performance costs across languages, and present a general technique for improving their efficiency while preserving their idiomatic abstractions.
+\jwc{needs a bit about RNG --- BCP thoughts?}
 We make the following contributions:
@@
     \item We identify two key sources of inefficiency in PBT generator libraries, which can significantly impact performance: 
-    namely and abstraction overhead and \rand choice. 
-    % \item We apply multi-stage programming to the novel domain of generators. \jwc{Less sure this should be a top-level conrib.} \secref{section:}
-    \item We implement \name{} in both OCaml and Scala 3, showcasing its cross-language applicability.
-    \item We demonstrate that both \name{} and choosing a faster \rand yield substantial performance improvements. Finally, we show
+    abstraction overhead and \rand choice. 
+    \item We present \name{}, a technique based on staging that eliminates the abstraction overhead of generators.
+We implement \name{} in both OCaml and Scala 3, showcasing its cross-language applicability.
+    \item \jwc{We need a top-level contrib about RNG}
+    \item We demonstrate that both \name{} and choosing a faster \rand yield substantial performance improvements to generator running time. Finally, we show
     that these performance improvements extend to significantly improved bug-finding ability. \jwc{This should be sell-ier: 2x speedups, etc. ``bug-finding ability'' isn't a race against time.}
@@
 
-\section{Sources of Inefficiency in Monadic Generator Libraries}
+\section{What are Generator Libraries, and why are they Slow?}
 \label{section:motiv}
-\jwc{In this section we answer: ``What are monadic generator libraries, and why are they slow?''}
-\bcp{That's actually not a bad section title!}
+% \jwc{In this section we answer: ``What are monadic generator libraries, and why are they slow?''}
+% \bcp{That's actually not a bad section title!}
 \todo{Intro of this section}
@@
 allocate memory cells to contain their output, which has a significant performance benefit.
-By building a version that uses much faster ``unboxed'' 64-bit integer arithmetic, we can demonstrate just how much bug-finding
-performance can be improved just by using a more performant randomness library.
+By building a version that uses much faster ``unboxed'' 64-bit integer arithmetic, we can demonstrate just how much faster bugs can be found just by using a more performant randomness library.
 

### 2025-03-19 paper/paper.tex
eval pass

@@
 Our single-pass BST strategy falls somewhere in between: it both samples frequently and makes extensive use of generator combinators.
-\tr{Goal is to replace this paragraph with the ref-counting chart---I think it will be WAY less boring.}
+\tr{Goal is to replace this paragraph with the ref-counting chart---I think it will be WAY less boring.
+But I need to do 3.7 first.}
 
@@
 
-As shown in \figref{fig:scala}, \scalaname achieves an even greater performance gain over ScalaCheck than \name does over \bq.
+As shown in \figref{fig:scala}, \scalaname achieves a greater performance gain over ScalaCheck than \name does over \bq.
 This discrepancy arises from ScalaCheck's representation of generators as functions of type \texttt{size -> seed -> Option[A]}.
@@
   \caption{Time to generate a value using each strategy, averaged over 10,000 trials.
-BQ is \bq; Staged SR is \name using \bq's \texttt{splittable\_random} RNG; Staged CSR is \name using \secref{subsection:faster-rng}'s \csplitmix.
-Performance benefits scale with generation size; lower is better.}
+Performance benefits scale with generation size; lower is better.
+BQ is \bq; Staged SR is \name using \bq's \texttt{splittable\_random} RNG; Staged CSR is \name using \secref{subsection:faster-rng}'s \csplitmix.}
   \label{fig:time}
@@
 
-The average of individual-task speedups for each strategy and benchmark, across all seeds, is shown in \figref{fig:etna}.
-Trends in bug-finding ability closely reproduce trends in performance from \secref{subsection:benchmarking}.
-The distribution of individual-task speedups, shown in \figref{fig:swarm}, tell a more interesting story: benchmarks
-
+The average individual-task speedups for each strategy and benchmark are shown in 
+\figref{fig:etna}, with trends in bug-finding ability closely mirroring those in performance 
+from \secref{subsection:benchmarking}. The distribution of speedups in \figref{fig:swarm} reveals 
+substantial variability, with most tasks clustering near the median and a long tail of outliers 
+achieving much larger gains. STLC shows a more bimodal distribution of speedups
+than the other benchmarks, which we attribute to the heterogenous difficulty of its tasks:  
+some tasks regularly hit the 60-second timeout, while others finish in a fraction of a second. We find that tasks of intermediate difficulty 
+tend to achieve larger speedups, whereas very fast or slow tasks cluster near the mean.
+Overall, 
+these results give strong evidence that our technique consistently benefits bug-finding ability, sometimes drastically.
 \begin{figure}[h]
@@
   \includegraphics[scale=0.41]{swarm.png}
-  \caption{Distribution of individual-task speedups across strategies and benchmarks. Extreme outliers correspond to large speedups.}
+  \caption{Swarm distribution of individual-task speedups across strategies and benchmarks (each black dot is a task).
+Plots with extremely low variance have no swarm overlay; in plots with high variance, outliers correspond to large speedups.
+\tr{Make mean higher-contrast.}}
   \label{fig:swarm}

### 2025-03-19 paper/paper.tex
minor

@@
 \label{section:eval}
-
-\jwc{Initial section here about eval}\tr{todo}
-
-\begin{itemize}
-  \item \textbf{RQ1}: Do generators written using our technique run faster than those written with regular generator combinators?
-  \item \textbf{RQ2}: Do performance speedups translate to better bug-finding ability?
-  % \item \textbf{RQ3}: Does our technique generalize to other strict functional languages?
-\end{itemize}
-
-All experiments were run on a 64-bit Linux machine with 264 GB RAM and an Intel Xeon Platinum 8375C CPU, running Ubuntu 24.04.1 LTS.
-
-The OCaml version of \name and all OCaml benchmarks
+% \jwc{Initial section here about eval}\tr{todo}
+All experiments were run on a 64-bit Linux machine with 264 GB RAM and an Intel Xeon Platinum 8375C CPU, running Ubuntu 24.04.1 LTS. The OCaml version of \name and all OCaml benchmarks
 were compiled with 4.14.1+BER MetaOCaml \texttt{ocamlopt}, the native code compiler for OCaml, using compiler flag \texttt{-O3}.
@@
 
+To evaluate our contributions, this section answers the following research questions:
+\begin{itemize}
+  \item \textbf{RQ1}: Do generators written using our technique run faster than those written with regular generator combinators?
+  \item \textbf{RQ2}: Do performance speedups translate to better bug-finding ability?
+  % \item \textbf{RQ3}: Does our technique generalize to other strict functional languages?
+\end{itemize}
+
 \subsection{Benchmarking Generator Speed}
@@
 Our benchmarks consist of boolean lists, binary search trees (BST), and simply typed lambda calculus (STLC) terms. We implement generators for these benchmarks using a variety of 
-\textit{strategies}, which vary in structure and sophistication. These include a derived generators for BSTs and STLC terms, using the approach outlined in \secref{subsection:type-derived}; 
+\textit{strategies}, which vary in structure and sophistication. These include type-derived generators for BSTs and STLC terms, using the approach outlined in \secref{subsection:type-derived}; 
 two custom BST generators, one that repeatedly inserts values into an initially empty tree and one that materializes the tree in a single pass by generating keys, values, and subtrees at each step; a generator of boolean lists that operates similarly to the single-pass BST strategy; and
@@
 Sampling-heavy generators see greater improvements from faster randomness libraries, while combinator-heavy ones benefit more from staging.
-Since both factors influence performance in different ways, generator libraries should use both to be performant over diverse workloads.
+Since both factors influence performance in different ways, generator libraries should use both in order to handle diverse workloads.
 \subsubsection{Scala}
@@
 Each generator combinator must construct and then pattern-match on these \texttt{Option} values, introducing significant boxing and unboxing overhead at each step.
-Staging allows us to eliminate this overhead, resulting in substantial speedups.
+Staging allows us to eliminate this overhead, resulting in substantial speedups in aggregate.
 
@@
 It is worth noting that the time-to-failure of a given strategy on a given seed is not necessarily representative of 
-the strategy's average time-to-failure over a large number of trials. We normalize by computing relative performance, or the ``speedup''. While this works in most cases,
-it is not perfect.
+the strategy's average time-to-failure over a large number of trials. We normalize by computing the relative performance, or the ``speedup''. This works in most cases,
+but it is not perfect.
 For example, it is theoretically possible

### 2025-03-19 paper/paper.tex
intro name usage fixup, line wrap in eval

--- 2025-03-19 paper/paper.tex
@@
 window to find bugs --- e.g., between 50 milliseconds and 30 seconds~\cite{inpractice}.
-In other words, the goal is to fail as fast as possible. \jwc{Cut at ``fail fast''}
+In other words, the goal is to fail as fast as possible. \jwc{First cut at ``fail fast''}
 
@@
 Secondly, we demonstrate the impact of different choices of \rand on generator
-performance by conducting a controlled comparison. Together, these insights provide a recipe for improving existing generator libraries.
-
-To demonstrate the effectiveness of our approach, we apply it to two PBT generator libraries, namely \bq{}  --- a
-high-quality PBT library in OCaml, authored by software engineers at the trading firm Jane Street \cn{} ---
-and ScalaCheck --- a very mature PBT library and the de facto choice in Scala \cn{}.
-In both languages, our staged versions are direct drop-in replacements that are
-\textit{100\% semantically equivalent}: that is, given identical random seeds, a generator written with \name{}
+performance by conducting a controlled comparison.
+Together, we refer to this recipe for improving generator libraries as \name{} --- \jwc{something pithy here}.
+
+To demonstrate the effectiveness of our approach, we build two different PBT libraries in two different langauges.
+In OCaml, we build \camlname{}, based on \bq{}  --- a
+high-quality PBT library in OCaml, authored by software engineers at the trading firm Jane Street \cn{}.
+In Scala 3, we build \scalaname{}, based on ScalaCheck --- a very mature PBT library and the de facto choice in Scala \cn{}.
+In both languages, our libraries are direct drop-in replacements that are
+\textit{100\% semantically equivalent}: that is, given identical random seeds, a generator written with our technique
 produces exactly the same sequence of values as the same generator written with the original generator library.
-The semantic equivalence between the original generators and \name generators allows us to 
+The semantic equivalence between the original generators and \name{} generators allows us to 
 perform clean, fine-grained comparisons, as differences in generation speed are
@@
 
-We evaluate the optimizations in \name with a series of case studies in each
+We evaluate the optimizations with a series of case studies in each
 generator library, and show that the \name generators run faster than their unstaged equivalents.
@@
 We implement \name{} in both OCaml and Scala 3, showcasing its cross-language applicability.
-    \item \jwc{We need a top-level contrib about RNG}
     \item We demonstrate that both \name{} and choosing a faster \rand yield substantial performance improvements to generator running time. Finally, we show
@@
 Then, in Section~\ref{section:faster-generators}, we explain multi-stage programming,
-iteratively develop the \name{} technique, and discuss options for faster \rand{} \jwc{fix}.
+and develop \name{}, illustrated by \camlname{}.
 Section~\ref{section:eval} evaluates all of the previous work, demonstrating the generator performance improvements
@@
 generators is via an embedded {\em monadic} language. For the purposes of this
-paper, we use the syntax from \bq, which is presented in
+paper, we use the syntax from OCaml's \bq library, which is presented in
 Figure~\ref{fig:bq-api}, but these kinds of DSLs can also be found in languages
-like Haskell, Scala, Rust, and many more.
+like Haskell, Scala, Python, and many more.
 The library provides some basic generators, for
@@
 monadic syntax; this is the one provided by Jane Street's libraries.}, this
-looks like \texttt{let\%bind x = e in e'}, which is a macro that desugars to
+looks like \texttt{let\%bind x = e in e'}, which desugars to
 \texttt{bind e (fun x => e')}. Figure~\ref{fig:simple-macro} shows the same
-generator as before, now written with macros.
-
-\subsubsection{Subtleties of Bind}
-The \texttt{bind} function in monadic generator libraries sequences generators
-with a call-by-value semantics, regardless of the calling convention of the
-language in which it is implemented. Concretely, this means that sampling
-\begin{center}
-  \texttt{let\%bind x = gen\_int 0 10 in return (x + x)}
-\end{center}
-always produces an even number; \texttt{x} is only sampled once (when the
-\texttt{bind} is evaluated) and from that point on it is a pure value. This is
-in contrast to the semantics of some probabilistic programming languages that
-may have call-by-name binders. This is important because \hg{Joe why are we
-talking about this here? What are we going to use this fact for?}.
-
-Another subtle aspect of \texttt{bind} is that many implementations mutate the
-internal state of the random number generator\footnote{The original QuickCheck
-implementation in Haskell uses a splittable random number generator~\cn, and
-\texttt{bind} splits the seed.} This means that generators do not always obey
-equivalences that one might expect from a monadic program. For example,
-\begin{center}
-  \texttt{let\%bind \_x = return () in gen\_int 0 10}
-\end{center}
-does not necessarily produce the same sequence of values as \texttt{gen\_int 0
-10}; the \texttt{bind} updates the random seed in the background before
-geenrating the integer, even though it looks like a no-op. Critically, while
-these generators do not produce the same sequence of values, they do produce the
-same {\em distributions} of values. We say that generators obey standard monadic
-equalities up to ``distributional equivalence.''
-In Section~\hg{todo}, we discuss why distributional equivalence is often weaker
-than one might like when evaluating the differences between
-generators.\hg{Update this}
-
-\subsubsection{Weighted Union and Other Conveniences}
+generator as before, now written the monadic syntax.
+
+% \subsubsection{Subtleties of Bind}
+% The \texttt{bind} function in monadic generator libraries sequences generators
+% with a call-by-value semantics, regardless of the calling convention of the
+% language in which it is implemented. Concretely, this means that sampling
+% \begin{center}
+%   \texttt{let\%bind x = gen\_int 0 10 in return (x + x)}
+% \end{center}
+% always produces an even number; \texttt{x} is only sampled once (when the
+% \texttt{bind} is evaluated) and from that point on it is a pure integer value. This is
+% in contrast to the semantics of some probabilistic programming languages that
+% may have call-by-name sampling semantics \cn{}. This is important because \hg{Joe why are we
+% talking about this here? What are we going to use this fact for?}.
+% \jwc{Well this is why directly inlining the binds doesn't work.}
+
+% Another subtle aspect of \texttt{bind} is that many implementations mutate the
+% internal state of the random number generator\footnote{The original QuickCheck
+% implementation in Haskell uses a splittable random number generator~\cn, and
+% \texttt{bind} splits the seed.} This means that generators do not always obey
+% equivalences that one might expect from a monadic program. For example,
+% \begin{center}
+%   \texttt{let\%bind \_x = return () in gen\_int 0 10}
+% \end{center}
+% does not necessarily produce the same sequence of values as \texttt{gen\_int 0
+% 10}; the \texttt{bind} updates the random seed in the background before
+% geenrating the integer, even though it looks like a no-op. Critically, while
+% these generators do not produce the same sequence of values, they do produce the
+% same {\em distributions} of values. We say that generators obey standard monadic
+% equalities up to ``distributional equivalence.''
+% In Section~\hg{todo}, we discuss why distributional equivalence is often weaker
+% than one might like when evaluating the differences between
+% generators.\hg{Update this}
+
 Generator libraries also include other functions that make generator
@@
 
-For this paper, we have implemented staged PBT libraries in both OCaml --- using the MetaOCaml fork of the OCaml compiler \cn --- and Scala 3 --- using
-\texttt{scala.quoted}. For presentation purposes, all staged code presented in the body of this paper is written in OCaml.
-However, we believe all of the concepts presented are portable across languages with staging support, and we have reimplemented \name in Scala 3.
+For this paper, we have implemented staged PBT libraries both in OCaml --- using the MetaOCaml fork of the OCaml compiler \cn --- and Scala 3 --- using
+\texttt{scala.quoted}. For presentation purposes, all staged code presented in the body of this paper is \camlname{} code written in OCaml:
+\scalaname{} differs almost entirely in syntax alone.
 
@@
 All experiments were run on a 64-bit Linux machine with 264 GB RAM and an Intel Xeon Platinum 8375C CPU, running Ubuntu 24.04.1 LTS.
-
-The OCaml version of \name and all OCaml benchmarks
+\camlname{} and all OCaml benchmarks
 were compiled with 4.14.1+BER MetaOCaml \texttt{ocamlopt}, the native code compiler for OCaml, using compiler flag \texttt{-O3}.
 The baseline OCaml generators were written with \bq{} 0.16.
-The Scala 3 version of \name and all Scala benchmarks
+\scalaname{} and all Scala benchmarks
 were run on Scala 3.6.3
@@
 \label{subsection:benchmarking}
-To answer \textbf{RQ1}, we implement benchmarks in both \name and \bq. We vary the choice of \rand in our \name generators, 
-using both \bq's default \texttt{splittable\_random} and our improved SplitMix (\csplitmix), as discussed in \secref{subsection:faster-rng}.
-
-Our benchmarks consist of boolean lists, binary search trees (BST), and simply typed lambda calculus (STLC) terms. We implement generators for these benchmarks using a variety of 
-\textit{strategies}, which vary in structure and sophistication. These include a derived generators for BSTs and STLC terms, using the approach outlined in \secref{subsection:type-derived}; 
-two custom BST generators, one that repeatedly inserts values into an initially empty tree and one that materializes the tree in a single pass by generating keys, values, and subtrees at each step; a generator of boolean lists that operates similarly to the single-pass BST strategy; and
-a custom, correct-by-construction generator of STLC terms, whose behavior is the most complex.
-
-The runtime of these generators over a range of sizes is shown in \figref{fig:time}.
-Staging provides only a negligible speedup for type-derived and insertion-based BST strategies, whose overhead is dominated by random sampling.
-In this case, performance improvement is primarily attributable the \csplitmix \rand.
-In contrast, STLC strategies benefit more from staging, as they involve fewer sampling operations but rely heavily on generator combinators.
-The same holds for boolean lists, as sampling booleans is inexpensive relative to integers.
-Our single-pass BST strategy falls somewhere in between: it both samples frequently and makes extensive use of generator combinators.
-\tr{Goal is to replace this paragraph with the ref-counting chart---I think it will be WAY less boring.}
-
-These results show that staging and \rand choice serve distinct but complementary roles in generator performance.
-Sampling-heavy generators see greater improvements from faster randomness libraries, while combinator-heavy ones benefit more from staging.
-Since both factors influence performance in different ways, generator libraries should use both to be performant over diverse workloads.
-\subsubsection{Scala}
-To show that the benefit of staging extends to other strict functional languages, we apply it to ScalaCheck, a widely-used PBT library written in Scala.
-We implement a subset of our strategies from \secref{subsection:benchmarking} in both standard and staged ScalaCheck (``\scalaname'').
-
-As shown in \figref{fig:scala}, \scalaname achieves an even greater performance gain over ScalaCheck than \name does over \bq.
-This discrepancy arises from ScalaCheck's representation of generators as functions of type \texttt{size -> seed -> Option[A]}.
-Each generator combinator must construct and then pattern-match on these \texttt{Option} values, introducing significant boxing and unboxing overhead at each step.
-Staging allows us to eliminate this overhead, resulting in substantial speedups.
-
-These results illustrate that staging is a powerful \textit{and} general optimization technique, capable of delivering performance gains in strict functional languages beyond OCaml.
+To answer \textbf{RQ1}, we implement benchmarks in both \name and \bq. We vary
+the choice of \rand in our \name generators, using both \bq's default
+\texttt{splittable\_random} and our improved SplitMix (\csplitmix), as discussed
+in \secref{subsection:faster-rng}.
+
+Our benchmarks consist of boolean lists, binary search trees (BST), and simply
+typed lambda calculus (STLC) terms. We implement generators for these benchmarks
+using a variety of \textit{strategies}, which vary in structure and
+sophistication. These include a derived generators for BSTs and STLC terms,
+using the approach outlined in \secref{subsection:type-derived}; two custom BST
+generators, one that repeatedly inserts values into an initially empty tree and
+one that materializes the tree in a single pass by generating keys, values, and
+subtrees at each step; a generator of boolean lists that operates similarly to
+the single-pass BST strategy; and a custom, correct-by-construction generator of
+STLC terms, whose behavior is the most complex.
+
+The runtime of these generators over a range of sizes is shown in
+\figref{fig:time}. Staging provides only a negligible speedup for type-derived
+and insertion-based BST strategies, whose overhead is dominated by random
+sampling. In this case, performance improvement is primarily attributable the
+\csplitmix \rand.  In contrast, STLC strategies benefit more from staging, as
+they involve fewer sampling operations but rely heavily on generator
+combinators. The same holds for boolean lists, as sampling booleans is
+inexpensive relative to integers. Our single-pass BST strategy falls somewhere
+in between: it both samples frequently and makes extensive use of generator
+combinators. \tr{Goal is to replace this paragraph with the ref-counting
+chart---I think it will be WAY less boring.}
+
+These results show that staging and \rand choice serve distinct but
+complementary roles in generator performance. Sampling-heavy generators see
+greater improvements from faster randomness libraries, while combinator-heavy
+ones benefit more from staging. Since both factors influence performance in
+different ways, generator libraries should use both to be performant over
+diverse workloads.
+
+\subsubsection{Scala} To show that the benefit of staging
+extends to other strict functional languages, we apply it to ScalaCheck, a
+widely-used PBT library written in Scala. We implement a subset of our
+strategies from \secref{subsection:benchmarking} in both standard and staged
+ScalaCheck (``\scalaname'').
+
+As shown in \figref{fig:scala}, \scalaname achieves an even greater performance
+gain over ScalaCheck than \name does over \bq. This discrepancy arises from
+ScalaCheck's representation of generators as functions of type \texttt{size ->
+seed -> Option[A]}. Each generator combinator must construct and then
+pattern-match on these \texttt{Option} values, introducing significant boxing
+and unboxing overhead at each step.  Staging allows us to eliminate this
+overhead, resulting in substantial speedups.
+
+These results illustrate that staging is a powerful \textit{and} general
+optimization technique, capable of delivering performance gains in strict
+functional languages beyond OCaml.
 \begin{figure}[h]
@@
 \subsection{Benchmarking Bug-Finding Ability}
-To answer \textbf{RQ2}, we evaluate the bug-finding ability of our BST and STLC case studies using Etna, a platform for assessing and comparing PBT techniques. 
-Etna allows users to measure the effectiveness of different generator implementations by injecting bugs into the system under test and recording the time taken for a relevant 
-property to fail in response.
-For example, in a tree with an ordering property, a bug injection might consist of inserting a value into the tree that violates that ordering.
-Each case study includes a diverse set of \textit{tasks}, where a task consists of a specific bug-property pair designed to test different aspects of the system (e.g., BST
-includes tasks that test insertion, deletion, and union operations on binary search trees). In our case, BST has 43 tasks, and STLC has 18.
+To answer \textbf{RQ2}, we evaluate the bug-finding ability of our BST and STLC
+case studies using Etna, a platform for assessing and comparing PBT techniques. 
+Etna allows users to measure the effectiveness of different generator
+implementations by injecting bugs into the system under test and recording the
+time taken for a relevant property to fail in response. For example, in a tree
+with an ordering property, a bug injection might consist of inserting a value
+into the tree that violates that ordering. Each case study includes a diverse
+set of \textit{tasks}, where a task consists of a specific bug-property pair
+designed to test different aspects of the system (e.g., BST includes tasks that
+test insertion, deletion, and union operations on binary search trees). In our
+case, BST has 43 tasks, and STLC has 18.
 
 Strategies for BST and STLC are unchanged from \secref{subsection:benchmarking}.
-To ensure a fair comparison, we initialize each strategy with the same random seed, ensuring that they produce identical sequences of values.
-We run each strategy on all tasks using a 60-second timeout.
-If a bug is not found within this limit, the task is considered unsolved. 
-Tasks where all strategies fail to find the bug are excluded from our dataset, 
-as they provide no basis for comparison. 
-Similarly, we exclude tasks where the BQ strategy completes in under 5ms,
- as such negligible runtimes do not yield meaningful insights into relative performance. These filters ensure that reported speedups are based on cases where optimization has a measurable impact.
-
-It is worth noting that the time-to-failure of a given strategy on a given seed is not necessarily representative of 
-the strategy's average time-to-failure over a large number of trials. We normalize by computing relative performance, or the ``speedup''. While this works in most cases,
-it is not perfect.
-For example, it is theoretically possible
-to choose a seed such that the \textit{first} value produced by a generator discovers the bug, which would elide any differential speedup (and likewise
-if both generators fail to find the bug).
-To account for these cases, we repeat the above process over 10 seeds.\footnote{Although the variance \textit{between} seeds in Etna can be vast, timing results are replicable for a \textit{given} seed.
-Across 1,000 trials using the same seed,
+To ensure a fair comparison, we initialize each strategy with the same random
+seed, ensuring that they produce identical sequences of values. We run each
+strategy on all tasks using a 60-second timeout.  If a bug is not found within
+this limit, the task is considered unsolved.  Tasks where all strategies fail to
+find the bug are excluded from our dataset, as they provide no basis for
+comparison.  Similarly, we exclude tasks where the BQ strategy completes in
+under 5ms, as such negligible runtimes do not yield meaningful insights into
+relative performance. These filters ensure that reported speedups are based on
+cases where optimization has a measurable impact.
+
+It is worth noting that the time-to-failure of a given strategy on a given seed
+is not necessarily representative of the strategy's average time-to-failure over
+a large number of trials. We normalize by computing relative performance, or the
+``speedup''. While this works in most cases, it is not perfect.  For example, it
+is theoretically possible to choose a seed such that the \textit{first} value
+produced by a generator discovers the bug, which would elide any differential
+speedup (and likewise if both generators fail to find the bug).  To account for
+these cases, we repeat the above process over 10 seeds.\footnote{Although the
+variance \textit{between} seeds in Etna can be vast, timing results are
+replicable for a \textit{given} seed. Across 1,000 trials using the same seed, 
 the observed variance in time-to-failure was less than a nanosecond.}
 
-The average of individual-task speedups for each strategy and benchmark, across all seeds, is shown in \figref{fig:etna}.
-Trends in bug-finding ability closely reproduce trends in performance from \secref{subsection:benchmarking}.
-The distribution of individual-task speedups, shown in \figref{fig:swarm}, tell a more interesting story: benchmarks
+The average of individual-task speedups for each strategy and benchmark, across
+all seeds, is shown in \figref{fig:etna}. Trends in bug-finding ability closely
+reproduce trends in performance from \secref{subsection:benchmarking}. The
+distribution of individual-task speedups, shown in \figref{fig:swarm}, tell a
+more interesting story: benchmarks
 
@@
   sometimes to build eDSLs in contexts where one does not want to rely on the
-  optimizer's heuristics.  In short: it is possible to port \name{} to Haskell
+  optimizer's heuristics.  In short: it is possible to apply \name{} to Haskell
   QuickCheck, but the benefits would be much less pronounced.
@@
   For this reason, \name{} does not seem directly applicable to any of the PBT libraries in Rust.
-
-  \item \textbf{Python} is home to Hypothesis, the most widely used PBT library in any language.
-  \jwc{More here}
-
 \end{itemize}

### 2025-03-19 paper/paper.tex
fix

@@
 \label{subsection:benchmarking}
-To answer \textbf{RQ1}, we implement benchmarks in both \name and \bq. We vary
-the choice of \rand in our \name generators, using both \bq's default
-\texttt{splittable\_random} and our improved SplitMix (\csplitmix), as discussed
-in \secref{subsection:faster-rng}.
-
-Our benchmarks consist of boolean lists, binary search trees (BST), and simply
-typed lambda calculus (STLC) terms. We implement generators for these benchmarks
-using a variety of \textit{strategies}, which vary in structure and
-sophistication. These include a derived generators for BSTs and STLC terms,
-using the approach outlined in \secref{subsection:type-derived}; two custom BST
-generators, one that repeatedly inserts values into an initially empty tree and
-one that materializes the tree in a single pass by generating keys, values, and
-subtrees at each step; a generator of boolean lists that operates similarly to
-the single-pass BST strategy; and a custom, correct-by-construction generator of
-STLC terms, whose behavior is the most complex.
-
-The runtime of these generators over a range of sizes is shown in
-\figref{fig:time}. Staging provides only a negligible speedup for type-derived
-and insertion-based BST strategies, whose overhead is dominated by random
-sampling. In this case, performance improvement is primarily attributable the
-\csplitmix \rand.  In contrast, STLC strategies benefit more from staging, as
-they involve fewer sampling operations but rely heavily on generator
-combinators. The same holds for boolean lists, as sampling booleans is
-inexpensive relative to integers. Our single-pass BST strategy falls somewhere
-in between: it both samples frequently and makes extensive use of generator
-combinators. \tr{Goal is to replace this paragraph with the ref-counting
-chart---I think it will be WAY less boring.}
-
-These results show that staging and \rand choice serve distinct but
-complementary roles in generator performance. Sampling-heavy generators see
-greater improvements from faster randomness libraries, while combinator-heavy
-ones benefit more from staging. Since both factors influence performance in
-different ways, generator libraries should use both to be performant over
-diverse workloads.
-
-\subsubsection{Scala} To show that the benefit of staging
-extends to other strict functional languages, we apply it to ScalaCheck, a
-widely-used PBT library written in Scala. We implement a subset of our
-strategies from \secref{subsection:benchmarking} in both standard and staged
-ScalaCheck (``\scalaname'').
-
-As shown in \figref{fig:scala}, \scalaname achieves an even greater performance
-gain over ScalaCheck than \name does over \bq. This discrepancy arises from
-ScalaCheck's representation of generators as functions of type \texttt{size ->
-seed -> Option[A]}. Each generator combinator must construct and then
-pattern-match on these \texttt{Option} values, introducing significant boxing
-and unboxing overhead at each step.  Staging allows us to eliminate this
-overhead, resulting in substantial speedups.
-
-These results illustrate that staging is a powerful \textit{and} general
-optimization technique, capable of delivering performance gains in strict
-functional languages beyond OCaml.
+To answer \textbf{RQ1}, we implement benchmarks in both \name and \bq. We vary the choice of \rand in our \name generators, 
+using both \bq's default \texttt{splittable\_random} and our improved SplitMix (\csplitmix), as discussed in \secref{subsection:faster-rng}.
+
+Our benchmarks consist of boolean lists, binary search trees (BST), and simply typed lambda calculus (STLC) terms. We implement generators for these benchmarks using a variety of 
+\textit{strategies}, which vary in structure and sophistication. These include a derived generators for BSTs and STLC terms, using the approach outlined in \secref{subsection:type-derived}; 
+two custom BST generators, one that repeatedly inserts values into an initially empty tree and one that materializes the tree in a single pass by generating keys, values, and subtrees at each step; a generator of boolean lists that operates similarly to the single-pass BST strategy; and
+a custom, correct-by-construction generator of STLC terms, whose behavior is the most complex.
+
+The runtime of these generators over a range of sizes is shown in \figref{fig:time}.
+Staging provides only a negligible speedup for type-derived and insertion-based BST strategies, whose overhead is dominated by random sampling.
+In this case, performance improvement is primarily attributable the \csplitmix \rand.
+In contrast, STLC strategies benefit more from staging, as they involve fewer sampling operations but rely heavily on generator combinators.
+The same holds for boolean lists, as sampling booleans is inexpensive relative to integers.
+Our single-pass BST strategy falls somewhere in between: it both samples frequently and makes extensive use of generator combinators.
+\tr{Goal is to replace this paragraph with the ref-counting chart---I think it will be WAY less boring.}
+
+These results show that staging and \rand choice serve distinct but complementary roles in generator performance.
+Sampling-heavy generators see greater improvements from faster randomness libraries, while combinator-heavy ones benefit more from staging.
+Since both factors influence performance in different ways, generator libraries should use both to be performant over diverse workloads.
+\subsubsection{Scala}
+To show that the benefit of staging extends to other strict functional languages, we apply it to ScalaCheck, a widely-used PBT library written in Scala.
+We implement a subset of our strategies from \secref{subsection:benchmarking} in both standard and staged ScalaCheck (``\scalaname'').
+
+As shown in \figref{fig:scala}, \scalaname achieves an even greater performance gain over ScalaCheck than \name does over \bq.
+This discrepancy arises from ScalaCheck's representation of generators as functions of type \texttt{size -> seed -> Option[A]}.
+Each generator combinator must construct and then pattern-match on these \texttt{Option} values, introducing significant boxing and unboxing overhead at each step.
+Staging allows us to eliminate this overhead, resulting in substantial speedups.
+
+These results illustrate that staging is a powerful \textit{and} general optimization technique, capable of delivering performance gains in strict functional languages beyond OCaml.
 \begin{figure}[h]
@@
 \subsection{Benchmarking Bug-Finding Ability}
-To answer \textbf{RQ2}, we evaluate the bug-finding ability of our BST and STLC
-case studies using Etna, a platform for assessing and comparing PBT techniques. 
-Etna allows users to measure the effectiveness of different generator
-implementations by injecting bugs into the system under test and recording the
-time taken for a relevant property to fail in response. For example, in a tree
-with an ordering property, a bug injection might consist of inserting a value
-into the tree that violates that ordering. Each case study includes a diverse
-set of \textit{tasks}, where a task consists of a specific bug-property pair
-designed to test different aspects of the system (e.g., BST includes tasks that
-test insertion, deletion, and union operations on binary search trees). In our
-case, BST has 43 tasks, and STLC has 18.
+To answer \textbf{RQ2}, we evaluate the bug-finding ability of our BST and STLC case studies using Etna, a platform for assessing and comparing PBT techniques. 
+Etna allows users to measure the effectiveness of different generator implementations by injecting bugs into the system under test and recording the time taken for a relevant 
+property to fail in response.
+For example, in a tree with an ordering property, a bug injection might consist of inserting a value into the tree that violates that ordering.
+Each case study includes a diverse set of \textit{tasks}, where a task consists of a specific bug-property pair designed to test different aspects of the system (e.g., BST
+includes tasks that test insertion, deletion, and union operations on binary search trees). In our case, BST has 43 tasks, and STLC has 18.
 
 Strategies for BST and STLC are unchanged from \secref{subsection:benchmarking}.
-To ensure a fair comparison, we initialize each strategy with the same random
-seed, ensuring that they produce identical sequences of values. We run each
-strategy on all tasks using a 60-second timeout.  If a bug is not found within
-this limit, the task is considered unsolved.  Tasks where all strategies fail to
-find the bug are excluded from our dataset, as they provide no basis for
-comparison.  Similarly, we exclude tasks where the BQ strategy completes in
-under 5ms, as such negligible runtimes do not yield meaningful insights into
-relative performance. These filters ensure that reported speedups are based on
-cases where optimization has a measurable impact.
-
-It is worth noting that the time-to-failure of a given strategy on a given seed
-is not necessarily representative of the strategy's average time-to-failure over
-a large number of trials. We normalize by computing relative performance, or the
-``speedup''. While this works in most cases, it is not perfect.  For example, it
-is theoretically possible to choose a seed such that the \textit{first} value
-produced by a generator discovers the bug, which would elide any differential
-speedup (and likewise if both generators fail to find the bug).  To account for
-these cases, we repeat the above process over 10 seeds.\footnote{Although the
-variance \textit{between} seeds in Etna can be vast, timing results are
-replicable for a \textit{given} seed. Across 1,000 trials using the same seed, 
+To ensure a fair comparison, we initialize each strategy with the same random seed, ensuring that they produce identical sequences of values.
+We run each strategy on all tasks using a 60-second timeout.
+If a bug is not found within this limit, the task is considered unsolved. 
+Tasks where all strategies fail to find the bug are excluded from our dataset, 
+as they provide no basis for comparison. 
+Similarly, we exclude tasks where the BQ strategy completes in under 5ms,
+ as such negligible runtimes do not yield meaningful insights into relative performance. These filters ensure that reported speedups are based on cases where optimization has a measurable impact.
+
+It is worth noting that the time-to-failure of a given strategy on a given seed is not necessarily representative of 
+the strategy's average time-to-failure over a large number of trials. We normalize by computing relative performance, or the ``speedup''. While this works in most cases,
+it is not perfect.
+For example, it is theoretically possible
+to choose a seed such that the \textit{first} value produced by a generator discovers the bug, which would elide any differential speedup (and likewise
+if both generators fail to find the bug).
+To account for these cases, we repeat the above process over 10 seeds.\footnote{Although the variance \textit{between} seeds in Etna can be vast, timing results are replicable for a \textit{given} seed.
+Across 1,000 trials using the same seed,
 the observed variance in time-to-failure was less than a nanosecond.}
 
-The average of individual-task speedups for each strategy and benchmark, across
-all seeds, is shown in \figref{fig:etna}. Trends in bug-finding ability closely
-reproduce trends in performance from \secref{subsection:benchmarking}. The
-distribution of individual-task speedups, shown in \figref{fig:swarm}, tell a
-more interesting story: benchmarks
+The average of individual-task speedups for each strategy and benchmark, across all seeds, is shown in \figref{fig:etna}.
+Trends in bug-finding ability closely reproduce trends in performance from \secref{subsection:benchmarking}.
+The distribution of individual-task speedups, shown in \figref{fig:swarm}, tell a more interesting story: benchmarks
 

### 2025-03-19 paper/paper.tex

@@
 \label{section:eval}
-
-\jwc{Initial section here about eval}\tr{todo}
-
-\begin{itemize}
-  \item \textbf{RQ1}: Do generators written using our technique run faster than those written with regular generator combinators?
-  \item \textbf{RQ2}: Do performance speedups translate to better bug-finding ability?
-  % \item \textbf{RQ3}: Does our technique generalize to other strict functional languages?
-\end{itemize}
-
-All experiments were run on a 64-bit Linux machine with 264 GB RAM and an Intel Xeon Platinum 8375C CPU, running Ubuntu 24.04.1 LTS.
-\camlname{} and all OCaml benchmarks
+% \jwc{Initial section here about eval}\tr{todo}
+All experiments were run on a 64-bit Linux machine with 264 GB RAM and an Intel Xeon Platinum 8375C CPU, running Ubuntu 24.04.1 LTS. The OCaml version of \name and all OCaml benchmarks
 were compiled with 4.14.1+BER MetaOCaml \texttt{ocamlopt}, the native code compiler for OCaml, using compiler flag \texttt{-O3}.
@@
 
-As shown in \figref{fig:scala}, \scalaname achieves an even greater performance gain over ScalaCheck than \name does over \bq.
+As shown in \figref{fig:scala}, \scalaname achieves a greater performance gain over ScalaCheck than \name does over \bq.
 This discrepancy arises from ScalaCheck's representation of generators as functions of type \texttt{size -> seed -> Option[A]}.
 Each generator combinator must construct and then pattern-match on these \texttt{Option} values, introducing significant boxing and unboxing overhead at each step.
-Staging allows us to eliminate this overhead, resulting in substantial speedups.
+Staging allows us to eliminate this overhead, resulting in substantial speedups in aggregate.
 

### 2025-03-19 paper/paper.tex
fix wrapping

@@
 \label{subsection:benchmarking}
-To answer \textbf{RQ1}, we implement benchmarks in both \name and \bq. We vary the choice of \rand in our \name generators, 
-using both \bq's default \texttt{splittable\_random} and our improved SplitMix (\csplitmix), as discussed in \secref{subsection:faster-rng}.
-
-Our benchmarks consist of boolean lists, binary search trees (BST), and simply typed lambda calculus (STLC) terms. We implement generators for these benchmarks using a variety of 
-\textit{strategies}, which vary in structure and sophistication. These include type-derived generators for BSTs and STLC terms, using the approach outlined in \secref{subsection:type-derived}; 
-two custom BST generators, one that repeatedly inserts values into an initially empty tree and one that materializes the tree in a single pass by generating keys, values, and subtrees at each step; a generator of boolean lists that operates similarly to the single-pass BST strategy; and
-a custom, correct-by-construction generator of STLC terms, whose behavior is the most complex.
-
-The runtime of these generators over a range of sizes is shown in \figref{fig:time}.
-Staging provides only a negligible speedup for type-derived and insertion-based BST strategies, whose overhead is dominated by random sampling.
-In this case, performance improvement is primarily attributable the \csplitmix \rand.
-In contrast, STLC strategies benefit more from staging, as they involve fewer sampling operations but rely heavily on generator combinators.
-The same holds for boolean lists, as sampling booleans is inexpensive relative to integers.
-Our single-pass BST strategy falls somewhere in between: it both samples frequently and makes extensive use of generator combinators.
-\tr{Goal is to replace this paragraph with the ref-counting chart---I think it will be WAY less boring.
-But I need to do 3.7 first.}
-
-These results show that staging and \rand choice serve distinct but complementary roles in generator performance.
-Sampling-heavy generators see greater improvements from faster randomness libraries, while combinator-heavy ones benefit more from staging.
-Since both factors influence performance in different ways, generator libraries should use both in order to handle diverse workloads.
-\subsubsection{Scala}
-To show that the benefit of staging extends to other strict functional languages, we apply it to ScalaCheck, a widely-used PBT library written in Scala.
-We implement a subset of our strategies from \secref{subsection:benchmarking} in both standard and staged ScalaCheck (``\scalaname'').
-
-As shown in \figref{fig:scala}, \scalaname achieves a greater performance gain over ScalaCheck than \name does over \bq.
-This discrepancy arises from ScalaCheck's representation of generators as functions of type \texttt{size -> seed -> Option[A]}.
-Each generator combinator must construct and then pattern-match on these \texttt{Option} values, introducing significant boxing and unboxing overhead at each step.
-Staging allows us to eliminate this overhead, resulting in substantial speedups in aggregate.
+To answer \textbf{RQ1}, we implement benchmarks in both \name and \bq. We vary
+the choice of \rand in our \name generators, using both \bq's default
+\texttt{splittable\_random} and our improved SplitMix (\csplitmix), as discussed
+in \secref{subsection:faster-rng}.
+
+Our benchmarks consist of boolean lists, binary search trees (BST), and simply
+typed lambda calculus (STLC) terms. We implement generators for these benchmarks
+using a variety of \textit{strategies}, which vary in structure and
+sophistication. These include type-derived generators for BSTs and STLC terms,
+using the approach outlined in \secref{subsection:type-derived}; two custom BST
+generators, one that repeatedly inserts values into an initially empty tree and
+one that materializes the tree in a single pass by generating keys, values, and
+subtrees at each step; a generator of boolean lists that operates similarly to
+the single-pass BST strategy; and a custom, correct-by-construction generator of
+STLC terms, whose behavior is the most complex.
+
+The runtime of these generators over a range of sizes is shown in
+\figref{fig:time}. Staging provides only a negligible speedup for type-derived
+and insertion-based BST strategies, whose overhead is dominated by random
+sampling. In this case, performance improvement is primarily attributable the
+\csplitmix \rand.  In contrast, STLC strategies benefit more from staging, as
+they involve fewer sampling operations but rely heavily on generator
+combinators. The same holds for boolean lists, as sampling booleans is
+inexpensive relative to integers. Our single-pass BST strategy falls somewhere
+in between: it both samples frequently and makes extensive use of generator
+combinators. \tr{Goal is to replace this paragraph with the ref-counting
+chart---I think it will be WAY less boring. But I need to do 3.7 first.}
+
+These results show that staging and \rand choice serve distinct but
+complementary roles in generator performance. Sampling-heavy generators see
+greater improvements from faster randomness libraries, while combinator-heavy
+ones benefit more from staging. Since both factors influence performance in
+different ways, generator libraries should use both in order to handle diverse
+workloads.
+
+\subsubsection{Scala} To show that the benefit of staging extends to
+other strict functional languages, we apply it to ScalaCheck, a widely-used PBT
+library written in Scala. We implement a subset of our strategies from
+\secref{subsection:benchmarking} in both standard and staged ScalaCheck
+(``\scalaname'').
+
+As shown in \figref{fig:scala}, \scalaname achieves a greater performance gain
+over ScalaCheck than \name does over \bq. This discrepancy arises from
+ScalaCheck's representation of generators as functions of type \texttt{size ->
+seed -> Option[A]}. Each generator combinator must construct and then
+pattern-match on these \texttt{Option} values, introducing significant boxing
+and unboxing overhead at each step.  Staging allows us to eliminate this
+overhead, resulting in substantial speedups in aggregate.
 
@@
 \subsection{Benchmarking Bug-Finding Ability}
-To answer \textbf{RQ2}, we evaluate the bug-finding ability of our BST and STLC case studies using Etna, a platform for assessing and comparing PBT techniques. 
-Etna allows users to measure the effectiveness of different generator implementations by injecting bugs into the system under test and recording the time taken for a relevant 
-property to fail in response.
-For example, in a tree with an ordering property, a bug injection might consist of inserting a value into the tree that violates that ordering.
-Each case study includes a diverse set of \textit{tasks}, where a task consists of a specific bug-property pair designed to test different aspects of the system (e.g., BST
-includes tasks that test insertion, deletion, and union operations on binary search trees). In our case, BST has 43 tasks, and STLC has 18.
+To answer \textbf{RQ2}, we evaluate the bug-finding ability of our BST and STLC
+case studies using Etna, a platform for assessing and comparing PBT techniques. 
+Etna allows users to measure the effectiveness of different generator
+implementations by injecting bugs into the system under test and recording the
+time taken for a relevant property to fail in response. For example, in a tree
+with an ordering property, a bug injection might consist of inserting a value
+into the tree that violates that ordering. Each case study includes a diverse
+set of \textit{tasks}, where a task consists of a specific bug-property pair
+designed to test different aspects of the system (e.g., BST includes tasks that
+test insertion, deletion, and union operations on binary search trees). In our
+case, BST has 43 tasks, and STLC has 18.
 
 Strategies for BST and STLC are unchanged from \secref{subsection:benchmarking}.
-To ensure a fair comparison, we initialize each strategy with the same random seed, ensuring that they produce identical sequences of values.
-We run each strategy on all tasks using a 60-second timeout.
-If a bug is not found within this limit, the task is considered unsolved. 
-Tasks where all strategies fail to find the bug are excluded from our dataset, 
-as they provide no basis for comparison. 
-Similarly, we exclude tasks where the BQ strategy completes in under 5ms,
- as such negligible runtimes do not yield meaningful insights into relative performance. These filters ensure that reported speedups are based on cases where optimization has a measurable impact.
-
-It is worth noting that the time-to-failure of a given strategy on a given seed is not necessarily representative of 
-the strategy's average time-to-failure over a large number of trials. We normalize by computing the relative performance, or the ``speedup''. This works in most cases,
-but it is not perfect.
-For example, it is theoretically possible
-to choose a seed such that the \textit{first} value produced by a generator discovers the bug, which would elide any differential speedup (and likewise
-if both generators fail to find the bug).
-To account for these cases, we repeat the above process over 10 seeds.\footnote{Although the variance \textit{between} seeds in Etna can be vast, timing results are replicable for a \textit{given} seed.
-Across 1,000 trials using the same seed,
+To ensure a fair comparison, we initialize each strategy with the same random
+seed, ensuring that they produce identical sequences of values. We run each
+strategy on all tasks using a 60-second timeout.  If a bug is not found within
+this limit, the task is considered unsolved.  Tasks where all strategies fail to
+find the bug are excluded from our dataset, as they provide no basis for
+comparison.  Similarly, we exclude tasks where the BQ strategy completes in
+under 5ms, as such negligible runtimes do not yield meaningful insights into
+relative performance. These filters ensure that reported speedups are based on
+cases where optimization has a measurable impact.
+
+It is worth noting that the time-to-failure of a given strategy on a given seed
+is not necessarily representative of the strategy's average time-to-failure over
+a large number of trials. We normalize by computing the relative performance, or
+the ``speedup''. This works in most cases, but it is not perfect.  For example,
+it is theoretically possible to choose a seed such that the \textit{first} value
+produced by a generator discovers the bug, which would elide any differential
+speedup (and likewise if both generators fail to find the bug).  To account for
+these cases, we repeat the above process over 10 seeds.\footnote{Although the
+variance \textit{between} seeds in Etna can be vast, timing results are
+replicable for a \textit{given} seed. Across 1,000 trials using the same seed, 
 the observed variance in time-to-failure was less than a nanosecond.}

### 2025-03-20 paper/paper.tex
interleave

@@
 \label{subsection:benchmarking}
-To answer \textbf{RQ1}, we implement benchmarks in both \name and \bq. We vary
-the choice of \rand in our \name generators, using both \bq's default
+To answer \textbf{RQ1}, we microbenchmark generators, comparing generators in \camlname to identical ones in \bq,
+and generators in \scalaname to those in ScalaCheck.
+ We vary the choice of \rand in our \camlname generators, using both \bq's default
 \texttt{splittable\_random} and our improved SplitMix (\csplitmix), as discussed
@@
 
-Our benchmarks consist of boolean lists, binary search trees (BST), and simply
+For \camlname, our test cases consist of generators for boolean lists, binary search trees (BST), and simply
 typed lambda calculus (STLC) terms. We implement generators for these benchmarks
@@
 STLC terms, whose behavior is the most complex. 
-
-The runtime of these generators over a range of sizes is shown in
+For evaluating \scalaname, we implement a subset of the afformentioned strategies  ---
+only the boolean list generator, single-pass BST generator, and well-typed STLC term generator.
+
+The runtime of the \camlname generators over a range of sizes is shown in
 \figref{fig:time}. Staging provides only a negligible speedup for type-derived
@@
 
+As shown in \figref{fig:scala}, \scalaname achieves an even greater performance gain
+--- purely from staging --- over ScalaCheck than \name does over \bq. This discrepancy arises from
+ScalaCheck's representation of generators as functions of type \texttt{size ->
+seed -> Option[A]}. Each generator combinator must construct and then
+pattern-match on these \texttt{Option} values, introducing significant boxing
+and unboxing overhead at each step.  Staging allows us to eliminate this
+overhead, resulting in substantial speedups in aggregate.
+
 These results show that staging and \rand choice serve distinct but
@@
 
-\subsubsection{Scala} To show that the benefit of staging extends to
-other strict functional languages, we apply it to ScalaCheck, a widely-used PBT
-library written in Scala. We implement a subset of our strategies from
-\secref{subsection:benchmarking} in both standard and staged ScalaCheck
-(``\scalaname'').
-
-As shown in \figref{fig:scala}, \scalaname achieves a greater performance gain
-over ScalaCheck than \name does over \bq. This discrepancy arises from
-ScalaCheck's representation of generators as functions of type \texttt{size ->
-seed -> Option[A]}. Each generator combinator must construct and then
-pattern-match on these \texttt{Option} values, introducing significant boxing
-and unboxing overhead at each step.  Staging allows us to eliminate this
-overhead, resulting in substantial speedups in aggregate.
-
-These results illustrate that staging is a powerful \textit{and} general optimization technique, capable of delivering performance gains in strict functional languages beyond OCaml.
+% \subsubsection{Scala} To show that the benefit of staging extends to
+% other strict functional languages, we apply it to ScalaCheck, a widely-used PBT
+% library written in Scala.
+
 \begin{figure}[h]

### 2025-03-20 paper/paper.tex
dashes

@@
 This has a significant impact on the performance of generators. A single call to
-one of the \bq library functions --- like generating an integer uniformly in a range --- may call the \texttt{splittable\_random}
+one of the \bq library functions---like generating an integer uniformly in a range---may call the \texttt{splittable\_random}
 algorithm multiple times. Each sample from \texttt{splittable\_random} allocates 9
@@
 STLC terms, whose behavior is the most complex. 
-For evaluating \scalaname, we implement a subset of the afformentioned strategies  ---
-only the boolean list generator, single-pass BST generator, and well-typed STLC term generator.
+For evaluating \scalaname, we implement a subset of the afformentioned 
+strategies---only the boolean list generator, single-pass BST generator, and well-typed STLC term generator.
 
@@
 
-As shown in \figref{fig:scala}, \scalaname achieves an even greater performance gain
---- purely from staging --- over ScalaCheck than \name does over \bq. This discrepancy arises from
+As shown in \figref{fig:scala}, \scalaname achieves an even greater performance 
+gain---purely from staging---over ScalaCheck than \name does over \bq. This discrepancy arises from
 ScalaCheck's representation of generators as functions of type \texttt{size ->

### 2025-03-20 paper/paper.tex
polish through MSP section, some wibbles in other-langs

--- 2025-03-20 paper/paper.tex
@@
 
-\jwc{Ensure we have a signposting table-of-contents-ish few sentences here, if it's not covered by the contribs.}\todo{Todo Thia}
 In Section~\ref{section:motiv}, we introduce monadic generator DSLs, discuss how they work, and take an example-guided tour through some of the ways that
@@
 Then, in Section~\ref{section:faster-generators}, we explain multi-stage programming,
-and develop \name{}, illustrated by \camlname{}.
+and develop both facets of \name{}, using \camlname{} to illustrate.
 Section~\ref{section:eval} evaluates all of the previous work, demonstrating the generator performance improvements
@@
 % \bcp{That's actually not a bad section title!}
-\todo{Intro of this section}
-
-\subsection{Background: Monadic Generator DSLs}
+
+% \subsection{Background: Monadic Generator DSLs}
 
@@
 generators is via an embedded {\em monadic} language. For the purposes of this
-paper, we use the syntax from OCaml's \bq library, which is presented in
+paper, we use the syntax and types from OCaml's \bq library, which is presented in
 Figure~\ref{fig:bq-api}, but these kinds of DSLs can also be found in languages
@@
 \texttt{size} and \texttt{with\_size} that are used to control the sizes of
-generated values and \texttt{fixed\_point} that is used tto define recursive
+generated values and \texttt{fixed\_point} that is used to define recursive
 generators.
@@
 The monad functions \texttt{return} and \texttt{bind} are defined in the usual way
-for an instance of the reader monad \cn{}. \texttt{Return x} ignores the size and seed and just returns
-\texttt{x}, while \texttt{bind g k} first runs \texttt{g} and then passes the generated value to the continuation \texttt{k}.
-\bq{} uses a mutable seed, so the \texttt{random} passed to \texttt{(k a)} will be different from the one passed to \texttt{g}.
+for an instance of the reader monad \cn{}. 
+% \bq{} uses a mutable seed, so the \texttt{random} passed to \texttt{(k a)} will be different from the one passed to \texttt{g}.
 The \texttt{gen\_int} combinator simply calls out to the randomness library \texttt{Splittable\_random}, aliased as \texttt{SR} here.
-\jwc{Should we do this?}
-
-Different PBT libraries use variations on this basic design. Haskell's QuickCheck uses a reader monad with an immutable state
+
+Different PBT libraries use variations on this basic design. The largest source of variation is how the random seed is threaded through the generator.
+\bq{} uses a reader monad with a mutable seed, while Haskell's QuickCheck uses a reader monad with an immutable state
 that is ``split'' at \texttt{bind}s \cn. Meanwhile, ScalaCheck (a) uses an immutable state type and state monad \cn{} to thread
 the state through, and (b) has an \texttt{Option} as its return type, to allow generation to fail.
-\jwc{However, the principles are the same, and we will use \bq{} to illustrate.}
 
@@
 The first versions (\texttt{int\_pair} in \bq{} and \texttt{intPair} in ScalaCheck) are written with the monadic generator combinators from their respective libraries.
-\jwc{Maybe we can avoid rewriting these if we already showed them earlier? Also maybe we duplicate the tutorial discussion of monadic generator libraries to include Scala?}
 The second versions (\texttt{int\_pair\_inlined} and \texttt{intPairInlined})
@@
 even worse for recursive generators, as the heuristics are necessarily even more
-conservative for optimizing recursive functions. \jwc{unclear if we need this sentence, it's sorta obvious}
+conservative for optimizing recursive functions.
 
@@
 generators suffer further performance penalties due to \emph{closure
-allocation}. In cases where the compiler cannot statically eliminate it \jwc{``it'' = ``the allocation''},
+allocation}. In cases where the compiler cannot statically eliminate it,
 running a monadic bind allocates a short-lived closure: we allocate a closure
@@
 a compiler could unroll the loops in \texttt{weighted\_union} to depth $n$ and specialize the function to avoid allocating the list.
-Unfortunately, almost~\jwc{Once again, GHC is a notable exception here \cn{}} no compilers perform this kind of optimization.
+Unfortunately, almost~\footnote{Once again, GHC is a notable exception here, performing sophisticated list fusion optimizations \cn{}} no compilers perform this kind of optimization.
 This allocation (or rather, its tendency to cause GC pauses) --- as well as the cost of running the code 
@@
 \subsection{Inefficient Randomness Libraries}
-\jwc{We really need to have explained the effect ordering equivalence aspect of the evaluation before here.}
-
 The core of any PBT generator library is a source of randomness: to generate
@@
 
-\hg{This paragraph feels a bit clumsy right now. What if we started with a table
+\iflater\hg{This paragraph feels a bit clumsy right now. What if we started with a table
 of the RNGs used by a bunch of popular PBT frameworks and then argued that
-things like Lehmer are faster?}
+things like Lehmer are faster?}\fi
 
@@
 pointer and reading from memory (which can be pipelined trivially) will always
-be faster on modern CPUs than any random sampling algorithm with data-dependent arithmetic
-instructions.\hg{Should we just always be doing this? I think we want to argue
+be faster on modern CPUs.
+\hg{Should we just always be doing this? I think we want to argue
 that this is a bridge too far because you still need to generate the buffer
@@
 \jwc{HG: talk about the tradeoffs here.}
-
 \jwc{The point here is that ``you should be thinking about using a more performant RNG, we'll show you that it does have real benefits (and quantify them). But there are tradeoffs here.'}
@@
 \jwc{This section needs a much better title...}
-\jwc{Emphasize that this is an \emph{COMPLETELY EQUIVALENT drop in replacement!} We din't just build a different library with different distributions.}
-
-\jwc{Usual introduction to this section, corresponding to how we talked about it in the intro. ``We present a library that XYZ''.}
-
-\hg{TODO: Signposting --- I got a few paragraphs in and realized I wasn't sure what I was reading}
+% \jwc{Usual introduction to this section, corresponding to how we talked about it in the intro. ``We present a library that XYZ''.}
+
+\jwc{This section needs an intro, and some signposting to say where we're going.}
 
@@
 back to quasiquotation in LISP \cn{}.  For the purposes of this paper, we will
-primarily consider two stages: compile time and run time. Staged programs thus
+primarily consider two stages, namely compile time and run time. Staged programs thus
 execute twice: once at compile time, which produces more code, which is then
@@
 \jwc{Cite racket macros for DSLs here:}
-While embedded DSLs are great \jwc{words}, they have well-known drawback: 
+While embedded DSLs are powerful tools, they have well-known drawback: 
 the functional abstractions used to build eDSLs prevent compilers from generating efficient machine code \cn.
-\jwc{Do racket people also use macros to eliminate run time overhead? Or is it just for extensibility?}
-As alluded to previously \jwc{This sentence here because of the section-ordering transpose...
-may need to mention this earlier.}, this effect is known as ``abstraction overhead'' \cite{carrette05, moller20}: the layers of abstraction that
+This effect is known as ``abstraction overhead'' \cite{carrette05, moller20}: the layers of abstraction that 
 make eDSLs so usable are precisely what prevents them from being fast.
-\jwc{Something to the effect of ``the issues we noticed in BQ and SC exist across many eDSLs''}
+Indeed, the root causes of the abstraction overheads we discovered in Section~\ref{section:motiv} are
+not unique to generator libraries.
 In light of this issue, staging is often used as a lightweight compiler for DSLs. The compile-time evaluation stage
 transforms the DSL code, eliminating abstractions to produce code that the host language compiler can generate fast machine code for.
-\jwc{This previous paragraph is peak ``shitty first draft''.}
 This recipe has been used to great effect across domains to stage eDSLs for stream processing \cite{moller20,strymonas}, parser combinators \cite{sspc, staged-parsers, krishnaswami19, flap},
 and query processing \cite{rhyme}.
-\jwc{Cite racket things here}
 
 Many languages have some degree of staging functionality, though support varies
-from full native support (Scala 3 \cite{scalamacros}, Haskell \cite{templatehaskell}, Racket \cn{}) to compiler extensions (OCaml \cite{metaocaml,macocaml}, Java \cite{mint}).
+from full native support (Scala 3 \cite{scalamacros}, Haskell \cite{templatehaskell}, \jwc{Racket??}) to compiler extensions (OCaml \cite{metaocaml,macocaml}, Java \cite{mint}).
 Moreover, many more languages have lighter-weight staging-like metaprogramming support in the form of macros.
-See Section~\ref{subsection:other-langs} for a full discussion of just how portable these ideas are. \jwc{this sentence is a placeholder lol}
-
 For this paper, we have implemented staged PBT libraries both in OCaml --- using the MetaOCaml fork of the OCaml compiler \cn --- and Scala 3 --- using
@@
 \scalaname{} differs almost entirely in syntax alone.
+For a further discussion of the potential of building staged generator DSLs in other languages, see Section~\ref{section:other-langs}.
 
@@
 
-\section{Other Languages and Libraries}
-\tr{Subtitle: ``Beyond strict functional languages?''}
-\label{subsection:other-langs}
+\section{Beyond Strict Functional Languages}
+\label{section:other-langs}
 
@@
 might be imported to their favorite language.
-Below we discuss how \name{} could be applied in other languages.
-
-% While we have done our best to keep the proceeding discussion as
-% language-agnostic as possible, our claims about the \name{} technique have only
-% been thoroughly tested in OCaml and Scala 3, and we are only claiming (in the
-% scientific sense) that we have contributed a technique applicable to PBT
-% libraries in strict functional programming languages.
-
+
+We begin by noting that the \name{} principle of choosing fast random number generators
+is completely language-agnostic. In languages like OCaml, where runtime value representations
+make natively-implemented \rand inefficient, calling out to a C implementation or using standard libraries written in C is a surefire win.
+In other languages, serious thought should be put into using as fast of a \rand{} as possible,
+as opposed to picking up any suitable option off the shelf.
+
+Next, we discuss language-by-language the degree to which the \name{} insights about staging could be used.
 \begin{itemize}
@@
   there are at least two \cn{} PBT libraries in Racket.
-  \jwc{The Racket macros literature, while extensive, is not about staging!}
-  We have not yet tested the technique in Racket for the simple reason that the authorship team of this paper does not posess
-  much Racket expertise \jwc{can we say this without unblinding?}.
+  However, the racket macros literature, while extensive, does not concern itself with staging for performance purposes.
+  For this reason, we --- the authorship team of this paper, devoid of much Racket expertise --- decided to not use Racket as a test case in this paper.
 

### 2025-03-20 paper/paper.tex
more wibbling in beyond strict

@@
 % \jwc{Initial section here about eval}\tr{todo}
-All experiments were run on a 64-bit Linux machine with 264 GB RAM and an Intel Xeon Platinum 8375C CPU, running Ubuntu 24.04.1 LTS. The OCaml version of \name and all OCaml benchmarks
+All experiments were run on a 64-bit Linux machine with 264 GB RAM and an Intel Xeon Platinum 8375C CPU, running Ubuntu 24.04.1 LTS. \camlname and all OCaml benchmarks
 were compiled with 4.14.1+BER MetaOCaml \texttt{ocamlopt}, the native code compiler for OCaml, using compiler flag \texttt{-O3}.
@@
 As shown in \figref{fig:scala}, \scalaname achieves an even greater performance 
-gain---purely from staging---over ScalaCheck than \name does over \bq. This discrepancy arises from
+gain---purely from staging---over ScalaCheck than \camlname does over \bq. This discrepancy arises from
 ScalaCheck's representation of generators as functions of type \texttt{size ->
@@
 Performance benefits scale with generation size; lower is better.
-BQ is \bq; Staged SR is \name using \bq's \texttt{splittable\_random} RNG; Staged CSR is \name using \secref{subsection:faster-rng}'s \csplitmix.}
+BQ is \bq; Staged SR is \camlname using \bq's \texttt{splittable\_random} RNG; Staged CSR is \camlname using \secref{subsection:faster-rng}'s \csplitmix.}
   \label{fig:time}
@@
 \begin{itemize}
-  \item \textbf{Racket} is the next target that we intend to test the \name{} methodology on.
+  \item \textbf{Racket} is the next target that we intend to test on.
   The entire Racket philosophy is intertwined with
@@
   \item \textbf{Haskell} is home to the original PBT implementation \cn{}, and
-  so it is natural to ask why we have not yet applied \name{} to it. The basic
+  so it is natural to ask why we have not built a \name{} in it yet. The basic
   answer is that the Haskell compiler (GHC) is designed specially to eliminate
@@
   sometimes to build eDSLs in contexts where one does not want to rely on the
-  optimizer's heuristics.  In short: it is possible to apply \name{} to Haskell
+  optimizer's heuristics.  In short: it is possible to build a \name{} version of Haskell
   QuickCheck, but the benefits would be much less pronounced.
@@
   PBT libraries in Rust ask programmers to define generators directly as \texttt{seed -> value} functions.
-  For this reason, \name{} does not seem directly applicable to any of the PBT libraries in Rust.
+  For this reason, staging does not seem directly applicable to any of the PBT libraries in Rust.
 \end{itemize}

### 2025-03-20 paper/paper.tex
wibbles

@@
 The performance impact of this inling is large (Figure~\ref{fig:overhead-explanation-perf}). In both languages, the
-inlined version takes (on average) \textbf{half} as much time to generate a single pair of
+inlined version takes (on average) half as much time to generate a single pair of
 \texttt{int}s. 
-Microbenchmarks of more realistic generators (see Section~\ref{section:eval}
-\jwc{more precise ref}) show an even more dramatic performance boost for
-inlining in this manner.
+Microbenchmarks of more realistic generators (see Section~\ref{section:eval})
+show an even more dramatic performance boost for inlining in this manner.
 
@@
 use heuristics to determine if and when to perform this particular kind of
-simplification~\footnote{As we discuss in Section~\ref{asdf} \jwc{fixme}, purity
+simplification~\footnote{As we discuss in Section~\ref{section:other-langs}, purity
 means that Haskell is a slightly different story. GHC can and often does
@@
 generator is very costly because each allocation brings us closer to the next costly GC pause \cn{}.
-Similarly to \jwc{the previous one} this effect is magnified in recursive generators: each iteration
+This effect is also magnified in recursive generators: each iteration
 through the recursive loop re-allocates closures for \texttt{bind}s, so the amount of allocation
@@
 \subsubsection{Overhead of Choice Combinators}
-After the \texttt{return}, \texttt{bind}, and base-type generation combinators
-(such as \texttt{gen\_int}), the most frequently used combinators are ``choice''
-combinators like \texttt{weighted\_union} which randomly select between
-sub-generators.
-Like \texttt{return} and \texttt{bind}, these combinators incur a performance penalty
+Like \texttt{return} and \texttt{bind}, combinators like \texttt{weighted\_union} incur a performance penalty
 at run time. Aside from the previously-discussed issues that compilers
 cannot see through the abstraction boundary to optimize these programs,
-choice combinators come with their own particular abstraction overhead.
+choice combinators like \texttt{weighted\_union} come with their own particular abstraction overhead.
 
@@
 the options for which are always known.
-This list is allocated at the call site, and then \emph{immediately}
-destructed by the function. In principle, since the elements of the list and its length $n$ are known statically,
-a compiler could unroll the loops in \texttt{weighted\_union} to depth $n$ and specialize the function to avoid allocating the list.
+This list is allocated at the call site and then never needed after the call to \texttt{weighted\_union} returns\footnote{
+This overhead is fixable in languages with stack alloctation, which OCaml will soon be \cn{}.
+}.
+Since the elements of the list and its length $n$ are known,
+a compiler could in principle unroll the loops in \texttt{weighted\_union} to depth $n$ and specialize the function at each call site to avoid allocating the list.
 Unfortunately, almost~\footnote{Once again, GHC is a notable exception here, performing sophisticated list fusion optimizations \cn{}} no compilers perform this kind of optimization.

### 2025-03-20 paper/paper.tex
non-sequitur removal

@@
 generation; for that, we need an experiment. The most obvious experiment is to
-simply swap out the randomness library for \bq{} with a totally different, faster one. But, as
+simply swap out the randomness library of either \bq{} or ScalaCheck with a totally different, faster one. But, as
 discussed previously, generators with different generation orders can be
 difficult to compare. To get around this, we exploit a ``natural experiment:
-OCaml's \texttt{Splittable\_random} library is slow in a way that
-can be improved \emph{without} changing its extensional behavior.
+the \rand{} that \bq{} uses --- OCaml's \texttt{Splittable\_random} library --- is slow in a way that
+can be improved \emph{without} changing its behavior.
 In particular,
@@
 
-\section{Speeding up Generator DSLs by Staging and Faster Randomness Libraries}
+\section{\name{}}
 \label{section:faster-generators}
@@
 over the course of multiple stages, with each stage producing the code to be run
-in the next.  This is an old idea in programming languages, with roots going
-back to quasiquotation in LISP \cn{}.  For the purposes of this paper, we will
+in the next. For the purposes of this paper, we will
 primarily consider two stages, namely compile time and run time. Staged programs thus
@@
 use the natural experiment provided by OCaml's inefficient implementation of
-SplitMix. By replacing this slow randomness library with a faster but extensionally equivalent
+SplitMix. By replacing this slow randomness library with a faster but semantically equivalent
 implemetnation, we are able to precisely quantify the bugfinding speedup that using a randomness library gives across a range of PBT scenarios.

### 2025-03-20 paper/paper.tex
hows this

@@
 make eDSLs so usable are precisely what prevents them from being fast.
-Indeed, the root causes of the abstraction overheads we discovered in Section~\ref{section:motiv} are
-not unique to generator libraries.
+Indeed, the root causes of many of the issues we discovered in Section~\ref{section:motiv} are
+not unique to generator DSLs.
 In light of this issue, staging is often used as a lightweight compiler for DSLs. The compile-time evaluation stage
@@
 from full native support (Scala 3 \cite{scalamacros}, Haskell \cite{templatehaskell}, \jwc{Racket??}) to compiler extensions (OCaml \cite{metaocaml,macocaml}, Java \cite{mint}).
-Moreover, many more languages have lighter-weight staging-like metaprogramming support in the form of macros.
 For this paper, we have implemented staged PBT libraries both in OCaml --- using the MetaOCaml fork of the OCaml compiler \cn --- and Scala 3 --- using

### 2025-03-20 paper/paper.tex
technical section wibbling

@@
 \texttt{'b}. In particular, because \texttt{f} itself runs at compile time, the
-abstraction of using \texttt{f} is necessarily completely eliminated by
-run time.
+fact that the programmer called \texttt{f} does not matter at run time --- the abstraction
+that \texttt{f} defines has been eliminated.
 A code-transforming function \texttt{'a code -> 'b code} can also be converted \emph{code for a function} --- a value of type
@@
 the quoted \texttt{x}.
-\jwc{Does this make any sense?}
 
@@
 By composing the two code-transforming functions together at compile time, and only then turning them into a run-time function,
-the functions are fused together... \jwc{words}.
+the functions are fused together.
 This is the basis of how staging is used to eliminate the abstraction of DSLs
@@
 and only calling \texttt{eta} at the end on the completed DSL program ---  we
-can ensure that any overhead of using them is eliminated by run time.
+can ensure that any overhead of using the combinators is eliminated by run time.
 
@@
 
-Recall \jwc{did we talk about this} that staging a DSL involves changing the
-combinators to run at compile time by carefully annotating their types with \texttt{code}s.
-Deciding which types \texttt{t} can
-instead be \texttt{t code} --- in other words,
-figuring out which parts of the DSL can be determined statically (and can be
-part of the compile-time stage), and which parts are only known dynamically (and
-hence must be \texttt{code}) --- is an art known as ``binding-time analysis'' \cn.
-
-The crux of our binding time analysis is that the particular random seed and
-size parameter \jwc{if explained} are only known at run time (the later stage),
-but the code of the generator itself is known at compile time.\hg{Are we going
-to talk more about how we did this analysis?} \jwc{Well it's sort of right here: you just think about it, it's the same thing as staging a parser too.}
-Generators ---
-values of type \texttt{'a Bq.t} --- are always constructed statically in practice,
-so all of the combinators we use to build them can run at compile time.
-
+To build a staged version of a generator DSL, we must rewrite the the generator
+combinators so that they run during the compile time. Of course,
+the compile time stage does not do all the work: the DSL must run at compile time,
+producing inlined code that uses no combinators, which can then be compiled and run at run time
+with different sizes and seeds.
+
+Our job is thus to carefully bisect the DSL, determining which inputs to generator
+combinators are known statically (and can be part of the compile time stage), and
+which parts are only known at run time (and must be \texttt{code}).
+In the staging literature, this task is known as ``binding-time analysis'', and it is more of an art than a science.
+
+% carefully break up so that you push what you can into the compile time stage.
+% Figuring out which parts of the generator are known statically ()
+% Recall \jwc{did we talk about this} that staging a DSL involves changing the
+% combinators to run at compile time by carefully annotating their types with \texttt{code}s.
+% Deciding which types \texttt{t} can
+% instead be \texttt{t code} --- in other words,
+% figuring out which parts of the DSL can be determined statically (and can be
+% part of the compile-time stage), and which parts are only known dynamically (and
+% hence must be \texttt{code}) --- is an art known as ``binding-time analysis'' \cn.
+
+The crux of our binding time analysis is that the \emph{only} parts of a generator that are not
+known at compile time are the particular random seed and
+size parameters that will be passed in at run time. Generators themselves are always known entirely at compile time.
 This means our library's generator type \texttt{'a
@@
 This type, along with basic staged monadic generator DSL functionality can be found in Figure~\ref{fig:gen-staged-basic}.
-\texttt{Return} is the constant generator, but this time it 
+The constant generator \texttt{return} has the same code, but this time it 
 runs at compile time. Given \texttt{cx : 'a code}, the code for a \texttt{'a}, it
-returns the generator which always generates that value.
-\texttt{bind g k} sequences generators by passing the result of running the generator \texttt{g} to a continuation \texttt{k}. However, instead of getting access
+returns the generator which always generates that value. Similarly,
+\texttt{bind g k} sequences generators as usual by passing the result of running the generator \texttt{g} to a continuation \texttt{k}. However, instead of getting access
 to the particular value generated by \texttt{g}, the continuation \texttt{k} gets access to \texttt{code} for the value sampled from \texttt{g}:
@@
 from Section~\ref{subsection:msp}).
+\iflater
 \hg{TODO: Be really careful with formatting of the above paragraph. It's very
@@
 content is already pretty low-level}
+\fi
 

### 2025-03-20 paper/paper.tex
done through 3.4

@@
 Returning to our running example, Figure~\ref{fig:running-staged} shows the
-int-pair example written with the staged \texttt{Gen.t} monad, as well as the
+int-pair generator written with the staged \texttt{Gen.t} monad, as well as the
 inlined code that results from calling \texttt{Gen.to\_bq} (changing some identifier names for clarity).
@@
 
-\jwc{Is there anything more we need to say here?}
-
 \subsection{Staging Combinators}
@@
 $\dots$
-  let pick (acc : int code) (weighted_gens : (int code * 'a t) list) (size : int code) (random : Sr.t code) : 'a code =
+  let pick (acc : int code) (weighted_gens : (int code * 'a t) list) size random : 'a code =
     match weighted_gens with
@@
 so they are \texttt{code}s.
-\texttt{Gen.weighted\_union} begins by computing \texttt{sum\_code}, an \texttt{int code}
-that is the sum of the weights. Note that this happens at compile time: we fold over a list
-known at compile time to produce another code value. 
-We then call \texttt{SR.int} to sample a random number \texttt{r} between \texttt{0} and the sum.
-Finally, we splice in the result of calling the helper function \texttt{pick}. \texttt{pick}
-produces a tree of \texttt{if}s by again traversing the list of generators at compile time.
-This tree of \texttt{ifs} ``searches'' for the generator corresponding to the
-sampled value \texttt{r}, and then runs it.
-\hg{I'm not sure this helps me. I think I'd prefer a less detailed explanation
-of the code that conveys the intuition and let the reader actually read the code
-if they want to know specifics. As it stands the explanation is just too dense
-for me to process}
+Instead of building a histogram of the distribution described by the weights at run time and then traversing it,
+the compile time \texttt{weighted\_union} combinator generates a
+tree of \texttt{if}s --- specialized to the list of weights known at compile time --- that searches for the selected generator.
+
+% \texttt{Gen.weighted\_union} begins by computing \texttt{sum\_code}, an \texttt{int code}
+% that is the sum of the weights. Note that this happens at compile time: we fold over a list
+% known at compile time to produce another code value. 
+% We then call \texttt{SR.int} to sample a random number \texttt{r} between \texttt{0} and the sum.
+% Finally, we splice in the result of calling the helper function \texttt{pick}. \texttt{pick}
+% produces a tree of \texttt{if}s by again traversing the list of generators at compile time.
+% This tree of \texttt{ifs} ``searches'' for the generator corresponding to the
+% sampled value \texttt{r}, and then runs it.
+% \hg{I'm not sure this helps me. I think I'd prefer a less detailed explanation
+% of the code that conveys the intuition and let the reader actually read the code
+% if they want to know specifics. As it stands the explanation is just too dense
+% for me to process}
 
@@
 the generated code first computes the sum of these numbers, samples between \texttt{0} and the sum, and then
-traverses a tree of three \texttt{if}s to find the correct value to return.
+traverses the tree of three \texttt{if}s to find the correct value to return.
 
@@
     let r = SR.int random 0 sum in
-    if r <= size then 'a'
-    else
+    if r <= size then 'a' else
       let r' = r - size in
-      if r' <= 2 then 'b'
-      else
+      if r' <= 2 then 'b' else
         let r'' = r' - 2 in
-        if r'' <= 1 then 'c'
-        else
-          failwith "Error"
+        if r'' <= 1 then 'c' else failwith "Error"
 >.

### 2025-03-20 paper/paper.tex
blah

@@
 \section{Introduction}
+\label{section:intro}
 
@@
   let bind (g : 'a t) (k : 'a code -> 'b t) : 'b t =
-    fun size_c random_c ->
+    fun size random ->
       .<
-        let a = .~(g size_c random_c) in
-        .~(k .<a>. size_c random_c)
+        let a = .~(g size random) in
+        .~(k .<a>. size random)
       >.
@@
   let int (lo : int code) (hi : int code) : int t =
-    fun size_c random_c ->
-      .< SR.int .~random_c .~lo .~hi >.
+    fun size random ->
+      .< SR.int .~random .~lo .~hi >.
 
@@
 \subsection{Let-Insertion and Effect Ordering}
-Careful readers might note that the definition of \texttt{bind} in Section~\ref{subsection:basic-design}
-was more complicated than one might expect. In particular, why not define bind in a more standard way: as
-\texttt{let bind' g k = fun size random -> k (g size random) size random}, without the code block
-that let-binds the spliced code \texttt{.\~{}(g size random)}?\hg{Need to spell
-this out --- the vast majority of readers won't have that precise question in
-their heads. Might even be worth comparing the two implementations here side by side}
-Unfortunately, using \texttt{Gen.bind'} leads to incorrect code being generated.
+Careful readers might note that the definition of \texttt{bind} (shown in Figure~\ref{fig:bind-with-let-binding})
+was more complicated than one might expect. In particular, why not define bind in the standard way
+for a reader monad (Figure~\ref{fig:bind-without-let-binding})?
+Unfortunately, the \texttt{bind'} definition is wrong in our context as it leads to incorrect code being generated.
 For example, consider \texttt{Gen.bind' (Gen.int .<0>. .<1>.) (fun x -> Gen.return .<(.~x,.~x)>.)}.
 This generates the run time code \texttt{fun size random -> (SR.int random 0 1, int SR.random 0 1)},
-which is incorrect. 
-This is not equivalent to the behavior of writing the same code with base quickcheck \jwc{Which is a property we want to hold for our eval.}.
-Instead of generating a single integer and returning it twice, it samples two different integers.
+which is incorrect.  Instead of generating a single integer and returning it
+twice, it samples two different integers.
+This matters because, as described in Section~\ref{section:intro}, \camlname and
+\scalaname should be equivalent to their unstaged counterparts.
+
+\begin{figure}[h]
+  \begin{subfigure}{.45\textwidth}
+\begin{lstlisting}
+let bind (g : 'a t) (k : 'a code -> 'b t) : 'b t =
+    fun size random ->
+      .<
+        let a = .~(g size random) in
+        .~(k .<a>. size random)
+      >.
+\end{lstlisting}
+\caption{\texttt{bind}, with a let-binding}\label{fig:bind-with-let-binding}
+  \end{subfigure}
+  \begin{subfigure}{.45\textwidth}
+\begin{lstlisting}
+let bind' (g : 'a t) (k : 'a code -> 'b t) : 'b t =
+    fun size random -> k (g size random) size random
+\end{lstlisting}
+\caption{\texttt{bind'}, the ``standard'' bind for the reader monad}\label{fig:bind-without-let-binding}
+  \end{subfigure}
+  \caption{\texttt{bind}, two ways}
+  \label{fig:bind-two-ways}
+\end{figure}
 
@@
 but that the value can be used in the continuation multiple times, \texttt{bind} let-binds the result of generation to a variable,
-and then passes that to the continuation.
-
-The library \jwc{ensure this is consistent with the way we talk about the project, cf cross-language} is careully designed to 
-preserve exactly the effect order of base quickcheck \jwc{and the scala version to preserve the effect ordering of ScalaCheck}.
-\jwc{This fact should go earlier. Not really sure about where this subsection should actually land, but it needs to be said somehwere.}
-\hg{+1, I think this feels sort of out of place here, but I also think moving it
-up would make that part harder to understand too. Is there a way to tie this in
-to our discussion about maintaining the precise set of choices? The two issues
-are different, but they're related}
+and then passes it to the continuation.
+
+% The library \jwc{ensure this is consistent with the way we talk about the project, cf cross-language} is careully designed to 
+% preserve exactly the effect order of base quickcheck \jwc{and the scala version to preserve the effect ordering of ScalaCheck}.
+% \jwc{This fact should go earlier. Not really sure about where this subsection should actually land, but it needs to be said somehwere.}
+% \hg{+1, I think this feels sort of out of place here, but I also think moving it
+% up would make that part harder to understand too. Is there a way to tie this in
+% to our discussion about maintaining the precise set of choices? The two issues
+% are different, but they're related}
 
@@
 \label{subsection:codecps}
-This is all great so far,\hg{too informal} but there's a subtle issue that prevents the version of the library design discussed so far
-from being used as a proper drop-in replacement for an existing generator DSL: the types of \texttt{return : 'a code -> 'a Gen.t}
+Another subtle issue prevents the version of the library design discussed so far
+from being used as a proper drop-in replacement for an existing generator DSL. In particular, the types of \texttt{return : 'a code -> 'a Gen.t}
 and \texttt{bind : 'a Gen.t -> ('a code -> 'b Gen.t) -> 'b Gen.t} aren't quite right.
@@
 actually the proper monad function types.
+\iflater
 \hg{A pedantic reader may ask: Are we concerned with the monad {\em laws} as
 well, or just the type signature?}
+\fi
 
@@
 the functions \texttt{return (x : 'a) = fun \_ \_ -> x} and \texttt{bind (g : 'a t) (k : 'a -> 'b t) = fun size random -> k (g size random) size random}
-have the proper types for a monad instance.
-Then, any combinators of type \texttt{'a Gen.t} before simply become \texttt{'a code Gen.t} with this new version.
-\jwc{do we want a different name for the first cut?}
+have the proper types for a monad instance. Then, any combinators of type \texttt{'a Gen.t} before simply become \texttt{'a code Gen.t} with this new version.
 
@@
 wiht an example}
-\jwc{I could, but i'm just not sure this is critical.}
+\jwc{I could, but i'm just not sure this is critical.
+In particular, every time I've ever seen this presented in a paper they just drop it and move on.
+It's too deep in the weeds for it to be worth explaining, too...
+nobody's going to read this section except staging fans}
 
@@
 that a program like \texttt{Gen.bind Gen.int (fun x -> ...)} generates a let-binding for the result of sampling the randomness library.
-For example, \texttt{bind (int .<0>. .<1.>) (fun cx -> return .<(.~cx,.~cx)>.)} now correctly generates
-\texttt{.< fun size random -> let x = SR.int random 0 1 in (x,x) >.}. \jwc{Should we do out the whole reduction sequence here? It might be explanatory, but it also might be boring.}
+For example, \lstinline{bind (int .<0>. .<1.>) (fun cx -> return .<(.~cx,.~cx)>.)} now correctly generates
+\lstinline{.< fun size random -> let x = SR.int random 0 1 in (x,x) >.}.
 
@@
 If no discrepancy arises, we conclude that the generators exhibit equivalent behavior.}
-
-\jwc{... and all of this works identically in Scala, too.}
 

### 2025-03-20 paper/paper.tex
pass done

@@
 An initial attempt is to try \texttt{type 'a t = int code -> SR.t code -> 'a}. If we strip the \texttt{code} off the result type,
-the functions \texttt{return (x : 'a) = fun \_ \_ -> x} and \texttt{bind (g : 'a t) (k : 'a -> 'b t) = fun size random -> k (g size random) size random}
+the functions \texttt{return (x : 'a) = fun \_ \_ -> x} and \texttt{bind g k = fun size seed -> k (g size seed) size seed}
 have the proper types for a monad instance. Then, any combinators of type \texttt{'a Gen.t} before simply become \texttt{'a code Gen.t} with this new version.
@@
 However, this definition of bind doesn't have call-by value effect semantics, as
-discussed in the previous section!  And because the type of \texttt{g size random} is
-just \texttt{'a} (not necessarily \texttt{'a code}), we cannot perform the
+discussed in the previous section!  And because the type of \texttt{g size seed} is
+now just \texttt{'a} (not necessarily \texttt{'a code}), we cannot perform the
 let-insertion needed to preserve the CBV effects. To solve this problem, we
@@
 wiht an example}
-\jwc{I could, but i'm just not sure this is critical.
-In particular, every time I've ever seen this presented in a paper they just drop it and move on.
-It's too deep in the weeds for it to be worth explaining, too...
-nobody's going to read this section except staging fans}
+\jwc{I could, but i'm just not sure this is important. It's too deep in the weeds for it to be worth explaining, too... nobody's going to read this section except staging fans}
 
@@
 \lstinline{.< fun size random -> let x = SR.int random 0 1 in (x,x) >.}.
-
 This design is less obviously correct, and does require some care. Rather than \texttt{bind} ensuring correct evaluation order once and for all,
@@
 Different generator DSLs handle defining recursive generators differently. Some allow recursive generators
-to be defined as recursive functions (or in Haskell's case, recursive values), while others
-(like \bq) expose a fixpoint combinator to construct recursive generators.
-Generator fixpoint combinators are simliar to a usual \texttt{fix : ('a -> 'a) -> 'a} combinator,
-except that they operate at monadic type \texttt{fixed\_point : ('a Bq.t -> 'a Bq.t) -> 'a Bq.t}.
+to be defined as recursive functions, while others
+(like \bq and ScalaCheck) expose a fixpoint combinator to construct recursive generators.
 Given a step function that takes a ``handle'' to sample from a recursive generator call, it
@@
 Programmers can then define recursive generators by \texttt{fixed\_point}, which ties the recursive knot.
-\jwc{don't want to handle GADTs}
 

### 2025-03-20 paper/paper.tex
nits

@@
 STLC terms, whose behavior is the most complex. 
-For evaluating \scalaname, we implement a subset of the afformentioned 
+For evaluating \scalaname, we implement a subset of the aformentioned 
 strategies---only the boolean list generator, single-pass BST generator, and well-typed STLC term generator.
@@
 sampling. In this case, performance improvement is primarily attributable the
-\csplitmix \rand.  In contrast, STLC strategies benefit more from staging, as
+\csplitmix \rand.  In contrast, the STLC strategies benefit more from staging, as
 they involve fewer sampling operations but rely heavily on generator
@@
 designed to test different aspects of the system (e.g., BST includes tasks that
-test insertion, deletion, and union operations on binary search trees). In our
-case, BST has 43 tasks, and STLC has 18.
+test insertion, deletion, and union operations on binary search trees). In particular, BST has 43 tasks, and STLC has 18.
 
@@
 under 5ms, as such negligible runtimes do not yield meaningful insights into
-relative performance. These filters ensure that reported speedups are based on
+relative performance. \tr{We probably want to discuss how many tasks get filtered out...} These filters ensure that reported speedups are based on
 cases where optimization has a measurable impact.
@@
 speedup (and likewise if both generators fail to find the bug).  To account for
-these cases, we repeat the above process over 10 seeds.\footnote{Although the
+these cases, we repeat the above process over 10 \tr{(soon to be 30)} seeds.\footnote{Although the
 variance \textit{between} seeds in Etna can be vast, timing results are
@@
 tend to achieve larger speedups, whereas very fast or slow tasks cluster near the mean.
+\tr{To be honest, we don't know EXACTLY why STLC exhibits this bimodal distribution.
+Maybe it's better to leave this out, and just say ``look, there's high variance, but it's actually a good thing---sometimes you get more than you expect, and you almost never get less than you expect"?}
 Overall, 

### 2025-03-20 paper/paper.tex
minor changes to eval

@@
   \includegraphics[scale=0.45]{time.png}
-  \caption{Time to generate a value using each strategy, averaged over 10,000 trials.
+  \caption{Time to generate a varying-size value using each strategy, averaged over 10,000 trials.
 Performance benefits scale with generation size; lower is better.
@@
   \includegraphics[scale=0.32]{scala.png}
- \caption{Staging in ScalaCheck shows greater speedups than in \bq. SC is ScalaCheck; BQ is \bq.}
+ \caption{\scalaname has better performance relative to ScalaCheck than \camlname relative to \bq. SC is ScalaCheck; BQ is \bq. Comparison is for a fixed size of 100.}
  \label{fig:scala}
@@
   \includegraphics[scale=0.333]{speedup.png}
-  \caption{Geometric average of all speedups for each strategy and benchmark, showing that staging leads to better bug-finding ability across the board.}
+  \caption{Geometric average of all speedups---relative to BQ---for each strategy and benchmark, showing that staging leads to better bug-finding ability across the board.}
   \label{fig:etna}
@@
   \includegraphics[scale=0.41]{swarm.png}
-  \caption{Swarm distribution of individual-task speedups across strategies and benchmarks (each black dot is a task).
+  \caption{Swarm distribution of individual-task speedups across strategies and benchmarks, black dots represent tasks.
 Plots with extremely low variance have no swarm overlay; in plots with high variance, outliers correspond to large speedups.

### 2025-03-20 paper/paper.tex
new swarm

@@
 \subsection{Staged Type-Derived Generators}
+
+Generators are traditionally handwritten, but some libraries allow users to synthesize 
+them automatically from type definitions. Type-derived generators are highly 
+convenient---the derivation process requires no manual effort---but also limitated: 
+they are unable to account for constraints absent from the type. 
+As a result, generating correct values with complex preconditions is difficult.
+This trade-off can be deceptive, especially for inexperienced PBT users, who might derive a generator, run their test harness, find no bugs, and assume the code is correct---when in reality, most values are never tested at all because they do not satisfy the preconditions.
+
+This makes type-derivation a particularly elegant target of \name. Since the process of type derivation is completely automatic,
+the end user does not need to understand the staging syntax---but more importantly, \tr{TODO}
+
 \label{subsection:type-derived}
@@
 subtrees at each step; a generator of boolean lists that operates similarly to
-the single-pass BST strategy; and a custom, correct-by-construction generator of
+the single-pass BST strategy; and a well-typed generator of
 STLC terms, whose behavior is the most complex. 
@@
 \begin{figure}[h]
-  \includegraphics[scale=0.41]{swarm.png}
+  \includegraphics[scale=0.26]{swarm.png}
   \caption{Swarm distribution of individual-task speedups across strategies and benchmarks, black dots represent tasks.

### 2025-03-20 paper/paper.tex
newswarm
fix
eval

@@
 Generators are traditionally handwritten, but some libraries allow users to synthesize 
-them automatically from type definitions. Type-derived generators are highly 
-convenient---the derivation process requires no manual effort---but also limitated: 
-they are unable to account for constraints absent from the type. 
-As a result, generating correct values with complex preconditions is difficult.
-This trade-off can be deceptive, especially for inexperienced PBT users, who might derive a generator, run their test harness, find no bugs, and assume the code is correct---when in reality, most values are never tested at all because they do not satisfy the preconditions.
+them automatically from type definitions. Type-derived generators are 
+convenient---the derivation process requires no manual effort---but also limited: 
+they are unable to account for constraints not encoded in the type. 
+As a result, generating values with complex preconditions is statistically unlikely.
+This state of affairs is not ideal, as inexperienced PBT users, who might derive a generator, run their test harness, find no bugs, and assume the code is correct---when in reality, most values are never tested at all because they do not satisfy the preconditions.
 
@@
 designed to test different aspects of the system (e.g., BST includes tasks that
-test insertion, deletion, and union operations on binary search trees). In particular, BST has 43 tasks, and STLC has 18.
-
-Strategies for BST and STLC are unchanged from \secref{subsection:benchmarking}.
-To ensure a fair comparison, we initialize each strategy with the same random
-seed, ensuring that they produce identical sequences of values. We run each
-strategy on all tasks using a 60-second timeout.  If a bug is not found within
-this limit, the task is considered unsolved.  Tasks where all strategies fail to
-find the bug are excluded from our dataset, as they provide no basis for
-comparison.  Similarly, we exclude tasks where the BQ strategy completes in
-under 5ms, as such negligible runtimes do not yield meaningful insights into
-relative performance. \tr{We probably want to discuss how many tasks get filtered out...} These filters ensure that reported speedups are based on
-cases where optimization has a measurable impact.
+test insertion, deletion, and union operations on binary search trees). 
+In particular, BST has 37 tasks, and STLC has 20. Strategies for BST and STLC 
+are implemented in
+\camlname; they are unchanged from their description in \secref{subsection:benchmarking}.\footnote{Etna does 
+not support Scala, so we were unable to test \scalaname, but we see no reason
+we see no reason our results in this section would not extend to \scalaname.}
 
@@
 speedup (and likewise if both generators fail to find the bug).  To account for
-these cases, we repeat the above process over 10 \tr{(soon to be 30)} seeds.\footnote{Although the
+these cases, we repeat the above process over 30 random seeds---that is, all strategies
+are run on the same seed so that they produce identical sequences of values, and this process
+is repeated 30 times.\footnote{Although the
 variance \textit{between} seeds in Etna can be vast, timing results are
@@
 the observed variance in time-to-failure was less than a nanosecond.}
+
+We run each
+strategy on all tasks using a 60-second timeout.  If a bug is not found within
+this limit, the task is considered unsolved.  Tasks where all strategies fail to
+find the bug are excluded from our dataset, as they provide no basis for
+comparison.  Similarly, we exclude tasks where the BQ strategy completes in
+under 5ms, as such negligible runtimes do not yield meaningful insights into
+relative performance. These filters remove
+28/600 (4.67\%) of type-derived STLC's tasks, 
+168/600 (28\%) of STLC's tasks, 
+149/1110 (14.42\%) of type-derived BST's tasks,
+268/1110 (24.1\%) of single-pass BST's tasks, and
+371/1110 (33.42\%) of insertion-based BST's tasks. More sophisticated 
+strategies tend to find bugs faster, leading to a higher number of filtered 
+tasks. By applying these filters, we ensure that our reported speedups reflect 
+optimizations that meaningfully impact performance.
 
@@
 than the other benchmarks, which we attribute to the heterogenous difficulty of its tasks:  
-some tasks regularly hit the 60-second timeout, while others finish in a fraction of a second. We find that tasks of intermediate difficulty 
-tend to achieve larger speedups, whereas very fast or slow tasks cluster near the mean.
-\tr{To be honest, we don't know EXACTLY why STLC exhibits this bimodal distribution.
-Maybe it's better to leave this out, and just say ``look, there's high variance, but it's actually a good thing---sometimes you get more than you expect, and you almost never get less than you expect"?}
+some tasks regularly hit the 60-second timeout, while others finish in a fraction of a second. We 
+find that ``easier'' tasks---those that run on the order of milliseconds---regularly achieve speedups in the 
+range of 300 to 800\%, whereas tasks that run on the order of seconds achieve more moderate speedups of up to 250\%.
 Overall, 
-these results give strong evidence that our technique consistently benefits bug-finding ability, sometimes drastically.
+these results give strong evidence that \camlname consistently benefits bug-finding ability, sometimes drastically.
 \begin{figure}[h]
@@
 \end{figure}
-
 \begin{figure}[h]
-  \includegraphics[scale=0.26]{swarm.png}
-  \caption{Swarm distribution of individual-task speedups across strategies and benchmarks, black dots represent tasks.
+  \includegraphics[scale=0.35]{swarm.png}
+  \caption{Swarm distribution of individual-task speedups across strategies and benchmarks, gray dots represent tasks.
 Plots with extremely low variance have no swarm overlay; in plots with high variance, outliers correspond to large speedups.

### 2025-03-20 paper/paper.tex
numbers on scala barchart
speed not ability

@@
     \item We demonstrate that both \name{} and choosing a faster \rand yield substantial performance improvements to generator running time. Finally, we show
-    that these performance improvements extend to significantly improved bug-finding ability. \jwc{This should be sell-ier: 2x speedups, etc. ``bug-finding ability'' isn't a race against time.}
+    that these performance improvements extend to significantly improved bug-finding speed. \jwc{This should be sell-ier: 2x speedups, etc. ``bug-finding speed'' isn't a race against time.}
 \end{enumerate}
@@
   \item \textbf{RQ1}: Do generators written using our technique run faster than those written with regular generator combinators?
-  \item \textbf{RQ2}: Do performance speedups translate to better bug-finding ability?
+  \item \textbf{RQ2}: Do performance speedups translate to better bug-finding speed?
   % \item \textbf{RQ3}: Does our technique generalize to other strict functional languages?
@@
 %}
-\subsection{Benchmarking Bug-Finding Ability}
-To answer \textbf{RQ2}, we evaluate the bug-finding ability of our BST and STLC
+\subsection{Benchmarking bug-finding speed}
+To answer \textbf{RQ2}, we evaluate the bug-finding speed of our BST and STLC
 case studies using Etna, a platform for assessing and comparing PBT techniques. 
@@
 \camlname; they are unchanged from their description in \secref{subsection:benchmarking}.\footnote{Etna does 
-not support Scala, so we were unable to test \scalaname, but we see no reason
+not support Scala, so we were unable to test \scalaname's bug-finding speed in Etna, but we see no reason
 we see no reason our results in this section would not extend to \scalaname.}
@@
 The average individual-task speedups for each strategy and benchmark are shown in 
-\figref{fig:etna}, with trends in bug-finding ability closely mirroring those in performance 
+\figref{fig:etna}, with trends in bug-finding speed closely mirroring those in performance 
 from \secref{subsection:benchmarking}. The distribution of speedups in \figref{fig:swarm} reveals 
@@
 Overall, 
-these results give strong evidence that \camlname consistently benefits bug-finding ability, sometimes drastically.
+these results give strong evidence that \camlname consistently benefits bug-finding speed, sometimes drastically.
 \begin{figure}[h]
   \includegraphics[scale=0.333]{speedup.png}
-  \caption{Geometric average of all speedups---relative to BQ---for each strategy and benchmark, showing that staging leads to better bug-finding ability across the board.}
+  \caption{Geometric average of all speedups---relative to BQ---for each strategy and benchmark, showing that staging leads to better bug-finding speed across the board.}
   \label{fig:etna}

### 2025-03-20 paper/paper.tex
nit

@@
 \camlname; they are unchanged from their description in \secref{subsection:benchmarking}.\footnote{Etna does 
-not support Scala, so we were unable to test \scalaname's bug-finding speed in Etna, but we see no reason
+not support Scala, so we were unable to test \scalaname's bug-finding speed in Etna, but we
 we see no reason our results in this section would not extend to \scalaname.}

### 2025-03-20 paper/paper.tex
words

@@
 For evaluating \scalaname, we implement a subset of the aformentioned 
-strategies---only the boolean list generator, single-pass BST generator, and well-typed STLC term generator.
-
-The runtime of the \camlname generators over a range of sizes is shown in
-\figref{fig:time}. Staging provides only a negligible speedup for type-derived
-and insertion-based BST strategies, whose overhead is dominated by random
-sampling. In this case, performance improvement is primarily attributable the
-\csplitmix \rand.  In contrast, the STLC strategies benefit more from staging, as
-they involve fewer sampling operations but rely heavily on generator
-combinators. The same holds for boolean lists, as sampling booleans is
-inexpensive relative to integers. Our single-pass BST strategy falls somewhere
-in between: it both samples frequently and makes extensive use of generator
-combinators. \tr{Goal is to replace this paragraph with the ref-counting
-chart---I think it will be WAY less boring. But I need to do 3.7 first.}
+strategies---only the boolean list generator, single-pass BST generator, and 
+well-typed STLC term generator.
+
+The runtime of \camlname generators across different input sizes is summarized in \figref{fig:time}. 
+Our \camlname type-derived BST generator achieves speedups ranging from 8\% to 29.8\%, which increase 
+dramatically to 515--865\% when combined with \csplitmix. The insertion-based BST generator sees 
+13--21\% speedups with staging alone, and 332--435\% when also 
+using \csplitmix. The single-pass BST generator benefits more significantly from 
+staging, with speedups of 131--134\%, which rise to 761--795\% when combined with 
+\csplitmix. We attribute the differential in performance speedup to the fact that 
+BST strategies rely on relatively few calls to generator combinator functions, 
+but sample extensively from the \rand; notably, single-pass BST, which uses the largest
+number of calls to generator combinators, also sees the greatest benefit from staging.
+
+For STLC, staging accounts for the bulk of performance gains, yielding
+ 71--164\% speedups for the type-derived STLC generator, and 360--446\% for 
+ the well-typed generator. When adding \csplitmix, these numbers increase to 
+ 160--415\% and 383--675\%, respectively. This result makes sense because
+ STLC generators, particularly the well-typed generator, sample very little from
+ randomness libraries, but extensively on generator library functions.
+
+The boolean list generator experiences 261--610\% speedups with 
+staging, and its performance actually slightly \textit{decreases} when 
+combined with \csplitmix (243--583\%). This is likely because sampling 
+booleans is ``cheap'' enough that the overhead of crossing the FFI barrier is 
+greater than that of generating values directly in OCaml.
+
+\tr{We might put a graph of calls to generator library functions vs. speedup here, but
+I'm not sure it's worth it---thoughts?}
 
@@
   \includegraphics[scale=0.35]{swarm.png}
-  \caption{Swarm distribution of individual-task speedups across strategies and benchmarks, gray dots represent tasks.
+  \caption{Swarm distribution of individual-task speedups across strategies and benchmarks---gray dots represent tasks.
 Plots with extremely low variance have no swarm overlay; in plots with high variance, outliers correspond to large speedups.

### 2025-03-20 paper/paper.tex
intro

@@
 % \jwc{Initial section here about eval}\tr{todo}
+To assess the impact of our approach, we evaluate the performance and bug-finding speed of 
+\name generators across a range of benchmarks. Our experiments compare generators 
+built using our technique against those implemented with standard generator combinators, 
+both with and without \csplitmix. Specifically, we implement two fully semantically 
+equivalent staged generator libraries that replicate the behavior of \bq 
+in OCaml and ScalaCheck in Scala. This allows us to assess the effectiveness of our 
+technique across different languages and runtime environments, ensuring a fair comparison 
+with existing combinator-based approaches.
+
+This section presents our experimental setup, details on the benchmarking tools and environments, and answers to the following research questions:
+
+To evaluate our contributions, this section answers the following research questions:
+\begin{itemize}
+  \item \textbf{RQ1}: Do generators written using our technique run faster than those written with regular generator combinators?
+  \item \textbf{RQ2}: Do performance speedups translate to better bug-finding speed?
+  % \item \textbf{RQ3}: Does our technique generalize to other strict functional languages?
+\end{itemize}
+
 All experiments were run on a 64-bit Linux machine with 264 GB RAM and an Intel Xeon Platinum 8375C CPU, running Ubuntu 24.04.1 LTS. \camlname and all OCaml benchmarks
@@
 We used \texttt{core\_bench} 0.16 \cn in OCaml and \texttt{jmh} 0.4.7 \cn for generator performance microbenchmarking.
-
-To evaluate our contributions, this section answers the following research questions:
-\begin{itemize}
-  \item \textbf{RQ1}: Do generators written using our technique run faster than those written with regular generator combinators?
-  \item \textbf{RQ2}: Do performance speedups translate to better bug-finding speed?
-  % \item \textbf{RQ3}: Does our technique generalize to other strict functional languages?
-\end{itemize}
 

### 2025-03-20 paper/paper.tex
a little more succinct

--- 2025-03-20 paper/paper.tex
@@
 % \jwc{Initial section here about eval}\tr{todo}
+To assess the impact of our approach, we evaluate the performance and bug-finding speed of 
+\name generators across a range of benchmarks. Our experiments compare generators 
+built using our technique---with and without \csplitmix---against those implemented with 
+standard generator combinators and default randomness libraries. Specifically, we implement 
+two fully semantically 
+equivalent staged generator libraries that replicate the behavior of \bq 
+in OCaml and ScalaCheck in Scala, allows us to assess the effectiveness of our 
+technique across different languages and runtime environments.
+
+This section presents our experimental setup, details on the benchmarking tools and environments, and answers to the following research questions:
+
+\begin{itemize}
+  \item \textbf{RQ1}: Do generators written using our technique run faster than those written with regular generator combinators?
+  \item \textbf{RQ2}: Do performance speedups translate to better bug-finding speed?
+  % \item \textbf{RQ3}: Does our technique generalize to other strict functional languages?
+\end{itemize}
+
 All experiments were run on a 64-bit Linux machine with 264 GB RAM and an Intel Xeon Platinum 8375C CPU, running Ubuntu 24.04.1 LTS. \camlname and all OCaml benchmarks

### 2025-03-20 paper/paper.tex
minor

@@
 built using our technique---with and without \csplitmix---against those implemented with 
-standard generator combinators and default randomness libraries. Specifically, we implement 
+existing generator libraries (and their default randomness mechanisms). Specifically, 
+we implement 
 two fully semantically 

### 2025-03-20 paper/paper.tex
blah

@@
 equivalent staged generator libraries that replicate the behavior of \bq 
-in OCaml and ScalaCheck in Scala, allows us to assess the effectiveness of our 
+in OCaml and ScalaCheck in Scala, allowing us to assess the effectiveness of our 
 technique across different languages and runtime environments.

### 2025-03-20 paper/paper.tex
minor
Intro comments

--- 2025-03-20 paper/paper.tex
@@
 
+\bcp{Feels like we're missing a trick in the transition here.  Can we
+  say, ``Because performance is important, PBT libraries are already
+  quite tuned.  [Can we substantiate that??]  But more is
+  possible...'' }
+
 Given these strict timelines, the impact of performance in PBT cannot be overstated. One important locus for performance improvement
@@
 \name generators across a range of benchmarks. Our experiments compare generators 
-built using our technique---with and without \csplitmix---against those implemented with 
-standard generator combinators and default randomness libraries. Specifically, we implement 
-two fully semantically 
+built using our technique against those implemented with standard generator combinators, 
+both with and without \csplitmix. Specifically, we implement two fully semantically 
 equivalent staged generator libraries that replicate the behavior of \bq 
-in OCaml and ScalaCheck in Scala, allows us to assess the effectiveness of our 
-technique across different languages and runtime environments.
+in OCaml and ScalaCheck in Scala. This allows us to assess the effectiveness of our 
+technique across different languages and runtime environments, ensuring a fair comparison 
+with existing combinator-based approaches.
 
@@
 
+To evaluate our contributions, this section answers the following research questions:
 \begin{itemize}

### 2025-03-20 paper/paper.tex

--- 2025-03-20 paper/paper.tex
@@
 
-This section presents our experimental setup, details on the benchmarking tools and environments, and answers to the following research questions:
+This section presents our experimental setup, and answers to the following research questions:
 

### 2025-03-20 paper/paper.tex
pauls eval wibbles

@@
 
-\bcp{Feels like we're missing a trick in the transition here.  Can we
-  say, ``Because performance is important, PBT libraries are already
-  quite tuned.  [Can we substantiate that??]  But more is
-  possible...'' }
-
 Given these strict timelines, the impact of performance in PBT cannot be overstated. One important locus for performance improvement
@@
 
-All experiments were run on a 64-bit Linux machine with 264 GB RAM and an Intel Xeon Platinum 8375C CPU, running Ubuntu 24.04.1 LTS. \camlname and all OCaml benchmarks
+All experiments were run on a 64-bit Linux machine with 264 GB RAM and a 128-core Intel Xeon Platinum 8375C CPU, running Ubuntu 24.04.1 LTS. \camlname and all OCaml benchmarks
 were compiled with 4.14.1+BER MetaOCaml \texttt{ocamlopt}, the native code compiler for OCaml, using compiler flag \texttt{-O3}.
@@
   \includegraphics[scale=0.32]{scala.png}
- \caption{\scalaname has better performance relative to ScalaCheck than \camlname relative to \bq. SC is ScalaCheck; BQ is \bq. Comparison is for a fixed size of 100.}
+ \caption{
+  \jwc{Don't explain the chart in the caption, split this into two different charts --- purple on one, green on the other.}
+  \jwc{Also can we use the language's colors for this?}
+  \scalaname has better performance relative to ScalaCheck than \camlname relative to \bq. SC is ScalaCheck; BQ is \bq. Comparison is for a fixed size of 100.}
  \label{fig:scala}

### 2025-03-20 paper/paper.tex
Comments on intro (more coming on PDF)

--- 2025-03-20 paper/paper.tex
@@
 
-To address this question, we outline two key principles for designing generator libraries that balance efficiency with flexibility.
+To address this question, we outline two key principles for designing
+generator libraries that balance efficiency with flexibility.\bcp{We
+  just said we want both, not a ``balance'' that trades one off
+  against the other...}
 First, we describe a method of optimizing generator libraries using
@@
 performance by conducting a controlled comparison.
-Together, we refer to this recipe for improving generator libraries as \name{} --- \jwc{something pithy here}.
+Together, we refer to this recipe\bcp{``demonstrating the impact'' is
+  not a recipe} for improving generator libraries as \name{} --- \jwc{something pithy here}.
 
@@
 high-quality PBT library in OCaml, authored by software engineers at the trading firm Jane Street \cn{}.
-In Scala 3, we build \scalaname{}, based on ScalaCheck --- a very mature PBT library and the de facto choice in Scala \cn{}.
+In Scala 3, we build \scalaname{}, based on ScalaCheck --- a very
+mature PBT library and the {\em de facto} choice\bcp{the wording here
+  implies that there are other choices in Scala, some of which might
+  be reasonable, and tht base\_quickcheck is not the de facto choice
+  in OCaml...} in Scala \cn{}.
 In both languages, our libraries are direct drop-in replacements that are
@@
 % ability \jwc{The paper is called ``fail faster'' --- we shouldn't say ``bug finding ability'', we should say something like ``how fast you can squash bugs''. This is too passive }
-by running these case studies in the Etna framework~\cite{etna}., a
+by running these case studies in the Etna
+framework~\cite{etna}\bcp{``Etna framework'' sounds like more than
+  what it is}, a
 platform that simulates real-world PBT usage\hg{Don't love this phrase, I'll
@@
 library~\cite{claessenQuickCheckLightweightTool2000}, {\em generators} have been
-a key element of the PBT process. While properties are critical for correctly
+a key element of the PBT process. While properties\bcp{this concept
+  comes out of nowhere} are critical for correctly
 expressing a developer's intent for a function, the success of the testing
 ultimately rests on how well the random inputs to that property exercise the
-system under test. QuickCheck's solution to this problem, which has been copied
-dozens of frameworks across as many languages, is to provide a domain-specific
+system under test.\bcp{Is this trying to say that generators are more
+  important than properties?  This seems like a strange assertion.}
+QuickCheck's solution to this problem, which has been copied in
+dozens of frameworks across a wide variety of languages, is to provide a domain-specific
 language (DSL) for writing random data generators that gives developers complete
-control over their distribution of test inputs.
+control\bcp{that might be overstating the case a bit!} over the
+distribution of their test inputs.
 

### 2025-03-20 paper/paper.tex
more graph wibbles

--- 2025-03-20 paper/paper.tex
@@
   \includegraphics[scale=0.333]{speedup.png}
-  \caption{Geometric average of all speedups---relative to BQ---for each strategy and benchmark, showing that staging leads to better bug-finding speed across the board.}
+  \caption{Geometric average of all speedups---relative to BQ---for each strategy and benchmark, showing that staging leads to better bug-finding speed across the board.
+  \jwc{Charts start at one? Can eliminate the BQ bar}
+  }
   \label{fig:etna}

### 2025-03-20 paper/paper.tex
times not x

@@
 The runtime of \camlname generators across different input sizes is summarized in \figref{fig:time}. 
-Our \camlname type-derived BST generator achieves speedups ranging from 8\% to 29.8\%, which increase 
-dramatically to 515--865\% when combined with \csplitmix. The insertion-based BST generator sees 
-13--21\% speedups with staging alone, and 332--435\% when also 
+Our \camlname type-derived BST generator achieves speedups ranging from 1.08--1.29x, which increase 
+dramatically to 6.13--9.65x when combined with \csplitmix. The insertion-based BST generator sees 
+1.13--1.21x speedups with staging alone, and 4.32--5.36x when also 
 using \csplitmix. The single-pass BST generator benefits more significantly from 
-staging, with speedups of 131--134\%, which rise to 761--795\% when combined with 
+staging, with speedups of 2.31--2.34x, which rise to 8.61--8.95x when combined with 
 \csplitmix. We attribute the differential in performance speedup to the fact that 
@@
 For STLC, staging accounts for the bulk of performance gains, yielding
- 71--164\% speedups for the type-derived STLC generator, and 360--446\% for 
+ 1.71--2.64x speedups for the type-derived STLC generator, and 4.60--5.46x for 
  the well-typed generator. When adding \csplitmix, these numbers increase to 
- 160--415\% and 383--675\%, respectively. This result makes sense because
+ 2.60--5.15x and 4.83--7.75x, respectively. This result makes sense because
  STLC generators, particularly the well-typed generator, sample very little from
@@
 
-The boolean list generator experiences 261--610\% speedups with 
+The boolean list generator experiences 3.61--7.10x speedups with 
 staging, and its performance actually slightly \textit{decreases} when 
-combined with \csplitmix (243--583\%). This is likely because sampling 
+combined with \csplitmix (3.43--6.83x). This is likely because sampling 
 booleans is ``cheap'' enough that the overhead of crossing the FFI barrier is 
@@
 find that ``easier'' tasks---those that run on the order of milliseconds---regularly achieve speedups in the 
-range of 300 to 800\%, whereas tasks that run on the order of seconds achieve more moderate speedups of up to 250\%.
+range of 4--9X, whereas tasks that run on the order of seconds achieve more moderate speedups of up to 3.5X.
 Overall, 

### 2025-03-20 paper/paper.tex
figure pass
more
x

@@
   \centering
-  \includegraphics[scale=0.32]{scala.png}
+  \includegraphics[scale=0.4]{scala.png}
  \caption{
@@
 find that ``easier'' tasks---those that run on the order of milliseconds---regularly achieve speedups in the 
-range of 4--9X, whereas tasks that run on the order of seconds achieve more moderate speedups of up to 3.5X.
+range of 4--9x, whereas tasks that run on the order of seconds achieve more moderate speedups of up to 3.5x.
 Overall, 
@@
 \begin{figure}[h]
-  \includegraphics[scale=0.333]{speedup.png}
+  \includegraphics[scale=0.3]{speedup.png}
   \caption{Geometric average of all speedups---relative to BQ---for each strategy and benchmark, showing that staging leads to better bug-finding speed across the board.
@@
 \begin{figure}[h]
-  \includegraphics[scale=0.35]{swarm.png}
+  \includegraphics[scale=0.31]{swarm.png}
   \caption{Swarm distribution of individual-task speedups across strategies and benchmarks---gray dots represent tasks.

### 2025-03-20 paper/paper.tex
polishing s3

--- 2025-03-20 paper/paper.tex
@@
 
-In Muli-stage programming (also known more simply as staging), programs execute
-over the course of multiple stages, with each stage producing the code to be run
+In muli-stage programming (also known more simply as staging), programs execute
+in multiple stages, with each stage producing the code to be run
 in the next. For the purposes of this paper, we will
-primarily consider two stages, namely compile time and run time. Staged programs thus
-execute twice: once at compile time, which produces more code, which is then
+consider two stages: namely compile time and run time. In our context, staged programs
+execute once at compile time, which produces more code, which is then
 compiled and run at run time.
 
-\subsubsection{Uses of Staging in eDSL Construction}
+\paragraph{Uses of Staging in eDSL Construction}
 One of the primary uses of staging is in embedded DSL construction \cite{sheard99, lms, trattdsl}.
@@
 the functional abstractions used to build eDSLs prevent compilers from generating efficient machine code \cn.
-This effect is known as ``abstraction overhead'' \cite{carrette05, moller20}: the layers of abstraction that 
+This is abstraction overhead \cite{carrette05, moller20}: the layers of abstraction that 
 make eDSLs so usable are precisely what prevents them from being fast.
-Indeed, the root causes of many of the issues we discovered in Section~\ref{section:motiv} are
+Indeed, the root causes of many of the performance issues we discovered in Section~\ref{section:motiv} are
 not unique to generator DSLs.
@@
 transforms the DSL code, eliminating abstractions to produce code that the host language compiler can generate fast machine code for.
-This recipe has been used to great effect across domains to stage eDSLs for stream processing \cite{moller20,strymonas}, parser combinators \cite{sspc, staged-parsers, krishnaswami19, flap},
+This recipe has been used to great effect to stage eDSLs for stream processing \cite{moller20,strymonas}, parser combinators \cite{sspc, staged-parsers, krishnaswami19, flap},
 and query processing \cite{rhyme}.
 
-Many languages have some degree of staging functionality, though support varies
-from full native support (Scala 3 \cite{scalamacros}, Haskell \cite{templatehaskell}, \jwc{Racket??}) to compiler extensions (OCaml \cite{metaocaml,macocaml}, Java \cite{mint}).
+Many languages have some degree of staging functionality, including
+Scala 3 \cite{scalamacros}, Haskell \cite{templatehaskell}, \jwc{Racket??}, OCaml \cite{metaocaml,macocaml}, and Java \cite{mint}.
 For this paper, we have implemented staged PBT libraries both in OCaml --- using the MetaOCaml fork of the OCaml compiler \cn --- and Scala 3 --- using
 \texttt{scala.quoted}. For presentation purposes, all staged code presented in the body of this paper is \camlname{} code written in OCaml:
-\scalaname{} differs almost entirely in syntax alone.
-For a further discussion of the potential of building staged generator DSLs in other languages, see Section~\ref{section:other-langs}.
-
-\subsubsection{Staging in MetaOCaml}
-MetaOCaml's staging functionality is exposed through a type \texttt{'a code}. A
-value of type \texttt{t code} at compile time is a (potentially open) OCaml term
+\scalaname{} is very similar.
+For further discussion of the potential for building staged generator DSLs in other languages, see Section~\ref{section:other-langs}.
+
+\paragraph{Staging in MetaOCaml}
+MetaOCaml's staging functionality is exposed through the type \texttt{'a code}. A
+value of type \texttt{t code} is an OCaml term
 of type \texttt{t}.
-
 Values of \texttt{code} type are introduced by \emph{quotes}, written
-\texttt{.<$\ldots$>.}. Brackets delay execution of a program until run time. For
+\texttt{.<$\ldots$>.}. Quotes delay execution of a program until run time. For
 example, the program \texttt{.< 5 + 1 >.} has type \texttt{int code}.  Note that
@@
 the code is not \emph{executed} until the next stage (run time).
-Values of type \texttt{code} can be combined together using \emph{escape},
-written \texttt{.\~{}(e)} (or just \texttt{.\~{}x}, when \texttt{x} is a variable).  Escape lets you take a value of type
-\texttt{code}, and ``splice'' it directly into a quote.  For example, this
+Values of type \texttt{code} can be combined together using an \emph{escape},
+written \texttt{.\~{}(e)} (or just \texttt{.\~{}x}, when \texttt{x} is a variable).  Escaping lets you take a value of type
+\texttt{code}, and ``splice'' it directly into a quote.  For example, the
 program \texttt{let x = .<1 * 5>. in .< .\~{}(x) + .\~{}(x) >.} evaluates to
@@
 
-The power of staging for optimization away abstraction overheads comes from
+The power of staging for optimizing away abstraction overheads comes from
 defining functions that accept and return \texttt{code} values.  A function
@@
 % USE THIS: https://arxiv.org/pdf/2309.08207
-It returns a code for a function that takes an argument \texttt{x}, and then it splats in the result of calling \texttt{f} on just
+It returns a code for a function that takes an argument \texttt{x} and splices in the result of calling \texttt{f} on just
 the quoted \texttt{x}.
@@
 let succ x = .< 1 + .~x >. in
-eta (succ . is_even)
+eta (fun x -> succ (is_even x))
 \end{lstlisting}
@@
 can ensure that any overhead of using the combinators is eliminated by run time.
+\jwc{BCP: You wrote a ``?'' next to this section. What's the issue?}
 
@@
 To build a staged version of a generator DSL, we must rewrite the the generator
-combinators so that they run during the compile time. Of course,
-the compile time stage does not do all the work: the DSL must run at compile time,
-producing inlined code that uses no combinators, which can then be compiled and run at run time
+combinators to do as much as possible at compile time, 
+producing inlined code that uses no combinators, which can be compiled and run
 with different sizes and seeds.
@@
 Our job is thus to carefully bisect the DSL, determining which inputs to generator
-combinators are known statically (and can be part of the compile time stage), and
+combinators are known statically (and can be part of the compile time stage) and
 which parts are only known at run time (and must be \texttt{code}).
-In the staging literature, this task is known as ``binding-time analysis'', and it is more of an art than a science.
+In the staging literature, this task is known as ``binding-time analysis'' and it is more an art than a science.
 
@@
 size parameters that will be passed in at run time. Generators themselves are always known entirely at compile time.
-This means our library's generator type \texttt{'a
-Gen.t} should have the type \texttt{int code -> SR.t code -> 'a code}: a compile-time
+This leads us to define our library's generator type \texttt{'a
+Gen.t} as \texttt{int code -> SR.t code -> 'a code}: a compile-time
 function from dynamically-known size and seed to dynamically-determined result.
 
-This type, along with basic staged monadic generator DSL functionality can be found in Figure~\ref{fig:gen-staged-basic}.
+The basic staged monadic generator DSL functionality can be found in Figure~\ref{fig:gen-staged-basic}.
 The constant generator \texttt{return} has the same code, but this time it 
@@
 \caption{Basic Staged Generator Library}
-\ref{fig:gen-staged-basic}
+\label{fig:gen-staged-basic}
 \end{figure}
@@
 \caption{Pairs of Ints, Staged}
-\ref{fig:running-staged}
+\label{fig:running-staged}
 \end{figure}

### 2025-03-21 paper/paper.tex
through 3.7

@@
 The basic staged monadic generator DSL functionality can be found in Figure~\ref{fig:gen-staged-basic}.
-The constant generator \texttt{return} has the same code, but this time it 
-runs at compile time. Given \texttt{cx : 'a code}, the code for a \texttt{'a}, it
-returns the generator which always generates that value. Similarly,
+The constant generator \texttt{return} runs at compile time. Given \texttt{cx : 'a code}, the code for an \texttt{'a}, it
+returns the generator that runs \texttt{cx}. Similarly,
 \texttt{bind g k} sequences generators as usual by passing the result of running the generator \texttt{g} to a continuation \texttt{k}. However, instead of getting access
 to the particular value generated by \texttt{g}, the continuation \texttt{k} gets access to \texttt{code} for the value sampled from \texttt{g}:
-at compile time, we only know \texttt{g} will generate \emph{some} \texttt{'a}, but not which one \footnote{Readers familiar with OCaml may notice that \texttt{return : 'a code -> 'a Gen.t} and \texttt{bind : 'a Gen.t -> ('a code -> 'b Gen.t) -> 'b Gen.t}
-do not have the correct types for a monad instance, preventing us from using \texttt{let\%bind} notation.
+at compile time, we only know \texttt{g} will generate \emph{some} \texttt{'a}, but not which one
+% \footnote{Readers familiar with OCaml may notice that \texttt{return : 'a code -> 'a Gen.t} and \texttt{bind : 'a Gen.t -> ('a code -> 'b Gen.t) -> 'b Gen.t} do not have the correct types for a monad instance, preventing us from using \texttt{let\%bind} notation.
 We rectify this issue in Section~\ref{subsection:codecps} by importing some clever ideas from the staging literature.}.
-Operationally, bind takes code for the size and seed, and returns code that (1) let-binds a variable \texttt{a} to spliced-in code that runs \texttt{g}, and then
+\jwc{Fix explanation here: bcp says ``this explanation doesn't quite explain what's going on.
+after all, in a staged program the contnuatino k doesn't know which values is going to be passed to it.
+the difference in the staged setting is that \texttt{k} is not allowed to ask any questions about its argument, it has to treat it totally parametrically.}
+Operationally, bind takes code for the size and seed and returns code that (1) let-binds a variable \texttt{a} to spliced-in code that runs \texttt{g} and then
 (2) runs the spliced-in continuation \texttt{k}.
@@
 inlined code that results from calling \texttt{Gen.to\_bq} (changing some identifier names for clarity).
-The code generated is identical to the manually inlined version from Section~\ref{section:motiv}, and of course runs equally fast.
+The code generated is identical to the manually inlined version from Section~\ref{section:motiv}.
 
@@
 In Section~\ref{section:motiv}, we noted that generator combinators like \texttt{weighted\_union}
-often allocate lists in the hot path of the generator. Even though these lists are usually small ---
-at most a few dozen elements in practice --- each allocation takes us closer to the next garbage collection.
+allocate lists in the hot path of the generator. Even though these lists are usually small ---
+at most a few dozen elements --- each allocation takes us closer to the next garbage collection.
 
@@
 so they are \texttt{code}s.
-Instead of building a histogram of the distribution described by the weights at run time and then traversing it,
+Instead of building a data structure representing a histogram of the distribution described by the weights at run time and then traversing it,
 the compile time \texttt{weighted\_union} combinator generates a
-tree of \texttt{if}s --- specialized to the list of weights known at compile time --- that searches for the selected generator.
+tree of \texttt{if}s specialized to the list of weights known at compile time, that searches for the selected generator.
 
@@
 \subsection{Let-Insertion and Effect Ordering}
-Careful readers might note that the definition of \texttt{bind} (shown in Figure~\ref{fig:bind-with-let-binding})
-was more complicated than one might expect. In particular, why not define bind in the standard way
+Careful readers might note that the definition of \texttt{bind} (Figure~\ref{fig:bind-with-let-binding})
+was more complicated than one might expect. Why not define bind in the standard way
 for a reader monad (Figure~\ref{fig:bind-without-let-binding})?
-Unfortunately, the \texttt{bind'} definition is wrong in our context as it leads to incorrect code being generated.
+Unfortunately, the latter definition is wrong in our context as it leads to incorrect code being generated.
 For example, consider \texttt{Gen.bind' (Gen.int .<0>. .<1>.) (fun x -> Gen.return .<(.~x,.~x)>.)}.
@@
 This matters because, as described in Section~\ref{section:intro}, \camlname and
-\scalaname should be equivalent to their unstaged counterparts.
+\scalaname are intended to be equivalent to their unstaged counterparts.
 
@@
 In essence, the behavior of splice \texttt{.\~{}cx} in a staged function \texttt{f(cx : 'a code) = ...} is to \emph{copy}
-the entire block of code, effects and all. To ensure that the randomness effects of the first generator are executed only once,
-but that the value can be used in the continuation multiple times, \texttt{bind} let-binds the result of generation to a variable,
+the entire block of code, effects and all. To ensure that the randomness effects of the first generator are executed only once
+but that the value can be used in the continuation multiple times, the correct \texttt{bind} let-binds the result of generation to a variable
 and then passes it to the continuation.
@@
 Another subtle issue prevents the version of the library design discussed so far
-from being used as a proper drop-in replacement for an existing generator DSL. In particular, the types of \texttt{return : 'a code -> 'a Gen.t}
+from being used as a drop-in replacement for an existing generator DSL: the types of \texttt{return : 'a code -> 'a Gen.t}
 and \texttt{bind : 'a Gen.t -> ('a code -> 'b Gen.t) -> 'b Gen.t} aren't quite right.
 For the type \texttt{'a Gen.t} to actually be a monad, 
-the these types cannot mention \texttt{code}. This is not just a theoretical issue, it is a significant usability concern:
-the syntactic sugar for monadic programming (\texttt{let\%bind} in OCaml, \texttt{foreach} in Scala, \texttt{do} in Haskell, etc)
-that makes it so appealing can \emph{only} be used if the types involved are
-actually the proper monad function types.
+these types cannot mention \texttt{code}. This is not just a theoretical issue; it is a significant usability concern.
+The syntactic sugar for monadic programming (\texttt{let\%bind} in OCaml, \texttt{foreach} in Scala, \texttt{do} in Haskell, etc)
+that makes it smooth can \emph{only} be used if
+return has type \texttt{'a -> 'a Gen.t} and bind has type \texttt{'a Gen.t -> ('a -> 'b Gen.t) -> 'b Gen.t}
 \iflater
@@
 
-To support real monadic programming, we'll need to adjust the type of \texttt{'a Gen.t} slightly.
+To support convenient monadic programming, we need to adjust the type of \texttt{'a Gen.t} slightly.
 An initial attempt is to try \texttt{type 'a t = int code -> SR.t code -> 'a}. If we strip the \texttt{code} off the result type,
@@
 However, this definition of bind doesn't have call-by value effect semantics, as
-discussed in the previous section!  And because the type of \texttt{g size seed} is
+discussed in the previous section.  And because the type of \texttt{g size seed} is
 now just \texttt{'a} (not necessarily \texttt{'a code}), we cannot perform the
@@
 \jwc{I could, but i'm just not sure this is important. It's too deep in the weeds for it to be worth explaining, too... nobody's going to read this section except staging fans}
+\bcp{...
+but they are gonig to be sad if they don't understand the details.
+If we're not too desperate for space, my suggestion would be to put in some more details plus sign posting that says tehy can be skipped.
+If we're up against space limits, then maybe most of the section should go to the appendix.
+Or second thought, maybe we should lean into the cmplexity here (and everywhere else) to give the impressions that what we've done is not simply applying known ideas in sraightforward ways}
+\jwc{Unfortunately, I think this technique is sufficiently well-understood to staging experts that we can't claim novelty. We'll jsut cut it for space if we need it.}
 
@@
 the type \texttt{'a CodeCps.t = 'z. ('a -> 'z code) -> 'z code}: a polymorphic continuation transformer with the result type
-always in \texttt{code}
-\footnote{This is an instance of the \emph{codensity} monad \cite{janis08}, a fact which deserves further investigation.}.
+always in \texttt{code}\footnote{This is an instance of the \emph{codensity} monad \cite{janis08}, a fact which deserves further investigation.}.
 The monad instance for this type is the standard instance for a CPS monad with polymorphic return type.
-In prior work, this type is often referred to as the ``code generation'' monad
-(and sometimes, ironically, called ``\texttt{Gen}'').
+In prior work, this type is often referred to as the ``code generation'' monad.
 This is because a value of type \texttt{('a code) CodeCps.t} is like an ``action'' that generates \texttt{code}:
@@
 This design is less obviously correct, and does require some care. Rather than \texttt{bind} ensuring correct evaluation order once and for all,
-individual combinators must be carefully written to ensure that \texttt{'a code} values that contian effects are \texttt{let\_insert}'d.\footnote{To ensure that the OCaml code is written correctly, we employ a PBT harness to compare its outputs to the corresponding \bq generator over 1,000 random seeds.
-If no discrepancy arises, we conclude that the generators exhibit equivalent behavior.}
+individual combinators must be carefully written to ensure that \texttt{'a code} values that contian effects are \texttt{let\_insert}'d.
+To ensure that the OCaml code is written correctly, we employ a PBT harness to compare its outputs to the corresponding \bq generator over 1,000 random seeds. 
+If no discrepancy arises, we conclude that the generators exhibit equivalent behavior.
 
@@
 to be defined as recursive functions, while others
-(like \bq and ScalaCheck) expose a fixpoint combinator to construct recursive generators.
+(including both \bq and ScalaCheck) expose a fixpoint combinator to construct recursive generators.
 Given a step function that takes a ``handle'' to sample from a recursive generator call, it
@@
 
-In our case, letting programmers define recursive generators as recursive functions is out of the question.
+In our setting, letting programmers define recursive generators as recursive functions is out of the question.
 With staged programming, recursion must be handled with care: it is far to easy to accidentally recursively
@@
 Programmers can then define recursive generators by \texttt{fixed\_point}, which ties the recursive knot.
+\bcp{Again, we could lean into the technical contribution here}\jwc{This is also totally standard.}
 

### 2025-03-21 paper/paper.tex
done through 227

@@
 
-While many DSL designs are possible, the standard design for QuickCheck-style
-generators is via an embedded {\em monadic} language. For the purposes of this
-paper, we use the syntax and types from OCaml's \bq library, which is presented in
-Figure~\ref{fig:bq-api}, but these kinds of DSLs can also be found in languages
+The standard design for QuickCheck-style
+generators is via an embedded {\em monadic} language. We use the syntax and types from OCaml's \bq library, which is presented in
+Figure~\ref{fig:bq-api}, but similar DSLs can also be found in languages
 like Haskell, Scala, Python, and many more.
@@
 example \texttt{gen\_int} for generating random integers in a range, along with
-the ``monadic interface'' consisting of \texttt{return} and \texttt{bind}.  The
+the functions \texttt{return} and \texttt{bind}.  The
 generator \texttt{return x} is the constant generator, always generating the
@@
 built with these operations; it first samples an int between $0$ and $100$,
-names it $x$, samples another between $0$ an $x$, and then returns the pair of
-them both.
+names it $x$, samples another between $0$ and $x$, and then returns the pair of
+them.
 
@@
 \end{lstlisting}
-\caption{An equivalent generator using macros for \texttt{bind}.}\label{fig:simple-macro}
+\caption{An equivalent generator using the macro for \texttt{bind}.}\label{fig:simple-macro}
   \end{subfigure}
@@
 \texttt{bind e (fun x => e')}. Figure~\ref{fig:simple-macro} shows the same
-generator as before, now written the monadic syntax.
+generator, written the monadic syntax.
 
@@
 Generator libraries also include other functions that make generator
-construction easier; some examples of these are also shown in
+construction easier; some examples of these are included in
 Figure~\ref{fig:bq-api}.  The \texttt{weighted\_union} function is a
-particularly well-used one: it makes making a weighted choice between different
+particularly well used one: it makes making a weighted choice between different
 generators, allowing the developer to combine different sub-generators into a
@@
 
-The generator in Figure~\ref{fig:tree_of} uses all of the aforementioned
+The generator in Figure~\ref{fig:tree_of} uses all of these
 features. It uses \texttt{fixed\_point} to define a recursive generator that
 reads the current value of \texttt{size} to determine how to generate a tree.
-The generator uses \texttt{weighted\_union} to make a random choice between an
+It uses \texttt{weighted\_union} to make a random choice between an
 empty tree and a node, choosing a node with weight proportional to the current
-size and choosing a leaf otherwise. When generating a node, the generator uses
+size and choosing a leaf otherwise. When generating a node, it uses
 \texttt{with\_size} to reduce the value of the size parameter for future
@@
 A generator \texttt{'a Bq.t} is just a function of type~\texttt{int -> SR.t -> 'a},
+\jwc{introduce what SR is}
 taking an \texttt{int} representing the current size parameter and a random seed \texttt{SR.t},
-and returning a generated value \texttt{'a}. It is an invariant of the library that function of this type are
-deterministic: for a fixed size and seed, it will return the same value. All of the randomness in testing comes from varying the initial seed.
+and returning a generated value \texttt{'a}. It is an invariant of the generator library that every function of this type is
+deterministic: for a fixed size and seed, it will always return the same value, so all of the randomness in testing comes from varying the initial seed.
 The monad functions \texttt{return} and \texttt{bind} are defined in the usual way
-for an instance of the reader monad \cn{}. 
+for an instance of the reader monad \cn{} \jwc{explain better}.
 % \bq{} uses a mutable seed, so the \texttt{random} passed to \texttt{(k a)} will be different from the one passed to \texttt{g}.
 The \texttt{gen\_int} combinator simply calls out to the randomness library \texttt{Splittable\_random}, aliased as \texttt{SR} here.
+\jwc{lift this explanation, explain what splittable random is before here.}
 
 Different PBT libraries use variations on this basic design. The largest source of variation is how the random seed is threaded through the generator.
-\bq{} uses a reader monad with a mutable seed, while Haskell's QuickCheck uses a reader monad with an immutable state
+\bcp{Sentence starts with lowerase} \bq{} uses a reader monad with a mutable seed, while Haskell's QuickCheck uses a reader monad with an immutable state
 that is ``split'' at \texttt{bind}s \cn. Meanwhile, ScalaCheck (a) uses an immutable state type and state monad \cn{} to thread
 the state through, and (b) has an \texttt{Option} as its return type, to allow generation to fail.
+\bcp{This discussion should go elsewhere...}
 
@@
 integers, written in both \bq{} and ScalaCheck.
-Figure~\ref{fig:overhead-explanation-code} shows two versions of the generator, written in both languages.
+Figure~\ref{fig:simplify} shows two versions of the generator, written in both languages.
 The first versions (\texttt{int\_pair} in \bq{} and \texttt{intPair} in ScalaCheck) are written with the monadic generator combinators from their respective libraries.
@@
 by (1) inlining all generator combinator definitions, and then (2)
-repeatedly reducing simplifiable terms like \lstinline{(fun x -> e) e'} --- where an anonymous function is defined and then immediatly called~\footnote{This is usually known as a ``$\beta$-redex''.
-Note that when reducing, we do not reduce to \lstinline{e[e'/x]}, as this does not preserve the order of effects.
-}
---- to \lstinline{let x = e' in e}.
-
-\begin{figure}
+repeatedly reducing simplifiable terms like \lstinline{(fun x -> e) e'} where an anonymous function is defined and then immediately called, to \lstinline{let x = e' in e}.
+
+\begin{figure}[h]
+  \begin{subfigure}{.49\textwidth}
 \begin{lstlisting}
@@
     (x,y)
-
+\end{lstlisting}
+\caption{Simplifying a Generator in OCaml}\label{fig:simplify-ocaml}
+  \end{subfigure}
+  \begin{subfigure}{.49\textwidth}
+\begin{lstlisting}
 def intPair : Gen[(Long,Long)] = for {
@@
 \end{lstlisting}
-\caption{Int Pair Generators in \bq{} and ScalaCheck}
-\label{fig:overhead-explanation-code}
+\caption{Simplifying a Generator in Scala}\label{fig:simplify-scala}
+  \end{subfigure}
+  \caption{Simplifying Generators}\label{fig:simplify}
 \end{figure}
 
-The performance impact of this inling is large (Figure~\ref{fig:overhead-explanation-perf}). In both languages, the
-inlined version takes (on average) half as much time to generate a single pair of
+% \begin{figure}
+% \begin{lstlisting}
+% \end{lstlisting}
+% \caption{Int Pair Generators in \bq{} and ScalaCheck}
+% \label{fig:overhead-explanation-code}
+% \end{figure}
+
+The performance impact of this simplification is large (Figure~\ref{fig:overhead-explanation-perf}). In both languages, the
+inlined version takes on average half as much time to generate a single pair of
 \texttt{int}s. 
 Microbenchmarks of more realistic generators (see Section~\ref{section:eval})
-show an even more dramatic performance boost for inlining in this manner.
+show an even more dramatic performance boost.
 
@@
 transformations of this form.}.  Even in cases as simple as
-Figure~\ref{fig:overhead-explanation-code}, the indirection of \texttt{return}
+Figure~\ref{fig:simplify}, the indirection of \texttt{return}
 and \texttt{bind} causes these heuristics to not fire.  Moreover, the story is

### 2025-03-21 paper/paper.tex
eval

@@
 
-Property-based testing (PBT) is a software testing technique which uses
-random test inputs to validate logical specifications. Recent studies on PBT usage~\cite{inpractice} have shown that
+Property-based testing (PBT) is a software testing technique that uses
+random test inputs to validate logical specifications. Recent studies on PBT usage~\cite{inpractice} in industry have shown that
 practitioners often interleave PBT with their software development process, sometimes
@@
 window to find bugs --- e.g., between 50 milliseconds and 30 seconds~\cite{inpractice}.
-In other words, the goal is to fail as fast as possible. \jwc{First cut at ``fail fast''}
+The goal of PBT is to fail tests as fast as possible. \jwc{First cut at ``fail fast''}
 
@@
 
-Given these strict timelines, the impact of performance in PBT cannot be overstated. One important locus for performance improvement
+One important locus for performance improvement
 is the \textit{generators} that produce random inputs to the properties under
@@
 is possible. This is due to two major sources of inefficiency.
-First, the same high-level design that gives generator libraries their
-flexibility and expressiveness also introduces overhead: layered abstractions
-and indirections hinder compiler optimizations, shifting work to
-run time.
-Second, the \rand that underlies the generator's operation---the source of the
-random numbers that are used to make choices in the generator---can
-constitute an unexpectedly high proportion of a generator's running time.
+First, the same high-level design approach that gives generator libraries their
+flexibility also introduces layered abstractions and indirections that are difficult for
+compilers to optimize, shifting work to run time.
+Second, the \rand that the generator uses to make choices can
+constitute an unexpectedly high proportion of its runtime.
 \iflater
@@
 To address this question, we outline two key principles for designing
-generator libraries that balance efficiency with flexibility.\bcp{We
-  just said we want both, not a ``balance'' that trades one off
-  against the other...}
+generator libraries that marry efficiency with flexibility.
 First, we describe a method of optimizing generator libraries using
@@
 Secondly, we demonstrate the impact of different choices of \rand on generator
-performance by conducting a controlled comparison.
-Together, we refer to this recipe\bcp{``demonstrating the impact'' is
-  not a recipe} for improving generator libraries as \name{} --- \jwc{something pithy here}.
+performance by conducting a controlled comparison, showing that showing that certain 
+implementations significantly reduce overhead.
+Together, these insights highlight practical strategies for improving generator libraries, 
+combining the efficiency gains of staging with a careful evaluation of \rand to optimize performance.
 
@@
 
-This section presents our experimental setup, and answers to the following research questions:
+This section presents our experimental setup, as well as answers to the following research questions:
 
@@
 For evaluating \scalaname, we implement a subset of the aformentioned 
-strategies---only the boolean list generator, single-pass BST generator, and 
+strategies---the boolean list generator, single-pass BST generator, and 
 well-typed STLC term generator.

### 2025-03-21 paper/paper.tex
intro

@@
 testing their properties every time they save their code. This means that PBT is often given a very short time
-window to find bugs --- e.g., between 50 milliseconds and 30 seconds~\cite{inpractice}.
-The goal of PBT is to fail tests as fast as possible. \jwc{First cut at ``fail fast''}
-
+window to find bugs---e.g., between 50 milliseconds and 30 seconds~\cite{inpractice}.
+The faster a test fails, the better.
+% The goal of PBT is to fail tests as fast as possible. \jwc{First cut at ``fail fast''}
+\iflater 
 \bcp{Feels like we're missing a trick in the transition here.  Can we
@@
   possible...'' }
+\fi
 
@@
 libraries}, usually expressed as embedded domain-specific languages (eDSLs) %\cntxt{edsls}
-that provide combinators for building and composing generators. Optimizing these
-libraries is a key opportunity to speed up PBT: if programmers are
-provided with the tools to write fast generators, then they do not themselves
-need to be experts in performance tuning to find bugs fast.
-
-However, careful measurements show that the performance of existing generator libraries falls significantly short of what
-is possible. This is due to two major sources of inefficiency.
+that provide combinators for building and composing generators. However, careful measurements show that the performance of existing generator libraries falls far short of what
+is possible.
+
+This is due to two major sources of inefficiency.
 First, the same high-level design approach that gives generator libraries their
@@
 compilers to optimize, shifting work to run time.
-Second, the \rand that the generator uses to make choices can
-constitute an unexpectedly high proportion of its runtime.
+Second, the \rand that a generator uses to make choices can
+constitute an unexpectedly high proportion of its runtime. Combined, these factors limit 
+the effectiveness of PBT.
 \iflater
@@
 \fi
-Together, these observations raise a fundamental question: can generator libraries be expressive and high-level without compromising efficiency?
-
-To address this question, we outline two key principles for designing
-generator libraries that marry efficiency with flexibility.
-First, we describe a method of optimizing generator libraries using
-\textit{multi-stage programming}, or ``staging,'' which has long been used to
-build DSLs without abstraction overhead (``without regret''~\cn{}).
-We apply this technique to generators, completely eliminating the run-time overhead of
+
+We lay a path forward.
+First, we show how to use \textit{multi-stage programming}---a classic technique for 
+building fast DSLs “without regret”~\cn{}---to eliminate the overhead of 
 many common generator abstractions.
-Secondly, we demonstrate the impact of different choices of \rand on generator
-performance by conducting a controlled comparison, showing that showing that certain 
-implementations significantly reduce overhead.
-Together, these insights highlight practical strategies for improving generator libraries, 
-combining the efficiency gains of staging with a careful evaluation of \rand to optimize performance.
-
-To demonstrate the effectiveness of our approach, we build two different PBT libraries in two different langauges.
-In OCaml, we build \camlname{}, based on \bq{}  --- a
+Second, we precisely characterize the impact of \rand choice through a natural experiment that preserves 
+generator behavior while improving execution speed, demonstrating that the entropy source itself should be treated as a first-class 
+concern in generator library design.
+Taken together, these insights can be applied to create tomorrow's generator libraries: those that 
+combine high-level design with low-level performance.
+
+To demonstrate the effectiveness and generality of our approach, which we call \name, we implement staged PBT libraries in multiple languages.
+In OCaml, we build \camlname{}, based on \bq{}---a
 high-quality PBT library in OCaml, authored by software engineers at the trading firm Jane Street \cn{}.
-In Scala 3, we build \scalaname{}, based on ScalaCheck --- a very
+In Scala 3, we build \scalaname{}, based on ScalaCheck---a very
 mature PBT library and the {\em de facto} choice\bcp{the wording here

### 2025-03-21 paper/paper.tex
blah

@@
 many common generator abstractions.
-Second, we precisely characterize the impact of \rand choice through a natural experiment that preserves 
-generator behavior while improving execution speed, demonstrating that the entropy source itself should be treated as a first-class 
-concern in generator library design.
+Second, we characterize the impact of \rand choice through a natural experiment 
+that isolates the performance contribution of the \rand within a generator, underscoring its role as a first-class design consideration.
 Taken together, these insights can be applied to create tomorrow's generator libraries: those that 

### 2025-03-21 paper/paper.tex
do a little rewrite of 231-239

--- 2025-03-21 paper/paper.tex
@@
 
-Property-based testing (PBT) is a software testing technique that uses
-random test inputs to validate logical specifications. Recent studies on PBT usage~\cite{inpractice} in industry have shown that
+Property-based testing (PBT) is a software testing technique which uses
+random test inputs to validate logical specifications. Recent studies on PBT usage~\cite{inpractice} have shown that
 practitioners often interleave PBT with their software development process, sometimes
@@
 window to find bugs --- e.g., between 50 milliseconds and 30 seconds~\cite{inpractice}.
-The goal of PBT is to fail tests as fast as possible. \jwc{First cut at ``fail fast''}
+In other words, the goal is to fail as fast as possible. \jwc{First cut at ``fail fast''}
 
@@
 
-One important locus for performance improvement
+Given these strict timelines, the impact of performance in PBT cannot be overstated. One important locus for performance improvement
 is the \textit{generators} that produce random inputs to the properties under
@@
 is possible. This is due to two major sources of inefficiency.
-First, the same high-level design approach that gives generator libraries their
-flexibility also introduces layered abstractions and indirections that are difficult for
-compilers to optimize, shifting work to run time.
-Second, the \rand that the generator uses to make choices can
-constitute an unexpectedly high proportion of its runtime.
+First, the same high-level design that gives generator libraries their
+flexibility and expressiveness also introduces overhead: layered abstractions
+and indirections hinder compiler optimizations, shifting work to
+run time.
+Second, the \rand that underlies the generator's operation---the source of the
+random numbers that are used to make choices in the generator---can
+constitute an unexpectedly high proportion of a generator's running time.
 \iflater
@@
 To address this question, we outline two key principles for designing
-generator libraries that marry efficiency with flexibility.
+generator libraries that balance efficiency with flexibility.\bcp{We
+  just said we want both, not a ``balance'' that trades one off
+  against the other...}
 First, we describe a method of optimizing generator libraries using
@@
 Secondly, we demonstrate the impact of different choices of \rand on generator
-performance by conducting a controlled comparison, showing that showing that certain 
-implementations significantly reduce overhead.
-Together, these insights highlight practical strategies for improving generator libraries, 
-combining the efficiency gains of staging with a careful evaluation of \rand to optimize performance.
+performance by conducting a controlled comparison.
+Together, we refer to this recipe\bcp{``demonstrating the impact'' is
+  not a recipe} for improving generator libraries as \name{} --- \jwc{something pithy here}.
 
@@
 show an even more dramatic performance boost.
-
 Because the inlined versions of the generator are identical to the un-inlined
@@
 this performance difference is attributable solely to
-the different machine code generated by the compiler.
-Indeed, compilers of effectful and strict functional languages (including JIT
-compilers, in the case of Scala 3)
-use heuristics to determine if and when to perform this particular kind of
-simplification~\footnote{As we discuss in Section~\ref{section:other-langs}, purity
+the different machine code generated by the compiler (and not lucky choice of random seeds).
+
+It might be surprising to readers that this dramatic overhead exists --- shouldn't the compiler perform
+this simple optimization? Compilers of effectful and strict functional languages (including JIT
+compilers in the case of Scala 3)
+do in fact have heuristics to determine if and when to perform this particular kind of
+simplification\footnote{As we discuss in Section~\ref{section:other-langs}, purity
 means that Haskell is a slightly different story. GHC can and often does
-transformations of this form.}.  Even in cases as simple as
+transformations of this form.}. However even in cases as simple as
 Figure~\ref{fig:simplify}, the indirection of \texttt{return}
@@
 conservative for optimizing recursive functions.
+\jwc{Rewrite this para, BCP what do you think now?}
 
@@
 
-While neither of these microbenchmarks exhibit this behavior, more complex
+More complex
 generators suffer further performance penalties due to \emph{closure
 allocation}. In cases where the compiler cannot statically eliminate it,
-running a monadic bind allocates a short-lived closure: we allocate a closure
-for the continuation, and then immediately jump into it.
+running a monadic bind allocates a short-lived closure
+for the continuation, and then immediately jumps into it.
 In strict functional languages like OCaml and Scala,
@@
 
-This section presents our experimental setup, as well as answers to the following research questions:
+This section presents our experimental setup, and answers to the following research questions:
 
@@
 For evaluating \scalaname, we implement a subset of the aformentioned 
-strategies---the boolean list generator, single-pass BST generator, and 
+strategies---only the boolean list generator, single-pass BST generator, and 
 well-typed STLC term generator.

### 2025-03-21 paper/paper.tex
more wibbles

@@
   \end{tabular}
-\caption{Microbenchmarks of Generators in \bq{} (using \texttt{core\_bench} \cn) and ScalaCheck (using \texttt{jmh} \cn). Average over 10000 generations with random seeds, and a fixed size}
+\caption{Microbenchmarks of Generators in \bq{} and ScalaCheck. Average over 10000 generations with random seeds and a fixed size.}
 \label{fig:overhead-explanation-perf}
@@
 running a monadic bind allocates a short-lived closure
-for the continuation, and then immediately jumps into it.
+for the continuation and then immediately jumps into it.
 In strict functional languages like OCaml and Scala,
-Each individual closure allocation is relatively cheap, but doing lots of allocation in a
+each individual closure allocation is relatively cheap; but doing lots of allocation in a
 generator is very costly because each allocation brings us closer to the next costly GC pause \cn{}.
@@
 through the recursive loop re-allocates closures for \texttt{bind}s, so the amount of allocation
-\emph{per generated value} scales linearly with the number of recursive generator calls.
-
-\subsubsection{Overhead of Choice Combinators}
+per generated value scales linearly with the number of recursive generator calls.
+
+\paragraph{Overhead of Choice Combinators}
 Like \texttt{return} and \texttt{bind}, combinators like \texttt{weighted\_union} incur a performance penalty
@@
 
-In practice, \texttt{weighted\_union} is almost always\footnote{
+In practice, \texttt{weighted\_union} is (almost \footnote{
 There are some generators where \texttt{weighted\_union} is passed a list which
 was itself the result of a generator: the well-typed STLC term generator used in Section~\ref{section:eval} is an example of this.
-} called with an explicitly constructed list, like
+}) always called with an explicitly constructed list, like
 \texttt{weighted\_union [(w1,g1); (w2,g2); (w3,g3)]}. 
@@
 the options for which are always known.
-This list is allocated at the call site and then never needed after the call to \texttt{weighted\_union} returns\footnote{
-This overhead is fixable in languages with stack alloctation, which OCaml will soon be \cn{}.
-}.
+This list is allocated at the call site and then never needed after the call to \texttt{weighted\_union} returns.
 Since the elements of the list and its length $n$ are known,

### 2025-03-21 paper/paper.tex

--- 2025-03-21 paper/paper.tex
@@
 
-While neither of these microbenchmarks exhibit this behavior, more complex
+More complex
 generators suffer further performance penalties due to \emph{closure
 allocation}. In cases where the compiler cannot statically eliminate it,
-running a monadic bind allocates a short-lived closure: we allocate a closure
-for the continuation, and then immediately jump into it.
+running a monadic bind allocates a short-lived closure
+for the continuation and then immediately jumps into it.
 In strict functional languages like OCaml and Scala,
-Each individual closure allocation is relatively cheap, but doing lots of allocation in a
+each individual closure allocation is relatively cheap; but doing lots of allocation in a
 generator is very costly because each allocation brings us closer to the next costly GC pause \cn{}.

### 2025-03-21 paper/paper.tex
done through 2.1

@@
 was itself the result of a generator: the well-typed STLC term generator used in Section~\ref{section:eval} is an example of this.
-}) always called with an explicitly constructed list, like
+}) always called with an explicitly constructed list, as in
 \texttt{weighted\_union [(w1,g1); (w2,g2); (w3,g3)]}. 
@@
 This list is allocated at the call site and then never needed after the call to \texttt{weighted\_union} returns.
-Since the elements of the list and its length $n$ are known,
-a compiler could in principle unroll the loops in \texttt{weighted\_union} to depth $n$ and specialize the function at each call site to avoid allocating the list.
-Unfortunately, almost~\footnote{Once again, GHC is a notable exception here, performing sophisticated list fusion optimizations \cn{}} no compilers perform this kind of optimization.
-This allocation (or rather, its tendency to cause GC pauses) --- as well as the cost of running the code 
-to traverse arbitrary lists compared to unrolled loops --- has a significant impact on 
+Since the elements of the list and its length \texttt{n} are known,
+a compiler could in principle unroll the loops in \texttt{weighted\_union} to depth \texttt{n} and specialize the function at each call site to avoid allocating the list.
+Unfortunately, (almost\footnote{GHC is a notable exception here, performing sophisticated list fusion optimizations \cn{}}) no compilers perform this kind of optimization.
+This allocation (or rather, its tendency to cause GC pauses)---as well as the cost of running the code 
+to traverse arbitrary lists compared to unrolled loops---has a significant impact on 
 performance.
@@
 Last, many PBT libraries --- including both \bq and ScalaCheck --- implement their \texttt{weighted\_union} combinators
-in ways that are asymptotically more efficient but slower in practice than the naive algorithm.
-Weighted union works by the Fitness Proportionate Selection algorithm \cn{}, which (1) samples a number $r$
+in ways that are asymptotically more efficient but slower in common cases than a more naive algorithm.
+Weighted union uses the Fitness Proportionate Selection algorithm \cn{}, which (1) samples a number $r$
 between $0$ and the sum of the weights, and then (2) finds the first generator for which the cumulative
 weights in the list before it exceeds $r$. This second part can be accomplished in $O(\log_2 n)$ time by
-binary searching the list. However, since the lists are short in practice, a linear scan is almost always faster \cn{}.
+a binary search. However, since the lists are short in practice, a linear scan is almost always faster \cn{}.
 Moreover, both \bq{} and ScalaCheck allocate auxiliary data structures (an array in \bq{} and a BST in ScalaCheck) to
 perform this search, which incurs further run time overhead.
+\bcp{Are we advocating a linear scan? If not, then maybe we don't need to make this point}\jwc{We are, but we do need to make this point, since
+it's important to understanding why we get more speedup than you might expect}
 

### 2025-03-21 paper/paper.tex
through S3

@@
 \subsection{Inefficient Randomness Libraries}
-The core of any PBT generator library is a source of randomness: to generate
-random values of some datatype, we some access to random numbers!
+The core of any PBT generator library is a source of randomness.
 Different PBT libraries use different randomness libraries implementing different algorithms
@@
 Following the original Haskell QuickCheck implementation, \bq\ uses the SplitMix
-algorithm \cn, implemented as a (stateful) OCaml library called \texttt{Splittable\_random}. Meanwhile,
+algorithm \cn, implemented by \texttt{Splittable\_random}. Meanwhile,
 ScalaCheck uses the JSF algorithm \cn.
@@
 
-The randomness library sits at the heart of the hot path.
+The randomness library is the hottest part of the hot path.
 Even basic generators --- like ones generating a single \texttt{int} or \texttt{float} uniformly within a range --- can sample \emph{unboundedly many}
-random numbers, since they usually use versions of rejection sampling to find a value within the range \cn.
-Moreover, generator combinators like \texttt{list} usually make $O(n)$ calls to even those basic generators.
+random numbers, since they usually use versions of rejection sampling to find a value within the range \cn{}\bcp{Really?
+I want to know more about that!!}\jwc{Well this is totally standard, and it's how sampling in a bounded range works.
+You take a thing that samples uniformly over \texttt{int64\_t} and do a combination of clever things and basic rejection sampling to get a value in the range.}.
+Moreover, generator combinators like \texttt{list} usually make $O(n)$ calls to the \texttt{int} or \texttt{float} generators.
 Because of this, the speed of a single sample matters a great deal. Unfortunately,
@@
 For example, significantly faster algorithms than SplitMix or JSF exist,
-such as the Lehmer algorithm\cn. In microbenchmarks, the Lehmer algorithm runs almost 2x as fast as SplitMix \cn.
+for example the Lehmer algorithm\cn. In microbenchmarks, the Lehmer algorithm runs almost 2x as fast as SplitMix \cn.
 % https://github.com/lemire/testingRNG/tree/master
@@
 \jwc{The point here is that ``you should be thinking about using a more performant RNG, we'll show you that it does have real benefits (and quantify them). But there are tradeoffs here.'}
+\bcp{But this is not the point that the reader is set up to expect: we've toldthem that we're going to make generation faster, not that we're going to give them tips so that htey can make it faster by themselves}
+\jwc{This actually isn't the case. In the intro, we're very careful to say that we show that choosing a faster randomness library \emph{leads to faster bugfinding}. And we do demonstrate this.}
 
 Of course, simply arguing that the randomness library is on the hot
-path does not guarantee that a faster sampling leads to measurably faster
+path for PBT does not guarantee that a faster sampling leads to measurably faster
 generation; for that, we need an experiment. The most obvious experiment is to

### 2025-03-21 paper/paper.tex
consistent-ify the name usage, fixing joe mistakes

@@
 Then, in Section~\ref{section:faster-generators}, we explain multi-stage programming,
-and develop both facets of \name{}, using \camlname{} to illustrate.
+develop \name{} (using \camlname{} to illustrate), and discuss how we isolate the performance of \rand{} to benchmark their impact on bug-finding speed.
 Section~\ref{section:eval} evaluates all of the previous work, demonstrating the generator performance improvements
@@
 \name generators across a range of benchmarks. Our experiments compare generators 
-built using our technique---with and without \csplitmix---against those implemented with 
+built using our technique---with and without \csplitmix \jwc{this is the first mention of csplitmix!}---against those implemented with 
 existing generator libraries (and their default randomness mechanisms). Specifically, 
@@
 
-We begin by noting that the \name{} principle of choosing fast random number generators
+We begin by noting that the principle of choosing fast random number generators
 is completely language-agnostic. In languages like OCaml, where runtime value representations
@@
 
-Next, we discuss language-by-language the degree to which the \name{} insights about staging could be used.
+Next, we discuss language-by-language the degree to which the staging-related insights of this paper are portable.
 \begin{itemize}

### 2025-03-21 paper/paper.tex
fix words

@@
 many common generator abstractions.
-Second, we characterize the impact of \rand choice through a natural experiment 
+Second, we characterize the impact of \rand choice through a controlled experiment 
 that isolates the performance contribution of the \rand within a generator, underscoring its role as a first-class design consideration.
@@
 generation; for that, we need an experiment. The most obvious experiment is to
-simply swap out the randomness library of either \bq{} or ScalaCheck with a totally different, faster one. But, as
-discussed previously, generators with different generation orders can be
-difficult to compare. To get around this, we exploit a ``natural experiment:
+simply swap out the randomness library of either \bq{} or ScalaCheck with a totally different, faster one.
+But, as discussed previously, \camlname{} and \scalaname{} are designed to be 100\% semantically equivalent replacements for
+their unstaged counterparts, ensuring that any bug-finding speedups are solely attributable to generator performance and not lucky seeds.
+Choosing a different \rand{} would break this property, making it much more challenging to assess the performance compared to the baseline.
+But, as discussed previously, generators with different generation orders can be
+difficult to compare \bcp{This was not discussed previously!}. To get around this, we exploit a coincidence:
 the \rand{} that \bq{} uses --- OCaml's \texttt{Splittable\_random} library --- is slow in a way that
@@
 To demonstrate that faster random sampling can significanty impact bugfinding power, we
-use the natural experiment provided by OCaml's inefficient implementation of
+use the controlled experiment provided by OCaml's inefficient implementation of
 SplitMix. By replacing this slow randomness library with a faster but semantically equivalent

### 2025-03-21 paper/paper.tex
Updates to the updates

--- 2025-03-21 paper/paper.tex
@@
 
-Property-based testing (PBT) is a software testing technique that uses
+Property-based testing (PBT) is a software-testing technique that uses
 random test inputs to validate logical specifications. Recent studies on PBT usage~\cite{inpractice} in industry have shown that
@@
 window to find bugs---e.g., between 50 milliseconds and 30 seconds~\cite{inpractice}.
-The faster a test fails, the better.
+The faster it finds a failing test, the better.
 % The goal of PBT is to fail tests as fast as possible. \jwc{First cut at ``fail fast''}
@@
 libraries}, usually expressed as embedded domain-specific languages (eDSLs) %\cntxt{edsls}
-that provide combinators for building and composing generators. However, careful measurements show that the performance of existing generator libraries falls far short of what
-is possible.
+that provide combinators for building and composing
+generators. However, careful measurements show that the performance of
+existing generator libraries---even well-engineered and widely used
+ones---falls far short of what is possible.
 
@@
 Second, the \rand that a generator uses to make choices can
-constitute an unexpectedly high proportion of its runtime. Combined, these factors limit 
+be responsible for a surprisingly high proportion of its runtime. Combined, these factors limit
 the effectiveness of PBT.
@@
 
-We lay a path forward.
+We address both sources of inefficiency.
 First, we show how to use \textit{multi-stage programming}---a classic technique for
@@
 Second, we characterize the impact of \rand choice through a natural experiment
-that isolates the performance contribution of the \rand within a generator, underscoring its role as a first-class design consideration.
+that isolates the performance contribution of the \rand within a
+generator, underscoring its role as a first-order design
+consideration.\bcp{We've done more than this, though: We've also
+  demonstrated that you can get an X\% improvement in performance by
+  using a specific RNG.}
 Taken together, these insights can be applied to create tomorrow's generator libraries: those that
@@
 
-To demonstrate the effectiveness and generality of our approach, which we call \name, we implement staged PBT libraries in multiple languages.
+To demonstrate the effectiveness and generality of our approach\bcp{It
+feels a bit grand to call what we've done an ``approach''---it's
+really just a single performance-enhancing technique, staging, plus
+the observation that the RNG matters and you should use the fastest
+one you can.}, which we call \name, we implement staged PBT libraries in multiple languages.
 In OCaml, we build \camlname{}, based on \bq{}---a
@@
   in OCaml...} in Scala \cn{}.
-In both languages, our libraries are direct drop-in replacements that are
+In both languages, our libraries are direct drop-in
+replacements\bcp{that's not quite accurate, is it? I mean, they have
+  the same functionality, but can you take a generator that works with
+base\_quickcheck and just link it against our version without changing
+the generator itself? That's what ``drop-in replacement'' means to me.} that are
 \textit{100\% semantically equivalent}: that is, given identical random seeds, a generator written with our technique
@@
 Figure~\ref{fig:bq-api}.  The \texttt{weighted\_union} function is a
-particularly well used one: it makes making a weighted choice between different
+particularly well-used one: it makes making a weighted choice between different
 generators, allowing the developer to combine different sub-generators into a
@@
 show an even more dramatic performance boost.
+
 Because the inlined versions of the generator are identical to the un-inlined
@@
 this performance difference is attributable solely to
-the different machine code generated by the compiler (and not lucky choice of random seeds).
-
-It might be surprising to readers that this dramatic overhead exists --- shouldn't the compiler perform
-this simple optimization? Compilers of effectful and strict functional languages (including JIT
-compilers in the case of Scala 3)
-do in fact have heuristics to determine if and when to perform this particular kind of
-simplification\footnote{As we discuss in Section~\ref{section:other-langs}, purity
+the different machine code generated by the compiler.
+Indeed, compilers of effectful and strict functional languages (including JIT
+compilers, in the case of Scala 3)
+use heuristics to determine if and when to perform this particular kind of
+simplification~\footnote{As we discuss in Section~\ref{section:other-langs}, purity
 means that Haskell is a slightly different story. GHC can and often does
-transformations of this form.}. However even in cases as simple as
+transformations of this form.}.  Even in cases as simple as
 Figure~\ref{fig:simplify}, the indirection of \texttt{return}
@@
 conservative for optimizing recursive functions.
-\jwc{Rewrite this para, BCP what do you think now?}
 
@@
   \end{tabular}
-\caption{Microbenchmarks of Generators in \bq{} and ScalaCheck. Average over 10000 generations with random seeds and a fixed size.}
+\caption{Microbenchmarks of Generators in \bq{} (using \texttt{core\_bench} \cn) and ScalaCheck (using \texttt{jmh} \cn). Average over 10000 generations with random seeds, and a fixed size}
 \label{fig:overhead-explanation-perf}
@@
 
-More complex
+While neither of these microbenchmarks exhibit this behavior, more complex
 generators suffer further performance penalties due to \emph{closure
 allocation}. In cases where the compiler cannot statically eliminate it,
-running a monadic bind allocates a short-lived closure
-for the continuation and then immediately jumps into it.
+running a monadic bind allocates a short-lived closure: we allocate a closure
+for the continuation, and then immediately jump into it.
 In strict functional languages like OCaml and Scala,
-each individual closure allocation is relatively cheap; but doing lots of allocation in a
+Each individual closure allocation is relatively cheap, but doing lots of allocation in a
 generator is very costly because each allocation brings us closer to the next costly GC pause \cn{}.
@@
 through the recursive loop re-allocates closures for \texttt{bind}s, so the amount of allocation
-per generated value scales linearly with the number of recursive generator calls.
-
-\paragraph{Overhead of Choice Combinators}
+\emph{per generated value} scales linearly with the number of recursive generator calls.
+
+\subsubsection{Overhead of Choice Combinators}
 Like \texttt{return} and \texttt{bind}, combinators like \texttt{weighted\_union} incur a performance penalty
@@
 
-In practice, \texttt{weighted\_union} is (almost \footnote{
+In practice, \texttt{weighted\_union} is almost always\footnote{
 There are some generators where \texttt{weighted\_union} is passed a list which
 was itself the result of a generator: the well-typed STLC term generator used in Section~\ref{section:eval} is an example of this.
-}) always called with an explicitly constructed list, as in
+} called with an explicitly constructed list, like
 \texttt{weighted\_union [(w1,g1); (w2,g2); (w3,g3)]}.
@@
 the options for which are always known.
-This list is allocated at the call site and then never needed after the call to \texttt{weighted\_union} returns.
-Since the elements of the list and its length \texttt{n} are known,
-a compiler could in principle unroll the loops in \texttt{weighted\_union} to depth \texttt{n} and specialize the function at each call site to avoid allocating the list.
-Unfortunately, (almost\footnote{GHC is a notable exception here, performing sophisticated list fusion optimizations \cn{}}) no compilers perform this kind of optimization.
-This allocation (or rather, its tendency to cause GC pauses)---as well as the cost of running the code 
-to traverse arbitrary lists compared to unrolled loops---has a significant impact on 
+This list is allocated at the call site and then never needed after the call to \texttt{weighted\_union} returns\footnote{
+This overhead is fixable in languages with stack alloctation, which OCaml will soon be \cn{}.
+}.
+Since the elements of the list and its length $n$ are known,
+a compiler could in principle unroll the loops in \texttt{weighted\_union} to depth $n$ and specialize the function at each call site to avoid allocating the list.
+Unfortunately, almost~\footnote{Once again, GHC is a notable exception here, performing sophisticated list fusion optimizations \cn{}} no compilers perform this kind of optimization.
+This allocation (or rather, its tendency to cause GC pauses) --- as well as the cost of running the code
+to traverse arbitrary lists compared to unrolled loops --- has a significant impact on
 performance.
@@
 Last, many PBT libraries --- including both \bq and ScalaCheck --- implement their \texttt{weighted\_union} combinators
-in ways that are asymptotically more efficient but slower in common cases than a more naive algorithm.
-Weighted union uses the Fitness Proportionate Selection algorithm \cn{}, which (1) samples a number $r$
+in ways that are asymptotically more efficient but slower in practice than the naive algorithm.
+Weighted union works by the Fitness Proportionate Selection algorithm \cn{}, which (1) samples a number $r$
 between $0$ and the sum of the weights, and then (2) finds the first generator for which the cumulative
 weights in the list before it exceeds $r$. This second part can be accomplished in $O(\log_2 n)$ time by
-a binary search. However, since the lists are short in practice, a linear scan is almost always faster \cn{}.
+binary searching the list. However, since the lists are short in practice, a linear scan is almost always faster \cn{}.
 Moreover, both \bq{} and ScalaCheck allocate auxiliary data structures (an array in \bq{} and a BST in ScalaCheck) to
 perform this search, which incurs further run time overhead.
-\bcp{Are we advocating a linear scan? If not, then maybe we don't need to make this point}\jwc{We are, but we do need to make this point, since
-it's important to understanding why we get more speedup than you might expect}
 
@@
 \subsection{Inefficient Randomness Libraries}
-The core of any PBT generator library is a source of randomness.
+The core of any PBT generator library is a source of randomness: to generate
+random values of some datatype, we some access to random numbers!
 Different PBT libraries use different randomness libraries implementing different algorithms
@@
 Following the original Haskell QuickCheck implementation, \bq\ uses the SplitMix
-algorithm \cn, implemented by \texttt{Splittable\_random}. Meanwhile,
+algorithm \cn, implemented as a (stateful) OCaml library called \texttt{Splittable\_random}. Meanwhile,
 ScalaCheck uses the JSF algorithm \cn.
@@
 
-The randomness library is the hottest part of the hot path.
+The randomness library sits at the heart of the hot path.
 Even basic generators --- like ones generating a single \texttt{int} or \texttt{float} uniformly within a range --- can sample \emph{unboundedly many}
-random numbers, since they usually use versions of rejection sampling to find a value within the range \cn{}\bcp{Really?
-I want to know more about that!!}\jwc{Well this is totally standard, and it's how sampling in a bounded range works.
-You take a thing that samples uniformly over \texttt{int64\_t} and do a combination of clever things and basic rejection sampling to get a value in the range.}.
-Moreover, generator combinators like \texttt{list} usually make $O(n)$ calls to the \texttt{int} or \texttt{float} generators.
+random numbers, since they usually use versions of rejection sampling to find a value within the range \cn.
+Moreover, generator combinators like \texttt{list} usually make $O(n)$ calls to even those basic generators.
 Because of this, the speed of a single sample matters a great deal. Unfortunately,
@@
 For example, significantly faster algorithms than SplitMix or JSF exist,
-for example the Lehmer algorithm\cn. In microbenchmarks, the Lehmer algorithm runs almost 2x as fast as SplitMix \cn.
+such as the Lehmer algorithm\cn. In microbenchmarks, the Lehmer algorithm runs almost 2x as fast as SplitMix \cn.
 % https://github.com/lemire/testingRNG/tree/master
@@
 \jwc{The point here is that ``you should be thinking about using a more performant RNG, we'll show you that it does have real benefits (and quantify them). But there are tradeoffs here.'}
-\bcp{But this is not the point that the reader is set up to expect: we've toldthem that we're going to make generation faster, not that we're going to give them tips so that htey can make it faster by themselves}
-\jwc{This actually isn't the case. In the intro, we're very careful to say that we show that choosing a faster randomness library \emph{leads to faster bugfinding}. And we do demonstrate this.}
 
 Of course, simply arguing that the randomness library is on the hot
-path for PBT does not guarantee that a faster sampling leads to measurably faster
+path does not guarantee that a faster sampling leads to measurably faster
 generation; for that, we need an experiment. The most obvious experiment is to

### 2025-03-21 paper/paper.tex
wibble

--- 2025-03-21 paper/paper.tex
@@
 Second, we characterize the impact of \rand choice through a controlled experiment 
-that isolates the performance contribution of the \rand within a generator, underscoring its role as a first-class design consideration.
+that isolates the performance contribution of the \rand within a generator, underscoring its role as a first-class design consideration.\bcp{We've done more than this, though: We've also
+  demonstrated that you can get an X\% improvement in performance by
+  using a specific RNG.}
 Taken together, these insights can be applied to create tomorrow's generator libraries: those that 
@@
 in OCaml and ScalaCheck in Scala, allowing us to assess the effectiveness of our
-technique across different languages and runtime environments.
+technique across different languages and runtime
+environments.\bcp{This says ``assess the impact'' twice...}
 

### 2025-03-21 staged-scala/src/main/scala/BST/SingleBespoke.scala
oops, fixed bug in bespoke bst

@@ -11,12 +11,15 @@ object SingleBespoke {
       if (lo >= hi || sz <= 1) {
         Gen.const(Bst.E)
       } else {
-        for {
-        k <- Gen.Choose.chooseLong.choose(lo, hi)
-        v <- Nat.generator(bst_bespoke_limits)
-        l <- Gen.resize(sz / 2, go(lo, k - 1))
-        r <- Gen.resize (sz / 2, go(k + 1, hi))
-        } yield (Bst.Node (l,k,v,r))
+        Gen.frequency(
+          1 -> Gen.const(Bst.E),
+          sz -> (for {
+            k <- Gen.Choose.chooseLong.choose(lo, hi)
+            v <- Nat.generator(bst_bespoke_limits)
+            l <- Gen.resize(sz / 2, go(lo, k - 1))
+            r <- Gen.resize (sz / 2, go(k + 1, hi))
+            } yield (Bst.Node (l,k,v,r))
+        ))
       }
     )
   }

### 2025-03-21 paper/paper.tex
some words in intro from joe/bcp convo

@@
   possible...'' }
+\jwc{Agreed.
+Maybe one way to do this is to say someting in the next pagraph like ``much work has been poured building DSLs that let programers cleverly generate inputs to maximize the potential of finding bugs.
+Comparably less work has been poured into just making them fast.''}
 \fi
@@
     that these performance improvements extend to significantly improved bug-finding speed. \jwc{This should be sell-ier: 2x speedups, etc. ``bug-finding speed'' isn't a race against time.}
+    \jwc{We precisely quantify the performance effects of staging and faster rng across a range of testing scenarios --- i.e.
+a contrib is the fact that we can exactly pin down when one helps versus the other.
+The ``logically smarter'' your generator is, the more you get out of staging, the ``dumber'', the more you get out of rng.}
+    \bcp{Agree with this. We need to beef up our claims here about what we learned about PBT-specifics.}
+    \jwc{And this needs to be flowed through the rest of the paper.}
 \end{enumerate}

### 2025-03-21 paper/paper.tex
more stuff from convo
intro got clobbered

@@
 %\jwc{``Property based testing is a race against time...''}
-
-Property-based testing (PBT) is a software-testing technique that uses
+Property-based testing (PBT) is a software testing technique that uses
 random test inputs to validate logical specifications. Recent studies on PBT usage~\cite{inpractice} in industry have shown that
@@
 window to find bugs---e.g., between 50 milliseconds and 30 seconds~\cite{inpractice}.
-The faster it finds a failing test, the better.
+The faster a test fails, the better.
 % The goal of PBT is to fail tests as fast as possible. \jwc{First cut at ``fail fast''}
@@
   possible...'' }
-\jwc{Agreed.
-Maybe one way to do this is to say someting in the next pagraph like ``much work has been poured building DSLs that let programers cleverly generate inputs to maximize the potential of finding bugs.
-Comparably less work has been poured into just making them fast.''}
 \fi
@@
 libraries}, usually expressed as embedded domain-specific languages (eDSLs) %\cntxt{edsls}
-that provide combinators for building and composing
-generators. However, careful measurements show that the performance of
-existing generator libraries---even well-engineered and widely used
-ones---falls far short of what is possible.
+that provide combinators for building and composing generators. However, careful measurements show that the performance of existing generator libraries falls far short of what
+is possible.
 
@@
 Second, the \rand that a generator uses to make choices can
-be responsible for a surprisingly high proportion of its runtime. Combined, these factors limit
+constitute an unexpectedly high proportion of its runtime. Combined, these factors limit 
 the effectiveness of PBT.
@@
 
-We address both sources of inefficiency.
+We lay a path forward.
 First, we show how to use \textit{multi-stage programming}---a classic technique for 
@@
 many common generator abstractions.
-Second, we characterize the impact of \rand choice through a controlled experiment 
-that isolates the performance contribution of the \rand within a generator, underscoring its role as a first-class design consideration.\bcp{We've done more than this, though: We've also
-  demonstrated that you can get an X\% improvement in performance by
-  using a specific RNG.}
+Second, we characterize the impact of \rand choice through a natural experiment 
+that isolates the performance contribution of the \rand within a generator, underscoring its role as a first-class design consideration.
 Taken together, these insights can be applied to create tomorrow's generator libraries: those that 
@@
 solely attributable to the staging and \rand optimizations.
+\jwc{BCP thinks this should be a little stronger: ``and not obscured by the probablistic nature of generation'', or something like that.
+``Comparing distributions of speed is hard, comparing pointwise is easier.''.
+This needs to have its own sentence here.}
 

### 2025-03-21 paper/paper.tex
put back

@@
 libraries}, usually expressed as embedded domain-specific languages (eDSLs) %\cntxt{edsls}
-that provide combinators for building and composing generators. However, careful measurements show that the performance of existing generator libraries falls far short of what
+that provide combinators for building and composing generators. However, careful 
+measurements show that the performance of existing generator libraries falls far short of what
 is possible.
@@
 Second, the \rand that a generator uses to make choices can
-constitute an unexpectedly high proportion of its runtime. Combined, these factors limit 
+constitute an unexpectedly high proportion of its runtime. These factors combine to limit 
 the effectiveness of PBT.
@@
 many common generator abstractions.
-Second, we characterize the impact of \rand choice through a natural experiment 
-that isolates the performance contribution of the \rand within a generator, underscoring its role as a first-class design consideration.
-Taken together, these insights can be applied to create tomorrow's generator libraries: those that 
-combine high-level design with low-level performance.
-
-To demonstrate the effectiveness and generality of our approach\bcp{It
-feels a bit grand to call what we've done an ``approach''---it's
-really just a single performance-enhancing technique, staging, plus
-the observation that the RNG matters and you should use the fastest
-one you can.}, which we call \name, we implement staged PBT libraries in multiple languages.
+Second, we show that entropy source is an undervalued consideration in generator library design
+by isolating the performance contribution of the \rand in generation time, and providing an alternative
+that behaves equivalently with $5.5\times$ speedups.
+These insights culminate in \name, a recipe for building generator libraries that marry high-level design
+with low-level performance.
+
+% \bcp{It
+% feels a bit grand to call what we've done an ``approach''---it's
+% really just a single performance-enhancing technique, staging, plus
+% the observation that the RNG matters and you should use the fastest
+% one you can.}
+
+To evaluate our technique, we implement staged PBT libraries in two strict functional languages: OCaml and Scala 3.
 In OCaml, we build \camlname{}, based on \bq{}---a

### 2025-03-21 paper/paper.tex
blah

@@
 many common generator abstractions.
-Second, we show that entropy source is an undervalued consideration in generator library design
-by isolating the performance contribution of the \rand in generation time, and providing an alternative
-that behaves equivalently with $5.5\times$ speedups.
+Second, we demostrate that entropy source is an undervalued consideration in generator library design
+by isolating its performance contribution, and showing that significant
+speedups are possible.
 These insights culminate in \name, a recipe for building generator libraries that marry high-level design

### 2025-03-21 paper/paper.tex
blah

@@
 \section{What are Generator Libraries, and why are they Slow?}
+\tr{This needs to introduce PBT WAY better.}
 \label{section:motiv}
@@
 The runtime of \camlname generators across different input sizes is summarized in \figref{fig:time}.
-Our \camlname type-derived BST generator achieves speedups ranging from 1.08--1.29x, which increase
-dramatically to 6.13--9.65x when combined with \csplitmix. The insertion-based BST generator sees
-1.13--1.21x speedups with staging alone, and 4.32--5.36x when also
+Our \camlname type-derived BST generator achieves speedups ranging from $1.08-1.29\times$, which increase
+dramatically to $6.13-9.65\times$ when combined with \csplitmix. The insertion-based BST generator sees
+$1.13-1.21\times$ speedups with staging alone, and $4.32-5.36\times$ when also
 using \csplitmix. The single-pass BST generator benefits more significantly from
-staging, with speedups of 2.31--2.34x, which rise to 8.61--8.95x when combined with
+staging, with speedups of $2.31-2.34\times$, which rise to $8.61-8.95\times$ when combined with
 \csplitmix. We attribute the differential in performance speedup to the fact that
@@
 For STLC, staging accounts for the bulk of performance gains, yielding
- 1.71--2.64x speedups for the type-derived STLC generator, and 4.60--5.46x for
+ $1.71-2.64\times$ speedups for the type-derived STLC generator, and $4.60-5.46\times$ for
  the well-typed generator. When adding \csplitmix, these numbers increase to
- 2.60--5.15x and 4.83--7.75x, respectively. This result makes sense because
+ $2.60-5.15\times$ and $4.83-7.75\times$, respectively. This result makes sense because
  STLC generators, particularly the well-typed generator, sample very little from
@@
 
-The boolean list generator experiences 3.61--7.10x speedups with
+The boolean list generator experiences $3.61-7.10\times$ speedups with
 staging, and its performance actually slightly \textit{decreases} when
-combined with \csplitmix (3.43--6.83x). This is likely because sampling
+combined with \csplitmix ($3.43-6.83\times$). This is likely because sampling
 booleans is ``cheap'' enough that the overhead of crossing the FFI barrier is

### 2025-03-21 staged-scala/src/main/scala/benchmark/GenBm.scala
push to run this on the metal

@@ -117,6 +117,88 @@ val size50 = Gen.Parameters.default.withSize(50)
 val size100 = Gen.Parameters.default.withSize(100)
 val size1000 = Gen.Parameters.default.withSize(1000)
 
+// Bool List Bespoke - unstaged benchmarks
+@Benchmark
+@BenchmarkMode(Array(Mode.AverageTime))
+@OutputTimeUnit(TimeUnit.NANOSECONDS)
+@Warmup(iterations = 3, time = 5, timeUnit = TimeUnit.SECONDS)
+@Measurement(iterations = 5, time = 5, timeUnit = TimeUnit.SECONDS)
+@Fork(1)
+  def generateBoolListBespoke10() = {
+    BoolList.Bespoke.gen.apply(size10,Seed.random())
+  }
+
+@Benchmark
+@BenchmarkMode(Array(Mode.AverageTime))
+@OutputTimeUnit(TimeUnit.NANOSECONDS)
+@Warmup(iterations = 3, time = 5, timeUnit = TimeUnit.SECONDS)
+@Measurement(iterations = 5, time = 5, timeUnit = TimeUnit.SECONDS)
+@Fork(1)
+  def generateBoolListBespoke50() = {
+    BoolList.Bespoke.gen.apply(size50,Seed.random())
+  }
+
+@Benchmark
+@BenchmarkMode(Array(Mode.AverageTime))
+@OutputTimeUnit(TimeUnit.NANOSECONDS)
+@Warmup(iterations = 3, time = 5, timeUnit = TimeUnit.SECONDS)
+@Measurement(iterations = 5, time = 5, timeUnit = TimeUnit.SECONDS)
+@Fork(1)
+  def generateBoolListBespoke100() = {
+    BoolList.Bespoke.gen.apply(size100,Seed.random())
+  }
+
+@Benchmark
+@BenchmarkMode(Array(Mode.AverageTime))
+@OutputTimeUnit(TimeUnit.NANOSECONDS)
+@Warmup(iterations = 3, time = 5, timeUnit = TimeUnit.SECONDS)
+@Measurement(iterations = 5, time = 5, timeUnit = TimeUnit.SECONDS)
+@Fork(1)
+  def generateBoolListBespoke1000() = {
+    BoolList.Bespoke.gen.apply(size1000,Seed.random())
+  }
+
+// Bool List Bespoke - staged benchmarks
+@Benchmark
[…]

### 2025-03-21 paper/paper.tex

@@
 by isolating its performance contribution, and showing that significant
-speedups are possible.
+speedups are possible---even while maintaining output equivalence.
 These insights culminate in \name, a recipe for building generator libraries that marry high-level design
@@
 high-quality PBT library in OCaml, authored by software engineers at the trading firm Jane Street \cn{}.
-In Scala 3, we build \scalaname{}, based on ScalaCheck---a very
-mature PBT library and the {\em de facto} choice\bcp{the wording here
-  implies that there are other choices in Scala, some of which might
-  be reasonable, and tht base\_quickcheck is not the de facto choice
-  in OCaml...} in Scala \cn{}.
-In both languages, our libraries are direct drop-in
-replacements\bcp{that's not quite accurate, is it? I mean, they have
-  the same functionality, but can you take a generator that works with
-base\_quickcheck and just link it against our version without changing
-the generator itself? That's what ``drop-in replacement'' means to me.} that are
-\textit{100\% semantically equivalent}: that is, given identical random seeds, a generator written with our technique
-produces exactly the same sequence of values as the same generator written with the original generator library.
-The semantic equivalence between the original generators and \name{} generators allows us to
-perform clean, fine-grained comparisons, as differences in generation speed are
-solely attributable to the staging and \rand optimizations.
-\jwc{BCP thinks this should be a little stronger: ``and not obscured by the probablistic nature of generation'', or something like that.
+In Scala 3, we build \scalaname{}, based on ScalaCheck---the language's standard PBT library, and by far the most widely-used\cn{}.
+In both languages, our libraries provide identical functionality and behavior: given the same random seed, a generator written with our technique produces exactly the same sequence of values as an equivalent generator written with the original library.
+This semantic equivalence enables pointwise, rather than distributional, comparisons, ensuring that performance differences are attributable \textit{solely} to staging and \rand optimizations.
+% \jwc{BCP thinks this should be a little stronger: ``and not obscured by the probablistic nature of generation'', or something like that.
 ``Comparing distributions of speed is hard, comparing pointwise is easier.''.

### 2025-03-21 staged-scala/src/main/scala/benchmark/GenBm.scala
add another order of magnitude

@@ -116,6 +116,7 @@ val size10 = Gen.Parameters.default.withSize(10)
 val size50 = Gen.Parameters.default.withSize(50)
 val size100 = Gen.Parameters.default.withSize(100)
 val size1000 = Gen.Parameters.default.withSize(1000)
+val size10000 = Gen.Parameters.default.withSize(10000)  // Added new size parameter
 
 // Bool List Bespoke - unstaged benchmarks
 @Benchmark
@@ -156,6 +157,16 @@ val size1000 = Gen.Parameters.default.withSize(1000)
 @Fork(1)
   def generateBoolListBespoke1000() = {
     BoolList.Bespoke.gen.apply(size1000,Seed.random())
+  }
+
+@Benchmark
+@BenchmarkMode(Array(Mode.AverageTime))
+@OutputTimeUnit(TimeUnit.NANOSECONDS)
+@Warmup(iterations = 3, time = 5, timeUnit = TimeUnit.SECONDS)
+@Measurement(iterations = 5, time = 5, timeUnit = TimeUnit.SECONDS)
+@Fork(1)
+  def generateBoolListBespoke10000() = {
+    BoolList.Bespoke.gen.apply(size10000,Seed.random())
   }
 
 // Bool List Bespoke - staged benchmarks
@@ -199,6 +210,16 @@ val size1000 = Gen.Parameters.default.withSize(1000)
     BoolList.BespokeStaged.gen(1000)(Seed.random())
   }
 
+@Benchmark
+@BenchmarkMode(Array(Mode.AverageTime))
+@OutputTimeUnit(TimeUnit.NANOSECONDS)
+@Warmup(iterations = 3, time = 5, timeUnit = TimeUnit.SECONDS)
+@Measurement(iterations = 5, time = 5, timeUnit = TimeUnit.SECONDS)
+@Fork(1)
+  def generateBoolListBespokeStaged10000() = {
+    BoolList.BespokeStaged.gen(10000)(Seed.random())
+  }
+
 // BST Type Derived - unstaged benchmarks
 @Benchmark
 @BenchmarkMode(Array(Mode.AverageTime))
@@ -238,6 +259,16 @@ val size1000 = Gen.Parameters.default.withSize(1000)
 @Fork(1)
   def generateBstType1000(): Option[Bst] = {
[…]

### 2025-03-21 staged-scala/src/main/scala/benchmark/GenBm.scala
rm 50s

@@ -113,7 +113,6 @@ def chLng(l: Long, h: Long)(p: Gen.Parameters, seed: Seed): (Option[Long],Seed) 
 class GenBm {
 
 val size10 = Gen.Parameters.default.withSize(10)
-val size50 = Gen.Parameters.default.withSize(50)
 val size100 = Gen.Parameters.default.withSize(100)
 val size1000 = Gen.Parameters.default.withSize(1000)
 val size10000 = Gen.Parameters.default.withSize(10000)  // Added new size parameter
@@ -129,15 +128,6 @@ val size10000 = Gen.Parameters.default.withSize(10000)  // Added new size parame
     BoolList.Bespoke.gen.apply(size10,Seed.random())
   }
 
-@Benchmark
-@BenchmarkMode(Array(Mode.AverageTime))
-@OutputTimeUnit(TimeUnit.NANOSECONDS)
-@Warmup(iterations = 3, time = 5, timeUnit = TimeUnit.SECONDS)
-@Measurement(iterations = 5, time = 5, timeUnit = TimeUnit.SECONDS)
-@Fork(1)
-  def generateBoolListBespoke50() = {
-    BoolList.Bespoke.gen.apply(size50,Seed.random())
-  }
 
 @Benchmark
 @BenchmarkMode(Array(Mode.AverageTime))
@@ -178,16 +168,6 @@ val size10000 = Gen.Parameters.default.withSize(10000)  // Added new size parame
 @Fork(1)
   def generateBoolListBespokeStaged10() = {
     BoolList.BespokeStaged.gen(10)(Seed.random())
-  }
-
-@Benchmark
-@BenchmarkMode(Array(Mode.AverageTime))
-@OutputTimeUnit(TimeUnit.NANOSECONDS)
-@Warmup(iterations = 3, time = 5, timeUnit = TimeUnit.SECONDS)
-@Measurement(iterations = 5, time = 5, timeUnit = TimeUnit.SECONDS)
-@Fork(1)
-  def generateBoolListBespokeStaged50() = {
-    BoolList.BespokeStaged.gen(50)(Seed.random())
   }
 
 @Benchmark
@@ -237,16 +217,6 @@ val size10000 = Gen.Parameters.default.withSize(10000)  // Added new size parame
 @Warmup(iterations = 3, time = 5, timeUnit = TimeUnit.SECONDS)
[…]

### 2025-03-21 staged-scala/src/main/scala/BST/Limits.scala
oops, ranges too small were bottoming out trees

@@ -1,4 +1,4 @@
 package BST
 
-val bst_bespoke_limits = 10
-val bst_type_limits = 10
+val bst_bespoke_limits = Int.MaxValue
+val bst_type_limits = Int.MaxValue

### 2025-03-21 paper/paper.tex
PBT related work

@@
 \subsection{Speeding up Property-Based Testing}
-\hg{\begin{itemize}
-  \item Changing the generation order~\cite{runcimanSmallcheckLazySmallcheck2008,braquehaisSimpleIncrementalDevelopment2017}
-  \item Dealing with filtering~\cite{claessenGeneratingConstrainedRandom2015}
-  \item QuickerCheck~\cite{krookQuickerCheckImplementingEvaluating2024}
-  \item Evaluation of PBT generation~\cite{shiEtnaEvaluationPlatform2023}
-\end{itemize}}
+
+Our work is unique in its approach to speeding up PBT, but it is certainly not
+the only work focused on making PBT faster.
+
+Perhaps the best-explored approaches to speeding up bug-finding with PBT is by
+changing the order in which test inputs are sampled. Feedback-driven mechanisms
+like targeted PBT~\cite{loscherTargetedPropertybasedTesting2017} and
+coverage-guided PBT~\cite{lampropoulosCoverageGuidedProperty2019} have the
+potential to speed up testing by changing the generation order to try to find
+``interesting'' inputs faster. Separately, enumerative approaches to
+PBT~\cite{runcimanSmallcheckLazySmallcheck2008,braquehaisSimpleIncrementalDevelopment2017}
+try to get to bugs faster by leveraging the ``small-scope hypothesis'' and
+attempting to quickly exhaust the list of small inputs to the program under
+test. All of these approaches can have significant performance benefits in
+practice, but they are largely orthogonal to our contributions; we expect
+\name{} could be used to speed these existing techniques to varying degrees.
+
+Similarly orthogonal to our approach is techniques for quickly filtering out
+inputs that are invalid for testing. Some of these approaches work statically,
+by automatically deriving generators that produce valid inputs by
+construction~\cite{lampropoulosGeneratingGoodGenerators2017}, while others
+filter dynamically via laziness~\cite{claessenGeneratingConstrainedRandom2015}
+or by solving satisfiability
+problems~\cite{seidelTypeTargetedTesting2015,steinhofelInputInvariants2022,lampropoulosBeginnerLuckLanguage2017}.
+In all cases, these approaches could be further improved by insights from \name.
+
+Perhaps the most direct comparison to our work is
+QuickerCheck~\cite{krookQuickerCheckImplementingEvaluating2024}, an
+implementation of Haskell's QuickCheck that exploits the inherent parallelism of
+PBT to achieve significant performance gains. QuickerCheck, like \name, was
+designed with performance engineering in mind, taking seriously the idea that
+PBT is a performance critical task. Still, despite their similar motivations and
+considerations, the solutions are entirely different and complementary.
 

### 2025-03-21 paper/paper.tex
Nit

@@
 
-Perhaps the most direct comparison to our work is
+The most direct comparison to our work is
 QuickerCheck~\cite{krookQuickerCheckImplementingEvaluating2024}, an

### 2025-03-21 paper/paper.tex
Nit
minor eval edits

@@
 We lay a path forward.
-First, we show how to use \textit{multi-stage programming}---a classic technique for 
-building fast DSLs “without regret”~\cn{}---to eliminate the overhead of 
-many common generator abstractions.
-Second, we demostrate that entropy source is an undervalued consideration in generator library design
-by isolating its performance contribution, and showing that significant
-speedups are possible---even while maintaining output equivalence.
-These insights culminate in \name, a recipe for building generator libraries that marry high-level design
+First, we show how to use \textit{multi-stage programming}, a classic technique for 
+building fast DSLs “without regret”~\cn{}, to eliminate the overhead of 
+many common generator abstractions. We refer to this approach as \name.
+Second, we show that the source of entropy is an underappreciated factor in 
+generator library design by isolating its run-time cost and demonstrating that 
+substantial speedups are possible, while maintaining output equivalence.
+These insights give a recipe for building generator libraries that marry high-level design
 with low-level performance.
@@
 In Scala 3, we build \scalaname{}, based on ScalaCheck---the language's standard PBT library, and by far the most widely-used\cn{}.
-In both languages, our libraries provide identical functionality and behavior: given the same random seed, a generator written with our technique produces exactly the same sequence of values as an equivalent generator written with the original library.
-This semantic equivalence enables pointwise, rather than distributional, comparisons, ensuring that performance differences are attributable \textit{solely} to staging and \rand optimizations.
+In both languages, our libraries preserve the semantics of the original: given the same random seed, generators written using \camlname or \scalaname produce exactly the same sequence of values as equivalent generators using ScalaCheck or \bq, respectively.
+This semantic equivalence enables pointwise, as opposed to distributional, comparisons, ensuring that performance differences are attributable \textit{solely} to staging and \rand optimizations.
 % \jwc{BCP thinks this should be a little stronger: ``and not obscured by the probablistic nature of generation'', or something like that.
@@
 by running these case studies in the Etna
-framework~\cite{etna}\bcp{``Etna framework'' sounds like more than
-  what it is}, a
-platform that simulates real-world PBT usage\hg{Don't love this phrase, I'll
-think more about how we should talk about Etna} by injecting bugs into a system and
+platform~\cite{etna}, which simulates real-world PBT usage \iflater\hg{Don't love this phrase, I'll
+think more about how we should talk about Etna}\fi by injecting bugs into a system and
 measuring how quickly various generators detect them.
 
-In summary, we show that monadic PBT generator DSLs incur significant performance costs across languages, and present a general technique for improving their efficiency while preserving their idiomatic abstractions.
+In summary, we show that PBT generator DSLs incur significant performance costs across languages, and present a general technique for improving their efficiency while preserving their idiomatic abstractions.
 \jwc{needs a bit about RNG --- BCP thoughts?}
@@
     abstraction overhead and \rand choice.
-    \item We present \name{}, a technique based on staging that eliminates the abstraction overhead of generators.
-We implement \name{} in both OCaml and Scala 3, showcasing its cross-language applicability.
-    \item We demonstrate that both \name{} and choosing a faster \rand yield substantial performance improvements to generator running time. Finally, we show
+    \item We present \name{}, a technique for eliminates the abstraction overhead of generators. We implement \name{} in both OCaml and Scala 3, showcasing its cross-language applicability.
+    \item We demonstrate that programs written using \name{} yield performance improvements to generator running time. Finally, we show
     that these performance improvements extend to significantly improved bug-finding speed. \jwc{This should be sell-ier: 2x speedups, etc. ``bug-finding speed'' isn't a race against time.}
@@
 % \jwc{Initial section here about eval}\tr{todo}
-To assess the impact of our approach, we evaluate the performance and bug-finding speed of
+To assess the impact of our interventions, we evaluate the raw generator performance and bug-finding speed of
 \name generators across a range of benchmarks. Our experiments compare generators
 built using our technique---with and without \csplitmix \jwc{this is the first mention of csplitmix!}---against those implemented with
-existing generator libraries (and their default randomness mechanisms). Specifically,
-we implement
-two fully semantically
+existing generator libraries and their default randomness mechanisms. Specifically,
+we implement semantically
 equivalent staged generator libraries that replicate the behavior of \bq
@@
 
-This section presents our experimental setup, as well as answers to the following research questions:
+This section presents our experimental setup, and address the following research questions:
 
@@
   \item \textbf{RQ1}: Do generators written using our technique run faster than those written with regular generator combinators?
-  \item \textbf{RQ2}: Do performance speedups translate to better bug-finding speed?
+  \item \textbf{RQ2}: Do observed generation speedups translate to better bug-finding speed?
   % \item \textbf{RQ3}: Does our technique generalize to other strict functional languages?
@@
 and OpenJDK 21.0.6, using ScalaCheck version 1.17 as the baseline.
-We used \texttt{core\_bench} 0.16 \cn in OCaml and \texttt{jmh} 0.4.7 \cn for generator performance microbenchmarking.
+We used \texttt{core\_bench} 0.16 \cn in OCaml and \texttt{jmh} 0.4.7 \cn for performance microbenchmarking. \tr{Etna?}
 
@@
 \label{subsection:benchmarking}
-To answer \textbf{RQ1}, we microbenchmark generators, comparing generators in \camlname to identical ones in \bq,
+To answer \textbf{RQ1}, we microbenchmark generators, comparing generators in \camlname to semantically identical ones in \bq,
 and generators in \scalaname to those in ScalaCheck.
@@
 typed lambda calculus (STLC) terms. We implement generators for these benchmarks
-using a variety of \textit{strategies}, which vary in structure and
+variety of \textit{strategies}, varying in structure and
 sophistication. These include type-derived generators for BSTs and STLC terms,
@@
 subtrees at each step; a generator of boolean lists that operates similarly to
-the single-pass BST strategy; and a well-typed generator of
-STLC terms, whose behavior is the most complex.
+the single-pass BST strategy; and a generator of STLC terms that is correct by construction (``well-typed''), whose behavior is the most complex.
 For evaluating \scalaname, we implement a subset of the aformentioned
@@
 Our \camlname type-derived BST generator achieves speedups ranging from $1.08-1.29\times$, which increase
-dramatically to $6.13-9.65\times$ when combined with \csplitmix. The insertion-based BST generator sees
+dramatically, to $6.13-9.65\times$, when combined with \csplitmix. The insertion-based BST generator sees
 $1.13-1.21\times$ speedups with staging alone, and $4.32-5.36\times$ when also
 using \csplitmix. The single-pass BST generator benefits more significantly from
-staging, with speedups of $2.31-2.34\times$, which rise to $8.61-8.95\times$ when combined with
-\csplitmix. We attribute the differential in performance speedup to the fact that
+staging, with speedups of $2.31-2.34\times$, rising to $8.61-8.95\times$ when combined with
+\csplitmix. We attribute the difference in performance speedup to the fact that
 BST strategies rely on relatively few calls to generator combinator functions,
 but sample extensively from the \rand; notably, single-pass BST, which uses the largest
-number of calls to generator combinators, also sees the greatest benefit from staging.
+number of calls to generator combinators, also sees the greatest benefit from staging. \tr{Call this out earlier? Unless we have the fancy plot.}
 
 For STLC, staging accounts for the bulk of performance gains, yielding
- $1.71-2.64\times$ speedups for the type-derived STLC generator, and $4.60-5.46\times$ for
- the well-typed generator. When adding \csplitmix, these numbers increase to
- $2.60-5.15\times$ and $4.83-7.75\times$, respectively. This result makes sense because
+ $1.71-2.64\times$ speedups for the type-derived STLC generator and $4.60-5.46\times$ for
+ the well-typed generator. Adding \csplitmix, these numbers increase to
+ $2.60-5.15\times$ and $4.83-7.75\times$. This result makes sense because
  STLC generators, particularly the well-typed generator, sample very little from
- randomness libraries, but extensively on generator library functions.
+ randomness libraries but rely extensively on generator library functions.
 
@@
 combined with \csplitmix ($3.43-6.83\times$). This is likely because sampling
-booleans is ``cheap'' enough that the overhead of crossing the FFI barrier is
+booleans is cheap enough that the overhead of crossing the FFI barrier is
 greater than that of generating values directly in OCaml.
@@
 As shown in \figref{fig:scala}, \scalaname achieves an even greater performance
-gain---purely from staging---over ScalaCheck than \camlname does over \bq. This discrepancy arises from
+gain---purely from staging---over ScalaCheck than \camlname does over \bq. This difference arises from
 ScalaCheck's representation of generators as functions of type \texttt{size ->
@@
 pattern-match on these \texttt{Option} values, introducing significant boxing
-and unboxing overhead at each step.  Staging allows us to eliminate this
-overhead, resulting in substantial speedups in aggregate.
-
-These results show that staging and \rand choice serve distinct but
+and unboxing overhead at each step.  Staging eliminates this
+overhead.
+
+These results show that staging and \rand choice play distinct and
 complementary roles in generator performance. Sampling-heavy generators see
 greater improvements from faster randomness libraries, while combinator-heavy
-ones benefit more from staging. Since both factors influence performance in
-different ways, generator libraries should use both in order to handle diverse
+ones benefit more from staging. Since both factors influence performance 
+significantly, generator libraries should use both in order to handle diverse
 workloads.
@@
   \includegraphics[scale=0.45]{time.png}
-  \caption{Time to generate a varying-size value using each strategy, averaged over 10,000 trials.
-Performance benefits scale with generation size; lower is better.
+  \caption{Time to generate using each strategy.
+Performance benefits scale with generation size (lower lines translate to greater performance benefit).
 BQ is \bq; Staged SR is \camlname using \bq's \texttt{splittable\_random} RNG; Staged CSR is \camlname using \secref{subsection:faster-rng}'s \csplitmix.}
@@
 
-Similarly orthogonal to our approach is techniques for quickly filtering out
+Similarly orthogonal to our approach are techniques for quickly filtering out
 inputs that are invalid for testing. Some of these approaches work statically,

### 2025-03-21 paper/paper.tex
chart

@@
 
-We lay a path forward.
+We chart a path forward.
 First, we show how to use \textit{multi-stage programming}, a classic technique for 
@@
 many common generator abstractions. We refer to this approach as \name.
-Second, we show that the source of entropy is an underappreciated factor in 
-generator library design by isolating its run-time cost and demonstrating that 
-substantial speedups are possible, while maintaining output equivalence.
-These insights give a recipe for building generator libraries that marry high-level design
-with low-level performance.
+Second, we demonstrate that the source of entropy is an underappreciated design factor by 
+isolating its runtime cost and showing that significant speedups can be achieved 
+while preserving output equivalence.
+These insights lay the groundwork for generator libraries that combine high-level 
+design with low-level performance.
 

### 2025-03-21 paper/paper.tex
decouple ocaml and scala

@@
 % \jwc{Initial section here about eval}\tr{todo}
-To assess the impact of our interventions, we evaluate the raw generator performance and bug-finding speed of
+We evaluate the raw generator performance and bug-finding speed of
 \name generators across a range of benchmarks. Our experiments compare generators
-built using our technique---with and without \csplitmix \jwc{this is the first mention of csplitmix!}---against those implemented with
+built using our technique---with and without our improved SplitMix (``\csplitmix'')---against those implemented with
 existing generator libraries and their default randomness mechanisms. Specifically,
@@
 technique across different languages and runtime
-environments.\bcp{This says ``assess the impact'' twice...}
-
-This section presents our experimental setup, and address the following research questions:
+environments.
+
+This section presents our experimental setup, and addresses the following research questions:
 
@@
  We vary the choice of \rand in our \camlname generators, using both \bq's default
-\texttt{splittable\_random} and our improved SplitMix (\csplitmix), as discussed
-in \secref{subsection:faster-rng}.
-
-For \camlname, our test cases consist of generators for boolean lists, binary search trees (BST), and simply
+\texttt{splittable\_random} and \csplitmix, as discussed
+in \secref{subsection:faster-rng}. Our test cases consist of generators for boolean lists, binary search trees (BST), and simply
 typed lambda calculus (STLC) terms. We implement generators for these benchmarks
 variety of \textit{strategies}, varying in structure and
-sophistication. These include type-derived generators for BSTs and STLC terms,
+sophistication.
+
+\subsubsection{OCaml}
+\label{subsubsection:ocaml}
+
+In \camlname, our strategies include type-derived generators for BSTs and STLC terms
 using the approach outlined in \secref{subsection:type-derived}; two custom BST
@@
 subtrees at each step; a generator of boolean lists that operates similarly to
-the single-pass BST strategy; and a generator of STLC terms that is correct by construction (``well-typed''), whose behavior is the most complex.
-For evaluating \scalaname, we implement a subset of the aformentioned
-strategies---the boolean list generator, single-pass BST generator, and
-well-typed STLC term generator.
+the single-pass BST strategy; and a generator of STLC terms that is correct by construction (``well-typed'').
 
@@
 
-\tr{We might put a graph of calls to generator library functions vs. speedup here, but
-I'm not sure it's worth it---thoughts?}
+\subsubsection{Scala 3}
+In \scalaname, we implement a subset of the \camlname
+strategies: the boolean list generator, single-pass BST generator, and
+well-typed STLC term generator. ScalaCheck does not support type derivation,
+so those strategies are excluded.
 

### 2025-03-21 paper/paper.tex
nit

@@
 in \secref{subsection:faster-rng}. Our test cases consist of generators for boolean lists, binary search trees (BST), and simply
-typed lambda calculus (STLC) terms. We implement generators for these benchmarks
-variety of \textit{strategies}, varying in structure and
+typed lambda calculus (STLC) terms. We implement generators for these benchmarks using a
+variety of \textit{strategies} varying in structure and
 sophistication.
@@
 $1.13-1.21\times$ speedups with staging alone, and $4.32-5.36\times$ when also
-using \csplitmix. The single-pass BST generator benefits more significantly from
+using \csplitmix. The single-pass BST generator benefits more from
 staging, with speedups of $2.31-2.34\times$, rising to $8.61-8.95\times$ when combined with

### 2025-03-21 paper/paper.tex
methodology

@@
 
-\subsubsection{OCaml}
+\subsubsection{\camlname}
 \label{subsubsection:ocaml}
 
-In \camlname, our strategies include type-derived generators for BSTs and STLC terms
-using the approach outlined in \secref{subsection:type-derived}; two custom BST
-generators, one that repeatedly inserts values into an initially empty tree and
-one that materializes the tree in a single pass by generating keys, values, and
-subtrees at each step; a generator of boolean lists that operates similarly to
-the single-pass BST strategy; and a generator of STLC terms that is correct by construction (``well-typed'').
-
-The runtime of \camlname generators across different input sizes is summarized in \figref{fig:time}.
-Our \camlname type-derived BST generator achieves speedups ranging from $1.08-1.29\times$, which increase
+In \camlname, our strategies include type-derived generators for BSTs and STLC terms, 
+following the approach described in \secref{subsection:type-derived}; two custom 
+BST generators—--one that builds a tree incrementally by repeatedly inserting 
+values into an initially empty structure, and another that constructs the 
+entire tree in a single pass by generating keys, values, and subtrees at each step; 
+a boolean list generator that mirrors the single-pass BST strategy; and a 
+generator for STLC terms that is correct by construction (i.e., it produces 
+only well-typed terms).
+
+These strategies differ in their use of generator library combinators 
+and in how frequently they sample random values. The type-derived and 
+insertion-based BST generators sample frequently but use combinators 
+sparingly. The single-pass BST generator also samples frequently 
+but uses more combinators, striking a middle ground. 
+The type-derived 
+STLC generator samples infrequently and makes moderate 
+use of combinators, while the custom STLC generator uses 
+significantly more combinators than any other strategy and 
+also takes the fewest samples. The boolean list generator 
+makes moderate use of combinators 
+and samples booleans frequently---however, because booleans are 
+inexpensive to sample, its overall sampling overhead is lower 
+than that of the other strategies.
+
+For each strategy, we compare three treatments: our \bq baseline; 
+a \camlname version using \bq's randomness 
+library, \texttt{splittable\_random}; and a
+\camlname version using \csplitmix. For each treatment, we measure the
+time to generate a value---i.e., a BST, a STLC term, or a boolean list---using a random seed, over a 
+variety of different sizes (10, 100, 1,000, 10,000).\footnote{The way size is used in a generator is a domain-specific implementation detail---in a list generator, size might correspond to the desired length of the list, whereas in a tree generator it refer to number of nodes, number of leaves, depth of tree, etc.
+Regardless of how size is encoded, our goal is to show that performance trends scale.}
+We run each treatment for 5 seconds---enough time for several thousand generations---and compute the average generation time of all values produced in that interval.
+
+Our results are summarized in \figref{fig:time}.
+Our type-derived \camlname BST generator achieves speedups ranging from $1.08-1.29\times$, which increase
 dramatically, to $6.13-9.65\times$, when combined with \csplitmix. The insertion-based BST generator sees
@@
 staging, with speedups of $2.31-2.34\times$, rising to $8.61-8.95\times$ when combined with
-\csplitmix. We attribute the difference in performance speedup to the fact that
-BST strategies rely on relatively few calls to generator combinator functions,
-but sample extensively from the \rand; notably, single-pass BST, which uses the largest
-number of calls to generator combinators, also sees the greatest benefit from staging. \tr{Call this out earlier? Unless we have the fancy plot.}
+\csplitmix. 
+% We attribute the difference in performance speedup to the fact that
+% BST strategies rely on relatively few calls to generator combinator functions,
+% but sample extensively from the \rand; notably, single-pass BST, which uses the largest
+% number of calls to generator combinators, also sees the greatest benefit from staging. \tr{Call this out earlier? Unless we have the fancy plot.}
 
@@
 the well-typed generator. Adding \csplitmix, these numbers increase to
-$2.60-5.15\times$ and $4.83-7.75\times$. This result makes sense because
-STLC generators, particularly the well-typed generator, sample very little from
-randomness libraries but rely extensively on generator library functions.
+$2.60-5.15\times$ and $4.83-7.75\times$.
+
+% This result makes sense because
+% STLC generators, particularly the well-typed generator, sample very little from
+% randomness libraries but rely extensively on generator library functions.
 
@@
 
-\subsubsection{Scala 3}
+These results show that staging and \rand choice play distinct and
+complementary roles in generator performance. Sampling-heavy generators see
+greater improvements from faster randomness libraries, while combinator-heavy
+ones benefit more from staging. Since both factors influence performance 
+significantly, generator libraries should use both in order to handle diverse
+workloads.
+
+\subsubsection{\scalaname}
 In \scalaname, we implement a subset of the \camlname
@@
 overhead.
-
-These results show that staging and \rand choice play distinct and
-complementary roles in generator performance. Sampling-heavy generators see
-greater improvements from faster randomness libraries, while combinator-heavy
-ones benefit more from staging. Since both factors influence performance 
-significantly, generator libraries should use both in order to handle diverse
-workloads.
 

### 2025-03-21 paper/paper.tex
wibbles

--- 2025-03-21 paper/paper.tex
@@
 %\jwc{``Property based testing is a race against time...''}
-Property-based testing (PBT) is a software testing technique that uses
+
+Property-based testing (PBT) is a software-testing technique that uses
 random test inputs to validate logical specifications. Recent studies on PBT usage~\cite{inpractice} in industry have shown that
@@
 libraries}, usually expressed as embedded domain-specific languages (eDSLs) %\cntxt{edsls}
-that provide combinators for building and composing generators. However, careful 
-measurements show that the performance of existing generator libraries falls far short of what
-is possible.
+that provide combinators for building and composing
+generators. However, careful measurements show that the performance of
+existing generator libraries---even well-engineered and widely used
+ones---falls far short of what is possible.
 
@@
 Second, the \rand that a generator uses to make choices can
-constitute an unexpectedly high proportion of its runtime. These factors combine to limit 
+be responsible for a surprisingly high proportion of its runtime. Combined, these factors limit
 the effectiveness of PBT.
@@
 
-We chart a path forward.
-First, we show how to use \textit{multi-stage programming}, a classic technique for 
-building fast DSLs “without regret”~\cn{}, to eliminate the overhead of 
-many common generator abstractions. We refer to this approach as \name.
-Second, we demonstrate that the source of entropy is an underappreciated design factor by 
-isolating its runtime cost and showing that significant speedups can be achieved 
-while preserving output equivalence.
-These insights lay the groundwork for generator libraries that combine high-level 
-design with low-level performance.
-
-% \bcp{It
-% feels a bit grand to call what we've done an ``approach''---it's
-% really just a single performance-enhancing technique, staging, plus
-% the observation that the RNG matters and you should use the fastest
-% one you can.}
-
-To evaluate our technique, we implement staged PBT libraries in two strict functional languages: OCaml and Scala 3.
+We address both sources of inefficiency.
+First, we show how to use \textit{multi-stage programming}---a classic technique for
+building fast DSLs “without regret”~\cn{}---to eliminate the overhead of
+many common generator abstractions.
+Second, we characterize the impact of \rand choice through a controlled experiment 
+that isolates the performance contribution of the \rand within a generator, underscoring its role as a first-class design consideration.\bcp{We've done more than this, though: We've also
+  demonstrated that you can get an X\% improvement in performance by
+  using a specific RNG.}
+Taken together, these insights can be applied to create tomorrow's generator libraries: those that 
+combine high-level design with low-level performance.
+
+To demonstrate the effectiveness and generality of our approach\bcp{It
+feels a bit grand to call what we've done an ``approach''---it's
+really just a single performance-enhancing technique, staging, plus
+the observation that the RNG matters and you should use the fastest
+one you can.}, which we call \name, we implement staged PBT libraries in multiple languages.
 In OCaml, we build \camlname{}, based on \bq{}---a
 high-quality PBT library in OCaml, authored by software engineers at the trading firm Jane Street \cn{}.
-In Scala 3, we build \scalaname{}, based on ScalaCheck---the language's standard PBT library, and by far the most widely-used\cn{}.
-In both languages, our libraries preserve the semantics of the original: given the same random seed, generators written using \camlname or \scalaname produce exactly the same sequence of values as equivalent generators using ScalaCheck or \bq, respectively.
-This semantic equivalence enables pointwise, as opposed to distributional, comparisons, ensuring that performance differences are attributable \textit{solely} to staging and \rand optimizations.
-% \jwc{BCP thinks this should be a little stronger: ``and not obscured by the probablistic nature of generation'', or something like that.
+In Scala 3, we build \scalaname{}, based on ScalaCheck---a very
+mature PBT library and the {\em de facto} choice\bcp{the wording here
+  implies that there are other choices in Scala, some of which might
+  be reasonable, and tht base\_quickcheck is not the de facto choice
+  in OCaml...} in Scala \cn{}.
+In both languages, our libraries are direct drop-in
+replacements\bcp{that's not quite accurate, is it? I mean, they have
+  the same functionality, but can you take a generator that works with
+base\_quickcheck and just link it against our version without changing
+the generator itself? That's what ``drop-in replacement'' means to me.} that are
+\textit{100\% semantically equivalent}: that is, given identical random seeds, a generator written with our technique
+produces exactly the same sequence of values as the same generator written with the original generator library.
+The semantic equivalence between the original generators and \name{} generators allows us to
+perform clean, fine-grained comparisons, as differences in generation speed are
+solely attributable to the staging and \rand optimizations.
+\jwc{BCP thinks this should be a little stronger: ``and not obscured by the probablistic nature of generation'', or something like that.
 ``Comparing distributions of speed is hard, comparing pointwise is easier.''.
@@
 by running these case studies in the Etna
-platform~\cite{etna}, which simulates real-world PBT usage \iflater\hg{Don't love this phrase, I'll
-think more about how we should talk about Etna}\fi by injecting bugs into a system and
+framework~\cite{etna}\bcp{``Etna framework'' sounds like more than
+  what it is}, a
+platform that simulates real-world PBT usage\hg{Don't love this phrase, I'll
+think more about how we should talk about Etna} by injecting bugs into a system and
 measuring how quickly various generators detect them.
 
-In summary, we show that PBT generator DSLs incur significant performance costs across languages, and present a general technique for improving their efficiency while preserving their idiomatic abstractions.
+In summary, we show that monadic PBT generator DSLs incur significant performance costs across languages, and present a general technique for improving their efficiency while preserving their idiomatic abstractions.
 \jwc{needs a bit about RNG --- BCP thoughts?}
@@
     abstraction overhead and \rand choice.
-    \item We present \name{}, a technique for eliminates the abstraction overhead of generators. We implement \name{} in both OCaml and Scala 3, showcasing its cross-language applicability.
-    \item We demonstrate that programs written using \name{} yield performance improvements to generator running time. Finally, we show
+    \item We present \name{}, a technique based on staging that eliminates the abstraction overhead of generators.
+We implement \name{} in both OCaml and Scala 3, showcasing its cross-language applicability.
+    \item We demonstrate that both \name{} and choosing a faster \rand yield substantial performance improvements to generator running time. Finally, we show
     that these performance improvements extend to significantly improved bug-finding speed. \jwc{This should be sell-ier: 2x speedups, etc. ``bug-finding speed'' isn't a race against time.}
@@
 \section{What are Generator Libraries, and why are they Slow?}
-\tr{This needs to introduce PBT WAY better.}
 \label{section:motiv}
@@
 \jwc{Unclear what we should call this section}
+\bcp{``Performance opportunities in the RNG?''}
 As we discussed in Section~\ref{section:motiv}, choosing an inefficient randomness library is
-another bottleneck for finding bugs fast. While generator libraries by
+another impediment to finding bugs fast. While generator libraries by
 and large use sensible sources of randomness, they are not explicitly chosen
 with performance in mind. Indeed, much faster randomness libraries\cn and fast sources
-of entropy \cn exist. \jwc{... more here?}
+of entropy \cn exist. \bcp{... more here?}
 To demonstrate that faster random sampling can significanty impact bugfinding power, we
@@
 % \jwc{Initial section here about eval}\tr{todo}
-We evaluate the raw generator performance and bug-finding speed of
+To assess the impact of our approach, we evaluate the performance and bug-finding speed of
 \name generators across a range of benchmarks. Our experiments compare generators
-built using our technique---with and without our improved SplitMix (``\csplitmix'')---against those implemented with
-existing generator libraries and their default randomness mechanisms. Specifically,
-we implement semantically
+built using our technique---with and without \csplitmix \jwc{this is the first mention of csplitmix!}---against those implemented with
+existing generator libraries (and their default randomness mechanisms). Specifically,
+we implement
+two fully semantically
 equivalent staged generator libraries that replicate the behavior of \bq
@@
 technique across different languages and runtime
-environments.
-
-This section presents our experimental setup, and addresses the following research questions:
+environments.\bcp{This says ``assess the impact'' twice...}
+
+This section presents our experimental setup, as well as answers to the following research questions:
 
@@
   \item \textbf{RQ1}: Do generators written using our technique run faster than those written with regular generator combinators?
-  \item \textbf{RQ2}: Do observed generation speedups translate to better bug-finding speed?
+  \item \textbf{RQ2}: Do performance speedups translate to better bug-finding speed?
   % \item \textbf{RQ3}: Does our technique generalize to other strict functional languages?
@@
 and OpenJDK 21.0.6, using ScalaCheck version 1.17 as the baseline.
-We used \texttt{core\_bench} 0.16 \cn in OCaml and \texttt{jmh} 0.4.7 \cn for performance microbenchmarking. \tr{Etna?}
+We used \texttt{core\_bench} 0.16 \cn in OCaml and \texttt{jmh} 0.4.7 \cn for generator performance microbenchmarking.
 
@@
 \label{subsection:benchmarking}
-To answer \textbf{RQ1}, we microbenchmark generators, comparing generators in \camlname to semantically identical ones in \bq,
+To answer \textbf{RQ1}, we microbenchmark generators, comparing generators in \camlname to identical ones in \bq,
 and generators in \scalaname to those in ScalaCheck.
  We vary the choice of \rand in our \camlname generators, using both \bq's default
-\texttt{splittable\_random} and \csplitmix, as discussed
-in \secref{subsection:faster-rng}. Our test cases consist of generators for boolean lists, binary search trees (BST), and simply
-typed lambda calculus (STLC) terms. We implement generators for these benchmarks using a
-variety of \textit{strategies} varying in structure and
-sophistication.
-
-\subsubsection{OCaml}
-\label{subsubsection:ocaml}
-
-In \camlname, our strategies include type-derived generators for BSTs and STLC terms
+\texttt{splittable\_random} and our improved SplitMix (\csplitmix), as discussed
+in \secref{subsection:faster-rng}.
+
+For \camlname, our test cases consist of generators for boolean lists, binary search trees (BST), and simply
+typed lambda calculus (STLC) terms. We implement generators for these benchmarks
+using a variety of \textit{strategies}, which vary in structure and
+sophistication. These include type-derived generators for BSTs and STLC terms,
 using the approach outlined in \secref{subsection:type-derived}; two custom BST
@@
 subtrees at each step; a generator of boolean lists that operates similarly to
-the single-pass BST strategy; and a generator of STLC terms that is correct by construction (``well-typed'').
+the single-pass BST strategy; and a well-typed generator of
+STLC terms, whose behavior is the most complex.
+For evaluating \scalaname, we implement a subset of the aformentioned
+strategies---the boolean list generator, single-pass BST generator, and
+well-typed STLC term generator.
 
 The runtime of \camlname generators across different input sizes is summarized in \figref{fig:time}.
-Our \camlname type-derived BST generator achieves speedups ranging from $1.08-1.29\times$, which increase
-dramatically, to $6.13-9.65\times$, when combined with \csplitmix. The insertion-based BST generator sees
-$1.13-1.21\times$ speedups with staging alone, and $4.32-5.36\times$ when also
-using \csplitmix. The single-pass BST generator benefits more from
-staging, with speedups of $2.31-2.34\times$, rising to $8.61-8.95\times$ when combined with
-\csplitmix. We attribute the difference in performance speedup to the fact that
+Our \camlname type-derived BST generator achieves speedups ranging from 1.08--1.29x, which increase
+dramatically to 6.13--9.65x when combined with \csplitmix. The insertion-based BST generator sees
+1.13--1.21x speedups with staging alone, and 4.32--5.36x when also
+using \csplitmix. The single-pass BST generator benefits more significantly from
+staging, with speedups of 2.31--2.34x, which rise to 8.61--8.95x when combined with
+\csplitmix. We attribute the differential in performance speedup to the fact that
 BST strategies rely on relatively few calls to generator combinator functions,
 but sample extensively from the \rand; notably, single-pass BST, which uses the largest
-number of calls to generator combinators, also sees the greatest benefit from staging. \tr{Call this out earlier? Unless we have the fancy plot.}
+number of calls to generator combinators, also sees the greatest benefit from staging.
 
 For STLC, staging accounts for the bulk of performance gains, yielding
-$1.71-2.64\times$ speedups for the type-derived STLC generator and $4.60-5.46\times$ for
-the well-typed generator. Adding \csplitmix, these numbers increase to
-$2.60-5.15\times$ and $4.83-7.75\times$. This result makes sense because
+ 1.71--2.64x speedups for the type-derived STLC generator, and 4.60--5.46x for
+ the well-typed generator. When adding \csplitmix, these numbers increase to
+ 2.60--5.15x and 4.83--7.75x, respectively. This result makes sense because
  STLC generators, particularly the well-typed generator, sample very little from
-randomness libraries but rely extensively on generator library functions.
-
-The boolean list generator experiences $3.61-7.10\times$ speedups with
+ randomness libraries, but extensively on generator library functions.
+
+The boolean list generator experiences 3.61--7.10x speedups with
 staging, and its performance actually slightly \textit{decreases} when
-combined with \csplitmix ($3.43-6.83\times$). This is likely because sampling
-booleans is cheap enough that the overhead of crossing the FFI barrier is
+combined with \csplitmix (3.43--6.83x). This is likely because sampling
+booleans is ``cheap'' enough that the overhead of crossing the FFI barrier is
 greater than that of generating values directly in OCaml.
 
-\subsubsection{Scala 3}
-In \scalaname, we implement a subset of the \camlname
-strategies: the boolean list generator, single-pass BST generator, and
-well-typed STLC term generator. ScalaCheck does not support type derivation,
-so those strategies are excluded.
+\tr{We might put a graph of calls to generator library functions vs. speedup here, but
+I'm not sure it's worth it---thoughts?}
 
 As shown in \figref{fig:scala}, \scalaname achieves an even greater performance
-gain---purely from staging---over ScalaCheck than \camlname does over \bq. This difference arises from
+gain---purely from staging---over ScalaCheck than \camlname does over \bq. This discrepancy arises from
 ScalaCheck's representation of generators as functions of type \texttt{size ->
@@
 pattern-match on these \texttt{Option} values, introducing significant boxing
-and unboxing overhead at each step.  Staging eliminates this
-overhead.
-
-These results show that staging and \rand choice play distinct and
+and unboxing overhead at each step.  Staging allows us to eliminate this
+overhead, resulting in substantial speedups in aggregate.
+
+These results show that staging and \rand choice serve distinct but
 complementary roles in generator performance. Sampling-heavy generators see
 greater improvements from faster randomness libraries, while combinator-heavy
-ones benefit more from staging. Since both factors influence performance 
-significantly, generator libraries should use both in order to handle diverse
+ones benefit more from staging. Since both factors influence performance in
+different ways, generator libraries should use both in order to handle diverse
 workloads.
@@
   \includegraphics[scale=0.45]{time.png}
-  \caption{Time to generate using each strategy.
-Performance benefits scale with generation size (lower lines translate to greater performance benefit).
+  \caption{Time to generate a varying-size value using each strategy, averaged over 10,000 trials.
+Performance benefits scale with generation size; lower is better.
 BQ is \bq; Staged SR is \camlname using \bq's \texttt{splittable\_random} RNG; Staged CSR is \camlname using \secref{subsection:faster-rng}'s \csplitmix.}
@@
 
+  \item \textbf{F#} 
+  % https://learn.microsoft.com/en-us/dotnet/fsharp/language-reference/code-quotations
+
   \item \textbf{Haskell} is home to the original PBT implementation \cn{}, and
@@
 \subsection{Speeding up Property-Based Testing}
-
-Our work is unique in its approach to speeding up PBT, but it is certainly not
-the only work focused on making PBT faster.
-
-Perhaps the best-explored approaches to speeding up bug-finding with PBT is by
-changing the order in which test inputs are sampled. Feedback-driven mechanisms
-like targeted PBT~\cite{loscherTargetedPropertybasedTesting2017} and
-coverage-guided PBT~\cite{lampropoulosCoverageGuidedProperty2019} have the
-potential to speed up testing by changing the generation order to try to find
-``interesting'' inputs faster. Separately, enumerative approaches to
-PBT~\cite{runcimanSmallcheckLazySmallcheck2008,braquehaisSimpleIncrementalDevelopment2017}
-try to get to bugs faster by leveraging the ``small-scope hypothesis'' and
-attempting to quickly exhaust the list of small inputs to the program under
-test. All of these approaches can have significant performance benefits in
-practice, but they are largely orthogonal to our contributions; we expect
-\name{} could be used to speed these existing techniques to varying degrees.
-
-Similarly orthogonal to our approach are techniques for quickly filtering out
-inputs that are invalid for testing. Some of these approaches work statically,
-by automatically deriving generators that produce valid inputs by
-construction~\cite{lampropoulosGeneratingGoodGenerators2017}, while others
-filter dynamically via laziness~\cite{claessenGeneratingConstrainedRandom2015}
-or by solving satisfiability
-problems~\cite{seidelTypeTargetedTesting2015,steinhofelInputInvariants2022,lampropoulosBeginnerLuckLanguage2017}.
-In all cases, these approaches could be further improved by insights from \name.
-
-The most direct comparison to our work is
-QuickerCheck~\cite{krookQuickerCheckImplementingEvaluating2024}, an
-implementation of Haskell's QuickCheck that exploits the inherent parallelism of
-PBT to achieve significant performance gains. QuickerCheck, like \name, was
-designed with performance engineering in mind, taking seriously the idea that
-PBT is a performance critical task. Still, despite their similar motivations and
-considerations, the solutions are entirely different and complementary.
+\hg{\begin{itemize}
+  \item Changing the generation order~\cite{runcimanSmallcheckLazySmallcheck2008,braquehaisSimpleIncrementalDevelopment2017}
+  \item Dealing with filtering~\cite{claessenGeneratingConstrainedRandom2015}
+  \item QuickerCheck~\cite{krookQuickerCheckImplementingEvaluating2024}
+  \item Evaluation of PBT generation~\cite{shiEtnaEvaluationPlatform2023}
+\end{itemize}}
 

### 2025-03-21 paper/paper.tex
update figs

@@
 many common generator abstractions. We refer to this approach as \name.
-Second, we demonstrate that the source of entropy is an underappreciated design factor by 
+Second, we demonstrate that a generator's source of entropy is an underappreciated design factor by 
 isolating its runtime cost and showing that significant speedups can be achieved 
@@
   % \includegraphics[scale=0.33]{allocs.png}
-  \includegraphics[scale=0.45]{time.png}
+  \includegraphics[scale=0.32]{time.png}
   \caption{Time to generate using each strategy.
@@
   \centering
-  \includegraphics[scale=0.4]{scala.png}
+  \includegraphics[scale=0.46]{scala.png}
  \caption{

### 2025-03-22 paper/paper.tex
update figs
change colors
new graph

@@
   % \includegraphics[scale=0.33]{allocs.png}
-  \includegraphics[scale=0.32]{time.png}
-  \caption{Time to generate using each strategy.
-Performance benefits scale with generation size (lower lines translate to greater performance benefit).
-BQ is \bq; Staged SR is \camlname using \bq's \texttt{splittable\_random} RNG; Staged CSR is \camlname using \secref{subsection:faster-rng}'s \csplitmix.}
+  \includegraphics[scale=0.43]{time.png}
+  \caption{Time to generate values of varying sizes using each \camlname strategy.
+Lower lines translate to greater performance benefit.
+BQ is \bq; Staged SR is \camlname using \bq's \texttt{splittable\_random} RNG; Staged CSR is \camlname using \csplitmix.}
   \label{fig:time}
@@
   \centering
-  \includegraphics[scale=0.46]{scala.png}
- \caption{
-  \jwc{Don't explain the chart in the caption, split this into two different charts --- purple on one, green on the other.}
-  \jwc{Also can we use the language's colors for this?}
-  \scalaname has better performance relative to ScalaCheck than \camlname relative to \bq. SC is ScalaCheck; BQ is \bq. Comparison is for a fixed size of 100.}
+  \includegraphics[scale=0.49]{scala.png}
+ \caption{Time to generate values of varying sizes using each \scalaname strategy. SC is ScalaCheck; Staged is \scalaname.}
  \label{fig:scala}
@@
   \caption{Geometric average of all speedups---relative to BQ---for each strategy and benchmark, showing that staging leads to better bug-finding speed across the board.
-  \jwc{Charts start at one? Can eliminate the BQ bar}
   }
@@
   \includegraphics[scale=0.31]{swarm.png}
-  \caption{Swarm distribution of individual-task speedups across strategies and benchmarks---gray dots represent tasks.
-Plots with extremely low variance have no swarm overlay; in plots with high variance, outliers correspond to large speedups.
-\tr{Make mean higher-contrast.}}
+  \caption{Distribution of individual-task speedups across strategies and benchmarks.
+Gray dots represent tasks.
+Plots with extremely low variance have no swarm overlay; in plots with high variance, outliers correspond to large speedups.}
   \label{fig:swarm}
@@
 
-  \item \textbf{F#} 
+  \item \textbf{F\#} 
   % https://learn.microsoft.com/en-us/dotnet/fsharp/language-reference/code-quotations

### 2025-03-22 paper/paper.tex
blah

@@
 Our results are summarized in \figref{fig:time}.
-Our type-derived \camlname BST generator achieves speedups ranging from $1.08-1.29\times$, which increase
-dramatically, to $6.13-9.65\times$, when combined with \csplitmix. The insertion-based BST generator sees
-$1.13-1.21\times$ speedups with staging alone, and $4.32-5.36\times$ when also
+Our type-derived \camlname BST generator achieves speedups ranging from $1.30-1.38\times$, which increase
+dramatically, to $2.13-7.76\times$, when combined with \csplitmix. The insertion-based BST generator sees
+$1.18-1.22\times$ speedups with staging alone, and $3.83-6.31\times$ when also
 using \csplitmix. The single-pass BST generator benefits more from
-staging, with speedups of $2.31-2.34\times$, rising to $8.61-8.95\times$ when combined with
+staging, with speedups of $2.22-2.25\times$, rising to $9.05-9.82\times$ when combined with
 \csplitmix. 
@@
 For STLC, staging accounts for the bulk of performance gains, yielding
-$1.71-2.64\times$ speedups for the type-derived STLC generator and $4.60-5.46\times$ for
+$1.73-4.05\times$ speedups for the type-derived STLC generator and $4.10-5.71\times$ for
 the well-typed generator. Adding \csplitmix, these numbers increase to
-$2.60-5.15\times$ and $4.83-7.75\times$.
+$2.90-5.39\times$ and $5.33-8.33\times$.
 
@@
 
-The boolean list generator experiences $3.61-7.10\times$ speedups with
-staging, and its performance actually slightly \textit{decreases} when
-combined with \csplitmix ($3.43-6.83\times$). This is likely because sampling
+The boolean list generator experiences $2.77-7.47\times$ speedups with
+staging, and its performance changes very little when
+combined with \csplitmix ($2.77-7.57\times$). This is likely because sampling
 booleans is cheap enough that the overhead of crossing the FFI barrier is
-greater than that of generating values directly in OCaml.
+comparable to that of generating values directly in OCaml.
 
@@
   \centering
-  \includegraphics[scale=0.49]{scala.png}
+  \includegraphics[scale=0.41]{scala.png}
  \caption{Time to generate values of varying sizes using each \scalaname strategy. SC is ScalaCheck; Staged is \scalaname.}

### 2025-03-22 paper/paper.tex
Redoing the beginning of section 2

@@
 
-Since property-based testing was first introduced in Haskell's QuickCheck
-library~\cite{claessenQuickCheckLightweightTool2000}, {\em generators} have been
-a key element of the PBT process. While properties\bcp{this concept
-  comes out of nowhere} are critical for correctly
-expressing a developer's intent for a function, the success of the testing
-ultimately rests on how well the random inputs to that property exercise the
-system under test.\bcp{Is this trying to say that generators are more
-  important than properties?  This seems like a strange assertion.}
-QuickCheck's solution to this problem, which has been copied in
-dozens of frameworks across a wide variety of languages, is to provide a domain-specific
-language (DSL) for writing random data generators that gives developers complete
-control\bcp{that might be overstating the case a bit!} over the
-distribution of their test inputs.
+Property-based testing~\cite{claessenQuickCheckLightweightTool2000} is an approach to software testing that centers around
+executable specifications of programs called {\em properties}. For example, if a
+programmer wants to test an invariant of a binary search tree (BST)
+implementation that they are working on, they may write a property like
+\begin{center}
+  \texttt{prop\_insert\_invariant t x = isBST t ==> isBST (insert x t)}
+\end{center}
+to check that for any tree \texttt{t} and value \texttt{x}, if \texttt{t} is
+already a valid BST then inserting \texttt{x} into it also yields a valid BST.
+Once a developer has a property, they test that property by executing
+it on hundreds or thousands of random test inputs. These test inputs are usually
+produced by a {\em generator} --- a program written in some domain-specific
+language (DSL) that allows the developer to express precisely how values should
+be sampled.
 
@@
 
-The standard design for QuickCheck-style
-generators is via an embedded {\em monadic} language. We use the syntax and types from OCaml's \bq library, which is presented in
+The standard design for a generator DSL, introduced in Haskell's QuickCheck
+library and copied in dozens of other frameworks,
+is via an embedded {\em monadic} language. We use the syntax and types from OCaml's \bq library, which is presented in
 Figure~\ref{fig:bq-api}, but similar DSLs can also be found in languages

### 2025-03-22 paper/paper.tex
blah

@@
 
-These strategies differ in their use of generator library combinators 
-and in how frequently they sample random values. The type-derived and 
-insertion-based BST generators sample frequently but use combinators 
-sparingly. The single-pass BST generator also samples frequently 
-but uses more combinators, striking a middle ground. 
-The type-derived 
-STLC generator samples infrequently and makes moderate 
-use of combinators, while the custom STLC generator uses 
-significantly more combinators than any other strategy and 
-also takes the fewest samples. The boolean list generator 
-makes moderate use of combinators 
-and samples booleans frequently---however, because booleans are 
-inexpensive to sample, its overall sampling overhead is lower 
-than that of the other strategies.
+% These strategies differ in their use of generator library combinators 
+% and in how frequently they sample random values. The type-derived and 
+% insertion-based BST generators sample frequently but use combinators 
+% sparingly. The single-pass BST generator also samples frequently 
+% but uses more combinators, striking a middle ground. 
+% The type-derived 
+% STLC generator samples infrequently and makes moderate 
+% use of combinators, while the custom STLC generator uses 
+% significantly more combinators than any other strategy. 
+% The boolean list generator 
+% makes moderate use of combinators 
+% and samples booleans frequently---however, because booleans are 
+% inexpensive to sample, its overall sampling overhead is lower 
+% than that of the other strategies.
 
@@
 time to generate a value---i.e., a BST, a STLC term, or a boolean list---using a random seed, over a 
-variety of different sizes (10, 100, 1,000, 10,000).\footnote{The way size is used in a generator is a domain-specific implementation detail---in a list generator, size might correspond to the desired length of the list, whereas in a tree generator it refer to number of nodes, number of leaves, depth of tree, etc.
+variety of different generation sizes (10, 100, 1,000, 10,000).\footnote{The way size is used in a generator is a domain-specific implementation detail---in a list generator, size might correspond to the desired length of the list, whereas in a tree generator it refer to number of nodes, number of leaves, depth of tree, etc.
 Regardless of how size is encoded, our goal is to show that performance trends scale.}
 We run each treatment for 5 seconds---enough time for several thousand generations---and compute the average generation time of all values produced in that interval.
+
+\begin{figure}[h]
+  % \includegraphics[scale=0.33]{allocs.png}
+  \includegraphics[scale=0.43]{time.png}
+  \caption{Time to generate values of varying sizes using each \camlname strategy.
+Lower lines translate to greater performance benefit.
+BQ is \bq; Staged SR is \camlname using \bq's \texttt{splittable\_random} RNG; Staged CSR is \camlname using \csplitmix.}
+  \label{fig:time}
+\end{figure}
 
@@
 
+To investigate the relationship between a generator's use of library 
+abstractions and the performance benefits it gains from staging, 
+we focused on the \texttt{bind} function---a core component of monadic 
+generator libraries, frequently invoked both directly and indirectly 
+through other combinators. We instrumented \texttt{bind} to count how many 
+times it was called during a generator's execution. Our 
+hypothesis was that generators which make heavier use of 
+library abstractions like \texttt{bind} would benefit more from staging.
+
+We evaluated this hypothesis using four benchmarks with widely varying 
+speedups: insertion-based BSTs, single-pass BSTs, boolean lists, and 
+well-typed STLC terms. For each generator, we ran 10,000 trials, 
+using a fixed generation size of 100, and recorded the average number 
+of \texttt{bind} calls. We then compared these usage statistics to the 
+observed speedups.
+
+As shown in \figref{fig:binds}, the results reveal a clear linear 
+correlation between the number of \texttt{bind} calls and the speedup 
+achieved through staging. This suggests that generators that 
+rely more heavily on the library's abstractions tend to 
+benefit more from staging optimizations. \tr{TALK ABOUT RANDOMNESS HERE.}
+
+\begin{figure}[h]
+  \includegraphics[scale=0.4]{bind.png}
+  \caption{}
+  \label{fig:binds}
+\end{figure}
+
 These results show that staging and \rand choice play distinct and
@@
 strategies: the boolean list generator, single-pass BST generator, and
-well-typed STLC term generator. ScalaCheck does not support type derivation,
-so those strategies are excluded.
+well-typed STLC term generator\footnote{By default, ScalaCheck does not support type derivation,
+so those strategies are excluded.}. We did not implement an optimized version of ScalaCheck's
+\rand, but our experimental setup was otherwise the same as in \secref{subsubsection:ocaml}.
 
 As shown in \figref{fig:scala}, \scalaname achieves an even greater performance
-gain---purely from staging---over ScalaCheck than \camlname does over \bq. This difference arises from
+gain---purely from staging---over ScalaCheck than \camlname does over \bq. In particular,
+the single-pass BST strategy is $4.89-6.51\times$ faster, the boolean list generator is $6.86-13.41\times$ faster,
+and STLC is $5.43-9.70\times$ faster. This difference arises from
 ScalaCheck's representation of generators as functions of type \texttt{size ->
-seed -> Option[A]}. Each generator combinator must construct and then
+seed -> Option[A]}: each generator combinator must construct and then
 pattern-match on these \texttt{Option} values, introducing significant boxing
@@
 are implemented in
-\camlname; they are unchanged from their description in \secref{subsection:benchmarking}.\footnote{Etna does
-not support Scala, so we were unable to test \scalaname's bug-finding speed in Etna, but we
-we see no reason our results in this section would not extend to \scalaname.}
+\camlname; they are unchanged from their description in \secref{subsection:benchmarking}. 
+Etna does not support Scala, so we were unable to test \scalaname's bug-finding speed in Etna, but we
+we see no reason our results in this section would not extend to \scalaname.
 
@@
 
-The average individual-task speedups for each strategy and benchmark are shown in
-\figref{fig:etna}, with trends in bug-finding speed closely mirroring those in performance
+Our results are summarized in \figref{fig:etna}, which shows the geometric average of 
+individual-task speedups for each strategy and benchmark. Trends in bug-finding speed reflect trends in performance
 from \secref{subsection:benchmarking}. The distribution of speedups in \figref{fig:swarm} reveals
@@
 find that ``easier'' tasks---those that run on the order of milliseconds---regularly achieve speedups in the
-range of 4--9x, whereas tasks that run on the order of seconds achieve more moderate speedups of up to 3.5x.
+range of $4-9\times$, whereas tasks that run on the order of seconds achieve more moderate speedups of up to $3.5\times$.
 Overall,

### 2025-03-22 paper/paper.tex
blah
3.8

@@
 
-% \begin{figure}
-% \begin{lstlisting}
-% \end{lstlisting}
-% \caption{Int Pair Generators in \bq{} and ScalaCheck}
-% \label{fig:overhead-explanation-code}
-% \end{figure}
-
 The performance impact of this simplification is large (Figure~\ref{fig:overhead-explanation-perf}). In both languages, the
@@
 To demonstrate that faster random sampling can significanty impact bugfinding power, we
-use the controlled experiment provided by OCaml's inefficient implementation of
-SplitMix. By replacing this slow randomness library with a faster but semantically equivalent
-implemetnation, we are able to precisely quantify the bugfinding speedup that using a randomness library gives across a range of PBT scenarios.
-
-We emphasize that while we believe that the insight that faster sampilng translates to faster bug finding
-is a cross-langauge one, the specific technical contents of this section are OCaml-specific.
-In the case of most other PBT frameworks \cn, the randomness library used operates on by machine
+use the controlled experiment suggested by OCaml's inefficient implementation of
+SplitMix. By replacing this relatively slow randomness library with a faster but semantically equivalent
+implemetnation, we can precisely quantify the bugfinding speedup that a better randomness library gives across a range of PBT scenarios.
+\jwc{Ensure this gets pushed to a top level contrib}
+
+Note that while that the insight that faster sampilng translates to faster bug finding
+is a cross-language one, the performance intervention described here is OCaml-specific.
+In most other PBT frameworks \cn, the randomness library used operates on machine
 integers, so this \emph{particular} inefficiency does not exist.
 
-The precise details of how the SplitMix algorithm works \cn are unimportant for the
-present paper, but the critical component is that all of its operations are are
-defined in terms of arithmetic bitwise operations on 64-bit integers. In OCaml,
+The precise details of how the SplitMix algorithm works \cn are not important for
+present purposes; the key fact is that all of its operations are are
+defined in terms of arithmetic and bitwise operations on 64-bit integers. In OCaml,
 because of details related to the garbage collector, the 64-bit integer type \texttt{int64} is represented
 at run time as a \emph{pointer} to an unscanned block of memory containing (among other things)
-a 64-bit integer \cn. This means that all operations that return an \texttt{int64} must allocate this block of memory.
-This has a significant impact on the performance of generators. A single call to
-one of the \bq library functions---like generating an integer uniformly in a range---may call the \texttt{splittable\_random}
-algorithm multiple times. Each sample from \texttt{splittable\_random} allocates 9
-times \jwc{this is a call to \texttt{next\_int64}}, and each allocation brings us closer
-to the next garbage collection pause. While small allocations like these are \emph{very} fast to perform and subsequently collect in OCaml,
-\footnote{The OCaml GC is a generational collector \cn, and since these allocations are small and mostly very short lived, they will all be minor allocations, never to be promoted.}, we will see in Section~\ref{section:eval} that this can have
-a large performance impact on some generators that spend most of their time sampling data.
-To circumvent this allocation and provide an equivalent version of \texttt{splittable\_random}, we
-reimplement SplitMix in C, and call out to it with the OCaml FFI. The C version of the library uses
-proper \texttt{int64\_t} arithmetic, only boxing and unboxing integers at the call boundaries between OCaml and C code.
-Ideally in the future, one would not need to call out to C for this: the Jane Street bleeding-edge OCaml compiler has support
+a 64-bit integer \cn. This means that all operations that return an \texttt{int64} must allocate this block of memory,
+which has a significant impact on performance. A single call to
+one of the \bq library functions---like generating an arbitrary integer---may call into the \texttt{Splittable\_random}
+library times. Each call into \texttt{Splittable\_random}
+may sample many times from the core SplitMix sampling routine \texttt{next\_int64}, i.e. using rejection sampling to find a value within a range.
+Finally, each call to \texttt{next\_int64} allocates 9 times, and each allocation brings us closer
+to the next garbage collection pause. While small allocations like these are \emph{very} fast to perform and subsequently
+collect in OCaml\footnote{The OCaml GC is a generational collector \cn, and since these allocations are small and mostly very short lived, they will all be minor allocations, never to be promoted.}
+, we will see in Section~\ref{section:eval} that they can still have
+a large performance impact on generators that spend most of their time sampling data.
+To avoid this allocation, we
+reimplement SplitMix in C and call out to it with the OCaml FFI. The C version of the library uses
+proper \texttt{int64\_t} arithmetic, only boxing and unboxing integers at the call boundaries between OCaml and
+C code\footnote{Ideally in the future, one would not need to call out to C for this: the Jane Street bleeding-edge OCaml compiler has support
 for unboxed types \cn, which (among other things) would let us implement a version of SplitMix that does not allocate, directly
 in OCaml. Unfortunately, the Jane Street branch of the compiler is incompatible with MetaOCaml, which we use to implement the metaprogramming
-discussed in the previous sections.
+discussed in the previous sections.}.
 
@@
   \includegraphics[scale=0.4]{bind.png}
-  \caption{}
+  \caption{Relationship between the number of binds---a proxy for reliance on generator library abstractions---and speedup from staging, showing that the two relate linearly.}
   \label{fig:binds}

### 2025-03-22 paper/paper.tex
blah

@@
 following the approach described in \secref{subsection:type-derived}; two custom 
-BST generators—--one that builds a tree incrementally by repeatedly inserting 
+BST generators---one that builds a tree incrementally by repeatedly inserting 
 values into an initially empty structure, and another that constructs the 
@@
 \camlname version using \csplitmix. For each treatment, we measure the
-time to generate a value---i.e., a BST, a STLC term, or a boolean list---using a random seed, over a 
-variety of different generation sizes (10, 100, 1,000, 10,000).\footnote{The way size is used in a generator is a domain-specific implementation detail---in a list generator, size might correspond to the desired length of the list, whereas in a tree generator it refer to number of nodes, number of leaves, depth of tree, etc.
+time to generate a value (i.e., a BST, a STLC term, or a boolean list), using a random seed, varying generation sizes (10, 100, 1,000, 10,000).\footnote{The way size is used in a generator is a domain-specific implementation detail---in a list generator, size might correspond to the desired length of the list, whereas in a tree generator it refer to number of nodes, number of leaves, depth of tree, etc.
 Regardless of how size is encoded, our goal is to show that performance trends scale.}
-We run each treatment for 5 seconds---enough time for several thousand generations---and compute the average generation time of all values produced in that interval.
+We run each treatment for 5 seconds---enough for several thousand generations---and compute the average generation time of a value produced in that interval.
 
@@
 The boolean list generator experiences $2.77-7.47\times$ speedups with
-staging, and its performance changes very little when
+staging, and its performance changes minimally when
 combined with \csplitmix ($2.77-7.57\times$). This is likely because sampling
@@
 
-To investigate the relationship between a generator's use of library 
-abstractions and the performance benefits it gains from staging, 
-we focused on the \texttt{bind} function---a core component of monadic 
-generator libraries, frequently invoked both directly and indirectly 
-through other combinators. We instrumented \texttt{bind} to count how many 
-times it was called during a generator's execution. Our 
-hypothesis was that generators which make heavier use of 
-library abstractions like \texttt{bind} would benefit more from staging.
-
-We evaluated this hypothesis using four benchmarks with widely varying 
-speedups: insertion-based BSTs, single-pass BSTs, boolean lists, and 
-well-typed STLC terms. For each generator, we ran 10,000 trials, 
-using a fixed generation size of 100, and recorded the average number 
-of \texttt{bind} calls. We then compared these usage statistics to the 
-observed speedups.
-
-As shown in \figref{fig:binds}, the results reveal a clear linear 
+The variation in speedups raises a natural question: what 
+determines how much a generator benefits from staging? One 
+likely factor is its degree of reliance on generator library functions.
+This is hard to quantify directly, since not all functions 
+contribute equally to overhead. As a proxy, we count how many 
+times a generator calls \texttt{bind}---the central function in 
+monadic generator libraries, extensively used both directly and within 
+other combinators.
+
+We select four benchmarks with disparate speedups: insertion-based BSTs, single-pass BSTs, 
+boolean lists, and well-typed STLC terms. For each generator, 
+we run 10,000 trials using a fixed generation size of 100, and 
+record the average number of \texttt{bind} calls. Results in \figref{fig:binds} show a clear linear 
 correlation between the number of \texttt{bind} calls and the speedup 
-achieved through staging. This suggests that generators that 
-rely more heavily on the library's abstractions tend to 
-benefit more from staging optimizations. \tr{TALK ABOUT RANDOMNESS HERE.}
+achieved by staging, supporting the idea that generators which make heavier use of 
+generator library functions tend to benefit more.
 
@@
 \end{figure}
+
+Similarly, we investigate the impact of \csplitmix by tracking the number of calls to \texttt{next\_int64} in \bq's \texttt{splittable\_random} 
+library, which samples the next random integer.
+This, too, is an imperfect proxy, as the SplitMix algorithm's runtime does plenty of stuff that isn't 
+sampling which can dominate its runtime---which is all faster in \csplitmix. However, it too validates our intuition that sampling-heavy generators achieve
+greater benefit from \csplitmix than they do staging:
 

### 2025-03-22 paper/paper.tex
bleh
some wibbles on the intro

@@
 constitute an unexpectedly high proportion of its runtime. These factors combine to limit 
-the effectiveness of PBT.
+the effectiveness of PBT. \jwc{Unsure if ``constitute a high proportion of its runtime'' limits effectiveness. it's more that the choice \rand limits effectiveness.}
 \iflater
@@
 
-We chart a path forward.
+We study and address these issues separately.
 First, we show how to use \textit{multi-stage programming}, a classic technique for 
@@
 many common generator abstractions. We refer to this approach as \name.
-Second, we demonstrate that a generator's source of entropy is an underappreciated design factor by 
-isolating its runtime cost and showing that significant speedups can be achieved 
-while preserving output equivalence.
-These insights lay the groundwork for generator libraries that combine high-level 
-design with low-level performance.
+% Secondly, we demonstrate the
+% impact of diferent choices of randomness library on generator performance by conducting a
+% controlled comparison.
+Second, we demonstrate that a DSL's choice of \rand{} has a
+dramatic impact on its performance by performing a controlled experiment,
+isolating the performance benefits of faster randomness in a clearly quantifiable manner.
+\jwc{Another cut at this.}
+
+% These insights lay the groundwork for generator libraries that combine high-level 
+% design with low-level performance.
 
@@
 
-To evaluate our technique, we implement staged PBT libraries in two strict functional languages: OCaml and Scala 3.
+To demonstrate the multi-language applicability of \name{}, we implement staged generator DSLs in two strict functional languages: OCaml and Scala 3.
 In OCaml, we build \camlname{}, based on \bq{}---a
 high-quality PBT library in OCaml, authored by software engineers at the trading firm Jane Street \cn{}.
-In Scala 3, we build \scalaname{}, based on ScalaCheck---the language's standard PBT library, and by far the most widely-used\cn{}.
+In Scala 3, we build \scalaname{}, based on ScalaCheck---the language's standard PBT library\cn{}.
 In both languages, our libraries preserve the semantics of the original: given the same random seed, generators written using \camlname or \scalaname produce exactly the same sequence of values as equivalent generators using ScalaCheck or \bq, respectively.
-This semantic equivalence enables pointwise, as opposed to distributional, comparisons, ensuring that performance differences are attributable \textit{solely} to staging and \rand optimizations.
+This semantic equivalence enables pointwise---as opposed to distributional---bug-finding comparisons, ensuring that if a \name{} generator finds bugs faster, this is attributable \textit{solely} to staging and \rand optimizations, not luck with seeds.
 % \jwc{BCP thinks this should be a little stronger: ``and not obscured by the probablistic nature of generation'', or something like that.
@@
 We evaluate the optimizations with a series of case studies in each
-generator library, and show that the \name generators run faster than their unstaged equivalents.
+generator library, and show that the \name{} generators run faster than their unstaged equivalents.
 Further, we show generators that run faster also find bugs faster

### 2025-03-22 paper/paper.tex
work on intro

@@
 Further, we show generators that run faster also find bugs faster
-% ability \jwc{The paper is called ``fail faster'' --- we shouldn't say ``bug finding ability'', we should say something like ``how fast you can squash bugs''. This is too passive }
 by running these case studies in the Etna
@@
     \item We present \name{}, a technique for eliminates the abstraction overhead of generators. We implement \name{} in both OCaml and Scala 3, showcasing its cross-language applicability.
-    \item We demonstrate that programs written using \name{} yield performance improvements to generator running time. Finally, we show
-    that these performance improvements extend to significantly improved bug-finding speed. \jwc{This should be sell-ier: 2x speedups, etc. ``bug-finding speed'' isn't a race against time.}
-    \jwc{We precisely quantify the performance effects of staging and faster rng across a range of testing scenarios --- i.e.
-a contrib is the fact that we can exactly pin down when one helps versus the other.
-The ``logically smarter'' your generator is, the more you get out of staging, the ``dumber'', the more you get out of rng.}
-    \bcp{Agree with this. We need to beef up our claims here about what we learned about PBT-specifics.}
-    \jwc{And this needs to be flowed through the rest of the paper.}
+    \item We demonstrate that writing generators using \name{} and fast \rands{} yields performance improvements.
+    We precisely quantify the impact of each optimization separately, and explain how they play distinct yet complementary roles in improving generator performance.
+    Finally, we show that these performance improvements extend to significantly improved bug-finding speed, finding some bugs as much as \todo{\%\%} faster.
 \end{enumerate}

### 2025-03-22 paper/paper.tex
wibbling

@@
 Second, the \rand that a generator uses to make choices can
-constitute an unexpectedly high proportion of its runtime. These factors combine to limit 
-the effectiveness of PBT. \jwc{Unsure if ``constitute a high proportion of its runtime'' limits effectiveness. it's more that the choice \rand limits effectiveness.}
+constitute an large proportion of its running time, magnifying the impact of an inefficient one.
 \iflater
@@
 
-We study and address these issues separately.
+In this paper, we study and address these issues.
 First, we show how to use \textit{multi-stage programming}, a classic technique for 
@@
     \item We present \name{}, a technique for eliminates the abstraction overhead of generators. We implement \name{} in both OCaml and Scala 3, showcasing its cross-language applicability.
-    \item We demonstrate that writing generators using \name{} and fast \rands{} yields performance improvements.
+    \item We demonstrate that writing generators using \name{} and fast \rands{} yields significant performance improvements.
     We precisely quantify the impact of each optimization separately, and explain how they play distinct yet complementary roles in improving generator performance.
-    Finally, we show that these performance improvements extend to significantly improved bug-finding speed, finding some bugs as much as \todo{\%\%} faster.
+    Finally, we show that these performance improvements extend to improved bug-finding speed, finding some bugs as much as \todo{\%\%} faster.
 \end{enumerate}

### 2025-03-22 paper/paper.tex
wibbling in s2

@@
 isolating the performance benefits of faster randomness in a clearly quantifiable manner.
-\jwc{Another cut at this.}
 
@@
 A generator \texttt{'a Bq.t} is just a function of type~\texttt{int -> SR.t -> 'a},
-\jwc{introduce what SR is}
 taking an \texttt{int} representing the current size parameter and a random seed \texttt{SR.t},
-and returning a generated value \texttt{'a}. It is an invariant of the generator library that every function of this type is
+and returning a generated value \texttt{'a}.
+The random seed \texttt{SR.t} from \bq's \rand{} called \texttt{Splittable\_random}, henceforth aliased in code as \texttt{SR}.
+It is an invariant of the generator library that every function of this type is
 deterministic: for a fixed size and seed, it will always return the same value, so all of the randomness in testing comes from varying the initial seed.
-The monad functions \texttt{return} and \texttt{bind} are defined in the usual way
-for an instance of the reader monad \cn{} \jwc{explain better}.
-% \bq{} uses a mutable seed, so the \texttt{random} passed to \texttt{(k a)} will be different from the one passed to \texttt{g}.
+The monad functions \texttt{return} and \texttt{bind} are defined in the usual way:
+for an instance of the reader monad \cn{}: \texttt{return} ignores the size and seed and returns its argument, while \texttt{bind} runs \texttt{g} and passes the result to \texttt{k}.
 The \texttt{gen\_int} combinator simply calls out to the randomness library \texttt{Splittable\_random}, aliased as \texttt{SR} here.
-\jwc{lift this explanation, explain what splittable random is before here.}
 
@@
   fun _ sr ->
-    let x = Splittable_random.int sr ~lo:0 ~hi:100 in
-    let y = Splittable_random.int sr ~lo:0 ~hi:x in
+    let x = SR.int sr ~lo:0 ~hi:100 in
+    let y = SR.int sr ~lo:0 ~hi:x in
     (x,y)
@@
 difficult to compare \bcp{This was not discussed previously!}. To get around this, we exploit a coincidence:
-the \rand{} that \bq{} uses --- OCaml's \texttt{Splittable\_random} library --- is slow in a way that
+\texttt{Splittable\_random}---the \rand{} that \bq{} uses---is slow in a way that
 can be improved \emph{without} changing its behavior.

### 2025-03-22 paper/paper.tex
wobbles

@@
 The \texttt{gen\_int} combinator simply calls out to the randomness library \texttt{Splittable\_random}, aliased as \texttt{SR} here.
-
-Different PBT libraries use variations on this basic design. The largest source of variation is how the random seed is threaded through the generator.
-\bcp{Sentence starts with lowerase} \bq{} uses a reader monad with a mutable seed, while Haskell's QuickCheck uses a reader monad with an immutable state
-that is ``split'' at \texttt{bind}s \cn. Meanwhile, ScalaCheck (a) uses an immutable state type and state monad \cn{} to thread
-the state through, and (b) has an \texttt{Option} as its return type, to allow generation to fail.
-\bcp{This discussion should go elsewhere...}
+Different generator DSLs use variatoins on this basic design, but the basics are the same across the board.
+
+% Different PBT libraries use variations on this basic design. The largest source of variation is how the random seed is threaded through the generator.
+% \bcp{Sentence starts with lowerase} \bq{} uses a reader monad with a mutable seed, while Haskell's QuickCheck uses a reader monad with an immutable state
+% that is ``split'' at \texttt{bind}s \cn. Meanwhile, ScalaCheck (a) uses an immutable state type and state monad \cn{} to thread
+% the state through, and (b) has an \texttt{Option} as its return type, to allow generation to fail.
+% \bcp{This discussion should go elsewhere...}
 
@@
 perform this search, which incurs further run time overhead.
-\bcp{Are we advocating a linear scan? If not, then maybe we don't need to make this point}\jwc{We are, but we do need to make this point, since
-it's important to understanding why we get more speedup than you might expect}
 
@@
 Choosing a different \rand{} would break this property, making it much more challenging to assess the performance compared to the baseline.
-But, as discussed previously, generators with different generation orders can be
-difficult to compare \bcp{This was not discussed previously!}. To get around this, we exploit a coincidence:
+To get around this, we exploit a coincidence:
 \texttt{Splittable\_random}---the \rand{} that \bq{} uses---is slow in a way that

### 2025-03-22 paper/paper.tex
wibbling

@@
 Even basic generators --- like ones generating a single \texttt{int} or \texttt{float} uniformly within a range --- can sample \emph{unboundedly many}
-random numbers, since they usually use versions of rejection sampling to find a value within the range \cn{}\bcp{Really?
-I want to know more about that!!}\jwc{Well this is totally standard, and it's how sampling in a bounded range works.
-You take a thing that samples uniformly over \texttt{int64\_t} and do a combination of clever things and basic rejection sampling to get a value in the range.}.
+random numbers, since they usually use versions of rejection sampling \cn{} to find a value within the range.
 Moreover, generator combinators like \texttt{list} usually make $O(n)$ calls to the \texttt{int} or \texttt{float} generators.
@@
 \label{section:faster-generators}
-\jwc{This section needs a much better title...}
-% \jwc{Usual introduction to this section, corresponding to how we talked about it in the intro. ``We present a library that XYZ''.}
-
-\jwc{This section needs an intro, and some signposting to say where we're going.}
+With a sense of the inefficiences pervasive in generator libraries, we set out in this section to find solutions.
+\todo{JWC}
+
+\jwc{This paragraph is a first draft, and needs to be edited!!!}
 
@@
 \label{subsection:basic-design}
-\jwc{Needs a lot more signposting at the top here.}
-
 To build a staged version of a generator DSL, we must rewrite the the generator
-combinators to do as much as possible at compile time,
-producing inlined code that uses no combinators, which can be compiled and run
+combinators to do as much as possible at compile time.
+The compile time stage will then produce simplified code, free of any DSL combinators, which can be compiled and run
 with different sizes and seeds.
-
 Our job is thus to carefully bisect the DSL, determining which inputs to generator

### 2025-03-22 paper/paper.tex
rough stab

@@
 
-For STLC, staging accounts for the bulk of performance gains, yielding
+For STLC, staging accounts for larger performance gains, yielding
 $1.73-4.05\times$ speedups for the type-derived STLC generator and $4.10-5.71\times$ for
@@
 
-The variation in speedups raises a natural question: what 
-determines how much a generator benefits from staging? One 
-likely factor is its degree of reliance on generator library functions.
-This is hard to quantify directly, since not all functions 
-contribute equally to overhead. As a proxy, we count how many 
-times a generator calls \texttt{bind}---the central function in 
-monadic generator libraries, extensively used both directly and within 
-other combinators.
-
-We select four benchmarks with disparate speedups: insertion-based BSTs, single-pass BSTs, 
-boolean lists, and well-typed STLC terms. For each generator, 
-we run 10,000 trials using a fixed generation size of 100, and 
+\subsubsection{Teasing apart speedups} \tr{Working title...}
+
+The variation in speedups in \secref{subsubsection:ocaml} raises a natural question: what 
+determines how much a generator benefits from staging, \rand optimizations, or both?
+One likely explanation is that the two techniques address different performance bottlenecks:
+generators that sample more often benefit more from improved random sampling, while those that rely
+heavily on generator library combinators gain more from staging.
+To test this idea, we look for measurable proxies that approximate a generator's reliance on its generator and randomness libraries.
+For the former, we use the number of calls to \texttt{bind}---the central monadic generator function, extensively used both directly and within other combinators.
+For the latter, we count the number of times a generator invokes the sampling function \texttt{next\_int64}.
+We select four strategies from \secref{subsubsection:ocaml} with disparate speedups: insertion-based BSTs, single-pass BSTs,
+boolean lists, and well-typed STLC terms.
+
+To determine how often \texttt{bind} is invoked, we generate 10,000 values using a fixed generation size of 100, and 
 record the average number of \texttt{bind} calls. Results in \figref{fig:binds} show a clear linear 
 correlation between the number of \texttt{bind} calls and the speedup 
-achieved by staging, supporting the idea that generators which make heavier use of 
-generator library functions tend to benefit more.
+achieved by staging, supporting the idea that generators that make heavier use of 
+generator library functions benefit more.
 
@@
 
-Similarly, we investigate the impact of \csplitmix by tracking the number of calls to \texttt{next\_int64} in \bq's \texttt{splittable\_random} 
-library, which samples the next random integer.
-This, too, is an imperfect proxy, as the SplitMix algorithm's runtime does plenty of stuff that isn't 
-sampling which can dominate its runtime---which is all faster in \csplitmix. However, it too validates our intuition that sampling-heavy generators achieve
-greater benefit from \csplitmix than they do staging:
+To determine the number of random samples, we repeat the same test, but use Intel ProcessorTrace to capture processor activity.
+We then count the number of invocations of \texttt{next\_int64} and average this over the number of complete generations in the trace. As expected,
+the BST strategies, which achieve large speedups from \csplitmix, invoke the sampling function more: 353 times for the insertion-based strategy, and 271 for the single-pass strategy. 
+STLC, which sees moderate improvements from \csplitmix, invokes \texttt{next\_int64} 211 times, and the boolean list strategy, which sees no improvement from \csplitmix, invokes it only 100 times.
 
@@
 % library written in Scala.
-
-\begin{figure}[h]
-  % \includegraphics[scale=0.33]{allocs.png}
-  \includegraphics[scale=0.43]{time.png}
-  \caption{Time to generate values of varying sizes using each \camlname strategy.
-Lower lines translate to greater performance benefit.
-BQ is \bq; Staged SR is \camlname using \bq's \texttt{splittable\_random} RNG; Staged CSR is \camlname using \csplitmix.}
-  \label{fig:time}
-\end{figure}
 

### 2025-03-22 paper/paper.tex
wibble

@@
 strategies: the boolean list generator, single-pass BST generator, and
-well-typed STLC term generator\footnote{By default, ScalaCheck does not support type derivation,
+well-typed STLC term generator\footnote{By default, ScalaCheck does not support deriving gnenerators from types,
 so those strategies are excluded.}. We did not implement an optimized version of ScalaCheck's

### 2025-03-22 paper/paper.tex
words about why no other graph

@@
 
-To determine the number of random samples, we repeat the same test, but use Intel ProcessorTrace to capture processor activity.
-We then count the number of invocations of \texttt{next\_int64} and average this over the number of complete generations in the trace. As expected,
+To determine the number of random samples, we repeat the same test, but use Intel ProcessorTrace \cn{} to capture a 4ms trace processor activity.
+We then count the number of invocations of \texttt{next\_int64}---the actual SplitMix sampling routine---and average this over the number of complete generations in the trace. As expected,
 the BST strategies, which achieve large speedups from \csplitmix, invoke the sampling function more: 353 times for the insertion-based strategy, and 271 for the single-pass strategy. 
-STLC, which sees moderate improvements from \csplitmix, invokes \texttt{next\_int64} 211 times, and the boolean list strategy, which sees no improvement from \csplitmix, invokes it only 100 times.
+STLC, which sees moderate improvements from \csplitmix, invokes \texttt{next\_int64} 211 times, and the boolean list strategy, which sees no improvement from \csplitmix, invokes it only 100 times\footnote{
+We did not measure generator performance for \emph{un}staged generators with \csplitmix, so there is no equivalent of Figure~\ref{fig:binds} for speedup versus sampling.
+}.
 

### 2025-03-22 paper/paper.tex
rg
bind explanation

@@
 to the particular value generated by \texttt{g}, the continuation \texttt{k} gets access to \texttt{code} for the value sampled from \texttt{g}:
-at compile time, we only know \texttt{g} will generate \emph{some} \texttt{'a}, but not which one
-% \footnote{Readers familiar with OCaml may notice that \texttt{return : 'a code -> 'a Gen.t} and \texttt{bind : 'a Gen.t -> ('a code -> 'b Gen.t) -> 'b Gen.t} do not have the correct types for a monad instance, preventing us from using \texttt{let\%bind} notation.
-We rectify this issue in Section~\ref{subsection:codecps} by importing some clever ideas from the staging literature.}.
-\jwc{Fix explanation here: bcp says ``this explanation doesn't quite explain what's going on.
-after all, in a staged program the contnuatino k doesn't know which values is going to be passed to it.
-the difference in the staged setting is that \texttt{k} is not allowed to ask any questions about its argument, it has to treat it totally parametrically.}
-Operationally, bind takes code for the size and seed and returns code that (1) let-binds a variable \texttt{a} to spliced-in code that runs \texttt{g} and then
+we know that at run time the code \texttt{g} generates will produce \emph{some} \texttt{'a}, but at compile time we cannot inspect the value.
+Operationally, bind takes code for the size and seed and returns code that (1) let-binds a variable \texttt{a} to spliced-in code of \texttt{g} and then
 (2) runs the spliced-in continuation \texttt{k}.
@@
 \subsection{Recursive Generators}
-\jwc{Generating recursive datatypes requires recursive generators!}
-
-Different generator DSLs handle defining recursive generators differently. Some allow recursive generators
+Generating recursive datatypes requires recursive generators. Different generator DSLs handle defining recursive generators differently. Some allow recursive generators
 to be defined as recursive functions, while others

### 2025-03-22 paper/paper.tex
blah

@@
 
-We can then redefine our staged generator monad type to be
+We can then redefine\footnote{
+The type \texttt{'a Gen.t} from before is like \texttt{'a code Gen.t} with this new definition. This type change
+carries through all of our combinators. For example, \texttt{Gen.int} now returns \texttt{int code Gen.t}.
+} our staged generator monad type to be
 \texttt{'a Gen.t = int code -> SR.t code -> 'a CodeCPS.t}, as shown in Figure~\ref{fig:codecps-and-final-gen}.
 \jwc{any types that were \texttt{'a Gen.t} before are now \texttt{'a code Gen.t}}.
+
 This gives us the best of both worlds. First, we get a monad instance for \texttt{'a Gen.t} with the correct types, which lets us
@@
 that a program like \texttt{Gen.bind Gen.int (fun x -> ...)} generates a let-binding for the result of sampling the randomness library.
-For example, \lstinline{bind (int .<0>. .<1.>) (fun cx -> return .<(.~cx,.~cx)>.)} now correctly generates
-\lstinline{.< fun size random -> let x = SR.int random 0 1 in (x,x) >.}.
+For example, \texttt{bind (int .<0>. .<1.>) (fun cx -> return .<(.~cx,.~cx)>.)} now correctly generates
+\texttt{.< fun size random -> let x = SR.int random 0 1 in (x,x) >.}.
 This design is less obviously correct, and does require some care. Rather than \texttt{bind} ensuring correct evaluation order once and for all,
-individual combinators must be carefully written to ensure that \texttt{'a code} values that contian effects are \texttt{let\_insert}'d.
-To ensure that the OCaml code is written correctly, we employ a PBT harness to compare its outputs to the corresponding \bq generator over 1,000 random seeds.
-If no discrepancy arises, we conclude that the generators exhibit equivalent behavior.
+individual combinators must be carefully written to ensure that \texttt{'a code} values that contain effects are \texttt{let\_insert}'d.
+To ensure that the library is written correctly, we built a PBT harness to compare staged generators to their \bq equivalents over 1,000 random seeds.
+By differentially testing \cn{} a large suite of generators in this way, we gained confidence that \camlname is equivalent to \bq{}.
 

### 2025-03-22 paper/paper.tex
fix

@@
 flexibility also introduces layered abstractions and indirections that are difficult for
-compilers to optimize, shifting work to run time.
+compilers to optimize\proposecut{, shifting work to run time}.
 Second, the \rand that a generator uses to make choices can
-constitute an large proportion of its running time, magnifying the impact of an inefficient one.
+constitute an large proportion of its running time, magnifying the impact of an inefficient one. \jwc{Object of ``one'' unclear, intended to be ``\rand{}''.}
 \iflater
@@
 In Scala 3, we build \scalaname{}, based on ScalaCheck---the language's standard PBT library\cn{}.
-In both languages, our libraries preserve the semantics of the original: given the same random seed, generators written using \camlname or \scalaname produce exactly the same sequence of values as equivalent generators using ScalaCheck or \bq, respectively.
+In both languages, our libraries preserve the semantics of the original: given the same random seed, generators written using \camlname or \scalaname produce exactly the same sequence of values as equivalent generators using \bq or ScalaCheck, respectively.
 This semantic equivalence enables pointwise---as opposed to distributional---bug-finding comparisons, ensuring that if a \name{} generator finds bugs faster, this is attributable \textit{solely} to staging and \rand optimizations, not luck with seeds.
@@
 \label{section:faster-generators}
+
 With a sense of the inefficiences pervasive in generator libraries, we set out in this section to find solutions.

### 2025-03-22 paper/paper.tex
intro of s3

@@
   \hg{TODO: Abstract}
+  \jwc{Make sure we are clear about our results here. I.e. VERY QUICKLY in this abstract, we need to say ``finds bugs 5x faster''}
 \end{abstract}
@@
 Property-based testing (PBT) is a software testing technique that uses
-random test inputs to validate logical specifications. Recent studies on PBT usage~\cite{inpractice} in industry have shown that
+random test inputs to validate logical specifications. Recent studies on PBT usage in industry~\cite{inpractice} have shown that
 practitioners often interleave PBT with their software development process, sometimes
@@
 
-With a sense of the inefficiences pervasive in generator libraries, we set out in this section to find solutions.
-\todo{JWC}
+With a sense of the inefficiences pervasive in generator libraries, we set out in this section to investigate solutions.
+We begin with a quick tour through multi-stage programming, and then incrementally build \camlname, layering on features and functionality.
+Last, we present the controlled experiment we use to demonstrate how faster \rand{} leads to failing faster.
 

### 2025-03-22 paper/paper.tex
cut at a title
multi

@@
 
-\section{\name{}}
+\section{How can I make my Generator Library Faster?}
 \label{section:faster-generators}
@@
 
-In muli-stage programming (also known more simply as staging), programs execute
+In multi-stage programming (also known more simply as staging), programs execute
 in multiple stages, with each stage producing the code to be run

### 2025-03-23 paper/paper.tex
graph

@@
 \subsection{Inefficient Randomness Libraries}
+\label{subsection:ineff-rand}
 The core of any PBT generator library is a source of randomness.
@@
 Because of this, the speed of a single sample matters a great deal. Unfortunately,
-existing PBT libraries make relatively inefficient choices on this front,
-leading to worse bugfinding power than what is possible.
+while existing PBT libraries by and large make sensible choices for their randomness libraries,
+they are not chosen with performance in mind, leading to worse bugfinding power than what is possible.
 
@@
 \bcp{``Performance opportunities in the RNG?''}
-As we discussed in Section~\ref{section:motiv}, choosing an inefficient randomness library is
-another impediment to finding bugs fast. While generator libraries by
-and large use sensible sources of randomness, they are not explicitly chosen
-with performance in mind. Indeed, much faster randomness libraries\cn and fast sources
-of entropy \cn exist. \bcp{... more here?}
+As we discussed in Section~\ref{subsection:ineff-rand}, choosing an inefficient randomness library is
+another impediment to finding bugs fast. 
 To demonstrate that faster random sampling can significanty impact bugfinding power, we
@@
 implemetnation, we can precisely quantify the bugfinding speedup that a better randomness library gives across a range of PBT scenarios.
-\jwc{Ensure this gets pushed to a top level contrib}
 
@@
 \begin{figure}[h]
-  \includegraphics[scale=0.4]{bind.png}
+  \includegraphics[scale=0.28]{bind.png}
   \caption{Relationship between the number of binds---a proxy for reliance on generator library abstractions---and speedup from staging, showing that the two relate linearly.}

### 2025-03-23 paper/paper.tex
some more wibbles in 3
4.1 new

@@
 For the former, we use the number of calls to \texttt{bind}---the central monadic generator function, extensively used both directly and within other combinators.
-For the latter, we count the number of times a generator invokes the sampling function \texttt{next\_int64}.
+For the latter, we count the number of times a generator invokes the SplitMix sampling routine \texttt{next\_int64}.
 We select four strategies from \secref{subsubsection:ocaml} with disparate speedups: insertion-based BSTs, single-pass BSTs,
@@
 To determine how often \texttt{bind} is invoked, we generate 10,000 values using a fixed generation size of 100, and 
-record the average number of \texttt{bind} calls. Results in \figref{fig:binds} show a clear linear 
-correlation between the number of \texttt{bind} calls and the speedup 
-achieved by staging, supporting the idea that generators that make heavier use of 
-generator library functions benefit more.
+record the average number of \texttt{bind} calls.
+To determine the number of random samples, we repeat the same test, but use Intel ProcessorTrace \cn{} to capture a 4ms snapshot of processor activity.
+We then count the number of invocations of \texttt{next\_int64} and average this over the number of complete generations in the trace.
 
@@
 
-To determine the number of random samples, we repeat the same test, but use Intel ProcessorTrace \cn{} to capture a 4ms trace processor activity.
-We then count the number of invocations of \texttt{next\_int64}---the actual SplitMix sampling routine---and average this over the number of complete generations in the trace. As expected,
-the BST strategies, which achieve large speedups from \csplitmix, invoke the sampling function more: 353 times for the insertion-based strategy, and 271 for the single-pass strategy. 
-STLC, which sees moderate improvements from \csplitmix, invokes \texttt{next\_int64} 211 times, and the boolean list strategy, which sees no improvement from \csplitmix, invokes it only 100 times\footnote{
-We did not measure generator performance for \emph{un}staged generators with \csplitmix, so there is no equivalent of Figure~\ref{fig:binds} for speedup versus sampling.
-}.
-
-These results show that staging and \rand choice play distinct and
-complementary roles in generator performance. Sampling-heavy generators see
-greater improvements from faster randomness libraries, while combinator-heavy
-ones benefit more from staging. Since both factors influence performance 
-significantly, generator libraries should use both in order to handle diverse
-workloads.
+Results are summarized in \figref{fig:binds}. The left plot shows a clear linear 
+relationship between performance benefit from staging and number of calls 
+to \texttt{bind}. The right plot shows a more bimodal distribution, but still validates our intuition that
+generators that sample more also stand to gain more from optimized \rand libraries.
+These results suggest that staging and \rand optimizations address distinct and complimentary roles in generator performance: the former aids combinator-heavy generators, while the latter benefits sampling-heavy ones.
+To support diverse workloads, generator libraries should combine both techniques.
+
+% We conclude that staging and \rand choice play distinct and
+% complementary roles in generator performance. Sampling-heavy generators see
+% greater improvements from faster randomness libraries, while combinator-heavy
+% ones benefit more from staging. Since both factors influence performance 
+% significantly, generator libraries should use both in order to handle diverse
+% workloads.
 

### 2025-03-23 paper/paper.tex
chart

@@
   \includegraphics[scale=0.28]{bind.png}
-  \caption{Relationship between the number of binds---a proxy for reliance on generator library abstractions---and speedup from staging, showing that the two relate linearly.}
+  \caption{Left: Speedup from staging (compared to baseline \bq) versus the number of \texttt{bind} calls.
+  Right: Speedup from \csplitmix (compared to staging alone) versus the number of random samples.}
   \label{fig:binds}
@@
 
-Results are summarized in \figref{fig:binds}. The left plot shows a clear linear 
+\figref{fig:binds} summarizes our results. The left plot shows a clear linear 
 relationship between performance benefit from staging and number of calls 
@@
 and unboxing overhead at each step.  Staging eliminates this
-overhead.
+overhead. These results show that staging generalizes well across languages.
 

### 2025-03-23 paper/paper.tex
small

@@
 
-\subsubsection{Teasing apart speedups} \tr{Working title...}
-
-The variation in speedups in \secref{subsubsection:ocaml} raises a natural question: what 
+The variation in speedups raises a natural question: what 
 determines how much a generator benefits from staging, \rand optimizations, or both?
@@
 heavily on generator library combinators gain more from staging.
-To test this idea, we look for measurable proxies that approximate a generator's reliance on its generator and randomness libraries.
+To test this idea, we need measurable proxies that approximate a generator's reliance on its generator and randomness libraries.
 For the former, we use the number of calls to \texttt{bind}---the central monadic generator function, extensively used both directly and within other combinators.
 For the latter, we count the number of times a generator invokes the SplitMix sampling routine \texttt{next\_int64}.
-We select four strategies from \secref{subsubsection:ocaml} with disparate speedups: insertion-based BSTs, single-pass BSTs,
-boolean lists, and well-typed STLC terms.
-
-To determine how often \texttt{bind} is invoked, we generate 10,000 values using a fixed generation size of 100, and 
+
+We select four strategies with disparate speedups: insertion-based BSTs, single-pass BSTs, 
+boolean lists, and well-typed STLC terms. To determine how often \texttt{bind} is invoked, we generate 10,000 values using a fixed generation size of 100, and 
 record the average number of \texttt{bind} calls.
@@
 relationship between performance benefit from staging and number of calls 
-to \texttt{bind}. The right plot shows a more bimodal distribution, but still validates our intuition that
-generators that sample more also stand to gain more from optimized \rand libraries.
-These results suggest that staging and \rand optimizations address distinct and complimentary roles in generator performance: the former aids combinator-heavy generators, while the latter benefits sampling-heavy ones.
-To support diverse workloads, generator libraries should combine both techniques.
-
-% We conclude that staging and \rand choice play distinct and
-% complementary roles in generator performance. Sampling-heavy generators see
-% greater improvements from faster randomness libraries, while combinator-heavy
-% ones benefit more from staging. Since both factors influence performance 
-% significantly, generator libraries should use both in order to handle diverse
-% workloads.
+to \texttt{bind}. The right plot shows a clear separation between 
+generators that sample heavily (BST strategies) and those that 
+do not (STLC and Bool List), with the former seeing significantly 
+greater speedups from \csplitmix.
+
+From these results, we conclude that staging and \rand choice play distinct and
+complementary roles in generator performance. Sampling-heavy generators see
+greater improvements from faster randomness libraries, while combinator-heavy
+ones benefit more from staging. Since both factors influence performance 
+significantly, generator libraries should use both in order to handle diverse
+workloads.
 

### 2025-03-23 paper/paper.tex
blah

@@
 and unboxing overhead at each step.  Staging eliminates this
-overhead. These results show that staging generalizes well across languages.
+overhead. This result demonstrates the generality of \name generators.
 
@@
 under 5ms, as such negligible runtimes do not yield meaningful insights into
-relative performance. These filters remove
+relative performance. Across all 30 seeds, these filters remove
 28/600 (4.67\%) of type-derived STLC's tasks,

### 2025-03-23 paper/paper.tex
nit

@@
 relationship between performance benefit from staging and number of calls 
-to \texttt{bind}. The right plot shows a clear separation between 
+to \texttt{bind}. The right plot shows a separation between 
 generators that sample heavily (BST strategies) and those that 

### 2025-03-23 paper/paper.tex
typederived

@@
 strategies: the boolean list generator, single-pass BST generator, and
-well-typed STLC term generator\footnote{By default, ScalaCheck does not support deriving gnenerators from types,
-so those strategies are excluded.}. We did not implement an optimized version of ScalaCheck's
+well-typed STLC term generator.\footnote{By default, ScalaCheck does not support deriving gnenerators from types,
+so those strategies are excluded.} We did not implement an optimized version of ScalaCheck's
 \rand, but our experimental setup was otherwise the same as in \secref{subsubsection:ocaml}.
@@
 find the bug are excluded from our dataset, as they provide no basis for
-comparison.  Similarly, we exclude tasks where the BQ strategy completes in
+comparison.  Similarly, we exclude tasks where the \bq strategy completes in
 under 5ms, as such negligible runtimes do not yield meaningful insights into
@@
 optimizations that meaningfully impact performance.
+
+\begin{figure}[h]
+  \includegraphics[scale=0.3]{speedup.png}
+  \caption{Geometric average of all speedups---relative to \bq---for each strategy and benchmark, showing that staging leads to better bug-finding speed across the board.
+  }
+  \label{fig:etna}
+\end{figure}
 
@@
 these results give strong evidence that \camlname consistently benefits bug-finding speed, sometimes drastically.
-\begin{figure}[h]
-  \includegraphics[scale=0.3]{speedup.png}
-  \caption{Geometric average of all speedups---relative to BQ---for each strategy and benchmark, showing that staging leads to better bug-finding speed across the board.
-  }
-  \label{fig:etna}
-\end{figure}
+
 \begin{figure}[h]
   \includegraphics[scale=0.31]{swarm.png}
-  \caption{Distribution of individual-task speedups across strategies and benchmarks.
-Gray dots represent tasks.
-Plots with extremely low variance have no swarm overlay; in plots with high variance, outliers correspond to large speedups.}
+  \caption{Distribution of individual-task speedups across strategies and benchmarks. Gray dots represent tasks. Plots with extremely low variance have no swarm overlay.}
   \label{fig:swarm}
@@
 
+\subsubsection{Do staged type-derived generators reduce timeouts?}
+In addition to \camlname's overall bug-finding ability, we evaluate 
+the performance of its type-derived generators---specifically, 
+whether they can transform previously intractable tasks into 
+tractable ones. Out of 78 tasks that initially timed out, 
+50 (64\%) completed successfully after our interventions. 
+This suggests that staged type-derived generators can provide 
+significant performance improvements for free---that is, without 
+requiring any additional effort from the developer.
 \section{Beyond Strict Functional Languages}

### 2025-03-23 paper/paper.tex
blah

@@
 \subsubsection{Do staged type-derived generators reduce timeouts?}
-In addition to \camlname's overall bug-finding ability, we evaluate 
+In addition to \camlname's overall bug-finding speed, we evaluate 
 the performance of its type-derived generators---specifically, 

### 2025-03-23 paper/paper.tex
blah

@@
 significant performance improvements for free---that is, without 
-requiring any additional effort from the developer.
+requiring any additional effort from the user.
 \section{Beyond Strict Functional Languages}

### 2025-03-23 paper/paper.tex
blah

@@
 significant performance improvements for free---that is, without
-requiring any additional effort from the user.
+ any additional effort from the user.
 \section{Beyond Strict Functional Languages}

### 2025-03-23 paper/paper.tex
type

@@
 convenient---the derivation process requires no manual effort---but also limited:
-they are unable to account for constraints not encoded in the type.
+they are unable to account for constraints not encoded in the type (e.g., they can derive a
+tree, but not a binary search tree).
 As a result, generating values with complex preconditions is statistically unlikely.
-This state of affairs is not ideal, as inexperienced PBT users, who might derive a generator, run their test harness, find no bugs, and assume the code is correct---when in reality, most values are never tested at all because they do not satisfy the preconditions.
+This state of affairs is not ideal, as inexperienced PBT users are liable to derive a generator, run their test harness, find no bugs, and assume the code is correct---when in reality, most values are never tested at all because they do not satisfy the preconditions.
 
 This makes type-derivation a particularly elegant target of \name. Since the process of type derivation is completely automatic,
-the end user does not need to understand the staging syntax---but more importantly, \tr{TODO}
+the end user does not need to understand the staging syntax---but more importantly, applying \name to type-derivation
+means it is more likely for even naive strategies to find a bug. \tr{TODO}
 

### 2025-03-23 paper/paper.tex
type

@@
 This makes type-derivation a particularly elegant target of \name. Since the process of type derivation is completely automatic,
-the end user does not need to understand the staging syntax---but more importantly, applying \name to type-derivation
-means it is more likely for even naive strategies to find a bug. \tr{TODO}
+the end user does not need to understand the staging syntax---but more importantly, \name
+makes it it more likely that even naïve strategies will find a bug. \tr{TODO}
 
@@
 heavily on generator library combinators gain more from staging.
-To test this idea, we need measurable proxies that approximate a generator's reliance on its generator and randomness libraries.
+To test this idea, we use measurable proxies that approximate a generator's reliance on its randomness and generator libraries.
 For the former, we use the number of calls to \texttt{bind}---the central monadic generator function, extensively used both directly and within other combinators.
@@
 
-\subsubsection{Do staged type-derived generators reduce timeouts?}
+\subsubsection{Do improved type-derived generators reduce timeouts?}
 In addition to \camlname's overall bug-finding speed, we evaluate 

### 2025-03-23 paper/paper.tex
touchup
3.8 words in

@@
 the end user does not need to understand the staging syntax---but more importantly, \name
-makes it it more likely that even naïve strategies will find a bug. \tr{TODO}
+makes it it more likely that even naïve strategies will find a bug.
+
+The type-deriving algorithm follows a compositional pattern. Generators for complex types 
+are synthesized by structurally composing generators of their subtypes: 
+base types are mapped to primitive generators included in the generator library;
+product types---such as tuples and records---are handled by sampling each component 
+using \texttt{bind} and aggregating the results; sum types, or variants, are generated using 
+a \texttt{weighted\_union} of the generators for each case; and recursive types are handled by 
+tying the knot at the generator level---recursive references are treated as parameters during 
+derivation, allowing the generator's structure to be defined first and then closed using a 
+fixed-point combinator.
+
+This compositional approach maps naturally to staged generation: 
+to derive a staged generator, we replace each standard combinator 
+with its staged counterpart. The derivation algorithm remains unchanged.
+
+Our implementation in OCaml uses the preprocessor extension---``PPX''---system, but the 
+approach is portable to other languages with metaprogramming capabilities. 
+In Scala, for instance, the same strategy can be realized using typeclass 
+resolution. In both settings, the result is a three-stage process: a 
+metaprogram---via PPX or typeclasses---constructs a generator expression 
+composed of staged combinators; this expression is evaluated at compile 
+time to produce a specialized generator; and finally, the generator is 
+executed at runtime to produce values. The entire 
+pipeline remains invisible to the user.
 
@@
 tractable ones. Out of 78 tasks that initially timed out, 
-50 (64\%) completed successfully after our interventions. 
+30 (38.4\%) completed successfully using \camlname alone, and 
+50 (64.1\%) completed successfully after adding \csplitmix. 
 This suggests that staged type-derived generators can provide 

### 2025-03-23 paper/paper.tex
staging related work

--- 2025-03-23 paper/paper.tex
@@
 The randomness library is the hottest part of the hot path.
-Even basic generators --- like ones generating a single \texttt{int} or \texttt{float} uniformly within a range --- can sample \emph{unboundedly many}
+Indeed, even basic generators --- like ones generating a single \texttt{int} or \texttt{float} uniformly within a range --- can sample \emph{unboundedly many}
 random numbers, since they usually use versions of rejection sampling \cn{} to find a value within the range.
@@
 
+\jwc{This needs to be smith'd}
 For example, significantly faster algorithms than SplitMix or JSF exist,
-for example the Lehmer algorithm\cn. In microbenchmarks, the Lehmer algorithm runs almost 2x as fast as SplitMix \cn.
+such as the Lehmer algorithm\cn, which runs almost 2x as fast as SplitMix in microbenchmarks \cn.
 % https://github.com/lemire/testingRNG/tree/master
@@
 Programmers can then define recursive generators by \texttt{fixed\_point}, which ties the recursive knot.
-\bcp{Again, we could lean into the technical contribution here}\jwc{This is also totally standard.}
 
@@
 convenient---the derivation process requires no manual effort---but also limited:
-they are unable to account for constraints not encoded in the type (e.g., they can derive a
-tree, but not a binary search tree).
+they are unable to account for constraints not encoded in the type.
 As a result, generating values with complex preconditions is statistically unlikely.
-This state of affairs is not ideal, as inexperienced PBT users are liable to derive a generator, run their test harness, find no bugs, and assume the code is correct---when in reality, most values are never tested at all because they do not satisfy the preconditions.
+This state of affairs is not ideal, as inexperienced PBT users, who might derive a generator, run their test harness, find no bugs, and assume the code is correct---when in reality, most values are never tested at all because they do not satisfy the preconditions.
 
 This makes type-derivation a particularly elegant target of \name. Since the process of type derivation is completely automatic,
-the end user does not need to understand the staging syntax---but more importantly, applying \name to type-derivation
-means it is more likely for even naive strategies to find a bug. \tr{TODO}
+the end user does not need to understand the staging syntax---but more importantly, \tr{TODO}
 
@@
 relationship between performance benefit from staging and number of calls 
-to \texttt{bind}. The right plot shows a separation between 
+to \texttt{bind}. The right plot shows a clear separation between 
 generators that sample heavily (BST strategies) and those that 
@@
 strategies: the boolean list generator, single-pass BST generator, and
-well-typed STLC term generator.\footnote{By default, ScalaCheck does not support deriving gnenerators from types,
-so those strategies are excluded.} We did not implement an optimized version of ScalaCheck's
+well-typed STLC term generator\footnote{By default, ScalaCheck does not support deriving gnenerators from types,
+so those strategies are excluded.}. We did not implement an optimized version of ScalaCheck's
 \rand, but our experimental setup was otherwise the same as in \secref{subsubsection:ocaml}.
@@
 and unboxing overhead at each step.  Staging eliminates this
-overhead. This result demonstrates the generality of \name generators.
+overhead. These results show that \name generalizes well across languages.
 
@@
 To answer \textbf{RQ2}, we evaluate the bug-finding speed of our BST and STLC
-case studies using Etna, a platform for assessing and comparing PBT techniques.
+case studies using Etna\cn{}, a platform for assessing and comparing PBT techniques.
 Etna allows users to measure the effectiveness of different generator
@@
 find the bug are excluded from our dataset, as they provide no basis for
-comparison.  Similarly, we exclude tasks where the \bq strategy completes in
+comparison.  Similarly, we exclude tasks where the BQ strategy completes in
 under 5ms, as such negligible runtimes do not yield meaningful insights into
-relative performance. Across all 30 seeds, these filters remove
+relative performance. These filters remove
 28/600 (4.67\%) of type-derived STLC's tasks,
@@
 optimizations that meaningfully impact performance.
-
-\begin{figure}[h]
-  \includegraphics[scale=0.3]{speedup.png}
-  \caption{Geometric average of all speedups---relative to \bq---for each strategy and benchmark, showing that staging leads to better bug-finding speed across the board.
-  }
-  \label{fig:etna}
-\end{figure}
 
@@
 these results give strong evidence that \camlname consistently benefits bug-finding speed, sometimes drastically.
-
+\begin{figure}[h]
+  \includegraphics[scale=0.3]{speedup.png}
+  \caption{Geometric average of all speedups---relative to BQ---for each strategy and benchmark, showing that staging leads to better bug-finding speed across the board.
+  }
+  \label{fig:etna}
+\end{figure}
 \begin{figure}[h]
   \includegraphics[scale=0.31]{swarm.png}
-  \caption{Distribution of individual-task speedups across strategies and benchmarks. Gray dots represent tasks. Plots with extremely low variance have no swarm overlay.}
+  \caption{Distribution of individual-task speedups across strategies and benchmarks.
+Gray dots represent tasks.
+Plots with extremely low variance have no swarm overlay; in plots with high variance, outliers correspond to large speedups.}
   \label{fig:swarm}
@@
 
-\subsubsection{Do staged type-derived generators reduce timeouts?}
-In addition to \camlname's overall bug-finding speed, we evaluate 
-the performance of its type-derived generators---specifically, 
-whether they can transform previously intractable tasks into 
-tractable ones. Out of 78 tasks that initially timed out, 
-50 (64\%) completed successfully after our interventions. 
-This suggests that staged type-derived generators can provide 
-significant performance improvements for free---that is, without
- any additional effort from the user.
 \section{Beyond Strict Functional Languages}
@@
 as opposed to picking up any suitable option off the shelf.
-
 Next, we discuss language-by-language the degree to which the staging-related insights of this paper are portable.
-\begin{itemize}
-  \item \textbf{Racket} is the next target that we intend to test on.
+
+  \textbf{Racket} is the next target that we intend to test on.
   The entire Racket philosophy is intertwined with
@@
   there are at least two \cn{} PBT libraries in Racket.
-  However, the racket macros literature, while extensive, does not concern itself with staging for performance purposes.
+  However, the racket macros literature, while extensive, does not usually concern itself with staging for performance purposes.
   For this reason, we --- the authorship team of this paper, devoid of much Racket expertise --- decided to not use Racket as a test case in this paper.
 
-  \item \textbf{F\#} 
-  % https://learn.microsoft.com/en-us/dotnet/fsharp/language-reference/code-quotations
-
-  \item \textbf{Haskell} is home to the original PBT implementation \cn{}, and
+  \textbf{Haskell} is home to the original PBT implementation \cn{}, and
   so it is natural to ask why we have not built a \name{} in it yet. The basic
@@
 
-  \item \textbf{Rust} shares some similarities with functional languages, and
+  \textbf{F\#} has both multi-stage programming capabilities \cn{} and a well-used PBT library called FsCheck \cn{}.
+  The internals of FsCheck's generator eDSL are very similar to both \bq and ScalaCheck, so we expect that building a \name{}
+  in F\# is primarily an engineering effort.
+  % https://learn.microsoft.com/en-us/dotnet/fsharp/language-reference/code-quotations
+
+  \textbf{Rust} shares some similarities with functional languages, and
   indeed, it is host to a number of property-based testing libraries \cn{}. However Rust does not
@@
   For this reason, staging does not seem directly applicable to any of the PBT libraries in Rust.
-\end{itemize}
 
@@
 
-\subsection{Speeding up Property-Based Testing}
+\paragraph{Speeding up Property-Based Testing}
 
@@
 
-\subsection{Staging eDSLs}
-\jwc{A generator is a parser of
-randomness~\cite{goldsteinParsingRandomness2022}, and people have implemented
-lots of staged parser libraries!}
-\jwc{cite Lionel Parreaux  work, he's on PC}
-\jwc{Macros in Racket}
+\paragraph{Multi-Stage Programming}
+% \jwc{A generator is a parser of
+% randomness~\cite{goldsteinParsingRandomness2022}, and people have implemented
+% lots of staged parser libraries!}
+Staging's roots come from quasiquotation in LISP \cite{bawden99}, where the
+unified representation of data and code allows for sophisticated
+metaprogramming. Quasiquation-style macros were introduced to the ML family of
+languages by Nielson and Nielson's Two-Level Functional Languages~\cn{} and
+Taha's MetaML~\cn{}, which in turn quickly inspired MetaOCaml (and more recently
+MacoCaml) in OCaml~\cn{}, Lightweight Modular Staging in Scala \cn{}, Template
+Haskell in Haskell~\cn{}, and more.  In addition to implementations, researchers
+have studied the type-theoretic foundations of multi-stage languages \cite{cmtt,
+davies-pfenning, davies17}, as well as ways of combining multi-stage types with
+other sophisticated features like dependent types \cite{kovacs24}. Meanwhile,
+the Racket community continues the LISP tradition with a long and fruitful
+line of work studying how macros and metaprogramming can be used to build extensible
+DSLs \cite{flatt12, samth11, skthesis}.
+
+Back in the ML-family world, the primary use of staging is building embedded DSLs
+with minimal performance overhead, an idea which has come to be known as ``abstraction without regret''
+\cite{lms,parreaux17,parreauxthesis, sheard99, trattdsl}.
+This technique has been applied across a range of domains, including
+Big data processing \cite{jet12}, stream processing \cite{moller20,strymonas},
+query processing \cite{rhyme}, and parser combinators \cite{sspc, staged-parsers, krishnaswami19, flap}.
+This work on staged parsers is the closest analogue to ours. Recent PBT research has drawn an explicit link between parsing and generation, framing generators
+as ``parsers of randomness'' \cite{goldsteinParsingRandomness2022}.
+\jwc{Words here about the difference!!}
+
+% \paragraph{Sources of Randomness in Random Testing}
+% \jwc{
+%   \begin{itemize}
+%     \item original QC work used SplitMix \cn{}, to split at binds.
+%     \item the point of 
+%   \end{itemize}
+% }
 

### 2025-03-23 paper/paper.tex
change recursive explanation

--- 2025-03-23 paper/paper.tex
@@
 using \texttt{bind} and aggregating the results; sum types, or variants, are generated using 
-a \texttt{weighted\_union} of the generators for each case; and recursive types are handled by 
-tying the knot at the generator level---recursive references are treated as parameters during 
-derivation, allowing the generator's structure to be defined first and then closed using a 
-fixed-point combinator.
-
+a \texttt{weighted\_union} of the generators for each case; and for recursive types, 
+the entire generator is wrapped in a fixed-point combinator, and a recursive handle 
+is used as the component generator for all recursive occurrences of the type.
 This compositional approach maps naturally to staged generation: 
@@
 executed at runtime to produce values. The entire 
-pipeline remains invisible to the user.
+pipeline remains invisible to the user. \tr{TODO: Boolean list example in a figure...}
 

### 2025-03-23 paper/paper.tex
blah

@@
 the entire generator is wrapped in a fixed-point combinator, and a recursive handle 
-is used as the component generator for all recursive occurrences of the type.
+is used as the generator for all recursive occurrences of the type.
 This compositional approach maps naturally to staged generation: 

### 2025-03-23 paper/paper.tex
nit

@@
 the entire generator is wrapped in a fixed-point combinator, and a recursive handle 
-is used as the generator for all recursive occurrences of the type.
+is used as the generator for all recursive occurrences in the type.
 This compositional approach maps naturally to staged generation: 

### 2025-03-23 handwritten-rust/benches/gen_bm.rs
oops

@@ -1,79 +1,0 @@
-use bolero_generator::{gen, driver::{*}, ValueGenerator};
-use criterion::{black_box, criterion_group, criterion_main, Criterion};
-
-
-fn bench_v1(c: &mut Criterion) {
-    let mut group = c.benchmark_group("Pair of Integers");
-    
-    group.bench_function("v1 - and_then_gen", |b| {
-        let g = gen::<u32>().and_then_gen(|x: u32| (0..x).map_gen(move|y| (x,y)));
-        b.iter(|| {
-            let mut driver = ByteSliceDriver::new(&[0xF,0xA,3,4,6,7,8,9], &Options::default());
-            g.generate(&mut driver).unwrap()
-        })
-    });
-
-    group.bench_function("v2 - inline bind", |b| {
-        b.iter(|| {
-            let mut driver = ByteSliceDriver::new(&[0xF,0xA,3,4,6,7,8,9], &Options::default());
-            let x = gen::<u32>().generate(&mut driver).unwrap();
-            let y= (0..x).generate(&mut driver).unwrap();
-            (x,y);
-        })
-    });
-
-    group.bench_function("v3 - full direct", |b| {
-        b.iter(|| {
-            let mut driver = ByteSliceDriver::new(&[0xF,0xA,3,4,6,7,8,9], &Options::default());
-            let x = driver.gen_u32(std::ops::Bound::Unbounded,std::ops::Bound::Unbounded).unwrap();
-            let y = driver.gen_u32(std::ops::Bound::Included(&0),std::ops::Bound::Excluded(&x)).unwrap();
-            (x,y);
-        })
-    });
-
-    group.finish();
-
-}
-
-enum List{
-    Nil,
-    Cons(u32,Box<List>),
-}
-
-
-fn bench_list(c: &mut Criterion) {
-    let mut group = c.benchmark_group("List");
-    
-    group.bench_function("v1 - and_then_gen", |b| {
-        let g = gen::<u32>().and_then_gen(|x: u32| (0..x).map_gen(move|y| (x,y)));
-        b.iter(|| {
-            let mut driver = ByteSliceDriver::new(&[0xF,0xA,3,4,6,7,8,9], &Options::default());
-            g.generate(&mut driver).unwrap()
-        })
-    });
-
-    group.bench_function("v2 - inline bind", |b| {
-        b.iter(|| {
[…]

### 2025-03-23 handwritten-rust/src/main.rs

@@ -1,51 +1,0 @@
-
-use std::collections::LinkedList;
-use bolero::TypeGenerator;
-use bolero_generator::{driver::*, gen, one_of, ValueGenerator};
-
-#[derive(Clone)]
-enum LL {
-    Nil,
-    Cons(u32,Box<LL>)
-}
-
-
-impl ValueGenerator for LL {
-    type Output = LL;
-
-    fn generate<D : Driver>(&self, driver : &mut D) -> Option<LL> {
-        Some(self.clone())
-    }
-
-}
-
-fn g<D : Driver>(sz : u32) -> dyn TypeGenerator {
-    let g = gen::<bool>().and_then_gen(|b|
-        if(sz == 0 || b) {
-            LL::Nil
-        } else {
-            g(sz / 2).and_then_gen(|l|
-                LL::Nil
-            )
-        }
-    );
-    return g.generate(driver);
-}
-
-fn main() {
-    let mut driver = ByteSliceDriver::new(&[0xF,0xA,3,4,6,7,8,9], &Options::default());
-    let x = gen::<u32>().generate(&mut driver).unwrap();
-    let y= (0..x).generate(&mut driver).unwrap();
-    println!("{x},{y}");
-    let mut driver = ByteSliceDriver::new(&[0xF,0xA,3,4,6,7,8,9], &Options::default());
-    let (x,y) = gen::<u32>().and_then_gen(|x: u32| (0..x).map_gen(move|y| (x,y)) ).generate(&mut driver).unwrap();
-    println!("{x},{y}");
-
-    // let x = (1..10);
-    // let y = (1..10).map_gen(|x| x + 1)
-    // (1..10).generate
-
-    let x = gen::<LinkedList<u32>>();
-
-    let x = driver.gen_u32(std::ops::Bound::Unbounded,std::ops::Bound::Unbounded);
-}

### 2025-03-23 paper/paper.tex
conclusion and FW

@@
 such as the Lehmer algorithm\cn, which runs almost 2x as fast as SplitMix in microbenchmarks \cn.
+\jwc{also Wyhash, xoroshiro128plus}
 % https://github.com/lemire/testingRNG/tree/master
+% https://github.com/wangyi-fudan/wyhash?tab=readme-ov-file
 Moreover, PBT libraries could even consider eschewing the requirement that a source of
@@
 \jwc{This actually isn't the case. In the intro, we're very careful to say that we show that choosing a faster randomness library \emph{leads to faster bugfinding}. And we do demonstrate this.}
+\jwc{Ok --- we could get rid of this ``source of entropy'' business and just write that you can speed up any of these by pipelining your sampling --- sampling and saving multiple values at a time with SIMD instead of just one each.}
 
@@
 \label{section:other-langs}
+\jwc{Nother name for this section needed, it's mostly about strict functional languages}
 
@@
 \section{Conclusion \& Future Work}
-\jwc{Why we stopped where we stopped: can you squeeze more out of this?}
-\jwc{
-  \begin{itemize}
-    \item On the staging front, we could do a little with more clever staging to codegen better...
-Avoided using floating let-insertion to ensure we had precise control over effect ordering, but we could try to relax that.
-          Recursive generators have some extra overhead for packing/unpacking accumulators, but could do a GADT cf \texttt{SOP} in 2ltt.
-    \item On the RNG front, there's not much more we can squeeze out of the RNG without breakning equiv with splitmix.
-As discussed, there are much faster things you can do as alterative sources of entropy.
-  \end{itemize}
-}
+In this paper, we identified, studied, and proposed potential solutions to two
+important sources of inefficiency in PBT generator libraries: abstraction
+overhead and choice of \rand{}.  In the future, we hope to continue to investigate and push the boundaries of
+PBT generator performance.
+
+On the abstraction overhead side, we hope to employ some of the many tricks for better code generation that have been written
+about in the staging literature. For example, MetaOCaml includes a primitive floating let-insertion \cn{} that yields optimal let placement,
+which we did not use in \camlname{}.
+Another trick we hope to use is GADT-based techniques for unpacking the states of recursive functions. Currently, the \camlname{}
+recursion combinator introduces some overhead from boxing and unboxing the accumulator values at each recursive call. As noticed in other work \cite{kovacs24}, this
+overhead can be eliminated using a type-level heterogenous list to represent the state at compile time.
+
+On the \rands{} side we plan to investigate fast \rands{} that are not equivalent to SplitMix, such as Lehmer~\cn, WyHash~\cn,
+and variants of xoroshiro~\cn. We also want to try techniques to speed up sampling such as pipelining, using SIMD instructions,
+or even generating random numbers ahead of time. Last, we hope to investigate the degree to which actual statistical randomness
+matters in PBT.
 

### 2025-03-23 paper/paper.tex
edits to 2.2, addressing concerns

@@
 they are not chosen with performance in mind, leading to worse bugfinding power than what is possible.
-
 \iflater\hg{This paragraph feels a bit clumsy right now. What if we started with a table
@@
 
-\jwc{This needs to be smith'd}
 For example, significantly faster algorithms than SplitMix or JSF exist,
-such as the Lehmer algorithm\cn, which runs almost 2x as fast as SplitMix in microbenchmarks \cn.
-\jwc{also Wyhash, xoroshiro128plus}
+such as the Lehmer algorithm\cn, WyHash \cn, and the xoroshiro family.
+These all run between 1.2-1.6x faster per byte than SplitMix in microbenchmarks \cn.
+Moreover, plenty of other optimizations on top of algorithm choice exist,
+including pipelined or even ahead-of-time sampling.
 % https://github.com/lemire/testingRNG/tree/master
 % https://github.com/wangyi-fudan/wyhash?tab=readme-ov-file
-Moreover, PBT libraries could even consider eschewing the requirement that a source of
-entropy pass statistical tests like BigCrush\cn. \hg{This needs more discussion}
-This is common practice in other areas of testing already:
-fuzzers often simply use a buffer full of arbitrary bytes as a source of entropy
-\cn. Such approaches are also faster than algorithms like SplitMix: bumping a
-pointer and reading from memory (which can be pipelined trivially) will always
-be faster on modern CPUs.
-\hg{Should we just always be doing this? I think we want to argue
-that this is a bridge too far because you still need to generate the buffer
-ahead of time and you can run out of randomness?}
-\jwc{HG: talk about the tradeoffs here.}
-\jwc{The point here is that ``you should be thinking about using a more performant RNG, we'll show you that it does have real benefits (and quantify them). But there are tradeoffs here.'}
-\bcp{But this is not the point that the reader is set up to expect: we've toldthem that we're going to make generation faster, not that we're going to give them tips so that htey can make it faster by themselves}
-\jwc{This actually isn't the case. In the intro, we're very careful to say that we show that choosing a faster randomness library \emph{leads to faster bugfinding}. And we do demonstrate this.}
-\jwc{Ok --- we could get rid of this ``source of entropy'' business and just write that you can speed up any of these by pipelining your sampling --- sampling and saving multiple values at a time with SIMD instead of just one each.}
+
+% Moreover, PBT libraries could even consider eschewing the requirement that a source of
+% entropy pass statistical tests like BigCrush\cn. \hg{This needs more discussion}
+% This is common practice in other areas of testing already:
+% fuzzers often simply use a buffer full of arbitrary bytes as a source of entropy
+% \cn. Such approaches are also faster than algorithms like SplitMix: bumping a
+% pointer and reading from memory (which can be pipelined trivially) will always
+% be faster on modern CPUs.
+% \hg{Should we just always be doing this? I think we want to argue
+% that this is a bridge too far because you still need to generate the buffer
+% ahead of time and you can run out of randomness?}
+% \jwc{HG: talk about the tradeoffs here.}
+% \jwc{The point here is that ``you should be thinking about using a more performant RNG, we'll show you that it does have real benefits (and quantify them). But there are tradeoffs here.'}
+% \bcp{But this is not the point that the reader is set up to expect: we've toldthem that we're going to make generation faster, not that we're going to give them tips so that htey can make it faster by themselves}
+% \jwc{This actually isn't the case. In the intro, we're very careful to say that we show that choosing a faster randomness library \emph{leads to faster bugfinding}. And we do demonstrate this.}
+% \jwc{Ok --- we could get rid of this ``source of entropy'' business and just write that you can speed up any of these by pipelining your sampling --- sampling and saving multiple values at a time with SIMD instead of just one each.}
 
@@
 generation; for that, we need an experiment. The most obvious experiment is to
-simply swap out the randomness library of either \bq{} or ScalaCheck with a totally different, faster one.
+simply swap out the randomness library of either \bq{} or ScalaCheck one of the afformentioned faster algorithms.
 But, as discussed previously, \camlname{} and \scalaname{} are designed to be 100\% semantically equivalent replacements for
@@
 To get around this, we exploit a coincidence:
-\texttt{Splittable\_random}---the \rand{} that \bq{} uses---is slow in a way that
-can be improved \emph{without} changing its behavior.
+\texttt{Splittable\_random}---the OCaml implementation of SplitMix that \bq{} uses---is slow in a way that
+can be improved \emph{without} changing the algorithm.
 In particular,
@@
 allocate memory cells to contain their output, which has a significant performance benefit.
-By building a version that uses much faster ``unboxed'' 64-bit integer arithmetic, we can demonstrate just how much faster bugs can be found just by using a more performant randomness library.
+By building a version that uses much faster ``unboxed'' 64-bit integer
+arithmetic, we can (a) demonstrate just how much faster bugs can be found just by
+using a more performant \rand{}, and (b) explore which generators benefit the most from faster \rand{}.
 

### 2025-03-23 paper/paper.tex
further fixes on rng section

@@
 For example, significantly faster algorithms than SplitMix or JSF exist,
-such as the Lehmer algorithm\cn, WyHash \cn, and the xoroshiro family.
-These all run between 1.2-1.6x faster per byte than SplitMix in microbenchmarks \cn.
-Moreover, plenty of other optimizations on top of algorithm choice exist,
+such as the Lehmer algorithm~\cite{lehmerrng}, WyRand~\cite{wyrand}, and the xorshiro family of algorithms~\cite{xorshiro}.
+These all run between 1.2-1.6x faster per byte than SplitMix in microbenchmarks~\cite{lemiremicrobenchmarks}.
+Plenty of other known optimizations on top of algorithm choice could also be implemented,
 including pipelined or even ahead-of-time sampling.

### 2025-03-23 paper/paper.tex
Blah
pass on 3.7

@@
 Scala 3 \cite{scalamacros}, Haskell \cite{templatehaskell}, \jwc{Racket??}, OCaml \cite{metaocaml,macocaml}, and Java \cite{mint}.
-For this paper, we have implemented staged PBT libraries both in OCaml --- using the MetaOCaml fork of the OCaml compiler \cn --- and Scala 3 --- using
+For this paper, we have implemented staged PBT libraries both in OCaml --- using MetaOCaml \cn --- and Scala 3 --- using
 \texttt{scala.quoted}. For presentation purposes, all staged code presented in the body of this paper is \camlname{} code written in OCaml:
@@
 \subsection{Staged Type-Derived Generators}
+\label{subsection:type-derived}
 
@@
 This state of affairs is not ideal, as inexperienced PBT users are liable to derive a generator, run their test harness, find no bugs, and assume the code is correct---when in reality, most values are never tested at all because they do not satisfy the preconditions.
+\jwc{This paragraph reads as mostly a critique of type-deriving, and it was made already in PBT in practice...}
 
@@
 makes it it more likely that even naïve strategies will find a bug.
+\jwc{This is good --- I think we can just say that  type deriving is easier to use but less powerful, and we make it slightly more powerful, since we make it more likely that the naive strategy finds a bug.}
 
@@
 is used as the generator for all recursive occurrences in the type.
+\jwc{A bit of a long sentence....}
 This compositional approach maps naturally to staged generation: 
@@
 with its staged counterpart. The derivation algorithm remains unchanged.
-
-Our implementation in OCaml uses the preprocessor extension---``PPX''---system, but the 
-approach is portable to other languages with metaprogramming capabilities. 
-In Scala, for instance, the same strategy can be realized using typeclass 
+\jwc{Great.}
+
+Our implementation in OCaml uses the preprocessor extension (``PPX'') system \cn, but the 
+approach is portable to other languages with \proposechange{metaprogramming capabilities}{type-derived generators}\jwc{I.e.
+if you can do type-derived generators any way, you can do them with \name{} in the same way}.
+In Scala, for instance, the same strategy \proposechange{can}{could} be realized using typeclass 
 resolution. In both settings, the result is a three-stage process: a 
@@
 pipeline remains invisible to the user. \tr{TODO: Boolean list example in a figure...}
-
-\label{subsection:type-derived}
-\jwc{
-  \begin{itemize}
-    \item In PBT in practice, we learned that in many cases, programmers don't
-    even write custom generators, instead relying on type-derived generators.
-    \item These generators just produce arbitrary values of a given type, not necessarily enforcing validity conditions.
-    \item If
-    \item Let's talk about how type-deriving works. In languages with typeclasses, it works
-    by typeclass resolution. In OCaml it works by PPX system, but the principle
-    is the same. You define rules to go from generators of a subtypes to a generator of the larger type, and then apply those rules to build up a full generator.
-    \item \begin{itemize}
-      \item For base types, you just call the associated generator
-      \item For product types, you sample from all the component gnerators, bind the values, and then tuple up the values, and return the tuple
-      \item For variant types, you use a weighted union to choose one of the component generators with a weighted union.
-      \item For recursive types, wrap the whole thing in a fixedpoint and then use the recusive handle as the ``component generator'' for all recursive instance of the type.
-    \end{itemize}
-    \item This kind of generic deriving of generators works just as well for
-    staged generators. Just replace the standard combinators in question with
-    staged ones!
-    \item We built this in OCaml, but you can easily use typeclasses to do it in scala too.
-    \item Note that this is (essentially) 3-stage metaprograming.
+\jwc{Unclear exactly what we want the boolean list example for here... are we going to show the type and the derived staged generator? I think that's a little too deep in the weeds for this section.}
+
+\jwc{... We implemented this in \camlname, but not yet in \scalaname, and we benchmark the results in Section~\ref{section:eval}}
+
+% \jwc{
+%   \begin{itemize}
+%     \item In PBT in practice, we learned that in many cases, programmers don't
+%     even write custom generators, instead relying on type-derived generators.
+%     \item These generators just produce arbitrary values of a given type, not necessarily enforcing validity conditions.
+%     \item If
+%     \item Let's talk about how type-deriving works. In languages with typeclasses, it works
+%     by typeclass resolution. In OCaml it works by PPX system, but the principle
+%     is the same. You define rules to go from generators of a subtypes to a generator of the larger type, and then apply those rules to build up a full generator.
+%     \item \begin{itemize}
+%       \item For base types, you just call the associated generator
+%       \item For product types, you sample from all the component gnerators, bind the values, and then tuple up the values, and return the tuple
+%       \item For variant types, you use a weighted union to choose one of the component generators with a weighted union.
+%       \item For recursive types, wrap the whole thing in a fixedpoint and then use the recusive handle as the ``component generator'' for all recursive instance of the type.
+%     \end{itemize}
+%     \item This kind of generic deriving of generators works just as well for
+%     staged generators. Just replace the standard combinators in question with
+%     staged ones!
+%     \item We built this in OCaml, but you can easily use typeclasses to do it in scala too.
+%     \item Note that this is (essentially) 3-stage metaprograming.
 You're using either the PPX mechanism or typeclass resolution to generate a bunch of staged combinator calls, which then run at compile time, which then run at run time.
-  \end{itemize}
-}
-\jwc{Todo: Thia}
-
-\subsection{Faster SplitMix with Unboxed \texttt{int64}s}
+%   \end{itemize}
+% }
+
+\subsection{Performance Opportunities in Randomness Libraries}
 \label{subsection:faster-rng}
-
-\jwc{Unclear what we should call this section}
-\bcp{``Performance opportunities in the RNG?''}
+% \jwc{Unclear what we should call this section}
+% \bcp{``Performance opportunities in the RNG?''}
 As we discussed in Section~\ref{subsection:ineff-rand}, choosing an inefficient randomness library is

### 2025-03-23 paper/paper.tex
wibble
through 3.1

@@
 
-\section{What are Generator Libraries, and why are they Slow?}
+\section{What are Generator Libraries, and Why are They Slow?}
 \label{section:motiv}
@@
 
-\section{How can I make my Generator Library Faster?}
+\section{How can I Make My Generator Library Faster?}
 \label{section:faster-generators}
@@
 With a sense of the inefficiences pervasive in generator libraries, we set out in this section to investigate solutions.
-We begin with a quick tour through multi-stage programming, and then incrementally build \camlname, layering on features and functionality.
-Last, we present the controlled experiment we use to demonstrate how faster \rand{} leads to failing faster.
+We begin with a quick tour through multi-stage programming, then incrementally build \camlname, layering on features and functionality.
+Last, we present the controlled experiment we use to demonstrate how a faster \rand{} leads to failing faster.
 
@@
 
-In multi-stage programming (also known more simply as staging), programs execute
+In multi-stage programming (or simply staging), programs execute
 in multiple stages, with each stage producing the code to be run
-in the next. For the purposes of this paper, we will
-consider two stages: namely compile time and run time. In our context, staged programs
-execute once at compile time, which produces more code, which is then
-compiled and run at run time.
+in the next. For the purposes of this paper, we just need two stages: compile time and run time.
 
@@
 \jwc{Cite racket macros for DSLs here:}
-While embedded DSLs are powerful tools, they have well-known drawback:
-the functional abstractions used to build eDSLs prevent compilers from generating efficient machine code \cn.
-This is abstraction overhead \cite{carrette05, moller20}: the layers of abstraction that
-make eDSLs so usable are precisely what prevents them from being fast.
+While embedded DSLs are powerful tools, they have a well-known drawback:
+the functional abstractions used to build eDSLs tend to prevent compilers from generating efficient machine code \cn.
+This is \emph{abstraction overhead} \cite{carrette05, moller20}: the layers of abstraction that
+make eDSLs so pleasant to use are precisely what prevents them from being fast.
 Indeed, the root causes of many of the performance issues we discovered in Section~\ref{section:motiv} are
-not unique to generator DSLs.
+not unique to generator DSLs, but pervasive across eDSLs.
 In light of this issue, staging is often used as a lightweight compiler for DSLs. The compile-time evaluation stage
@@
 Scala 3 \cite{scalamacros}, Haskell \cite{templatehaskell}, \jwc{Racket??}, OCaml \cite{metaocaml,macocaml}, and Java \cite{mint}.
-For this paper, we have implemented staged PBT libraries both in OCaml --- using MetaOCaml \cn --- and Scala 3 --- using
-\texttt{scala.quoted}. For presentation purposes, all staged code presented in the body of this paper is \camlname{} code written in OCaml:
+For this paper, we have implemented staged PBT libraries both in OCaml (using MetaOCaml \cn) and Scala 3 (using
+\texttt{scala.quoted}). All staged code presented in the body of this paper is \camlname{} code written in OCaml:
 \scalaname{} is very similar.
-For further discussion of the potential for building staged generator DSLs in other languages, see Section~\ref{section:other-langs}.
+For further discussion of the potential for staged generator DSLs in other languages, see Section~\ref{section:other-langs}.
 
@@
 of type \texttt{t}.
-Values of \texttt{code} type are introduced by \emph{quotes}, written
-\texttt{.<$\ldots$>.}. Quotes delay execution of a program until run time. For
+Values of \texttt{code} type are introduced by \emph{quotes} (written
+\texttt{.<$\ldots$>.}). Quotes delay execution of a program until run time. For
 example, the program \texttt{.< 5 + 1 >.} has type \texttt{int code}.  Note that
-this is not the same as \texttt{.< 6 >.}. Because brackets delay computation,
+this is not the same as \texttt{.< 6 >.} Because brackets delay computation,
 the code is not \emph{executed} until the next stage (run time).
-Values of type \texttt{code} can be combined together using an \emph{escape},
+Values of type \texttt{code} can be combined using an \emph{escape},
 written \texttt{.\~{}(e)} (or just \texttt{.\~{}x}, when \texttt{x} is a variable).  Escaping lets you take a value of type
-\texttt{code}, and ``splice'' it directly into a quote.  For example, the
+\texttt{code} and ``splice'' it directly into a quote.  For example, the
 program \texttt{let x = .<1 * 5>. in .< .\~{}(x) + .\~{}(x) >.} evaluates to
-\texttt{.<(1 * 5) + (1 * 5)>.}.  MetaOCaml ensures correct scoping and macro
+\texttt{.<(1 * 5) + (1 * 5)>.}  MetaOCaml enforces correct scoping and macro
 hygiene, ensuring that variables are not shadowed when open terms are spliced.
@@
 defining functions that accept and return \texttt{code} values.  A function
-\texttt{f : 'a code -> 'b code} is a function that takes a program computing a
+\texttt{f : 'a code -> 'b code} takes a program computing a
 run-time \texttt{'a} and transforms it into a program computing a run-time
 \texttt{'b}. In particular, because \texttt{f} itself runs at compile time, the
-fact that the programmer called \texttt{f} does not matter at run time --- the abstraction
+fact that the programmer called \texttt{f} does not matter at run time---the abstraction
 that \texttt{f} defines has been eliminated.
 A code-transforming function \texttt{'a code -> 'b code} can also be converted \emph{code for a function} --- a value of type
-\texttt{('a -> 'b) code} --- with the following program:
+\texttt{('a -> 'b) code}---with the following program:
 \begin{lstlisting}
@@
 
-This program is known as ``the trick'' in the partial evaluation and multi-stage programming literature.
-\cn{}
+This program is known as ``the trick'' in the partial evaluation and multi-stage programming literature \cn{}.
 % USE THIS: https://arxiv.org/pdf/2309.08207
@@
 
-The trick is best illustrated by an example. The following program reduces to \texttt{.< fun x -> (1 + x) mod 2 == 0 >}.
+The trick is best illustrated by an example. The following program reduces at compile time to \texttt{.< fun x -> (1 + x) mod 2 == 0 >.}:
 \begin{lstlisting}
@@
 By composing the two code-transforming functions together at compile time, and only then turning them into a run-time function,
-the functions are fused together.
-This is the basis of how staging is used to eliminate the abstraction of DSLs
-(like a generator DSL). By writing DSL combinators as compile-time functions ---
-and only calling \texttt{eta} at the end on the completed DSL program ---  we
+the two functions are fused together.
+This is the basis of how staging is used to eliminate the abstraction overhead of DSLs.
+By writing DSL combinators as compile-time functions ---
+and calling \texttt{eta} at the end on the completed DSL program ---  we
 can ensure that any overhead of using the combinators is eliminated by run time.
-\jwc{BCP: You wrote a ``?'' next to this section. What's the issue?}
 
@@
 combinators to do as much as possible at compile time.
-The compile time stage will then produce simplified code, free of any DSL combinators, which can be compiled and run
+The compile-time stage will then produce simplified code, free of any DSL combinators, which can be compiled and run
 with different sizes and seeds.

### 2025-03-23 paper/paper.tex
through 3.5

@@
 combinators to do as much as possible at compile time.
-The compile-time stage will then produce simplified code, free of any DSL combinators, which can be compiled and run
+The compile-time stage will then produce simplified code, free of any DSL abstraction, that can be compiled and run
 with different sizes and seeds.
@@
 combinators are known statically (and can be part of the compile time stage) and
-which parts are only known at run time (and must be \texttt{code}).
-In the staging literature, this task is known as ``binding-time analysis'' and it is more an art than a science.
+which parts are only known at run time (and must be treated as \texttt{code}).
+In the staging literature, this task is known as ``binding-time analysis,'' and it is more art than science.
 
@@
 The crux of our binding time analysis is that the \emph{only} parts of a generator that are not
-known at compile time are the particular random seed and
-size parameters that will be passed in at run time. Generators themselves are always known entirely at compile time.
+known at compile time are the random seed and
+size parameters. In practice, generators themselves are entirely known at compile time.
 This leads us to define our library's generator type \texttt{'a
 Gen.t} as \texttt{int code -> SR.t code -> 'a code}: a compile-time
-function from dynamically-known size and seed to dynamically-determined result.
+function from dynamically known size and seed to dynamically determined result.
 
@@
 returns the generator that runs \texttt{cx}. Similarly,
-\texttt{bind g k} sequences generators as usual by passing the result of running the generator \texttt{g} to a continuation \texttt{k}. However, instead of getting access
+\texttt{bind g k} sequences generators by passing the result of running the generator \texttt{g} to the continuation \texttt{k}. However, instead of getting access
 to the particular value generated by \texttt{g}, the continuation \texttt{k} gets access to \texttt{code} for the value sampled from \texttt{g}:
@@
 (2) runs the spliced-in continuation \texttt{k}.
-Both function applications \texttt{g size\_c random\_c}
-and \texttt{k .<a>. size\_c random\_c} run at compile time.
+Both function applications \texttt{g size random}
+and \texttt{k .<a>. size random} run at compile time.
 \texttt{Gen.int} is the generator that samples an int from the randomness library.
@@
 might not be known at compile time --- they may themselves be the results of calling \texttt{Gen.int} ---
-the arguments \texttt{lo} and \texttt{hi} are of type \texttt{int code}, and get spliced into the code block as arguments to \texttt{SR.int}.
+the arguments \texttt{lo} and \texttt{hi} are of type \texttt{int code} and get spliced into the code block as arguments to \texttt{SR.int}.
 Lastly, \texttt{to\_bq} turns a staged generator into code for a normal \bq\
-generator. This function is just a 2-argument version of ``The Trick'' (\texttt{eta}
-from Section~\ref{subsection:msp}).
+generator. This function is just a 2-argument version of ``The Trick'' 
+from Section~\ref{subsection:msp}.
 \iflater
@@
 
-  let return (cx : 'a code) : 'a t = fun size_c random_c -> cx
+  let return (cx : 'a code) : 'a t = fun size random -> cx
 
@@
 allocate lists in the hot path of the generator. Even though these lists are usually small ---
-at most a few dozen elements --- each allocation takes us closer to the next garbage collection.
+at most a few dozen elements --- each allocation takes us closer to the next garbage collection, which is bad for performance.
 
@@
 Instead of building a data structure representing a histogram of the distribution described by the weights at run time and then traversing it,
-the compile time \texttt{weighted\_union} combinator generates a
-tree of \texttt{if}s specialized to the list of weights known at compile time, that searches for the selected generator.
+the compile-time \texttt{weighted\_union} combinator generates a
+tree of \texttt{if}s, specialized to the list of weights known at compile time, that picks out the selected generator.
 
@@
 Figure~\ref{fig:staged-weighted-union-example} demonstrates a use of this staged weighted union.
-Given a list (in this case constant) generators with weights ``the current size parameter'', \texttt{2}, and \texttt{1},
-the generated code first computes the sum of these numbers, samples between \texttt{0} and the sum, and then
+Given a list of (in this case constant) generators with weights ``the current size parameter'', \texttt{2}, and \texttt{1},
+the generated code first computes the sum of these numbers, samples between \texttt{0} and the sum, and
 traverses the tree of three \texttt{if}s to find the correct value to return.
@@
 
-In Figure~\ref{fig:codecps-and-final-gen}, we follow prior work \cite{kovacs24, carrette05} and define
+In Figure~\ref{fig:codecps-and-final-gen}, following prior work \cite{kovacs24, carrette05}, we define
 the type \texttt{'a CodeCps.t = 'z. ('a -> 'z code) -> 'z code}: a polymorphic continuation transformer with the result type
-always in \texttt{code}\footnote{This is an instance of the \emph{codensity} monad \cite{janis08}, a fact which deserves further investigation.}.
+always in \texttt{code}\footnote{This is an instance of the \emph{codensity} monad \cite{janis08}---a fact that deserves further investigation.}.
 The monad instance for this type is the standard instance for a CPS monad with polymorphic return type.
@@
 
-We can then redefine\footnote{
-The type \texttt{'a Gen.t} from before is like \texttt{'a code Gen.t} with this new definition. This type change
+We can then redefine
+our staged generator monad type to be
+\texttt{'a Gen.t = int code -> SR.t code -> 'a CodeCPS.t}, as shown in Figure~\ref{fig:codecps-and-final-gen}.
+The (old) type \texttt{'a Gen.t} is now written as the (new) type \texttt{'a code Gen.t}, and this type change
 carries through all of our combinators. For example, \texttt{Gen.int} now returns \texttt{int code Gen.t}.
-} our staged generator monad type to be
-\texttt{'a Gen.t = int code -> SR.t code -> 'a CodeCPS.t}, as shown in Figure~\ref{fig:codecps-and-final-gen}.
-\jwc{any types that were \texttt{'a Gen.t} before are now \texttt{'a code Gen.t}}.
 
@@
 that a program like \texttt{Gen.bind Gen.int (fun x -> ...)} generates a let-binding for the result of sampling the randomness library.
-For example, \texttt{bind (int .<0>. .<1.>) (fun cx -> return .<(.~cx,.~cx)>.)} now correctly generates
-\texttt{.< fun size random -> let x = SR.int random 0 1 in (x,x) >.}.
+For example, \texttt{bind (int .<0>. .<1>.) (fun cx -> return .<(.~cx,.~cx)>.)} now correctly generates
+\texttt{.< fun size random -> let x = SR.int random 0 1 in (x,x) >.}
 This design is less obviously correct, and does require some care. Rather than \texttt{bind} ensuring correct evaluation order once and for all,
-individual combinators must be carefully written to ensure that \texttt{'a code} values that contain effects are \texttt{let\_insert}'d.
-To ensure that the library is written correctly, we built a PBT harness to compare staged generators to their \bq equivalents over 1,000 random seeds.
+individual combinators must be carefully written to ensure that \texttt{'a code} values that contain effects are \texttt{let\_insert}ed.
+To validate the library, we built a PBT harness to compare staged generators to their \bq equivalents over 1,000 random seeds.
 By differentially testing \cn{} a large suite of generators in this way, we gained confidence that \camlname is equivalent to \bq{}.
@@
 \end{lstlisting}
-\caption{CodeCPS and The Final Gen Monad}
+\caption{CodeCPS and the Final Gen Monad}
 \label{fig:codecps-and-final-gen}

### 2025-03-23 paper/paper.tex
done through 3.7

@@
 \subsection{Recursive Generators}
-Generating recursive datatypes requires recursive generators. Different generator DSLs handle defining recursive generators differently. Some allow recursive generators
+Generating values of recursive datatypes requires recursive generators. Different generator DSLs support recursive generators differently. Some allow recursive generators
 to be defined as recursive functions, while others
@@
 To this end, we develop a staged recursive generator combinator\footnote{
-In reality, we actually have a more general API that allows programmers to define \emph{parameterized}
-recursive combinators, of type \texttt{'r code -> 'a code Gen.t}, for any type \texttt{'r}. See Appendix~\jwc{appendix} for details.
+We actually provide a more general API that allows programmers to define \emph{parameterized}
+recursive combinators, of type \texttt{'r code -> 'a code Gen.t}, for any type \texttt{'r}.
 }, whose API is shown in Figure~\ref{fig:staged-recursive-generator}.
-The code for this combinator can be found in Appendix~\jwc{appendix}.
 The recursion API consists of an opaque type \texttt{'a handle}, and a function \texttt{recurse} to perform recursive calls.

### 2025-03-24 paper/paper.tex
later is now
transpose

@@
 \newif\ifdraft\drafttrue{}
-\newif\iflater\laterfalse{}
+\newif\iflater\latertrue{}
 
@@
 \footnote{The common term for such an algorithm or library is a ``Random Number
-Generator'' (RNG) we will avoid this term and instead say ``randomness library''
+Generator'' (RNG). We will avoid this term and instead say ``randomness library''
 to avoid confusing RNG implementations with the PBT gnenerator libraries that
@@
 they are not chosen with performance in mind, leading to worse bugfinding power than what is possible.
-\iflater\hg{This paragraph feels a bit clumsy right now. What if we started with a table
-of the RNGs used by a bunch of popular PBT frameworks and then argued that
-things like Lehmer are faster?}\fi
 
@@
 
-\jwc{This paragraph is a first draft, and needs to be edited!!!}
-
 \subsection{Background: Multi-Stage Programming}
@@
 \begin{lstlisting}
-let eta (f : 'a code -> 'b code) : ('a -> 'b) code = .<fun x -> .\~(f .<x>.)>.
+let eta (f : 'a code -> 'b code) : ('a -> 'b) code = .<fun x -> .~(f .<x>.)>.
 \end{lstlisting}
 
-This program is known as ``the trick'' in the partial evaluation and multi-stage programming literature \cn{}.
+This program is known as ``The Trick'' in the partial evaluation and multi-stage programming literature \cn{}.
 % USE THIS: https://arxiv.org/pdf/2309.08207
@@
     fun size random ->
-      CodeCps.bind (g size random) (fun x ->
-        (f x) size random
-      )
+      CodeCps.bind (g size random) (fun x -> (f x) size random)
 
@@
 
-Note that while that the insight that faster sampilng translates to faster bug finding
-is a cross-language one, the performance intervention described here is OCaml-specific.
+Note that the performance intervention described here is OCaml-specific.
 In most other PBT frameworks \cn, the randomness library used operates on machine
-integers, so this \emph{particular} inefficiency does not exist.
-
-The precise details of how the SplitMix algorithm works \cn are not important for
-present purposes; the key fact is that all of its operations are are
+integers, so this \emph{particular} inefficiency does not exist. \bcp{This paragraph should go on to explain why it is still interesting to talk about or measure it.}
+However, the insights that we will derive from this experiment (in Section~\ref{section:eval})---
+about which kinds of generators benefit the most from faster randomness and how it leads to faster bug-finding---
+are applicable to all languages.
+\jwc{How's that}
+
+The precise details of the SplitMix algorithm~\cn are not important for
+present purposes; the key fact is that all of its operations are
 defined in terms of arithmetic and bitwise operations on 64-bit integers. In OCaml,
@@
 library times. Each call into \texttt{Splittable\_random}
-may sample many times from the core SplitMix sampling routine \texttt{next\_int64}, i.e. using rejection sampling to find a value within a range.
+may sample many times from the core SplitMix sampling routine \texttt{next\_int64}, i.e., using rejection sampling to find a value within a range.
 Finally, each call to \texttt{next\_int64} allocates 9 times, and each allocation brings us closer
@@
 a large performance impact on generators that spend most of their time sampling data.
-To avoid this allocation, we
+To avoid these allocations, we
 reimplement SplitMix in C and call out to it with the OCaml FFI. The C version of the library uses
@@
  any additional effort from the user.
-\section{Beyond Strict Functional Languages}
+
+\section{Beyond OCaml and Scala}
 \label{section:other-langs}
-\jwc{Nother name for this section needed, it's mostly about strict functional languages}
 

### 2025-03-24 paper/paper.tex
tentative subtitleg
type deriving

@@
 \title{Fail Faster}
-\subtitle{\hg{TODO: Subtitle}}
+\subtitle{Multi-Stage Programming and Fast Randomness for High-Performance PBT}
 
@@
 
-\subsection{Staged Type-Derived Generators}
+\subsection{Staging Type-Derived Generators}
 \label{subsection:type-derived}
 
-Generators are traditionally handwritten, but some libraries allow users to synthesize
+Generators are traditionally handwritten, but some PBT libraries allow users to synthesize
 them automatically from type definitions. Type-derived generators are
 convenient---the derivation process requires no manual effort---but also limited:
-they are unable to account for constraints not encoded in the type (e.g., they can derive a
-tree, but not a binary search tree).
-As a result, generating values with complex preconditions is statistically unlikely.
-This state of affairs is not ideal, as inexperienced PBT users are liable to derive a generator, run their test harness, find no bugs, and assume the code is correct---when in reality, most values are never tested at all because they do not satisfy the preconditions.
-\jwc{This paragraph reads as mostly a critique of type-deriving, and it was made already in PBT in practice...}
-
-This makes type-derivation a particularly elegant target of \name. Since the process of type derivation is completely automatic,
-the end user does not need to understand the staging syntax---but more importantly, \name
-makes it it more likely that even naïve strategies will find a bug.
-\jwc{This is good --- I think we can just say that  type deriving is easier to use but less powerful, and we make it slightly more powerful, since we make it more likely that the naive strategy finds a bug.}
+they are unable to account for constraints not encoded in the type. For example, they can 
+generate arbitrary trees, but not binary search trees. When such constraints are present, 
+type-derived generators are usually less effective than hand-crafted ones, since most 
+generated values will be invalid.
+
+The speed of generation becomes particularly important in
+this setting. Faster generators will be able to explore more of the search 
+space, produce more valid inputs to the system
+under test, and uncover more bugs. Additionally,
+since type derivation is a fully automatic process, the user is not required to understand 
+staging in any way.
+
+% As a result, producing values with complex preconditions is statistically unlikely.
+% This state of affairs is not ideal, as inexperienced PBT users are liable to derive a generator, run their tests, find no bugs, and assume the code is correct---when in reality, most generated values are never tested at all because they do not satisfy the preconditions.
+% \jwc{This paragraph reads as mostly a critique of type-deriving, and it was made already in PBT in practice...}
 
 The type-deriving algorithm follows a compositional pattern. Generators for complex types 
-are synthesized by structurally composing generators of their subtypes: 
+are synthesized by structurally composing generators for their subtypes: 
 base types are mapped to primitive generators included in the generator library;
-product types---such as tuples and records---are handled by sampling each component 
+product types such as tuples and records are handled by sampling each component 
 using \texttt{bind} and aggregating the results; sum types, or variants, are generated using 
-a \texttt{weighted\_union} of the generators for each case; and for recursive types, 
-the entire generator is wrapped in a fixed-point combinator, and a recursive handle 
+a \texttt{weighted\_union} of the generators for each case; for recursive types, 
+the entire generator is wrapped in a fixed-point combinator and a recursive handle 
 is used as the generator for all recursive occurrences in the type.
@@
 with its staged counterpart. The derivation algorithm remains unchanged.
-\jwc{Great.}
-
-Our implementation in OCaml uses the preprocessor extension (``PPX'') system \cn, but the 
-approach is portable to other languages with \proposechange{metaprogramming capabilities}{type-derived generators}\jwc{I.e.
-if you can do type-derived generators any way, you can do them with \name{} in the same way}.
-In Scala, for instance, the same strategy \proposechange{can}{could} be realized using typeclass 
+
+Our implementation in OCaml uses the PPX (PreProcessor eXtension) system to synthesize 
+staged generators from type definitions. However, any language that supports type-derived 
+generators can implement a staged version using a similar implementation to the original. In Scala, 
+for instance, the same strategy could be realized using typeclass 
 resolution. In both settings, the result is a three-stage process: a 
@@
 executed at runtime to produce values. The entire 
-pipeline remains invisible to the user. \tr{TODO: Boolean list example in a figure...}
-\jwc{Unclear exactly what we want the boolean list example for here... are we going to show the type and the derived staged generator? I think that's a little too deep in the weeds for this section.}
-
-\jwc{... We implemented this in \camlname, but not yet in \scalaname, and we benchmark the results in Section~\ref{section:eval}}
+pipeline remains invisible to the user. 
+% \tr{TODO: Boolean list example in a figure...}
+% \jwc{Unclear exactly what we want the boolean list example for here...
+are we going to show the type and the derived staged generator?
+I think that's a little too deep in the weeds for this section.}
+
+% \jwc{... We implemented this in \camlname, but not yet in \scalaname, and we benchmark the results in Section~\ref{section:eval}}
 

### 2025-03-24 paper/paper.tex
type deriving

@@
 is used as the generator for all recursive occurrences in the type.
-\jwc{A bit of a long sentence....}
 This compositional approach maps naturally to staged generation: 

### 2025-03-24 paper/paper.tex
3.7

@@
 
-The speed of generation becomes particularly important in
-this setting. Faster generators will be able to explore more of the search 
-space, produce more valid inputs to the system
-under test, and uncover more bugs. Additionally,
-since type derivation is a fully automatic process, the user is not required to understand 
-staging in any way.
-
-% As a result, producing values with complex preconditions is statistically unlikely.
-% This state of affairs is not ideal, as inexperienced PBT users are liable to derive a generator, run their tests, find no bugs, and assume the code is correct---when in reality, most generated values are never tested at all because they do not satisfy the preconditions.
-% \jwc{This paragraph reads as mostly a critique of type-deriving, and it was made already in PBT in practice...}
+The speed of generation becomes particularly important in this setting. In a generator 
+that produces only valid values, only a subset of them will trigger a bug; in 
+a type-derived generator, however, only a subset of generated values will be valid, 
+and only a subset of \textit{those} will find a bug. As a result, many more values must be produced, 
+making it important to produce them as quickly as possible.
 
@@
 time to produce a specialized generator; and finally, the generator is 
-executed at runtime to produce values. The entire 
-pipeline remains invisible to the user. 
+executed at runtime to produce values.
+
 % \tr{TODO: Boolean list example in a figure...}

### 2025-03-24 paper/paper.tex
blah
eval feedback

@@
 and only a subset of \textit{those} will find a bug. As a result, many more values must be produced, 
-making it important to produce them as quickly as possible.
+making it important to do so as quickly as possible.
 
@@
 
-This section presents our experimental setup, and addresses the following research questions:
+This section presents our experimental setup and addresses the following research questions:
 
@@
 and OpenJDK 21.0.6, using ScalaCheck version 1.17 as the baseline.
-We used \texttt{core\_bench} 0.16 \cn in OCaml and \texttt{jmh} 0.4.7 \cn for performance microbenchmarking. \tr{Etna?}
+We used \texttt{core\_bench} 0.16 \cn in OCaml and \texttt{jmh} 0.4.7 \cn for performance microbenchmarking. For assessing and comparing PBT techniques, we used the Etna\cn platform.
 
@@
 \label{subsection:benchmarking}
-To answer \textbf{RQ1}, we microbenchmark generators, comparing generators in \camlname to semantically identical ones in \bq,
+To answer \textbf{RQ1}, we microbenchmark generators, comparing generators written in \camlname to semantically identical ones in \bq 
 and generators in \scalaname to those in ScalaCheck.
@@
 \texttt{splittable\_random} and \csplitmix, as discussed
-in \secref{subsection:faster-rng}. Our test cases consist of generators for boolean lists, binary search trees (BST), and simply
-typed lambda calculus (STLC) terms. We implement generators for these benchmarks using a
-variety of \textit{strategies} varying in structure and
+in \secref{subsection:faster-rng}. Our test cases consist of generators for boolean lists, binary search trees (BSTs), and simply
+typed lambda-calculus (STLC) terms. We implement generators for these benchmarks using a
+variety of \textit{strategies}, varying in structure and
 sophistication.
 
-\subsubsection{\camlname}
-\label{subsubsection:ocaml}
-
-In \camlname, our strategies include type-derived generators for BSTs and STLC terms, 
+In \camlname, our strategies include: type-derived generators for BSTs and STLC terms, 
 following the approach described in \secref{subsection:type-derived}; two custom 
@@
 \camlname version using \csplitmix. For each treatment, we measure the
-time to generate a value (i.e., a BST, a STLC term, or a boolean list), using a random seed, varying generation sizes (10, 100, 1,000, 10,000).\footnote{The way size is used in a generator is a domain-specific implementation detail---in a list generator, size might correspond to the desired length of the list, whereas in a tree generator it refer to number of nodes, number of leaves, depth of tree, etc.
-Regardless of how size is encoded, our goal is to show that performance trends scale.}
-We run each treatment for 5 seconds---enough for several thousand generations---and compute the average generation time of a value produced in that interval.
+time to generate a value (i.e., a BST, STLC term, or boolean list), using a random seed, varying generation sizes (10, 100, 1,000, 10,000).\footnote{The way size is used in a generator is a domain-specific implementation detail---in a list generator, size might correspond to the desired length of the list, whereas in a tree generator it refer to number of nodes, number of leaves, depth of tree, etc.
+Regardless of how size is used, our goal is to show that performance trends scale.}
+We run each treatment for 5 seconds and compute the average generation time of a value produced in that interval.
 
@@
   \caption{Time to generate values of varying sizes using each \camlname strategy.
-Lower lines translate to greater performance benefit.
+Lower is better.
+Both axes are logarithmic.
 BQ is \bq; Staged SR is \camlname using \bq's \texttt{splittable\_random} RNG; Staged CSR is \camlname using \csplitmix.}
@@
 Our results are summarized in \figref{fig:time}.
-Our type-derived \camlname BST generator achieves speedups ranging from $1.30-1.38\times$, which increase
+The type-derived \camlname BST generator achieves speedups ranging from $1.30-1.38\times$, which increase
 dramatically, to $2.13-7.76\times$, when combined with \csplitmix. The insertion-based BST generator sees
-$1.18-1.22\times$ speedups with staging alone, and $3.83-6.31\times$ when also
+$1.18-1.22\times$ speedups with staging alone and $3.83-6.31\times$ when also
 using \csplitmix. The single-pass BST generator benefits more from
@@
 The boolean list generator experiences $2.77-7.47\times$ speedups with
-staging, and its performance changes minimally when
+staging, and its performance changes only minimally when
 combined with \csplitmix ($2.77-7.57\times$). This is likely because sampling
@@
 To test this idea, we use measurable proxies that approximate a generator's reliance on its randomness and generator libraries.
-For the former, we use the number of calls to \texttt{bind}---the central monadic generator function, extensively used both directly and within other combinators.
-For the latter, we count the number of times a generator invokes the SplitMix sampling routine \texttt{next\_int64}.
-
-We select four strategies with disparate speedups: insertion-based BSTs, single-pass BSTs, 
-boolean lists, and well-typed STLC terms. To determine how often \texttt{bind} is invoked, we generate 10,000 values using a fixed generation size of 100, and 
-record the average number of \texttt{bind} calls.
-To determine the number of random samples, we repeat the same test, but use Intel ProcessorTrace \cn{} to capture a 4ms snapshot of processor activity.
-We then count the number of invocations of \texttt{next\_int64} and average this over the number of complete generations in the trace.
+For the former, we count the number of times a generator invokes the SplitMix sampling routine \texttt{next\_int64}. 
+For the latter, we use the number of calls to \texttt{bind}---the central monadic generator function, extensively used both directly and within other combinators.
+
+We select four strategies with disparate speedup profiles: insertion-based BSTs, single-pass BSTs, 
+boolean lists, and well-typed STLC terms. To determine how often \texttt{bind} is invoked, we generate 10,000 values using a fixed generation size of 100 and 
+record the average number of \texttt{bind} calls. To determine the number of random samples, we repeat the same test but use Intel ProcessorTrace \cn{} to capture a 4ms snapshot of processor activity.
+We then count the number of invocations of \texttt{next\_int64} and average this over the number of generator calls in the trace.
 
@@
 
-\figref{fig:binds} summarizes our results. The left plot shows a clear linear 
+\figref{fig:binds} summarizes the results. The left plot shows a clear linear 
 relationship between performance benefit from staging and number of calls 
@@
 
-\subsubsection{\scalaname}
 In \scalaname, we implement a subset of the \camlname
@@
 so those strategies are excluded.} We did not implement an optimized version of ScalaCheck's
-\rand, but our experimental setup was otherwise the same as in \secref{subsubsection:ocaml}.
+\rand, but our experimental setup was otherwise the same as in OCaml.
 
@@
 and unboxing overhead at each step.  Staging eliminates this
-overhead. These results show that \name generalizes well across languages.
+overhead. These results show that the \name approach generalizes well across languages.
 
@@
   \includegraphics[scale=0.41]{scala.png}
- \caption{Time to generate values of varying sizes using each \scalaname strategy. SC is ScalaCheck; Staged is \scalaname.}
+ \caption{Time to generate values of varying sizes using each \scalaname strategy. Lower is better. Both axes are logarithmic. SC is ScalaCheck; Staged is \scalaname.}
  \label{fig:scala}
@@
 To answer \textbf{RQ2}, we evaluate the bug-finding speed of our BST and STLC
-case studies using Etna\cn{}, a platform for assessing and comparing PBT techniques.
+case studies using Etna\cn{}.
 Etna allows users to measure the effectiveness of different generator
 implementations by injecting bugs into the system under test and recording the
-time taken for a relevant property to fail in response. For example, in a tree
-with an ordering property, a bug injection might consist of inserting a value
-into the tree that violates that ordering. Each case study includes a diverse
+time taken for a relevant property to fail in response. Each case study includes a diverse
 set of \textit{tasks}, where a task consists of a specific bug-property pair
-designed to test different aspects of the system (e.g., BST includes tasks that
-test insertion, deletion, and union operations on binary search trees).
+designed to test a specific aspect of the system (e.g., BST includes tasks that
+test insertion, deletion, and union operations).
 In particular, BST has 37 tasks, and STLC has 20. Strategies for BST and STLC
@@
 Etna does not support Scala, so we were unable to test \scalaname's bug-finding speed in Etna, but we
-we see no reason our results in this section would not extend to \scalaname.
+expect the results in this section should extend to \scalaname.
 
@@
 it is theoretically possible to choose a seed such that the \textit{first} value
-produced by a generator discovers the bug, which would elide any differential
-speedup (and likewise if both generators fail to find the bug).  To account for
+produced by a generator discovers the bug, which would eliminate any differential
+speedup, and likewise if both generators fail to find the bug.  To account for
 these cases, we repeat the above process over 30 random seeds---that is, all strategies
-are run on the same seed so that they produce identical sequences of values, and this process
-is repeated 30 times.\footnote{Although the
+run on the same seed so that they produce identical sequences of values, and this process
+is repeated 30 times. Although the
 variance \textit{between} seeds in Etna can be vast, timing results are
 replicable for a \textit{given} seed. Across 1,000 trials using the same seed,
-the observed variance in time-to-failure was less than a nanosecond.}
+the observed variance in time-to-failure was less than a nanosecond.
 
@@
 strategy on all tasks using a 60-second timeout.  If a bug is not found within
-this limit, the task is considered unsolved.  Tasks where all strategies fail to
-find the bug are excluded from our dataset, as they provide no basis for
+this limit, the task is considered ``unsolved.''  We exclude tasks where all strategies fail to
+find the bug from our dataset, as they provide no basis for
 comparison.  Similarly, we exclude tasks where the \bq strategy completes in
@@
 371/1110 (33.42\%) of insertion-based BST's tasks. More sophisticated
-strategies tend to find bugs faster, leading to a higher number of filtered
+strategies tend to find bugs very quickly, leading to a higher number of filtered
 tasks. By applying these filters, we ensure that our reported speedups reflect
@@
 
-\subsubsection{Do improved type-derived generators reduce timeouts?}
 In addition to \camlname's overall bug-finding speed, we evaluate 
@@
 50 (64.1\%) completed successfully after adding \csplitmix. 
-This suggests that staged type-derived generators can provide 
-significant performance improvements for free---that is, without
+This shows that staged type-derived generators can provide 
+significant performance improvements for free, without
  any additional effort from the user.

### 2025-03-24 paper/paper.tex
3.7 nit
nits

@@
 
+We implemented this in \camlname; it is not yet implemented in \scalaname. We benchmark the results in \secref{section:eval}.
+
 % \tr{TODO: Boolean list example in a figure...}
@@
 \camlname version using \csplitmix. For each treatment, we measure the
-time to generate a value (i.e., a BST, STLC term, or boolean list), using a random seed, varying generation sizes (10, 100, 1,000, 10,000).\footnote{The way size is used in a generator is a domain-specific implementation detail---in a list generator, size might correspond to the desired length of the list, whereas in a tree generator it refer to number of nodes, number of leaves, depth of tree, etc.
-Regardless of how size is used, our goal is to show that performance trends scale.}
+time to generate a value (i.e., a BST, STLC term, or boolean list), using a random seed, varying generation sizes (10, 100, 1,000, 10,000).\footnote{The specific meaning of generation size is a domain-specific implementation detail---in a list generator, size might correspond to the desired length of the list, whereas in a tree generator it refer to number of nodes, number of leaves, depth of tree, etc.
+Regardless, our goal is to show that performance trends scale.}
 We run each treatment for 5 seconds and compute the average generation time of a value produced in that interval.
@@
 strategies: the boolean list generator, single-pass BST generator, and
-well-typed STLC term generator.\footnote{By default, ScalaCheck does not support deriving gnenerators from types,
-so those strategies are excluded.} We did not implement an optimized version of ScalaCheck's
+well-typed STLC term generator. The type derived strategies are excluded, since
+\scalaname does not currently support type derivation. We did not implement an optimized version of ScalaCheck's
 \rand, but our experimental setup was otherwise the same as in OCaml.

### 2025-03-24 paper/paper.tex
13x

@@
 % The goal of PBT is to fail tests as fast as possible. \jwc{First cut at ``fail fast''}
-\iflater 
-\bcp{Feels like we're missing a trick in the transition here.  Can we
-  say, ``Because performance is important, PBT libraries are already
-  quite tuned.  [Can we substantiate that??]  But more is
-  possible...'' }
-\fi
+% \iflater 
+% \bcp{Feels like we're missing a trick in the transition here.  Can we
+%   say, ``Because performance is important, PBT libraries are already
+%   quite tuned.  [Can we substantiate that??]  But more is
+%   possible...'' }
+% \fi
 
@@
 flexibility also introduces layered abstractions and indirections that are difficult for
-compilers to optimize\proposecut{, shifting work to run time}.
-Second, the \rand that a generator uses to make choices can
-constitute an large proportion of its running time, magnifying the impact of an inefficient one. \jwc{Object of ``one'' unclear, intended to be ``\rand{}''.}
-\iflater
-\hg{TODO later, I'm not 100\% sure it's clear how this particular question is
-raised by the above. We could strengthen that connection}
-\fi
+compilers to optimize.
+Second, the calls a generator makes to its \rand{} can constitute a large 
+proportion of its running time, magnifying the cost of an inefficient 
+implementation.
 
@@
     We precisely quantify the impact of each optimization separately, and explain how they play distinct yet complementary roles in improving generator performance.
-    Finally, we show that these performance improvements extend to improved bug-finding speed, finding some bugs as much as \todo{\%\%} faster.
+    Finally, we show that these performance improvements extend to improved bug-finding speed, finding some bugs as much as $13\times$ faster.
 \end{enumerate}

### 2025-03-24 paper/paper.tex
abs

@@
 \begin{abstract}
-  \hg{TODO: Abstract}
-  \jwc{Make sure we are clear about our results here. I.e. VERY QUICKLY in this abstract, we need to say ``finds bugs 5x faster''}
+  Property-based testing relies on random input generators to uncover bugs, often under tight time constraints. These generators are built 
+  using generator libraries, which provide expressive combinators for constructing complex inputs. 
+  Although convenient,
+  today's generator libraries suffer from substantial performance costs, limited by abstraction overhead and inefficient randomness libraries.
+  In this paper, we identify and address these bottlenecks. We introduce \name, a multi-stage programming technique that eliminates generator overhead
+  without sacrificing the flexibility and expressiveness of existing generator library functions. We augment leading generator libraries in OCaml and 
+  Scala 3 to use the \name approach, preserving the original library's semantics and enabling precise performance comparisons. We also show that \rand
+  has a measurable impact on generator performance, quantify its contribution, and show that speedups of $5.5\times$ are possible. Together, these contributions
+  lead to up to $13\times$ faster bug finding, showing that high-level generator abstractions need not come at the cost of speed.
 \end{abstract}

### 2025-03-24 paper/paper.tex
abs

@@
   without sacrificing the flexibility and expressiveness of existing generator library functions. We augment leading generator libraries in OCaml and 
-  Scala 3 to use the \name approach, preserving the original library's semantics and enabling precise performance comparisons. We also show that \rand
-  has a measurable impact on generator performance, quantify its contribution, and show that speedups of $5.5\times$ are possible. Together, these contributions
-  lead to up to $13\times$ faster bug finding, showing that high-level generator abstractions need not come at the cost of speed.
+  Scala 3 to use the \name approach, preserving the original library's semantics and enabling precise, pointwise performance comparisons. We also show that \rand
+  has a measurable impact on generator performance, quantify its contribution, and show that speedups of $5.5\times$ are possible. Together, these 
+  improvements lead to bug-finding up to $13\times$ faster---showing that high-level generator abstractions need not come at the cost of speed.
 \end{abstract}

### 2025-03-24 paper/paper.tex
blah

@@
   has a measurable impact on generator performance, quantify its contribution, and show that speedups of $5.5\times$ are possible. Together, these 
-  improvements lead to bug-finding up to $13\times$ faster---showing that high-level generator abstractions need not come at the cost of speed.
+  improvements lead to bug-finding up to $13\times$ faster---showing that high-level generator abstractions need not come at the cost of speed. \tr{Needs a lot of work. No spice.}
 \end{abstract}
@@
 Second, the calls a generator makes to its \rand{} can constitute a large 
-proportion of its running time, magnifying the cost of an inefficient 
+proportion of its run-time, magnifying the cost of an inefficient 
 implementation.

### 2025-03-24 paper/paper.tex
blah

@@
 strategies: the boolean list generator, single-pass BST generator, and
-well-typed STLC term generator. The type derived strategies are excluded, since
-\scalaname does not currently support type derivation. We did not implement an optimized version of ScalaCheck's
+well-typed STLC term generator.\footnote{The type-derived strategies are excluded, since
+\scalaname does not currently support type derivation.} We did not implement an optimized version of ScalaCheck's
 \rand, but our experimental setup was otherwise the same as in OCaml.

### 2025-03-24 paper/paper.tex
staging
wibble
CP

@@
 \begin{abstract}
-  Property-based testing relies on random input generators to uncover bugs, often under tight time constraints. These generators are built 
-  using generator libraries, which provide expressive combinators for constructing complex inputs. 
-  Although convenient,
-  today's generator libraries suffer from substantial performance costs, limited by abstraction overhead and inefficient randomness libraries.
-  In this paper, we identify and address these bottlenecks. We introduce \name, a multi-stage programming technique that eliminates generator overhead
-  without sacrificing the flexibility and expressiveness of existing generator library functions. We augment leading generator libraries in OCaml and 
-  Scala 3 to use the \name approach, preserving the original library's semantics and enabling precise, pointwise performance comparisons. We also show that \rand
-  has a measurable impact on generator performance, quantify its contribution, and show that speedups of $5.5\times$ are possible. Together, these 
-  improvements lead to bug-finding up to $13\times$ faster---showing that high-level generator abstractions need not come at the cost of speed. \tr{Needs a lot of work. No spice.}
+Property-based testing (PBT) relies on generators for random test
+cases, often constructed using embedded domain specific languages,
+which provide expressive combinators for building and composing
+generators.  The effectiveness of PBT depends critically on the speed
+of these generators.
+%
+However, careful measurements show that the generator performance of
+two widely used PBT libraries falls well short of what is possible,
+due principally to (1) the abstraction overhead of their
+combinator-heavy style and (2) suboptimal sources of randomness.
+We characterize, quantify, and address these bottlenecks.
+
+To eliminate abstraction overheads, we propose a technique based on
+multi-stage programming, dubbed \name.  We apply this technique to
+leading generator libraries in OCaml and Scala 3, significantly
+improving performance.
+%
+To quantify the performance impact of the randomness source, we carry
+out a controlled experiment, replacing the randomness in the OCaml PBT
+library with an optimized version.
+%
+Both interventions exactly preserve the semantics of generators,
+enabling precise, pointwise comparisons.
+%
+Together, these improvements make bug finding up to $13\times$ faster.
 \end{abstract}
@@
 
-\section{How can I Make My Generator Library Faster?}
+\section{How Can I Make My Generator Library Faster?}
 \label{section:faster-generators}

### 2025-03-24 paper/paper.tex
CP

@@
 \title{Fail Faster}
-\subtitle{Multi-Stage Programming and Fast Randomness for High-Performance PBT}
+\subtitle{Staging and Fast Randomness for High-Performance PBT}
 
@@
 %
-Together, these improvements make bug finding up to $13\times$ faster.
+Together, these improvements find bugs up to $13\times$ faster.
 \end{abstract}
@@
 %\jwc{``Property based testing is a race against time...''}
-Property-based testing (PBT) is a software testing technique that uses
-random test inputs to validate logical specifications. Recent studies on PBT usage in industry~\cite{inpractice} have shown that
-practitioners often interleave PBT with their software development process, sometimes
-testing their properties every time they save their code. This means that PBT is often given a very short time
-window to find bugs---e.g., between 50 milliseconds and 30 seconds~\cite{inpractice}.
-The faster a test fails, the better.
-% The goal of PBT is to fail tests as fast as possible. \jwc{First cut at ``fail fast''}
-% \iflater
-% \bcp{Feels like we're missing a trick in the transition here.  Can we
-%   say, ``Because performance is important, PBT libraries are already
-%   quite tuned.  [Can we substantiate that??]  But more is
-%   possible...'' }
-% \fi
-
-One important locus for performance improvement
+Property-based testing is a software testing technique that uses
+random test inputs to validate logical specifications. A recent study
+on PBT usage in industry~\cite{inpractice} shows that many
+practitioners run their property-based tests very frequently, and with
+a very short time budget.  The faster the tests fail, the better.
+
+One important opportunity for performance improvement
 is the \textit{generators} that produce random inputs to the properties under
@@
 libraries}, usually expressed as embedded domain-specific languages (eDSLs) %\cntxt{edsls}
-that provide combinators for building and composing generators. However, careful
+that provide combinators for building and composing generators.
+
+However, careful
 measurements show that the performance of existing generator libraries falls far short of what
-is possible.
-
-This is due to two major sources of inefficiency.
-First, the same high-level design approach that gives generator libraries their
-flexibility also introduces layered abstractions and indirections that are difficult for
+is possible,
+for two reasons.
+%
+First, the high-level design approach followed by generator combinator
+libraries introduces layered abstractions  that are difficult for
 compilers to optimize.
@@
 implementation.
-
-In this paper, we study and address these issues.
-First, we show how to use \textit{multi-stage programming}, a classic technique for
-building fast DSLs “without regret”~\cn{}, to eliminate the overhead of
-many common generator abstractions. We refer to this approach as \name.
-% Secondly, we demonstrate the
-% impact of diferent choices of randomness library on generator performance by conducting a
-% controlled comparison.
-Second, we demonstrate that a DSL's choice of \rand{} has a
-dramatic impact on its performance by performing a controlled experiment,
-isolating the performance benefits of faster randomness in a clearly quantifiable manner.
-
-% These insights lay the groundwork for generator libraries that combine high-level
-% design with low-level performance.
-
-% \bcp{It
-% feels a bit grand to call what we've done an ``approach''---it's
-% really just a single performance-enhancing technique, staging, plus
-% the observation that the RNG matters and you should use the fastest
-% one you can.}
-
-To demonstrate the multi-language applicability of \name{}, we implement staged generator DSLs in two strict functional languages: OCaml and Scala 3.
-In OCaml, we build \camlname{}, based on \bq{}---a
-high-quality PBT library in OCaml, authored by software engineers at the trading firm Jane Street \cn{}.
-In Scala 3, we build \scalaname{}, based on ScalaCheck---the language's standard PBT library\cn{}.
+%
+In this paper, we characterize, quantify, and address these issues.
+
+We address the overhead of
+many common generator abstractions using \textit{multi-stage
+  programming} (or {\em staging}), a well-studied technique for
+building fast DSLs “without regret”~\cn{}.  We refer to this approach as \name.
+%
+To demonstrate its generality, we implement staged generator DSLs in
+two strict functional languages: OCaml and Scala 3.  In OCaml, we
+build \camlname{}, based on \bq{}---a high-quality PBT library in
+OCaml, authored by software engineers at the trading firm Jane Street
+\cn{}.  In Scala 3, we build \scalaname{}, based on ScalaCheck---the
+language's standard PBT library\cn{}.
+
+We isolate and quantify the performance benefits of faster randomness,
+showing that a DSL's choice of \rand{} has a dramatic impact on its
+performance.  \bcp{describe the experiment}
+
 In both languages, our libraries preserve the semantics of the original: given the same random seed, generators written using \camlname or \scalaname produce exactly the same sequence of values as equivalent generators using \bq or ScalaCheck, respectively.
 This semantic equivalence enables pointwise---as opposed to distributional---bug-finding comparisons, ensuring that if a \name{} generator finds bugs faster, this is attributable \textit{solely} to staging and \rand optimizations, not luck with seeds.
-% \jwc{BCP thinks this should be a little stronger: ``and not obscured by the probablistic nature of generation'', or something like that.
-``Comparing distributions of speed is hard, comparing pointwise is easier.''.
-This needs to have its own sentence here.}
 
@@
 measuring how quickly various generators detect them.
+\jwc{Beef up this paragraph, explain the results of our eval.}
 

### 2025-03-24 paper/paper.tex
CP

@@
 
-In both languages, our libraries preserve the semantics of the original: given the same random seed, generators written using \camlname or \scalaname produce exactly the same sequence of values as equivalent generators using \bq or ScalaCheck, respectively.
-This semantic equivalence enables pointwise---as opposed to distributional---bug-finding comparisons, ensuring that if a \name{} generator finds bugs faster, this is attributable \textit{solely} to staging and \rand optimizations, not luck with seeds.
+Both interventions preserve semantics on the nose: given the same
+random seed, generators written using \camlname or \scalaname produce
+exactly the same sequence of values as equivalent generators using \bq
+or ScalaCheck.  This semantic equivalence enables pointwise---as
+opposed to distributional---comparisons of bug-finding effectiveness,
+ensuring that, if a \name{} generator finds bugs faster, this is
+attributable \textit{solely} to the staging and \rand interventions,
+not lucky choice of seeds.
 
 We evaluate the optimizations with a series of case studies in each
-generator library, and show that the \name{} generators run faster than their unstaged equivalents.
-Further, we show generators that run faster also find bugs faster
-by running these case studies in the Etna
-platform~\cite{etna}, which simulates real-world PBT usage \iflater\hg{Don't love this phrase, I'll
-think more about how we should talk about Etna}\fi by injecting bugs into a system and
-measuring how quickly various generators detect them.
-\jwc{Beef up this paragraph, explain the results of our eval.}
-
-In summary, we show that PBT generator DSLs incur significant performance costs across languages, and present a general technique for improving their efficiency while preserving their idiomatic abstractions.
-\jwc{needs a bit about RNG --- BCP thoughts?}
-We make the following contributions:
+generator library, and show that the \name{} generators run faster
+than their unstaged equivalents.  Further, we show generators that run
+faster also find bugs faster by running these case studies in the Etna
+platform~\cite{etna}, which injects bugs into a system and measures
+how quickly various generators detect them.  \jwc{Beef up this
+  paragraph, explain the results of our eval.}  \todo{We precisely
+  quantify the impact of each intervention and explain how they play
+  distinct yet complementary roles in improving generator performance.
+}
+
+In summary, we show that PBT generator DSLs incur significant
+performance costs across languages and present two interventions that
+significantly improve their efficiency while preserving their
+idiomatic style.  Concretely, we offer the following contributions:
 \begin{enumerate}
-    \item We identify two key sources of inefficiency in PBT generator libraries, which can significantly impact performance:
-    abstraction overhead and \rand choice.
-    \item We present \name{}, a technique for eliminates the abstraction overhead of generators. We implement \name{} in both OCaml and Scala 3, showcasing its cross-language applicability.
-    \item We demonstrate that writing generators using \name{} and fast \rands{} yields significant performance improvements.
-    We precisely quantify the impact of each optimization separately, and explain how they play distinct yet complementary roles in improving generator performance.
-    Finally, we show that these performance improvements extend to improved bug-finding speed, finding some bugs as much as $13\times$ faster.
+\item We identify two key sources of inefficiency in PBT generator
+libraries---abstraction overhead and choice of \rand---both of which
+significantly impact performance (Section~\ref{section:motiv}).
+\item We present \name{}, a staging technique that eliminates the
+abstraction overhead of generators. We apply \name{} to standard
+generator libraries in both OCaml and Scala 3, showcasing its
+generality (Section~\ref{section:faster-generators}).
+\item We demonstrate that writing generators using \name{} and fast
+\rands{} yields substantial performance improvements, both separately
+and in combination, and we show that these performance improvements
+extend to significantly improved bug-finding speed (Section~\ref{section:eval}).
 \end{enumerate}
 
-In Section~\ref{section:motiv}, we introduce monadic generator DSLs, discuss how they work, and take an example-guided tour through some of the ways that
-they are slower than is possible.
-Then, in Section~\ref{section:faster-generators}, we explain multi-stage programming,
-develop \name{} (using \camlname{} to illustrate), and discuss how we isolate the performance of \rand{} to benchmark their impact on bug-finding speed.
-Section~\ref{section:eval} evaluates all of the previous work, demonstrating the generator performance improvements
-and how they lead to much faster bug finding.
+Section~\ref{section:other-langs} discusses how the \name{}
+technique could be applied to PBT libraries in other
+languages---Racket, F\#, Haskell, and Rust.
+%
+Sections~\ref{section:related} and~\ref{section:concl} present related
+and future work.
 
@@
 
-With a sense of the inefficiences pervasive in generator libraries, we set out in this section to investigate solutions.
-We begin with a quick tour through multi-stage programming, then incrementally build \camlname, layering on features and functionality.
-Last, we present the controlled experiment we use to demonstrate how a faster \rand{} leads to failing faster.
+With a sense of two main inefficiences in generator libraries, we set out in this section to investigate solutions.
+We begin with a quick tour through multi-stage programming (\ref{xx}),
+then incrementally build \camlname, layering on features and
+functionality  (\ref{xx}).
+Last, we present the controlled experiment we use to demonstrate how a
+faster \rand{} leads to failing faster  (\ref{xx}).
 
@@
 
-\paragraph{Uses of Staging in eDSL Construction}
+\paragraph{Staging eDSLs}
 One of the primary uses of staging is in embedded DSL construction \cite{sheard99, lms, trattdsl}.
@@
 \section{Related Work}
+\label{section:related}
 
@@
 \section{Conclusion \& Future Work}
+\label{section:concl}
 In this paper, we identified, studied, and proposed potential solutions to two

### 2025-03-24 paper/paper.tex
fuck you jvn

@@
 Property-based testing is a software testing technique that uses
-random test inputs to validate logical specifications. A recent study
+random test inputs to validate logical specifications~\cite{quickcheck00}. A recent study
 on PBT usage in industry~\cite{inpractice} shows that many
@@
   programming} (or {\em staging}), a well-studied technique for
-building fast DSLs “without regret”~\cn{}.  We refer to this approach as \name.
+building fast DSLs “without regret”~\cite{lms}.  We refer to this approach as \name.
 %
@@
 build \camlname{}, based on \bq{}---a high-quality PBT library in
-OCaml, authored by software engineers at the trading firm Jane Street
-\cn{}.  In Scala 3, we build \scalaname{}, based on ScalaCheck---the
-language's standard PBT library\cn{}.
+OCaml, authored by software engineers at the trading firm Jane Street~\cite{bq}.  In Scala 3, we build \scalaname{}, based on ScalaCheck---the
+language's standard PBT library~\cite{sc}.
 
@@
 The standard design for a generator DSL, introduced in Haskell's QuickCheck
-library and copied in dozens of other frameworks,
-is via an embedded {\em monadic} language. We use the syntax and types from OCaml's \bq library, which is presented in
+library~\cite{quickcheck00} and copied in dozens of other frameworks,
+is via an embedded {\em monadic} language~\cite{moggi91}. We use the syntax and types from OCaml's \bq library, which is presented in
 Figure~\ref{fig:bq-api}, but similar DSLs can also be found in languages
@@
 
-% \subsubsection{Subtleties of Bind}
-% The \texttt{bind} function in monadic generator libraries sequences generators
-% with a call-by-value semantics, regardless of the calling convention of the
-% language in which it is implemented. Concretely, this means that sampling
-% \begin{center}
-%   \texttt{let\%bind x = gen\_int 0 10 in return (x + x)}
-% \end{center}
-% always produces an even number; \texttt{x} is only sampled once (when the
-% \texttt{bind} is evaluated) and from that point on it is a pure integer value. This is
-% in contrast to the semantics of some probabilistic programming languages that
-% may have call-by-name sampling semantics \cn{}. This is important because \hg{Joe why are we
-% talking about this here? What are we going to use this fact for?}.
-% \jwc{Well this is why directly inlining the binds doesn't work.}
-
-% Another subtle aspect of \texttt{bind} is that many implementations mutate the
-% internal state of the random number generator\footnote{The original QuickCheck
-% implementation in Haskell uses a splittable random number generator~\cn, and
-% \texttt{bind} splits the seed.} This means that generators do not always obey
-% equivalences that one might expect from a monadic program. For example,
-% \begin{center}
-%   \texttt{let\%bind \_x = return () in gen\_int 0 10}
-% \end{center}
-% does not necessarily produce the same sequence of values as \texttt{gen\_int 0
-% 10}; the \texttt{bind} updates the random seed in the background before
-% geenrating the integer, even though it looks like a no-op. Critically, while
-% these generators do not produce the same sequence of values, they do produce the
-% same {\em distributions} of values. We say that generators obey standard monadic
-% equalities up to ``distributional equivalence.''
-% In Section~\hg{todo}, we discuss why distributional equivalence is often weaker
-% than one might like when evaluating the differences between
-% generators.\hg{Update this}
-
 Generator libraries also include other functions that make generator
@@
 The monad functions \texttt{return} and \texttt{bind} are defined in the usual way:
-for an instance of the reader monad \cn{}: \texttt{return} ignores the size and seed and returns its argument, while \texttt{bind} runs \texttt{g} and passes the result to \texttt{k}.
+for an instance of the reader monad~\cite{moggi91}: \texttt{return} ignores the size and seed and returns its argument, while \texttt{bind} runs \texttt{g} and passes the result to \texttt{k}.
 The \texttt{gen\_int} combinator simply calls out to the randomness library \texttt{Splittable\_random}, aliased as \texttt{SR} here.
@@
 each individual closure allocation is relatively cheap; but doing lots of allocation in a
-generator is very costly because each allocation brings us closer to the next costly GC pause \cn{}.
+generator is very costly because each allocation brings us closer to the next costly GC pause.
 This effect is also magnified in recursive generators: each iteration
@@
 a compiler could in principle unroll the loops in \texttt{weighted\_union} to depth \texttt{n} and specialize the function at each call site to avoid allocating the list.
-Unfortunately, (almost\footnote{GHC is a notable exception here, performing sophisticated list fusion optimizations \cn{}}) no compilers perform this kind of optimization.
+Unfortunately, (almost\footnote{GHC is a notable exception here, performing sophisticated list fusion optimizations \cite{gill93}}) no compilers perform this kind of optimization.
 This allocation (or rather, its tendency to cause GC pauses)---as well as the cost of running the code
@@
 in ways that are asymptotically more efficient but slower in common cases than a more naive algorithm.
-Weighted union uses the Fitness Proportionate Selection algorithm \cn{}, which (1) samples a number $r$
+Weighted union uses the Fitness Proportionate Selection algorithm~\cite{geneticalgorithm}, which (1) samples a number $r$
 between $0$ and the sum of the weights, and then (2) finds the first generator for which the cumulative
 weights in the list before it exceeds $r$. This second part can be accomplished in $O(\log_2 n)$ time by
-a binary search. However, since the lists are short in practice, a linear scan is almost always faster \cn{}.
+a binary search. However, since the lists are short in practice, a linear scan is almost always faster.
 Moreover, both \bq{} and ScalaCheck allocate auxiliary data structures (an array in \bq{} and a BST in ScalaCheck) to
@@
 Following the original Haskell QuickCheck implementation, \bq\ uses the SplitMix
-algorithm \cn, implemented by \texttt{Splittable\_random}. Meanwhile,
-ScalaCheck uses the JSF algorithm \cn.
+algorithm~\cite{splitmix}, implemented by \texttt{Splittable\_random}. Meanwhile,
+ScalaCheck uses the JSF algorithm~\cite{jsf}.
 % https://www.pcg-random.org/posts/bob-jenkins-small-prng-passes-practrand.html
@@
 Indeed, even basic generators --- like ones generating a single \texttt{int} or \texttt{float} uniformly within a range --- can sample \emph{unboundedly many}
-random numbers, since they usually use versions of rejection sampling \cn{} to find a value within the range.
+random numbers, since they usually use versions of rejection sampling~\cite{vn51} to find a value within the range.
 Moreover, generator combinators like \texttt{list} usually make $O(n)$ calls to the \texttt{int} or \texttt{float} generators.

### 2025-03-24 paper/paper.tex
newfigs

@@
   % \includegraphics[scale=0.33]{allocs.png}
-  \includegraphics[scale=0.43]{time.png}
-  \caption{Time to generate values of varying sizes using each \camlname strategy.
-Lower is better.
-Both axes are logarithmic.
-BQ is \bq; Staged SR is \camlname using \bq's \texttt{splittable\_random} RNG; Staged CSR is \camlname using \csplitmix.}
+  \includegraphics[scale=0.57]{time.png}
+  \caption{Time to generate values of varying sizes using each \camlname strategy. Lower is better. Both axes are logarithmic. BQ is \bq.}
   \label{fig:time}
@@
 \begin{figure}[h]
-  \includegraphics[scale=0.28]{bind.png}
-  \caption{Left: Speedup from staging (compared to baseline \bq) versus the number of \texttt{bind} calls.
-  Right: Speedup from \csplitmix (compared to staging alone) versus the number of random samples.}
+  \includegraphics[scale=0.43]{bind.png}
+  \caption{Left: Speedup from staging (compared to \bq) versus the number of \texttt{bind} calls.
+  Right: Speedup from \csplitmix (compared to \camlname alone) versus the number of random samples.}
   \label{fig:binds}
@@
   \centering
-  \includegraphics[scale=0.41]{scala.png}
+  \includegraphics[scale=0.25]{scala.png}
  \caption{Time to generate values of varying sizes using each \scalaname strategy. Lower is better. Both axes are logarithmic. SC is ScalaCheck; Staged is \scalaname.}
@@
 \begin{figure}[h]
-  \includegraphics[scale=0.3]{speedup.png}
-  \caption{Geometric average of all speedups---relative to \bq---for each strategy and benchmark, showing that staging leads to better bug-finding speed across the board.
+  \includegraphics[scale=0.33]{speedup.png}
+  \caption{Geometric average of all speedups---relative to \bq---for each strategy and benchmark, showing that staging leads to better bug-finding speed across the board. CSM is \csplitmix.
   }
@@
 \begin{figure}[h]
-  \includegraphics[scale=0.31]{swarm.png}
-  \caption{Distribution of individual-task speedups across strategies and benchmarks. Gray dots represent tasks. Plots with extremely low variance have no swarm overlay.}
+  \includegraphics[scale=0.49]{swarm.png}
+  \caption{Distribution of individual-task speedups across strategies and benchmarks. Gray dots represent tasks. Plots with extremely low variance have no swarm overlay. CSM is \csplitmix.}
   \label{fig:swarm}

### 2025-03-24 paper/paper.tex
most

--- 2025-03-24 paper/paper.tex
@@
 
-% Moreover, PBT libraries could even consider eschewing the requirement that a source of
-% entropy pass statistical tests like BigCrush\cn. \hg{This needs more discussion}
-% This is common practice in other areas of testing already:
-% fuzzers often simply use a buffer full of arbitrary bytes as a source of entropy
-% \cn. Such approaches are also faster than algorithms like SplitMix: bumping a
-% pointer and reading from memory (which can be pipelined trivially) will always
-% be faster on modern CPUs.
-% \hg{Should we just always be doing this? I think we want to argue
-% that this is a bridge too far because you still need to generate the buffer
-% ahead of time and you can run out of randomness?}
-% \jwc{HG: talk about the tradeoffs here.}
-% \jwc{The point here is that ``you should be thinking about using a more performant RNG, we'll show you that it does have real benefits (and quantify them). But there are tradeoffs here.'}
-% \bcp{But this is not the point that the reader is set up to expect: we've toldthem that we're going to make generation faster, not that we're going to give them tips so that htey can make it faster by themselves}
-% \jwc{This actually isn't the case. In the intro, we're very careful to say that we show that choosing a faster randomness library \emph{leads to faster bugfinding}. And we do demonstrate this.}
-% \jwc{Ok --- we could get rid of this ``source of entropy'' business and just write that you can speed up any of these by pipelining your sampling --- sampling and saving multiple values at a time with SIMD instead of just one each.}
-
 Of course, simply arguing that the randomness library is on the hot
@@
 While embedded DSLs are powerful tools, they have a well-known drawback:
-the functional abstractions used to build eDSLs tend to prevent compilers from generating efficient machine code \cn.
+the functional abstractions used to build eDSLs tend to prevent compilers from generating efficient machine code.
 This is \emph{abstraction overhead} \cite{carrette05, moller20}: the layers of abstraction that
@@
 Scala 3 \cite{scalamacros}, Haskell \cite{templatehaskell}, \jwc{Racket??}, OCaml \cite{metaocaml,macocaml}, and Java \cite{mint}.
-For this paper, we have implemented staged PBT libraries both in OCaml (using MetaOCaml \cn) and Scala 3 (using
+For this paper, we have implemented staged PBT libraries both in OCaml (using MetaOCaml \cite{metaocaml}) and Scala 3 (using
 \texttt{scala.quoted}). All staged code presented in the body of this paper is \camlname{} code written in OCaml:
@@
 
-This program is known as ``The Trick'' in the partial evaluation and multi-stage programming literature \cn{}.
+This program is known as ``The Trick'' in the partial evaluation and multi-stage programming literature~\cite{metaocamlimpl}.
 % USE THIS: https://arxiv.org/pdf/2309.08207
@@
 
-This problem is intimately related to nondeterministic effects in the presence of CBN/CBV evaluation ordering \cn.
 In essence, the behavior of splice \texttt{.\~{}cx} in a staged function \texttt{f(cx : 'a code) = ...} is to \emph{copy}
@@
 To validate the library, we built a PBT harness to compare staged generators to their \bq equivalents over 1,000 random seeds.
-By differentially testing \cn{} a large suite of generators in this way, we gained confidence that \camlname is equivalent to \bq{}.
+By differentially testing~\cite{difftesting} a large suite of generators in this way, we gained confidence that \camlname is equivalent to \bq{}.
 
@@
 Note that the performance intervention described here is OCaml-specific.
-In most other PBT frameworks \cn, the randomness library used operates on machine
-integers, so this \emph{particular} inefficiency does not exist. \bcp{This paragraph should go on to explain why it is still interesting to talk about or measure it.}
-However, the insights that we will derive from this experiment (in Section~\ref{section:eval})---
+In most other PBT frameworks, the randomness library used operates on machine
+integers, so this \emph{particular} inefficiency does not exist.
+However, the insights that we will derive from this experiment in Section~\ref{section:eval}---
 about which kinds of generators benefit the most from faster randomness and how it leads to faster bug-finding---
 are applicable to all languages.
-\jwc{How's that}
-
-The precise details of the SplitMix algorithm~\cn are not important for
+
+The precise details of the SplitMix algorithm~\cite{splitmix} are not important for
 present purposes; the key fact is that all of its operations are
@@
 at run time as a \emph{pointer} to an unscanned block of memory containing (among other things)
-a 64-bit integer \cn. This means that all operations that return an \texttt{int64} must allocate this block of memory,
+a 64-bit integer~\cite{ocamlruntimerep}. This means that all operations that return an \texttt{int64} must allocate this block of memory,
 which has a significant impact on performance. A single call to
@@
 to the next garbage collection pause. While small allocations like these are \emph{very} fast to perform and subsequently
-collect in OCaml\footnote{The OCaml GC is a generational collector \cn, and since these allocations are small and mostly very short lived, they will all be minor allocations, never to be promoted.}
+collect in OCaml\footnote{The OCaml GC is a generational collector~\cite{realworldocaml}, and since these allocations are small and mostly very short lived, they will all be minor allocations, never to be promoted.}
 , we will see in Section~\ref{section:eval} that they can still have

### 2025-03-24 paper/paper.tex
its like high school

@@
 test. These generators are written with the help of \textit{generator
-libraries}, usually expressed as embedded domain-specific languages (eDSLs) %\cntxt{edsls}
+libraries}, usually expressed as embedded domain-specific languages (eDSLs)
 that provide combinators for building and composing generators.

### 2025-03-24 paper/paper.tex
remove commented cns
pdffig
cites

@@
 % \bcp{Sentence starts with lowerase} \bq{} uses a reader monad with a mutable seed, while Haskell's QuickCheck uses a reader monad with an immutable state
-% that is ``split'' at \texttt{bind}s \cn. Meanwhile, ScalaCheck (a) uses an immutable state type and state monad \cn{} to thread
+% that is ``split'' at \texttt{bind}s. Meanwhile, ScalaCheck (a) uses an immutable state type and state monad to thread
 % the state through, and (b) has an \texttt{Option} as its return type, to allow generation to fail.
@@
 One of the primary uses of staging is in embedded DSL construction \cite{sheard99, lms, trattdsl}.
-\jwc{Cite racket macros for DSLs here:}
 While embedded DSLs are powerful tools, they have a well-known drawback:
@@
 Many languages have some degree of staging functionality, including
-Scala 3 \cite{scalamacros}, Haskell \cite{templatehaskell}, \jwc{Racket??}, OCaml \cite{metaocaml,macocaml}, and Java \cite{mint}.
-For this paper, we have implemented staged PBT libraries both in OCaml (using MetaOCaml \cite{metaocaml}) and Scala 3 (using
+Scala 3~\cite{scalamacros}, Haskell~\cite{templatehaskell}, Racket~\cite{flatt12}, OCaml~\cite{metaocaml,macocaml}, and Java~\cite{mint}.
+For this paper, we have implemented staged PBT libraries both in OCaml (using MetaOCaml~\cite{metaocaml}) and Scala 3 (using
 \texttt{scala.quoted}). All staged code presented in the body of this paper is \camlname{} code written in OCaml:
@@
 % part of the compile-time stage), and which parts are only known dynamically (and
-% hence must be \texttt{code}) --- is an art known as ``binding-time analysis'' \cn.
+% hence must be \texttt{code}) --- is an art known as ``binding-time analysis''
 
@@
 C code\footnote{Ideally in the future, one would not need to call out to C for this: the Jane Street bleeding-edge OCaml compiler has support
-for unboxed types \cn, which (among other things) would let us implement a version of SplitMix that does not allocate, directly
+for unboxed types~\cite{unboxedtypes}, which (among other things) would let us implement a version of SplitMix that does not allocate, directly
 in OCaml. Unfortunately, the Jane Street branch of the compiler is incompatible with MetaOCaml, which we use to implement the metaprogramming
@@
 and OpenJDK 21.0.6, using ScalaCheck version 1.17 as the baseline.
-We used \texttt{core\_bench} 0.16 \cn in OCaml and \texttt{jmh} 0.4.7 \cn for performance microbenchmarking. For assessing and comparing PBT techniques, we used the Etna\cn platform.
+We used \texttt{core\_bench} 0.16~\cite{corebench} in OCaml and \texttt{jmh} 0.4.7~\cite{jmh} for performance microbenchmarking.
+For assessing and comparing PBT techniques, we used the Etna platform~\cite{etna}.
 
@@
 boolean lists, and well-typed STLC terms. To determine how often \texttt{bind} is invoked, we generate 10,000 values using a fixed generation size of 100 and
-record the average number of \texttt{bind} calls. To determine the number of random samples, we repeat the same test but use Intel ProcessorTrace \cn{} to capture a 4ms snapshot of processor activity.
+record the average number of \texttt{bind} calls.
+To determine the number of random samples, we repeat the same test but use Intel ProcessorTrace~\cite{intelpt} to capture a 4ms snapshot of processor activity.
 We then count the number of invocations of \texttt{next\_int64} and average this over the number of generator calls in the trace.
@@
 To answer \textbf{RQ2}, we evaluate the bug-finding speed of our BST and STLC
-case studies using Etna\cn{}.
+case studies using Etna~\cite{etna}.
 Etna allows users to measure the effectiveness of different generator
@@
 
-In the years since the original QuickCheck paper \cn{}, PBT has had remarkable success
-in languages outside of the Haskell world from which it came.
+In the years since the original QuickCheck paper~\cite{quickcheck00}, PBT has had remarkable success
+in languages outside of the Haskell world from which it came~\cite{bq,sc,hypothesis}.
 For this reason, readers who are users and developers of PBT libraries in \emph{other} languages ---
@@
   The entire Racket philosophy is intertwined with
-  using macros to build small eDSLs like the ones we use to write generators, and so it is a natural fit. Indeed,
-  there are at least two \cn{} PBT libraries in Racket.
+  using macros to build small eDSLs like the ones we use to write generators, and so it seems a natural fit.
   However, the racket macros literature, while extensive, does not usually concern itself with staging for performance purposes.
@@
 
-  \textbf{Haskell} is home to the original PBT implementation \cn{}, and
+  \textbf{Haskell} is home to the original PBT implementation, and
   so it is natural to ask why we have not built a \name{} in it yet. The basic
@@
   to performance overheads for monadic code.  For this reason, Haskell does have
-  a multi-stage programming system called Template Haskell \cn{}, which is
+  a multi-stage programming system called Template Haskell~\cite{templatehaskell}, which is
   sometimes to build eDSLs in contexts where one does not want to rely on the
@@
 
-  \textbf{F\#} has both multi-stage programming capabilities \cn{} and a well-used PBT library called FsCheck \cn{}.
-  The internals of FsCheck's generator eDSL are very similar to both \bq and ScalaCheck, so we expect that building a \name{}
+  \textbf{F\#} has both multi-stage programming capabilities~\cite{fsmeta} and a well-used PBT library called FsCheck~\cite{fscheck}.
+  The internals of FsCheck's generator eDSL are very similar to both \bq and ScalaCheck, so we expect that building an \name{}
   in F\# is primarily an engineering effort.
@@
   \textbf{Rust} shares some similarities with functional languages, and
-  indeed, it is host to a number of property-based testing libraries \cn{}. However Rust does not
+  indeed, it is host to a number of property-based testing libraries~\cite{Bolero,ProptestBook}. However Rust does not
   structurally encourage monadic eDSLs --- it does not have a special monadic syntax --- and so
@@
 metaprogramming. Quasiquation-style macros were introduced to the ML family of
-languages by Nielson and Nielson's Two-Level Functional Languages~\cn{} and
-Taha's MetaML~\cn{}, which in turn quickly inspired MetaOCaml (and more recently
-MacoCaml) in OCaml~\cn{}, Lightweight Modular Staging in Scala \cn{}, Template
-Haskell in Haskell~\cn{}, and more.  In addition to implementations, researchers
+languages by Nielson and Nielson's Two-Level Functional Languages~\cite{nielson92} and
+MetaML~\cite{taha00}, which in turn quickly inspired MetaOCaml~\cite{metaocaml} (and more recently
+MacoCaml~\cite{macocaml}) in OCaml, Lightweight Modular Staging in Scala~\cite{lms}, Template
+Haskell in Haskell~\cite{templatehaskell}, and more.  In addition to implementations, researchers
 have studied the type-theoretic foundations of multi-stage languages \cite{cmtt,
@@
 On the abstraction overhead side, we hope to employ some of the many tricks for better code generation that have been written
-about in the staging literature. For example, MetaOCaml includes a primitive floating let-insertion \cn{} that yields optimal let placement,
+about in the staging literature. For example, MetaOCaml includes a primitive floating let-insertion~\cite{kiselyov2022letrecinsertioneffects} that yields optimal let placement,
 which we did not use in \camlname{}.
@@
 
-On the \rands{} side we plan to investigate fast \rands{} that are not equivalent to SplitMix, such as Lehmer~\cn, WyHash~\cn,
-and variants of xoroshiro~\cn. We also want to try techniques to speed up sampling such as pipelining, using SIMD instructions,
+On the \rands{} side we plan to investigate fast \rands{} that are not equivalent to SplitMix, such as Lehmer~\cite{lehmerrng}, WyHash~\cite{wyrand},
+and variants of xoroshiro~\cite{xorshiro}. We also want to try techniques to speed up sampling such as pipelining, using SIMD instructions,
 or even generating random numbers ahead of time. Last, we hope to investigate the degree to which actual statistical randomness

### 2025-03-24 paper/paper.tex
binding time analysis
fig
beautiful beautiful swarm
bind
rng para

@@
 
-We isolate and quantify the performance benefits of faster randomness,
-showing that a DSL's choice of \rand{} has a dramatic impact on its
-performance.  \bcp{describe the experiment}
+We isolate and quantify the performance benefits of faster randomness, showing
+that a DSL's choice of \rand{} has a dramatic effect on its performance. Based
+on the observation that the OCaml implementation of \bq{}'s \rand{} is slow in
+an easy-to-fix way, we perform a controlled experiment, comparing fast and slow
+versions of the same library to analyze the impact of faster \rand{} on
+generator performance.
 
@@
 which parts are only known at run time (and must be treated as \texttt{code}).
-In the staging literature, this task is known as ``binding-time analysis,'' and it is more art than science.
+In the staging literature, this task is known as ``binding-time analysis,'' \cite{jones1993partial} and it is more art than science.
 
@@
   % \includegraphics[scale=0.33]{allocs.png}
-  \includegraphics[scale=0.57]{time.png}
+  \includegraphics[scale=0.60]{time.pdf}
   \caption{Time to generate values of varying sizes using each \camlname strategy. Lower is better. Both axes are logarithmic. BQ is \bq.}
@@
 \begin{figure}[h]
-  \includegraphics[scale=0.33]{speedup.png}
+  \includegraphics[scale=0.46]{speedup.pdf}
   \caption{Geometric average of all speedups---relative to \bq---for each strategy and benchmark, showing that staging leads to better bug-finding speed across the board. CSM is \csplitmix.

### 2025-03-24 paper/paper.tex
blah
devoid of thoughts

@@
 on the observation that the OCaml implementation of \bq{}'s \rand{} is slow in
-an easy-to-fix way, we perform a controlled experiment, comparing fast and slow
+a way that is simple to fix, we perform a controlled experiment, comparing fast and slow
 versions of the same library to analyze the impact of faster \rand{} on
@@
   % \includegraphics[scale=0.33]{allocs.png}
-  \includegraphics[scale=0.60]{time.pdf}
+  \includegraphics[scale=0.66]{time.pdf}
   \caption{Time to generate values of varying sizes using each \camlname strategy. Lower is better. Both axes are logarithmic. BQ is \bq.}
@@
 \begin{figure}[h]
-  \includegraphics[scale=0.43]{bind.png}
+  \includegraphics[scale=0.46]{bind.pdf}
   \caption{Left: Speedup from staging (compared to \bq) versus the number of \texttt{bind} calls.
@@
   \centering
-  \includegraphics[scale=0.25]{scala.png}
+  \includegraphics[scale=0.66]{scala.pdf}
  \caption{Time to generate values of varying sizes using each \scalaname strategy. Lower is better. Both axes are logarithmic. SC is ScalaCheck; Staged is \scalaname.}
@@
 \begin{figure}[h]
-  \includegraphics[scale=0.46]{speedup.pdf}
+  \includegraphics[scale=0.45]{speedup.pdf}
   \caption{Geometric average of all speedups---relative to \bq---for each strategy and benchmark, showing that staging leads to better bug-finding speed across the board. CSM is \csplitmix.
@@
 \begin{figure}[h]
-  \includegraphics[scale=0.49]{swarm.png}
+  \includegraphics[scale=0.35]{swarm.pdf}
   \caption{Distribution of individual-task speedups across strategies and benchmarks. Gray dots represent tasks. Plots with extremely low variance have no swarm overlay. CSM is \csplitmix.}

### 2025-03-24 paper/paper.tex
blah

@@
 a way that is simple to fix, we perform a controlled experiment, comparing fast and slow
-versions of the same library to analyze the impact of faster \rand{} on
+versions of the same library to analyze the impact of faster \rands{} on
 generator performance.

### 2025-03-24 paper/paper.tex
capsg

@@
 changing the order in which test inputs are sampled. Feedback-driven mechanisms
-like targeted PBT~\cite{loscherTargetedPropertybasedTesting2017} and
-coverage-guided PBT~\cite{lampropoulosCoverageGuidedProperty2019} have the
+like Targeted PBT~\cite{loscherTargetedPropertybasedTesting2017} and
+Coverage-Guided PBT~\cite{lampropoulosCoverageGuidedProperty2019} have the
 potential to speed up testing by changing the generation order to try to find

### 2025-03-24 paper/paper.tex
rm stale commentsg

@@
 a}.
-\iflater
-\jwc{\texttt{gen\_int} here is really \texttt{int\_uniform\_inclusive}. I also
-think we should completely omit named arguments in the paper so we don't confuse
-with splices.}
-\fi
 
@@
 return has type \texttt{'a -> 'a Gen.t} and bind has type \texttt{'a Gen.t -> ('a -> 'b Gen.t) -> 'b Gen.t}
-\iflater
-\hg{A pedantic reader may ask: Are we concerned with the monad {\em laws} as
-well, or just the type signature?}
-\fi
 
@@
 our staged programs in continuation-passing style \cite{bondorf92}.
-\hg{Slow down! This is all very interesting and very technical! Try making each
-of these sentences two sentences, add some citations, and maybe spell this out
-wiht an example}
-\jwc{I could, but i'm just not sure this is important. It's too deep in the weeds for it to be worth explaining, too... nobody's going to read this section except staging fans}
-\bcp{...
-but they are gonig to be sad if they don't understand the details.
-If we're not too desperate for space, my suggestion would be to put in some more details plus sign posting that says tehy can be skipped.
-If we're up against space limits, then maybe most of the section should go to the appendix.
-Or second thought, maybe we should lean into the cmplexity here (and everywhere else) to give the impressions that what we've done is not simply applying known ideas in sraightforward ways}
-\jwc{Unfortunately, I think this technique is sufficiently well-understood to staging experts that we can't claim novelty. We'll jsut cut it for space if we need it.}
 

### 2025-03-24 paper/paper.tex
scala
more consistent figure text size
gnenerators

@@
 Generator'' (RNG). We will avoid this term and instead say ``randomness library''
-to avoid confusing RNG implementations with the PBT gnenerator libraries that
+to avoid confusing RNG implementations with the PBT generator libraries that
 use them.}.
@@
 
+\subsection*{Data Availabilty Statement}
+For artifact evaluation we will submit (1) the source code of \camlname{} and \scalaname{} as well as (2) the code for our version of Etna,
+which we have modified slightly to pass deterministic seeds to all versions of a single strategy.
+These programs are not available publicly or anonymized at the time of submission.
+
 %%

### 2025-03-24 paper/paper.tex
data avail
relwork

@@
 \begin{figure}[h]
-  \includegraphics[scale=0.46]{bind.pdf}
+  \includegraphics[scale=0.44]{bind.pdf}
   \caption{Left: Speedup from staging (compared to \bq) versus the number of \texttt{bind} calls.
@@
   \centering
-  \includegraphics[scale=0.66]{scala.pdf}
+  \includegraphics[scale=0.68]{scala.pdf}
  \caption{Time to generate values of varying sizes using each \scalaname strategy. Lower is better. Both axes are logarithmic. SC is ScalaCheck; Staged is \scalaname.}
@@
 \begin{figure}[h]
-  \includegraphics[scale=0.45]{speedup.pdf}
+  \includegraphics[scale=0.59]{speedup.pdf}
   \caption{Geometric average of all speedups---relative to \bq---for each strategy and benchmark, showing that staging leads to better bug-finding speed across the board. CSM is \csplitmix.
@@
 \begin{figure}[h]
-  \includegraphics[scale=0.35]{swarm.pdf}
+  \includegraphics[scale=0.36]{swarm.pdf}
   \caption{Distribution of individual-task speedups across strategies and benchmarks. Gray dots represent tasks. Plots with extremely low variance have no swarm overlay. CSM is \csplitmix.}
@@
 query processing \cite{rhyme}, and parser combinators \cite{sspc, staged-parsers, krishnaswami19, flap}.
-This work on staged parsers is the closest analogue to ours. Recent PBT research has drawn an explicit link between parsing and generation, framing generators
+This work on staged parsers is the closest analogue to ours---indeed, PBT research has drawn an explicit link between parsing and generation, framing generators
 as ``parsers of randomness'' \cite{goldsteinParsingRandomness2022}.
-\jwc{Words here about the difference!!}
+However, this equivalence is primarily theoretical. PBT generators are (a) not actually implemented as parsers
+of random sequences, and (b) support choice-based combinators like \texttt{weighted\_union} that parsers do not.
+For this reason, we must build staged generator libraries from scratch and cannot simply use existing staged parser libraries
 
@@
 These programs are not available publicly or anonymized at the time of submission.
+The dataset for our evaluation is produced by a combination of Etna and microbenchmarks in the \camlname{} and \scalaname{} sources.
+It is also not available publicly at the time of submission, but the artifact will contain the means and instructions to reproduce it.
 

### 2025-03-24 paper/paper.tex
blah
eval para

@@
 generator library, and show that the \name{} generators run faster
-than their unstaged equivalents.  Further, we show generators that run
-faster also find bugs faster by running these case studies in the Etna
+than their unstaged equivalents. In \camlname, they run up to $7\times$
+faster, and in \scalaname, up to $13\times$ faster. Using our improved version
+of \bq's \rand in \camlname, we see even greater gains---more than $12\times$---showing that
+both staging and fast randomness play distinct and complementary roles in improving generator performance.
+Further,
+we show that generators that run
+faster also find bugs faster by running our case studies in the Etna
 platform~\cite{etna}, which injects bugs into a system and measures
-how quickly various generators detect them.  \jwc{Beef up this
-  paragraph, explain the results of our eval.}  \todo{We precisely
-  quantify the impact of each intervention and explain how they play
-  distinct yet complementary roles in improving generator performance.
-}
+how quickly various generators detect them. In Etna, we find bugs up 
+to $3.8\times$ faster on average, and up to $13\times$ in some cases.
 
@@
 of random sequences, and (b) support choice-based combinators like \texttt{weighted\_union} that parsers do not.
-For this reason, we must build staged generator libraries from scratch and cannot simply use existing staged parser libraries
+For this reason, we must build staged generator libraries from scratch and cannot simply use existing staged parser libraries.
 

### 2025-03-24 paper/paper.tex
attempt at eval para

@@
 we show that generators that run
-faster also find bugs faster by running our case studies in the Etna
-platform~\cite{etna}, which injects bugs into a system and measures
-how quickly various generators detect them. In Etna, we find bugs up 
-to $3.8\times$ faster on average, and up to $13\times$ in some cases.
+faster also find bugs faster by running our case studies in the Etna platform, which uses generated values as
+inputs to a pre-defined buggy system and measures how quickly different generators can detect the bugs.
+Combined with fast randomness, \camlname generators find bugs 
+to $2.3-3.8\times$ faster on average, and up to $13\times$ in some cases.
 

### 2025-03-24 paper/paper.tex

--- 2025-03-24 paper/paper.tex
@@
 Further,
-we show that generators that run
-faster also find bugs faster by running our case studies in the Etna
-platform~\cite{etna}, which injects bugs into a system and measures
-how quickly various generators detect them. In Etna, we find bugs up 
-to $3.8\times$ faster on average, and up to $13\times$ in some cases.
+we show that \camlname generators find bugs faster by running our case studies 
+in the Etna platform, which uses generated values as
+inputs to a pre-defined buggy system and measures how quickly different 
+generators can detect the bugs.
+\camlname generators find bugs up to $2.5\times$ faster on average than 
+their \bq counterparts. With fast randomness, 
+the average speedup rises to $3.8\times$, 
+with some cases exceeding $13\times$.
 

### 2025-03-24 paper/paper.tex
Polishing on 3.1 and 3.2

--- 2025-03-24 paper/paper.tex
@@
 \newif\iflater\latertrue{}
+\newif\ifaftersubmission\aftersubmissionfalse{}
+\newif\ifsilly\sillyfalse{}
 
@@
 functionality  (\ref{xx}).
-Last, we present the controlled experiment we use to demonstrate how a
+Last, we present a controlled experiment demonstrating how a
 faster \rand{} leads to failing faster  (\ref{xx}).
@@
 
-In multi-stage programming (or simply staging), programs execute
-in multiple stages, with each stage producing the code to be run
+In multi-stage programming (or simply ``staging''), programs execute
+in multiple stages, with each stage producing code to be run
 in the next. For the purposes of this paper, we just need two stages: compile time and run time.
@@
 This is \emph{abstraction overhead} \cite{carrette05, moller20}: the layers of abstraction that
-make eDSLs so pleasant to use are precisely what prevents them from being fast.
+make eDSLs pleasant to use are precisely what prevents them from being fast.
 Indeed, the root causes of many of the performance issues we discovered in Section~\ref{section:motiv} are
@@
 Scala 3~\cite{scalamacros}, Haskell~\cite{templatehaskell}, Racket~\cite{flatt12}, OCaml~\cite{metaocaml,macocaml}, and Java~\cite{mint}.
-For this paper, we have implemented staged PBT libraries both in OCaml (using MetaOCaml~\cite{metaocaml}) and Scala 3 (using
-\texttt{scala.quoted}). All staged code presented in the body of this paper is \camlname{} code written in OCaml:
+For this paper, we have implemented staged PBT libraries in both OCaml (using MetaOCaml~\cite{metaocaml}) and Scala 3 (using
+\texttt{scala.quoted}). All staged code presented in the body of this paper is \camlname{} code written in OCaml;
 \scalaname{} is very similar.
-For further discussion of the potential for staged generator DSLs in other languages, see Section~\ref{section:other-langs}.
+We discuss the potential for staged generator DSLs in other languages in Section~\ref{section:other-langs}.
 
@@
 example, the program \texttt{.< 5 + 1 >.} has type \texttt{int code}.  Note that
-this is not the same as \texttt{.< 6 >.} Because brackets delay computation,
+this is not the same as \texttt{.< 6 >.}: because brackets delay computation,
 the code is not \emph{executed} until the next stage (run time).
@@
 \texttt{.<(1 * 5) + (1 * 5)>.}  MetaOCaml enforces correct scoping and macro
-hygiene, ensuring that variables are not shadowed when open terms are spliced.
+hygiene, ensuring that variables are not shadowed when open terms are
+spliced in under binders.
 
@@
 the quoted \texttt{x}.
-
-The trick is best illustrated by an example. The following program reduces at compile time to \texttt{.< fun x -> (1 + x) mod 2 == 0 >.}:
+%
+For example, the following program
 \begin{lstlisting}
@@
 \end{lstlisting}
-By composing the two code-transforming functions together at compile time, and only then turning them into a run-time function,
+reduces at compile time to the more efficient \texttt{.< fun x -> (1 + x) mod 2 == 0 >.}
+%
+By composing the two code-transforming functions together at compile time and only then turning them into a run-time function,
 the two functions are fused together.
 This is the basis of how staging is used to eliminate the abstraction overhead of DSLs.
-By writing DSL combinators as compile-time functions ---
-and calling \texttt{eta} at the end on the completed DSL program ---  we
-can ensure that any overhead of using the combinators is eliminated by run time.
+By writing DSL combinators as compile-time functions---and calling \texttt{eta} at the end on the completed DSL program---we
+can ensure that any overhead of using the combinators is eliminated before run time.
 
@@
 \label{subsection:basic-design}
-To build a staged version of a generator DSL, we must rewrite the the generator
+To build a staged version of a generator DSL, we want to rewrite the the generator
 combinators to do as much as possible at compile time.
@@
 which parts are only known at run time (and must be treated as \texttt{code}).
-In the staging literature, this task is known as ``binding-time analysis,'' \cite{jones1993partial} and it is more art than science.
+In the staging literature, this task is known as ``binding-time
+analysis,'' \cite{jones1993partial}, and it is more art than
+science.\bcp{Why do we need to say that last part?}
 
@@
 
-The crux of our binding time analysis is that the \emph{only} parts of a generator that are not
+The crux of our binding time analysis\bcp{the wording here is
+  misleading: it feels like we're implying we've {\em implemented} a
+  binding-time analysis...} is that the \emph{only} parts of a
+generator that are not
 known at compile time are the random seed and
@@
 This leads us to define our library's generator type \texttt{'a
-Gen.t} as \texttt{int code -> SR.t code -> 'a code}: a compile-time
+  Gen.t} as
+\[
+\texttt{'a Gen.t} \ \ = \ \  \texttt{int code -> SR.t code -> 'a code}.
+\]
+That is, a \texttt{Gen.t} is a compile-time
 function from dynamically known size and seed to dynamically determined result.
 
-The basic staged monadic generator DSL functionality can be found in Figure~\ref{fig:gen-staged-basic}.
+The basic functionality of the staged monadic generator DSL can be
+found in Figure~\ref{fig:gen-staged-basic}.
 The constant generator \texttt{return} runs at compile time. Given \texttt{cx : 'a code}, the code for an \texttt{'a}, it
-returns the generator that runs \texttt{cx}. Similarly,
+returns the generator that ignores its \texttt{size} and
+\texttt{random} arguments and simply runs\bcp{returns the code for?} \texttt{cx}. Similarly,
 \texttt{bind g k} sequences generators by passing the result of running the generator \texttt{g} to the continuation \texttt{k}. However, instead of getting access
-to the particular value generated by \texttt{g}, the continuation \texttt{k} gets access to \texttt{code} for the value sampled from \texttt{g}:
+to the particular value generated by \texttt{g}, the continuation
+\texttt{k} only gets access to \texttt{code} for the value sampled from \texttt{g}:
 we know that at run time the code \texttt{g} generates will produce \emph{some} \texttt{'a}, but at compile time we cannot inspect the value.
@@
 Lastly, \texttt{to\_bq} turns a staged generator into code for a normal \bq\
-generator. This function is just a 2-argument version of ``The Trick''
+generator. This is just a 2-argument version of ``The Trick''
 from Section~\ref{subsection:msp}.
-\iflater
+\ifaftersubmission
 \hg{TODO: Be really careful with formatting of the above paragraph. It's very
@@
 \end{lstlisting}
-\caption{Basic Staged Generator Library}
+\caption{Basic Staged Generator Library\ifaftersubmission\bcp{We could
+    save some space if needed by eliminating a couple of linebreaks
+    here.}\fi}
 \label{fig:gen-staged-basic}
@@
 let int_pair : (int * int) Bq.t code = Gen.to_bq int_pair_staged
-(* .< fun size random ->
+(* int_pair = .< fun size random ->
                  let x = SR.int random 0 100 in
@@
                  (x,y)
-    >.
-*)
+               >.   *)
 \end{lstlisting}
@@
   % \includegraphics[scale=0.33]{allocs.png}
-  \includegraphics[scale=0.66]{time.pdf}
+\ifsilly  \includegraphics[scale=0.66]{time.pdf} \fi
   \caption{Time to generate values of varying sizes using each \camlname strategy. Lower is better. Both axes are logarithmic. BQ is \bq.}
@@
 \begin{figure}[h]
-  \includegraphics[scale=0.44]{bind.pdf}
+\ifsilly  \includegraphics[scale=0.44]{bind.pdf} \fi
   \caption{Left: Speedup from staging (compared to \bq) versus the number of \texttt{bind} calls.
@@
   \centering
-  \includegraphics[scale=0.68]{scala.pdf}
+\ifsilly  \includegraphics[scale=0.68]{scala.pdf}\fi
  \caption{Time to generate values of varying sizes using each \scalaname strategy. Lower is better. Both axes are logarithmic. SC is ScalaCheck; Staged is \scalaname.}
@@
 \begin{figure}[h]
-  \includegraphics[scale=0.59]{speedup.pdf}
+\ifsilly  \includegraphics[scale=0.59]{speedup.pdf}\fi
   \caption{Geometric average of all speedups---relative to \bq---for each strategy and benchmark, showing that staging leads to better bug-finding speed across the board. CSM is \csplitmix.
@@
 \begin{figure}[h]
-  \includegraphics[scale=0.36]{swarm.pdf}
+\ifsilly  \includegraphics[scale=0.36]{swarm.pdf} \fi
   \caption{Distribution of individual-task speedups across strategies and benchmarks. Gray dots represent tasks. Plots with extremely low variance have no swarm overlay. CSM is \csplitmix.}

### 2025-03-24 paper/paper.tex

@@
 Further,
-we show that generators that run
-faster also find bugs faster by running our case studies in the Etna platform, which uses generated values as
-inputs to a pre-defined buggy system and measures how quickly different generators can detect the bugs.
-Combined with fast randomness, \camlname generators find bugs
-to $2.3-3.8\times$ faster on average, and up to $13\times$ in some cases.
+we show that \camlname generators find bugs faster by running our case studies
+in the Etna platform, which uses generated values as
+inputs to a pre-defined buggy system and measures how quickly different
+generators can detect the bugs.
+\camlname generators find bugs up to $2.5\times$ faster on average than
+their \bq counterparts. With fast randomness,
+the average speedup rises to $3.8\times$,
+with some cases exceeding $13\times$.
 

### 2025-03-24 paper/paper.tex
fix figs
wibbling

@@
 \newif\ifaftersubmission\aftersubmissionfalse{}
-\newif\ifsilly\sillyfalse{}
+\newif\ifsilly\sillytrue{}
 
@@
 With a sense of two main inefficiences in generator libraries, we set out in this section to investigate solutions.
-We begin with a quick tour through multi-stage programming (\ref{xx}),
+We begin with a quick tour through multi-stage programming (Section~\ref{subsection:msp}),
 then incrementally build \camlname, layering on features and
-functionality  (\ref{xx}).
+functionality  (Sections~\ref{subsection:basic-design} through \ref{subsection:type-derived}).
 Last, we present a controlled experiment demonstrating how a
-faster \rand{} leads to failing faster  (\ref{xx}).
+faster \rand{} leads to failing faster  (\ref{subsection:faster-rng}).
 
@@
 In the staging literature, this task is known as ``binding-time
-analysis,'' \cite{jones1993partial}, and it is more art than
-science.\bcp{Why do we need to say that last part?}
+analysis,'' \cite{jones1993partial}.
 

### 2025-03-24 paper/paper.tex
wibble
nit

@@
 \end{lstlisting}
-reduces at compile time to the more efficient \texttt{.< fun x -> (1 + x) mod 2 == 0 >.}
+reduces at compile time to \texttt{.< fun x -> (1 + x) mod 2 == 0 >.}
 %
@@
 Etna does not support Scala, so we were unable to test \scalaname's bug-finding speed in Etna, but we
-expect the results in this section should extend to \scalaname.
+expect the results in this section to extend to \scalaname.
 

### 2025-03-24 paper/paper.tex
yay

@@
 in the Etna platform, which uses generated values as
-inputs to a pre-defined buggy system and measures how quickly different
-generators can detect the bugs.
+inputs to a buggy system and measures how quickly different
+generators can detect the bugs. In Etna,
 \camlname generators find bugs up to $2.5\times$ faster on average than
@@
 Most languages in which monadic APIs are common expose some sort of syntactic
-sugar for them. In OCaml\footnote{OCaml actually has a few ways to implement
+sugar for them. In OCaml, \footnote{OCaml actually has a few ways to implement
 monadic syntax; this is the one provided by Jane Street's libraries.}, this
-looks like \texttt{let\%bind x = e in e'}, which desugars to
+looks like \texttt{let\%bind x = e in e'} which desugars to
 \texttt{bind e (fun x => e')}. Figure~\ref{fig:simple-macro} shows the same
@@
 do in fact have heuristics to determine if and when to perform this particular kind of
-simplification\footnote{As we discuss in Section~\ref{section:other-langs}, purity
+simplification.\footnote{As we discuss in Section~\ref{section:other-langs}, purity
 means that Haskell is a slightly different story. GHC can and often does
-transformations of this form.}. However even in cases as simple as
+transformations of this form.} However even in cases as simple as
 Figure~\ref{fig:simplify}, the indirection of \texttt{return}
@@
 The core of any PBT generator library is a source of randomness.
-Different PBT libraries use different randomness libraries implementing different algorithms
+Different PBT libraries use different randomness libraries implementing different algorithms.
 \footnote{The common term for such an algorithm or library is a ``Random Number
@@
 to avoid confusing RNG implementations with the PBT generator libraries that
-use them.}.
+use them.}
 Following the original Haskell QuickCheck implementation, \bq\ uses the SplitMix
@@
 the type \texttt{'a CodeCps.t = 'z. ('a -> 'z code) -> 'z code}: a polymorphic continuation transformer with the result type
-always in \texttt{code}\footnote{This is an instance of the \emph{codensity} monad \cite{janis08}---a fact that deserves further investigation.}.
+always in \texttt{code}.\footnote{This is an instance of the \emph{codensity} monad \cite{janis08}---a fact that deserves further investigation.}
 The monad instance for this type is the standard instance for a CPS monad with polymorphic return type.
@@
 a \texttt{code} representing a recursive program.
-To this end, we develop a staged recursive generator combinator\footnote{
+To this end, we develop a staged recursive generator combinator.\footnote{
 We actually provide a more general API that allows programmers to define \emph{parameterized}
 recursive combinators, of type \texttt{'r code -> 'a code Gen.t}, for any type \texttt{'r}.
-}, whose API is shown in Figure~\ref{fig:staged-recursive-generator}.
+}, whose API is shown in Figure~\ref{fig:staged-recursive-generator}
 The recursion API consists of an opaque type \texttt{'a handle}, and a function \texttt{recurse} to perform recursive calls.
@@
 proper \texttt{int64\_t} arithmetic, only boxing and unboxing integers at the call boundaries between OCaml and
-C code\footnote{Ideally in the future, one would not need to call out to C for this: the Jane Street bleeding-edge OCaml compiler has support
+C code.\footnote{Ideally in the future, one would not need to call out to C for this: the Jane Street bleeding-edge OCaml compiler has support
 for unboxed types~\cite{unboxedtypes}, which (among other things) would let us implement a version of SplitMix that does not allocate, directly
 in OCaml. Unfortunately, the Jane Street branch of the compiler is incompatible with MetaOCaml, which we use to implement the metaprogramming
-discussed in the previous sections.}.
+discussed in the previous sections.}
 
@@
 a large number of trials. We normalize by computing the relative performance, or
-the ``speedup''. This works in most cases, but it is not perfect.  For example,
+the ``speedup.'' This works in most cases, but it is not perfect.  For example,
 it is theoretically possible to choose a seed such that the \textit{first} value

### 2025-03-24 paper/paper.tex
hyphens

@@
 it on hundreds or thousands of random test inputs. These test inputs are usually
-produced by a {\em generator} --- a program written in some domain-specific
+produced by a {\em generator}---a program written in some domain-specific
 language (DSL) that allows the developer to express precisely how values should
@@
 
-It might be surprising to readers that this dramatic overhead exists --- shouldn't the compiler perform
+It might be surprising to readers that this dramatic overhead exists---shouldn't the compiler perform
 this simple optimization? Compilers of effectful and strict functional languages (including JIT
@@
 
-Last, many PBT libraries --- including both \bq and ScalaCheck --- implement their \texttt{weighted\_union} combinators
+Last, many PBT libraries---including both \bq and ScalaCheck---implement their \texttt{weighted\_union} combinators
 in ways that are asymptotically more efficient but slower in common cases than a more naive algorithm.
@@
 The randomness library is the hottest part of the hot path.
-Indeed, even basic generators --- like ones generating a single \texttt{int} or \texttt{float} uniformly within a range --- can sample \emph{unboundedly many}
+Indeed, even basic generators---like ones generating a single \texttt{int} or \texttt{float} uniformly within a range---can sample \emph{unboundedly many}
 random numbers, since they usually use versions of rejection sampling~\cite{vn51} to find a value within the range.
@@
 that \texttt{f} defines has been eliminated.
-A code-transforming function \texttt{'a code -> 'b code} can also be converted \emph{code for a function} --- a value of type
+A code-transforming function \texttt{'a code -> 'b code} can also be converted \emph{code for a function}---a value of type
 \texttt{('a -> 'b) code}---with the following program:
@@
 % Deciding which types \texttt{t} can
-% instead be \texttt{t code} --- in other words,
+% instead be \texttt{t code}---in other words,
 % figuring out which parts of the DSL can be determined statically (and can be
 % part of the compile-time stage), and which parts are only known dynamically (and
-% hence must be \texttt{code}) --- is an art known as ``binding-time analysis''
+% hence must be \texttt{code})---is an art known as ``binding-time analysis''
 
@@
 Given any size and random seed, it returns a code block that calls \texttt{SR.int} with that random seed. Because the lower and upper bounds
-might not be known at compile time --- they may themselves be the results of calling \texttt{Gen.int} ---
-the arguments \texttt{lo} and \texttt{hi} are of type \texttt{int code} and get spliced into the code block as arguments to \texttt{SR.int}.
+might not be known at compile time---they may themselves be the results of calling \texttt{Gen.int}---the arguments 
+\texttt{lo} and \texttt{hi} are of type \texttt{int code} and get spliced into the code block as arguments to \texttt{SR.int}.
 Lastly, \texttt{to\_bq} turns a staged generator into code for a normal \bq\
@@
 In Section~\ref{section:motiv}, we noted that generator combinators like \texttt{weighted\_union}
-allocate lists in the hot path of the generator. Even though these lists are usually small ---
-at most a few dozen elements --- each allocation takes us closer to the next garbage collection, which is bad for performance.
+allocate lists in the hot path of the generator.
+Even though these lists are usually small---at most a few dozen elements---each allocation takes us closer to the next garbage collection, which is bad for performance.
 
@@
 Crucially, it takes a \emph{compile-time} list \texttt{weighted\_gens} of
-generators and weights. The weights themselves might only be known at run time --- it is
-common to use the current size parameter as a weight, for instance ---
-so they are \texttt{code}s.
+generators and weights. The weights themselves might only be known at run time---it is
+common to use the current size parameter as a weight, for instance---so they are \texttt{code}s.
 Instead of building a data structure representing a histogram of the distribution described by the weights at run time and then traversing it,
@@
 
-% \jwc{Describe split --- this section is approximately experts-only, but you can just point at the Andras paper and the things it cites.}
+% \jwc{Describe split---this section is approximately experts-only, but you can just point at the Andras paper and the things it cites.}
 
@@
 in languages outside of the Haskell world from which it came~\cite{bq,sc,hypothesis}.
-For this reason, readers who are users and developers of PBT libraries in \emph{other} languages ---
-including non-strict or procedural ones --- might be curious about how the
+For this reason, readers who are users and developers of PBT libraries in \emph{other} languages---including non-strict or procedural ones---might be curious about how the
 techniques brought to bear on OCaml's \bq{} and Scala's ScalaCheck in this paper
@@
   However, the racket macros literature, while extensive, does not usually concern itself with staging for performance purposes.
-  For this reason, we --- the authorship team of this paper, devoid of much Racket expertise --- decided to not use Racket as a test case in this paper.
+  For this reason, we---the authorship team of this paper, devoid of much Racket expertise---decided to not use Racket as a test case in this paper.
 
@@
   indeed, it is host to a number of property-based testing libraries~\cite{Bolero,ProptestBook}. However Rust does not
-  structurally encourage monadic eDSLs --- it does not have a special monadic syntax --- and so
+  structurally encourage monadic eDSLs---it does not have a special monadic syntax---and so
   PBT libraries in Rust ask programmers to define generators directly as \texttt{seed -> value} functions.

### 2025-03-24 paper/paper.tex
two

@@
 However, careful measurements show that the generator performance of
-two widely used PBT libraries falls well short of what is possible,
+widely used PBT libraries falls well short of what is possible,
 due principally to (1) the abstraction overhead of their

### 2025-03-24 paper/paper.tex
two

@@
 Second, the calls a generator makes to its \rand{} can constitute a large
-proportion of its run-time, magnifying the cost of an inefficient
+proportion of its run time, magnifying the cost of an inefficient
 implementation.
@@
 a way that is simple to fix, we perform a controlled experiment, comparing fast and slow
-versions of the same library to analyze the impact of faster \rands{} on
+versions of the library to analyze the impact of faster \rands{} on
 generator performance.
@@
 Moreover, both \bq{} and ScalaCheck allocate auxiliary data structures (an array in \bq{} and a BST in ScalaCheck) to
-perform this search, which incurs further run time overhead.
+perform this search, which incurs further run-time overhead.
 
@@
 For example, consider \texttt{Gen.bind' (Gen.int .<0>. .<1>.) (fun x -> Gen.return .<(.~x,.~x)>.)}.
-This generates the run time code \texttt{fun size random -> (SR.int random 0 1, int SR.random 0 1)},
+This generates the run-time code \texttt{fun size random -> (SR.int random 0 1, int SR.random 0 1)},
 which is incorrect.  Instead of generating a single integer and returning it
@@
 time to produce a specialized generator; and finally, the generator is
-executed at runtime to produce values.
+executed at run time to produce values.
 
@@
 comparison.  Similarly, we exclude tasks where the \bq strategy completes in
-under 5ms, as such negligible runtimes do not yield meaningful insights into
+under 5ms, as such negligible run times do not yield meaningful insights into
 relative performance. Across all 30 seeds, these filters remove

### 2025-03-25 paper/paper.tex
paper

@@
 Second, the calls a generator makes to its \rand{} can constitute a large
-proportion of its run time, magnifying the cost of an inefficient
+proportion of its run-time, magnifying the cost of an inefficient
 implementation.
@@
 The \texttt{gen\_int} combinator simply calls out to the randomness library \texttt{Splittable\_random}, aliased as \texttt{SR} here.
-Different generator DSLs use variatoins on this basic design, but the basics are the same across the board.
+Different generator DSLs use variations on this basic design, but the basics are the same across the board.
 
@@
 while existing PBT libraries by and large make sensible choices for their randomness libraries,
-they are not chosen with performance in mind, leading to worse bugfinding power than what is possible.
+they are not chosen with performance in mind, leading to worse bug-finding power than what is possible.
 
@@
 generation; for that, we need an experiment. The most obvious experiment is to
-simply swap out the randomness library of either \bq{} or ScalaCheck one of the afformentioned faster algorithms.
+simply swap out the randomness library of either \bq{} or ScalaCheck one of the aforementioned faster algorithms.
 But, as discussed previously, \camlname{} and \scalaname{} are designed to be 100\% semantically equivalent replacements for
@@
 to be defined as recursive functions, while others
-(including both \bq and ScalaCheck) expose a fixpoint combinator to construct recursive generators.
+(including both \bq and ScalaCheck) expose a fixed-point combinator to construct recursive generators.
 Given a step function that takes a ``handle'' to sample from a recursive generator call, it
@@
 generators can implement a staged version using a similar implementation to the original. In Scala,
-for instance, the same strategy could be realized using typeclass
+for instance, the same strategy could be realized using type class
 resolution. In both settings, the result is a three-stage process: a
-metaprogram---via PPX or typeclasses---constructs a generator expression
+metaprogram---via PPX or type classes---constructs a generator expression
 composed of staged combinators; this expression is evaluated at compile
@@
 another impediment to finding bugs fast.
-To demonstrate that faster random sampling can significanty impact bugfinding power, we
+To demonstrate that faster random sampling can significantly impact bug-finding power, we
 use the controlled experiment suggested by OCaml's inefficient implementation of
 SplitMix. By replacing this relatively slow randomness library with a faster but semantically equivalent
-implemetnation, we can precisely quantify the bugfinding speedup that a better randomness library gives across a range of PBT scenarios.
+implementation, we can precisely quantify the bug-finding speedup that a better randomness library gives across a range of PBT scenarios.
 
@@
 \camlname version using \csplitmix. For each treatment, we measure the
-time to generate a value (i.e., a BST, STLC term, or boolean list), using a random seed, varying generation sizes (10, 100, 1,000, 10,000).\footnote{The specific meaning of generation size is a domain-specific implementation detail---in a list generator, size might correspond to the desired length of the list, whereas in a tree generator it refer to number of nodes, number of leaves, depth of tree, etc.
-Regardless, our goal is to show that performance trends scale.}
+time to generate a value (i.e., a BST, STLC term, or boolean list), using 
+a random seed, varying generation sizes (10, 100, 1,000, 10,000).\footnote{The 
+specific meaning of generation size is a domain-specific implementation 
+detail---in a list generator, size might correspond to the desired length of the list, 
+whereas in a tree generator it refer to number of nodes, number of leaves, depth of tree, 
+etc. Regardless, our goal is to show that performance trends scale.}
 We run each treatment for 5 seconds and compute the average generation time of a value produced in that interval.
@@
 For artifact evaluation we will submit (1) the source code of \camlname{} and \scalaname{} as well as (2) the code for our version of Etna,
-which we have modified slightly to pass deterministic seeds to all versions of a single strategy.
+which we have modified slightly to (a) support parallelism, and (b) pass deterministic seeds to all versions of a single strategy.
 These programs are not available publicly or anonymized at the time of submission.

### 2025-03-25 paper/paper.tex
scoot

@@
 
+\lstset{xleftmargin=0pt}
 %%
@@
 \begin{figure}[h]
+  \begin{tabular}{lll}
+  Library    & Generator          & Average Time per Generation (ns) \\
+  \bq{}         & \texttt{int\_pair}          & 70                       \\
+  \bq{}         & \texttt{int\_pair\_inlined} & 35                       \\
+  ScalaCheck & \texttt{intPair}            & 458                      \\
+  ScalaCheck & \texttt{intPairInlined}     & 266
+  \end{tabular}
+\caption{Microbenchmarks of generators in \bq{} and ScalaCheck. Average over 10000 generations with random seeds and a fixed size.}
+\label{fig:overhead-explanation-perf}
+\end{figure}
+
+The performance impact of this simplification is large (Figure~\ref{fig:overhead-explanation-perf}). In both languages, the
+inlined version takes on average half as much time to generate a single pair of
+\texttt{int}s.
+Microbenchmarks of more realistic generators (see Section~\ref{section:eval})
+show an even more dramatic performance boost.
+Because the inlined versions of the generator are identical to the un-inlined
+versions except for mechanical, semantics-preserving transformations,
+this performance difference is attributable solely to
+the different machine code generated by the compiler (and not lucky choice of random seeds).
+
+\begin{figure}[h]
   \begin{subfigure}{.49\textwidth}
@@
 
-def intPairInlined : (Gen.Parameters, Seed) => (Option[(Long,Long)],Seed) = {
+def intPairInlined : (Gen.Parameters, Seed) => 
+(Option[(Long,Long)],Seed) = {
  (p,seed) =>
@@
 
-The performance impact of this simplification is large (Figure~\ref{fig:overhead-explanation-perf}). In both languages, the
-inlined version takes on average half as much time to generate a single pair of
-\texttt{int}s.
-Microbenchmarks of more realistic generators (see Section~\ref{section:eval})
-show an even more dramatic performance boost.
-Because the inlined versions of the generator are identical to the un-inlined
-versions except for mechanical, semantics-preserving transformations,
-this performance difference is attributable solely to
-the different machine code generated by the compiler (and not lucky choice of random seeds).
-
 It might be surprising to readers that this dramatic overhead exists---shouldn't the compiler perform
@@
 conservative for optimizing recursive functions.
-
-\begin{figure}
-  \begin{tabular}{lll}
-  Library    & Generator          & Average Time per Generation (ns) \\
-  \bq{}         & \texttt{int\_pair}          & 70                       \\
-  \bq{}         & \texttt{int\_pair\_inlined} & 35                       \\
-  ScalaCheck & \texttt{intPair}            & 458                      \\
-  ScalaCheck & \texttt{intPairInlined}     & 266
-  \end{tabular}
-\caption{Microbenchmarks of Generators in \bq{} and ScalaCheck. Average over 10000 generations with random seeds and a fixed size.}
-\label{fig:overhead-explanation-perf}
-\end{figure}
 
@@
   \begin{subfigure}{.45\textwidth}
-\begin{lstlisting}
+\begin{lstlisting}[xleftmargin=0pt, xrightmargin=-3pt, aboveskip=0pt, belowskip=0pt]
 let bind (g : 'a t) (k : 'a code -> 'b t) : 'b t =
@@
   \begin{subfigure}{.45\textwidth}
-\begin{lstlisting}
+\begin{lstlisting}[xleftmargin=0pt, xrightmargin=-27pt, aboveskip=0pt, belowskip=0pt]
 let bind' (g : 'a t) (k : 'a code -> 'b t) : 'b t =

### 2025-03-25 paper/paper.tex
figs

@@
 
-\begin{wrapfigure}{r}{.5\textwidth}
+\begin{figure}[h]
 \begin{lstlisting}
@@
 \label{fig:bq-api}
-\end{wrapfigure}
+\end{figure}
 
@@
 
-\begin{wrapfigure}{r}{.5\textwidth}
-\begin{lstlisting}
+\begin{wrapfigure}{r}{.49\textwidth}
+  \vspace{1.5em}
+\begin{lstlisting}[
+    xleftmargin=0pt,
+    xrightmargin=0pt,
+    aboveskip=0pt,
+    belowskip=0pt,
+    basicstyle=\ttfamily\footnotesize,
+    lineskip=0pt
+  ]
 let tree_of g = fixed_point (fun rg ->
@@
 found in Figure~\ref{fig:gen-staged-basic}.
+
+\begin{figure}[h]
+  \begin{lstlisting}
+  module Gen = struct
+    type 'a t = int code -> Random.t code -> 'a code
+
+    let return (cx : 'a code) : 'a t = fun size random -> cx
+
+    let bind (g : 'a t) (k : 'a code -> 'b t) : 'b t =
+      fun size random ->
+        .<
+          let a = .~(g size random) in
+          .~(k .<a>. size random)
+        >.
+
+    let int (lo : int code) (hi : int code) : int t =
+      fun size random ->
+        .< SR.int .~random .~lo .~hi >.
+
+    let to_bq (g : 'a code Gen.t) : ('a Bq.t) code =
+      .<
+        fun size random -> .~(g .<size>. .<random>.)
+      >.
+  end
+  \end{lstlisting}
+  \caption{Basic Staged Generator Library\ifaftersubmission\bcp{We could
+      save some space if needed by eliminating a couple of linebreaks
+      here.}\fi}
+  \label{fig:gen-staged-basic}
+  \end{figure}
+
 The constant generator \texttt{return} runs at compile time. Given \texttt{cx : 'a code}, the code for an \texttt{'a}, it
@@
 
-\begin{figure}
-\begin{lstlisting}
-module Gen = struct
-  type 'a t = int code -> Random.t code -> 'a code
-
-  let return (cx : 'a code) : 'a t = fun size random -> cx
-
-  let bind (g : 'a t) (k : 'a code -> 'b t) : 'b t =
-    fun size random ->
-      .<
-        let a = .~(g size random) in
-        .~(k .<a>. size random)
-      >.
-
-  let int (lo : int code) (hi : int code) : int t =
-    fun size random ->
-      .< SR.int .~random .~lo .~hi >.
-
-  let to_bq (g : 'a code Gen.t) : ('a Bq.t) code =
-    .<
-      fun size random -> .~(g .<size>. .<random>.)
-    >.
-end
-\end{lstlisting}
-\caption{Basic Staged Generator Library\ifaftersubmission\bcp{We could
-    save some space if needed by eliminating a couple of linebreaks
-    here.}\fi}
-\label{fig:gen-staged-basic}
-\end{figure}
-
 Returning to our running example, Figure~\ref{fig:running-staged} shows the
@@
 
-\begin{figure}
+\begin{figure}[h]
 \begin{lstlisting}
@@
 
+\begin{figure}[h]
+  \begin{lstlisting}
+  module CodeCps = struct
+    type 'a t = { cps : 'z. ('a -> 'z code) -> 'z code }
+
+    let return x = {cps = fun k -> k x}
+
+    let bind (x : 'a t) (f : 'a -> 'b t) : 'b t =
+      {cps = fun k -> x.cps (fun a -> (f a).cps k)}
+
+    let run (t : ('a code) t) : 'a code = t.cps (fun x -> x)
+
+    let let_insert (cx : 'a code) : 'a code t =
+      {cps = fun k -> k .< let x = .~cx in .~(k .<x>.) >.}
+  end
+
+  module Gen = struct
+    type 'a t = int code -> SR.t code -> 'a CodeCps.t
+
+    let return (x : 'a) : 'a t = fun _ _ -> Codecps.return x
+
+    let bind (g : 'a t) (f : 'a -> 'b t) =
+      fun size random ->
+        CodeCps.bind (g size random) (fun x -> (f x) size random)
+
+    let int (lo : int code) (hi : int code) : int code t =
+      fun size random -> let_insert .< SR.int .~random .~lo .~hi >.
+  end
+  \end{lstlisting}
+  \caption{CodeCPS and the Final Gen Monad}
+  \label{fig:codecps-and-final-gen}
+  \end{figure}
+
 In Figure~\ref{fig:codecps-and-final-gen}, following prior work \cite{kovacs24, carrette05}, we define
@@
 
-\begin{figure}
-\begin{lstlisting}
-module CodeCps = struct
-  type 'a t = { cps : 'z. ('a -> 'z code) -> 'z code }
-
-  let return x = {cps = fun k -> k x}
-
-  let bind (x : 'a t) (f : 'a -> 'b t) : 'b t =
-    {cps = fun k -> x.cps (fun a -> (f a).cps k)}
-
-  let run (t : ('a code) t) : 'a code = t.cps (fun x -> x)
-
-  let let_insert (cx : 'a code) : 'a code t =
-    {cps = fun k -> k .< let x = .~cx in .~(k .<x>.) >.}
-end
-
-module Gen = struct
-  type 'a t = int code -> SR.t code -> 'a CodeCps.t
-
-  let return (x : 'a) : 'a t = fun _ _ -> Codecps.return x
-
-  let bind (g : 'a t) (f : 'a -> 'b t) =
-    fun size random ->
-      CodeCps.bind (g size random) (fun x -> (f x) size random)
-
-  let int (lo : int code) (hi : int code) : int code t =
-    fun size random -> let_insert .< SR.int .~random .~lo .~hi >.
-end
-\end{lstlisting}
-\caption{CodeCPS and the Final Gen Monad}
-\label{fig:codecps-and-final-gen}
-\end{figure}
-
 % \subsubsection{The Trick at Other Types}

### 2025-03-25 paper/paper.tex
nit

@@
 
-In practice, \texttt{weighted\_union} is (almost \footnote{
+In practice, \texttt{weighted\_union} is (almost\footnote{
 There are some generators where \texttt{weighted\_union} is passed a list which

### 2025-03-25 paper/paper.tex
a -> an

@@
 opposed to distributional---comparisons of bug-finding effectiveness,
-ensuring that, if a \name{} generator finds bugs faster, this is
+ensuring that, if an \name{} generator finds bugs faster, this is
 attributable \textit{solely} to the staging and \rand interventions,

### 2025-03-25 paper/paper.tex
indentation

@@
 Figure~\ref{fig:bq-api}.  The \texttt{weighted\_union} function is a
-particularly well-used one: it makes making a weighted choice between different
+particularly well-used one: it makes a weighted choice between different
 generators, allowing the developer to combine different sub-generators into a
@@
 Figure~\ref{fig:simplify}, the indirection of \texttt{return}
-and \texttt{bind} causes these heuristics to not fire.  Moreover, the story is
+and \texttt{bind} causes these heuristics to not fire.  The story is
 even worse for recursive generators, as the heuristics are necessarily even more
@@
 each individual closure allocation is relatively cheap; but doing lots of allocation in a
-generator is very costly because each allocation brings us closer to the next costly GC pause.
-This effect is also magnified in recursive generators: each iteration
+generator is very expensive because each allocation brings us closer to the next costly GC pause.
+This effect is magnified in recursive generators: each iteration
 through the recursive loop re-allocates closures for \texttt{bind}s, so the amount of allocation
@@
 Like \texttt{return} and \texttt{bind}, combinators like \texttt{weighted\_union} incur a performance penalty
-at run time. Aside from the previously-discussed issues that compilers
+at run time. Aside from the previously-discussed issue that compilers
 cannot see through the abstraction boundary to optimize these programs,
@@
 The core of any PBT generator library is a source of randomness.
-Different PBT libraries use different randomness libraries implementing different algorithms.
-\footnote{The common term for such an algorithm or library is a ``Random Number
+Different PBT libraries use different randomness libraries implementing different algorithms.\footnote{The common term for such an algorithm or library is a ``Random Number
 Generator'' (RNG). We will avoid this term and instead say ``randomness library''
@@
 By building a version that uses much faster ``unboxed'' 64-bit integer
-arithmetic, we can (a) demonstrate just how much faster bugs can be found just by
-using a more performant \rand{}, and (b) explore which generators benefit the most from faster \rand{}.
+arithmetic, we (a) demonstrate how much faster bugs can be found just by
+using a more performant \rand{}, and (b) explore which generators benefit the most from faster \rands{}.
 

### 2025-03-25 paper/paper.tex
daniel comments

@@
 \texttt{bind e (fun x => e')}. Figure~\ref{fig:simple-macro} shows the same
-generator, written the monadic syntax.
+generator, written in the monadic syntax.
 
@@
 and returning a generated value \texttt{'a}.
-The random seed \texttt{SR.t} from \bq's \rand{} called \texttt{Splittable\_random}, henceforth aliased in code as \texttt{SR}.
+The random seed \texttt{SR.t} from \bq's \rand{} is called \texttt{Splittable\_random}, henceforth aliased in code as \texttt{SR}.
 It is an invariant of the generator library that every function of this type is
@@
   \end{tabular}
-\caption{Microbenchmarks of generators in \bq{} and ScalaCheck. Average over 10000 generations with random seeds and a fixed size.}
+\caption{Microbenchmarks of generators in \bq{} and ScalaCheck. Average over 10,000 generations with random seeds and a fixed size.}
 \label{fig:overhead-explanation-perf}
@@
 
-In multi-stage programming (or simply ``staging''), programs execute
+In multi-stage programming, or simply ``staging,'' programs execute
 in multiple stages, with each stage producing code to be run
-in the next. For the purposes of this paper, we just need two stages: compile time and run time.
+in the next. For the purposes of this paper, we need just two stages: compile time and run time.
 
@@
 that \texttt{f} defines has been eliminated.
-A code-transforming function \texttt{'a code -> 'b code} can also be converted \emph{code for a function}---a value of type
+A code-transforming function \texttt{'a code -> 'b code} can also be converted to \emph{code for a function}---a value of type
 \texttt{('a -> 'b) code}---with the following program:
@@
 
-This gives us the best of both worlds. First, we get a monad instance for \texttt{'a Gen.t} with the correct types, which lets us
+This approach gives us the best of both worlds. First, we get a monad instance for \texttt{'a Gen.t} with the correct types, which lets us
 use the monadic syntactic sugar of our chosen language. Moreover, we also get to maintain the correct
@@
 and only a subset of \textit{those} will find a bug. As a result, many more values must be produced,
-making it important to do so as quickly as possible.
+making generation speed particularly crucial.
 
@@
   using macros to build small eDSLs like the ones we use to write generators, and so it seems a natural fit.
-  However, the racket macros literature, while extensive, does not usually concern itself with staging for performance purposes.
+  However, the Racket macros literature, while extensive, does not usually concern itself with staging for performance purposes.
   For this reason, we---the authorship team of this paper, devoid of much Racket expertise---decided to not use Racket as a test case in this paper.

### 2025-03-25 paper/paper.tex
editing pass complete

@@
 In the staging literature, this task is known as ``binding-time
-analysis,'' \cite{jones1993partial}.
+analysis\cite{jones1993partial}.''
 
@@
 Lastly, \texttt{to\_bq} turns a staged generator into code for a normal \bq\
-generator. This is just a 2-argument version of ``The Trick''
+generator. This is just a two-argument version of ``The Trick''
 from Section~\ref{subsection:msp}.
@@
 allocate lists in the hot path of the generator.
-Even though these lists are usually small---at most a few dozen elements---each allocation takes us closer to the next garbage collection, which is bad for performance.
+Even though these lists are usually small---at most a few dozen elements---each allocation takes us closer to the next garbage collection\proposecut{, which is bad for performance}.
 
@@
 the functions \texttt{return (x : 'a) = fun \_ \_ -> x} and \texttt{bind g k = fun size seed -> k (g size seed) size seed}
-have the proper types for a monad instance. Then, any combinators of type \texttt{'a Gen.t} before simply become \texttt{'a code Gen.t} with this new version.
+have the proper types for a monad instance. Then, any combinators of type \texttt{'a Gen.t} become \texttt{'a code Gen.t} in this new version.
 
@@
 In our setting, letting programmers define recursive generators as recursive functions is out of the question.
-With staged programming, recursion must be handled with care: it is far to easy to accidentally recursively
-define an infinite \texttt{code} value and have the program diverge at compile time, when trying to write
-a \texttt{code} representing a recursive program.
-To this end, we develop a staged recursive generator combinator.\footnote{
+With staged programming, recursion must be handled with care: it is far too easy to accidentally recursively
+define an infinite \texttt{code} value and have the program diverge at compile time\proposecut{, when trying to write
+a \texttt{code} representing a recursive program.}
+To this end, we develop a staged recursive generator combinator\footnote{
 We actually provide a more general API that allows programmers to define \emph{parameterized}
 recursive combinators, of type \texttt{'r code -> 'a code Gen.t}, for any type \texttt{'r}.
-}, whose API is shown in Figure~\ref{fig:staged-recursive-generator}
+}, whose API is shown in Figure~\ref{fig:staged-recursive-generator}.
 The recursion API consists of an opaque type \texttt{'a handle}, and a function \texttt{recurse} to perform recursive calls.
@@
 
-Perhaps the best-explored approaches to speeding up bug-finding with PBT is by
-changing the order in which test inputs are sampled. Feedback-driven mechanisms
+Reordering test input sampling is perhaps the most well-studied way to accelerate bug-finding in PBT.
+ Feedback-driven mechanisms
 like Targeted PBT~\cite{loscherTargetedPropertybasedTesting2017} and

### 2025-03-25 paper/paper.tex
A few quick fixes, plus polishing on last two sections

@@
 \end{lstlisting}
-\caption{Simplifying a Generator in OCaml}\label{fig:simplify-ocaml}
+\caption{Simplifying a generator in OCaml}\label{fig:simplify-ocaml}
   \end{subfigure}
@@
 \end{lstlisting}
-\caption{Simplifying a Generator in Scala}\label{fig:simplify-scala}
+\caption{Simplifying a generator in Scala}\label{fig:simplify-scala}
   \end{subfigure}
@@
 \subsection{Benchmarking bug-finding speed}
+\bcp{Make sure that capitalization of section titles is consistent}
 To answer \textbf{RQ2}, we evaluate the bug-finding speed of our BST and STLC
@@
 For this reason, readers who are users and developers of PBT libraries in \emph{other} languages---including non-strict or procedural ones---might be curious about how the
-techniques brought to bear on OCaml's \bq{} and Scala's ScalaCheck in this paper
+techniques brought to bear here on OCaml's \bq{} and Scala's ScalaCheck
 might be imported to their favorite language.
@@
   \textbf{Haskell} is home to the original PBT implementation, and
-  so it is natural to ask why we have not built a \name{} in it yet. The basic
+  so it is natural to ask why we have not applied \name{} to it yet. The basic
   answer is that the Haskell compiler (GHC) is designed specially to eliminate
@@
   its normal optimiation steps.  This means that in many cases, the impact of a
-  staged PBT library (and indeed, all staged monadic DSLs) negligible.  For
-  complex enough programs however, the heuristics GHC uses can degrade, leading
+  staged PBT library (and indeed, all staged monadic DSLs) is negligible.  For
+  complex enough programs, however, the heuristics GHC uses can degrade, leading
   to performance overheads for monadic code.  For this reason, Haskell does have
-  a multi-stage programming system called Template Haskell~\cite{templatehaskell}, which is
+  a multi-stage programming system, Template Haskell~\cite{templatehaskell}, that is
   sometimes to build eDSLs in contexts where one does not want to rely on the
-  optimizer's heuristics.  In short: it is possible to build a \name{} version of Haskell
+  optimizer's heuristics.  In short: it is possible to build an \name{} version of Haskell
   QuickCheck, but the benefits would be much less pronounced.
@@
   \textbf{F\#} has both multi-stage programming capabilities~\cite{fsmeta} and a well-used PBT library called FsCheck~\cite{fscheck}.
-  The internals of FsCheck's generator eDSL are very similar to both \bq and ScalaCheck, so we expect that building an \name{}
-  in F\# is primarily an engineering effort.
+  The internals of FsCheck's generator eDSL are very similar to both
+  \bq and ScalaCheck, so we expect that building \name{}-style generators
+  in F\# should be a straightforward engineering effort.
   % https://learn.microsoft.com/en-us/dotnet/fsharp/language-reference/code-quotations
 
-  \textbf{Rust} shares some similarities with functional languages, and
-  indeed, it is host to a number of property-based testing libraries~\cite{Bolero,ProptestBook}. However Rust does not
+  \textbf{Rust} shares some similarities with functional languages, and,
+  indeed, it features a number of property-based testing libraries~\cite{Bolero,ProptestBook}. However, Rust does not
   structurally encourage monadic eDSLs---it does not have a special monadic syntax---and so
   PBT libraries in Rust ask programmers to define generators directly as \texttt{seed -> value} functions.
-  For this reason, staging does not seem directly applicable to any of the PBT libraries in Rust.
+  For this reason, staging does not seem directly applicable to any of
+  the PBT libraries in Rust.
 
@@
 
-Our work is unique in its approach to speeding up PBT, but it is certainly not
-the only work focused on making PBT faster.
-
-Reordering test input sampling is perhaps the most well-studied way to accelerate bug-finding in PBT.
+Our work is unique in its approach to speeding up
+PBT, but it is certainly not the only work focused on making PBT faster.
+
+Reordering\bcp{``reordering'' sounds like rearranging a deterministic
+  order; this is changing the shape of a distribution, right?} test input sampling is perhaps the most well-studied way to accelerate bug-finding in PBT.
  Feedback-driven mechanisms
 like Targeted PBT~\cite{loscherTargetedPropertybasedTesting2017} and
-Coverage-Guided PBT~\cite{lampropoulosCoverageGuidedProperty2019} have the
-potential to speed up testing by changing the generation order to try to find
+Coverage-Guided PBT~\cite{lampropoulosCoverageGuidedProperty2019}
+speed up testing by changing the generation order to find
 ``interesting'' inputs faster. Separately, enumerative approaches to
 PBT~\cite{runcimanSmallcheckLazySmallcheck2008,braquehaisSimpleIncrementalDevelopment2017}
-try to get to bugs faster by leveraging the ``small-scope hypothesis'' and
-attempting to quickly exhaust the list of small inputs to the program under
-test. All of these approaches can have significant performance benefits in
+try to get to bugs faster by leveraging the ``small-scope
+hypothesis,'' which posits that most bugs will be triggered by smallish
+inputs. All of these approaches can have significant performance
+benefits in
 practice, but they are largely orthogonal to our contributions; we expect
-\name{} could be used to speed these existing techniques to varying degrees.
+\name{} could be used to speed these existing techniques to varying
+degrees. \bcp{Namely: the staging part would be equally effective, but
+the RNG part would be either less or not relevant.}
 
 Similarly orthogonal to our approach are techniques for quickly filtering out
-inputs that are invalid for testing. Some of these approaches work statically,
-by automatically deriving generators that produce valid inputs by
+invalid inputs. Some of these approaches work statically,
+by automatically deriving generators that produce only valid inputs by
 construction~\cite{lampropoulosGeneratingGoodGenerators2017}, while others
@@
 problems~\cite{seidelTypeTargetedTesting2015,steinhofelInputInvariants2022,lampropoulosBeginnerLuckLanguage2017}.
-In all cases, these approaches could be further improved by insights from \name.
+In all cases, these approaches could be further improved using
+techniques from \name. \bcp{Why not also RNG improvements?  (I think
+  \name{} refers {\em only} to the staging part of what we've done,
+  but I'm not certain about that, and I'm not certain we've been consistent...)}
 
@@
 designed with performance engineering in mind, taking seriously the idea that
-PBT is a performance critical task. Still, despite their similar motivations and
-considerations, the solutions are entirely different and complementary.
+PBT is a performance critical task. Still, despite their similar
+motivations, the solutions are entirely different and complementary.
 
@@
 unified representation of data and code allows for sophisticated
-metaprogramming. Quasiquation-style macros were introduced to the ML family of
+metaprogramming. Quasiquotation-style macros were introduced to the ML family of
 languages by Nielson and Nielson's Two-Level Functional Languages~\cite{nielson92} and
-MetaML~\cite{taha00}, which in turn quickly inspired MetaOCaml~\cite{metaocaml} (and more recently
-MacoCaml~\cite{macocaml}) in OCaml, Lightweight Modular Staging in Scala~\cite{lms}, Template
+MetaML~\cite{taha00}, which in turn quickly inspired MetaOCaml~\cite{metaocaml} and more recently
+MacoCaml~\cite{macocaml}, Lightweight Modular Staging in Scala~\cite{lms}, Template
 Haskell in Haskell~\cite{templatehaskell}, and more.  In addition to implementations, researchers
@@
 line of work studying how macros and metaprogramming can be used to build extensible
-DSLs \cite{flatt12, samth11, skthesis}.
-
-Back in the ML-family world, the primary use of staging is building embedded DSLs
-with minimal performance overhead, an idea which has come to be known as ``abstraction without regret''
+DSLs \cite[etc.]{flatt12, samth11, skthesis}.
+
+In ML-like languages, the primary use of staging is building embedded DSLs
+with minimal performance overhead, an idea that has come to be known as ``abstraction without regret''
 \cite{lms,parreaux17,parreauxthesis, sheard99, trattdsl}.
 This technique has been applied across a range of domains, including
-Big data processing \cite{jet12}, stream processing \cite{moller20,strymonas},
+big-data processing \cite{jet12}, stream processing \cite{moller20,strymonas},
 query processing \cite{rhyme}, and parser combinators \cite{sspc, staged-parsers, krishnaswami19, flap}.
@@
 as ``parsers of randomness'' \cite{goldsteinParsingRandomness2022}.
-However, this equivalence is primarily theoretical. PBT generators are (a) not actually implemented as parsers
-of random sequences, and (b) support choice-based combinators like \texttt{weighted\_union} that parsers do not.
-For this reason, we must build staged generator libraries from scratch and cannot simply use existing staged parser libraries.
+However, this equivalence is primarily theoretical. PBT generators (a)
+are not actually implemented as parsers
+of random sequences and (b) support choice-based combinators like \texttt{weighted\_union} that parsers do not.
+For this reason, we must build staged generator libraries from scratch
+and cannot simply use existing staged parser libraries.  \bcp{Maybe
+  cut that last sentence?}
 
@@
 \label{section:concl}
-In this paper, we identified, studied, and proposed potential solutions to two
+We have identified, studied, and proposed general solutions to two
 important sources of inefficiency in PBT generator libraries: abstraction
-overhead and choice of \rand{}.  In the future, we hope to continue to investigate and push the boundaries of
+overhead and choice of \rand{}.
+
+In the future, we hope to continue to investigate and push the boundaries of
 PBT generator performance.
-
+%
 On the abstraction overhead side, we hope to employ some of the many tricks for better code generation that have been written
@@
 which we did not use in \camlname{}.
-Another trick we hope to use is GADT-based techniques for unpacking the states of recursive functions. Currently, the \camlname{}
+Another trick we hope to use is GADT-based techniques for unpacking the states of recursive functions\bcp{citation?}. Currently, the \camlname{}
 recursion combinator introduces some overhead from boxing and unboxing the accumulator values at each recursive call. As noticed in other work \cite{kovacs24}, this
@@
 
-On the \rands{} side we plan to investigate fast \rands{} that are not equivalent to SplitMix, such as Lehmer~\cite{lehmerrng}, WyHash~\cite{wyrand},
+On the randomness side we plan to investigate fast \rands{} that are not equivalent to SplitMix, such as Lehmer~\cite{lehmerrng}, WyHash~\cite{wyrand},
 and variants of xoroshiro~\cite{xorshiro}. We also want to try techniques to speed up sampling such as pipelining, using SIMD instructions,
-or even generating random numbers ahead of time. Last, we hope to investigate the degree to which actual statistical randomness
-matters in PBT.
+or generating large buffers of random numbers ahead of time. Finally, we hope to investigate the degree to which statistical randomness
+matters in PBT, since some of these ideas trade statistical guarantees for speed.
 

### 2025-03-25 paper/paper.tex
fixing some bcp comments

@@
 returns the generator that ignores its \texttt{size} and
-\texttt{random} arguments and simply runs\bcp{returns the code for?} \texttt{cx}. Similarly,
+\texttt{random} arguments and simply returns \texttt{cx}. Similarly,
 \texttt{bind g k} sequences generators by passing the result of running the generator \texttt{g} to the continuation \texttt{k}. However, instead of getting access
@@
 In all cases, these approaches could be further improved using
-techniques from \name. \bcp{Why not also RNG improvements?  (I think
-  \name{} refers {\em only} to the staging part of what we've done,
-  but I'm not certain about that, and I'm not certain we've been consistent...)}
+techniques from \name and faster \rands{}.
 
@@
 of random sequences and (b) support choice-based combinators like \texttt{weighted\_union} that parsers do not.
-For this reason, we must build staged generator libraries from scratch
-and cannot simply use existing staged parser libraries.  \bcp{Maybe
-  cut that last sentence?}
 
@@
 which we did not use in \camlname{}.
-Another trick we hope to use is GADT-based techniques for unpacking the states of recursive functions\bcp{citation?}. Currently, the \camlname{}
+Another trick we hope to use is GADT-based techniques for unpacking the states of recursive functions~\cite{kovacs24}. Currently, the \camlname{}
 recursion combinator introduces some overhead from boxing and unboxing the accumulator values at each recursive call. As noticed in other work \cite{kovacs24}, this

### 2025-03-25 paper/paper.tex
section titles
fix bcp nit

@@
 
-Reordering\bcp{``reordering'' sounds like rearranging a deterministic
-  order; this is changing the shape of a distribution, right?} test input sampling is perhaps the most well-studied way to accelerate bug-finding in PBT.
+Changing the shape of the test input distribution is perhaps the most well-studied way to accelerate bug-finding in PBT.
  Feedback-driven mechanisms

### 2025-03-25 paper/paper.tex
wibble

@@
 
-The crux of our binding time analysis\bcp{the wording here is
-  misleading: it feels like we're implying we've {\em implemented} a
-  binding-time analysis...} is that the \emph{only} parts of a
+The crux of our binding time analysis is that the \emph{only} parts of a
 generator that are not
@@
 %}
-\subsection{Benchmarking bug-finding speed}
-\bcp{Make sure that capitalization of section titles is consistent}
+\subsection{Benchmarking Bug-Finding Speed}
+% \bcp{Make sure that capitalization of section titles is consistent}
 To answer \textbf{RQ2}, we evaluate the bug-finding speed of our BST and STLC

### 2025-03-25 paper/paper.tex
last bcp nits

@@
 practice, but they are largely orthogonal to our contributions; we expect
-\name{} could be used to speed these existing techniques to varying
-degrees. \bcp{Namely: the staging part would be equally effective, but
-the RNG part would be either less or not relevant.}
+both \name{} and faster \rands{} could be used to speed these existing techniques to varying
+degrees.
 

### 2025-03-25 paper/paper.tex
nit

@@
 Figure~\ref{fig:staged-weighted-union-example} demonstrates a use of this staged weighted union.
-Given a list of (in this case constant) generators with weights ``the current size parameter'', \texttt{2}, and \texttt{1},
+Given a list of (in this case constant) generators with weights ``the current size parameter,'' \texttt{2}, and \texttt{1},
 the generated code first computes the sum of these numbers, samples between \texttt{0} and the sum, and

### 2025-03-25 paper/paper.tex
TERRIBLE DATA ERROR

@@
 whether they can transform previously intractable tasks into
-tractable ones. Out of 78 tasks that initially timed out,
-30 (38.4\%) completed successfully using \camlname alone, and
-50 (64.1\%) completed successfully after adding \csplitmix.
+tractable ones. Out of 79 tasks that initially timed out,
+31 (38.4\%) completed successfully using \camlname alone, and
+51 (64.1\%) completed successfully after adding \csplitmix.
 This shows that staged type-derived generators can provide

### 2025-03-25 paper/paper.tex
swarm plot labels
final edits

@@
-\documentclass[sigplan,screen,acmsmall,anonymous,review]{acmart}%%
+\documentclass[acmsmall,screen,review,anonymous]{acmart}
 %% \BibTeX command to typeset BibTeX logo in the docs
@@
 allocate lists in the hot path of the generator.
-Even though these lists are usually small---at most a few dozen elements---each allocation takes us closer to the next garbage collection\proposecut{, which is bad for performance}.
+Even though these lists are usually small---at most a few dozen elements---each allocation takes us closer to the next garbage collection, which is bad for performance.
 
@@
 continuation.
-\hg{I'm getting lost in the inline code again}
+% \hg{I'm getting lost in the inline code again}
 
@@
 With staged programming, recursion must be handled with care: it is far too easy to accidentally recursively
-define an infinite \texttt{code} value and have the program diverge at compile time\proposecut{, when trying to write
-a \texttt{code} representing a recursive program.}
+define an infinite \texttt{code} value and have the program diverge at compile time, when trying to write
+a \texttt{code} representing a recursive program.
 To this end, we develop a staged recursive generator combinator\footnote{

### 2025-03-25 paper/paper.tex
nit

@@
 a compiler could in principle unroll the loops in \texttt{weighted\_union} to depth \texttt{n} and specialize the function at each call site to avoid allocating the list.
-Unfortunately, (almost\footnote{GHC is a notable exception here, performing sophisticated list fusion optimizations \cite{gill93}}) no compilers perform this kind of optimization.
+Unfortunately, (almost\footnote{GHC is a notable exception here, performing sophisticated list fusion optimizations\cite{gill93}.}) no compilers perform this kind of optimization.
 This allocation (or rather, its tendency to cause GC pauses)---as well as the cost of running the code

### 2025-03-25 paper/paper.tex
oops, rwo

--- 2025-03-25 paper/paper.tex
@@
 at run time as a \emph{pointer} to an unscanned block of memory containing (among other things)
-a 64-bit integer~\cite{ocamlruntimerep}. This means that all operations that return an \texttt{int64} must allocate this block of memory,
+a 64-bit integer~\cite{realworldocaml}. This means that all operations that return an \texttt{int64} must allocate this block of memory,
 which has a significant impact on performance. A single call to

### 2025-03-25 paper/paper.tex
nit
nits

@@
     Bib\TeX}}}
-
 \newif\ifdraft\drafttrue{}
@@
 a compiler could in principle unroll the loops in \texttt{weighted\_union} to depth \texttt{n} and specialize the function at each call site to avoid allocating the list.
-Unfortunately, (almost\footnote{GHC is a notable exception here, performing sophisticated list fusion optimizations\cite{gill93}.}) no compilers perform this kind of optimization.
+Unfortunately, (almost\footnote{GHC is a notable exception here, performing sophisticated list fusion optimizations \cite{gill93}.}) no compilers perform this kind of optimization.
 This allocation (or rather, its tendency to cause GC pauses)---as well as the cost of running the code
@@
 \ifsilly  \includegraphics[scale=0.36]{swarm.pdf} \fi
-  \caption{Distribution of individual-task speedups across strategies and benchmarks. Gray dots represent tasks. Plots with extremely low variance have no swarm overlay. CSM is \csplitmix.}
+  \caption{Distribution of individual-task speedups across strategies and benchmarks.
+Gray dots represent tasks.
+Plots with extremely low variance have no swarm overlay.
+CSM is \csplitmix; AOC is \camlname.}
   \label{fig:swarm}
@@
 We begin by noting that the principle of choosing fast random number generators
-is completely language-agnostic. In languages like OCaml, where runtime value representations
+is completely language-agnostic. In languages like OCaml, where run-time value representations
 make natively-implemented \rand inefficient, calling out to a C implementation or using standard libraries written in C is a surefire win.

### 2025-10-06 paper/paper.tex
draft 3.3

@@
 
-\subsection{Staging Combinators}
-
-In Section~\ref{section:motiv}, we noted that generator combinators like \texttt{weighted\_union}
-allocate lists in the hot path of the generator.
-Even though these lists are usually small---at most a few dozen elements---each allocation takes us closer to the next garbage collection, which is bad for performance.
-
-This is an ideal opportunity to exercise another feature of staging: compile-time specialization.
-Since we almost always know the particular list of choices at compile time, a staged version of \texttt{weighted\_union}
-can generate \emph{different code} depending on the number of generators in the union.
-If we use weighted union on a compile-time list of generators \texttt{g1}, \texttt{g2}, and \texttt{g3},
-we can emit code that picks between the generators without realizing the list at
-run time.
+\subsection{PBT-Specific Staging}
+While the techniques in the previous section eliminates monadic
+DSL overhead using standard ideas in staging, it cannot address some of the
+more subtle and domain-specific performance issues identified in Section~\ref{section:motiv},
+like hot-path allocations induced by generator combinators such as \texttt{weighted\_union}.
+Removing these allocations requires a observation that transcends simple 
+inlining: though the \emph{elements} of the lists may be dynamic, their
+\emph{spine}---the structure of the list itself---is in practice always statically
+known. This is an instance of \emph{partially-static data} \cite{partially-static}:
+a data structure whose shape is static but whose contents are dynamic.
+
+This observation enables an optimization unique to the PBT generator library setting.
+Because the spine is static, we can unfold the choice-selection
+logic over the list at compile time time, specializing \texttt{weighted\_union} to the exact number
+and weights of choices. The result is a specialized generator that branches directly to the chosen
+alternative, without ever materializing the list of choices at run time.
+
+% In Section~\ref{section:motiv}, we noted that generator combinators like
+% \texttt{weighted\_union} allocate lists in the hot path of the generator. Even
+% though these lists are usually small---at most a few dozen elements---each
+% allocation takes us closer to the next garbage collection, which is bad for
+% performance.
+
+% The simple staging we've used thusfar to remove the overhead of \texttt{return}
+% and \texttt{bind} cannot eliminate this allocation --- we must use structure specific
+% to PBT generator combinators to our advantage.
+
+% In particular, the \emph{spine} of the list of choices that \texttt{weighted\_union} picks from is
+% in practice always statically known, even if the elements themselves may be determined dynamically
+% --- the list is \emph{partially-static data} \cite{partially-static}.
+% Because of this, we can statically unroll the outer loop of the \texttt{weighted\_union} combinator
+% and specialize to the potential choices at compile-time.
+
+% This is an ideal opportunity to exercise another feature of staging: compile-time specialization.
+
+% Since we almost always know the particular list of choices at compile time, a staged version of \texttt{weighted\_union}
+% can generate \emph{different code} depending on the number of generators in the union.
+% If we use weighted union on a compile-time list of generators \texttt{g1}, \texttt{g2}, and \texttt{g3},
+% we can emit code that picks between the generators without realizing the list at
+% run time.
 
@@
 the compile-time \texttt{weighted\_union} combinator generates a
-tree of \texttt{if}s, specialized to the list of weights known at compile time, that picks out the selected generator.
+tree of \texttt{if}s that picks out the selected generator.
 

### 2025-10-06 paper/paper.tex
paper

@@
 the average speedup rises to $3.8\times$,
-with some cases exceeding $13\times$.
+with some cases exceeding $13\times$. Finally, we review related libraries
+in other languages including Racket, F\#, Haskell, Rust, Clojure, and Swift,
+and discuss how the \name{} technique could be applied to them.
 
@@
 generality (Section~\ref{section:faster-generators}).
+\item We present a novel methodology for quantifying 
+      the precise speedup achieved using staging and fast randomness,
+      in which the performance benefit of each intervention can be
+      measured in isolation with absolutely no noise (Section~\ref{section:eval}).
 \item We demonstrate that writing generators using \name{} and fast
@@
 technique could be applied to PBT libraries in other
-languages---Racket, F\#, Haskell, and Rust.
+languages---Racket, F\#, Haskell, Rust, Clojure, and Swift.
 %
@@
 
-\section{Beyond OCaml and Scala}
+\section{Other Languages \& Libraries}
 \label{section:other-langs}

### 2025-10-06 paper/paper.tex
update intro

@@
 In this paper, we characterize, quantify, and address these issues.
-
+% We address the overhead of
+% many common generator abstractions using \textit{multi-stage
+%   programming} (or {\em staging}), a well-studied technique for
+% building fast DSLs “without regret”~\cite{lms}.  We refer to this approach as \name.
+% %
+
+%
 We address the overhead of
@@
   programming} (or {\em staging}), a well-studied technique for
-building fast DSLs “without regret”~\cite{lms}.  We refer to this approach as \name.
-%
+building fast DSLs "without regret"~\cite{lms}. Beyond the standard
+staging transformations, we apply advanced staging techniques~\cite{partially-static,kovacs24, carrette05}
+that let us leverage our domain-specific understanding of PBT generators to achieve the fastest possible code.
+We refer to this approach as \name.
 To demonstrate its generality, we implement staged generator DSLs in

### 2025-10-07 paper/paper.tex
changes

@@
       in which the performance benefit of each intervention can be 
-      measured in isolation with absolutely no noise (Section~\ref{section:eval}).
+      attributed to the interventions themselves, as opposed to 
+      random chance (Section~\ref{section:eval}).
 \item We demonstrate that writing generators using \name{} and fast
@@
 that \emph{all} \texttt{int64} operations (both arithmetic and bitwise) must
-allocate memory cells to contain their output, which has a significant performance benefit.
+allocate memory cells to contain their output, which has a significant performance penalty.
 By building a version that uses much faster ``unboxed'' 64-bit integer
@@
 \label{subsection:basic-design}
-To build a staged version of a generator DSL, we want to rewrite the the generator
+To build a staged version of a generator DSL, we want to rewrite the generator
 combinators to do as much as possible at compile time.
@@
 
-  \textbf{Racket} is the next target that we intend to test on.
-  The entire Racket philosophy is intertwined with
+  \textbf{Racket}'s entire philosophy is intertwined with
   using macros to build small eDSLs like the ones we use to write generators, and so it seems a natural fit.
@@
   the PBT libraries in Rust.
+
+  \textbf{Clojure}'s PBT library test.check\cite{test.check} is a straightforward port
+  of Haskell's QuickCheck. Like Racket, 
+  Clojure has a powerful macro system which could be employed to build staged generators. Further, since
+  Clojure runs on the JVM, we anticipate that the performance benefits of staging
+  would be similar to those observed in ScalaCheck.
+
+  \textbf{Swift}'s PBT library SwiftCheck\cite{swiftcheck} is likewise a direct port of Haskell's QuickCheck,
+  and Swift's macro system\cite{swiftmac} is expressive enough to support staged generators. Swift's inlining is
+  not as robust as GHC's, so we expect that it would also see some benefit from the \name{} technique. However,
+  SwiftCheck is not maintained, so we did not pursue this direction.
+
+This covers most of the major functional and functional-adjacent languages with both PBT libraries and metaprogramming capabilities.
 

### 2025-10-08 paper/paper.tex
typo
meeting nits

@@
   programming} (or {\em staging}), a well-studied technique for
-building fast DSLs "without regret"~\cite{lms}. Beyond the standard
-staging transformations, we apply advanced staging techniques~\cite{partially-static,kovacs24, carrette05}
-that let us leverage our domain-specific understanding of PBT generators to achieve the fastest possible code.
+building fast DSLs "without regret"~\cite{lms}. 
+%
+The simple monadic structure of typical PBT generators lend themselves naturally
+to well-known staging transformations. Beyond these basic techniques, we observe
+that common usage patterns of fundemental PBT-specific combinators---e.g., weighted choices with statically known options---are amenable
+to advanced staging techniques~\cite{partially-static,kovacs24, carrette05}.
+
+% Beyond the standard
+% staging transformations, we apply advanced 
+% that let us leverage our domain-specific understanding of PBT generators to achieve the fastest possible code.
+
 We refer to this approach as \name.
@@
 generality (Section~\ref{section:faster-generators}).
-\item We present a novel methodology for quantifying 
-      the precise speedup achieved using staging and fast randomness,
-      in which the performance benefit of each intervention can be 
-      attributed to the interventions themselves, as opposed to 
-      random chance (Section~\ref{section:eval}).
+% \item We present a novel methodology---enabled by the pointwise semantic equivalence of original and \name{}-optimized generators---for quantifying 
+%       the precise speedup achieved using staging and fast randomness,
+%       in which the performance benefit of each intervention can be 
+%       attributed to the interventions themselves, as opposed to 
+%       random chance (Section~\ref{section:eval}).
 \item We demonstrate that writing generators using \name{} and fast
@@
     let let_insert (cx : 'a code) : 'a code t =
-      {cps = fun k -> k .< let x = .~cx in .~(k .<x>.) >.}
+      {cps = fun k -> .< let x = .~cx in .~(k .<x>.) >.}
   end

### 2025-12-02 paper/paper.tex

@@
 functionality  (Sections~\ref{subsection:basic-design} through \ref{subsection:type-derived}).
+We begin with simple staging tricks, then move on to more advanced techniques that
+leverage our domain-specific understanding of PBT generators.
 Last, we present a controlled experiment demonstrating how a
@@
 
-This covers most of the major functional and functional-adjacent languages with both PBT libraries and metaprogramming capabilities.
+This covers a large swath of functional and functional-adjacent languages with both PBT libraries and metaprogramming capabilities,
+suggesting that the Allegro technique is widely applicable.
 

### 2025-12-02 paper/reviews/rebuttal_oopsla26.md
add things

Reviewer A, Q1:
The main novelty of this work lies in applying staging to PBT. While staging
techniques are well established, their application to the abstractions and
performance bottlenecks of PBT libraries has not been explored, nor is it
straightforward. Our paper is intended to promote staging as a tool in the PBT
developer’s toolkit by showing that it can erase abstraction overhead, leading
to substantial performance improvements.

Prior to this work, neither abstraction overhead nor
sampling costs were viewed as major targets for optimization in PBT. As
evidence, we point to SOTA libraries like `Base_quickcheck`, which is highly
performance sensitive yet retains design choices (monadic combinators, expensive
randomness) that introduce significant overhead.

Additional novelty lies in Section 3.7, where we stage type-derived generators.
Unlike most staged libraries, which require users to understand metaprogramming,
our approach is fully automatic. Since type-derived generators are constructed
at compile time, they can be staged without altering user experience. This is a
rare example of staging "for free."

---

Reviewer B also asks (Q2) if the testing-time speedups are a benefit, given (a) the time required to compile the staged code, and (b) the relative balance between input-generating and property-running durations in practice.

These are both excellent points...

For (a), we do not believe that this is an issue in practice, for two reasons.

1. Generator code changes extremely infrequently compared to code being
tested, and the generator metaprogram needs to be re-run and the output code
recompiled only as the generator itself changes. This is somewhat different from
many other uses of run-time metaprogramming, where the application code itself is
written as a metaprogram. Our intended mode of usage therefore is more like compile time metaprogramming
than run-time metaprogramming, where programmers save the code output to disk to cache it.
2. Generators are quite small, and compile quite quickly.

<!-- ``The authors need to use more workloads and justify this more clearly.'' You're saying... give us *more* workloads? -->

<!-- These benchmarks are relatively challenging to build: one needs many a program, many properties, and many mutants of varying known difficulties -->

<!-- 
The second criticism is with the evaluation. It is unclear to me if, in the software development cycle, the proposed improvement will provide any actual benefit for two main reasons. First, the compiler time is not taken into account. Multi-stage programming adds a runtime code generation time; the extent to which it is taken into account is unclear. It has an overhead that can be amortized if the computation itself is very time-consuming. The authors need to clarify if this time is taken into account. Second, based on Amdahl's Law, does the proposed improvement bring actual benefits in practice? RQ2 shows results for finding bugs in the Etna platform. However, it is still unclear how many practical applications one would see in which a massive proportion of overhead for the testing infrastructure vs. the time each function itself takes. The authors need to use more workloads and justify this more clearly. -->

---

Reviewer C:

1. Can you provide an example of how the recursive generator API in Section 3.6 is used?
Yes, it's used in Figure 3 --- we can include a figure down at section 3.6 with a small example in the camera ready.

2. How were the experimental benchmarks chosen?
We used the subset of the pre-existing Etna suite of benchmarks. Etna was
designed exactly for this purpose: to test the bugfinding capabilities of
competing PBT generators.  There are two benchmarks from the original Etna we
did not use. One tests Red-Black trees, and the other tests a lambda calculus
typed with System FSub, a more sophisticated type system.  We felt that the RBT eval was
redundant with the BTS eval we already used: the generators are very similar.
We chose not to use the FSub eval for two reasons. First, it had not been implemented
in OCaml, and second, we felt this would be redundant with the STLC
evals. The FSub generator is even more complex with more combinators than the STLC one, but the testing code and properties are essentially the same:
we would see an exaggerated version of the speedups in the STLC eval.

3. What do you mean by "semantically identical", and are the generators also syntactically similar?

Yes, the AllegrOCaml and Base_quickcheck generors are as you imagine: essentially syntactically identical, modulo
some extra syntactic cruft for staging. We plan to submit all of our code with the artifact evaluation, but for now, we
include an example from our evals of a generator and its staged counterpart below the fold.

By "semantically identical", we mean that given the same random seeds, the two
generators produce the same generated output.  This property is important
because it guarantees that all of our bug-finding speedups are solely
attributable to the improved performance of staged generators, and not because
the staged generators happen to luckily create bug-squashing outputs. The
sequence of outputs is the same between the staged and unstaged versions, the staged
generators just produce them faster.

We further note that this is not an obvious or simple property to maintain:
staged and unstaged programs can look syntactically similar but behave
non-equivalently, especially in the presence of effects (for example, `let x =
print "hi" in x;x` and `let x = .<print "hi">. in .<$x;$x>. --- the former
prints once, the latter twice). Section 3.4 is essentially about ensuring that
programmers need not worry about this, and can write their generators in a
manner that is syntactically as similar as possible to the non-staged versions.

```
let rec gen ~(lo: int) ~(hi: int)  =
  let%bind sz = size in
  let should_stop = lo >= hi || sz <= 1  in
  if should_stop
    then return E
  else
    weighted_union [
      (1., return E);
      (float_of_int sz , (
        let%bind k = int_inclusive ~lo ~hi in
        let%bind v = Nat.quickcheck_generator_parameterized bst_bespoke_limits in
        let%bind left = with_size ~size_c:(sz / 2) (gen ~lo:lo ~hi:(k - 1))  in
        let%bind right = with_size ~size_c:(sz / 2) (gen ~lo:(k + 1) ~hi:hi) in
        return (T (left, k, v, right))  
      ))
    ]

let staged_quickcheck_generator (lo: int code) (hi: int code) : tree code G.t =
  recursive (.< (.~lo, .~hi ) >.) 
  (fun go lohi -> 
    let%bind sz = size in
    let%bind (lo, hi) = split_pair lohi in
    let%bind should_stop = split_bool .< .~hi <= .~lo || .~sz <= 1 >. in
    if should_stop
      then return .< E >.
    else
      weighted_union [
        (.< 1. >., return .< E >.);
        ((G.C.i2f sz), (
          let%bind k = int_inclusive ~lo ~hi in
          let%bind v = Nat.staged_quickcheck_generator_sr_t (G.C.lift bst_bespoke_limits) in
          let%bind left = with_size ~size_c:(G.C.div2 sz) (recurse go .<(.~lo, .~k - 1) >.) in
          let%bind right = with_size ~size_c:(G.C.div2 sz) (recurse go .<(.~k + 1, .~hi) >.) in
          return (.< T (.~left, .~k, .~v, .~right) >.)))
      ]
  )
```

-------------

Reviewer B:

> Typos on l379, l475, l621, l653, l718, l824

p4: It's odd that Figure 4 appears before Figure 3.

: "returns a code" should be "returns code"

: "a observation"

: "a 'a code" - perhaps you mean "a value of type 'a code"?

> l119: I'm curious whether OCaml 5 might benefit from using it's native
effect handlers in place of a monadic generator DSL.

We are also now curious about this! Very interesting idea, direct-style generators
could potentially be very performant and also have the benefit of being more idiomatic.

> l989: You have conjectured that because GHC is set up to perform aggressive optimisations without staging then it is likely to benefit less from the kind of optimisations you exploit. It would be worth investigating to what extent this is indeed the case. Given that GHC’s QuickCheck is the canonical PBT framework it seems particularly worthwhile to perform further experiments with it. I wonder to what extent it would be possible to disable some of GHC’s aggressive inlining, both in order to assess how much it is really paying off, and to compare its robustness to your staging approach.

This is a fair point. Our conjecture is based on anecdotal evidence; one of the authors — an experienced Haskell developer — manually applied some of Allegro’s optimizations to a few Haskell generators, and they found that they could really only manage to make performance worse, not better. This is nowhere near a proof, but it discouraged us from exploring that path in the short term. We would be happy to mention this anecdotal experience and/or go into more detail about wanting to do this experiment as future work

### 2025-12-02 staged-scala/src/main/scala/stagedgen/StGen.scala
blah

@@ -191,18 +191,5 @@ object StGen {
     )
   }
 
-  def wgImpl (using q : Quotes): Expr[Int => Seed => Int] = {
-    val e = StGen.splat(
-      StGen.frequency(
-        '{2} -> StGen.pure('{999}),
-        '{3} -> StGen.pure('{111})
-      )
-    )
-    e
-  }
-
-  inline def wg = {
-    ${wgImpl}
-  }
 }
 

### 2025-12-02 paper/reviews/rebuttal_oopsla26.md
nit

@@
-Reviewer A, Q1:
+## Reviewer A
+
+### Q1) Can you clarify the novelty of your paper? Does it introduce any optimization from the multi-stage programming not known before?
+
 The main novelty of this work lies in applying staging to PBT. While staging
@@
 rare example of staging "for free."
+
+### Q2) Can you elaborate on the end-to-end effectiveness of your proposed solution?
+
+The end-to-end effectiveness of our solution depends on (a) the complexity 
+of the inputs being generated, and (b) the amount of time generation takes 
+relative to other parts of the testing process—most obviously, running the 
+system on the generated input.
+
+In general, more complex generators contain more calls to bind that are 
+fused away using the Allegro technique; therefore, more complex generators 
+see a larger speedup. Figure 15 showcases this relationship.
+
+The effectiveness of efficient input generation on total testing time is 
+dependent on how long everything that isn’t input generation takes. In systems 
+where running the tests themselves is slow, speeding up input generation won’t 
+have as large an impact; in systems where testing is fast, it will constitute a 
+major speedup. Our end-to-end tests on bug-finding speed, shown in Fig. 17 and 
+Fig. 18, show the time it takes for PBTs to find bugs in mutated programs. In 
+these examples, faster input generation has a measurable end-to-end impact on 
+all of our benchmarks—up to 2.65X using staging alone, and 3.40X in combination 
+with fast randomness.
+
+### Q3) What would it take to implement type-derived generators in Scala? Why did you decide not to implement it?
 

### 2025-12-02 paper/reviews/rebuttal_oopsla26.md
cleanup
C4

@@
-## Reviewer A
+# Reviewer A
 
@@
 
+# Reviewer B
+
 Reviewer B also asks (Q2) if the testing-time speedups are a benefit, given (a) the time required to compile the staged code, and (b) the relative balance between input-generating and property-running durations in practice.
@@
 2. Generators are quite small, and compile quite quickly.
+
+> Typos on l379, l475, l621, l653, l718, l824
+
+p4: It's odd that Figure 4 appears before Figure 3.
+
+: "returns a code" should be "returns code"
+
+: "a observation"
+
+: "a 'a code" - perhaps you mean "a value of type 'a code"?
+
+> l119: I'm curious whether OCaml 5 might benefit from using it's native
+effect handlers in place of a monadic generator DSL.
+
+We are also now curious about this! Very interesting idea, direct-style generators
+could potentially be very performant and also have the benefit of being more idiomatic.
+
+> l989: You have conjectured that because GHC is set up to perform aggressive optimisations without staging then it is likely to benefit less from the kind of optimisations you exploit.
+It would be worth investigating to what extent this is indeed the case.
+Given that GHC’s QuickCheck is the canonical PBT framework it seems particularly worthwhile to perform further experiments with it.
+I wonder to what extent it would be possible to disable some of GHC’s aggressive inlining, both in order to assess how much it is really paying off, and to compare its robustness to your staging approach.
+
+This is a fair point.
+Our conjecture is based on anecdotal evidence; one of the authors — an experienced Haskell developer — manually applied some of Allegro’s optimizations to a few Haskell generators, and they found that they could really only manage to make performance worse, not better.
+This is nowhere near a proof, but it discouraged us from exploring that path in the short term.
+We would be happy to mention this anecdotal experience and/or go into more detail about wanting to do this experiment as future work
 
@@
 
----
+# Reviewer C
 
-Reviewer C:
-
-1. Can you provide an example of how the recursive generator API in Section 3.6 is used?
+> 1. Can you provide an example of how the recursive generator API in Section 3.6 is used?
 Yes, it's used in Figure 3 --- we can include a figure down at section 3.6 with a small example in the camera ready.
 
-2. How were the experimental benchmarks chosen?
+> 2. How were the experimental benchmarks chosen?
+
 We used the subset of the pre-existing Etna suite of benchmarks. Etna was
@@
 
-3. What do you mean by "semantically identical", and are the generators also syntactically similar?
+> 3. What do you mean by "semantically identical", and are the generators also syntactically similar?
 
@@
 
--------------
+> 4. Why was BST (Repeated Insert) not used for the ScAllegro experiments in Figure 16?
 
-Reviewer B:
-
-> Typos on l379, l475, l621, l653, l718, l824
-
-p4: It's odd that Figure 4 appears before Figure 3.
-
-: "returns a code" should be "returns code"
-
-: "a observation"
-
-: "a 'a code" - perhaps you mean "a value of type 'a code"?
-
-> l119: I'm curious whether OCaml 5 might benefit from using it's native
-effect handlers in place of a monadic generator DSL.
-
-We are also now curious about this! Very interesting idea, direct-style generators
-could potentially be very performant and also have the benefit of being more idiomatic.
-
-> l989: You have conjectured that because GHC is set up to perform aggressive optimisations without staging then it is likely to benefit less from the kind of optimisations you exploit.
-It would be worth investigating to what extent this is indeed the case.
-Given that GHC’s QuickCheck is the canonical PBT framework it seems particularly worthwhile to perform further experiments with it.
-I wonder to what extent it would be possible to disable some of GHC’s aggressive inlining, both in order to assess how much it is really paying off, and to compare its robustness to your staging approach.
-
-This is a fair point.
-Our conjecture is based on anecdotal evidence; one of the authors — an experienced Haskell developer — manually applied some of Allegro’s optimizations to a few Haskell generators, and they found that they could really only manage to make performance worse, not better.
-This is nowhere near a proof, but it discouraged us from exploring that path in the short term.
-We would be happy to mention this anecdotal experience and/or go into more detail about wanting to do this experiment as future work
+We picked a representative from each category of generators: trees, lists, and STLC terms, representing low, medium, and high numbers of binds, respectively.
+We anticipate BST (Repeated Insert) would see a slightly smaller speedup due to its lower number of binds, and we would be happy to include this information in a revision.

### 2025-12-02 paper/reviews/rebuttal_oopsla26.md
harry

@@
 
-### Q1) Can you clarify the novelty of your paper? Does it introduce any optimization from the multi-stage programming not known before?
+> Q1) Can you clarify the novelty of your paper? Does it introduce any optimization from the multi-stage programming not known before?
 
@@
 
-### Q2) Can you elaborate on the end-to-end effectiveness of your proposed solution?
+> Q2) Can you elaborate on the end-to-end effectiveness of your proposed solution?
 
@@
 
-### Q3) What would it take to implement type-derived generators in Scala? Why did you decide not to implement it?
+> Q3) What would it take to implement type-derived generators in Scala? Why did you decide not to implement it?
 
@@
 -->
+
+> One somewhat open-ended question I am wondering about is to what
+extent we can bridge the performance gap between bespoke
+highly-optimised fuzzing approaches and the much more generic PBT
+libraries exemplified by QuickCheck. Can we hope to build fuzzers on
+top of generic PBT libraries which perform as well as the hand-coded
+fuzzers used in practice? Does your work shed any light on how to
+tackle this question?
+
+This is a really interesting question, and we think that our work speaks directly to part of it.
+Specifically, hand-coded fuzzers that construct inputs satisfying specific structural and semantic constraints are performing precisely the same task that PBT generators are designed for.
+Hand-coded fuzzers might be faster than PBT generators if users choose to optimize them aggressively, but ideally those optimizations should not need to be done manually.
+Our hope is that Allegro (and other advances from the PBT literature) can make PBT generators fast enough that developers could use them as highly-performant fuzzers, even though they are expressed in a higher-level language that is easier to write in and reason about.
+The question gets a bit less clear when considering fuzzers that use mutation and other techniques to obtain interesting inputs, rather than a hand-coded program, although ideas in papers like Coverage Guided Property-Based Testing and Parsing Randomness may provide paths forward in those domains as well.
 

### 2025-12-02 paper/reviews/rebuttal_oopsla26.md
nit
reviews

@@
-# Reviewer A
+Reviewer A:
 
-> Q1) Can you clarify the novelty of your paper? Does it introduce any optimization from the multi-stage programming not known before?
+> Q1) Can you clarify the novelty of your paper? Does it introduce any 
+optimization from the multi-stage programming not known before?
 
@@
 
-Prior to this work, neither abstraction overhead nor
-sampling costs were viewed as major targets for optimization in PBT. As
-evidence, we point to SOTA libraries like `Base_quickcheck`, which is highly
-performance sensitive yet retains design choices (monadic combinators, expensive
-randomness) that introduce significant overhead.
+Prior to this work, neither DSL abstraction overhead nor sampling costs were
+viewed as targets for optimization in PBT. As evidence, we point to SOTA
+libraries like `Base_quickcheck`, which is has had a lot of optimization effort
+put into it --- it uses imperative features like mutable arrays and a stateful
+RNG for performance reasons --- yet retains design choices (monadic combinators,
+expensive randomness) that introduce significant overhead.
 
@@
 
-> Q2) Can you elaborate on the end-to-end effectiveness of your proposed solution?
+>  Q2) Can you elaborate on the end-to-end effectiveness of your proposed 
+solution?
 
-The end-to-end effectiveness of our solution depends on (a) the complexity 
+[JWC: reframe this to be more about the eval.]
+
+The effectiveness depends on (a) the complexity 
 of the inputs being generated, and (b) the amount of time generation takes 
@@
 
-> Q3) What would it take to implement type-derived generators in Scala? Why did you decide not to implement it?
+As for the compilation time of the emitted generator code, we do not believe
+that this is an issue, for two reasons. The first is that our metaprogrammed
+generators compile extremely quickly in AllegrOCaml: 33ms for the simple Bool List
+generator, up to 54ms for the STLC generator (a full table of the AllegrOCaml generators and their compilation times can be found after the fold).  Amortized across testing multiple
+properties (which frequently take on the order of seconds to run), this is
+negligible. Second, generator code changes extremely infrequently compared to
+code being tested, and the generator metaprogram needs to be re-run and the
+output code recompiled only when the generator itself changes. This is somewhat
+different from many other uses of run-time metaprogramming, where the
+application code itself is written as a metaprogram. Our intended mode of usage
+therefore is more like compile time metaprogramming than run-time
+metaprogramming, where programmers save the code output to disk to cache it.
 
----
+> Q3) What would it take to implement type-derived generators in Scala? 
+Why did you decide not to implement it?
 
-# Reviewer B
+...
 
-Reviewer B also asks (Q2) if the testing-time speedups are a benefit, given (a) the time required to compile the staged code, and (b) the relative balance between input-generating and property-running durations in practice.
+-------------
 
-These are both excellent points...
-
-For (a), we do not believe that this is an issue in practice, for two reasons.
-
-1. Generator code changes extremely infrequently compared to code being
-tested, and the generator metaprogram needs to be re-run and the output code
-recompiled only as the generator itself changes. This is somewhat different from
-many other uses of run-time metaprogramming, where the application code itself is
-written as a metaprogram. Our intended mode of usage therefore is more like compile time metaprogramming
-than run-time metaprogramming, where programmers save the code output to disk to cache it.
-2. Generators are quite small, and compile quite quickly.
+Reviewer B:
 
@@
 It would be worth investigating to what extent this is indeed the case.
-Given that GHC’s QuickCheck is the canonical PBT framework it seems particularly worthwhile to perform further experiments with it.
-I wonder to what extent it would be possible to disable some of GHC’s aggressive inlining, both in order to assess how much it is really paying off, and to compare its robustness to your staging approach.
 
@@
 This is nowhere near a proof, but it discouraged us from exploring that path in the short term.
-We would be happy to mention this anecdotal experience and/or go into more detail about wanting to do this experiment as future work
+We would be happy to mention this anecdotal experience and/or go into more detail about wanting to do this experiment as future work.
 
-<!-- ``The authors need to use more workloads and justify this more clearly.'' You're saying... give us *more* workloads? -->
+> Can we can bridge the performance gap between bespoke
+highly-optimised fuzzing approaches and the much more generic PBT
+libraries exemplified by QuickCheck?
 
-<!-- These benchmarks are relatively challenging to build: one needs many a program, many properties, and many mutants of varying known difficulties -->
+This is a really interesting question, and we think that our work speaks 
+directly to part of it. Specifically, hand-coded fuzzers that construct 
+inputs satisfying specific structural and semantic constraints are performing 
+precisely the same task that PBT generators are designed for. Hand-coded fuzzers 
+might be faster than PBT generators if users choose to optimize them aggressively, 
+but ideally those optimizations should not need to be done manually. Our hope is 
+that Allegro (and other advances from the PBT literature) can make PBT generators 
+fast enough that developers could use them as highly-performant fuzzers, even 
+though they are expressed in a higher-level language that is easier to write 
+in and reason about. The question gets a bit less clear when considering fuzzers 
+that use mutation and other techniques to obtain interesting inputs, rather than 
+a hand-coded program, although ideas in papers like Coverage Guided Property-Based 
+Testing and Parsing Randomness may provide paths forward in those domains as well.
 
-<!-- 
-The second criticism is with the evaluation.
-It is unclear to me if, in the software development cycle, the proposed improvement will provide any actual benefit for two main reasons.
-First, the compiler time is not taken into account.
-Multi-stage programming adds a runtime code generation time; the extent to which it is taken into account is unclear.
-It has an overhead that can be amortized if the computation itself is very time-consuming.
-The authors need to clarify if this time is taken into account.
-Second, based on Amdahl's Law, does the proposed improvement bring actual benefits in practice?
-RQ2 shows results for finding bugs in the Etna platform.
-However, it is still unclear how many practical applications one would see in which a massive proportion of overhead for the testing infrastructure vs.
-the time each function itself takes.
-The authors need to use more workloads and justify this more clearly.
--->
+----
 
-> One somewhat open-ended question I am wondering about is to what
-extent we can bridge the performance gap between bespoke
-highly-optimised fuzzing approaches and the much more generic PBT
-libraries exemplified by QuickCheck. Can we hope to build fuzzers on
-top of generic PBT libraries which perform as well as the hand-coded
-fuzzers used in practice? Does your work shed any light on how to
-tackle this question?
-
-This is a really interesting question, and we think that our work speaks directly to part of it.
-Specifically, hand-coded fuzzers that construct inputs satisfying specific structural and semantic constraints are performing precisely the same task that PBT generators are designed for.
-Hand-coded fuzzers might be faster than PBT generators if users choose to optimize them aggressively, but ideally those optimizations should not need to be done manually.
-Our hope is that Allegro (and other advances from the PBT literature) can make PBT generators fast enough that developers could use them as highly-performant fuzzers, even though they are expressed in a higher-level language that is easier to write in and reason about.
-The question gets a bit less clear when considering fuzzers that use mutation and other techniques to obtain interesting inputs, rather than a hand-coded program, although ideas in papers like Coverage Guided Property-Based Testing and Parsing Randomness may provide paths forward in those domains as well.
-
-# Reviewer C
+Reviewer C:
 
 > 1. Can you provide an example of how the recursive generator API in Section 3.6 is used?
-Yes, it's used in Figure 3 --- we can include a figure down at section 3.6 with a small example in the camera ready.
+Yes, it's used in Figure 3 --- we can include a figure down at Section 3.6 with a small example in the camera-ready version.
 
 > 2. How were the experimental benchmarks chosen?
-
 We used the subset of the pre-existing Etna suite of benchmarks. Etna was
@@
 
-We picked a representative from each category of generators: trees, lists, and STLC terms, representing low, medium, and high numbers of binds, respectively.
-We anticipate BST (Repeated Insert) would see a slightly smaller speedup due to its lower number of binds, and we would be happy to include this information in a revision.
+We picked a representative from each category of generators: trees, lists, and STLC terms, 
+representing low, medium, and high numbers of binds, respectively. We anticipate BST 
+(Repeated Insert) would see a slightly smaller speedup due to its lower number of binds, 
+and we would be happy to include this information in a revision. Our goal in this evaluation subsection was not to point-by-point reproduce the AllegrOCaml evaluation in ScAllegro, but
+simply to demonstrate that the performance improvments of the Allegro technique are portable
+across languges.
+
+----
+
+Generator Compilation Times
+
+Bool List: 33.183ms
+BST Repated Insert: 45.453ms
+BST Type: 46.051ms
+BST Single Pass: 50.147ms
+STLC Type: 44.663ms
+STLC Bespoke: 53.699ms

### 2025-12-02 paper/reviews/rebuttal_oopsla26.md

@@
-Reviewer A:
+# Reviewer A
 
-> Q1) Can you clarify the novelty of your paper? Does it introduce any 
-optimization from the multi-stage programming not known before?
+> Q1) Can you clarify the novelty of your paper? Does it introduce any optimization from the multi-stage programming not known before?
 
@@
 
--------------
-
-Reviewer B:
+# Reviewer B:
 
@@
 It would be worth investigating to what extent this is indeed the case.
+Given that GHC’s QuickCheck is the canonical PBT framework it seems particularly worthwhile to perform further experiments with it.
+I wonder to what extent it would be possible to disable some of GHC’s aggressive inlining, both in order to assess how much it is really paying off, and to compare its robustness to your staging approach.
 
@@
 This is nowhere near a proof, but it discouraged us from exploring that path in the short term.
-We would be happy to mention this anecdotal experience and/or go into more detail about wanting to do this experiment as future work.
+We would be happy to mention this anecdotal experience and/or go into more detail about wanting to do this experiment as future work
 
@@
 
-Reviewer C:
+This is a really interesting question, and we think that our work speaks directly to part of it.
+Specifically, hand-coded fuzzers that construct inputs satisfying specific structural and semantic constraints are performing precisely the same task that PBT generators are designed for.
+Hand-coded fuzzers might be faster than PBT generators if users choose to optimize them aggressively, but ideally those optimizations should not need to be done manually.
+Our hope is that Allegro (and other advances from the PBT literature) can make PBT generators fast enough that developers could use them as highly-performant fuzzers, even though they are expressed in a higher-level language that is easier to write in and reason about.
+The question gets a bit less clear when considering fuzzers that use mutation and other techniques to obtain interesting inputs, rather than a hand-coded program, although ideas in papers like Coverage Guided Property-Based Testing and Parsing Randomness may provide paths forward in those domains as well.
+
+# Reviewer C
+
+> 1. Can you provide an example of how the recursive generator API in Section 3.6 is used?
 

### 2025-12-02 paper/reviews/rebuttal_oopsla26.md
blah
oops

@@
 > Q1) Can you clarify the novelty of your paper? Does it introduce any optimization from the multi-stage programming not known before?
+
+[JWC: ... and noticing that these sources of inefficiency exist in PBT]
 
@@
 [JWC: reframe this to be more about the eval.]
-
+<!-- 
 The effectiveness depends on (a) the complexity 
@@
 all of our benchmarks—up to 2.65X using staging alone, and 3.40X in combination 
-with fast randomness.
+with fast randomness. -->
 

### 2025-12-02 paper/reviews/rebuttal_oopsla26.md
??
blah

@@
 rare example of staging "for free."
+
+[JWC can we also say that this is an instance of two-level metaprogramming? It's a metaprogram that generates metaprograms!]
 
@@
 
-...
+None of the authors are Scala experts.
+We are much more familiar with OCaml, so we chose to focus our implementation efforts there.
+That said, we see no fundamental barrier to implementing type-derived staged generators in Scala.
+Indeed, the scalacheck-derived project already uses Scala's runtime reflection to type-derive standard (unstaged) ScalaCheck generators.
+Adapting this approach to emit ScAllegro would require engineering effort, but we anticipate no fundamental obstacles beyond that.
 
@@
 Testing and Parsing Randomness may provide paths forward in those domains as well.
-
-----
-
-This is a really interesting question, and we think that our work speaks directly to part of it.
-Specifically, hand-coded fuzzers that construct inputs satisfying specific structural and semantic constraints are performing precisely the same task that PBT generators are designed for.
-Hand-coded fuzzers might be faster than PBT generators if users choose to optimize them aggressively, but ideally those optimizations should not need to be done manually.
-Our hope is that Allegro (and other advances from the PBT literature) can make PBT generators fast enough that developers could use them as highly-performant fuzzers, even though they are expressed in a higher-level language that is easier to write in and reason about.
-The question gets a bit less clear when considering fuzzers that use mutation and other techniques to obtain interesting inputs, rather than a hand-coded program, although ideas in papers like Coverage Guided Property-Based Testing and Parsing Randomness may provide paths forward in those domains as well.
 

### 2025-12-02 paper/reviews/rebuttal_oopsla26.md
responses to b

@@
 
-None of the authors are Scala experts.
-We are much more familiar with OCaml, so we chose to focus our implementation efforts there.
-That said, we see no fundamental barrier to implementing type-derived staged generators in Scala.
-Indeed, the scalacheck-derived project already uses Scala's runtime reflection to type-derive standard (unstaged) ScalaCheck generators.
-Adapting this approach to emit ScAllegro would require engineering effort, but we anticipate no fundamental obstacles beyond that.
+None of the authors are Scala experts: we are much more familiar with OCaml, so
+we chose to focus our implementation efforts there. That said, we see no
+fundamental barrier to implementing type-derived staged generators in Scala.
+Indeed, the scalacheck-derived project already uses Scala's reflection features
+to type-derive standard (unstaged) ScalaCheck generators. Adapting this approach
+to emit ScAllegro would require engineering effort, but we anticipate no
+fundamental obstacles beyond that.
 
@@
 
-p4: It's odd that Figure 4 appears before Figure 3.
-
-: "returns a code" should be "returns code"
-
-: "a observation"
-
-: "a 'a code" - perhaps you mean "a value of type 'a code"?
+Thank you! We'll fix these for the camera ready.
 
@@
 could potentially be very performant and also have the benefit of being more idiomatic.
+
+> l639: Testing is all well and good, but have you not also considered
+trying to construct a more rigorous proof that AllegrOCaml is correct
+by program transformation / calculation?
+
+We had not considered this, but it would be interesting to investigate.
+
+> l751: Given that MetaOCaml is compatible with OCaml 5, why are you
+still using OCaml 4?
+
+We were not aware that MetaOCaml was now compatible with OCaml 5, this is good
+to know!  When we began the project earlier this year, the MetaOCaml homepage
+instructed users to install a version compatible with OCaml 4.14.1, which is
+what we did (it has since been updated to recommend 5.3.0).
+
+> l824: The text reads a bit oddly to me at this point, as you are
+talking about an experiment you performed in the past and yet you
+insist on using the present tense.
+
+Thanks for the feedback, we'll workshop this prose to make it less awkward.
 
@@
 
-> 1. Can you provide an example of how the recursive generator API in Section 3.6 is used?
 Yes, it's used in Figure 3 --- we can include a figure down at Section 3.6 with a small example in the camera-ready version.
@@
 > 2. How were the experimental benchmarks chosen?
+
 We used the subset of the pre-existing Etna suite of benchmarks. Etna was

### 2025-12-02 paper/reviews/rebuttal_oopsla26.md
attempt

@@
 with fast randomness. -->
+
+Regarding end-to-end effectiveness: Etna is the standard platform for 
+evaluating the bug-finding speed of PBT generators. It was designed 
+as a testbench for comparing generator strategies against each other, and, 
+to our knowledge, no more comprehensive benchmark suite exists for this 
+purpose. Our results show consistent speedups across Etna's tasks, with 
+generation time translating directly to faster bug-finding.
+
+We acknowledge that characterizing the proportion of testing time spent 
+on generation across real-world PBT workloads is an interesting question 
+that merits further study, and doing so would be an important research 
+contribution in itself; we view this as future work outside 
+the scope of this paper.
 

### 2025-12-02 paper/reviews/rebuttal_oopsla26.md
more words

--- 2025-12-02 paper/reviews/rebuttal_oopsla26.md
@@
+Thank you for your comments and questions! We start by discussing some common high-level concerns, then address all comments in order.
+
 # Reviewer A
@@
 
-[JWC: ... and noticing that these sources of inefficiency exist in PBT]
+The main novelties are (1) identifying previously unrecognized sources of inefficiency in PBT, and (2) applying staging to address them.
 
-The main novelty of this work lies in applying staging to PBT. While staging
-techniques are well established, their application to the abstractions and
-performance bottlenecks of PBT libraries has not been explored, nor is it
-straightforward. Our paper is intended to promote staging as a tool in the PBT
-developer’s toolkit by showing that it can erase abstraction overhead, leading
-to substantial performance improvements.
+Prior to this work, neither DSL abstraction overhead nor sampling costs were viewed as optimization targets in PBT.
+As evidence, SOTA libraries like `Base_quickcheck` use imperative features (mutable arrays, stateful RNG) to try to improve performance, yet retain design choices (monadic combinators, expensive randomness) that introduce significant overhead.
 
-Prior to this work, neither DSL abstraction overhead nor sampling costs were
-viewed as targets for optimization in PBT. As evidence, we point to SOTA
-libraries like `Base_quickcheck`, which is has had a lot of optimization effort
-put into it --- it uses imperative features like mutable arrays and a stateful
-RNG for performance reasons --- yet retains design choices (monadic combinators,
-expensive randomness) that introduce significant overhead.
+While our staging techniques are well established, their application to PBT abstractions and bottlenecks has not been explored.
+This paper promotes staging as a tool for PBT developers by showing it can erase abstraction overhead, yielding substantial performance improvements.
 
-Additional novelty lies in Section 3.7, where we stage type-derived generators.
+<!-- Additional novelty lies in Section 3.7, where we stage type-derived generators.
 Unlike most staged libraries, which require users to understand metaprogramming,
@@
 at compile time, they can be staged without altering user experience. This is a
-rare example of staging "for free."
-
-[JWC can we also say that this is an instance of two-level metaprogramming? It's a metaprogram that generates metaprograms!]
+rare example of staging "for free." -->
+<!-- 
+[JWC can we also say that this is an instance of two-level metaprogramming? It's a metaprogram that generates metaprograms!] -->
 
@@
 
-As for the compilation time of the emitted generator code, we do not believe
-that this is an issue, for two reasons. The first is that our metaprogrammed
-generators compile extremely quickly in AllegrOCaml: 33ms for the simple Bool List
-generator, up to 54ms for the STLC generator (a full table of the AllegrOCaml generators and their compilation times can be found after the fold).  Amortized across testing multiple
-properties (which frequently take on the order of seconds to run), this is
-negligible. Second, generator code changes extremely infrequently compared to
-code being tested, and the generator metaprogram needs to be re-run and the
-output code recompiled only when the generator itself changes. This is somewhat
-different from many other uses of run-time metaprogramming, where the
-application code itself is written as a metaprogram. Our intended mode of usage
-therefore is more like compile time metaprogramming than run-time
-metaprogramming, where programmers save the code output to disk to cache it.
+As for compilation time of emitted generator code, we do not believe this is an issue for two reasons.
+First, our metaprogrammed generators compile extremely quickly: 33ms for Bool List up to 54ms for STLC (full table after the fold).
+Amortized across testing multiple properties (which usually take seconds each), this is negligible.
+Second, generators change infrequently compared to code being tested, and the generator only needs to be recompiled when it changes.
+Indeed,
+we anticipate that a production-ready instantiation of Allegro could let programmers
+cache generator code to disk, to avoid compilation at testing-time altogether.
+<!-- Unlike typical run-time metaprogramming which must be frequently recompiled because application code itself is written as a metaprogram, our usage is more like compile-time metaprogramming, where programmers can cache generated code to disk.
+ -->
 
@@
 
-We picked a representative from each category of generators: trees, lists, and STLC terms, 
-representing low, medium, and high numbers of binds, respectively. We anticipate BST 
-(Repeated Insert) would see a slightly smaller speedup due to its lower number of binds, 
-and we would be happy to include this information in a revision. Our goal in this evaluation subsection was not to point-by-point reproduce the AllegrOCaml evaluation in ScAllegro, but
-simply to demonstrate that the performance improvments of the Allegro technique are portable
-across languges.
+We picked a representative from each category of generators: trees, lists, and
+STLC terms, representing low, medium, and high numbers of binds, respectively.
+We anticipate BST (Repeated Insert) would see a slightly smaller than the other
+generators speedup due to its lower number of binds, and we would be happy to
+include this information in a revision. Our goal in this evaluation subsection
+was not to point-by-point reproduce the AllegrOCaml evaluation in ScAllegro, but
+simply to demonstrate that the performance improvments of the Allegro technique
+are portable across languges.
+
+> I found the repeated shifting of focus between OCaml and Scala to be distracting.
+An alternative approach might have been to describe the entirety of the approach for OCaml, and then summarize the Scala-specific differences.
+
+Thanks for this feedback. We'll try to further isolate the discussion of Scala to just the intro and the bits of the evaluation where we evaluate ScAllegro.
+
+> On a related note, the paper might be easier to read if there was a single table / figure that summarized the types of various values and objects used throughout the paper.
+I found myself making this list as I read the paper.
+
+This is a great idea, we'll do this with the extra space in the camera-ready version.
 

### 2025-12-02 paper/reviews/rebuttal_oopsla26.md

@@
 
-As for compilation time of emitted generator code, we do not believe this is an issue for two reasons.
-First, our metaprogrammed generators compile extremely quickly: 33ms for Bool List up to 54ms for STLC (full table after the fold).
-Amortized across testing multiple properties (which usually take seconds each), this is negligible.
-Second, generators change infrequently compared to code being tested, and the generator only needs to be recompiled when it changes.
-Indeed,
-we anticipate that a production-ready instantiation of Allegro could let programmers
-cache generator code to disk, to avoid compilation at testing-time altogether.
 <!-- Unlike typical run-time metaprogramming which must be frequently recompiled because application code itself is written as a metaprogram, our usage is more like compile-time metaprogramming, where programmers can cache generated code to disk.
  -->
+Regarding end-to-end effectiveness: Etna is the standard platform for 
+evaluating the bug-finding speed of PBT generators. It was designed 
+as a testbench for comparing generator strategies against each other, and, 
+to our knowledge, no more comprehensive benchmark suite exists for this 
+purpose. Our results show consistent speedups across Etna's tasks, with 
+generation time translating directly to faster bug-finding.
+
+We acknowledge that characterizing the proportion of testing time spent 
+on generation across real-world PBT workloads is an interesting question 
+that merits further study, and doing so would be an important research 
+contribution in itself; we view this as future work outside 
+the scope of this paper.
+
+As for compilation time of emitted generator code, we do not believe this is an
+issue for two reasons. First, our metaprogrammed generators compile extremely
+quickly: 33ms for Bool List up to 54ms for STLC (full table after the fold).
+Amortized across testing multiple properties (which usually take seconds each),
+this is negligible. Second, generators change infrequently compared to code
+being tested, and the generator only needs to be recompiled when it changes.
+Indeed, we anticipate that a production-ready instantiation of Allegro could let
+programmers cache generator code to disk, to avoid compilation at testing-time
+altogether.
 

### 2025-12-02 paper/reviews/rebuttal_oopsla26.md
bloop

@@
 Prior to this work, neither DSL abstraction overhead nor sampling costs were viewed as optimization targets in PBT.
-As evidence, SOTA libraries like `Base_quickcheck` use imperative features (mutable arrays, stateful RNG) to try to improve performance, yet retain design choices (monadic combinators, expensive randomness) that introduce significant overhead.
+As evidence, SOTA libraries like `Base_quickcheck` use imperative features (mutable arrays, stateful instead of functional RNG) to try to improve performance, yet retain design choices (monadic combinators, relatively expensive randomness) that introduce significant overhead.
 
-While our staging techniques are well established, their application to PBT abstractions and bottlenecks has not been explored.
+While our staging techniques are well established, their application to PBT has not been explored.
 This paper promotes staging as a tool for PBT developers by showing it can erase abstraction overhead, yielding substantial performance improvements.
@@
 
-[JWC: reframe this to be more about the eval.]
-<!-- 
-The effectiveness depends on (a) the complexity 
-of the inputs being generated, and (b) the amount of time generation takes 
-relative to other parts of the testing process—most obviously, running the 
-system on the generated input.
-
-In general, more complex generators contain more calls to bind that are 
-fused away using the Allegro technique; therefore, more complex generators 
-see a larger speedup. Figure 15 showcases this relationship.
-
-The effectiveness of efficient input generation on total testing time is 
-dependent on how long everything that isn’t input generation takes. In systems 
-where running the tests themselves is slow, speeding up input generation won’t 
-have as large an impact; in systems where testing is fast, it will constitute a 
-major speedup. Our end-to-end tests on bug-finding speed, shown in Fig. 17 and 
-Fig. 18, show the time it takes for PBTs to find bugs in mutated programs. In 
-these examples, faster input generation has a measurable end-to-end impact on 
-all of our benchmarks—up to 2.65X using staging alone, and 3.40X in combination 
-with fast randomness. -->
-
-<!-- Unlike typical run-time metaprogramming which must be frequently recompiled because application code itself is written as a metaprogram, our usage is more like compile-time metaprogramming, where programmers can cache generated code to disk.
- -->
-Regarding end-to-end effectiveness: Etna is the standard platform for 
+Our evaluation uses Etna, the standard platform for 
 evaluating the bug-finding speed of PBT generators. It was designed 
@@
 
-None of the authors are Scala experts: we are much more familiar with OCaml, so
-we chose to focus our implementation efforts there. That said, we see no
-fundamental barrier to implementing type-derived staged generators in Scala.
-Indeed, the scalacheck-derived project already uses Scala's reflection features
-to type-derive standard (unstaged) ScalaCheck generators. Adapting this approach
-to emit ScAllegro would require engineering effort, but we anticipate no
-fundamental obstacles beyond that.
+We see no fundamental barrier to implementing type-derived staged generators in
+Scala.  Indeed, the scalacheck-derived project already uses Scala's reflection
+features to type-derive standard (unstaged) ScalaCheck generators. Adapting this
+approach to emit ScAllegro would require engineering effort, but we anticipate
+no obstacles beyond that.  We chose to implement our type-derived generators in
+OCaml simply because none of the authors are Scala experts: we are much more
+familiar with OCaml, so we chose to focus our implementation efforts there.
 
@@
 
-> Typos on l379, l475, l621, l653, l718, l824
+> Typos and formatting issues on p4, l379, l475, l621, l653, l718, l824, l996
 

### 2025-12-02 paper/reviews/rebuttal_oopsla26.md
Comments

--- 2025-12-02 paper/reviews/rebuttal_oopsla26.md
@@
 As evidence, SOTA libraries like `Base_quickcheck` use imperative features (mutable arrays, stateful RNG) to try to improve performance, yet retain design choices (monadic combinators, expensive randomness) that introduce significant overhead.
+[BCP: Yes!]
 
@@
 This paper promotes staging as a tool for PBT developers by showing it can erase abstraction overhead, yielding substantial performance improvements.
+[BCP: ... but was it hard?]
 
@@
 rare example of staging "for free." -->
+[BCP: Why not include this? Seems strong.]
 <!-- 
@@
 
-The effectiveness of efficient input generation on total testing time is 
+As the reviewer notes,
+the effectiveness of efficient input generation on total testing time is 
 dependent on how long everything that isn’t input generation takes. In systems 
@@
 the scope of this paper.
+[BCP: Maybe explain why it's not easy?]
 

### 2025-12-02 paper/reviews/rebuttal_oopsla26.md

--- 2025-12-02 paper/reviews/rebuttal_oopsla26.md
@@
 As evidence, SOTA libraries like `Base_quickcheck` use imperative features (mutable arrays, stateful instead of functional RNG) to try to improve performance, yet retain design choices (monadic combinators, relatively expensive randomness) that introduce significant overhead.
-
+[BCP: Yes!]
 While our staging techniques are well established, their application to PBT has not been explored.
 This paper promotes staging as a tool for PBT developers by showing it can erase abstraction overhead, yielding substantial performance improvements.
+[BCP: ... but was it hard?]
 

### 2025-12-02 paper/reviews/rebuttal_oopsla26.md
Comments

@@
 rare example of staging "for free." -->
-[BCP: Why not include this? Seems strong.]
+[BCP: include this! Seems strong.]
 <!-- 
@@
 
+[BCP: IMO we can elide low-level comments and responses.]
+
 > l119: I'm curious whether OCaml 5 might benefit from using it's native
@@
 
-We are also now curious about this! Very interesting idea, direct-style generators
-could potentially be very performant and also have the benefit of being more idiomatic.
+Now we are also curious about this! Very interesting idea: direct-style generators
+could potentially be very performant (and more idiomatic).
 
@@
 
-We were not aware that MetaOCaml was now compatible with OCaml 5, this is good
+We were not aware that MetaOCaml was now compatible with OCaml 5 -- this is good
 to know!  When we began the project earlier this year, the MetaOCaml homepage

### 2025-12-02 paper/reviews/rebuttal_oopsla26.md
wibble

--- 2025-12-02 paper/reviews/rebuttal_oopsla26.md
@@
 As evidence, SOTA libraries like `Base_quickcheck` use imperative features (mutable arrays, stateful instead of functional RNG) to try to improve performance, yet retain design choices (monadic combinators, relatively expensive randomness) that introduce significant overhead.
-[BCP: Yes!]
+
 While our staging techniques are well established, their application to PBT has not been explored.
 This paper promotes staging as a tool for PBT developers by showing it can erase abstraction overhead, yielding substantial performance improvements.
-[BCP: ... but was it hard?]
 
@@
 rare example of staging "for free." -->
-[BCP: Why not include this? Seems strong.]
 <!-- 
@@
 the scope of this paper.
-[BCP: Maybe explain why it's not easy?]
 
@@
 This is a fair point.
-Our conjecture is based on anecdotal evidence; one of the authors — an experienced Haskell developer — manually applied some of Allegro’s optimizations to a few Haskell generators, and they found that they could really only manage to make performance worse, not better.
+Our conjecture is based on anecdotal evidence; one of the authors — an experienced Haskell developer — manually applied some of Allegro’s optimizations to a few Haskell generators, and they found that they could really only manage to make performance identical or worse.
 This is nowhere near a proof, but it discouraged us from exploring that path in the short term.
@@
 
-We used the subset of the pre-existing Etna suite of benchmarks. Etna was
-designed exactly for this purpose: to test the bugfinding capabilities of
-competing PBT generators.  There are two benchmarks from the original Etna we
-did not use. One tests Red-Black trees, and the other tests a lambda calculus
-typed with System FSub, a more sophisticated type system.  We felt that the RBT eval was
-redundant with the BTS eval we already used: the generators are very similar.
-We chose not to use the FSub eval for two reasons. First, it had not been implemented
-in OCaml, and second, we felt this would be redundant with the STLC
-evals. The FSub generator is even more complex with more combinators than the STLC one, but the testing code and properties are essentially the same:
-we would see an exaggerated version of the speedups in the STLC eval.
+We used benchmarks from Etna, which was specifically designed to test the
+bugfinding capabilities of competing PBT generators.  There are two benchmarks
+from the original Etna we did not use. One tests Red-Black trees, and the other
+tests a lambda calculus typed with System FSub, a more sophisticated type
+system.  We felt that the RBT eval was redundant with the BTS eval we already
+used: the generators are very similar.  We chose not to use the FSub eval for
+two reasons. First, it had not been implemented in OCaml, and second, we felt
+this would be redundant with the STLC evals. The FSub generator is even more
+complex with more combinators than the STLC one, but the testing code and
+properties are essentially the same: we would see an exaggerated version of the
+speedups in the STLC eval.
 

### 2025-12-02 paper/reviews/rebuttal_oopsla26.md
Comments

--- 2025-12-02 paper/reviews/rebuttal_oopsla26.md
@@
 what we did (it has since been updated to recommend 5.3.0).
+[BCP: How hard would it be to upgrade?]
 
@@
 This is a fair point.
-Our conjecture is based on anecdotal evidence; one of the authors — an experienced Haskell developer — manually applied some of Allegro’s optimizations to a few Haskell generators, and they found that they could really only manage to make performance worse, not better.
+Our conjecture is based on anecdotal evidence; one of the authors — an experienced Haskell developer — manually applied some of Allegro’s optimizations to a few Haskell generators, and they only managed to make performance worse, not better.
 This is nowhere near a proof, but it discouraged us from exploring that path in the short term.
@@
 This is a really interesting question, and we think that our work speaks 
-directly to part of it. Specifically, hand-coded fuzzers that construct 
+directly to part of it.[BCP: Blah.] Specifically, hand-coded fuzzers that construct 
 inputs satisfying specific structural and semantic constraints are performing 
@@
 
-We used the subset of the pre-existing Etna suite of benchmarks. Etna was
+We used the subset of the pre-existing Etna benchmark suite. Etna was
 designed exactly for this purpose: to test the bugfinding capabilities of
@@
 
-Yes, the AllegrOCaml and Base_quickcheck generors are as you imagine: essentially syntactically identical, modulo
-some extra syntactic cruft for staging. We plan to submit all of our code with the artifact evaluation, but for now, we
-include an example from our evals of a generator and its staged counterpart below the fold.
+Yes, the AllegrOCaml and Base_quickcheck generors are as you imagine:
+essentially identical syntactically, modulo
+some extra cruft for staging. We plan to submit all of our code with the artifact evaluation, but for now, we
+include an example from our evals of a generator and its staged counterpart below.
 
-By "semantically identical", we mean that given the same random seeds, the two
+By "semantically identical", we mean that, given the same random seeds, the two
 generators produce the same generated output.  This property is important
@@
 the staged generators happen to luckily create bug-squashing outputs. The
-sequence of outputs is the same between the staged and unstaged versions, the staged
+sequence of outputs is the same between the staged and unstaged versions; the staged
 generators just produce them faster.
 
-We further note that this is not an obvious or simple property to maintain:
+This is not an obvious or simple property to maintain:
 staged and unstaged programs can look syntactically similar but behave
@@
 prints once, the latter twice). Section 3.4 is essentially about ensuring that
-programmers need not worry about this, and can write their generators in a
+programmers need not worry about this and can write their generators in a
 manner that is syntactically as similar as possible to the non-staged versions.
 
+[BCP: Brief reminder what this is:]
 ```
@@
 
-This is a great idea, we'll do this with the extra space in the camera-ready version.
+This is a great idea: we'll do it with the extra space in the camera-ready version.
 

### 2025-12-02 paper/reviews/rebuttal_oopsla26.md

@@
 This is a fair point.
-Our conjecture is based on anecdotal evidence; one of the authors — an experienced Haskell developer — manually applied some of Allegro’s optimizations to a few Haskell generators, and they only managed to make performance worse, not better.
+Our conjecture is based on anecdotal evidence; one of the authors — an experienced Haskell developer — manually applied some of Allegro’s optimizations to a few Haskell generators, and they found that they could really only manage to make performance identical or worse.
 This is nowhere near a proof, but it discouraged us from exploring that path in the short term.
-We would be happy to mention this anecdotal experience and/or go into more detail about wanting to do this experiment as future work
+We would be happy to mention this anecdotal experience and/or go into more detail about wanting to do this experiment as future work.
 
@@
 
-We used the subset of the pre-existing Etna benchmark suite. Etna was
-designed exactly for this purpose: to test the bugfinding capabilities of
-competing PBT generators.  There are two benchmarks from the original Etna we
-did not use. One tests Red-Black trees, and the other tests a lambda calculus
-typed with System FSub, a more sophisticated type system.  We felt that the RBT eval was
-redundant with the BTS eval we already used: the generators are very similar.
-We chose not to use the FSub eval for two reasons. First, it had not been implemented
-in OCaml, and second, we felt this would be redundant with the STLC
-evals. The FSub generator is even more complex with more combinators than the STLC one, but the testing code and properties are essentially the same:
-we would see an exaggerated version of the speedups in the STLC eval.
+We used benchmarks from Etna, which was specifically designed to test the
+bugfinding capabilities of competing PBT generators.  There are two benchmarks
+from the original Etna we did not use. One tests Red-Black trees, and the other
+tests a lambda calculus typed with System FSub, a more sophisticated type
+system.  We felt that the RBT eval was redundant with the BTS eval we already
+used: the generators are very similar.  We chose not to use the FSub eval for
+two reasons. First, it had not been implemented in OCaml, and second, we felt
+this would be redundant with the STLC evals. The FSub generator is even more
+complex with more combinators than the STLC one, but the testing code and
+properties are essentially the same: we would see an exaggerated version of the
+speedups in the STLC eval.
 

### 2025-12-02 paper/reviews/rebuttal_oopsla26.md
attempt
prefold

--- 2025-12-02 paper/reviews/rebuttal_oopsla26.md
@@
 Thank you for your comments and questions! We start by discussing some common high-level concerns, then address all comments in order.
+
+Reviewr A about novelty
+
+Reviewrs A and C about generality and choice of our benchmarks.
+
+Reviewer A further about compilation time
+
+Reviewer B about fuzzing
+
+Reviewer C about sematnically identical
 
@@
 
-<!-- Additional novelty lies in Section 3.7, where we stage type-derived generators.
+Additional novelty lies in Section 3.7, where we stage type-derived generators.
 Unlike most staged libraries, which require users to understand metaprogramming,
@@
 at compile time, they can be staged without altering user experience. This is a
-rare example of staging "for free." -->
-[BCP: include this! Seems strong.]
+rare example of staging "for free."
+
+<!-- We say more after the fold After the fold -->
 <!-- 
@@
 issue for two reasons. First, our metaprogrammed generators compile extremely
-quickly: 33ms for Bool List up to 54ms for STLC (full table after the fold).
+quickly: a average of 33ms for Bool List up to 54ms for STLC, plus or minus 5ms for each (full table after the fold).
 Amortized across testing multiple properties (which usually take seconds each),
@@
 Testing and Parsing Randomness may provide paths forward in those domains as well.
+
+``we think so, making tenerators more performmant increases the chance they can be used in fuzzing workflows --- we say more belowt he fold''
 

### 2025-12-02 paper/reviews/rebuttal_oopsla26.md
wibble

--- 2025-12-02 paper/reviews/rebuttal_oopsla26.md
@@
-Thank you for your comments and questions! We start by discussing some common high-level concerns, then address all comments in order.
+Thank you for your comments and questions! We start by summarizing our responses to some high-level concerns, and then discuss all reviewer questions in detail after the fold.
+
+Reviewr A about novelty
+
+Reviewrs A and C about generality and choice of our benchmarks.
+
+Reviewer A further about compilation time
+
+Reviewer B about fuzzing
+
+Reviewer C about sematnically identical
 
@@
 OCaml simply because none of the authors are Scala experts: we are much more
-familiar with OCaml, so we chose to focus our implementation efforts there.
+familiar with OCaml, so we chose to focus our implementation efforts there. In
+the camera ready, we will make it clear earlier that we only implemented type-derived
+generators in AllegrOCaml.
 

### 2025-12-03 paper/reviews/rebuttal_oopsla26.md
lah

@@
 
-Reviewr A about novelty
+## Reviewer A asks us to clarify the novelty of our paper.
+The first novel contribution is our identification of previously unrecognized
+sources of inefficiency in PBT: DSL abstraction overhead and sampling costs.
+Prior work on speeding up PBT has never targeted these; even production-grade
+PBT libraries like Base_quickcheck retain monadic combinators and expensive
+randomness. Our second source of novelty is the application of staging to PBT
+generators.  While using staging to eliminate the abstraction overhead of DSLs
+is well-studied, it has never before been applied to PBT. Last, we provide a way
+to synthesize staged generators from type definitions (type-deriving).
+Type-derived generators are standard, but type-derived *staged* generators are a
+two-level metaprogram: based on the type, we generate code that generates a generator.
 
-Reviewrs A and C about generality and choice of our benchmarks.
+## Reviewers A and C ask about our benchmarks, how we chose them, and the balance between testing time versus generation time in practice
 
-Reviewer A further about compilation time
+Our evaluation uses Etna, the standard benchmark suite for comparing the
+bug-finding speed of PBT generators.  Etna includes generators from simple BST
+generators to sophisticated well-typed STLC term generators, paired with
+properties ranging from very cheap to check --- BST operations maintain the
+invariant --- to more expensive --- well typed STLC programs don't get stuck.
+In all of these cases, we find a consistent bug-finding speed improvement from
+Allegro.  Of course, real world uses of PBT vary even more wildly than those in
+Etna.  Characterizing what proportion of testing time real-world PBT workloads
+spend on generation is an interesting question that merits further study, though
+we view this as future work.  We did omit two Etna of the benchmarks: Red-Black
+trees --- which we thougth were redundant with our BST evaluation --- and System
+FSub --- redundant with STLC, and not implemented in OCaml.
 
-Reviewer B about fuzzing
+## Reviewer A asks about compilation time overhead.
+All the generators from the eval compile in under 55ms, which is amortized
+across all the tests, each of which takes on the order of seconds. Moreover,
+generators change very infrequently compared to application code, so
+recompilation is rare. A production-ready Allegro implementation could cache
+compiled generators to disk, avoiding compilation at test time entirely.
 
-Reviewer C about sematnically identical
+## Reviewer B asks if our work can be used to bridge the performance gap between PBT and fast hand-rolled fuzzers
+
+We think that our work speaks 
+directly to part of this question. Hand-coded fuzzers that construct 
+inputs satisfying specific structural and semantic constraints are performing 
+precisely the same task that PBT generators are designed for. Hand-coded fuzzers 
+might be faster than PBT generators if users choose to optimize them aggressively, 
+but ideally those optimizations should not need to be done manually. Our hope is 
+that Allegro (and other advances from the PBT literature) can make PBT generators 
+fast enough that developers could use them as highly-performant fuzzers, even 
+though they are expressed in a higher-level language that is easier to write 
+in and reason about. The question gets a bit less clear when considering fuzzers 
+that use mutation and other techniques to obtain interesting inputs, rather than 
+a hand-coded program, although ideas in papers like Coverage Guided Property-Based 
+Testing and Parsing Randomness may provide paths forward in those domains as well.
+
+## Reviewer C inquires if the staged generators are syntactically similar to their unstaged counterparts, and asks for clarification about the term "semantically identical"
+Indeed, the staged and unstaged generators are syntactically very similar,
+modulo some extra cruft required by MetaOCaml and Scala to write staged code. By
+"semantically identical," we mean that given the same random seeds, both produce
+the same outputs. This guarantees our speedups come purely from faster
+generation, not from lucky outputs that happen to find bugs sooner.
+
+------------------------------------------------------
 
@@
 As evidence, SOTA libraries like `Base_quickcheck` use imperative features (mutable arrays, stateful instead of functional RNG) to try to improve performance, yet retain design choices (monadic combinators, relatively expensive randomness) that introduce significant overhead.
-[BCP: Yes!]
-While our staging techniques are well established, their application to PBT has not been explored.
-This paper promotes staging as a tool for PBT developers by showing it can erase abstraction overhead, yielding substantial performance improvements.
-[BCP: ... but was it hard?]
+
+While staging techniques for optimizing DSLs are well established, their
+application to PBT is novel. Doing so was not easy, and posed some technical
+challenges --- Sections 3.5 and 3.6 describe how simply applying naive staging
+is not sufficient, and more sophisticated techniques must be used to make
+programming in a staged generator DSL ergonomic.
 
 Additional novelty lies in Section 3.7, where we stage type-derived generators.
-Unlike most staged libraries, which require users to understand metaprogramming,
-our approach is fully automatic. Since type-derived generators are constructed
-at compile time, they can be staged without altering user experience. This is a
-rare example of staging "for free."
-
-<!-- We say more after the fold After the fold -->
-<!-- 
-[JWC can we also say that this is an instance of two-level metaprogramming? It's a metaprogram that generates metaprograms!] -->
+Unlike most staged DSLs, which require users to understand metaprogramming,
+type-deriving a staged generator is fully automatic. Since type-derived
+generators are constructed at compile time, they can be staged without altering
+user experience. This is a rare example of staging "for free."
 
@@
 issue for two reasons. First, our metaprogrammed generators compile extremely
-quickly: a average of 33ms for Bool List up to 54ms for STLC, plus or minus 5ms for each (full table after the fold).
-Amortized across testing multiple properties (which usually take seconds each),
-this is negligible. Second, generators change infrequently compared to code
-being tested, and the generator only needs to be recompiled when it changes.
-Indeed, we anticipate that a production-ready instantiation of Allegro could let
+quickly: a average of 33ms for Bool List up to 54ms for STLC, plus or minus 5ms
+for each (a full table can be found at the bottom of our response).  Amortized
+across testing multiple properties (which usually take seconds each), this is
+negligible. Second, generators change infrequently compared to code being
+tested, and the generator only needs to be recompiled when it changes.  Indeed,
+we anticipate that a production-ready instantiation of Allegro could let
 programmers cache generator code to disk, to avoid compilation at testing-time
@@
 
-[BCP: IMO we can elide low-level comments and responses.]
-
 > l119: I'm curious whether OCaml 5 might benefit from using it's native
@@
 what we did (it has since been updated to recommend 5.3.0).
-[BCP: How hard would it be to upgrade?]
 
@@
 
+<!-- 
 > Can we can bridge the performance gap between bespoke
@@
 
-This is a really interesting question, and we think that our work speaks 
-directly to part of it.[BCP: Blah.] Specifically, hand-coded fuzzers that construct 
+We think that our work speaks 
+directly to part of this question. Hand-coded fuzzers that construct 
 inputs satisfying specific structural and semantic constraints are performing 
@@
 a hand-coded program, although ideas in papers like Coverage Guided Property-Based 
-Testing and Parsing Randomness may provide paths forward in those domains as well.
+Testing and Parsing Randomness may provide paths forward in those domains as well. -->
 
-``we think so, making tenerators more performmant increases the chance they can be used in fuzzing workflows --- we say more belowt he fold''
+<!-- ``we think so, making tenerators more performmant increases the chance they can be used in fuzzing workflows --- we say more belowt he fold'' -->
 
@@
 some extra cruft for staging. We plan to submit all of our code with the artifact evaluation, but for now, we
-include an example from our evals of a generator and its staged counterpart below.
+include an example from our evals of a generator and its staged counterpart at the bottom of this response.
 
@@
 
-[BCP: Brief reminder what this is:]
-```
-let rec gen ~(lo: int) ~(hi: int)  =
-  let%bind sz = size in
-  let should_stop = lo >= hi || sz <= 1  in
-  if should_stop
-    then return E
-  else
-    weighted_union [
-      (1., return E);
-      (float_of_int sz , (
-        let%bind k = int_inclusive ~lo ~hi in
-        let%bind v = Nat.quickcheck_generator_parameterized bst_bespoke_limits in
-        let%bind left = with_size ~size_c:(sz / 2) (gen ~lo:lo ~hi:(k - 1))  in
-        let%bind right = with_size ~size_c:(sz / 2) (gen ~lo:(k + 1) ~hi:hi) in
-        return (T (left, k, v, right))  
-      ))
-    ]
-
-let staged_quickcheck_generator (lo: int code) (hi: int code) : tree code G.t =
-  recursive (.< (.~lo, .~hi ) >.) 
-  (fun go lohi -> 
-    let%bind sz = size in
-    let%bind (lo, hi) = split_pair lohi in
-    let%bind should_stop = split_bool .< .~hi <= .~lo || .~sz <= 1 >. in
-    if should_stop
-      then return .< E >.
-    else
-      weighted_union [
-        (.< 1. >., return .< E >.);
-        ((G.C.i2f sz), (
-          let%bind k = int_inclusive ~lo ~hi in
-          let%bind v = Nat.staged_quickcheck_generator_sr_t (G.C.lift bst_bespoke_limits) in
-          let%bind left = with_size ~size_c:(G.C.div2 sz) (recurse go .<(.~lo, .~k - 1) >.) in
-          let%bind right = with_size ~size_c:(G.C.div2 sz) (recurse go .<(.~k + 1, .~hi) >.) in
-          return (.< T (.~left, .~k, .~v, .~right) >.)))
-      ]
-  )
-```
-
+`
 > 4. Why was BST (Repeated Insert) not used for the ScAllegro experiments in Figure 16?
@@
 
-----
+-------
 
@@
 STLC Bespoke: 53.699ms
+
+-------
+
+Unstaged and Staged Single-Pass BST Generator
+
+```
+let rec gen ~(lo: int) ~(hi: int)  =
+  let%bind sz = size in
+  let should_stop = lo >= hi || sz <= 1  in
+  if should_stop
+    then return E
+  else
+    weighted_union [
+      (1., return E);
+      (float_of_int sz , (
+        let%bind k = int_inclusive ~lo ~hi in
+        let%bind v = Nat.quickcheck_generator_parameterized 1000 in
+        let%bind left = with_size ~size_c:(sz / 2) (gen ~lo:lo ~hi:(k - 1))  in
+        let%bind right = with_size ~size_c:(sz / 2) (gen ~lo:(k + 1) ~hi:hi) in
+        return (T (left, k, v, right))  
+      ))
+    ]
+
+let gen (lo: int code) (hi: int code) : tree code G.t =
+  recursive (.< (.~lo, .~hi ) >.) 
+  (fun go lohi -> 
+    let%bind sz = size in
+    let%bind (lo, hi) = split_pair lohi in
+    let%bind should_stop = split_bool .< .~hi <= .~lo || .~sz <= 1 >. in
+    if should_stop
+      then return .< E >.
+    else
+      weighted_union [
+        (.< 1. >., return .< E >.);
+        ((G.C.i2f sz), (
+          let%bind k = int_inclusive ~lo ~hi in
+          let%bind v = Nat.staged_quickcheck_generator_sr_t (G.C.lift 1000) in
+          let%bind left = with_size ~size_c:(G.C.div2 sz) (recurse go .<(.~lo, .~k - 1) >.) in
+          let%bind right = with_size ~size_c:(G.C.div2 sz) (recurse go .<(.~k + 1, .~hi) >.) in
+          return (.< T (.~left, .~k, .~v, .~right) >.)))
+      ]
+  )
+```

### 2025-12-03 paper/reviews/rebuttal_oopsla26.md
re butt tal

@@
 
-## Reviewer A asks us to clarify the novelty of our paper.
+## Reviewer A asks us to clarify the novelty of our paper (Q1).
+
 The first novel contribution is our identification of previously unrecognized
@@
 PBT libraries like Base_quickcheck retain monadic combinators and expensive
-randomness. Our second source of novelty is the application of staging to PBT
+randomness.
+
+Our second source of novelty is the application of staging to PBT
 generators. While using staging to eliminate the abstraction overhead of DSLs
-is well-studied, it has never before been applied to PBT. Last, we provide a way
+is well-studied, it has never before been applied to PBT. 
+While staging techniques for optimizing DSLs are well established, their
+application to PBT is novel. Doing so was not easy, and posed some technical
+challenges --- Sections 3.5 and 3.6 describe how simply applying naive staging
+is not sufficient, and more sophisticated techniques must be used to make
+programming in a staged generator DSL ergonomic.
+
+Last, we provide a way
 to synthesize staged generators from type definitions (type-deriving).
@@
 Our evaluation uses Etna, the standard benchmark suite for comparing the
-bug-finding speed of PBT generators.  Etna includes generators from simple BST
-generators to sophisticated well-typed STLC term generators, paired with
-properties ranging from very cheap to check --- BST operations maintain the
-invariant --- to more expensive --- well typed STLC programs don't get stuck.
+bug-finding speed of PBT generators. Etna includes generators ranging from simple (BST)
+to sophisticated (well-typed STLC terms), and properties
+from very cheap to check (BST operations maintain the
+invariant) to more expensive (well typed STLC programs don't get stuck).
 In all of these cases, we find a consistent bug-finding speed improvement from
@@
 # Reviewer A
-
-> Q1) Can you clarify the novelty of your paper? Does it introduce any optimization from the multi-stage programming not known before?
-
-The main novelties are (1) identifying previously unrecognized sources of inefficiency in PBT, and (2) applying staging to address them.
-
-Prior to this work, neither DSL abstraction overhead nor sampling costs were viewed as optimization targets in PBT.
-As evidence, SOTA libraries like `Base_quickcheck` use imperative features (mutable arrays, stateful instead of functional RNG) to try to improve performance, yet retain design choices (monadic combinators, relatively expensive randomness) that introduce significant overhead.
-
-While staging techniques for optimizing DSLs are well established, their
-application to PBT is novel. Doing so was not easy, and posed some technical
-challenges --- Sections 3.5 and 3.6 describe how simply applying naive staging
-is not sufficient, and more sophisticated techniques must be used to make
-programming in a staged generator DSL ergonomic.
-
-Additional novelty lies in Section 3.7, where we stage type-derived generators.
-Unlike most staged DSLs, which require users to understand metaprogramming,
-type-deriving a staged generator is fully automatic. Since type-derived
-generators are constructed at compile time, they can be staged without altering
-user experience. This is a rare example of staging "for free."
 

### 2025-12-03 paper/reviews/rebuttal_oopsla26.md
wow

@@
 sources of inefficiency in PBT: DSL abstraction overhead and sampling costs.
-Prior work on speeding up PBT has never targeted these; even production-grade
-PBT libraries like Base_quickcheck retain monadic combinators and expensive
-randomness.
+Prior work on speeding up PBT has never targeted these. Indeed, even
+production-grade PBT libraries like Base_quickcheck are optimized for
+performance using imperative language features yet retain monadic combinators and
+expensive randomness.
 
@@
 application to PBT is novel. Doing so was not easy, and posed some technical
-challenges --- Sections 3.5 and 3.6 describe how simply applying naive staging
+challenges–Sections 3.5 and 3.6 describe how simply applying naive staging
 is not sufficient, and more sophisticated techniques must be used to make
@@
 to synthesize staged generators from type definitions (type-deriving).
-Type-derived generators are standard, but type-derived *staged* generators are a
-two-level metaprogram: based on the type, we generate code that generates a generator.
+Type-derived generators are standard in the PBT literature, but type-derived *staged* generators are a
+two-level metaprogram: based on a type, we generate code that generates a PBT generator.
 
@@
 we view this as future work.  We did omit two Etna of the benchmarks: Red-Black
-trees --- which we thougth were redundant with our BST evaluation --- and System
-FSub --- redundant with STLC, and not implemented in OCaml.
+trees–which we thougth were redundant with our BST evaluation–and System
+FSub–redundant with STLC, and not implemented in OCaml.
 
@@
 
-# Reviewer A
+# Reviewer A:
 
@@
 
-<!-- 
-> Can we can bridge the performance gap between bespoke
-highly-optimised fuzzing approaches and the much more generic PBT
-libraries exemplified by QuickCheck?
-
-We think that our work speaks 
-directly to part of this question. Hand-coded fuzzers that construct 
-inputs satisfying specific structural and semantic constraints are performing 
-precisely the same task that PBT generators are designed for. Hand-coded fuzzers 
-might be faster than PBT generators if users choose to optimize them aggressively, 
-but ideally those optimizations should not need to be done manually. Our hope is 
-that Allegro (and other advances from the PBT literature) can make PBT generators 
-fast enough that developers could use them as highly-performant fuzzers, even 
-though they are expressed in a higher-level language that is easier to write 
-in and reason about. The question gets a bit less clear when considering fuzzers 
-that use mutation and other techniques to obtain interesting inputs, rather than 
-a hand-coded program, although ideas in papers like Coverage Guided Property-Based 
-Testing and Parsing Randomness may provide paths forward in those domains as well. -->
-
-<!-- ``we think so, making tenerators more performmant increases the chance they can be used in fuzzing workflows --- we say more belowt he fold'' -->
-
-# Reviewer C
+# Reviewer C:
 
@@
 
-Yes, it's used in Figure 3 --- we can include a figure down at Section 3.6 with a small example in the camera-ready version.
+Yes, it's used in Figure 3–we can include a figure down at Section 3.6 with a small example in the camera-ready version.
 
@@
 
-Yes, the AllegrOCaml and Base_quickcheck generors are as you imagine:
+Yes, the AllegrOCaml and Base_quickcheck generators are as you imagine:
 essentially identical syntactically, modulo
@@
 non-equivalently, especially in the presence of effects (for example, `let x =
-print "hi" in x;x` and `let x = .<print "hi">. in .<$x;$x>. --- the former
+print "hi" in x;x` and `let x = .<print "hi">. in .<$x;$x>.–the former
 prints once, the latter twice). Section 3.4 is essentially about ensuring that
@@
 
-`
 > 4. Why was BST (Repeated Insert) not used for the ScAllegro experiments in Figure 16?
@@
 
-Generator Compilation Times
+# Generator Compilation Times
 
-Bool List: 33.183ms
-BST Repated Insert: 45.453ms
-BST Type: 46.051ms
-BST Single Pass: 50.147ms
-STLC Type: 44.663ms
-STLC Bespoke: 53.699ms
+| Generator | Compilation Time |
+|---|---|
+|Bool List|33.183ms|
+|BST Repated Insert|45.453ms|
+|BST Type|46.051ms|
+|BST Single Pass|50.147ms|
+|STLC Type|44.663ms|
+|STLC Bespoke|53.699ms|
 
@@
 
-Unstaged and Staged Single-Pass BST Generator
+# Unstaged and Staged Single-Pass BST Generator
 
+## Unstaged
 ```
@@
     ]
+```
 
+## Staged
+```
 let gen (lo: int code) (hi: int code) : tree code G.t =

### 2025-12-03 paper/reviews/rebuttal_oopsla26.md
spellcheck

@@
 we view this as future work.  We did omit two Etna of the benchmarks: Red-Black
-trees–which we thougth were redundant with our BST evaluation–and System
+trees–which we thought were redundant with our BST evaluation–and System
 FSub–redundant with STLC, and not implemented in OCaml.
@@
 
-> l989: You have conjectured that because GHC is set up to perform aggressive optimisations without staging then it is likely to benefit less from the kind of optimisations you exploit.
+> l989: You have conjectured that because GHC is set up to perform aggressive optimizations without staging then it is likely to benefit less from the kind of optimizations you exploit.
 It would be worth investigating to what extent this is indeed the case.
@@
 was not to point-by-point reproduce the AllegrOCaml evaluation in ScAllegro, but
-simply to demonstrate that the performance improvments of the Allegro technique
-are portable across languges.
+simply to demonstrate that the performance improvements of the Allegro technique
+are portable across languages.
 
@@
 |Bool List|33.183ms|
-|BST Repated Insert|45.453ms|
+|BST Repeated Insert|45.453ms|
 |BST Type|46.051ms|

### 2025-12-03 paper/reviews/rebuttal_oopsla26.md
spellcheck

@@
-Thank you for your comments and questions! We start by summarizing our responses to some high-level concerns, and then discuss all reviewer questions in detail after the fold.
+Thank you for your comments and questions! We begin by summarizing our responses to some high-level concerns, then discuss all reviewer questions in detail after the fold.
 
@@
 
-The first novel contribution is our identification of previously unrecognized
+Our first novel contribution is our identification of previously unrecognized
 sources of inefficiency in PBT: DSL abstraction overhead and sampling costs.
 Prior work on speeding up PBT has never targeted these. Indeed, even
-production-grade PBT libraries like Base_quickcheck are optimized for
-performance using imperative language features yet retain monadic combinators and
+production-grade PBT libraries like Base_quickcheck, which are optimized for
+performance using imperative language features, retain monadic combinators and
 expensive randomness.
@@
 challenges–Sections 3.5 and 3.6 describe how simply applying naive staging
-is not sufficient, and more sophisticated techniques must be used to make
+is not sufficient, and more sophisticated techniques must be employed to make
 programming in a staged generator DSL ergonomic.
 
-Last, we provide a way
-to synthesize staged generators from type definitions (type-deriving).
-Type-derived generators are standard in the PBT literature, but type-derived *staged* generators are a
-two-level metaprogram: based on a type, we generate code that generates a PBT generator.
+Last, we provide a way to synthesize staged generators from type definitions (type-deriving).
+Type-derived generators are standard in the PBT literature, but type-derived *staged* generators are a two-level metaprogram: based on a type, we generate code that generates a PBT generator.
 
@@
 Our evaluation uses Etna, the standard benchmark suite for comparing the
-bug-finding speed of PBT generators. Etna includes generators ranging from simple (BST)
-to sophisticated (well-typed STLC terms), and properties
-from very cheap to check (BST operations maintain the
-invariant) to more expensive (well typed STLC programs don't get stuck).
-In all of these cases, we find a consistent bug-finding speed improvement from
-Allegro.  Of course, real world uses of PBT vary even more wildly than those in
-Etna.  Characterizing what proportion of testing time real-world PBT workloads
-spend on generation is an interesting question that merits further study, though
-we view this as future work.  We did omit two Etna of the benchmarks: Red-Black
-trees–which we thought were redundant with our BST evaluation–and System
+bug-finding speed of PBT generators.
+Etna includes both simple generators (e.g., BST) and sophisticated ones (well-typed STLC terms), as well as properties that range from cheap to check (BST operations maintain the invariant) to expensive (well-typed STLC programs don't get stuck).
+In all of these cases, we find a consistent bug-finding speed improvement from Allegro.
+Of course, real world uses of PBT vary even more widely than those in Etna.
+Characterizing what proportion of testing time real-world PBT workloads spend on generation is an interesting question that merits further study, though we view this as future work.
+We did omit two Etna of the benchmarks: Red-Black trees–which we thought were redundant with our BST evaluation–and System
 FSub–redundant with STLC, and not implemented in OCaml.
@@
 ## Reviewer A asks about compilation time overhead.
-All the generators from the eval compile in under 55ms, which is amortized
-across all the tests, each of which takes on the order of seconds. Moreover,
+All the generators from the evaluation compile in under 55ms, which is amortized
+across all the tests, each of which take on the order of seconds. Moreover,
 generators change very infrequently compared to application code, so
@@
 
-As for compilation time of emitted generator code, we do not believe this is an
+As for the compilation time of emitted generator code, we do not believe this is an
 issue for two reasons. First, our metaprogrammed generators compile extremely
@@
 negligible. Second, generators change infrequently compared to code being
-tested, and the generator only needs to be recompiled when it changes.  Indeed,
-we anticipate that a production-ready instantiation of Allegro could let
-programmers cache generator code to disk, to avoid compilation at testing-time
+tested, and the generator only needs to be recompiled when it changes.
+We anticipate that a production-ready instantiation of Allegro could let
+programmers cache generator code to disk, avoiding compilation at testing-time
 altogether.
@@
 We see no fundamental barrier to implementing type-derived staged generators in
-Scala. Indeed, the scalacheck-derived project already uses Scala's reflection
+Scala. Indeed, the ScalaCheck-derived project already uses Scala's reflection
 features to type-derive standard (unstaged) ScalaCheck generators. Adapting this
@@
 
-We used benchmarks from Etna, which was specifically designed to test the
-bugfinding capabilities of competing PBT generators.  There are two benchmarks
+We used benchmarks from Etna, a platform specifically designed to test the
+bug-finding capabilities of competing PBT generators.  There are two benchmarks
 from the original Etna we did not use. One tests Red-Black trees, and the other
-tests a lambda calculus typed with System FSub, a more sophisticated type
+tests a lambda calculus typed with System FSub—a more sophisticated type
 system. We felt that the RBT eval was redundant with the BTS eval we already
-used: the generators are very similar.  We chose not to use the FSub eval for
-two reasons. First, it had not been implemented in OCaml, and second, we felt
-this would be redundant with the STLC evals. The FSub generator is even more
+used: the generators are very similar, and so were the results. We chose not to use the FSub eval for
+two reasons: first, it had not been implemented in OCaml; second, we felt
+it would be redundant with the STLC evals. The FSub generator is even more
 complex with more combinators than the STLC one, but the testing code and

### 2025-12-03 paper/reviews/rebuttal_oopsla26.md
Polishing and a BCP comment or two

@@
-Thank you for your comments and questions! We begin by summarizing our responses to some high-level concerns, then discuss all reviewer questions in detail after the fold.
+Thank you for your comments and questions! We begin with brief to some
+high-level concerns, then address all reviewer questions in detail
+"below the fold."
 
@@
 
-Our first novel contribution is our identification of previously unrecognized
-sources of inefficiency in PBT: DSL abstraction overhead and sampling costs.
-Prior work on speeding up PBT has never targeted these. Indeed, even
-production-grade PBT libraries like Base_quickcheck, which are optimized for
-performance using imperative language features, retain monadic combinators and
-expensive randomness.
+Our first novel contribution is the identification of previously
+unrecognized sources of inefficiency in PBT: DSL abstraction overhead
+and sampling costs.  Prior work on speeding up PBT has never targeted
+these. Indeed, even production-grade PBT libraries like
+Base_quickcheck, which are optimized for performance using imperative
+language features, retain monadic combinators and expensive
+randomness.
 
-Our second source of novelty is the application of staging to PBT
-generators. While using staging to eliminate the abstraction overhead of DSLs
-is well-studied, it has never before been applied to PBT. 
-While staging techniques for optimizing DSLs are well established, their
-application to PBT is novel. Doing so was not easy, and posed some technical
-challenges–Sections 3.5 and 3.6 describe how simply applying naive staging
-is not sufficient, and more sophisticated techniques must be employed to make
-programming in a staged generator DSL ergonomic.
+A second source of novelty is the application of staging to PBT
+generators. While using staging to eliminate the abstraction overhead
+of DSLs is well studied, it has never before been applied to PBT.
+While staging techniques for optimizing DSLs are well established,
+their application to PBT is novel. Indeed, doing so posed some
+nontrivial technical challenges. Sections 3.5 and 3.6 describe how
+simply applying naive staging is not sufficient: more sophisticated
+techniques must be employed to make programming in a staged generator
+DSL ergonomic.
 
-Last, we provide a way to synthesize staged generators from type definitions (type-deriving).
-Type-derived generators are standard in the PBT literature, but type-derived *staged* generators are a two-level metaprogram: based on a type, we generate code that generates a PBT generator.
+Finally, our method for synthesizing staged generators from type
+definitions deploys multi-stage programming in an unusually strong
+form. Type-derived generators are standard in the PBT literature, but
+type-derived *staged* generators are two-level metaprograms: from a
+type we generate code that generates a PBT generator.
 
@@
 
-Our evaluation uses Etna, the standard benchmark suite for comparing the
-bug-finding speed of PBT generators.
-Etna includes both simple generators (e.g., BST) and sophisticated ones (well-typed STLC terms), as well as properties that range from cheap to check (BST operations maintain the invariant) to expensive (well-typed STLC programs don't get stuck).
-In all of these cases, we find a consistent bug-finding speed improvement from Allegro.
-Of course, real world uses of PBT vary even more widely than those in Etna.
-Characterizing what proportion of testing time real-world PBT workloads spend on generation is an interesting question that merits further study, though we view this as future work.
-We did omit two Etna of the benchmarks: Red-Black trees–which we thought were redundant with our BST evaluation–and System
+Our evaluation uses Etna, a standard benchmark suite for comparing the
+bug-finding speed of PBT generators. Etna includes both simple
+generators (e.g., BST) and sophisticated ones (well-typed STLC terms),
+and properties that range from cheap-to-check (BST operations maintain
+the BST invariant) to expensive (well-typed STLC programs don't get
+stuck). In all of these cases, we find a consistent end-to-end
+bug-finding speed improvement from Allegro.
+
+Of course, real world uses of PBT will vary even more widely than
+those in Etna.  Characterizing what proportion of testing time is
+spent on test generation in real-world PBT workloads definitely merits
+further study, but we view this as future work.
+
+We did omit two Etna of the benchmarks: Red-Black trees–which we
+thought were redundant with our BST evaluation–and System
 FSub–redundant with STLC, and not implemented in OCaml.
 
-## Reviewer A asks about compilation time overhead.
-All the generators from the evaluation compile in under 55ms, which is amortized
-across all the tests, each of which take on the order of seconds. Moreover,
-generators change very infrequently compared to application code, so
-recompilation is rare. A production-ready Allegro implementation could cache
-compiled generators to disk, avoiding compilation at test time entirely.
+## Reviewer A asks about compilation-time overhead.
+
+All the generators from the evaluation compile in under 55ms, which is
+amortized across all the tests, each of which takes on the order of
+seconds. Moreover, generators change infrequently compared to
+application code, so a production Allegro implementation could cache
+compiled generators, avoiding compilation at test time almost
+entirely.
 
@@
 
-We think that our work speaks 
-directly to part of this question. Hand-coded fuzzers that construct 
-inputs satisfying specific structural and semantic constraints are performing 
-precisely the same task that PBT generators are designed for. Hand-coded fuzzers 
-might be faster than PBT generators if users choose to optimize them aggressively, 
-but ideally those optimizations should not need to be done manually. Our hope is 
-that Allegro (and other advances from the PBT literature) can make PBT generators 
-fast enough that developers could use them as highly-performant fuzzers, even 
-though they are expressed in a higher-level language that is easier to write 
-in and reason about. The question gets a bit less clear when considering fuzzers 
-that use mutation and other techniques to obtain interesting inputs, rather than 
-a hand-coded program, although ideas in papers like Coverage Guided Property-Based 
-Testing and Parsing Randomness may provide paths forward in those domains as well.
+[BCP: This part of the response still feels mushy, in part because I
+don't think we've really gotten clear in our minds what the real
+question is or what our real answer is.]
+
+Our work speaks to part of this question.
+
+Hand-coded fuzzers that construct inputs satisfying specific
+structural and semantic constraints are performing precisely the same
+task that PBT generators are designed for. Hand-coded fuzzers might be
+faster than PBT generators if users choose to optimize them
+aggressively, but ideally such optimizations should not need to be
+done manually. Our hope is that Allegro (and other advances from the
+PBT literature) can make PBT generators fast enough that developers
+could use them as highly-performant fuzzers, even though they are
+expressed in a higher-level language that is easier to write in and
+reason about.
+
+(The question gets a bit less clear when considering fuzzers that use
+mutation and other techniques to obtain interesting inputs, rather
+than a hand-coded program, although ideas in papers like Coverage
+Guided Property-Based Testing and Parsing Randomness may provide paths
+forward in those domains as well.)
 
 ## Reviewer C inquires if the staged generators are syntactically similar to their unstaged counterparts, and asks for clarification about the term "semantically identical"
-Indeed, the staged and unstaged generators are syntactically very similar,
-modulo some extra cruft required by MetaOCaml and Scala to write staged code. By
-"semantically identical," we mean that given the same random seeds, both produce
-the same outputs. This guarantees our speedups come purely from faster
-generation, not from lucky outputs that happen to find bugs sooner.
 
-------------------------------------------------------
+Indeed, the staged and unstaged generators are syntactically very
+similar, modulo some extra cruft required by MetaOCaml and Scala to
+write staged code.
+
+By "semantically identical" we mean that, given the same random seed,
+both produce the same outputs. This ensures that our speedups come
+purely from faster generation, not from lucky choices of random test
+generation that happen to find bugs sooner.
+
+-------------------------------------------------------------------------
 
@@
 
->  Q2) Can you elaborate on the end-to-end effectiveness of your proposed 
-solution?
+##  Q2) Can you elaborate on the end-to-end effectiveness of your proposed solution?
 
-Our evaluation uses Etna, the standard platform for 
-evaluating the bug-finding speed of PBT generators. It was designed 
-as a testbench for comparing generator strategies against each other, and, 
-to our knowledge, no more comprehensive benchmark suite exists for this 
-purpose. Our results show consistent speedups across Etna's tasks, with 
-generation time translating directly to faster bug-finding.
+Our evaluation uses Etna, a standard platform for evaluating the
+bug-finding speed of PBT generators. It was designed as a testbench
+for comparing generator strategies against each other, and, to our
+knowledge, no more comprehensive benchmark suite exists for this
+purpose. Our results show consistent speedups across Etna's tasks,
+with generation time translating directly to faster bug finding.
 
-We acknowledge that characterizing the proportion of testing time spent 
-on generation across real-world PBT workloads is an interesting question 
-that merits further study, and doing so would be an important research 
-contribution in itself; we view this as future work outside 
-the scope of this paper.
+Characterizing the proportion of testing time spent on generation
+across real-world PBT workloads is a question that definitely merits
+further study; we view this as future work outside the scope of this
+paper.
 
-As for the compilation time of emitted generator code, we do not believe this is an
-issue for two reasons. First, our metaprogrammed generators compile extremely
-quickly: a average of 33ms for Bool List up to 54ms for STLC, plus or minus 5ms
-for each (a full table can be found at the bottom of our response).  Amortized
-across testing multiple properties (which usually take seconds each), this is
-negligible. Second, generators change infrequently compared to code being
-tested, and the generator only needs to be recompiled when it changes.
-We anticipate that a production-ready instantiation of Allegro could let
-programmers cache generator code to disk, avoiding compilation at testing-time
-altogether.
+As for the compilation time of emitted generator code, we do not
+believe this is an issue for two reasons. First, our metaprogrammed
+generators compile very quickly: a average of 33ms for Bool List up to
+54ms for STLC, plus or minus 5ms for each (a full table can be found
+at the bottom of our response). Amortized across testing multiple
+properties (which usually take seconds each), this cost is
+negligible. Second, generators change infrequently compared to code
+being tested, and the generator really only needs to be recompiled
+when it changes. A production instantiation of Allegro could let
+programmers cache generator code, avoiding compilation at testing time
+almost completely.
 
-> Q3) What would it take to implement type-derived generators in Scala? 
-Why did you decide not to implement it?
+## Q3) What would it take to implement type-derived generators in Scala? Why did you decide not to implement it?
 
-We see no fundamental barrier to implementing type-derived staged generators in
-Scala. Indeed, the ScalaCheck-derived project already uses Scala's reflection
-features to type-derive standard (unstaged) ScalaCheck generators. Adapting this
-approach to emit ScAllegro would require engineering effort, but we anticipate
-no obstacles beyond that. We chose to implement our type-derived generators in
-OCaml simply because none of the authors are Scala experts: we are much more
-familiar with OCaml, so we chose to focus our implementation efforts there. In
-the camera ready, we will make it clear earlier that we only implemented type-derived
-generators in AllegrOCaml.
+We see no fundamental barrier to implementing type-derived staged
+generators in Scala. Indeed, the ScalaCheck-derived project already
+uses Scala's reflection features to type-derive standard (unstaged)
+ScalaCheck generators. Adapting this approach to emit ScAllegro would
+require engineering effort, but we anticipate no obstacles beyond
+that. We chose to implement our type-derived generators in OCaml
+simply because none of the authors are Scala experts: we are much more
+familiar with OCaml, so we chose to focus our implementation efforts
+there.
+
+In the camera ready, we will make it clear earlier that we only
+implemented type-derived generators in AllegrOCaml.
+
+-------------------------------------------------------------------------
 
@@
 
-> Typos and formatting issues on p4, l379, l475, l621, l653, l718, l824, l996
-
-Thank you! We'll fix these for the camera ready.
-
-> l119: I'm curious whether OCaml 5 might benefit from using it's native
+## l119: I'm curious whether OCaml 5 might benefit from using it's native
 effect handlers in place of a monadic generator DSL.
 
-Now we are also curious about this! Very interesting idea: direct-style generators
-could potentially be very performant (and more idiomatic).
+Now we are also curious about this!  It's a very interesting idea:
+direct-style generators could potentially be very performant and more
+idiomatic.
 
-> l639: Testing is all well and good, but have you not also considered
-trying to construct a more rigorous proof that AllegrOCaml is correct
-by program transformation / calculation?
+## l639: Testing is all well and good, but have you not also considered trying to construct a more rigorous proof that AllegrOCaml is correct by program transformation / calculation?
 
-We had not considered this, but it would be interesting to investigate.
+We had not considered this, but it would be interesting to try.
 
-> l751: Given that MetaOCaml is compatible with OCaml 5, why are you
-still using OCaml 4?
+## l751: Given that MetaOCaml is compatible with OCaml 5, why are you still using OCaml 4?
 
@@
 
-> l824: The text reads a bit oddly to me at this point, as you are
-talking about an experiment you performed in the past and yet you
-insist on using the present tense.
-
-Thanks for the feedback, we'll workshop this prose to make it less awkward.
-
-> l989: You have conjectured that because GHC is set up to perform aggressive optimizations without staging then it is likely to benefit less from the kind of optimizations you exploit.
+## l989: You have conjectured that because GHC is set up to perform aggressive optimizations without staging then it is likely to benefit less from the kind of optimizations you exploit.
 It would be worth investigating to what extent this is indeed the case.
@@
 
-This is a fair point.
-Our conjecture is based on anecdotal evidence; one of the authors — an experienced Haskell developer — manually applied some of Allegro’s optimizations to a few Haskell generators, and they found that they could really only manage to make performance identical or worse.
-This is nowhere near a proof, but it discouraged us from exploring that path in the short term.
-We would be happy to mention this anecdotal experience and/or go into more detail about wanting to do this experiment as future work.
+This is a fair point. Our conjecture is based on anecdotal evidence;
+one of the authors — an experienced Haskell developer — manually
+applied some of Allegro’s optimizations to a few Haskell generators,
+and they found that they could really only manage to make performance
+identical or worse. This is nowhere near a proof, but it discouraged
+us from exploring that path in the short term. We would be happy to
+mention this anecdotal experience and/or go into more detail about
+wanting to do this experiment as future work.
+
+-------------------------------------------------------------------------
 
@@
 
-> 1. Can you provide an example of how the recursive generator API in Section 3.6 is used?
+## 1. Can you provide an example of how the recursive generator API in Section 3.6 is used?
 
-Yes, it's used in Figure 3–we can include a figure down at Section 3.6 with a small example in the camera-ready version.
+Yes, it's used in Figure 3. We can include a figure down at Section
+3.6 with a small example in the next version.
 
-> 2. How were the experimental benchmarks chosen?
+## 2. How were the experimental benchmarks chosen?
 
-We used benchmarks from Etna, a platform specifically designed to test the
-bug-finding capabilities of competing PBT generators.  There are two benchmarks
-from the original Etna we did not use. One tests Red-Black trees, and the other
-tests a lambda calculus typed with System FSub—a more sophisticated type
-system. We felt that the RBT eval was redundant with the BTS eval we already
-used: the generators are very similar, and so were the results. We chose not to use the FSub eval for
-two reasons: first, it had not been implemented in OCaml; second, we felt
-it would be redundant with the STLC evals. The FSub generator is even more
-complex with more combinators than the STLC one, but the testing code and
-properties are essentially the same: we would see an exaggerated version of the
-speedups in the STLC eval.
+We used benchmarks from Etna, a platform specifically designed to test
+the bug-finding capabilities of competing PBT generators.  There are
+two benchmarks from the original Etna we did not use. One tests
+Red-Black trees, and the other tests a lambda calculus typed with
+System FSub—a more sophisticated type system. We felt that the RBT
+eval was redundant with the BTS eval we already used: the generators
+are very similar, and so were the results. We chose not to use the
+FSub eval for two reasons: first, it had not been implemented in
+OCaml; second, we felt it would be redundant with the STLC evals. The
+FSub generator is even more complex with more combinators than the
+STLC one, but the testing code and properties are essentially the
+same: we would see an exaggerated version of the speedups in the STLC
+eval.
 
-> 3. What do you mean by "semantically identical", and are the generators also syntactically similar?
+## 3. What do you mean by "semantically identical", and are the generators also syntactically similar?
 
-Yes, the AllegrOCaml and Base_quickcheck generators are as you imagine:
-essentially identical syntactically, modulo
-some extra cruft for staging. We plan to submit all of our code with the artifact evaluation, but for now, we
-include an example from our evals of a generator and its staged counterpart at the bottom of this response.
+Yes, the AllegrOCaml and Base_quickcheck generators are as you
+imagine: essentially identical syntactically, modulo some extra cruft
+for staging. We plan to submit all of our code with the artifact
+evaluation, but, for now, we include an example from our evals of a
+generator and its staged counterpart at the bottom of this response.
 
-By "semantically identical", we mean that, given the same random seeds, the two
-generators produce the same generated output.  This property is important
-because it guarantees that all of our bug-finding speedups are solely
-attributable to the improved performance of staged generators, and not because
-the staged generators happen to luckily create bug-squashing outputs. The
-sequence of outputs is the same between the staged and unstaged versions; the staged
-generators just produce them faster.
+By "semantically identical", we mean that, given the same random
+seeds, the two generators produce the same generated output.  This
+property is important because it guarantees that all of our
+bug-finding speedups are solely attributable to the improved
+performance of staged generators, and not because the staged
+generators happen to luckily create bug-squashing outputs. The
+sequence of outputs is the same between the staged and unstaged
+versions; the staged generators just produce them faster.
 
-This is not an obvious or simple property to maintain:
-staged and unstaged programs can look syntactically similar but behave
-non-equivalently, especially in the presence of effects (for example, `let x =
-print "hi" in x;x` and `let x = .<print "hi">. in .<$x;$x>.–the former
-prints once, the latter twice). Section 3.4 is essentially about ensuring that
-programmers need not worry about this and can write their generators in a
-manner that is syntactically as similar as possible to the non-staged versions.
+This is not an obvious property or an easy one to maintain: staged and
+unstaged programs can look syntactically similar but behave
+non-equivalently, especially in the presence of effects (for example,
+`let x = print "hi" in x;x` and `let x = .<print "hi">. in
+.<$x;$x>. -- the former prints once, the latter twice). Section 3.4 is
+essentially about ensuring that programmers need not worry about this
+and can write their generators in a manner that is syntactically as
+similar as possible to the non-staged versions.
 
-> 4. Why was BST (Repeated Insert) not used for the ScAllegro experiments in Figure 16?
+## 4. Why was BST (Repeated Insert) not used for the ScAllegro experiments in Figure 16?
 
-We picked a representative from each category of generators: trees, lists, and
-STLC terms, representing low, medium, and high numbers of binds, respectively.
-We anticipate BST (Repeated Insert) would see a slightly smaller than the other
-generators speedup due to its lower number of binds, and we would be happy to
-include this information in a revision. Our goal in this evaluation subsection
-was not to point-by-point reproduce the AllegrOCaml evaluation in ScAllegro, but
-simply to demonstrate that the performance improvements of the Allegro technique
-are portable across languages.
+We picked a representative from each category of generators: trees,
+lists, and STLC terms, representing low, medium, and high numbers of
+binds, respectively.  We anticipate BST (Repeated Insert) would see a
+slightly smaller than the other generators speedup due to its lower
+number of binds, and we would be happy to include this information in
+a revision. Our goal in this evaluation subsection was not to
+point-by-point reproduce the AllegrOCaml evaluation in ScAllegro, but
+simply to demonstrate that the performance improvements of the Allegro
+technique are portable across languages.
 
-> I found the repeated shifting of focus between OCaml and Scala to be distracting.
+## I found the repeated shifting of focus between OCaml and Scala to be distracting.
 An alternative approach might have been to describe the entirety of the approach for OCaml, and then summarize the Scala-specific differences.
 
-Thanks for this feedback. We'll try to further isolate the discussion of Scala to just the intro and the bits of the evaluation where we evaluate ScAllegro.
+Thanks for this feedback. We'll try to further isolate the discussion
+of Scala to just the intro and the bits of the evaluation where we
+evaluate ScAllegro.
 
-> On a related note, the paper might be easier to read if there was a single table / figure that summarized the types of various values and objects used throughout the paper.
+## On a related note, the paper might be easier to read if there was a single table / figure that summarized the types of various values and objects used throughout the paper.
 I found myself making this list as I read the paper.
 
-This is a great idea: we'll do it with the extra space in the camera-ready version.
+This is a great idea and a good use of the extra space in a
+camera-ready version.
 
@@
 # Generator Compilation Times
+
+[BCP: Format for easy reading in ascii too?]
 
@@
 
-# Unstaged and Staged Single-Pass BST Generator
+# Unstaged and Staged Single-Pass BST Generators
 

### 2025-12-03 paper/reviews/rebuttal_oopsla26.md
format table

@@
 
-[BCP: Format for easy reading in ascii too?]
-
 |     Generator     | Compilation Time |
-|---|---|
-|Bool List|33.183ms|
-|BST Repeated Insert|45.453ms|
-|BST Type|46.051ms|
-|BST Single Pass|50.147ms|
-|STLC Type|44.663ms|
-|STLC Bespoke|53.699ms|
+|-------------------|------------------|
+|Bool List          |33.183ms          |
+|BST Repeated Insert|45.453ms          |
+|BST Type           |46.051ms          |
+|BST Single Pass    |50.147ms          |
+|STLC Type          |44.663ms          |
+|STLC Bespoke       |53.699ms          |
 

### 2025-12-04 paper/reviews/rebuttal_oopsla26.md
Reviewer B - try again

@@
-Thank you for your comments and questions! We begin with brief to some
+Thank you for the comments and questions! We begin with brief to some
 high-level concerns, then address all reviewer questions in detail
@@
 
-[BCP: This part of the response still feels mushy, in part because I
-don't think we've really gotten clear in our minds what the real
-question is or what our real answer is.]
+We're not 100% sure we understand the phrasing of the question, but we
+think the reviewer is probably asking about the gap between PBT
+generators, which are generally expressed in a high-level, often
+monadic domain-specific language, and hand-coded fuzzers that are
+written in a lower-level language and more heavily optimized for a
+particular use-case. If so, then our work represents (we believe) a
+significant step in closing that gap. Our hope is that Allegro can
+make monadic PBT generators fast enough that developers could use them
+as highly-performant fuzzers, even though they are expressed in a
+higher-level language that is easier to write in and reason about.
 
-Our work speaks to part of this question.
+Alternatively, the reviewer may also be asking about the gap between
+standard PBT generators and fuzzers that use mutation, coverage
+feedback, and other techniques to obtain interesting inputs. Allegro
+doesn't directly impact these kinds of approaches, although ideas in
+papers like _Coverage Guided Property-Based Testing_ and _Parsing
+Randomness_ may provide paths forward in those domains as well.
 
-Hand-coded fuzzers that construct inputs satisfying specific
-structural and semantic constraints are performing precisely the same
-task that PBT generators are designed for. Hand-coded fuzzers might be
-faster than PBT generators if users choose to optimize them
-aggressively, but ideally such optimizations should not need to be
-done manually. Our hope is that Allegro (and other advances from the
-PBT literature) can make PBT generators fast enough that developers
-could use them as highly-performant fuzzers, even though they are
-expressed in a higher-level language that is easier to write in and
-reason about.
-
-(The question gets a bit less clear when considering fuzzers that use
-mutation and other techniques to obtain interesting inputs, rather
-than a hand-coded program, although ideas in papers like Coverage
-Guided Property-Based Testing and Parsing Randomness may provide paths
-forward in those domains as well.)
+Or (covering all the bases we can think of :-) the reviewer might be
+wondering about the gap between the "cheap but generic" generators
+that Haskell QuickCheck can synthesize automatically from the types of
+the data being generated and the hand-build generators that
+sophisticated users write to achieve maximum testing
+effectiveness. This gap can be arbitrarily large, so it probably can't
+pbe closed in general, but there has been a ton of work already on
+_generating good generators_ automatically from formal descriptions of
+test data satisfying given preconditions, and we believe there is much
+more to do in this direction.
 
@@
 
-## l119: I'm curious whether OCaml 5 might benefit from using it's native
-effect handlers in place of a monadic generator DSL.
+## l119: I'm curious whether OCaml 5 might benefit from using it's native effect handlers in place of a monadic generator DSL.
 

### 2025-12-04 paper/reviews/rebuttal_oopsla26.md
move much of rb response below the fold

@@
 Thank you for the comments and questions! We begin with brief to some
-high-level concerns, then address all reviewer questions in detail
+high-level concerns, then address all questions in detail
 "below the fold."
@@
 A second source of novelty is the application of staging to PBT
-generators. While using staging to eliminate the abstraction overhead
-of DSLs is well studied, it has never before been applied to PBT.
-While staging techniques for optimizing DSLs are well established,
-their application to PBT is novel. Indeed, doing so posed some
-nontrivial technical challenges. Sections 3.5 and 3.6 describe how
+generators. While using staging to eliminate abstraction overhead
+of DSLs is well studied, it has never been applied to PBT.
+Indeed, doing so posed nontrivial technical challenges. Sections 3.5 and 3.6 describe how
 simply applying naive staging is not sufficient: more sophisticated
@@
 
-Finally, our method for synthesizing staged generators from type
-definitions deploys multi-stage programming in an unusually strong
-form. Type-derived generators are standard in the PBT literature, but
-type-derived *staged* generators are two-level metaprograms: from a
-type we generate code that generates a PBT generator.
+Finally, our method for synthesizing staged generators from type definitions is
+novel and technically interesting. Type-derived generators are standard in the
+PBT literature, but type-derived *staged* generators are two-level metaprograms:
+from a type we generate code that generates a PBT generator.
 
@@
 
-Alternatively, the reviewer may also be asking about the gap between
-standard PBT generators and fuzzers that use mutation, coverage
-feedback, and other techniques to obtain interesting inputs. Allegro
-doesn't directly impact these kinds of approaches, although ideas in
-papers like _Coverage Guided Property-Based Testing_ and _Parsing
-Randomness_ may provide paths forward in those domains as well.
-
-Or (covering all the bases we can think of :-) the reviewer might be
-wondering about the gap between the "cheap but generic" generators
-that Haskell QuickCheck can synthesize automatically from the types of
-the data being generated and the hand-build generators that
-sophisticated users write to achieve maximum testing
-effectiveness. This gap can be arbitrarily large, so it probably can't
-pbe closed in general, but there has been a ton of work already on
-_generating good generators_ automatically from formal descriptions of
-test data satisfying given preconditions, and we believe there is much
-more to do in this direction.
+In case the reviewer had another question in mind, we answer some alternate
+interpretations below the fold.
 
@@
 
-Indeed, the staged and unstaged generators are syntactically very
-similar, modulo some extra cruft required by MetaOCaml and Scala to
-write staged code.
+Indeed, the staged and unstaged generators are syntactically very similar,
+modulo some extra cruft required by MetaOCaml and Scala to write staged code ---
+we give a full example of a generator in its staged and unstaged forms at the
+bottom of this response.
 
@@
 wanting to do this experiment as future work.
+
+## ...
+to what extent we can bridge the performance gap between bespoke highly-optimised fuzzing approaches and the much more generic PBT libraries exemplified by QuickCheck.
+Can we hope to build fuzzers on top of generic PBT libraries which perform as well as the hand-coded fuzzers used in practice?
+Does your work shed any light on how to tackle this question?
+
+We believe the reviewer may also be asking about the gap between
+standard PBT generators and fuzzers that use mutation, coverage
+feedback, and other techniques to obtain interesting inputs. Allegro
+doesn't directly impact these kinds of approaches, although ideas in
+papers like _Coverage Guided Property-Based Testing_ and _Parsing
+Randomness_ may provide paths forward in those domains as well.
+
+Or (covering all the bases we can think of :-)) the reviewer might be
+wondering about the gap between the "cheap but generic" generators
+that Haskell QuickCheck can synthesize automatically from the types of
+the data being generated and the hand-build generators that
+sophisticated users write to achieve maximum testing
+effectiveness. This gap can be arbitrarily large, so it probably can't
+be closed in general, but there has been a ton of work already on
+_generating good generators_ automatically from formal descriptions of
+test data satisfying given preconditions, and we believe there is much
+more to do in this direction.
 
